---
title: "Examine estrogen and androgen receptor gene expression"
author: "Emma Jones"
date: "2025-06-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The purpose of this script is to examine estrogen and andorgen expressing cells


First, we load packages. We might want more later, but for now I know I need the following

```{r load packages}
library(tidyverse)
library(Seurat)
library(readxl)
library(Signac)
library(glue)
library(cowplot)
library(edgeR)
library(limma)
library(patchwork)
```


## Load data

```{r load seurat object and metabolomics excel spreadsheet}
# load 10x multiome data
SSD_data_v5 <- readRDS("../../data/SSD_data_v5.rds")
# remove doublets
SSD_data_v5 <- subset(SSD_data_v5, subset = Droplet == "SNG")

# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx("../../doc/targeted/Targeted Comparison.xlsx")
```

## Examine esrogen and androgen recpetor expression per cell type

```{r make violin plots}

DefaultAssay(SSD_data_v5) <- "SCT"

VlnPlot(SSD_data_v5, "AR")

VlnPlot(SSD_data_v5, "ESR1")

VlnPlot(SSD_data_v5, "ESR2")

VlnPlot(SSD_data_v5, "PGR")

FeaturePlot(SSD_data_v5, "AR")
FeaturePlot(SSD_data_v5, "ESR1")
FeaturePlot(SSD_data_v5, "ESR2") # most highly expressed
FeaturePlot(SSD_data_v5, "PGR")

DimPlot(SSD_data_v5)

table(SSD_data_v5$cluster_call)

```


```{r get esr2 cells as metadata}

# Identify ESR2-expressing cells
esr2_cells <- WhichCells(SSD_data_v5, expression = ESR2 != 0)

# Add metadata column for labeling
SSD_data_v5 <- AddMetaData(object = SSD_data_v5, metadata = ifelse(colnames(SSD_data_v5) %in% esr2_cells, "ESR2_positive", "ESR2_negative"), col.name = "ESR2_expression_status")

# Figure out which cell type have the most ESR2
SSD_data_v5$ESR2_positive_SCT <- SSD_data_v5@assays$SCT@data["ESR2",] > 0

percentage_ESR2_by_cell_type_SCT <- SSD_data_v5@meta.data %>%
    group_by(cluster_call) %>%
    summarize(percent_ESR2_positive = mean(ESR2_positive_SCT) * 100)

# Visualize the data
ggplot(percentage_ESR2_by_cell_type_SCT, aes(x = cluster_call, y = percent_ESR2_positive, fill = percent_ESR2_positive)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(title = "Percentage of ESR2-expressing Cells (any) by Cell Type",
         x = "Cell Type",
         y = "Percentage of ESR2-expressing Cells")

# try for "high"
SSD_data_v5$ESR2_positive_SCT_high <- SSD_data_v5@assays$SCT@data["ESR2",] > 1.5

percentage_ESR2_by_cell_type_SCT_high <- SSD_data_v5@meta.data %>%
    group_by(cluster_call) %>%
    summarize(percent_ESR2_positive = mean(ESR2_positive_SCT_high) * 100)

# Visualize the data
ggplot(percentage_ESR2_by_cell_type_SCT_high, aes(x = cluster_call, y = percent_ESR2_positive, fill = percent_ESR2_positive)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(title = "Percentage of ESR2-expressing Cells (high) by Cell Type",
         x = "Cell Type",
         y = "Percentage of ESR2-expressing Cells")

# try for "med"
SSD_data_v5$ESR2_positive_SCT_med <- SSD_data_v5@assays$SCT@data["ESR2",] > 1

percentage_ESR2_by_cell_type_SCT_med <- SSD_data_v5@meta.data %>%
    group_by(cluster_call) %>%
    summarize(percent_ESR2_positive = mean(ESR2_positive_SCT_med) * 100)

# Visualize the data
ggplot(percentage_ESR2_by_cell_type_SCT_med, aes(x = cluster_call, y = percent_ESR2_positive, fill = percent_ESR2_positive)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(title = "Percentage of ESR2-expressing Cells (med) by Cell Type",
         x = "Cell Type",
         y = "Percentage of ESR2-expressing Cells")

```

# rerun limma on only esr2-expressing cells

```{r filter data}
esr2_data <- subset(SSD_data_v5, ESR2_expression_status == "ESR2_positive")
```


## Remove Outliers
```{r remove outliers}
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) + 1.5 * IQR(data[[variable]], na.rm = TRUE)
  data %>% filter(.data[[variable]] <= outlier_limit)
}

targeted_comparision_filtered_E2 <- detect_outliers(targeted_comparision_raw, "17B-E2")
```

## Metadata and Pseudobulk Functions

```{r metadata and pseudobulk}
prepare_metadata_for_analysis <- function(pb_metadata, targeted_comparision_filtered) {
  targeted_comparision_filtered$Sample <- paste0("tissue", targeted_comparision_filtered$Sample)
  pb_metadata %>%
    as.data.frame() %>%
    rownames_to_column("pseudobulk_id") %>%
    left_join(targeted_comparision_filtered %>% select(Sample, Sex, TT, `17B-E2`, P4, GW),
              by = c("sample_id" = "Sample")) %>%
    filter(!is.na(Sex), cell_count >= 5) %>%
    mutate(group = paste(cell_type, Sex, sep = "_")) %>%
    column_to_rownames("pseudobulk_id")
}

pseudobulk_data <- function(seurat_obj, assay = "RNA", layer = "counts", sample_col = "sample_id", celltype_col = "cell_type") {
  counts <- GetAssayData(seurat_obj, assay = assay, slot = layer)
  meta <- seurat_obj@meta.data
  meta$cell_type <- meta$cluster_call
  group_ids <- interaction(meta[[sample_col]], meta[[celltype_col]], drop = TRUE)
  group_levels <- levels(group_ids)
  aggregated_list <- list()
  metadata_list <- list()
  
  for (gid in group_levels) {
  group_cells <- which(group_ids == gid)
  if (length(group_cells) < 1) next
  group_counts <- Matrix::rowSums(counts[, group_cells, drop = FALSE])
  aggregated_list[[gid]] <- group_counts
    metadata_list[[gid]] <- data.frame(
      sample_id = meta[[sample_col]][group_cells[1]],
      cell_type = meta[[celltype_col]][group_cells[1]],
      cell_count = length(group_cells),
      row.names = gid
    )
  }
  
  pseudobulk_mat <- do.call(rbind, aggregated_list)
  metadata_df <- do.call(rbind, metadata_list)
  
  pseudobulk_mat_log <- log1p(as.matrix(pseudobulk_mat))
  
  list(counts = pseudobulk_mat, logcounts = pseudobulk_mat_log, metadata = metadata_df)
}
```

## Generate and Filter Pseudobulk Data


```{r generate and filter pseudobulk data}

generate_and_filter_pseudobulk <- function(seurat_obj, targeted_filtered) {
  seurat_obj$cell_type <- seurat_obj$cluster_call  # ensure column exists
  seurat_obj$sample_id <- seurat_obj$sample        # ensure column exists
  
  pb <- pseudobulk_data(seurat_obj)
  pb_metadata <- prepare_metadata_for_analysis(pb$metadata, targeted_filtered)
  
  common_ids <- intersect(rownames(pb$logcounts), rownames(pb_metadata))
  pb$logcounts <- pb$logcounts[common_ids, , drop = FALSE]
  pb$counts <- pb$counts[common_ids, , drop = FALSE]
  pb$metadata <- pb_metadata[common_ids, , drop = FALSE]
  
  list(pb = pb, metadata = pb_metadata)
}

pb_E2 <- generate_and_filter_pseudobulk(esr2_data, targeted_comparision_filtered_E2)
```

## Limma analysis

```{r limma anlysis}
run_voom_lm <- function(counts, metadata, predictor) {
  stopifnot(ncol(counts) == nrow(metadata))
  if (!"GW" %in% colnames(metadata)) metadata$GW <- NA
  design <- model.matrix(reformulate(c(predictor, "cell_count", "GW")), data = metadata)
  v <- limma::voom(counts, design)
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)
  limma::topTable(fit, coef = predictor, number = Inf, sort.by = "none") %>%
    rownames_to_column("gene")
}

analyze_hormone <- function(pb_counts, metadata_merged, hormone, output_dir) {
  split_ids <- split(rownames(metadata_merged), interaction(metadata_merged$cell_type, metadata_merged$Sex, drop = TRUE))
  
  results_by_group <- purrr::map(split_ids, function(ids) {
    counts_sub <- t(pb_counts)[, ids, drop = FALSE] 
    metadata_sub <- metadata_merged[ids, , drop = FALSE]
    
    if (nrow(metadata_sub) < 5) return(NULL)
    keep_genes <- colSums(counts_sub >= 2) >= 3
    counts_sub <- counts_sub[, keep_genes, drop = FALSE]
    if (ncol(counts_sub) == 0) return(NULL)
    
    result <- run_voom_lm(counts_sub, metadata_sub, hormone)
    result$cell_type <- metadata_sub$cell_type[1]
    result$sex <- metadata_sub$Sex[1]
    result
  })
  
  bind_rows(results_by_group[!sapply(results_by_group, is.null)])
}
```

# Run analyses

```{r merge metadata and run analyses}
# E2
results_all_E2 <- analyze_hormone(pb_E2$pb$counts, pb_E2$metadata, "`17B-E2`", "pseudobulk_plots_estrogen")
```

## Plot Top Genes

```{r plot top genes}
plot_top_genes <- function(results, metadata_merged, log_cpm_mat, output_dir, hormone) {
  cell_types <- unique(results$cell_type)
  dir.create(output_dir, showWarnings = FALSE)
  
  walk(cell_types, function(group_of_interest) {
    results_subset <- results %>%
      filter(cell_type == group_of_interest) %>%
      arrange(-abs(t)) %>%
      slice_head(n = 6)
    
    samples_in_group <- metadata_merged %>% filter(cell_type == group_of_interest)
    counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]
    
    plot_list <- map(results_subset$gene, function(gene) {
      if (!gene %in% colnames(counts_in_group)) return(NULL)
      expr_vals <- counts_in_group[, gene]
      if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
      
      df <- samples_in_group %>% mutate(expr = expr_vals)
      
      ggplot(df, aes_string(x = hormone, y = "expr", color = "Sex", fill = "Sex")) +
        geom_point(size = 2, alpha = 0.8) +
        geom_smooth(method = "lm", se = TRUE) +
        scale_color_manual(values = c("M" = "steelblue", "F" = "violetred3")) +
        scale_fill_manual(values = c("M" = "steelblue", "F" = "violetred3")) +
        labs(title = paste0(gene, " in ", group_of_interest),
             x = hormone, y = "log1p(CPM)") +
        theme_minimal(base_size = 14)
    }) %>% compact()
    
    if (length(plot_list) == 0) return()
    
    combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
    clean_name <- gsub("/", "-", group_of_interest)
    ggsave(filename = file.path(output_dir, paste0(clean_name, "_top_genes.png")),
           plot = combined_plot, width = 10, height = 6 + ceiling(length(plot_list)/2), dpi = 300)
  })
}

plot_top_genes(results_all_E2, pb_E2$metadata, pb_E2$pb$logcounts, "../../results/targeted/pseudobulk_plots_esr2_pos_estrogen", "`17B-E2`")
```

## Summarize and save results

```{r summarize and save results}

summarize_sig_genes <- function(results, fdr_cutoff = 0.05, logfc_cutoff = 0.25) {
  sig <- results %>% filter(adj.P.Val < fdr_cutoff)
  if (!is.null(logfc_cutoff)) {
    sig <- sig %>% filter(abs(logFC) >= logfc_cutoff)
  }
  sig %>%
    count(cell_type, sex, name = "num_sig_genes") %>%
    arrange(desc(num_sig_genes))
}

plot_sig_gene_summary <- function(summary_df, title) {
  ggplot(summary_df, aes(x = cell_type, y = num_sig_genes, fill = sex)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(title = title, x = "Cell Type", y = "# Significant Genes") +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("M" = "steelblue", "F" = "violetred3"))
}

# Summarize
summary_E2 <- summarize_sig_genes(results_all_E2)

# Save CSVs
write_csv(summary_E2, "../../results/targeted/summary_esr2_pos_E2_sig_genes.csv")

# Plot and save
ggsave("../../results/targeted/esr2_pos_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(summary_E2, "Estradiol — ESR2-positive cells"), width = 8, height = 5)

```

