---
title: "SFARI Overlap with TT signficiant DEGs"
author: "Emma Jones"
date: "2025-06-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of his script is to determine if any SFARI genes (https://gene.sfari.org/database/human-gene/) are included in the significant TT genes for males. I also comapres out TT-response gene to known gene lists of TT response genes. The papers for the following gene lists are as folows:

human brain organoid androgen response paper: https://www.nature.com/articles/s41586-021-04330-4
sex-specific mouse Embryonic Neural Stem/Progenitor Cells response paper: https://www.nature.com/articles/srep36916 
single cell, 17 tissue androgen mouse paper: https://www.nature.com/articles/s41586-024-07291-6 

##Load Packages

```{r load package}
library(tidyverse)
library(Seurat)
library(readxl)
library(Signac)
library(glue)
library(cowplot)
library(edgeR)
library(limma)
library(patchwork)
library(Matrix)
library(readxl)

set.seed(123)
```

## Import data

```{r import SFARI gene list, androgen response genes, and TT results}
SFARI_Gene <- read_csv("../../doc/targeted/SFARI-Gene_genes_07-08-2025release_07-22-2025export.csv", col_types = cols(status = col_skip()))

androgen_response_up <- read_csv("../../doc/targeted/Upregulated_DEGs.csv", col_names = FALSE)
androgen_response_down <- read_csv("../../doc/targeted/Downregulated_DEGs.csv", col_names = FALSE)

mouse_response <- read_csv("../../doc/targeted/converted_sex_tt_mouse.csv") #already converted by a web app, and values were copied form pdf manually

singlecell_response <- read_xlsx("../../doc/targeted/li_2024_sex_diff_degs.xlsx", sheet = "AASB-DEGs", skip = 1) # these are still mouse symbols

TT_results <- read_csv("../../results/targeted/TT_results.csv")
```


```{r filter gene list and genes}
SFARI_Gene_filt <- SFARI_Gene %>% dplyr::filter(`gene-score` <= 2)

TT_sig_results <- res_TT$combined_results %>%
  dplyr::filter(sex == "M") %>%
  dplyr::filter(adj.P.Val < 0.05 & abs(logFC) >= 0.25)

E2_sig_results <- res_E2$combined_results %>%
  dplyr::filter(sex == "M") %>%
  dplyr::filter(adj.P.Val < 0.05 & abs(logFC) >= 0.25)

P4_sig_results <- res_P4$combined_results %>%
  dplyr::filter(sex == "M") %>%
  dplyr::filter(adj.P.Val < 0.05 & abs(logFC) >= 0.25)

```

```{r check for SFARI intersect}
TT_SFARI_shared_genes <- intersect(SFARI_Gene_filt$`gene-symbol`, TT_sig_results$gene)

TT_SFARI_shared_sig_results <- TT_sig_results %>%
  dplyr::filter(gene %in% TT_SFARI_shared_genes)

E2_SFARI_shared_genes <- intersect(SFARI_Gene_filt$`gene-symbol`, E2_sig_results$gene)

E2_SFARI_shared_sig_results <- E2_sig_results %>%
  dplyr::filter(gene %in% E2_SFARI_shared_genes)

P4_SFARI_shared_genes <- intersect(SFARI_Gene_filt$`gene-symbol`, P4_sig_results$gene)

P4_SFARI_shared_sig_results <- P4_sig_results %>%
  dplyr::filter(gene %in% P4_SFARI_shared_genes)
```

```{r visualize SFARI genes}
plot_SFARI_genes <- function(by_group_data, output_dir, hormone, top_n = 6) {
  dir.create(output_dir, showWarnings = FALSE)
  
  message("Number of groups to plot: ", length(by_group_data))
  
  purrr::walk(by_group_data, function(group_list) {
    if (is.null(group_list)) {
      message("  Skipping NULL group")
      return()
    }
    
    limma_res <- group_list$limma_results
    counts_sub <- group_list$counts_sub  # genes x samples
    metadata_sub <- group_list$metadata_sub

    if (nrow(metadata_sub) < 1 || ncol(counts_sub) < 1) {
      message("  Skipping empty subset: ", metadata_sub$cell_type[1], "_", metadata_sub$Sex[1])
      return()
    }

    # Compute log1p(CPM)
    counts_cpm <- edgeR::cpm(counts_sub, log = FALSE)
    counts_log <- log1p(counts_cpm)
    
    # Transpose counts: samples x genes
    counts_log <- t(counts_log)

    top_genes <- limma_res %>%
      filter(adj.P.Val < 0.05 & abs(logFC) >= 0.25) %>%
      filter(gene %in% SFARI_Gene_filt$`gene-symbol`) %>%
      pull(gene)

    message("  Plotting group: ", metadata_sub$cell_type[1], "_", metadata_sub$Sex[1])
    message("  Top genes: ", paste(top_genes, collapse = ", "))

    plot_list <- purrr::map(top_genes, function(gene) {
      if (!gene %in% colnames(counts_log)) {
        message("    Gene ", gene, " not in counts — skipping")
        return(NULL)
      }

      expr_vals <- counts_log[, gene]
      if (all(is.na(expr_vals)) || all(expr_vals == 0)) {
        message("    Gene ", gene, " has all NA or zero expression — skipping")
        return(NULL)
      }

      df <- metadata_sub %>% mutate(expr = expr_vals)

      ggplot(df, aes_string(x = hormone, y = "expr", color = "Sex", fill = "Sex")) +
        geom_point(size = 2, alpha = 0.8) +
        geom_smooth(method = "lm", se = TRUE) +
        scale_color_manual(values = c("M" = "steelblue", "F" = "violetred3")) +
        scale_fill_manual(values = c("M" = "steelblue", "F" = "violetred3")) +
        labs(title = paste0(gene, " in ", metadata_sub$cell_type[1], " (", metadata_sub$Sex[1], ")"),
             x = hormone, y = "log1p(CPM)") +
        theme_minimal(base_size = 14)
    }) %>% compact()

    if (length(plot_list) == 0) {
      message("    No valid plots for group")
      return()
    }

    combined_plot <- patchwork::wrap_plots(plotlist = plot_list, ncol = 2)
    clean_name <- paste0(metadata_sub$cell_type[1], "_", metadata_sub$Sex[1])
    clean_name <- gsub("/", "-", clean_name)

    ggsave(file.path(output_dir, paste0(clean_name, "_top_genes.png")),
           plot = combined_plot, width = 10, height = 6 + ceiling(length(plot_list)/2), dpi = 300)
    message("  ✅ Saved plot for ", clean_name)
  })
}

plot_SFARI_genes(res_TT$by_group_data, "../../results/targeted/SFARI_testosterone_plots", hormone = "TT")

plot_SFARI_genes(res_E2$by_group_data, "../../results/targeted/SFARI_estradiol_plots", hormone = "E2")

plot_SFARI_genes(res_P4$by_group_data, "../../results/targeted/SFARI_progesterone_plots", hormone = "P4")

```


```{r make stacked bar chart with all cell types}

summarize_sig_genes <- function(results, SFARI_Gene, cell_type, sex, fdr_cutoff = 0.05, logfc_cutoff = 0.25) {
  required_cols <- c("adj.P.Val", "logFC", "gene")
  if (!all(required_cols %in% colnames(results))) {
    warning("Missing required columns in results. Skipping.")
    return(NULL)
  }
  
  sig <- results %>% filter(adj.P.Val < fdr_cutoff)
  if (!is.null(logfc_cutoff)) {
    sig <- sig %>% filter(abs(logFC) >= logfc_cutoff)
  }

  if (nrow(sig) == 0) return(NULL)

  sig %>%
    mutate(SFARI_Status = ifelse(gene %in% SFARI_Gene$`gene-symbol`, "SFARI", "Other"),
           cell_type = cell_type,
           sex = sex) %>%
    count(cell_type, sex, SFARI_Status, name = "num_sig_genes") %>%
    arrange(desc(num_sig_genes))
}

plot_sig_gene_summary <- function(summary_df) {
  ggplot(summary_df, aes(x = cell_type, y = num_sig_genes, fill = SFARI_Status)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(aes(label = num_sig_genes), 
              position = position_stack(vjust = 0.5), 
              color = "white", size = 4) +
    labs(x = "Cell Type", y = "# Significant Genes")+
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("SFARI" = "steelblue", "Other" = "lightblue")) +
    guides(fill = guide_legend(title = "SFARI Status")) +
    theme_minimal(base_size = 14) 
}



plot_SFARI_genes_from_subsets <- function(by_group_data, SFARI_Gene, output_dir, hormone, top_n = 6) {
  dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
  message("Number of groups to plot: ", length(by_group_data))

  all_summaries <- map_dfr(by_group_data, function(group_list) {
    if (is.null(group_list)) {
      message("Skipping NULL group")
      return(NULL)
    }

    limma_res <- group_list$limma_results
    counts_sub <- group_list$counts_sub  # genes x samples
    metadata_sub <- group_list$metadata_sub

    if (is.null(metadata_sub) || is.null(counts_sub) ||
        nrow(metadata_sub) < 1 || ncol(counts_sub) < 1) {
      message("  Skipping empty or malformed subset")
      return(NULL)
    }

    if (!all(c("cell_type", "Sex") %in% colnames(metadata_sub))) {
      warning("  Missing 'cell_type' or 'Sex' in metadata_sub. Skipping group.")
      return(NULL)
    }

    result_df <- summarize_sig_genes(
      limma_res,
      SFARI_Gene,
      metadata_sub$cell_type[1],
      metadata_sub$Sex[1]
    )
    
    print(head(result_df))  # Debugging output
    return(result_df)
  })
  

  # Debug 'Sex' column added correctly
  print(head(all_summaries))
      
  male_summaries <- all_summaries %>%
    filter(Sex == "M")
  
  female_summaries <- all_summaries %>%
    filter(Sex == "F")
  
  if (is.null(male_summaries) || nrow(male_summaries) == 0) {
    message("No significant genes found across male groups")
  } else {
    plot_title_male <- paste("Significant Gene Summary Across Male Groups (", hormone, ")", sep = "")
    summary_plot_male <- plot_sig_gene_summary(male_summaries)
    ggsave(file.path(output_dir, "combined_sig_gene_summary_male.png"),
           plot = summary_plot_male, width = 12, height = 8, dpi = 300)
    message("  ✅ Saved combined summary plot (male only)")
  }

  if (is.null(female_summaries) || nrow(female_summaries) == 0) {
    message("No significant genes found across female groups")
  } else {
    plot_title_female <- paste("Significant Gene Summary Across Female Groups (", hormone, ")", sep = "")
    summary_plot_female <- plot_sig_gene_summary(female_summaries)
    ggsave(file.path(output_dir, "combined_sig_gene_summary_female.png"),
           plot = summary_plot_female, width = 12, height = 8, dpi = 300)
    message("  ✅ Saved combined summary plot (female only)")
  }
}

# Call function 
plot_SFARI_genes_from_subsets(
  by_group_data = res_TT$by_group_data,
  SFARI_Gene = SFARI_Gene,
  output_dir = "../../results/targeted/SFARI_testosterone_plots",
  hormone = "TT"
)

# Call function 
plot_SFARI_genes_from_subsets(
  by_group_data = res_E2$by_group_data,
  SFARI_Gene = SFARI_Gene,
  output_dir = "../../results/targeted/SFARI_estradiol_plots",
  hormone = "E2"
)

# Call function 
plot_SFARI_genes_from_subsets(
  by_group_data = res_P4$by_group_data,
  SFARI_Gene = SFARI_Gene,
  output_dir = "../../results/targeted/SFARI_progesterone_plots",
  hormone = "P4"
)


```


```{r join SFARI information}
summarize_sig_genes <- function(results, SFARI_Gene, cell_type, sex, fdr_cutoff = 0.05, logfc_cutoff = 0.25) {
  required_cols <- c("adj.P.Val", "logFC", "gene")
  if (!all(required_cols %in% colnames(results))) {
    warning("Missing required columns in results. Skipping.")
    return(NULL)
  }

  sig <- results %>% filter(adj.P.Val < fdr_cutoff)
  if (!is.null(logfc_cutoff)) {
    sig <- sig %>% filter(abs(logFC) >= logfc_cutoff)
  }

  if (nrow(sig) == 0) return(NULL)

  sig <- sig %>%
    left_join(SFARI_Gene %>% select(`gene-symbol`, `gene-score`), by = c("gene" = "gene-symbol")) %>%
    mutate(score_group = ifelse(is.na(`gene-score`), "Not SFARI", as.character(`gene-score`)),
           cell_type = cell_type,
           sex = sex)

  sig %>%
    count(cell_type, sex, score_group, name = "num_sig_genes") %>%
    arrange(desc(num_sig_genes))
}

plot_sig_gene_summary <- function(summary_df) {
  # Set factor order for coloring if needed
  summary_df <- summary_df %>%
    mutate(score_group = factor(score_group, levels = sort(unique(score_group))))

  ggplot(summary_df, aes(x = cell_type, y = num_sig_genes, fill = score_group)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(aes(label = num_sig_genes),
              position = position_stack(vjust = 0.5),
              color = "white", size = 4) +
    labs(x = "Cell Type", y = "# Significant Genes", fill = "SFARI Gene Score") +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c(
      "1" = "#bdd7e7",  # light blue
      "2" = "#6baed6",  # medium blue
      "3" = "#3182bd",  # darker blue
      "4" = "#08519c",  # darkest blue
      "Not SFARI" = "gray80"  # non-SFARI genes
    )) +
    guides(fill = guide_legend(title = "SFARI Gene Score"))
}

plot_SFARI_genes_from_subsets <- function(by_group_data, SFARI_Gene, output_dir, hormone, top_n = 6) {
  dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
  message("Number of groups to plot: ", length(by_group_data))

  all_summaries <- map_dfr(by_group_data, function(group_list) {
    if (is.null(group_list)) {
      message("  Skipping NULL group")
      return(NULL)
    }

    limma_res <- group_list$limma_results
    counts_sub <- group_list$counts_sub  # genes x samples
    metadata_sub <- group_list$metadata_sub

    if (is.null(metadata_sub) || is.null(counts_sub) ||
        nrow(metadata_sub) < 1 || ncol(counts_sub) < 1) {
      message("  Skipping empty or malformed subset")
      return(NULL)
    }

    if (!all(c("cell_type", "Sex") %in% colnames(metadata_sub))) {
      warning("  Missing 'cell_type' or 'Sex' in metadata_sub. Skipping group.")
      return(NULL)
    }

    if (metadata_sub$Sex[1] != "M") {
      message("  Skipping non-male group: ", metadata_sub$cell_type[1], "_", metadata_sub$Sex[1])
      return(NULL)
    }

    summarize_sig_genes(
      limma_res,
      SFARI_Gene,
      metadata_sub$cell_type[1],
      metadata_sub$Sex[1]
    )
  })

  if (is.null(all_summaries) || nrow(all_summaries) == 0) {
    message("No significant genes found across male groups")
    return(NULL)
  }

  # Plot the combined summary of significant genes for males only
  plot_title <- paste("Significant Gene Summary Across Male Groups (", hormone, ")", sep = "")
  summary_plot <- plot_sig_gene_summary(all_summaries)
  ggsave(file.path(output_dir, "combined_SFARI_score_summary_male.png"),
         plot = summary_plot, width = 12, height = 8, dpi = 300)
  message("  ✅ Saved combined summary plot (male only)")
}

# Plot the combined summary of significant genes for females only
  plot_title <- paste("Significant Gene Summary Across Female Groups (", hormone, ")", sep = "")
  summary_plot <- plot_sig_gene_summary(all_summaries)
  ggsave(file.path(output_dir, "combined_SFARI_score_summary_male.png"),
         plot = summary_plot, width = 12, height = 8, dpi = 300)
  message("  ✅ Saved combined summary plot (male only)")
}


plot_SFARI_genes_from_subsets(
  by_group_data = res_TT$by_group_data,
  SFARI_Gene = SFARI_Gene,
  output_dir = "../../results/targeted/SFARI_testosterone_plots",
  hormone = "TT"
)

```

## Check other datasets for overlap

Make sure you run this code after plotting, biomaRt and annotationDBI mess up some tidyverse functions (like select)

```{r prepare other data}
androgen_response_all <- c(androgen_response_up$X1, androgen_response_down$X1)

mouse_response_genes <- mouse_response$symbol_human

singlecell_needs_converting <- singlecell_response$Gene_symbol
```


```{r convert genes from mouse to human symbols}
library(biomaRt)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(org.Hs.eg.db)

# Load the file with fread
url <- "https://ftp.ensembl.org/pub/current_tsv/ensembl-compara/homologies/mus_musculus/Compara.114.protein_default.homologies.tsv.gz"
orthologs <- fread(url)

# Check the column names
head(colnames(orthologs))

orthologs <- as.data.frame(orthologs)

# Filter for human orthologs only
mouse2human <- orthologs %>%
  filter(homology_species == "homo_sapiens")

# Optional: keep only 1:1 orthologs
mouse2human_1to1 <- filter(mouse2human, homology_type == "ortholog_one2one")

# Step 3: Map mouse Ensembl gene IDs to symbols
mouse_symbols <- mapIds(
  org.Mm.eg.db,
  keys = mouse2human_1to1$gene_stable_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

# Step 4: Map human Ensembl gene IDs to symbols
human_symbols <- mapIds(
  org.Hs.eg.db,
  keys = mouse2human_1to1$homology_gene_stable_id,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

# Step 5: Attach symbols to the table
mouse2human_1to1$mouse_symbol <- mouse_symbols
mouse2human_1to1$human_symbol <- human_symbols

# Step 7: Subset to your gene list
mouse_genes <- singlecell_needs_converting
converted <- mouse2human_1to1 %>%
  filter(mouse_symbol %in% mouse_genes)

# Optional: just get human symbols
converted_human_genes <- unique(converted$human_symbol)

```


```{r check for overlap}

# again for androgen response
ar_shared_genes <- intersect(androgen_response_all, TT_sig_results$gene)
ar_shared_genes

# again for mouse stem cells
mouse_shared_genes <- intersect(mouse_response_genes, TT_sig_results$gene)
mouse_shared_genes

# again for single cell mouse data
converted_human_genes_shared <- intersect(converted_human_genes, TT_sig_results$gene)
converted_human_genes_shared

```



