---
title: "Correlation of Hormone Concentration and Gene Expression Data"
author: "Emma Jones"
date: "2025-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of his script is to explore single cell RNA-seq results (and potentially scATAC), and correlate these levels with testosterone, progesterone, and estradiol concentrations from fetal human brain samples.

## Load packages

First, we load packages. We might want more later, but for now I know I need the following

```{r load packages}
library(tidyverse)
library(Seurat)
library(readxl)
library(Signac)
library(glue)
library(cowplot)
library(edgeR)
library(limma)
library(patchwork)
```


## Load data

```{r load seurat object and excel spreadsheet}
# load 10x multiome data
SSD_data_v5 <- readRDS("../../data/SSD_data_v5.rds")

# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx("../../doc/targeted/Targeted Comparison.xlsx")

```

## Remove Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

```{r detect and remove female outlier}
#filter for only females
targeted_comparision_females <- targeted_comparision_raw[targeted_comparision_raw$Sex == "F",]

IQR(targeted_comparision_females$TT)

outlier_limit <- quantile(targeted_comparision_females$TT)[4] + 1.5*IQR(targeted_comparision_females$TT)

max(targeted_comparision_females$TT) > outlier_limit 

```

So, yes, this is above the typical outlier limit so we can remove from analysis. It is sample number 38, which has an XX Xist genotype and sex_genome_studio asignment of XX as well, so it seems to be a true female sample and not a swap.

```{r remove outlier}
targeted_comparision_filtered <- targeted_comparision_raw[-34,]
```


## how does this correlate with age?
```{r look at overall relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  geom_point() + theme_minimal()

all_simple <- lm(TT~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```

so its not signficant, but the graph clearly supports that males differ from females, so I will separate the groups

```{r separate groups and find a model}
targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_males) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_males) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_males)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_males$GW)

# Base plot - males
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 


# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run for females}
#filter for only females
targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_females) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_females) 

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 


# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 

```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_m <- loess(TT ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_f <- loess(TT ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = TT, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outlier removed)", x = "Gestational Week (GW)", y = "Testosterone") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

# Save
ggsave("../../results/TT_loess_graph.png")

```


```{r model using best-fitting equation}
male_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.01292

female_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.7959

# Create prediction frame - male
new_male <- data.frame(GW = seq(min(targeted_comparision_males$GW),
                                max(targeted_comparision_males$GW), length.out = 100))
new_male$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_males),
                            newdata = new_male)

# Create prediction frame - female
new_female <- data.frame(GW = seq(min(targeted_comparision_females$GW),
                                max(targeted_comparision_females$GW), length.out = 100))
new_female$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_females),
                            newdata = new_female)

# Plot 1
ggplot(targeted_comparision_males, aes(x = GW, y = TT)) +
  geom_point(color = "dodgerblue", alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Males", x = "GW", y = "TT") +
  theme_minimal()

# Plot 2
ggplot(targeted_comparision_females, aes(x = GW, y = TT)) +
  geom_point(color = "violetred2", alpha = 0.6) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Females", x = "GW", y = "TT") +
  theme_minimal()

# Combine information
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² + I(GW²) Fit in Males and Females", subtitle = "Male p-value: 0.01292\nFemale p-value:0.7959", x = "GW", y = "TT") +
  theme_minimal()

# move the text
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² + I(GW²) Fit in Males and Females", x = "Gestational Week (GW)", y = "Testosterone") +
  annotate("text", x = 16, y = 0.4, label = "Male p-value: 0.01292", color = "dodgerblue4", hjust = 0) +
  annotate("text", x = 16, y = 0.35, label = "Female p-value: 0.7959", color = "violetred4", hjust = 0) +
  theme_minimal()

# save new plot
ggsave("../../results/TT_GW_modelfit.png")

```

# Work on understanding expression patterns and TT level - DO NOT SKIP!

If there is a quadratic equation that explains expression over time, it is possible the relationship between TT and gene expression is also nonlinear. Will start with linear first, since that is more straightforward.

```{r Combine metadata and expression matrix and Prepare and join sample-level metadata with TT concentrations}
cell_metadata <- SSD_data_v5@meta.data %>%
  select(sample, cluster_call)
cell_metadata$cell_id <- rownames(cell_metadata) 

targeted_comparision_filtered$Sample <- paste0("tissue", targeted_comparision_filtered$Sample)

cell_metadata <- cell_metadata %>%
  left_join(targeted_comparision_filtered, by = c("sample" = "Sample"))

# Rename cell types for clarity
cell_metadata$cell_type <- cell_metadata$cluster_call

# remove rows where testosterone is not measured
cell_metadata <- cell_metadata[!is.na(cell_metadata$TT),]
```

# Try pseudobulk approach

```{r psuedobulk data function}
pseudobulk_data <- function(seurat_obj, assay = "RNA", layer = "counts", 
                            sample_col = "sample_id", celltype_col = "cell_type") {
  # Get raw counts
  counts <- GetAssayData(seurat_obj, assay = assay, layer = layer)
  meta <- seurat_obj@meta.data
  
  # Group cells by sample and cell type
  group_ids <- interaction(meta[[sample_col]], meta[[celltype_col]], drop = TRUE)
  group_levels <- levels(group_ids)
  
  # Initialize list to store aggregated matrices
  aggregated_list <- list()
  metadata_list <- list()
  
  for (gid in group_levels) {
    # Get cells for this group
    group_cells <- which(group_ids == gid)
    if (length(group_cells) < 1) next
    
    # Sum counts across cells
    group_counts <- Matrix::rowSums(counts[, group_cells, drop = FALSE])
    
    # Store as data frame row
    aggregated_list[[gid]] <- group_counts
    metadata_list[[gid]] <- data.frame(
      sample_id = meta[[sample_col]][group_cells[1]],
      cell_type = meta[[celltype_col]][group_cells[1]],
      cell_count = length(group_cells),
      row.names = gid
    )
  }
  
  # Combine into matrices and metadata frame
  pseudobulk_mat <- do.call(rbind, aggregated_list)
  metadata_df <- do.call(rbind, metadata_list)
  
  # Log-transform
  pseudobulk_mat_log <- log1p(pseudobulk_mat)
  
  return(list(counts = pseudobulk_mat, logcounts = pseudobulk_mat_log, metadata = metadata_df))
}
```


```{r pseudobulk}
# Run pseudobulk
pb <- pseudobulk_data(SSD_data_v5, sample_col = "sample", celltype_col = "cluster_call")

# View structure
str(pb)
```

```{r prepare metadata}
# Convert rownames to a column so they survive the join
pb_metadata <- pb$metadata %>%
  rownames_to_column("pseudobulk_id")

# Join with testosterone/sex data
metadata_merged <- pb_metadata %>%
  left_join(targeted_comparision_filtered %>% select(Sample, Sex), 
            by = c("sample_id" = "Sample")) %>%
  filter(!is.na(Sex)) %>%
  mutate(group = paste(cell_type, Sex, sep = "_")) %>%
  column_to_rownames("pseudobulk_id")

# Filter out samples with missing sex or too few cells
metadata_merged <- metadata_merged %>%
  filter(!is.na(Sex)) %>%
  filter(cell_count >= 10)  # Remove sparse pseudobulks

# Subset logcounts and counts to only include valid samples
pb$logcounts <- pb$logcounts[rownames(pb$logcounts) %in% rownames(metadata_merged), , drop = FALSE]
pb$counts <- pb$counts[rownames(pb$counts) %in% rownames(metadata_merged), , drop = FALSE]

# Reorder metadata to match logcounts and counts
metadata_merged <- metadata_merged[match(rownames(pb$logcounts), rownames(metadata_merged)), ]
metadata_merged <- metadata_merged[match(rownames(pb$counts), rownames(metadata_merged)), ]

# Split into pseudobulk expression matrices by cell type + sex
pseudobulk_expr_by_group <- split(seq_len(nrow(pb$counts)), metadata_merged$group) %>%
  purrr::map(~ {
    mat <- pb$counts[.x, , drop = FALSE]
    
    # Filter genes: retain only those with at least three samples with count >= 3
    keep_genes <- which(colSums(mat >= 3) >= 3)
    mat[, keep_genes, drop = FALSE]
  })
```

## Run limma to model everything

```{r prepare to run limma}
metadata_merged <- left_join(metadata_merged, targeted_comparision_filtered, by = c("sample_id" = "Sample", "Sex" = "Sex"))

# Create unique rownames by combining sample_id and cell_type
metadata_merged <- metadata_merged %>%
  mutate(pseudobulk_id = paste(sample_id, cell_type, sep = "_")) %>%
  column_to_rownames("pseudobulk_id")

# Update pb$counts and pb$logcounts rownames accordingly
rownames(pb$counts) <- rownames(metadata_merged)
rownames(pb$logcounts) <- rownames(metadata_merged)

# Confirm alignment
stopifnot(identical(rownames(pb$counts), rownames(metadata_merged)))

run_voom_lm <- function(counts, metadata) {
  # Transpose counts: genes as rows, samples as columns
  counts_t <- t(counts)

  # Check dimensions
  stopifnot(ncol(counts_t) == nrow(metadata))

  # Create design matrix
  design <- model.matrix(~ TT + cell_count, data = metadata)

  # Run voom
  v <- limma::voom(counts_t, design)

  # Fit model
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)

  # Tidy results
  results <- limma::topTable(fit, coef = "TT", number = Inf, sort.by = "none") %>%
    tibble::rownames_to_column("gene")

  return(results)
}

# Ensure sample order matches
stopifnot(identical(rownames(pb$counts), rownames(metadata_merged)))

# Split by both cell type and sex
split_indices <- split(seq_len(nrow(pb$counts)), interaction(metadata_merged$cell_type, metadata_merged$Sex, drop = TRUE))
```

```{r run limma for each group}
# Run model per group
results_by_group <- purrr::map(split(seq_len(nrow(pb$counts)), metadata_merged$group), function(idx) {
  counts_sub <- pb$counts[idx, , drop = FALSE]
  metadata_sub <- metadata_merged[idx, , drop = FALSE]

  # Remove samples with zero library size
  keep_samples <- rowSums(counts_sub) > 0
  counts_sub <- counts_sub[keep_samples, , drop = FALSE]
  metadata_sub <- metadata_sub[keep_samples, , drop = FALSE]

  # Skip group if too few samples remain
  if (nrow(metadata_sub) < 5) {
    message("Skipping group ", unique(metadata_sub$group), " due to too few samples.")
    return(NULL)
  }

  # Optional: filter genes with low expression across samples
  keep_genes <- colSums(counts_sub >= 3) >= 5
  counts_sub <- counts_sub[, keep_genes, drop = FALSE]

  if (ncol(counts_sub) == 0) {
    message("Skipping group ", unique(metadata_sub$group), " due to no genes passing filter.")
    return(NULL)
  }

  # Run voom + lm with error handling
  tryCatch({
    result <- run_voom_lm(counts_sub, metadata_sub)
    result$cell_type <- metadata_sub$cell_type[1]
    result$sex <- metadata_sub$Sex[1]
    result
  }, error = function(e) {
    message("Skipping group ", unique(metadata_sub$group), " due to error: ", e$message)
    NULL
  })
})
```

```{r examine limma results}
# Combine all results (remove NULLs)
results_all <- bind_rows(results_by_group[!sapply(results_by_group, is.null)])

# examine female results
results_female <- results_all %>%
  filter(sex == "F")

# examine male results
results_male <- results_all %>%
  filter(sex == "M")

# Get top genes
top_genes <- results_all %>%
  filter(adj.P.Val < 0.05) %>%
  group_by(cell_type) %>%
  arrange(desc(abs(t))) %>%
  slice_head(n = 5) %>%
  ungroup()

  
str(pb$counts)
summary(pb$counts)
anyNA(pb$counts)
any(rowSums(pb$counts) == 0)
any(colSums(pb$counts) == 0)

# Remove genes (columns) with zero total count
pb_counts_filtered <- pb$counts[, colSums(pb$counts) > 0]

# make log(cpm) matrix for plotting
log_cpm_mat <- cpm(pb_counts_filtered, log = TRUE, prior.count = 1)
```


```{r make and save male plots}
# Get unique male cell types from results
cell_types <- unique(results_male$cell_type)

# Directory to save plots
output_dir <- "../../results/targeted/pseudobulk_plots"
dir.create(output_dir, showWarnings = FALSE)

# Loop through groups and save combined plots
walk(cell_types, function(group_of_interest) {
  
  # === Filter for this group ===
  results_subset <- results_male %>%
    filter(cell_type == group_of_interest) %>%
    arrange(P.Value) %>%
    slice_head(n = n_top)
  
  # === Subset expression and metadata ===
  samples_in_group <- metadata_merged %>%
    filter(cell_type == group_of_interest)
  
  if (nrow(samples_in_group) == 0) return()
  counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]
  
  # === Plot top genes ===
  plot_list <- map(results_subset$gene, function(gene) {
    if (!gene %in% colnames(counts_in_group)) return(NULL)
    
    expr_vals <- counts_in_group[, gene]
    if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
    
    df <- samples_in_group %>%
      mutate(expr = expr_vals) %>%
      filter(Sex == "M")
    
    ggplot(df, aes(x = TT, y = expr)) +
      geom_point() +
      geom_smooth(method = "lm", se = TRUE, color = "steelblue", fill = "steelblue") +
      labs(
        title = paste0(gene, " in ", group_of_interest, ": Males"),
        x = "Testosterone",
        y = "LogCPM"
      ) +
      theme_minimal()
  }) %>% compact()  # remove NULLs
  
  if (length(plot_list) == 0) return()
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
  
  # Clean group name for filename
clean_name <- gsub("/", "-", group_of_interest)
  
  # Save to PNG
  ggsave(
    filename = file.path(output_dir, paste0(clean_name, "_male_top_genes.png")),
    plot = combined_plot,
    width = 10,
    height = 6 + ceiling(length(plot_list) / 2),
    dpi = 300
  )
})

```



```{r make and save female plots}
# Get unique male cell types from results
cell_types <- unique(results_female$cell_type)

# Directory to save plots
output_dir <- "../../results/targeted/pseudobulk_plots"
dir.create(output_dir, showWarnings = FALSE)

# Loop through groups and save combined plots
walk(cell_types, function(group_of_interest) {
  
  # === Filter for this group ===
  results_subset <- results_female %>%
    filter(cell_type == group_of_interest) %>%
    arrange(P.Value) %>%
    slice_head(n = n_top)
  
  # === Subset expression and metadata ===
  samples_in_group <- metadata_merged %>%
    filter(cell_type == group_of_interest)
  
  if (nrow(samples_in_group) == 0) return()
  counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]

  
  # === Plot top genes ===
  plot_list <- map(results_subset$gene, function(gene) {
    if (!gene %in% colnames(counts_in_group)) return(NULL)
    
    expr_vals <- counts_in_group[, gene]
    if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
    
    df <- samples_in_group %>%
      mutate(expr = expr_vals) %>%
      filter(Sex == "F")
    
    ggplot(df, aes(x = TT, y = expr)) +
      geom_point() +
      geom_smooth(method = "lm", se = TRUE, color = "violetred3", fill = "violetred3") +
      labs(
        title = paste0(gene, " in ", group_of_interest, ": Females"),
        x = "Testosterone",
        y = "LogCPM"
      ) +
      theme_minimal()
  }) %>% compact()  # remove NULLs
  
  if (length(plot_list) == 0) return()
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
  
  # Clean group name for filename
clean_name <- gsub("/", "-", group_of_interest)
  
  # Save to PNG
  ggsave(
    filename = file.path(output_dir, paste0(clean_name, "_female_top_genes.png")),
    plot = combined_plot,
    width = 10,
    height = 6 + ceiling(length(plot_list) / 2),
    dpi = 300
  )
})

```
