---
title: "Correlate Age and Metabolite Concentrations"
author: "Emma Jones"
date: "2025-06-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


First, we load packages. We might want more later, but for now I know I need the following

```{r load packages}
library(tidyverse)
library(Seurat)
library(readxl)
library(Signac)
library(glue)
library(cowplot)
library(edgeR)
library(limma)
library(patchwork)
```

## Load data

```{r load seurat object and excel spreadsheet}
# load 10x multiome data
SSD_data_v5 <- readRDS("../../data/SSD_data_v5.rds")
# remove doublets
SSD_data_v5 <- subset(SSD_data_v5, subset = Droplet == "SNG")

# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx("../../doc/targeted/Targeted Comparison.xlsx")

#rename estrogen
targeted_comparision_raw$E2 <- targeted_comparision_raw$`17B-E2`

```

# testosterone analysis

## Remove Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

```{r detect and remove female outlier}
#filter for only females
targeted_comparision_females <- targeted_comparision_raw[targeted_comparision_raw$Sex == "F",]

outlier_limit <- quantile(targeted_comparision_raw$TT)[4] + 1.5*IQR(targeted_comparision_raw$TT)

max(targeted_comparision_females$TT) > outlier_limit 

targeted_comparision_females$TT > outlier_limit
```

So, yes, this is above the typical outlier limit so we can remove from analysis. It is sample number 38, which has an XX Xist genotype and sex_genome_studio asignment of XX as well, so it seems to be a true female sample and not a swap. However, there are other samples that are also outliers by this definition, so we may want to be cautious by how this is represented

```{r remove outlier}
targeted_comparision_filtered <- targeted_comparision_raw[-34,]
```


## how does this correlate with age?
```{r look at overall relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  geom_point() + theme_minimal()

all_simple <- lm(TT~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```

so its not signficant, but the graph clearly supports that males differ from females, so I will separate the groups

```{r separate groups and find a model}
targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_males) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_males) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_males)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_males$GW)

# Base plot - males
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 

# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run for females}
#filter for only females
targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_females) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_females) 

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 


# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 

```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_m <- loess(TT ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_f <- loess(TT ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = TT, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outlier removed)", x = "Gestational Week (GW)", y = "Testosterone") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

# Save
ggsave("../../results/TT_loess_graph.png")

```


```{r model using best-fitting equation}
male_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.01292

female_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.7959

# Create prediction frame - male
new_male <- data.frame(GW = seq(min(targeted_comparision_males$GW),
                                max(targeted_comparision_males$GW), length.out = 100))
new_male$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_males),
                            newdata = new_male)

# Create prediction frame - female
new_female <- data.frame(GW = seq(min(targeted_comparision_females$GW),
                                max(targeted_comparision_females$GW), length.out = 100))
new_female$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_females),
                            newdata = new_female)

# Plot 1
ggplot(targeted_comparision_males, aes(x = GW, y = TT)) +
  geom_point(color = "dodgerblue", alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Males", x = "GW", y = "TT") +
  theme_minimal()

# Plot 2
ggplot(targeted_comparision_females, aes(x = GW, y = TT)) +
  geom_point(color = "violetred2", alpha = 0.6) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Females", x = "GW", y = "TT") +
  theme_minimal()

# Combine information
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² + I(GW²) Fit in Males and Females", subtitle = "Male p-value: 0.01292\nFemale p-value:0.7959", x = "GW", y = "TT") +
  theme_minimal()

# move the text
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² + I(GW²) Fit in Males and Females", x = "Gestational Week (GW)", y = "Testosterone") +
  annotate("text", x = 16, y = 0.4, label = "Male p-value: 0.01292", color = "dodgerblue4", hjust = 0) +
  annotate("text", x = 16, y = 0.35, label = "Female p-value: 0.7959", color = "violetred4", hjust = 0) +
  theme_minimal()

# save new plot
ggsave("../../results/TT_GW_modelfit.png")

```

# do again for estrogen

## Detect Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

```{r detect and remove female outlier}
#filter for only females and males
targeted_comparision_females <- targeted_comparision_raw[targeted_comparision_raw$Sex == "F",]
targeted_comparision_males <- targeted_comparision_raw[targeted_comparision_raw$Sex == "M",]

outlier_limit <- quantile(targeted_comparision_raw$E2)[4] + 1.5*IQR(targeted_comparision_raw$E2)

to_keep <- targeted_comparision_raw$E2 > outlier_limit
```

So, yes, we have 5 outliers here, but we can remove them all

```{r remove outliers}
targeted_comparision_filtered <- targeted_comparision_raw[!to_keep,]
```

## how does this correlate with age?

```{r look at overall relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered, aes(x = GW, y = E2, color = Sex)) +
  geom_point() + theme_minimal()

all_simple <- lm(TT~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```

im not seeing a strong difference for males or females

```{r separate groups and find a model}
targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(E2~GW, data=targeted_comparision_females) 
linear_model2 <- lm(E2~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(E2~poly(GW,3,raw=TRUE), data=targeted_comparision_females)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$E2, main="Polynomial Fits", xlab="GW", ylab="E2")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 

# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$E2, main="Loess Fit", xlab="GW", ylab="E2")
loess_fit_m <- loess(E2 ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$E2, main="Loess Fit", xlab="GW", ylab="E2")
loess_fit_f <- loess(E2 ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = E2, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Estradiol") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

# Save
ggsave("../../results/E2_loess_graph.png")

```


```{r model using best-fitting equation}
male_model <- lm(E2 ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.2672

female_model <- lm(E2 ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.6758

```

There seems to be no significant relationship with estrogen and age, but the high outliers (5) might be influencing other analyses. I should also run outlier analysis on progesterone.

# Progesterone outlier analysis!!!

## Detect Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

```{r detect and remove female outlier}
#filter for only females and males
targeted_comparision_females <- targeted_comparision_raw[targeted_comparision_raw$Sex == "F",]
targeted_comparision_males <- targeted_comparision_raw[targeted_comparision_raw$Sex == "M",]

outlier_limit <- quantile(targeted_comparision_raw$P4)[4] + 1.5*IQR(targeted_comparision_raw$P4)

to_keep <- targeted_comparision_raw$P4 > outlier_limit

to_keep
```

So, yes, we have 5 outliers here, but we can remove them all

```{r remove outliers}
targeted_comparision_filtered <- targeted_comparision_raw[!to_keep,]
```

## how does this correlate with age?

```{r look at overall relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered, aes(x = GW, y = P4, color = Sex)) +
  geom_point() + theme_minimal()

all_simple <- lm(TT~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```

im not seeing a strong difference for males or females

```{r separate groups and find a model}
targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(P4~GW, data=targeted_comparision_females) 
linear_model2 <- lm(P4~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(P4~poly(GW,3,raw=TRUE), data=targeted_comparision_females)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$P4, main="Polynomial Fits", xlab="GW", ylab="P4")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 

# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$P4, main="Loess Fit", xlab="GW", ylab="P4")
loess_fit_m <- loess(P4 ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$P4, main="Loess Fit", xlab="GW", ylab="P4")
loess_fit_f <- loess(P4 ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = P4, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Estradiol") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

# Save
ggsave("../../results/P4_loess_graph.png")

```


```{r model using best-fitting equation}
male_model <- lm(P4 ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.3699

female_model <- lm(P4 ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.6672

```


