---
title: "Correlation of Hormone Concentration and Gene Expression Data"
author: "Emma Jones"
date: "2025-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of his script is to explore single cell RNA-seq results (and potentially scATAC), and correlate these levels with testosterone, progesterone, and estradiol concentrations from fetal human brain samples.

## Load packages

First, we load packages. We might want more later, but for now I know I need the following

```{r load packages}
library(tidyverse)
library(Seurat)
library(readxl)
library(Signac)
library(glue)
library(cowplot)

```


## Load data

```{r load seurat object and excel spreadsheet}
# load 10x multiome data
SSD_data_v5 <- readRDS("../../data/SSD_data_v5.rds")

# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx("../../doc/targeted/Targeted Comparison.xlsx")

```

## Remove Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

```{r detect and remove female outlier}
#filter for only females
targeted_comparision_females <- targeted_comparision_raw[targeted_comparision_raw$Sex == "F",]

IQR(targeted_comparision_females$TT)

outlier_limit <- quantile(targeted_comparision_females$TT)[4] + 1.5*IQR(targeted_comparision_females$TT)

max(targeted_comparision_females$TT) > outlier_limit 

```

So, yes, this is above the typical outlier limit so we can remove from analysis. It is sample number 38, which has an XX Xist genotype and sex_genome_studio asignment of XX as well, so it seems to be a true female sample and not a swap.

```{r remove outlier}
targeted_comparision_filtered <- targeted_comparision_raw[-34,]
```


## how does this correlate with age?
```{r look at overall relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  geom_point() + theme_minimal()

all_simple <- lm(TT~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```

so its not signficant, but the graph clearly supports that males differ from females, so I will separate the groups

```{r separate groups and find a model}
targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_males) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_males) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_males)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_males$GW)

# Base plot - males
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 


# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run for females}
#filter for only females
targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_females) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_females) 

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 


# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 

```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_m <- loess(TT ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_f <- loess(TT ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = TT, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outlier removed)", x = "Gestational Week (GW)", y = "Testosterone") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

# Save
ggsave("../../results/TT_loess_graph.png")

```


```{r model using best-fitting equation}
male_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.01292

female_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.7959

# Create prediction frame - male
new_male <- data.frame(GW = seq(min(targeted_comparision_males$GW),
                                max(targeted_comparision_males$GW), length.out = 100))
new_male$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_males),
                            newdata = new_male)

# Create prediction frame - female
new_female <- data.frame(GW = seq(min(targeted_comparision_females$GW),
                                max(targeted_comparision_females$GW), length.out = 100))
new_female$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_females),
                            newdata = new_female)

# Plot 1
ggplot(targeted_comparision_males, aes(x = GW, y = TT)) +
  geom_point(color = "dodgerblue", alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Males", x = "GW", y = "TT") +
  theme_minimal()

# Plot 2
ggplot(targeted_comparision_females, aes(x = GW, y = TT)) +
  geom_point(color = "violetred2", alpha = 0.6) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Females", x = "GW", y = "TT") +
  theme_minimal()

# Combine information
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² + I(GW²) Fit in Males and Females", subtitle = "Male p-value: 0.01292\nFemale p-value:0.7959", x = "GW", y = "TT") +
  theme_minimal()

# move the text
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² + I(GW²) Fit in Males and Females", x = "Gestational Week (GW)", y = "Testosterone") +
  annotate("text", x = 16, y = 0.4, label = "Male p-value: 0.01292", color = "dodgerblue4", hjust = 0) +
  annotate("text", x = 16, y = 0.35, label = "Female p-value: 0.7959", color = "violetred4", hjust = 0) +
  theme_minimal()

# save new plot
ggsave("../../results/TT_GW_modelfit.png")

```

# Work on understanding expression patterns and TT level

If there is a quadratic equation that explains expression over time, it is possible the relationship between TT and gene expression is also nonlinear. Will start with linear first, since that is more straightforward.

```{r Combine metadata and expression matrix and Prepare and join sample-level metadata with TT concentrations}
expr_mat <- as.matrix(GetAssayData(SSD_data_v5[["SCT"]], layer = "data"))  # use SCT-transformed expression

cell_metadata <- SSD_data_v5@meta.data %>%
  select(sample, cluster_call)
cell_metadata$cell_id <- rownames(cell_metadata) 

targeted_comparision_filtered$Sample <- paste0("tissue", targeted_comparision_filtered$Sample)

cell_metadata <- cell_metadata %>%
  left_join(targeted_comparision_filtered, by = c("sample" = "Sample"))

# Rename cell types for clarity
cell_metadata$cell_type <- cell_metadata$cluster_call

# remove rows where testosterone is not measured
cell_metadata <- cell_metadata[!is.na(cell_metadata$TT),]
```


```{r For each cell type, compute median expression of each gene per cell type and sample}
# Set minimum cell count threshold
min_cells <- 10

# Initialize list to store dropped samples
dropped_samples_info <- list()

# Main processing
median_expr_by_group <- cell_metadata %>%
  group_by(cell_type, Sex) %>%
  group_split() %>%
  set_names(map_chr(., ~ paste(unique(.x$cell_type), unique(.x$Sex), sep = "_"))) %>%
  imap(function(df, group_name) {
    # Count cells per sample
    cell_counts <- df %>%
      count(sample, name = "n_cells")

    # Identify low-cell samples
    low_count_samples <- cell_counts %>% filter(n_cells < min_cells)
    if (nrow(low_count_samples) > 0) {
      dropped_samples_info[[group_name]] <<- low_count_samples
    }

    # Keep only valid samples
    valid_samples <- cell_counts %>% filter(n_cells >= min_cells) %>% pull(sample)
    df_filtered <- df %>% filter(sample %in% valid_samples)
    if (length(unique(df_filtered$sample)) < 2) return(NULL)

    samples <- unique(df_filtered$sample)
    expr_sub <- expr_mat[, df_filtered$cell_id, drop = FALSE]

    median_expr_list <- map(samples, function(sid) {
      cells_in_sample <- df_filtered %>% filter(sample == sid) %>% pull(cell_id)
      matrixStats::rowMedians(expr_sub[, cells_in_sample, drop = FALSE])
    })

    median_expr <- do.call(cbind, median_expr_list)
    colnames(median_expr) <- as.character(samples)
    rownames(median_expr) <- rownames(expr_sub)

    # Remove genes with sum == 0
    median_expr <- median_expr[rowSums(median_expr, na.rm = TRUE) != 0, , drop = FALSE]

    return(median_expr)
  }) %>%
  compact()  # Remove NULLs

```


```{r For each cell type and gene, correlate or regress with testosterone}
# Helper function to calculate correlation and p-value
calculate_corr_pval <- function(gene_expr, test_vals) {
  valid_idx <- gene_expr != 0 & !is.na(gene_expr) & test_vals != 0 & !is.na(test_vals)
  
  #if (sum(valid_idx) < 2 || sd(gene_expr[valid_idx]) == 0 || sd(test_vals[valid_idx]) == 0) {
   # return(list(cor = NA, pval = NA))
 # }
  
  cor_value <- cor(gene_expr[valid_idx], test_vals[valid_idx], use = "complete.obs")
  model <- lm(gene_expr[valid_idx] ~ test_vals[valid_idx])
  pval_value <- summary(model)$coefficients[2, 4]
  
  list(cor = cor_value, pval = pval_value)
}

# Generate results by group using precomputed median expressions
results_by_group <- map2_dfr(median_expr_by_group, names(median_expr_by_group), function(expr_df, groupname) {
  samples <- colnames(expr_df)

  # Split the group name to extract cell type and sex labels
  group_labels <- strsplit(groupname, "_")[[1]]
  celltype_label <- group_labels[1]
  sex_label <- group_labels[2]

  # Obtain testosterone values for the current group of samples
  test_vals <- targeted_comparision_filtered %>%
    filter(Sample %in% samples) %>%
    arrange(match(Sample, samples)) %>%
    pull(TT)

  # Check if there is sufficient variation in testosterone values
  if (all(test_vals == 0) || sum(test_vals != 0) < 3) {
    message(glue("Skipping {groupname}: no variation in testosterone or not enough non-zero values"))
    return(NULL)  # Skip this group if conditions are met
  }
  
  # Remove low TT sample counts
  if (sum(test_vals != 0 & !is.na(test_vals)) < 4) {
  message(glue("Skipping {groupname}: < 4 non-zero testosterone values"))
  return(NULL)
    }
  
  # Apply calculate_corr_pval function to each gene expression profile
  results <- apply(expr_df, 1, function(gene_expr) calculate_corr_pval(gene_expr, test_vals))
  
  # Extract correlation and p-value from the results
  cor_values <- sapply(results, `[[`, "cor")
  pval_values <- sapply(results, `[[`, "pval")
  
  # Create a tibble to store the results
  tibble(
    gene = rownames(expr_df),
    cell_type = celltype_label,
    Sex = sex_label,
    cor = cor_values,
    pval = pval_values
  )
})
```

```{r add fdr correction and save}
results_by_group_median <- results_by_group %>%
  filter(!is.na(pval))  %>%
  group_by(cell_type, Sex) %>%
  mutate(padj = p.adjust(pval, method = "fdr")) %>%
  ungroup()

write_csv(results_by_group, "../../results/all_TT_corr_res.csv")
```

```{r get top results for males and females}
sig_results_males <- results_by_group %>% filter(Sex == "M", padj < 0.05)
sig_results_females <- results_by_group %>% filter(Sex == "F", padj < 0.05)

sig_overlap <- intersect(unique(sig_results_males$gene),unique(sig_results_females$gene))

write_csv(sig_results_males, "../../results/sig_TT_results_males.csv")
write_csv(sig_results_females, "../../results/sig_TT_results_females.csv")
```


```{r get top 3 genes per sex}
top_genes_per_sex <- results_by_group %>%
  filter(padj < 0.05) %>%
  group_by(Sex, cell_type) %>%
  arrange(desc(abs(cor))) %>%
  slice_head(n = 5) %>%
  ungroup()
```


```{r plot nicely}
# Function to generate one plot
make_scatter_plot <- function(gene, cell_type, sex, expr_df, test_vals) {
  expr_vals <- expr_df[gene, ]
  data <- tibble(
    Expression = as.numeric(expr_vals),
    Testosterone = test_vals
  ) %>%
    filter(!is.na(Expression) & !is.na(Testosterone) & Expression != 0)

  ggplot(data, aes(x = Testosterone, y = Expression)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = glue::glue("{gene} in {cell_type} ({sex})"),
      x = "Testosterone",
      y = "Median Gene Expression"
    ) +
    theme_minimal()
}

# Plot all top genes per sex
top_plots <- pmap(
  top_genes_per_sex,
  function(gene, cell_type, Sex, cor, pval, padj) {
    groupname <- paste(cell_type, Sex, sep = "_")
    expr_df <- median_expr_by_group[[groupname]]
    
    samples <- colnames(expr_df)
    test_vals <- targeted_comparision_filtered %>%
      filter(Sample %in% samples) %>%
      arrange(match(Sample, samples)) %>%
      pull(TT)

    make_scatter_plot(gene, cell_type, Sex, expr_df, test_vals)
  }
)

top_plots
```


```{r plot median gene expr results}
# Function to generate one plot with correlation and p-value
make_scatter_plot <- function(gene, cell_type, sex, expr_df, test_vals) {
  expr_vals <- expr_df[gene, ]
  data <- tibble(
    Expression = as.numeric(expr_vals),
    Testosterone = test_vals
  ) %>%
    filter(!is.na(Expression) & !is.na(Testosterone) & Expression != 0)
  
  # Calculate correlation and p-value
  valid_idx <- !is.na(data$Expression) & !is.na(data$Testosterone)
  cor_value <- cor(data$Expression[valid_idx], data$Testosterone[valid_idx])
  model <- lm(Expression ~ Testosterone, data = data)
  pval_value <- summary(model)$coefficients[2, 4]

  # Create plot
  plot <- ggplot(data, aes(x = Testosterone, y = Expression)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = glue::glue("{gene} in {cell_type} ({sex})"),
      x = "Testosterone",
      y = "Gene Expression",
      subtitle = glue::glue("Correlation: {round(cor_value, 4)}, p-value: {round(pval_value, 8)}")
    ) +
    theme_minimal()
  
  return(plot)
}

# Plot all top genes per sex
top_plots <- pmap(
  top_genes_per_sex,
  function(gene, cell_type, Sex, cor, pval, padj) {
    groupname <- paste(cell_type, Sex, sep = "_")
    expr_df <- median_expr_by_group[[groupname]]
    
    samples <- colnames(expr_df)
    test_vals <- targeted_comparision_filtered %>%
      filter(Sample %in% samples) %>%
      arrange(match(Sample, samples)) %>%
      pull(TT)

    make_scatter_plot(gene, cell_type, Sex, expr_df, test_vals)
  }
)

top_plots

```

# Try pseudobulk approach

```{r psuedobulk data function}
pseudobulk_data <- function(seurat_obj, assay = "RNA", slot = "counts", 
                            sample_col = "sample_id", celltype_col = "cell_type") {
  # Get raw counts
  counts <- GetAssayData(seurat_obj, assay = assay, slot = slot)
  meta <- seurat_obj@meta.data
  
  # Group cells by sample and cell type
  group_ids <- interaction(meta[[sample_col]], meta[[celltype_col]], drop = TRUE)
  group_levels <- levels(group_ids)
  
  # Initialize list to store aggregated matrices
  aggregated_list <- list()
  metadata_list <- list()
  
  for (gid in group_levels) {
    # Get cells for this group
    group_cells <- which(group_ids == gid)
    if (length(group_cells) < 1) next
    
    # Sum counts across cells
    group_counts <- Matrix::rowSums(counts[, group_cells, drop = FALSE])
    
    # Store as data frame row
    aggregated_list[[gid]] <- group_counts
    metadata_list[[gid]] <- data.frame(
      sample_id = meta[[sample_col]][group_cells[1]],
      cell_type = meta[[celltype_col]][group_cells[1]],
      cell_count = length(group_cells),
      row.names = gid
    )
  }
  
  # Combine into matrices and metadata frame
  pseudobulk_mat <- do.call(rbind, aggregated_list)
  metadata_df <- do.call(rbind, metadata_list)
  
  # Log-transform
  pseudobulk_mat_log <- log1p(pseudobulk_mat)
  
  return(list(counts = pseudobulk_mat, logcounts = pseudobulk_mat_log, metadata = metadata_df))
}
```



```{r pseudobulk}
# Run pseudobulk
pb <- pseudobulk_data(SSD_data_v5, sample_col = "sample", celltype_col = "cluster_call")

# View structure
str(pb)
```


```{r prepare for correlations}
# Convert rownames to a column so they survive the join
pb_metadata <- pb$metadata %>%
  tibble::rownames_to_column("pseudobulk_id")

# Join with testosterone/sex data
metadata_merged <- pb_metadata %>%
  left_join(targeted_comparision_filtered %>% select(Sample, Sex), 
            by = c("sample_id" = "Sample")) %>%
  filter(!is.na(Sex)) %>%
  mutate(group = paste(cell_type, Sex, sep = "_")) %>%
  tibble::column_to_rownames("pseudobulk_id")

# Drop samples with missing sex info (i.e., no testosterone measurement)
metadata_merged <- metadata_merged %>%
  filter(!is.na(Sex))

# Subset logcounts to only include valid samples
pb$logcounts <- pb$logcounts[rownames(pb$logcounts) %in% rownames(metadata_merged), , drop = FALSE]

# Reorder metadata to match logcounts
metadata_merged <- metadata_merged[match(rownames(pb$logcounts), rownames(metadata_merged)), ]

# Confirm alignment
stopifnot(identical(rownames(pb$logcounts), rownames(metadata_merged)))

# Split into pseudobulk expression matrices by cell type + sex
pseudobulk_expr_by_group <- split(seq_len(nrow(pb$logcounts)), metadata_merged$group) %>%
  purrr::map(~ pb$logcounts[.x, , drop = FALSE])
```



```{r run regression}
calculate_corr_pval <- function(gene_expr, test_vals) {
  valid_idx <- !is.na(gene_expr) & !is.na(test_vals)
  
  # Check if there are enough valid data points
  if (sum(valid_idx) < 2) {
    return(list(cor = NA, pval = NA, message = "Not enough valid data points"))
  }
  
  # Check if either variable has zero variance
  if (sd(gene_expr[valid_idx]) == 0 || sd(test_vals[valid_idx]) == 0) {
    return(list(cor = NA, pval = NA, message = "Zero variance in data"))
  }

  # Ensure proper alignment of data
  test_vals <- test_vals[valid_idx]
  gene_expr <- gene_expr[valid_idx]
  
  # Calculate correlation
  cor_value <- cor(gene_expr, test_vals, use = "complete.obs")
  
  # Fit linear regression
  model <- lm(gene_expr ~ test_vals)
  pval_value <- summary(model)$coefficients[2, 4]
  
  # Return results
  list(cor = cor_value, pval = pval_value, message = "Success")
}

# Run correlation/regression for each celltype-sex pseudobulk group
results_by_group_pseudo <- map2_dfr(pseudobulk_expr_by_group, names(pseudobulk_expr_by_group), function(expr_mat, groupname) {
  samples <- rownames(expr_mat)
  samples <- substr(samples, 1, regexpr("\\.", samples) - 1)

  # Extract cell type and sex
  group_parts <- strsplit(groupname, "_")[[1]]
  celltype_label <- group_parts[1]
  sex_label <- group_parts[2]

  # Get testosterone values for these samples
  test_vals <- targeted_comparision_filtered %>%
    filter(Sample %in% samples) %>%
    arrange(match(Sample, samples)) %>%
    pull(TT)

  # Ensure alignment
  if (length(test_vals) != length(samples)) {
    warning(glue("Sample mismatch in group {groupname}"))
    return(NULL)
  }

  # Filter for all NA or zero/low variance
  non_na <- !is.na(test_vals)
  if (sum(non_na) < 4 || length(unique(test_vals[non_na])) < 2) {
    message(glue("Skipping {groupname}: insufficient testosterone data"))
    return(NULL)
  }

  # Run regression for each gene
  results <- apply(expr_mat, 2, function(gene_expr) calculate_corr_pval(gene_expr, test_vals))
  cor_values <- sapply(results, `[[`, "cor")
  pval_values <- sapply(results, `[[`, "pval")

  tibble(
    gene = colnames(expr_mat),
    cell_type = celltype_label,
    sex = sex_label,
    cor = cor_values,
    pval = pval_values
  )
})
```



```{r pseudo results}
results_by_group_pseudo <- results_by_group_pseudo %>%
  filter(!is.na(pval))  %>%
  group_by(cell_type, sex) %>%
  mutate(padj = p.adjust(pval, method = "fdr")) %>%
  ungroup()

top_genes_per_sex_pseudo <- results_by_group_pseudo %>%
  filter(padj < 0.05) %>%
  group_by(sex, cell_type) %>%
  arrange(desc(abs(cor))) %>%
  slice_head(n = 5) %>%
  ungroup()

sig_results_males_pseudo <- results_by_group_pseudo %>% filter(sex == "M", padj < 0.05)
sig_results_females_pseudo <- results_by_group_pseudo %>% filter(sex == "F", padj < 0.05)

```



```{r look at results again}
# Function to generate one plot with correlation and p-value
make_scatter_plot <- function(gene, cell_type, sex, pseudobulk_expr_df, test_vals) {
  expr_vals <- pseudobulk_expr_df[, gene]
  data <- tibble(
    Expression = as.numeric(expr_vals),
    Testosterone = test_vals
  ) %>%
    filter(!is.na(Expression) & !is.na(Testosterone) & Expression != 0)
  
  # Calculate correlation and p-value
  valid_idx <- !is.na(data$Expression) & !is.na(data$Testosterone)
  cor_value <- cor(data$Expression[valid_idx], data$Testosterone[valid_idx])
  model <- lm(Expression ~ Testosterone, data = data)
  pval_value <- summary(model)$coefficients[2, 4]

  # Create plot
  plot <- ggplot(data, aes(x = Testosterone, y = Expression)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = glue::glue("{gene} in {cell_type} ({sex})"),
      x = "Testosterone",
      y = "Gene Expression",
      subtitle = glue::glue("Correlation: {round(cor_value, 2)}, p-value: {round(pval_value, 3)}")
    ) +
    theme_minimal()
  
  return(plot)
}

# Plot all top genes per sex using pseudobulk expression data
top_plots <- pmap(
  top_genes_per_sex_pseudo,
  function(gene, cell_type, sex, cor, pval, padj) {
    groupname <- paste(cell_type, sex, sep = "_")
    expr_df <- pseudobulk_expr_by_group[[groupname]]

    # Get sample names and strip suffix like ".RG" to match testosterone data
    samples <- rownames(expr_df)
    samples_clean <- sub("\\..*", "", samples)

    # Get testosterone values in the same order as the expression matrix
    test_vals_df <- targeted_comparision_filtered %>%
      filter(Sample %in% samples_clean) %>%
      mutate(row_id = match(Sample, samples_clean)) %>%
      arrange(row_id)

    test_vals <- test_vals_df$TT

    # Now plot
    make_scatter_plot(gene, cell_type, sex, expr_df, test_vals)
  }
)


top_plots


```


```{r try one more time before giving up}

calculate_corr_pval <- function(expr_vals, test_vals) {
  valid_idx <- !is.na(expr_vals) & !is.na(test_vals)
  expr_vals <- expr_vals[valid_idx]
  test_vals <- test_vals[valid_idx]

  # Early return if insufficient data
  if (length(expr_vals) < 3 || length(unique(test_vals)) < 2 || length(unique(expr_vals)) < 2) {
    return(list(cor = NA, pval = NA))
  }

  # Safe correlation and linear model
  cor_value <- cor(expr_vals, test_vals)
  model <- lm(expr_vals ~ test_vals)
  pval_value <- summary(model)$coefficients[2, 4]

  return(list(cor = cor_value, pval = pval_value))
}

# Function to generate one plot with correlation and p-value
make_scatter_plot <- function(gene, cell_type, sex, pseudobulk_expr_df, test_vals) {
  expr_vals <- pseudobulk_expr_df[, gene]
  data <- tibble(
    Expression = as.numeric(expr_vals),
    Testosterone = test_vals
  ) %>%
    filter(!is.na(Expression) & !is.na(Testosterone) & Expression != 0)
  
  # Calculate correlation and p-value using the helper function
  corr_pval_results <- calculate_corr_pval(data$Expression, data$Testosterone)
  cor_value <- corr_pval_results$cor
  pval_value <- corr_pval_results$pval
  
  # Create plot
  plot <- ggplot(data, aes(x = Testosterone, y = Expression)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = glue::glue("{gene} in {cell_type} ({sex})"),
      x = "Testosterone",
      y = "Gene Expression",
      subtitle = glue::glue("Correlation: {round(cor_value, 2)}, p-value: {round(pval_value, 3)}")
    ) +
    theme_minimal()
  
  return(plot)
}

# Run correlation/regression for each celltype-sex pseudobulk group
results_by_group_pseudo <- map2_dfr(pseudobulk_expr_by_group, names(pseudobulk_expr_by_group), function(expr_mat, groupname) {
  samples <- rownames(expr_mat)
  samples <- substr(samples, 1, regexpr("\\.", samples) - 1)

  # Extract cell type and sex
  group_parts <- strsplit(groupname, "_")[[1]]
  celltype_label <- group_parts[1]
  sex_label <- group_parts[2]

  # Get testosterone values for these samples
  test_vals <- targeted_comparision_filtered %>%
    filter(Sample %in% samples) %>%
    arrange(match(Sample, samples)) %>%
    pull(TT)

  # Ensure alignment
  if (length(test_vals) != length(samples)) {
    warning(glue("Sample mismatch in group {groupname}"))
    return(NULL)
  }

  # Filter for all NA or zero/low variance
  non_na <- !is.na(test_vals)
  if (sum(non_na) < 4 || length(unique(test_vals[non_na])) < 2) {
    message(glue("Skipping {groupname}: insufficient testosterone data"))
    return(NULL)
  }

  # Run regression for each gene (calculate correlation and p-value)
  results <- apply(expr_mat, 2, function(gene_expr) {
    calculate_corr_pval(gene_expr, test_vals)
  })

  # Extract correlation and p-value from the results
  cor_values <- sapply(results, `[[`, "cor")
  pval_values <- sapply(results, `[[`, "pval")

  # Create a tibble to store the results
  tibble(
    gene = colnames(expr_mat),
    cell_type = celltype_label,
    sex = sex_label,
    cor = cor_values,
    pval = pval_values
  )
})

top_plots <- pmap(
  top_genes_per_sex_pseudo,
  function(gene, cell_type, sex, cor, pval, padj) {
    groupname <- paste(cell_type, sex, sep = "_")
    expr_df <- pseudobulk_expr_by_group[[groupname]]

    # Get sample names and strip suffix like ".RG" to match testosterone data
    samples <- rownames(expr_df)
    samples_clean <- sub("\\..*", "", samples)

    # Get testosterone values in the same order as the expression matrix
    test_vals_df <- targeted_comparision_filtered %>%
      filter(Sample %in% samples_clean) %>%
      mutate(row_id = match(Sample, samples_clean)) %>%
      arrange(row_id)

    # Ensure sample alignment (in case of missing values)
    aligned_idx <- match(test_vals_df$Sample, samples_clean)
    expr_df <- expr_df[aligned_idx, , drop = FALSE]

    test_vals <- test_vals_df$TT

    # Now plot
    make_scatter_plot(gene, cell_type, sex, expr_df, test_vals)
  }
)


top_plots
```
