---
title: "Calculate Relationship Between Age and Gonadal Hormone Concentrations"
author: "Emma Jones, Sneha Chaturvedi"
date: "Most recent update 9.23.25"
output: 
  html_document: 
    toc: TRUE
---

The purpose of this analysis is threefold: 1) to find outliers in the distribution of hormone concentrations, 2) to understand the relationship between age (gestational week) and hormone concentration, and 3) to determine if the hormones themselves are correlated with each other. This is important for downstream analyses, to see if GW should be added as a co-variate, and if outliers should be removed.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!require("pacman")) {
  install.packages("pacman")
  library(pacman)
} 
p_load(tidyverse,readxl,stats,ggpubr,svglite,knitr)

svg_path_age <- "../../results/targeted/age_analysis_plots"
svg_path_correlation <- "../../results/targeted/hormone_hormone_correlation_plots"
```

## Load data

```{r load seurat object and excel spreadsheet}
# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx("../../doc/targeted/targeted_hormones.xlsx")

#create new column for estradiol labeled E2 to prevent future coding difficulties with a column label starting with a number
targeted_comparision_raw$`E2 (ng/mL)`<- targeted_comparision_raw$`17B-E2 (ng/mL)`

```

## Remove Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

```{r remove outliers}
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) + 1.5 * IQR(data[[variable]], na.rm = TRUE)
  data %>% filter(.data[[variable]] <= outlier_limit)
}

targeted_comparision_filtered_TT <- detect_outliers(targeted_comparision_raw, "TT (ng/mL)")
targeted_comparision_filtered_E2 <- detect_outliers(targeted_comparision_raw, "E2 (ng/mL)")
targeted_comparision_filtered_P4 <- detect_outliers(targeted_comparision_raw, "P4 (ng/mL)")

# save only samples with observations for all 3, will be used for correlations later
all_obs <- intersect(targeted_comparision_filtered_TT$Sample, targeted_comparision_filtered_E2$Sample)
all_obs <- intersect(all_obs, targeted_comparision_filtered_P4$Sample)

targeted_comparision_filtered <- subset(targeted_comparision_raw, Sample %in% all_obs)
```

# Hormone Correlations
## Testosterone and Estradiol
```{r look at overall hormone relationships - tt and e2}
# look at relationship with tt and e2
ggplot(targeted_comparision_filtered, aes(x = `TT (ng/mL)` , y = `E2 (ng/mL)` , color = GW)) +
  geom_point() + 
  theme_minimal() 
ggsave("TT_E2_scatter_gw.svg", path = svg_path_correlation)

ggplot(targeted_comparision_filtered, aes(x = `TT (ng/mL)`, y = `E2 (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("TT_E2_scatter_sex.svg", path = svg_path_correlation)

# linear model
tt_e2 <- lm(`TT (ng/mL)`~`E2 (ng/mL)`, data=targeted_comparision_filtered) 
summary(tt_e2)

targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]
tt_e2_m <- lm(`TT (ng/mL)`~`E2 (ng/mL)`, data=targeted_comparision_males) 
summary(tt_e2_m) 
```

## Progesterone and Estradiol
```{r look at overall hormone relationships - p4 and e2}
# look at relationship with p4 and e2
ggplot(targeted_comparision_filtered, aes(x = `P4 (ng/mL)`, y = `E2 (ng/mL)`, color = GW)) +
  geom_point() + theme_minimal()
ggsave("P4_E2_scatter_gw.svg", path = svg_path_correlation)

ggplot(targeted_comparision_filtered, aes(x = `P4 (ng/mL)`, y = `E2 (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("P4_E2_scatter_sex.svg", path = svg_path_correlation)

# linear model
p4_e2 <- lm(`P4 (ng/mL)`~`E2 (ng/mL)`, data=targeted_comparision_filtered) 
summary(p4_e2) 

targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
p4_e2_f <- lm(`P4 (ng/mL)`~`E2 (ng/mL)`, data=targeted_comparision_females) 
summary(p4_e2_f)

targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]
p4_e2_m <- lm(`P4 (ng/mL)`~`E2 (ng/mL)`, data=targeted_comparision_males) 
summary(p4_e2_m)
```

## Progesterone and Testosterone
```{r look at overall hormone relationships - p4 and tt}
# look at relationship with p4 and tt
ggplot(targeted_comparision_filtered, aes(x = `P4 (ng/mL)`, y = `TT (ng/mL)`, color = GW)) +
  geom_point() + theme_minimal()
ggsave("P4_TT_scatter_gw.svg", path = svg_path_correlation)

ggplot(targeted_comparision_filtered, aes(x = `P4 (ng/mL)`, y = `TT (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("P4_TT_scatter_gw.svg", path = svg_path_correlation)

# linear model
p4_tt <- lm(`P4 (ng/mL)`~`TT (ng/mL)`, data=targeted_comparision_filtered) 
summary(p4_tt) 

targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
p4_tt_f <- lm(`P4 (ng/mL)`~`TT (ng/mL)`, data=targeted_comparision_females) 
summary(p4_tt_f) 

targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]
p4_tt_m <- lm(`P4 (ng/mL)`~`TT (ng/mL)`, data=targeted_comparision_males) 
summary(p4_tt_m) 
```

# Hormone/Gestational Week Analysis 

## Testosterone
### Combined sex
```{r look at age (gestational week) relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered_TT, aes(x = GW, y = `TT (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("TT_gw.svg", path = svg_path_age)

all_simple <- lm(`TT (ng/mL)`~GW, data=targeted_comparision_filtered_TT) 
summary(all_simple)
```

Not significant with males and females combined, but the graph visually supports  males differ from females.

### Separate Sex
```{r loess - tt}
# LOESS (Locally Estimated Scatterplot Smoothing)
plot(targeted_comparision_males$GW, targeted_comparision_males$`TT (ng/mL)`, main="Loess Fit", xlab="GW", ylab="`TT (ng/mL)`")
loess_fit_m <- loess(`TT (ng/mL)` ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$`TT (ng/mL)`, main="Loess Fit", xlab="GW", ylab="`TT (ng/mL)`")
loess_fit_f <- loess(`TT (ng/mL)` ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = `TT (ng/mL)`, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Testosterone") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()
ggsave("TT_Loess_GW.svg", path = svg_path_age)

```
```{r model using best-fitting equation - tt}
male_model <- lm(`TT (ng/mL)` ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.002309

female_model <- lm(`TT (ng/mL)` ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.7959
```

Re-run with time points with n < 3 removed
```{r loess-subset}
#create subset of data
subset_data <- targeted_comparision_filtered %>%
  group_by(GW) %>%
  filter(n() >= 3) %>%
  ungroup()

subset_data_males <- subset_data[subset_data$Sex == "M",]
subset_data_females <- subset_data[subset_data$Sex == "F",]

# LOESS 
plot(subset_data_males$GW, subset_data_males$`TT (ng/mL)`, main="Loess Fit", xlab="GW", ylab="`TT (ng/mL)`")
loess_fit_m <- loess(`TT (ng/mL)` ~ GW, data = subset_data_males, span = 0.85)
lines(sort(subset_data_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(subset_data_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(subset_data_females$GW, subset_data_females$`TT (ng/mL)`, main="Loess Fit", xlab="GW", ylab="`TT (ng/mL)`")
loess_fit_f <- loess(`TT (ng/mL)` ~ GW, data = subset_data_females, span = 0.85)
lines(sort(subset_data_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(subset_data_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
subset_data_males$sex <- "male"
subset_data_females$sex <- "female"
combined_data <- rbind(subset_data_males, subset_data_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = `TT (ng/mL)`, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Testosterone") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()
ggsave("subset_TT_Loess_GW.svg", path = svg_path_age)

```
```{r model using best-fitting equation - tt subset}
male_model <- lm(`TT (ng/mL)` ~ GW + I(GW^2), data = subset_data_males)
summary(male_model) #p-value: 0.002309

female_model <- lm(`TT (ng/mL)` ~ GW + I(GW^2), data = subset_data_females)
summary(female_model) #p-value: 0.7959
```
# Estrogen
### Combined Sex

```{r look at overall relationship - e2}
# look at relationship with age
ggplot(targeted_comparision_filtered_E2, aes(x = GW, y = `E2 (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("E2_gw.svg", path = svg_path_age)

all_simple <- lm(`E2 (ng/mL)`~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```
### Separate Sex
```{r loess - e2}
# LOESS
plot(targeted_comparision_males$GW, targeted_comparision_males$`E2 (ng/mL)`, main="Loess Fit", xlab="GW", ylab="`E2 (ng/mL)`")
loess_fit_m <- loess(`E2 (ng/mL)` ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$`E2 (ng/mL)`, main="Loess Fit", xlab="GW", ylab="`E2 (ng/mL)`")
loess_fit_f <- loess(`E2 (ng/mL)` ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = `E2 (ng/mL)`, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Estradiol") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()
ggsave("E2_Loess_GW.svg", path = svg_path_age)

```


```{r model using best-fitting equation - e2}
male_model <- lm(`E2 (ng/mL)` ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.2672

female_model <- lm(`E2 (ng/mL)` ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.6758

```


## Progesterone

### Combined Sex 

```{r look at overall relationship - p4}
# look at relationship with age
ggplot(targeted_comparision_filtered_P4, aes(x = GW, y = `P4 (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("P4_gw.svg", path = svg_path_age)

all_simple <- lm(`P4 (ng/mL)`~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```

### Separate Sex
```{r loess - p4}
# loess
plot(targeted_comparision_males$GW, targeted_comparision_males$`P4 (ng/mL)`, main="Loess Fit", xlab="GW", ylab="`P4 (ng/mL)`")
loess_fit_m <- loess(`P4 (ng/mL)` ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$`P4 (ng/mL)`, main="Loess Fit", xlab="GW", ylab="P4")
loess_fit_f <- loess(`P4 (ng/mL)` ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = `P4 (ng/mL)`, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Progesterone") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()
ggsave("P4_Loess_GW.svg", path = svg_path_age)

```


```{r model using best-fitting equation - p4}
male_model <- lm(`P4 (ng/mL)` ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.3699

female_model <- lm(`P4 (ng/mL)` ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.6672

```


