---
title: "Correlation of Hormone Concentration and Gene Expression Data"
author: "Emma Jones"
date: "2025-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of his script is to explore single cell RNA-seq results (and potentially scATAC), and correlate these levels with testosterone, progesterone, and estradiol concentrations from fetal human brain samples.

## Load packages

First, we load packages. We might want more later, but for now I know I need the following

```{r load packages}
library(tidyverse)
library(Seurat)
library(readxl)
library(Signac)

```


## Load data

```{r load seurat object and excel spreadsheet}

# load 10x multiome data
SSD_data_v5 <- readRDS("../data/SSD_data_v5.rds")

# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx("../doc/Targeted Comparison.xlsx")

```

## Explore data

```{r make histograms of hormone concentrations}

hist(targeted_comparision_raw$TT, breaks = 20)

hist(targeted_comparision_raw$`17B-E2`, breaks = 20)

hist(targeted_comparision_raw$P4, breaks = 20)

```

## how does this correlate with age?

```{r ggplot}

ggplot(targeted_comparision_raw, aes(x = GW, y = TT, color = Sex)) +
  geom_point() + theme_minimal()

targeted_comparision_males <- targeted_comparision_raw[targeted_comparision_raw$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_males) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_males) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_males)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_males$GW)

# Base plot - males
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 


# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run for females}
# try again for females
targeted_comparision_females <- targeted_comparision_raw[targeted_comparision_raw$Sex == "F",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_females) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_females) 

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 


# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 

```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_m <- loess(TT ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_f <- loess(TT ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = TT, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex", x = "GW", y = "TT") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

```


```{r model using equation}


male_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.01292

female_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.6589

# Create prediction frame - male
new_male <- data.frame(GW = seq(min(targeted_comparision_males$GW),
                                max(targeted_comparision_males$GW), length.out = 100))
new_male$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_males),
                            newdata = new_male)

# Create prediction frame - female
new_female <- data.frame(GW = seq(min(targeted_comparision_females$GW),
                                max(targeted_comparision_females$GW), length.out = 100))
new_female$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_females),
                            newdata = new_female)

# Plot 1
ggplot(targeted_comparision_males, aes(x = GW, y = TT)) +
  geom_point(color = "dodgerblue", alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Males", x = "GW", y = "TT") +
  theme_minimal()

# Plot 2
ggplot(targeted_comparision_females, aes(x = GW, y = TT)) +
  geom_point(color = "violetred2", alpha = 0.6) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Females", x = "GW", y = "TT") +
  theme_minimal()

# Combine information
ggplot(targeted_comparision_raw, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Males and Females", subtitle = "Male p-value: 0.01292\nFemale p-value:0.6589", x = "GW", y = "TT") +
  theme_minimal()

```

# Work on understanding expression patterns and TT level

If there is a quadratic equation that explains expression over time, it is possible the relationship between TT and gene expression is also nonlinear. Will start with linear first, since that is more straightforward.

```{r Combine metadata and expression matrix and Prepare and join sample-level metadata with TT concentrations}
expr_mat <- as.matrix(GetAssayData(SSD_data_v5[["SCT"]], slot = "data"))  # use SCT-transformed expression

cell_metadata <- SSD_data_v5@meta.data %>%
  select(sample, cluster_call)
cell_metadata$cell_id <- rownames(cell_metadata) 

targeted_comparision_raw$Sample <- paste0("tissue", targeted_comparision_raw$Sample)

cell_metadata <- cell_metadata %>%
  left_join(targeted_comparision_raw, by = c("sample" = "Sample"))

# Rename cell types for clarity
cell_metadata$cell_type <- cell_metadata$cluster_call
```


```{r For each cell type, compute median expression of each gene per sample}
# Split cell metadata by cell_type and Sex
median_expr_by_group <- cell_metadata %>%
  group_by(cell_type, Sex) %>%
  group_split() %>%
  set_names(map_chr(., ~ paste(unique(.x$cell_type), unique(.x$Sex), sep = "_"))) %>%
  map(function(df) {
    samples <- unique(df$sample)
    expr_sub <- expr_mat[, df$cell_id, drop = FALSE]
    median_expr_list <- map(samples, function(sid) {
      cells_in_sample <- df %>% filter(sample == sid) %>% pull(cell_id)
      if (length(cells_in_sample) == 0) {
        return(rep(NA, nrow(expr_sub)))
        }
      matrixStats::rowMedians(expr_sub[, cells_in_sample, drop = FALSE])
})

# Combine to matrix and name columns
median_expr <- do.call(cbind, median_expr_list)
colnames(median_expr) <- as.character(samples)
    rownames(median_expr) <- rownames(expr_sub)
    return(median_expr)
  })

```

```{r remove samples 14 and 16 because there are no testosterone measurements?}
```


```{r For each cell type and gene, correlate or regress with testosterone}
# Regression/correlation per group (cell_type + sex)
results_by_group <- map2_dfr(median_expr_by_group, names(median_expr_by_group), function(expr_df, groupname) {
  samples <- colnames(expr_df)
  sex_label <- strsplit(groupname, "_")[[1]][2]
  celltype_label <- strsplit(groupname, "_")[[1]][1]
  
  test_vals <- targeted_comparision_raw %>%
    filter(Sample %in% samples) %>%
    arrange(match(Sample, samples)) %>%
    pull(TT)
  
  if (sd(test_vals, na.rm = TRUE) == 0 || all(is.na(test_vals))) {
    message(glue::glue("Skipping {groupname}: no variation in testosterone"))
    return(NULL)
  }
  
  tibble(
    gene = rownames(expr_df),
    cell_type = celltype_label,
    Sex = sex_label,
    cor = apply(expr_df, 1, function(gene_expr) {
  if (sd(gene_expr, na.rm = TRUE) == 0) return(NA)
  cor(gene_expr, test_vals, use = "complete.obs")
}),
    pval = apply(expr_df, 1, function(gene_expr) {
  if (sd(gene_expr, na.rm = TRUE) == 0) return(NA)
  model <- lm(gene_expr ~ test_vals)
  summary(model)$coefficients[2, 4]
})
  )
})
```

```{r add fdr correction}

results_by_group <- results_by_group %>%
  group_by(cell_type, Sex) %>%
  mutate(padj = p.adjust(pval, method = "fdr")) %>%
  ungroup()

sig_results_males <- results_by_group %>% filter(Sex == "M", padj < 0.05)
length(unique(sig_results_males$gene))

sig_results_females <- results_by_group %>% filter(Sex == "F", padj < 0.05)
length(unique(sig_results_females$gene))

sig_overlap <- intersect(unique(sig_results_males$gene),unique(sig_results_females$gene))

write_csv(sig_results_males, "../results/sig_TT_results_males.csv")
write_csv(sig_results_females, "../results/sig_TT_results_females.csv")

```





