---
title: "Gonadal Hormone Age Relationships and Inter-Hormone Correlations"
author: "Emma Jones, Sneha Chaturvedi"
date: "Last updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

Outlier detection for each hormone (TT, E2, P4), Pearson correlations between all hormone pairs (combined and sex-stratified), and linear vs. polynomial model comparison of each hormone's relationship with gestational week.

| | |
|---|---|
| **Inputs** | `doc/targeted/targeted_hormones.xlsx` |
| **Outputs** | Figures in `results/targeted/age_analysis_plots/` and `results/targeted/hormone_hormone_correlation_plots/`, model comparison CSVs in `results/targeted/hormone_csv/` |
| **Upstream** | None (reads raw hormone data) |
| **Downstream** | `hormone_sex_comparison.Rmd` |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Packages <- c("tidyverse", "readxl", "stats", "ggpubr", "svglite", "knitr", "gridExtra", "broom")
lapply(Packages, library, character.only=TRUE)

find_project_root <- function(marker = ".git") {
  d <- normalizePath(getwd())
  repeat {
    if (dir.exists(file.path(d, marker))) return(d)
    parent <- dirname(d)
    if (parent == d) stop("Project root not found (no ", marker, " above ", getwd(), ")")
    d <- parent
  }
}

root <- find_project_root()
svg_path_age <- file.path(root, "results/targeted/age_analysis_plots")
svg_path_correlation <- file.path(root, "results/targeted/hormone_hormone_correlation_plots")
csv_path <- file.path(root, "results/targeted/hormone_csv")
dir.create(svg_path_age, recursive = TRUE, showWarnings = FALSE)
dir.create(svg_path_correlation, recursive = TRUE, showWarnings = FALSE)
dir.create(csv_path, recursive = TRUE, showWarnings = FALSE)
```

## Load data

```{r load seurat object and excel spreadsheet}
# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx(file.path(root, "doc/targeted/targeted_hormones.xlsx"))

#create new column for estradiol labeled E2
targeted_comparision_raw$`E2 (ng/mL)`<- targeted_comparision_raw$`17B-E2 (ng/mL)`

```

``` {r gw histogram}
# Count samples per gestational week
sample_counts <- targeted_comparision_raw %>%
  group_by(GW) %>%
  summarise(n_samples = n()) %>%
  ungroup()

# Print sample counts
cat("\nSample counts by gestational week:\n")
print(sample_counts)

# Create histogram
p_histogram <- ggplot(targeted_comparision_raw, aes(x = GW)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Sample Distribution Across Gestational Weeks",
       x = "Gestational Week (GW)",
       y = "Number of Samples") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

# Display plot
print(p_histogram)

# Save plot
ggsave("sample_distribution_GW_histogram.svg", 
       plot = p_histogram, 
       path = svg_path_age,
       width = 8, 
       height = 6)

# Count samples per gestational week and sex
sample_counts_sex <- targeted_comparision_raw %>%
  group_by(GW, Sex) %>%
  summarise(n_samples = n(), .groups = "drop") %>%
  ungroup()

cat("\nSample counts by gestational week and sex:\n")
print(sample_counts_sex)

# Create histogram with sex breakdown
p_histogram_sex <- ggplot(targeted_comparision_raw, aes(x = GW, fill = Sex)) +
  geom_histogram(binwidth = 1, color = "black", alpha = 0.7, position = "stack") +
  labs(title = "Sample Distribution Across Gestational Weeks by Sex",
       x = "Gestational Week (GW)",
       y = "Number of Samples") +
  scale_fill_manual(values = c("M" = "dodgerblue", "F" = "violetred2"),
                    labels = c("M" = "Male", "F" = "Female")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "right")

print(p_histogram_sex)

ggsave("sample_distribution_GW_by_sex_histogram.svg", 
       plot = p_histogram_sex, 
       path = svg_path_age,
       width = 8, 
       height = 6)

```
## Remove Outliers

Q3 + 1.5*IQR 


```{r remove outliers}
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) + 1.5 * IQR(data[[variable]], na.rm = TRUE)
  data %>% filter(.data[[variable]] <= outlier_limit)
}

targeted_comparision_filtered_TT <- detect_outliers(targeted_comparision_raw, "TT (ng/mL)")
targeted_comparision_filtered_E2 <- detect_outliers(targeted_comparision_raw, "E2 (ng/mL)")
targeted_comparision_filtered_P4 <- detect_outliers(targeted_comparision_raw, "P4 (ng/mL)")

# save only samples with observations for all 3, will be used for correlations later
all_obs <- intersect(targeted_comparision_filtered_TT$Sample, targeted_comparision_filtered_E2$Sample)
all_obs <- intersect(all_obs, targeted_comparision_filtered_P4$Sample)

targeted_comparision_filtered <- subset(targeted_comparision_raw, Sample %in% all_obs)
```

# Hormone Correlations
## Testosterone and Estradiol
```{r look at overall hormone relationships - tt and e2}
# look at relationship with tt and e2
ggplot(targeted_comparision_filtered, aes(x = `TT (ng/mL)` , y = `E2 (ng/mL)` , color = GW)) +
  geom_point() + 
  theme_minimal() 
ggsave("TT_E2_scatter_gw.svg", path = svg_path_correlation)
ggplot(targeted_comparision_filtered, aes(x = `TT (ng/mL)`, y = `E2 (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("TT_E2_scatter_sex.svg", path = svg_path_correlation)

# Correlation test
tt_e2 <- cor.test(targeted_comparision_filtered$`TT (ng/mL)`, 
                  targeted_comparision_filtered$`E2 (ng/mL)`, 
                  method = "pearson")
print(tt_e2)

targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]
tt_e2_m <- cor.test(targeted_comparision_males$`TT (ng/mL)`, 
                    targeted_comparision_males$`E2 (ng/mL)`, 
                    method = "pearson")
print(tt_e2_m)

targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
tt_e2_f <- cor.test(targeted_comparision_females$`TT (ng/mL)`, 
                    targeted_comparision_females$`E2 (ng/mL)`, 
                    method = "pearson")
print(tt_e2_f)
```

## Progesterone and Estradiol
```{r look at overall hormone relationships - p4 and e2}
# look at relationship with p4 and e2
ggplot(targeted_comparision_filtered, aes(x = `P4 (ng/mL)`, y = `E2 (ng/mL)`, color = GW)) +
  geom_point() + theme_minimal()
ggsave("P4_E2_scatter_gw.svg", path = svg_path_correlation)
ggplot(targeted_comparision_filtered, aes(x = `P4 (ng/mL)`, y = `E2 (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("P4_E2_scatter_sex.svg", path = svg_path_correlation)

# Correlation test
p4_e2 <- cor.test(targeted_comparision_filtered$`P4 (ng/mL)`, 
                  targeted_comparision_filtered$`E2 (ng/mL)`, 
                  method = "pearson")
print(p4_e2)

targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
p4_e2_f <- cor.test(targeted_comparision_females$`P4 (ng/mL)`, 
                    targeted_comparision_females$`E2 (ng/mL)`, 
                    method = "pearson")
print(p4_e2_f)

targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]
p4_e2_m <- cor.test(targeted_comparision_males$`P4 (ng/mL)`, 
                    targeted_comparision_males$`E2 (ng/mL)`, 
                    method = "pearson")
print(p4_e2_m)
```

## Progesterone and Testosterone
```{r look at overall hormone relationships - p4 and tt}
# look at relationship with p4 and tt
ggplot(targeted_comparision_filtered, aes(x = `P4 (ng/mL)`, y = `TT (ng/mL)`, color = GW)) +
  geom_point() + theme_minimal()
ggsave("P4_TT_scatter_gw.svg", path = svg_path_correlation)
ggplot(targeted_comparision_filtered, aes(x = `P4 (ng/mL)`, y = `TT (ng/mL)`, color = Sex)) +
  geom_point() + theme_minimal()
ggsave("P4_TT_scatter_sex.svg", path = svg_path_correlation)

# Correlation test
p4_tt <- cor.test(targeted_comparision_filtered$`P4 (ng/mL)`, 
                  targeted_comparision_filtered$`TT (ng/mL)`, 
                  method = "pearson")
print(p4_tt)

targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
p4_tt_f <- cor.test(targeted_comparision_females$`P4 (ng/mL)`, 
                    targeted_comparision_females$`TT (ng/mL)`, 
                    method = "pearson")
print(p4_tt_f)

targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]
p4_tt_m <- cor.test(targeted_comparision_males$`P4 (ng/mL)`, 
                    targeted_comparision_males$`TT (ng/mL)`, 
                    method = "pearson")
print(p4_tt_m)
```
## Correlation Summary Table
## Summary Table of Hormone Correlations
```{r hormone correlation summary table}
# Create comprehensive correlation summary table
correlation_summary <- data.frame(
  Comparison = c(
    "TT vs E2", "TT vs E2", "TT vs E2",
    "P4 vs E2", "P4 vs E2", "P4 vs E2",
    "P4 vs TT", "P4 vs TT", "P4 vs TT"
  ),
  Sex = c(
    "Combined", "Male", "Female",
    "Combined", "Female", "Male",
    "Combined", "Female", "Male"
  ),
  N = c(
    nrow(targeted_comparision_filtered),
    nrow(targeted_comparision_males),
    nrow(targeted_comparision_females),
    nrow(targeted_comparision_filtered),
    nrow(targeted_comparision_females),
    nrow(targeted_comparision_males),
    nrow(targeted_comparision_filtered),
    nrow(targeted_comparision_females),
    nrow(targeted_comparision_males)
  ),
  Correlation_r = c(
    tt_e2$estimate,
    tt_e2_m$estimate,
    tt_e2_f$estimate,
    p4_e2$estimate,
    p4_e2_f$estimate,
    p4_e2_m$estimate,
    p4_tt$estimate,
    p4_tt_f$estimate,
    p4_tt_m$estimate
  ),
  CI_Lower = c(
    tt_e2$conf.int[1],
    tt_e2_m$conf.int[1],
    tt_e2_f$conf.int[1],
    p4_e2$conf.int[1],
    p4_e2_f$conf.int[1],
    p4_e2_m$conf.int[1],
    p4_tt$conf.int[1],
    p4_tt_f$conf.int[1],
    p4_tt_m$conf.int[1]
  ),
  CI_Upper = c(
    tt_e2$conf.int[2],
    tt_e2_m$conf.int[2],
    tt_e2_f$conf.int[2],
    p4_e2$conf.int[2],
    p4_e2_f$conf.int[2],
    p4_e2_m$conf.int[2],
    p4_tt$conf.int[2],
    p4_tt_f$conf.int[2],
    p4_tt_m$conf.int[2]
  ),
  P_Value = c(
    tt_e2$p.value,
    tt_e2_m$p.value,
    tt_e2_f$p.value,
    p4_e2$p.value,
    p4_e2_f$p.value,
    p4_e2_m$p.value,
    p4_tt$p.value,
    p4_tt_f$p.value,
    p4_tt_m$p.value
  ),
  Significant = c(
    tt_e2$p.value < 0.05,
    tt_e2_m$p.value < 0.05,
    tt_e2_f$p.value < 0.05,
    p4_e2$p.value < 0.05,
    p4_e2_f$p.value < 0.05,
    p4_e2_m$p.value < 0.05,
    p4_tt$p.value < 0.05,
    p4_tt_f$p.value < 0.05,
    p4_tt_m$p.value < 0.05
  )
)

# Round numeric columns for readability
correlation_summary$Correlation_r <- round(correlation_summary$Correlation_r, 4)
correlation_summary$CI_Lower <- round(correlation_summary$CI_Lower, 4)
correlation_summary$CI_Upper <- round(correlation_summary$CI_Upper, 4)
correlation_summary$P_Value <- formatC(correlation_summary$P_Value, format = "e", digits = 3)

# Print the table
print(correlation_summary)

# Save as CSV
write.csv(correlation_summary, file.path(csv_path, "hormone_correlation_summary.csv"), row.names = FALSE)

# Optional: Create a formatted table using knitr/kable if you're in RMarkdown
library(knitr)
kable(correlation_summary, 
      caption = "Summary of Hormone Correlation Analyses (Pearson)",
      align = c('l', 'l', 'r', 'r', 'r', 'r', 'r', 'c'))
```
# Hormone/Gestational Week Analysis 

Functions
``` {r age_analysis_functions}
# Function to filter data keeping only timepoints with n >= 3
filter_by_n <- function(data, min_n = 3) {
  data %>%
    group_by(GW) %>%
    filter(n() >= min_n) %>%
    ungroup()
}

# Function to create model comparison table
create_model_table <- function(linear_model, poly_model, model_name) {
  linear_stats <- glance(linear_model)
  poly_stats <- glance(poly_model)
  
  data.frame(
    Model = c(paste(model_name, "- Linear"), paste(model_name, "- Polynomial")),
    R_squared = c(linear_stats$r.squared, poly_stats$r.squared),
    Adj_R_squared = c(linear_stats$adj.r.squared, poly_stats$adj.r.squared),
    P_value = c(linear_stats$p.value, poly_stats$p.value),
    AIC = c(AIC(linear_model), AIC(poly_model))
  )
}


```
## Testosterone
### Combined sex
```{r look at age (gestational week) relationship}
cat("\n=== TESTOSTERONE ANALYSIS ===\n")

# Filter data (n >= 3 per timepoint)
TT_filtered <- filter_by_n(targeted_comparision_filtered_TT, min_n = 3)
TT_males <- TT_filtered %>% filter(Sex == "M")
TT_females <- TT_filtered %>% filter(Sex == "F")

# --- COMBINED SEX ANALYSIS ---
cat("\n--- Testosterone: Combined Sex ---\n")

# LOESS plot - Combined
p_tt_loess_combined <- ggplot(TT_filtered, aes(x = GW, y = `TT (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE) +
  labs(title = "Testosterone: LOESS Fit (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Testosterone (ng/mL)") +
  theme_minimal()
ggsave("TT_LOESS_combined.svg", plot = p_tt_loess_combined, path = svg_path_age)

# Linear model - Combined
tt_linear_combined <- lm(`TT (ng/mL)` ~ GW, data = TT_filtered)
cat("\nLinear Model Summary:\n")
print(summary(tt_linear_combined))

p_tt_linear_combined <- ggplot(TT_filtered, aes(x = GW, y = `TT (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "Testosterone: Linear Model (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Testosterone (ng/mL)") +
  theme_minimal()
ggsave("TT_Linear_combined.svg", plot = p_tt_linear_combined, path = svg_path_age)

# Polynomial model - Combined
tt_poly_combined <- lm(`TT (ng/mL)` ~ GW + I(GW^2), data = TT_filtered)
cat("\nPolynomial Model Summary:\n")
print(summary(tt_poly_combined))

TT_filtered$pred_poly <- predict(tt_poly_combined)
p_tt_poly_combined <- ggplot(TT_filtered, aes(x = GW, y = `TT (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_poly), color = "black", size = 1) +
  labs(title = "Testosterone: Polynomial Model (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Testosterone (ng/mL)") +
  theme_minimal()
ggsave("TT_Polynomial_combined.svg", plot = p_tt_poly_combined, path = svg_path_age)

# Model comparison table
tt_combined_table <- create_model_table(tt_linear_combined, tt_poly_combined, "TT Combined")
print(tt_combined_table)
write.csv(tt_combined_table, file.path(csv_path, "TT_combined_model_comparison.csv"), row.names = FALSE)
```


### Separate Sex
```{r loess - tt}
cat("\n--- Testosterone: Sex-Separate ---\n")

# Prepare combined data for plotting
TT_males$sex <- "Male"
TT_females$sex <- "Female"
TT_sex_combined <- rbind(TT_males, TT_females)

# LOESS plot - Sex separate
p_tt_loess_sep <- ggplot(TT_sex_combined, aes(x = GW, y = `TT (ng/mL)`, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "Testosterone: LOESS Fit by Sex", 
       x = "Gestational Week (GW)", 
       y = "Testosterone (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  scale_fill_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("TT_LOESS_by_sex.svg", plot = p_tt_loess_sep, path = svg_path_age)

# Linear models - Males
tt_linear_male <- lm(`TT (ng/mL)` ~ GW, data = TT_males)
cat("\nMales - Linear Model Summary:\n")
print(summary(tt_linear_male))

# Linear models - Females
tt_linear_female <- lm(`TT (ng/mL)` ~ GW, data = TT_females)
cat("\nFemales - Linear Model Summary:\n")
print(summary(tt_linear_female))

# Linear plot - Sex separate
TT_males$pred_linear <- predict(tt_linear_male)
TT_females$pred_linear <- predict(tt_linear_female)
TT_sex_combined_linear <- rbind(TT_males, TT_females)

p_tt_linear_sep <- ggplot(TT_sex_combined_linear, aes(x = GW, y = `TT (ng/mL)`, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_linear), size = 1) +
  labs(title = "Testosterone: Linear Model by Sex", 
       x = "Gestational Week (GW)", 
       y = "Testosterone (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("TT_Linear_by_sex.svg", plot = p_tt_linear_sep, path = svg_path_age)

# Polynomial models - Males
tt_poly_male <- lm(`TT (ng/mL)` ~ GW + I(GW^2), data = TT_males)
cat("\nMales - Polynomial Model Summary:\n")
print(summary(tt_poly_male))

# Polynomial models - Females
tt_poly_female <- lm(`TT (ng/mL)` ~ GW + I(GW^2), data = TT_females)
cat("\nFemales - Polynomial Model Summary:\n")
print(summary(tt_poly_female))

# Polynomial plot - Sex separate
TT_males$pred_poly <- predict(tt_poly_male)
TT_females$pred_poly <- predict(tt_poly_female)
TT_sex_combined_poly <- rbind(TT_males, TT_females)

p_tt_poly_sep <- ggplot(TT_sex_combined_poly, aes(x = GW, y = `TT (ng/mL)`, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_poly), size = 1) +
  labs(title = "Testosterone: Polynomial Model by Sex", 
       x = "Gestational Week (GW)", 
       y = "Testosterone (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("TT_Polynomial_by_sex.svg", plot = p_tt_poly_sep, path = svg_path_age)

# Model comparison tables - Sex separate
tt_male_table <- create_model_table(tt_linear_male, tt_poly_male, "TT Male")
tt_female_table <- create_model_table(tt_linear_female, tt_poly_female, "TT Female")
tt_sex_table <- rbind(tt_male_table, tt_female_table)
print(tt_sex_table)
write.csv(tt_sex_table, file.path(csv_path, "TT_sex_model_comparison.csv"), row.names = FALSE)
```

# Estrogen
### Combined Sex

```{r look at overall relationship - e2}
cat("\n=== ESTRADIOL ANALYSIS ===\n")

# Filter data (n >= 3 per timepoint)
E2_filtered <- filter_by_n(targeted_comparision_filtered_E2, min_n = 3)
E2_males <- E2_filtered %>% filter(Sex == "M")
E2_females <- E2_filtered %>% filter(Sex == "F")

# --- COMBINED SEX ANALYSIS ---
cat("\n--- Estradiol: Combined Sex ---\n")

# LOESS plot - Combined
p_e2_loess_combined <- ggplot(E2_filtered, aes(x = GW, y = `E2 (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE) +
  labs(title = "Estradiol: LOESS Fit (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Estradiol (ng/mL)") +
  theme_minimal()
ggsave("E2_LOESS_combined.svg", plot = p_e2_loess_combined, path = svg_path_age)

# Linear model - Combined
e2_linear_combined <- lm(`E2 (ng/mL)` ~ GW, data = E2_filtered)
cat("\nLinear Model Summary:\n")
print(summary(e2_linear_combined))

p_e2_linear_combined <- ggplot(E2_filtered, aes(x = GW, y = `E2 (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "Estradiol: Linear Model (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Estradiol (ng/mL)") +
  theme_minimal()
ggsave("E2_Linear_combined.svg", plot = p_e2_linear_combined, path = svg_path_age)

# Polynomial model - Combined
e2_poly_combined <- lm(`E2 (ng/mL)` ~ GW + I(GW^2), data = E2_filtered)
cat("\nPolynomial Model Summary:\n")
print(summary(e2_poly_combined))

E2_filtered$pred_poly <- predict(e2_poly_combined)
p_e2_poly_combined <- ggplot(E2_filtered, aes(x = GW, y = `E2 (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_poly), color = "black", size = 1) +
  labs(title = "Estradiol: Polynomial Model (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Estradiol (ng/mL)") +
  theme_minimal()
ggsave("E2_Polynomial_combined.svg", plot = p_e2_poly_combined, path = svg_path_age)

# Model comparison table
e2_combined_table <- create_model_table(e2_linear_combined, e2_poly_combined, "E2 Combined")
print(e2_combined_table)
write.csv(e2_combined_table, file.path(csv_path, "E2_combined_model_comparison.csv"), row.names = FALSE)

```
### Separate Sex
```{r loess - e2}
cat("\n--- Estradiol: Sex-Separate ---\n")

# Prepare combined data for plotting
E2_males$sex <- "Male"
E2_females$sex <- "Female"
E2_sex_combined <- rbind(E2_males, E2_females)

# LOESS plot - Sex separate
p_e2_loess_sep <- ggplot(E2_sex_combined, aes(x = GW, y = `E2 (ng/mL)`, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "Estradiol: LOESS Fit by Sex", 
       x = "Gestational Week (GW)", 
       y = "Estradiol (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  scale_fill_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("E2_LOESS_by_sex.svg", plot = p_e2_loess_sep, path = svg_path_age)

# Linear models - Males
e2_linear_male <- lm(`E2 (ng/mL)` ~ GW, data = E2_males)
cat("\nMales - Linear Model Summary:\n")
print(summary(e2_linear_male))

# Linear models - Females
e2_linear_female <- lm(`E2 (ng/mL)` ~ GW, data = E2_females)
cat("\nFemales - Linear Model Summary:\n")
print(summary(e2_linear_female))

# Linear plot - Sex separate
E2_males$pred_linear <- predict(e2_linear_male)
E2_females$pred_linear <- predict(e2_linear_female)
E2_sex_combined_linear <- rbind(E2_males, E2_females)

p_e2_linear_sep <- ggplot(E2_sex_combined_linear, aes(x = GW, y = `E2 (ng/mL)`, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_linear), size = 1) +
  labs(title = "Estradiol: Linear Model by Sex", 
       x = "Gestational Week (GW)", 
       y = "Estradiol (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("E2_Linear_by_sex.svg", plot = p_e2_linear_sep, path = svg_path_age)

# Polynomial models - Males
e2_poly_male <- lm(`E2 (ng/mL)` ~ GW + I(GW^2), data = E2_males)
cat("\nMales - Polynomial Model Summary:\n")
print(summary(e2_poly_male))

# Polynomial models - Females
e2_poly_female <- lm(`E2 (ng/mL)` ~ GW + I(GW^2), data = E2_females)
cat("\nFemales - Polynomial Model Summary:\n")
print(summary(e2_poly_female))

# Polynomial plot - Sex separate
E2_males$pred_poly <- predict(e2_poly_male)
E2_females$pred_poly <- predict(e2_poly_female)
E2_sex_combined_poly <- rbind(E2_males, E2_females)

p_e2_poly_sep <- ggplot(E2_sex_combined_poly, aes(x = GW, y = `E2 (ng/mL)`, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_poly), size = 1) +
  labs(title = "Estradiol: Polynomial Model by Sex", 
       x = "Gestational Week (GW)", 
       y = "Estradiol (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("E2_Polynomial_by_sex.svg", plot = p_e2_poly_sep, path = svg_path_age)

# Model comparison tables - Sex separate
e2_male_table <- create_model_table(e2_linear_male, e2_poly_male, "E2 Male")
e2_female_table <- create_model_table(e2_linear_female, e2_poly_female, "E2 Female")
e2_sex_table <- rbind(e2_male_table, e2_female_table)
print(e2_sex_table)
write.csv(e2_sex_table, file.path(csv_path, "E2_sex_model_comparison.csv"), row.names = FALSE)
```

# Progesterone

## Combined Sex 

```{r look at overall relationship - p4}
cat("\n=== PROGESTERONE ANALYSIS ===\n")

# Filter data (n >= 3 per timepoint)
P4_filtered <- filter_by_n(targeted_comparision_filtered_P4, min_n = 3)
P4_males <- P4_filtered %>% filter(Sex == "M")
P4_females <- P4_filtered %>% filter(Sex == "F")

# --- COMBINED SEX ANALYSIS ---
cat("\n--- Progesterone: Combined Sex ---\n")

# LOESS plot - Combined
p_p4_loess_combined <- ggplot(P4_filtered, aes(x = GW, y = `P4 (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE) +
  labs(title = "Progesterone: LOESS Fit (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Progesterone (ng/mL)") +
  theme_minimal()
ggsave("P4_LOESS_combined.svg", plot = p_p4_loess_combined, path = svg_path_age)

# Linear model - Combined
p4_linear_combined <- lm(`P4 (ng/mL)` ~ GW, data = P4_filtered)
cat("\nLinear Model Summary:\n")
print(summary(p4_linear_combined))

p_p4_linear_combined <- ggplot(P4_filtered, aes(x = GW, y = `P4 (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "Progesterone: Linear Model (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Progesterone (ng/mL)") +
  theme_minimal()
ggsave("P4_Linear_combined.svg", plot = p_p4_linear_combined, path = svg_path_age)

# Polynomial model - Combined
p4_poly_combined <- lm(`P4 (ng/mL)` ~ GW + I(GW^2), data = P4_filtered)
cat("\nPolynomial Model Summary:\n")
print(summary(p4_poly_combined))

P4_filtered$pred_poly <- predict(p4_poly_combined)
p_p4_poly_combined <- ggplot(P4_filtered, aes(x = GW, y = `P4 (ng/mL)`, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_poly), color = "black", size = 1) +
  labs(title = "Progesterone: Polynomial Model (Combined)", 
       x = "Gestational Week (GW)", 
       y = "Progesterone (ng/mL)") +
  theme_minimal()
ggsave("P4_Polynomial_combined.svg", plot = p_p4_poly_combined, path = svg_path_age)

# Model comparison table
p4_combined_table <- create_model_table(p4_linear_combined, p4_poly_combined, "P4 Combined")
print(p4_combined_table)
write.csv(p4_combined_table, file.path(csv_path, "P4_combined_model_comparison.csv"), row.names = FALSE)

```

## Separate Sex
```{r loess - p4}
cat("\n--- Progesterone: Sex-Separate ---\n")

# Prepare combined data for plotting
P4_males$sex <- "Male"
P4_females$sex <- "Female"
P4_sex_combined <- rbind(P4_males, P4_females)

# LOESS plot - Sex separate
p_p4_loess_sep <- ggplot(P4_sex_combined, aes(x = GW, y = `P4 (ng/mL)`, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "Progesterone: LOESS Fit by Sex", 
       x = "Gestational Week (GW)", 
       y = "Progesterone (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  scale_fill_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("P4_LOESS_by_sex.svg", plot = p_p4_loess_sep, path = svg_path_age)

# Linear models - Males
p4_linear_male <- lm(`P4 (ng/mL)` ~ GW, data = P4_males)
cat("\nMales - Linear Model Summary:\n")
print(summary(p4_linear_male))

# Linear models - Females
p4_linear_female <- lm(`P4 (ng/mL)` ~ GW, data = P4_females)
cat("\nFemales - Linear Model Summary:\n")
print(summary(p4_linear_female))

# Linear plot - Sex separate
P4_males$pred_linear <- predict(p4_linear_male)
P4_females$pred_linear <- predict(p4_linear_female)
P4_sex_combined_linear <- rbind(P4_males, P4_females)

p_p4_linear_sep <- ggplot(P4_sex_combined_linear, aes(x = GW, y = `P4 (ng/mL)`, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_linear), size = 1) +
  labs(title = "Progesterone: Linear Model by Sex", 
       x = "Gestational Week (GW)", 
       y = "Progesterone (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("P4_Linear_by_sex.svg", plot = p_p4_linear_sep, path = svg_path_age)

# Polynomial models - Males
p4_poly_male <- lm(`P4 (ng/mL)` ~ GW + I(GW^2), data = P4_males)
cat("\nMales - Polynomial Model Summary:\n")
print(summary(p4_poly_male))

# Polynomial models - Females
p4_poly_female <- lm(`P4 (ng/mL)` ~ GW + I(GW^2), data = P4_females)
cat("\nFemales - Polynomial Model Summary:\n")
print(summary(p4_poly_female))

# Polynomial plot - Sex separate
P4_males$pred_poly <- predict(p4_poly_male)
P4_females$pred_poly <- predict(p4_poly_female)
P4_sex_combined_poly <- rbind(P4_males, P4_females)

p_p4_poly_sep <- ggplot(P4_sex_combined_poly, aes(x = GW, y = `P4 (ng/mL)`, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_line(aes(y = pred_poly), size = 1) +
  labs(title = "Progesterone: Polynomial Model by Sex", 
       x = "Gestational Week (GW)", 
       y = "Progesterone (ng/mL)") +
  scale_color_manual(values = c("Male" = "dodgerblue", "Female" = "violetred2")) +
  theme_minimal()
ggsave("P4_Polynomial_by_sex.svg", plot = p_p4_poly_sep, path = svg_path_age)

# Model comparison tables - Sex separate
p4_male_table <- create_model_table(p4_linear_male, p4_poly_male, "P4 Male")
p4_female_table <- create_model_table(p4_linear_female, p4_poly_female, "P4 Female")
p4_sex_table <- rbind(p4_male_table, p4_female_table)
print(p4_sex_table)
write.csv(p4_sex_table, file.path(csv_path, "P4_sex_model_comparison.csv"), row.names = FALSE)

```

``` {r summary_age}

cat("\n=== CREATING SUMMARY TABLE ===\n")

# Combined all model comparison tables
all_models_summary <- rbind(
  tt_combined_table,
  tt_sex_table,
  e2_combined_table,
  e2_sex_table,
  p4_combined_table,
  p4_sex_table
)

print(all_models_summary)
write.csv(all_models_summary, file.path(csv_path, "ALL_hormones_model_comparison.csv"), row.names = FALSE)
```
# AI Assistance Disclosure

Code in this script was developed with assistance from Claude (Anthropic). All
AI-generated code was reviewed, validated, and adapted by the author.

# Session Info

```{r session-info}
sessionInfo()
```
