---
title: "Correlation of Hormone Concentration and Gene Expression Data"
author: "Emma Jones"
date: "2025-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of his script is to explore single cell RNA-seq results (and potentially scATAC), and correlate these levels with testosterone, progesterone, and estradiol concentrations from fetal human brain samples.

## Load packages

First, we load packages. We might want more later, but for now I know I need the following

```{r load packages}
library(tidyverse)
library(Seurat)
library(readxl)
library(Signac)
library(glue)
library(cowplot)
library(edgeR)
library(limma)
library(patchwork)
```

## Load data

```{r load seurat object and excel spreadsheet}
# load 10x multiome data
SSD_data_v5 <- readRDS("../../data/SSD_data_v5.rds")
# remove doublets
SSD_data_v5 <- subset(SSD_data_v5, subset = Droplet == "SNG")

# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx("../../doc/targeted/Targeted Comparison.xlsx")

```

## Remove Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

```{r detect and remove female outlier}
#filter for only females
targeted_comparision_females <- targeted_comparision_raw[targeted_comparision_raw$Sex == "F",]

outlier_limit <- quantile(targeted_comparision_raw$TT)[4] + 1.5*IQR(targeted_comparision_raw$TT)

max(targeted_comparision_females$TT) > outlier_limit 

targeted_comparision_females$TT > outlier_limit
```

So, yes, this is above the typical outlier limit so we can remove from analysis. It is sample number 38, which has an XX Xist genotype and sex_genome_studio asignment of XX as well, so it seems to be a true female sample and not a swap.

```{r remove outlier}
targeted_comparision_filtered <- targeted_comparision_raw[-34,]
```

# Pseudobulk: understanding the realtionship between expression patterns and hormone/TT level

If there is a quadratic equation that explains expression over time, it is possible the relationship between TT and gene expression is also nonlinear. Will start with linear first, since that is more straightforward.

```{r Combine metadata and expression matrix and Prepare and join sample-level metadata with TT concentrations}
cell_metadata <- SSD_data_v5@meta.data %>%
  select(sample, cluster_call)
cell_metadata$cell_id <- rownames(cell_metadata) 

targeted_comparision_filtered$Sample <- paste0("tissue", targeted_comparision_filtered$Sample)

cell_metadata <- cell_metadata %>%
  left_join(targeted_comparision_filtered, by = c("sample" = "Sample"))

# Rename cell types for clarity
cell_metadata$cell_type <- cell_metadata$cluster_call

# remove rows where testosterone is not measured, commented out because this removes samples from estrogen analysis
#cell_metadata <- cell_metadata[!is.na(cell_metadata$TT),]
```

# prepare function

```{r psuedobulk data function}
pseudobulk_data <- function(seurat_obj, assay = "RNA", layer = "counts", 
                            sample_col = "sample_id", celltype_col = "cell_type") {
  # Get raw counts
  counts <- GetAssayData(seurat_obj, assay = assay, layer = layer)
  meta <- seurat_obj@meta.data
  
  # Group cells by sample and cell type
  group_ids <- interaction(meta[[sample_col]], meta[[celltype_col]], drop = TRUE)
  group_levels <- levels(group_ids)
  
  # Initialize list to store aggregated matrices
  aggregated_list <- list()
  metadata_list <- list()
  
  for (gid in group_levels) {
    # Get cells for this group
    group_cells <- which(group_ids == gid)
    if (length(group_cells) < 1) next
    
    # Sum counts across cells
    group_counts <- Matrix::rowSums(counts[, group_cells, drop = FALSE])
    
    # Store as data frame row
    aggregated_list[[gid]] <- group_counts
    metadata_list[[gid]] <- data.frame(
      sample_id = meta[[sample_col]][group_cells[1]],
      cell_type = meta[[celltype_col]][group_cells[1]],
      cell_count = length(group_cells),
      row.names = gid
    )
  }
  
  # Combine into matrices and metadata frame
  pseudobulk_mat <- do.call(rbind, aggregated_list)
  metadata_df <- do.call(rbind, metadata_list)
  
  # Log-transform
  pseudobulk_mat_log <- log1p(pseudobulk_mat)
  
  return(list(counts = pseudobulk_mat, logcounts = pseudobulk_mat_log, metadata = metadata_df))
}
```


```{r pseudobulk}
# Run pseudobulk
pb <- pseudobulk_data(SSD_data_v5, sample_col = "sample", celltype_col = "cluster_call")

# View structure
str(pb)
```

```{r prepare metadata}
# Convert rownames to a column so they survive the join
pb_metadata <- pb$metadata %>%
  rownames_to_column("pseudobulk_id")

# Join with testosterone/sex data
metadata_merged <- pb_metadata %>%
  left_join(targeted_comparision_filtered %>% select(Sample, Sex), 
            by = c("sample_id" = "Sample")) %>%
  filter(!is.na(Sex)) %>%
  mutate(group = paste(cell_type, Sex, sep = "_")) %>%
  column_to_rownames("pseudobulk_id")

# Filter out samples with missing sex or too few cells
metadata_merged <- metadata_merged %>%
  filter(!is.na(Sex)) %>%
  filter(cell_count >= 5)  # Remove sparse pseudobulks

# Subset logcounts and counts to only include valid samples
pb$logcounts <- pb$logcounts[rownames(pb$logcounts) %in% rownames(metadata_merged), , drop = FALSE]
pb$counts <- pb$counts[rownames(pb$counts) %in% rownames(metadata_merged), , drop = FALSE]

# Reorder metadata to match logcounts and counts
metadata_merged <- metadata_merged[match(rownames(pb$logcounts), rownames(metadata_merged)), ]
metadata_merged <- metadata_merged[match(rownames(pb$counts), rownames(metadata_merged)), ]
```

```{r split pseduobulk}
# Split into pseudobulk expression matrices by cell type + sex
pseudobulk_expr_by_group <- split(seq_len(nrow(pb$counts)), metadata_merged$group) %>%
  purrr::map(~ {
    mat <- pb$counts[.x, , drop = FALSE]
    
    # Filter genes: retain only those with at least three samples with count >= 3
    keep_genes <- which(colSums(mat >= 3) >= 3)
    mat[, keep_genes, drop = FALSE]
  })
```

```{r prepare to run limma}
metadata_merged <- left_join(metadata_merged, targeted_comparision_filtered, by = c("sample_id" = "Sample", "Sex" = "Sex"))

# Create unique rownames by combining sample_id and cell_type
metadata_merged <- metadata_merged %>%
  mutate(pseudobulk_id = paste(sample_id, cell_type, sep = "_")) %>%
  column_to_rownames("pseudobulk_id")

# Update pb$counts and pb$logcounts rownames accordingly
rownames(pb$counts) <- rownames(metadata_merged)
rownames(pb$logcounts) <- rownames(metadata_merged)

# Confirm alignment
stopifnot(identical(rownames(pb$counts), rownames(metadata_merged)))
```

## Run limma to model everything, start with testosterone

```{r run limma on TT}
run_voom_lm <- function(counts, metadata) {
  # Transpose counts: genes as rows, samples as columns
  counts_t <- t(counts)

  # Check dimensions
  stopifnot(ncol(counts_t) == nrow(metadata))

  # Create design matrix
  design <- model.matrix(~ TT + cell_count + GW, data = metadata)

  # Run voom
  v <- limma::voom(counts_t, design)

  # Fit model
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)

  # Tidy results
  results <- limma::topTable(fit, coef = "TT", number = Inf, sort.by = "none") %>%
    tibble::rownames_to_column("gene")

  return(results)
}

# Ensure sample order matches
stopifnot(identical(rownames(pb$counts), rownames(metadata_merged)))

# Split by both cell type and sex
split_indices <- split(seq_len(nrow(pb$counts)), interaction(metadata_merged$cell_type, metadata_merged$Sex, drop = TRUE))
```

```{r run limma for each group}
# Run model per group
results_by_group <- purrr::map(split(seq_len(nrow(pb$counts)), metadata_merged$group), function(idx) {
  counts_sub <- pb$counts[idx, , drop = FALSE]
  metadata_sub <- metadata_merged[idx, , drop = FALSE]

  # Remove samples with zero library size
  keep_samples <- rowSums(counts_sub) > 0
  counts_sub <- counts_sub[keep_samples, , drop = FALSE]
  metadata_sub <- metadata_sub[keep_samples, , drop = FALSE]

  # Skip group if too few samples remain
  if (nrow(metadata_sub) < 5) {
    message("Skipping group ", unique(metadata_sub$group), " due to too few samples.")
    return(NULL)
  }

  # Optional: filter genes with low expression across samples
  keep_genes <- colSums(counts_sub >= 2) >= 3
  counts_sub <- counts_sub[, keep_genes, drop = FALSE]

  if (ncol(counts_sub) == 0) {
    message("Skipping group ", unique(metadata_sub$group), " due to no genes passing filter.")
    return(NULL)
  }

  # Run voom + lm with error handling
  tryCatch({
    result <- run_voom_lm(counts_sub, metadata_sub)
    result$cell_type <- metadata_sub$cell_type[1]
    result$sex <- metadata_sub$Sex[1]
    result
  }, error = function(e) {
    message("Skipping group ", unique(metadata_sub$group), " due to error: ", e$message)
    NULL
  })
})
```

```{r examine limma results}
# Combine all results (remove NULLs)
results_all <- bind_rows(results_by_group[!sapply(results_by_group, is.null)])

# examine female results
results_female <- results_all %>%
  filter(sex == "F")

write_csv(results_female, "../../results/targeted/res_TT_genes_female.csv")

sig_counts_female <- results_female %>%
  filter(adj.P.Val < 0.05) %>%
  count(cell_type, name = "num_sig_genes") %>%
  arrange(desc(num_sig_genes))

ggplot(sig_counts_female, aes(x = cell_type, y = num_sig_genes)) +
  geom_bar(stat = "identity", fill = "violetred3") +
  geom_text(aes(label = num_sig_genes), vjust = -0.5) +
  theme_minimal() +
  labs(x = "Cell Type", y = "Number of Significant Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../results/targeted/sig_TT_genes_female.png")

# examine male results
results_male <- results_all %>%
  filter(sex == "M")

write_csv(results_male, "../../results/targeted/res_TT_genes_male.csv")

sig_counts_male <- results_male %>%
  filter(adj.P.Val < 0.05) %>%
  count(cell_type, name = "num_sig_genes") %>%
  arrange(desc(num_sig_genes))

ggplot(sig_counts_male, aes(x = cell_type, y = num_sig_genes)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = num_sig_genes), vjust = -0.5) +
  theme_minimal() +
  labs(x = "Cell Type", y = "Number of Significant Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../results/targeted/sig_TT_genes_male.png")

# Remove genes (columns) with zero total count
pb_counts_filtered <- pb$counts[, colSums(pb$counts) > 0]

# make log(cpm) matrix for plotting
log_cpm_mat <- cpm(pb_counts_filtered, log = TRUE, prior.count = 1)
```
## plot top TT genes for male

```{r make and save male plots}
# Get unique male cell types from results
cell_types <- unique(results_male$cell_type)

# Directory to save plots
output_dir <- "../../results/targeted/pseudobulk_plots_testosterone"
dir.create(output_dir, showWarnings = FALSE)

# Loop through groups and save combined plots
walk(cell_types, function(group_of_interest) {
  
  # === Filter for this group ===
  results_subset <- results_male %>%
    filter(cell_type == group_of_interest) %>%
    arrange(-abs(t)) %>%
    slice_head(n = 6)
  
  # === Subset expression and metadata ===
  samples_in_group <- metadata_merged %>%
    filter(cell_type == group_of_interest)
  
  if (nrow(samples_in_group) == 0) return()
  counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]
  
  # === Plot top genes ===
  plot_list <- map(results_subset$gene, function(gene) {
    if (!gene %in% colnames(counts_in_group)) return(NULL)
    
    expr_vals <- counts_in_group[, gene]
    if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
    
    df <- samples_in_group %>%
      mutate(expr = expr_vals) %>%
      filter(Sex == "M")
    
    ggplot(df, aes(x = TT, y = expr)) +
      geom_point() +
      geom_smooth(method = "lm", se = TRUE, color = "steelblue", fill = "steelblue") +
      labs(
        title = paste0(gene, " in ", group_of_interest, ": Males"),
        x = "Testosterone",
        y = "LogCPM"
      ) +
      theme_minimal()
  }) %>% compact()  # remove NULLs
  
  if (length(plot_list) == 0) return()
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
  
  # Clean group name for filename
clean_name <- gsub("/", "-", group_of_interest)
  
  # Save to PNG
  ggsave(
    filename = file.path(output_dir, paste0(clean_name, "_male_top_genes.png")),
    plot = combined_plot,
    width = 10,
    height = 6 + ceiling(length(plot_list) / 2),
    dpi = 300
  )
})

```

## repeat plotting top TT genes for female

```{r make and save female plots}
# Get unique male cell types from results
cell_types <- unique(results_female$cell_type)

# Directory to save plots
output_dir <- "../../results/targeted/pseudobulk_plots_testosterone"
dir.create(output_dir, showWarnings = FALSE)

# Loop through groups and save combined plots
walk(cell_types, function(group_of_interest) {
  
  # === Filter for this group ===
  results_subset <- results_female %>%
    filter(cell_type == group_of_interest) %>%
    arrange(-abs(t)) %>%
    slice_head(n = 6)
  
  # === Subset expression and metadata ===
  samples_in_group <- metadata_merged %>%
    filter(cell_type == group_of_interest)
  
  if (nrow(samples_in_group) == 0) return()
  counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]

  
  # === Plot top genes ===
  plot_list <- map(results_subset$gene, function(gene) {
    if (!gene %in% colnames(counts_in_group)) return(NULL)
    
    expr_vals <- counts_in_group[, gene]
    if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
    
    df <- samples_in_group %>%
      mutate(expr = expr_vals) %>%
      filter(Sex == "F")
    
    ggplot(df, aes(x = TT, y = expr)) +
      geom_point() +
      geom_smooth(method = "lm", se = TRUE, color = "violetred3", fill = "violetred3") +
      labs(
        title = paste0(gene, " in ", group_of_interest, ": Females"),
        x = "Testosterone",
        y = "LogCPM"
      ) +
      theme_minimal()
  }) %>% compact()  # remove NULLs
  
  if (length(plot_list) == 0) return()
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
  
  # Clean group name for filename
clean_name <- gsub("/", "-", group_of_interest)
  
  # Save to PNG
  ggsave(
    filename = file.path(output_dir, paste0(clean_name, "_female_top_genes.png")),
    plot = combined_plot,
    width = 10,
    height = 6 + ceiling(length(plot_list) / 2),
    dpi = 300
  )
})

```

# Do everything again for estrogen - can make this into better function later

```{r prepare to run limma}
colnames(metadata_merged)[colnames(metadata_merged) == "17B-E2"] <- "E2"

run_voom_lm <- function(counts, metadata) {
  # Transpose counts: genes as rows, samples as columns
  counts_t <- t(counts)

  # Check dimensions
  stopifnot(ncol(counts_t) == nrow(metadata))

  # Create design matrix
  design <- model.matrix(~ E2 + cell_count + GW, data = metadata)

  # Run voom
  v <- limma::voom(counts_t, design)

  # Fit model
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)

  # Tidy results
  results <- limma::topTable(fit, coef = "E2", number = Inf, sort.by = "none") %>%
    tibble::rownames_to_column("gene")

  return(results)
}

# Ensure sample order matches
stopifnot(identical(rownames(pb$counts), rownames(metadata_merged)))

# Split by both cell type and sex
split_indices <- split(seq_len(nrow(pb$counts)), interaction(metadata_merged$cell_type, metadata_merged$Sex, drop = TRUE))
```

```{r run limma for each group}
# Run model per group
results_by_group <- purrr::map(split(seq_len(nrow(pb$counts)), metadata_merged$group), function(idx) {
  counts_sub <- pb$counts[idx, , drop = FALSE]
  metadata_sub <- metadata_merged[idx, , drop = FALSE]

  # Remove samples with zero library size
  keep_samples <- rowSums(counts_sub) > 0
  counts_sub <- counts_sub[keep_samples, , drop = FALSE]
  metadata_sub <- metadata_sub[keep_samples, , drop = FALSE]

  # Skip group if too few samples remain
  if (nrow(metadata_sub) < 5) {
    message("Skipping group ", unique(metadata_sub$group), " due to too few samples.")
    return(NULL)
  }

  # Optional: filter genes with low expression across samples
  keep_genes <- colSums(counts_sub >= 2) >= 3
  counts_sub <- counts_sub[, keep_genes, drop = FALSE]

  if (ncol(counts_sub) == 0) {
    message("Skipping group ", unique(metadata_sub$group), " due to no genes passing filter.")
    return(NULL)
  }

  # Run voom + lm with error handling
  tryCatch({
    result <- run_voom_lm(counts_sub, metadata_sub)
    result$cell_type <- metadata_sub$cell_type[1]
    result$sex <- metadata_sub$Sex[1]
    result
  }, error = function(e) {
    message("Skipping group ", unique(metadata_sub$group), " due to error: ", e$message)
    NULL
  })
})
```

```{r examine limma results}
# Combine all results (remove NULLs)
results_all <- bind_rows(results_by_group[!sapply(results_by_group, is.null)])

# examine female results
results_female <- results_all %>%
  filter(sex == "F")

write_csv(results_female, "../../results/targeted/res_E2_genes_female.csv")

sig_counts_female <- results_female %>%
  filter(adj.P.Val < 0.05) %>%
  count(cell_type, name = "num_sig_genes") %>%
  arrange(desc(num_sig_genes))

ggplot(sig_counts_female, aes(x = cell_type, y = num_sig_genes)) +
  geom_bar(stat = "identity", fill = "violetred3") +
  geom_text(aes(label = num_sig_genes), vjust = -0.5) +
  theme_minimal() +
  labs(x = "Cell Type", y = "Number of Significant Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../results/targeted/sig_E2_genes_female.png")

# examine male results
results_male <- results_all %>%
  filter(sex == "M")

write_csv(results_male, "../../results/targeted/res_E2_genes_male.csv")

sig_counts_male <- results_male %>%
  filter(adj.P.Val < 0.05) %>%
  count(cell_type, name = "num_sig_genes") %>%
  arrange(desc(num_sig_genes))

ggplot(sig_counts_male, aes(x = cell_type, y = num_sig_genes)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = num_sig_genes), vjust = -0.5) +
  theme_minimal() +
  labs(x = "Cell Type", y = "Number of Significant Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../results/targeted/sig_E2_genes_male.png")

# Remove genes (columns) with zero total count
pb_counts_filtered <- pb$counts[, colSums(pb$counts) > 0]

# make log(cpm) matrix for plotting
log_cpm_mat <- cpm(pb_counts_filtered, log = TRUE, prior.count = 1)
```
## plot estrogen in males

```{r make and save male plots}
# Get unique male cell types from results
cell_types <- unique(results_male$cell_type)

# Directory to save plots
output_dir <- "../../results/targeted/pseudobulk_plots_estrogen"
dir.create(output_dir, showWarnings = FALSE)

# Loop through groups and save combined plots
walk(cell_types, function(group_of_interest) {
  
  # === Filter for this group ===
  results_subset <- results_male %>%
    filter(cell_type == group_of_interest) %>%
    arrange(-abs(t)) %>%
    slice_head(n = 6)
  
  # === Subset expression and metadata ===
  samples_in_group <- metadata_merged %>%
    filter(cell_type == group_of_interest)
  
  if (nrow(samples_in_group) == 0) return()
  counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]
  
  # === Plot top genes ===
  plot_list <- map(results_subset$gene, function(gene) {
    if (!gene %in% colnames(counts_in_group)) return(NULL)
    
    expr_vals <- counts_in_group[, gene]
    if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
    
    df <- samples_in_group %>%
      mutate(expr = expr_vals) %>%
      filter(Sex == "M")
    
    ggplot(df, aes(x = E2, y = expr)) +
      geom_point() +
      geom_smooth(method = "lm", se = TRUE, color = "steelblue", fill = "steelblue") +
      labs(
        title = paste0(gene, " in ", group_of_interest, ": Males"),
        x = "Estrogen",
        y = "LogCPM"
      ) +
      theme_minimal()
  }) %>% compact()  # remove NULLs
  
  if (length(plot_list) == 0) return()
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
  
  # Clean group name for filename
clean_name <- gsub("/", "-", group_of_interest)
  
  # Save to PNG
  ggsave(
    filename = file.path(output_dir, paste0(clean_name, "_male_top_genes.png")),
    plot = combined_plot,
    width = 10,
    height = 6 + ceiling(length(plot_list) / 2),
    dpi = 300
  )
})

```

## plot top estrogen genes in females

```{r make and save female plots}
# Get unique male cell types from results
cell_types <- unique(results_female$cell_type)

# Directory to save plots
output_dir <- "../../results/targeted/pseudobulk_plots_estrogen"
dir.create(output_dir, showWarnings = FALSE)

# Loop through groups and save combined plots
walk(cell_types, function(group_of_interest) {
  
  # === Filter for this group ===
  results_subset <- results_female %>%
    filter(cell_type == group_of_interest) %>%
    arrange(-abs(t)) %>%
    slice_head(n = 6)
  
  # === Subset expression and metadata ===
  samples_in_group <- metadata_merged %>%
    filter(cell_type == group_of_interest)
  
  if (nrow(samples_in_group) == 0) return()
  counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]

  
  # === Plot top genes ===
  plot_list <- map(results_subset$gene, function(gene) {
    if (!gene %in% colnames(counts_in_group)) return(NULL)
    
    expr_vals <- counts_in_group[, gene]
    if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
    
    df <- samples_in_group %>%
      mutate(expr = expr_vals) %>%
      filter(Sex == "F")
    
    ggplot(df, aes(x = E2, y = expr)) +
      geom_point() +
      geom_smooth(method = "lm", se = TRUE, color = "violetred3", fill = "violetred3") +
      labs(
        title = paste0(gene, " in ", group_of_interest, ": Females"),
        x = "Estrogen",
        y = "LogCPM"
      ) +
      theme_minimal()
  }) %>% compact()  # remove NULLs
  
  if (length(plot_list) == 0) return()
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
  
  # Clean group name for filename
clean_name <- gsub("/", "-", group_of_interest)
  
  # Save to PNG
  ggsave(
    filename = file.path(output_dir, paste0(clean_name, "_female_top_genes.png")),
    plot = combined_plot,
    width = 10,
    height = 6 + ceiling(length(plot_list) / 2),
    dpi = 300
  )
})

```

# Do everything again for progesterone - can make this into better function later

```{r prepare to run limma}
run_voom_lm <- function(counts, metadata) {
  # Transpose counts: genes as rows, samples as columns
  counts_t <- t(counts)

  # Check dimensions
  stopifnot(ncol(counts_t) == nrow(metadata))

  # Create design matrix
  design <- model.matrix(~ P4 + cell_count + GW, data = metadata)

  # Run voom
  v <- limma::voom(counts_t, design)

  # Fit model
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)

  # Tidy results
  results <- limma::topTable(fit, coef = "P4", number = Inf, sort.by = "none") %>%
    tibble::rownames_to_column("gene")

  return(results)
}

# Ensure sample order matches
stopifnot(identical(rownames(pb$counts), rownames(metadata_merged)))

# Split by both cell type and sex
split_indices <- split(seq_len(nrow(pb$counts)), interaction(metadata_merged$cell_type, metadata_merged$Sex, drop = TRUE))
```

```{r run limma for each group}
# Run model per group
results_by_group <- purrr::map(split(seq_len(nrow(pb$counts)), metadata_merged$group), function(idx) {
  counts_sub <- pb$counts[idx, , drop = FALSE]
  metadata_sub <- metadata_merged[idx, , drop = FALSE]

  # Remove samples with zero library size
  keep_samples <- rowSums(counts_sub) > 0
  counts_sub <- counts_sub[keep_samples, , drop = FALSE]
  metadata_sub <- metadata_sub[keep_samples, , drop = FALSE]

  # Skip group if too few samples remain
  if (nrow(metadata_sub) < 5) {
    message("Skipping group ", unique(metadata_sub$group), " due to too few samples.")
    return(NULL)
  }

  # Optional: filter genes with low expression across samples
  keep_genes <- colSums(counts_sub >= 2) >= 3
  counts_sub <- counts_sub[, keep_genes, drop = FALSE]

  if (ncol(counts_sub) == 0) {
    message("Skipping group ", unique(metadata_sub$group), " due to no genes passing filter.")
    return(NULL)
  }

  # Run voom + lm with error handling
  tryCatch({
    result <- run_voom_lm(counts_sub, metadata_sub)
    result$cell_type <- metadata_sub$cell_type[1]
    result$sex <- metadata_sub$Sex[1]
    result
  }, error = function(e) {
    message("Skipping group ", unique(metadata_sub$group), " due to error: ", e$message)
    NULL
  })
})
```

```{r examine limma results}
# Combine all results (remove NULLs)
results_all <- bind_rows(results_by_group[!sapply(results_by_group, is.null)])

# examine female results
results_female <- results_all %>%
  filter(sex == "F")

write_csv(results_female, "../../results/targeted/res_P4_genes_female.csv")

sig_counts_female <- results_female %>%
  filter(adj.P.Val < 0.05) %>%
  count(cell_type, name = "num_sig_genes") %>%
  arrange(desc(num_sig_genes))

ggplot(sig_counts_female, aes(x = cell_type, y = num_sig_genes)) +
  geom_bar(stat = "identity", fill = "violetred3") +
  geom_text(aes(label = num_sig_genes), vjust = -0.5) +
  theme_minimal() +
  labs(x = "Cell Type", y = "Number of Significant Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../results/targeted/sig_P4_genes_female.png")

# examine male results
results_male <- results_all %>%
  filter(sex == "M")

write_csv(results_male, "../../results/targeted/res_P4_genes_male.csv")

sig_counts_female <- results_male %>%
  filter(adj.P.Val < 0.05) %>%
  count(cell_type, name = "num_sig_genes") %>%
  arrange(desc(num_sig_genes))

ggplot(sig_counts_female, aes(x = cell_type, y = num_sig_genes)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = num_sig_genes), vjust = -0.5) +
  theme_minimal() +
  labs(x = "Cell Type", y = "Number of Significant Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../results/targeted/sig_P4_genes_male.png")

# Remove genes (columns) with zero total count
pb_counts_filtered <- pb$counts[, colSums(pb$counts) > 0]

# make log(cpm) matrix for plotting
log_cpm_mat <- cpm(pb_counts_filtered, log = TRUE, prior.count = 1)
```
## plot top P4 genes for males

```{r make and save male plots}
# Get unique male cell types from results
cell_types <- unique(results_male$cell_type)

# Directory to save plots
output_dir <- "../../results/targeted/pseudobulk_plots_progesterone"
dir.create(output_dir, showWarnings = FALSE)

# Loop through groups and save combined plots
walk(cell_types, function(group_of_interest) {
  
  # === Filter for this group ===
  results_subset <- results_male %>%
    filter(cell_type == group_of_interest) %>%
    arrange(-abs(t)) %>%
    slice_head(n = 6)
  
  # === Subset expression and metadata ===
  samples_in_group <- metadata_merged %>%
    filter(cell_type == group_of_interest)
  
  if (nrow(samples_in_group) == 0) return()
  counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]
  
  # === Plot top genes ===
  plot_list <- map(results_subset$gene, function(gene) {
    if (!gene %in% colnames(counts_in_group)) return(NULL)
    
    expr_vals <- counts_in_group[, gene]
    if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
    
    df <- samples_in_group %>%
      mutate(expr = expr_vals) %>%
      filter(Sex == "M")
    
    ggplot(df, aes(x = E2, y = expr)) +
      geom_point() +
      geom_smooth(method = "lm", se = TRUE, color = "steelblue", fill = "steelblue") +
      labs(
        title = paste0(gene, " in ", group_of_interest, ": Males"),
        x = "Progesterone",
        y = "LogCPM"
      ) +
      theme_minimal()
  }) %>% compact()  # remove NULLs
  
  if (length(plot_list) == 0) return()
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
  
  # Clean group name for filename
clean_name <- gsub("/", "-", group_of_interest)
  
  # Save to PNG
  ggsave(
    filename = file.path(output_dir, paste0(clean_name, "_male_top_genes.png")),
    plot = combined_plot,
    width = 10,
    height = 6 + ceiling(length(plot_list) / 2),
    dpi = 300
  )
})

```

## plot top P4 genes for females

```{r make and save female plots}
# Get unique female cell types from results
cell_types <- unique(results_female$cell_type)

# Directory to save plots
output_dir <- "../../results/targeted/pseudobulk_plots_progesterone"
dir.create(output_dir, showWarnings = FALSE)

# Loop through groups and save combined plots
walk(cell_types, function(group_of_interest) {
  
  # === Filter for this group ===
  results_subset <- results_female %>%
    filter(cell_type == group_of_interest) %>%
    arrange(-abs(t)) %>%
    slice_head(n = 6)
  
  # === Subset expression and metadata ===
  samples_in_group <- metadata_merged %>%
    filter(cell_type == group_of_interest)
  
  if (nrow(samples_in_group) == 0) return()
  counts_in_group <- log_cpm_mat[rownames(samples_in_group), , drop = FALSE]

  
  # === Plot top genes ===
  plot_list <- map(results_subset$gene, function(gene) {
    if (!gene %in% colnames(counts_in_group)) return(NULL)
    
    expr_vals <- counts_in_group[, gene]
    if (all(is.na(expr_vals)) || all(expr_vals == 0)) return(NULL)
    
    df <- samples_in_group %>%
      mutate(expr = expr_vals) %>%
      filter(Sex == "F")
    
    ggplot(df, aes(x = E2, y = expr)) +
      geom_point() +
      geom_smooth(method = "lm", se = TRUE, color = "violetred3", fill = "violetred3") +
      labs(
        title = paste0(gene, " in ", group_of_interest, ": Females"),
        x = "Progesterone",
        y = "LogCPM"
      ) +
      theme_minimal()
  }) %>% compact()  # remove NULLs
  
  if (length(plot_list) == 0) return()
  
  # Combine plots using patchwork
  combined_plot <- wrap_plots(plotlist = plot_list, ncol = 2)
  
  # Clean group name for filename
clean_name <- gsub("/", "-", group_of_interest)
  
  # Save to PNG
  ggsave(
    filename = file.path(output_dir, paste0(clean_name, "_female_top_genes.png")),
    plot = combined_plot,
    width = 10,
    height = 6 + ceiling(length(plot_list) / 2),
    dpi = 300
  )
})

```
