---
title: "Correlate Age and Metabolite Concentrations"
author: "Emma Jones"
date: "2025-06-20"
output: html_document
---

The purpose of this analysis is threefold: 1) to understand the relationship between age (gestational week) and hormone concentration, 2) to find outliers in the distribution of hormone concentrations, and 3) to determine if the hormones themselves are correlated with each other. This is important for downstream analyses, to see if GW should be added as a covariate, and if outliers should be removed.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, we load packages. We might want more later, but for now I know I need the following

```{r load packages}
library(tidyverse)
library(Seurat)
library(readxl)
library(Signac)
library(glue)
library(cowplot)
library(edgeR)
library(limma)
library(patchwork)
```

## Load data

```{r load seurat object and excel spreadsheet}
# load 10x multiome data
SSD_data_v5 <- readRDS("../../data/SSD_data_v5.rds")
# remove doublets
SSD_data_v5 <- subset(SSD_data_v5, subset = Droplet == "SNG")

# load hormone metabolomics data
targeted_comparision_raw <- read_xlsx("../../doc/targeted/Targeted Comparison.xlsx")

#rename estrogen
targeted_comparision_raw$E2 <- targeted_comparision_raw$`17B-E2`

```

## Remove Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

```{r remove outliers}
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) + 1.5 * IQR(data[[variable]], na.rm = TRUE)
  data %>% filter(.data[[variable]] <= outlier_limit)
}

targeted_comparision_filtered_TT <- detect_outliers(targeted_comparision_raw, "TT")
targeted_comparision_filtered_E2 <- detect_outliers(targeted_comparision_raw, "17B-E2")
targeted_comparision_filtered_P4 <- detect_outliers(targeted_comparision_raw, "P4")

# save only samples with observations for all 3, will be used for correlations later
all_obs <- intersect(targeted_comparision_filtered_TT$Sample, targeted_comparision_filtered_E2$Sample)
all_obs <- intersect(all_obs, targeted_comparision_filtered_P4$Sample)

targeted_comparision_filtered <- subset(targeted_comparision_raw, Sample %in% all_obs)
```

# Are hormones correlated with each other?

We can make a fairly quick scatterplot for each pairwise relationship

```{r look at overall hormone relationships - tt and e2}
# look at relationship with tt and e2
ggplot(targeted_comparision_filtered, aes(x = TT, y = E2, color = GW)) +
  geom_point() + theme_minimal()
ggplot(targeted_comparision_filtered, aes(x = TT, y = E2, color = Sex)) +
  geom_point() + theme_minimal()
# Save
ggsave("../../results/targeted/hormone_hormone_plots/TT_E2_scatter.png")

# linear model
tt_e2 <- lm(TT~E2, data=targeted_comparision_filtered) 
summary(tt_e2) #pval is 0.006437

targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]
tt_e2_m <- lm(TT~E2, data=targeted_comparision_males) 
summary(tt_e2_m) #pval is 0.2538
```
```{r look at overall hormone relationships - p4 and e2}
# look at relationship with p4 and e2
ggplot(targeted_comparision_filtered, aes(x = P4, y = E2, color = GW)) +
  geom_point() + theme_minimal()
ggplot(targeted_comparision_filtered, aes(x = P4, y = E2, color = Sex)) +
  geom_point() + theme_minimal()
# Save
ggsave("../../results/targeted/hormone_hormone_plots/P4_E2_scatter.png")

# linear model
p4_e2 <- lm(P4~E2, data=targeted_comparision_filtered) 
summary(p4_e2) #pval is 0.06845

targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
p4_e2_f <- lm(P4~E2, data=targeted_comparision_females) 
summary(p4_e2_f) #pval is 0.02831
```

```{r look at overall hormone relationships - p4 and tt}
# look at relationship with p4 and tt
ggplot(targeted_comparision_filtered, aes(x = P4, y = TT, color = GW)) +
  geom_point() + theme_minimal()
ggplot(targeted_comparision_filtered, aes(x = P4, y = TT, color = Sex)) +
  geom_point() + theme_minimal()
# Save
ggsave("../../results/targeted/hormone_hormone_plots/P4_TT_scatter.png")

# linear model
p4_tt <- lm(P4~TT, data=targeted_comparision_filtered) 
summary(p4_tt) #pval is 0.4612

targeted_comparision_females <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "F",]
p4_tt_f <- lm(P4~TT, data=targeted_comparision_females) 
summary(p4_tt_f) #pval is 0.3497

targeted_comparision_males <- targeted_comparision_filtered[targeted_comparision_filtered$Sex == "M",]
p4_tt_m <- lm(P4~TT, data=targeted_comparision_males) 
summary(p4_tt_m) #pval is 0.5836
```

# testosterone analysis

## how does this correlate with age?
```{r look at age (gestational week) relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered_TT, aes(x = GW, y = TT, color = Sex)) +
  geom_point() + theme_minimal()

all_simple <- lm(TT~GW, data=targeted_comparision_filtered_TT) 
summary(all_simple)
```

so its not significant, but the graph clearly supports that males differ from females, so I will separate the groups

```{r separate groups and find a model}
targeted_comparision_males <- targeted_comparision_filtered_TT[targeted_comparision_filtered_TT$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_males) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_males) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_males)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_males$GW)

# Base plot - males
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 

# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run for females}
#filter for only females
targeted_comparision_females <- targeted_comparision_filtered_TT[targeted_comparision_filtered_TT$Sex == "F",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(TT~GW, data=targeted_comparision_females) 
linear_model2 <- lm(TT~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(TT~poly(GW,3,raw=TRUE), data=targeted_comparision_females) 

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Polynomial Fits", xlab="GW", ylab="TT")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 


# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 

```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_m <- loess(TT ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$TT, main="Loess Fit", xlab="GW", ylab="TT")
loess_fit_f <- loess(TT ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = TT, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Testosterone") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

# Save
ggsave("../../results/targeted/hormone_age_plots/TT_loess_graph.png")

```


```{r model using best-fitting equation}
male_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.002309

female_model <- lm(TT ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.7959

# Create prediction frame - male
new_male <- data.frame(GW = seq(min(targeted_comparision_males$GW),
                                max(targeted_comparision_males$GW), length.out = 100))
new_male$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_males),
                            newdata = new_male)

# Create prediction frame - female
new_female <- data.frame(GW = seq(min(targeted_comparision_females$GW),
                                max(targeted_comparision_females$GW), length.out = 100))
new_female$TT_pred <- predict(lm(TT ~ GW + I(GW^2), data = targeted_comparision_females),
                            newdata = new_female)

# Plot 1
ggplot(targeted_comparision_males, aes(x = GW, y = TT)) +
  geom_point(color = "dodgerblue", alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Males", x = "GW", y = "TT") +
  theme_minimal()

# Plot 2
ggplot(targeted_comparision_females, aes(x = GW, y = TT)) +
  geom_point(color = "violetred2", alpha = 0.6) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² Fit in Females", x = "GW", y = "TT") +
  theme_minimal()

# Combine information
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² + I(GW²) Fit in Males and Females", subtitle = "Male p-value: 0.01292\nFemale p-value:0.7959", x = "GW", y = "TT") +
  theme_minimal()

# move the text
ggplot(targeted_comparision_filtered, aes(x = GW, y = TT, color = Sex)) +
  scale_color_manual(values = c("F" = "violetred2", "M" = "dodgerblue")) +
  geom_point(alpha = 0.6) +
  geom_line(data = new_male, aes(x = GW, y = TT_pred), color = "dodgerblue4", size = 1.2) +
  geom_line(data = new_female, aes(x = GW, y = TT_pred), color = "violetred4", size = 1.2) +
  labs(title = "TT ~ GW² + I(GW²) Fit in Males and Females", x = "Gestational Week (GW)", y = "Testosterone") +
  annotate("text", x = 16, y = 0.4, label = "Male p-value: 0.002309", color = "dodgerblue4", hjust = 0) +
  annotate("text", x = 16, y = 0.35, label = "Female p-value: 0.7959", color = "violetred4", hjust = 0) +
  theme_minimal()

# save new plot
ggsave("../../results/targeted/hormone_age_plots/TT_GW_modelfit.png")

```

# do again for estrogen

## Detect Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

## how does this correlate with age?

```{r look at overall relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered_E2, aes(x = GW, y = E2, color = Sex)) +
  geom_point() + theme_minimal()

all_simple <- lm(E2~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```

im not seeing a strong difference for males or females

```{r separate groups and find a model}
targeted_comparision_females <- targeted_comparision_filtered_E2[targeted_comparision_filtered_E2$Sex == "F",]
targeted_comparision_males <- targeted_comparision_filtered_E2[targeted_comparision_filtered_E2$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(E2~GW, data=targeted_comparision_females) 
linear_model2 <- lm(E2~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(E2~poly(GW,3,raw=TRUE), data=targeted_comparision_females)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$E2, main="Polynomial Fits", xlab="GW", ylab="E2")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 

# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$E2, main="Loess Fit", xlab="GW", ylab="E2")
loess_fit_m <- loess(E2 ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$E2, main="Loess Fit", xlab="GW", ylab="E2")
loess_fit_f <- loess(E2 ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = E2, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Estradiol") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

# Save
ggsave("../../results/targeted/hormone_age_plots/E2_loess_graph.png")

```


```{r model using best-fitting equation}
male_model <- lm(E2 ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.2672

female_model <- lm(E2 ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.6758

```

There seems to be no significant relationship with estrogen and age, but the high outliers (5) might be influencing other analyses. I should also run outlier analysis on progesterone.

# Progesterone outlier analysis!!!

## Detect Outliers

The statistical definition of an outlier differs by field, but generally Q3 + 1.5*IQR is accepted as a way to detect outliers

Let's calculate below:

## how does this correlate with age?

```{r look at overall relationship}
# look at relationship with age
ggplot(targeted_comparision_filtered_P4, aes(x = GW, y = P4, color = Sex)) +
  geom_point() + theme_minimal()

all_simple <- lm(P4~GW, data=targeted_comparision_filtered) 
summary(all_simple)
```

im not seeing a strong difference for males or females

```{r separate groups and find a model}
targeted_comparision_females <- targeted_comparision_filtered_P4[targeted_comparision_filtered_P4$Sex == "F",]
targeted_comparision_males <- targeted_comparision_filtered_P4[targeted_comparision_filtered_P4$Sex == "M",]

# fit polynomial regression models up to degree 5 
linear_model1 <- lm(P4~GW, data=targeted_comparision_females) 
linear_model2 <- lm(P4~poly(GW,2,raw=TRUE), data=targeted_comparision_females) 
linear_model3 <- lm(P4~poly(GW,3,raw=TRUE), data=targeted_comparision_females)

# Sort the x values and get corresponding predicted values
x_axis_sorted <- sort(targeted_comparision_females$GW)

# Base plot - males
plot(targeted_comparision_females$GW, targeted_comparision_females$P4, main="Polynomial Fits", xlab="GW", ylab="P4")

# Add model lines using sorted x
lines(x_axis_sorted, predict(linear_model1, data.frame(GW=x_axis_sorted)), col='green') 
lines(x_axis_sorted, predict(linear_model2, data.frame(GW=x_axis_sorted)), col='red') 
lines(x_axis_sorted, predict(linear_model3, data.frame(GW=x_axis_sorted)), col='purple') 

# calculated adjusted R-squared of each model 
summary(linear_model1)$adj.r.squared 
summary(linear_model2)$adj.r.squared 
summary(linear_model3)$adj.r.squared 
```

```{r run loess instead}
# try loess
plot(targeted_comparision_males$GW, targeted_comparision_males$P4, main="Loess Fit", xlab="GW", ylab="P4")
loess_fit_m <- loess(P4 ~ GW, data = targeted_comparision_males, span = 0.85)
lines(sort(targeted_comparision_males$GW),
      predict(loess_fit_m, newdata = data.frame(GW = sort(targeted_comparision_males$GW))),
      col = "dodgerblue", lwd = 2)

plot(targeted_comparision_females$GW, targeted_comparision_females$P4, main="Loess Fit", xlab="GW", ylab="P4")
loess_fit_f <- loess(P4 ~ GW, data = targeted_comparision_females, span = 0.85)
lines(sort(targeted_comparision_females$GW),
      predict(loess_fit_f, newdata = data.frame(GW = sort(targeted_comparision_females$GW))),
      col = "violetred2", lwd = 2)

# Combine male and female data (assuming you already split them)
targeted_comparision_males$sex <- "male"
targeted_comparision_females$sex <- "female"
combined_data <- rbind(targeted_comparision_males, targeted_comparision_females)

# Plot with LOESS curves
ggplot(combined_data, aes(x = GW, y = P4, color = sex, fill = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.85, se = TRUE, size = 1.2) +
  labs(title = "LOESS Fit by Sex (outliers removed)", x = "Gestational Week (GW)", y = "Progesterone") +
  scale_color_manual(values = c("male" = "dodgerblue", "female" = "violetred2")) +
  theme_minimal()

# Save
ggsave("../../results/targeted/hormone_age_plots/P4_loess_graph.png")

```


```{r model using best-fitting equation}
male_model <- lm(P4 ~ GW + I(GW^2), data = targeted_comparision_males)
summary(male_model) #p-value: 0.3699

female_model <- lm(P4 ~ GW + I(GW^2), data = targeted_comparision_females)
summary(female_model) #p-value: 0.6672

```


