---
title: "Gonadal Hormone and Gene Expression Relationship Analysis - Combined Sex"
author: "Emma Jones, Sneha Chaturvedi"
date: "Last updated 11.06.25"
output: 
  html_document:
    toc: TRUE
---
# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Packages <- c("tidyverse", "Seurat", "Signac", "glue", "readxl", "cowplot", "edgeR", "limma", "patchwork", "svglite", "Matrix", "ggpubr")

lapply(Packages, library, character.only=TRUE)

set.seed(123)
```

The purpose of this script is to 1) explore single cell RNA-seq results and 2) analyze for relationships between gene expression and testosterone, progesterone, and estradiol concentrations from developing human brain samples. This script analyzes males and females combined. 

## Load Data
```{r load data}
SSD_data_v5 <- readRDS("../../data/SSD_data_v5.rds")
SSD_data_v5 <- subset(SSD_data_v5, subset = Droplet == "SNG")

targeted_comparison_raw <- read_xlsx("../../doc/targeted/targeted_hormones.xlsx")
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`
```

## Show Cell Type UMAP
```{r display umap}
DefaultAssay(SSD_data_v5) <- "SCT"
DimPlot(SSD_data_v5, label = TRUE, label.box = TRUE, repel = TRUE, label.color = "white", label.size = 5) + NoLegend()

ggsave("../../results/targeted/single_cell_plots/celltype_umap.png", width = 7, height = 4)

#Create Cell Type Count Table
sample_cluster_table <- table(SSD_data_v5$sample, SSD_data_v5$cluster_call)
write.csv(sample_cluster_table, "../../results/targeted/gene_hormone_csv/sample_cluster_table.csv", row.names = TRUE)
```

# Pseudobulk
## Metadata and Pseudobulk Functions

```{r metadata and pseudobulk}
prepare_metadata_for_analysis <- function(pb_metadata, targeted_comparison_filtered) {
  targeted_comparison_filtered$Sample <- paste0("tissue", targeted_comparison_filtered$Sample)
  pb_metadata %>%
    as.data.frame() %>%
    rownames_to_column("pseudobulk_id") %>%
    left_join(targeted_comparison_filtered %>% select(Sample, Sex, TT, E2, P4, GW),
              by = c("sample_id" = "Sample")) %>%
    filter(!is.na(Sex), cell_count >= 10) %>%
    mutate(group = paste(cell_type, Sex, sep = "_")) %>%
    column_to_rownames("pseudobulk_id")
}

pseudobulk_data <- function(seurat_obj, assay = "RNA", layer = "counts", sample_col = "sample_id", celltype_col = "cell_type") {
  counts <- GetAssayData(seurat_obj, assay = assay, slot = layer)
  meta <- seurat_obj@meta.data
  meta$cell_type <- meta$cluster_call
  group_ids <- interaction(meta[[sample_col]], meta[[celltype_col]], drop = TRUE)
  group_levels <- levels(group_ids)
  aggregated_list <- list()
  metadata_list <- list()
  
  
  for (gid in group_levels) {
  group_cells <- which(group_ids == gid)
  if (length(group_cells) < 1) next
  group_counts <- Matrix::rowSums(counts[, group_cells, drop = FALSE])
  aggregated_list[[gid]] <- group_counts
    metadata_list[[gid]] <- data.frame(
      sample_id = meta[[sample_col]][group_cells[1]],
      cell_type = meta[[celltype_col]][group_cells[1]],
      cell_count = length(group_cells),
      row.names = gid
    )
  }
  
  pseudobulk_mat <- do.call(rbind, aggregated_list)
  metadata_df <- do.call(rbind, metadata_list)
  
  pseudobulk_mat_log <- log1p(as.matrix(pseudobulk_mat))
  
  list(counts = pseudobulk_mat, logcounts = pseudobulk_mat_log, metadata = metadata_df)
}

generate_and_filter_pseudobulk <- function(seurat_obj, targeted_filtered) {
  seurat_obj$cell_type <- seurat_obj$cluster_call  # ensure column exists
  seurat_obj$sample_id <- seurat_obj$sample        # ensure column exists
  
  pb <- pseudobulk_data(seurat_obj)
  pb_metadata <- prepare_metadata_for_analysis(pb$metadata, targeted_filtered)
  
  common_ids <- intersect(rownames(pb$logcounts), rownames(pb_metadata))
  pb$logcounts <- pb$logcounts[common_ids, , drop = FALSE]
  pb$counts <- pb$counts[common_ids, , drop = FALSE]
  pb$metadata <- pb_metadata[common_ids, , drop = FALSE]
  
  list(pb = pb, metadata = pb_metadata)
}
```

## Total Pseudobulk

```{r generate total pseudobulk data, warnings = FALSE}
pb_total <- generate_and_filter_pseudobulk(SSD_data_v5, targeted_comparison_raw)

```
## Covariate Testing
Does gestational week predict gene expression? 

``` {r covariate testing}
# Linear regression: gestational week predicts gene expression
perform_gw_regression <- function(pb_object, analysis_name = "Pseudobulk") {
  
  # Extract components
  logcounts <- pb_object$pb$logcounts
  metadata <- pb_object$metadata
  
  # Ensure GW column exists
  if (!"GW" %in% colnames(metadata)) {
    stop("GW (gestational week) column not found in metadata")
  }
  
  # Initialize results data frame
  results <- data.frame(
    gene = character(),
    estimate = numeric(),
    std_error = numeric(),
    t_value = numeric(),
    p_value = numeric(),
    adj_p_value = numeric(),
    r_squared = numeric(),
    n_samples = integer(),
    stringsAsFactors = FALSE
  )
  
  # Get all genes
  genes <- colnames(logcounts)
  
  # Progress tracking
  cat(sprintf("Running linear regression for %d genes in %s...\n", 
              length(genes), analysis_name))
  
  # Loop through each gene
  for (i in seq_along(genes)) {
    gene <- genes[i]
    
    # Progress indicator
    if (i %% 500 == 0) {
      cat(sprintf("  Processed %d / %d genes (%.1f%%)\n", 
                  i, length(genes), 100 * i / length(genes)))
    }
    
    # Prepare data for regression
    gene_expr <- logcounts[, gene]
    
    # Remove NA values
    valid_idx <- !is.na(gene_expr) & !is.na(metadata$GW)
    
    if (sum(valid_idx) < 3) {
      next  # Skip if too few samples
    }
    
    expr_valid <- gene_expr[valid_idx]
    gw_valid <- metadata$GW[valid_idx]
    
    # Fit linear model
    tryCatch({
      model <- lm(expr_valid ~ gw_valid)
      model_summary <- summary(model)
      
      # Extract coefficients for GW (slope)
      coef_table <- coef(model_summary)
      gw_coef <- coef_table["gw_valid", ]
      
      # Store results
      results <- rbind(results, data.frame(
        gene = gene,
        estimate = gw_coef["Estimate"],
        std_error = gw_coef["Std. Error"],
        t_value = gw_coef["t value"],
        p_value = gw_coef["Pr(>|t|)"],
        adj_p_value = NA,
        r_squared = model_summary$r.squared,
        n_samples = sum(valid_idx),
        stringsAsFactors = FALSE
      ))
      
    }, error = function(e) {
      return(NULL)
    })
  }
  
  # Adjust p-values for multiple testing (FDR correction)
  if (nrow(results) > 0) {
    results$adj_p_value <- p.adjust(results$p_value, method = "fdr")
    results <- results[order(results$adj_p_value), ]
    rownames(results) <- NULL
  }
  
  cat(sprintf("Completed! Found %d genes with valid results.\n", nrow(results)))
  
  return(results)
}

# Run analysis on the single pseudobulk
cat("\n=== Running GW regression analysis ===\n\n")

results_gw_regression <- perform_gw_regression(pb_total)

# Summary function
summarize_gw_results <- function(results, alpha = 0.05) {
  cat(sprintf("\n=== SUMMARY ===\n"))
  cat(sprintf("Total genes tested: %d\n", nrow(results)))
  cat(sprintf("Significant genes (adj. p < %.2f): %d\n", 
              alpha, sum(results$adj_p_value < alpha, na.rm = TRUE)))
  cat(sprintf("Positive associations: %d\n", 
              sum(results$adj_p_value < alpha & results$estimate > 0, na.rm = TRUE)))
  cat(sprintf("Negative associations: %d\n", 
              sum(results$adj_p_value < alpha & results$estimate < 0, na.rm = TRUE)))
  cat("\nTop 10 most significant genes:\n")
  
  top_results <- head(results[, c("gene", "estimate", "p_value", "adj_p_value", "r_squared")], 10)
  
  cat(sprintf("%-15s %12s %12s %12s %10s\n", 
              "Gene", "Estimate", "P-value", "Adj.P-value", "R-squared"))
  cat(paste(rep("-", 70), collapse = ""), "\n")
  
  for (i in 1:nrow(top_results)) {
    cat(sprintf("%-15s %12.6f %12.2e %12.2e %10.4f\n",
                top_results$gene[i],
                top_results$estimate[i],
                top_results$p_value[i],
                top_results$adj_p_value[i],
                top_results$r_squared[i]))
  }
  cat("\n")
}

# Print summary
cat("\n=== GW Regression ===")
summarize_gw_results(results_gw_regression)

# SUMMARY FIGURE: Effect of GW across genes
plot_gw_summary <- function(results, alpha = 0.05, output_file = NULL) {
  
  if (!is.null(output_file)) {
    svg(output_file, width = 10, height = 7)
  }
  
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
  
  # 1. Volcano plot
  plot(results$estimate, -log10(results$p_value),
       xlab = "Effect Size (GW coefficient)",
       ylab = "-log10(p-value)",
       main = "Volcano Plot",
       pch = 16, cex = 0.5, col = rgb(0, 0, 0, 0.3))
  
  sig_idx <- results$adj_p_value < alpha
  points(results$estimate[sig_idx], -log10(results$p_value[sig_idx]),
         pch = 16, cex = 0.6, col = "red")
  
  abline(h = -log10(0.05), lty = 2, col = "blue")
  abline(v = 0, lty = 1, col = "gray50")
  
  legend("topright", 
         legend = c(sprintf("Sig. (FDR < %.2f)", alpha), "Non-sig."),
         col = c("red", rgb(0, 0, 0, 0.3)),
         pch = 16, cex = 0.8, bty = "n")
  
  # 2. Distribution of effect sizes
  hist(results$estimate, breaks = 50,
       xlab = "Effect Size (GW coefficient)",
       ylab = "Number of Genes",
       main = "Effect Size Distribution",
       col = "lightblue", border = "white")
  abline(v = 0, lty = 2, col = "red", lwd = 2)
  
  mean_effect <- mean(results$estimate, na.rm = TRUE)
  abline(v = mean_effect, lty = 2, col = "darkblue", lwd = 2)
  legend("topright",
         legend = c(sprintf("Mean = %.4f", mean_effect),
                   sprintf("Positive = %d", sum(results$estimate > 0)),
                   sprintf("Negative = %d", sum(results$estimate < 0))),
         bty = "n", cex = 0.8)
  
  # 3. P-value distribution
  hist(results$p_value, breaks = 50,
       xlab = "P-value",
       ylab = "Number of Genes",
       main = "P-value Distribution",
       col = "lightgreen", border = "white")
  abline(v = 0.05, lty = 2, col = "red", lwd = 2)
  
  # 4. R-squared distribution
  hist(results$r_squared, breaks = 50,
       xlab = "R²",
       ylab = "Number of Genes",
       main = "R² Distribution",
       col = "lightyellow", border = "white")
  
  median_r2 <- median(results$r_squared, na.rm = TRUE)
  abline(v = median_r2, lty = 2, col = "darkgreen", lwd = 2)
  legend("topright",
         legend = sprintf("Median R² = %.3f", median_r2),
         bty = "n", cex = 0.8)
  
  mtext("Gestational Week Effect on Gene Expression", 
        outer = TRUE, cex = 1.3, font = 2)
  
  if (!is.null(output_file)) {
    dev.off()
    cat(sprintf("\nSummary figure saved to: %s\n", output_file))
  }
}

# Generate summary figure
plot_gw_summary(results_gw_regression, alpha = 0.05, 
                output_file = "../../results/targeted/gene_expression_plots/GW_regression_summary_pseudobulk.svg")

# Plot top genes
plot_top_genes <- function(pb_object, results, n_genes = 6, output_file = NULL) {
  
  top_genes <- head(results$gene, n_genes)
  
  if (!is.null(output_file)) {
    svg(output_file, width = 12, height = 8)
  }
  
  par(mfrow = c(2, 3), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
  
  for (gene in top_genes) {
    logcounts <- pb_object$pb$logcounts
    metadata <- pb_object$metadata
    
    expr <- logcounts[, gene]
    gw <- metadata$GW
    
    valid_idx <- !is.na(expr) & !is.na(gw)
    expr <- expr[valid_idx]
    gw <- gw[valid_idx]
    
    plot(gw, expr, 
         xlab = "Gestational Week", 
         ylab = "Log Expression",
         main = gene,
         pch = 16, col = rgb(0, 0, 0, 0.5), cex = 1.2)
    
    model <- lm(expr ~ gw)
    abline(model, col = "red", lwd = 2)
    
    gene_res <- results[results$gene == gene, ]
    legend("topleft", 
           legend = c(sprintf("R² = %.3f", gene_res$r_squared),
                     sprintf("p = %.2e", gene_res$adj_p_value),
                     sprintf("n = %d", gene_res$n_samples)),
           bty = "n", cex = 0.9)
  }
  
  mtext("Top Genes Associated with Gestational Week", 
        outer = TRUE, cex = 1.2, font = 2)
  
  if (!is.null(output_file)) {
    dev.off()
    cat(sprintf("\nTop genes figure saved to: %s\n", output_file))
  }
}

plot_top_genes(pb_total, results_gw_regression, n_genes = 6,
               output_file = "../../results/targeted/gene_expression_plots/GW_top_genes_pseudobulk.svg")

cat("\n=== Pseudobulk GW Analysis Complete ===\n")

# Export results to CSV
write.csv(results_gw_regression, "../../results/targeted/gene_hormone_csv/gwreg_results_pseudobulk.csv", row.names = FALSE)

```
Does cell count predict gene expression? 
``` {r cell count regression}
# Linear regression: cell count predicts gene expression
# Working on pseudobulked data (single pseudobulk, no conditions)

perform_cellcount_regression <- function(pb_object, analysis_name = "Pseudobulk", 
                                         cellcount_column = "cell_count") {
  
  # Extract components
  logcounts <- pb_object$pb$logcounts
  metadata <- pb_object$metadata
  
  # Ensure cell_count column exists
  if (!cellcount_column %in% colnames(metadata)) {
    stop(sprintf("%s column not found in metadata", cellcount_column))
  }
  
  # Initialize results data frame
  results <- data.frame(
    gene = character(),
    estimate = numeric(),
    std_error = numeric(),
    t_value = numeric(),
    p_value = numeric(),
    adj_p_value = numeric(),
    r_squared = numeric(),
    n_samples = integer(),
    stringsAsFactors = FALSE
  )
  
  # Get all genes
  genes <- colnames(logcounts)
  
  # Progress tracking
  cat(sprintf("Running cell count regression for %d genes in %s...\n", 
              length(genes), analysis_name))
  
  # Loop through each gene
  for (i in seq_along(genes)) {
    gene <- genes[i]
    
    # Progress indicator
    if (i %% 500 == 0) {
      cat(sprintf("  Processed %d / %d genes (%.1f%%)\n", 
                  i, length(genes), 100 * i / length(genes)))
    }
    
    # Prepare data for regression
    gene_expr <- logcounts[, gene]
    
    # Remove NA values
    valid_idx <- !is.na(gene_expr) & !is.na(metadata[[cellcount_column]])
    
    if (sum(valid_idx) < 3) {
      next  # Skip if too few samples
    }
    
    expr_valid <- gene_expr[valid_idx]
    cellcount_valid <- metadata[[cellcount_column]][valid_idx]
    
    # Fit linear model
    tryCatch({
      model <- lm(expr_valid ~ cellcount_valid)
      model_summary <- summary(model)
      
      # Extract coefficients for cell count (slope)
      coef_table <- coef(model_summary)
      cc_coef <- coef_table["cellcount_valid", ]
      
      # Store results
      results <- rbind(results, data.frame(
        gene = gene,
        estimate = cc_coef["Estimate"],
        std_error = cc_coef["Std. Error"],
        t_value = cc_coef["t value"],
        p_value = cc_coef["Pr(>|t|)"],
        adj_p_value = NA,
        r_squared = model_summary$r.squared,
        n_samples = sum(valid_idx),
        stringsAsFactors = FALSE
      ))
      
    }, error = function(e) {
      return(NULL)
    })
  }
  
  # Adjust p-values for multiple testing (FDR correction)
  if (nrow(results) > 0) {
    results$adj_p_value <- p.adjust(results$p_value, method = "fdr")
    results <- results[order(results$adj_p_value), ]
    rownames(results) <- NULL
  }
  
  cat(sprintf("Completed! Found %d genes with valid results.\n", nrow(results)))
  
  return(results)
}

# Check what cell count column exists in your pseudobulk metadata
cat("Available metadata columns:\n")
print(colnames(pb_total$metadata))

# Run analysis on the single pseudobulk
# Adjust cellcount_column parameter to match your metadata column name
cat("\n=== Running Cell Count regression analysis ===\n\n")

results_cellcount_regression <- perform_cellcount_regression(pb_total, cellcount_column = "cell_count")

# Summary function
summarize_cellcount_results <- function(results, alpha = 0.05) {
  cat(sprintf("\n=== SUMMARY ===\n"))
  cat(sprintf("Total genes tested: %d\n", nrow(results)))
  cat(sprintf("Significant genes (adj. p < %.2f): %d\n", 
              alpha, sum(results$adj_p_value < alpha, na.rm = TRUE)))
  cat(sprintf("Positive associations: %d\n", 
              sum(results$adj_p_value < alpha & results$estimate > 0, na.rm = TRUE)))
  cat(sprintf("Negative associations: %d\n", 
              sum(results$adj_p_value < alpha & results$estimate < 0, na.rm = TRUE)))
  cat("\nTop 10 most significant genes:\n")
  
  top_results <- head(results[, c("gene", "estimate", "p_value", "adj_p_value", "r_squared")], 10)
  
  cat(sprintf("%-15s %12s %12s %12s %10s\n", 
              "Gene", "Estimate", "P-value", "Adj.P-value", "R-squared"))
  cat(paste(rep("-", 70), collapse = ""), "\n")
  
  for (i in 1:nrow(top_results)) {
    cat(sprintf("%-15s %12.6f %12.2e %12.2e %10.4f\n",
                top_results$gene[i],
                top_results$estimate[i],
                top_results$p_value[i],
                top_results$adj_p_value[i],
                top_results$r_squared[i]))
  }
  cat("\n")
}

# Print summary
cat("\n=== Cell Count Regression ===")
summarize_cellcount_results(results_cellcount_regression)

# SUMMARY FIGURE: Effect of cell count across genes
plot_cellcount_summary <- function(results, alpha = 0.05, output_file = NULL) {
  
  if (!is.null(output_file)) {
    svg(output_file, width = 10, height = 7)
  }
  
  par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
  
  # 1. Volcano plot
  plot(results$estimate, -log10(results$p_value),
       xlab = "Effect Size (Cell Count coefficient)",
       ylab = "-log10(p-value)",
       main = "Volcano Plot",
       pch = 16, cex = 0.5, col = rgb(0, 0, 0, 0.3))
  
  sig_idx <- results$adj_p_value < alpha
  points(results$estimate[sig_idx], -log10(results$p_value[sig_idx]),
         pch = 16, cex = 0.6, col = "red")
  
  abline(h = -log10(0.05), lty = 2, col = "blue")
  abline(v = 0, lty = 1, col = "gray50")
  
  legend("topright", 
         legend = c(sprintf("Sig. (FDR < %.2f)", alpha), "Non-sig."),
         col = c("red", rgb(0, 0, 0, 0.3)),
         pch = 16, cex = 0.8, bty = "n")
  
  # 2. Distribution of effect sizes
  hist(results$estimate, breaks = 50,
       xlab = "Effect Size (Cell Count coefficient)",
       ylab = "Number of Genes",
       main = "Effect Size Distribution",
       col = "lightblue", border = "white")
  abline(v = 0, lty = 2, col = "red", lwd = 2)
  
  mean_effect <- mean(results$estimate, na.rm = TRUE)
  abline(v = mean_effect, lty = 2, col = "darkblue", lwd = 2)
  legend("topright",
         legend = c(sprintf("Mean = %.4f", mean_effect),
                   sprintf("Positive = %d", sum(results$estimate > 0)),
                   sprintf("Negative = %d", sum(results$estimate < 0))),
         bty = "n", cex = 0.8)
  
  # 3. P-value distribution
  hist(results$p_value, breaks = 50,
       xlab = "P-value",
       ylab = "Number of Genes",
       main = "P-value Distribution",
       col = "lightgreen", border = "white")
  abline(v = 0.05, lty = 2, col = "red", lwd = 2)
  
  # 4. R-squared distribution
  hist(results$r_squared, breaks = 50,
       xlab = "R²",
       ylab = "Number of Genes",
       main = "R² Distribution",
       col = "lightyellow", border = "white")
  
  median_r2 <- median(results$r_squared, na.rm = TRUE)
  abline(v = median_r2, lty = 2, col = "darkgreen", lwd = 2)
  legend("topright",
         legend = sprintf("Median R² = %.3f", median_r2),
         bty = "n", cex = 0.8)
  
  mtext("Cell Count Effect on Gene Expression", 
        outer = TRUE, cex = 1.3, font = 2)
  
  if (!is.null(output_file)) {
    dev.off()
    cat(sprintf("\nSummary figure saved to: %s\n", output_file))
  }
}

# Generate summary figure
plot_cellcount_summary(results_cellcount_regression, alpha = 0.05, 
                       output_file = "../../results/targeted/gene_expression_plots/CellCount_regression_summary_pseudobulk.svg")

# Plot top genes
plot_top_genes_cellcount <- function(pb_object, results, n_genes = 6, 
                                     output_file = NULL, cellcount_column = "cell_count") {
  
  top_genes <- head(results$gene, n_genes)
  
  if (!is.null(output_file)) {
    svg(output_file, width = 12, height = 8)
  }
  
  par(mfrow = c(2, 3), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
  
  for (gene in top_genes) {
    logcounts <- pb_object$pb$logcounts
    metadata <- pb_object$metadata
    
    expr <- logcounts[, gene]
    cellcount <- metadata[[cellcount_column]]
    
    valid_idx <- !is.na(expr) & !is.na(cellcount)
    expr <- expr[valid_idx]
    cellcount <- cellcount[valid_idx]
    
    plot(cellcount, expr, 
         xlab = "Cell Count", 
         ylab = "Log Expression",
         main = gene,
         pch = 16, col = rgb(0, 0, 0, 0.5), cex = 1.2)
    
    model <- lm(expr ~ cellcount)
    abline(model, col = "red", lwd = 2)
    
    gene_res <- results[results$gene == gene, ]
    legend("topleft", 
           legend = c(sprintf("R² = %.3f", gene_res$r_squared),
                     sprintf("p = %.2e", gene_res$adj_p_value),
                     sprintf("n = %d", gene_res$n_samples)),
           bty = "n", cex = 0.9)
  }
  
  mtext("Top Genes Associated with Cell Count", 
        outer = TRUE, cex = 1.2, font = 2)
  
  if (!is.null(output_file)) {
    dev.off()
    cat(sprintf("\nTop genes figure saved to: %s\n", output_file))
  }
}

plot_top_genes_cellcount(pb_total, results_cellcount_regression, n_genes = 6,
                         output_file = "../../results/targeted/gene_expression_plots/CellCount_top_genes_pseudobulk.svg",
                         cellcount_column = "cell_count")

cat("\n=== Pseudobulk Cell Count Analysis Complete ===\n")

# Export results to CSV
write.csv(results_cellcount_regression, "../../results/targeted/gene_hormone_csv/cellcount_results_pseudobulk.csv", row.names = FALSE)
```

## Remove Outliers
Removing Samples based on hormone concentration
```{r remove outliers}
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) + 1.5 * IQR(data[[variable]], na.rm = TRUE)
  
  # Identify outliers
  outliers <- data %>% filter(.data[[variable]] > outlier_limit)
  
  # Print outlier information
  if(nrow(outliers) > 0) {
    cat("\n=== Outliers removed for", variable, "===\n")
    cat("Outlier limit:", outlier_limit, "\n")
    cat("Number of outliers removed:", nrow(outliers), "\n")
    print(outliers)
  } else {
    cat("\nNo outliers detected for", variable, "\n")
  }
  
  # Return filtered data
  data %>% filter(.data[[variable]] <= outlier_limit)
}

targeted_comparison_filtered_TT <- detect_outliers(targeted_comparison_raw, "TT")
targeted_comparison_filtered_E2 <- detect_outliers(targeted_comparison_raw, "E2")
targeted_comparison_filtered_P4 <- detect_outliers(targeted_comparison_raw, "P4")
```

## Hormone Specific Pseudobulk

```{r generate and filter hormone specific pseudobulk data, warnings = FALSE}

generate_and_filter_pseudobulk <- function(seurat_obj, targeted_filtered) {
  seurat_obj$cell_type <- seurat_obj$cluster_call  # ensure column exists
  seurat_obj$sample_id <- seurat_obj$sample        # ensure column exists
  
  pb <- pseudobulk_data(seurat_obj)
  pb_metadata <- prepare_metadata_for_analysis(pb$metadata, targeted_filtered)
  
  common_ids <- intersect(rownames(pb$logcounts), rownames(pb_metadata))
  pb$logcounts <- pb$logcounts[common_ids, , drop = FALSE]
  pb$counts <- pb$counts[common_ids, , drop = FALSE]
  pb$metadata <- pb_metadata[common_ids, , drop = FALSE]
  
  list(pb = pb, metadata = pb_metadata)
}

pb_TT <- generate_and_filter_pseudobulk(SSD_data_v5, targeted_comparison_filtered_TT)
pb_E2 <- generate_and_filter_pseudobulk(SSD_data_v5, targeted_comparison_filtered_E2)
pb_P4 <- generate_and_filter_pseudobulk(SSD_data_v5, targeted_comparison_filtered_P4)
```

## Create Hormone Specific Pseudobulk Visualization

```{r cell count histograms}
# Function to get global max sample count across all groups and cell types
get_global_y_limit <- function(pb_TT, pb_E2, pb_P4) {
  
  # Get max counts per cell type for each group
  max_counts <- c()
  
  for (pb in list(pb_TT, pb_E2, pb_P4)) {
    cell_types <- unique(pb$pb$metadata$cell_type)
    for (ct in cell_types) {
      ct_data <- pb$pb$metadata %>% filter(cell_type == ct)
      # Count how many samples fall in each bin
      h <- hist(ct_data$cell_count, breaks = 30, plot = FALSE)
      max_counts <- c(max_counts, max(h$counts))
    }
  }
  
  ceiling(max(max_counts))
}
# Function to create histograms of cell counts by cell type
create_cell_count_histograms <- function(pb_result, title_prefix = "", y_limit = NULL) {
  
  # Extract metadata from pseudobulk result
  metadata <- pb_result$pb$metadata
  
  # Get unique cell types
  cell_types <- unique(metadata$cell_type)
  
  # Create histogram for each cell type
  plot_list <- lapply(cell_types, function(ct) {
    
    # Filter data for current cell type
    ct_data <- metadata %>%
      filter(cell_type == ct)
    
    # Create histogram
    p <- ggplot(ct_data, aes(x = cell_count)) +
      geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
      labs(
        title = paste0(title_prefix, ct),
        x = "Cell Count per Sample",
        y = "Number of Samples"
      ) +
      theme_bw() +
      theme(
        plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 9)
      ) +
      geom_vline(aes(xintercept = median(cell_count)), 
                 color = "red", linetype = "dashed", linewidth = 0.8) +
      annotate("text", 
               x = Inf, y = Inf, 
               label = paste0("n = ", nrow(ct_data), "\nMedian = ", round(median(ct_data$cell_count), 0)),
               hjust = 1.1, vjust = 1.5, size = 3) +
      scale_y_continuous(
        breaks = function(x) {
          if (!is.null(y_limit)) {
            # Create breaks based on y_limit
            if (y_limit <= 10) {
              seq(0, y_limit, by = 1)
            } else if (y_limit <= 50) {
              seq(0, y_limit, by = 5)
            } else if (y_limit <= 100) {
              seq(0, y_limit, by = 10)
            } else {
              seq(0, y_limit, by = 20)
            }
          } else {
            pretty(x, n = 10)
          }
        },
        limits = if (!is.null(y_limit)) c(0, y_limit) else NULL,
        expand = c(0, 0)
      )
    
    return(p)
  })
  
  names(plot_list) <- cell_types
  return(plot_list)
}

# Function to create bar plots showing cell count per sample for each cell type
create_sample_bar_plots <- function(pb_result, title_prefix = "", y_limit = NULL, sample_order = NULL) {
  
  # Extract metadata from pseudobulk result
  metadata <- pb_result$pb$metadata
  
  # Get unique cell types
  cell_types <- unique(metadata$cell_type)
  
  # If no sample order provided, create one from the data
  if (is.null(sample_order)) {
    sample_order <- sort(unique(metadata$sample_id))
  }
  
  # Create bar plot for each cell type
  plot_list <- lapply(cell_types, function(ct) {
    
    # Filter data for current cell type
    ct_data <- metadata %>%
      filter(cell_type == ct)
    
    # Create a complete data frame with all samples (fill missing with 0)
    complete_data <- data.frame(sample_id = sample_order) %>%
      left_join(ct_data %>% select(sample_id, cell_count), by = "sample_id") %>%
      mutate(cell_count = ifelse(is.na(cell_count), 0, cell_count))
    
    # Set factor levels to maintain order
    complete_data$sample_id <- factor(complete_data$sample_id, levels = sample_order)
    
    # Create bar plot
    p <- ggplot(complete_data, aes(x = sample_id, y = cell_count)) +
      geom_bar(stat = "identity", fill = "steelblue", color = "black", alpha = 0.7, width = 0.8) +
      labs(
        title = paste0(title_prefix, ct),
        x = "Sample ID",
        y = "Cell Count"
      ) +
      theme_bw() +
      theme(
        plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 9),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
      ) +
      geom_hline(aes(yintercept = median(cell_count[cell_count > 0])), 
                 color = "red", linetype = "dashed", linewidth = 0.8) +
      annotate("text", 
               x = Inf, y = Inf, 
               label = paste0("n = ", sum(complete_data$cell_count > 0), 
                            "\nMedian = ", round(median(complete_data$cell_count[complete_data$cell_count > 0]), 0),
                            "\nMin = ", min(complete_data$cell_count[complete_data$cell_count > 0]),
                            "\nMax = ", max(complete_data$cell_count)),
               hjust = 1.1, vjust = 1.5, size = 3) +
      scale_y_continuous(
        breaks = function(x) {
          if (!is.null(y_limit)) {
            range <- y_limit
            if (range <= 50) {
              seq(0, y_limit, by = 5)
            } else if (range <= 200) {
              seq(0, y_limit, by = 20)
            } else if (range <= 500) {
              seq(0, y_limit, by = 50)
            } else {
              seq(0, y_limit, by = 100)
            }
          } else {
            pretty(x, n = 10)
          }
        },
        limits = if (!is.null(y_limit)) c(0, y_limit) else NULL,
        expand = c(0, 0)
      )
    
    return(p)
  })
  
  names(plot_list) <- cell_types
  return(plot_list)
}

# Function to get global max cell count for consistent y-axis on sample bar plots
get_global_cell_count_limit <- function(pb_TT, pb_E2, pb_P4) {
  all_counts <- c(
    pb_TT$pb$metadata$cell_count,
    pb_E2$pb$metadata$cell_count,
    pb_P4$pb$metadata$cell_count
  )
  
  ceiling(max(all_counts))
}

# Function to get consistent sample order across all groups
get_sample_order <- function(pb_TT, pb_E2, pb_P4) {
  all_samples <- unique(c(
    pb_TT$pb$metadata$sample_id,
    pb_E2$pb$metadata$sample_id,
    pb_P4$pb$metadata$sample_id
  ))
  
  sort(all_samples)
}

# Function to create a combined summary plot
create_summary_boxplot <- function(pb_result, title_prefix = "") {
  
  metadata <- pb_result$pb$metadata
  
  p <- ggplot(metadata, aes(x = reorder(cell_type, cell_count, FUN = median), 
                             y = cell_count, fill = cell_type)) +
    geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
    coord_flip() +
    labs(
      title = paste0(title_prefix, "Cell Count Distribution Across Cell Types"),
      x = "Cell Type",
      y = "Cell Count per Sample"
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      legend.position = "none"
    ) +
    scale_fill_viridis_d(option = "turbo") +
    scale_y_continuous(breaks = function(x) {
      range <- diff(range(x))
      if (range <= 50) {
        seq(floor(x[1]), ceiling(x[2]), by = 5)
      } else if (range <= 200) {
        seq(floor(x[1]), ceiling(x[2]), by = 20)
      } else if (range <= 500) {
        seq(floor(x[1]), ceiling(x[2]), by = 50)
      } else {
        seq(floor(x[1]), ceiling(x[2]), by = 100)
      }
    })
  
  return(p)
}

# Function to save plots as SVG files
save_plots_svg <- function(plot_list, prefix, output_dir) {
  # Create output directory
  dir.create(output_dir, showWarnings = FALSE)
  
  # Save each plot as individual SVG file
  for (cell_type in names(plot_list)) {
    # Create safe filename (replace spaces and special characters)
    safe_filename <- gsub("[^A-Za-z0-9_]", "_", cell_type)
    filename <- file.path(output_dir, paste0(prefix, "_", safe_filename, ".svg"))
    
    ggsave(filename, plot_list[[cell_type]], width = 10, height = 8, device = "svg")
    cat("Saved:", filename, "\n")
  }
}

# Function to calculate and return summary statistics
calculate_summary_stats <- function(pb_result) {
  pb_result$pb$metadata %>%
    group_by(cell_type) %>%
    summarise(
      n_samples = n(),
      mean_count = mean(cell_count),
      median_count = median(cell_count),
      sd_count = sd(cell_count),
      min_count = min(cell_count),
      max_count = max(cell_count),
      samples_below_25 = sum(cell_count < 25),
      pct_below_25 = 100 * samples_below_25 / n()
    ) %>%
    arrange(desc(median_count))
}

# Get global y-axis limit for histograms
global_y_limit <- get_global_y_limit(pb_TT, pb_E2, pb_P4)
cat("Global histogram y-axis limit set to:", global_y_limit, "\n")

# Get global y-axis limit for sample bar plots
global_cell_count_limit <- get_global_cell_count_limit(pb_TT, pb_E2, pb_P4)
cat("Global cell count y-axis limit set to:", global_cell_count_limit, "\n\n")

# Get consistent sample order across all groups
sample_order <- get_sample_order(pb_TT, pb_E2, pb_P4)
cat("Number of unique samples:", length(sample_order), "\n\n")

# Generate histograms for each treatment group with consistent y-axis
histograms_TT <- create_cell_count_histograms(pb_TT, "TT - ", global_y_limit)
histograms_E2 <- create_cell_count_histograms(pb_E2, "E2 - ", global_y_limit)
histograms_P4 <- create_cell_count_histograms(pb_P4, "P4 - ", global_y_limit)

# Generate sample bar plots for each treatment group with consistent y-axis
sample_bars_TT <- create_sample_bar_plots(pb_TT, "TT - ", y_limit = NULL, sample_order)
sample_bars_E2 <- create_sample_bar_plots(pb_E2, "E2 - ", y_limit = NULL, sample_order)
sample_bars_P4 <- create_sample_bar_plots(pb_P4, "P4 - ", y_limit = NULL, sample_order)

# Generate summary boxplots
summary_TT <- create_summary_boxplot(pb_TT, "TT - ")
summary_E2 <- create_summary_boxplot(pb_E2, "E2 - ")
summary_P4 <- create_summary_boxplot(pb_P4, "P4 - ")

# Display summary plots
print(summary_TT)
print(summary_E2)
print(summary_P4)

# Save all histograms as SVG files
save_histograms_svg(histograms_TT, "TT_histogram", "../../results/targeted/pseudobulk_testosterone_plots/cell_count_histograms_TT")
save_histograms_svg(histograms_E2, "E2_histogram", "../../results/targeted/pseudobulk_estrogen_plots/cell_count_histograms_E2")
save_histograms_svg(histograms_P4, "P4_histogram", "../../results/targeted/pseudobulk_progesterone_plots/cell_count_histograms_P4")

# Save all sample bar plots as SVG files
save_plots_svg(sample_bars_TT, "TT_sample_bars", "../../results/targeted/pseudobulk_testosterone_plots/cell_count_sample_bars_TT")
save_plots_svg(sample_bars_E2, "E2_sample_bars", "../../results/targeted/pseudobulk_estrogen_plots/cell_count_sample_bars_E2")
save_plots_svg(sample_bars_P4, "P4_sample_bars", "../../results/targeted/pseudobulk_progesterone_plots/cell_count_sample_bars_P4")

# Save summary plots as SVG
ggsave("cell_count_summary_TT.svg", summary_TT, path = "../../results/targeted/pseudobulk_testosterone_plots/cell_count_histograms_TT", width = 10, height = 8, device = "svg")
ggsave("cell_count_summary_E2.svg", summary_E2, path = "../../results/targeted/pseudobulk_estrogen_plots/cell_count_histograms_E2", width = 10, height = 8, device = "svg")
ggsave("cell_count_summary_P4.svg", summary_P4, path = "../../results/targeted/pseudobulk_progesterone_plots/cell_count_histograms_P4", width = 10, height = 8, device = "svg")

# Calculate summary statistics for each group
cat("\n=== Calculating Summary Statistics ===\n")
stats_TT <- calculate_summary_stats(pb_TT)
stats_E2 <- calculate_summary_stats(pb_E2)
stats_P4 <- calculate_summary_stats(pb_P4)

# Print summary statistics for each group
cat("\n=== TT Group Summary Statistics ===\n")
print(stats_TT)

cat("\n=== E2 Group Summary Statistics ===\n")
print(stats_E2)

cat("\n=== P4 Group Summary Statistics ===\n")
print(stats_P4)

# Save summary statistics as CSV files
write.csv(stats_TT, "../../results/targeted/gene_hormone_csv/cell_count_summary_stats_TT.csv", row.names = FALSE)
write.csv(stats_E2, "../../results/targeted/gene_hormone_csv/cell_count_summary_stats_E2.csv", row.names = FALSE)
write.csv(stats_P4, "../../results/targeted/gene_hormone_csv/cell_count_summary_stats_P4.csv", row.names = FALSE)

```

# Differential Gene Expression Analysis
Note: the default min samples is increased form 8 from single-sex analysis to 10 for both sexes, since including both sexes should increase numbers of samples for each observation. 

## Define Functions
### Filtering Function
```{r filtering function}
add_outlier_analyze_hormone <- function(pb_counts, metadata_merged, hormone, overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, model = run_voom_lm) {
   # Split the metadata based on cell type
  split_ids <- split(rownames(metadata_merged), metadata_merged$cell_type)
  
  results_by_group <- purrr::map(split_ids, function(ids) {
    
    # Subset counts and metadata based on ids
    counts_sub <- t(pb_counts)[, ids, drop = FALSE]
    metadata_sub <- metadata_merged[ids, , drop = FALSE]
    
    initial_sample_size <- nrow(metadata_sub)
    cat("\nGroup:", metadata_sub$cell_type[1], "\n")
    cat("Initial number of samples:", initial_sample_size, "\n")
    
    # Check if the group has a minimum number of samples
    if (nrow(metadata_sub) < overall_min_samples) {
      cat("Group removed: insufficient samples", nrow(metadata_sub), "\n")
      return(NULL)
    }
    
    # Gene-level filtering to retain genes with counts above certain thresholds
    keep_genes <- colSums(counts_sub >= count_threshold) >= gene_sample_threshold
    
    counts_sub <- counts_sub[, keep_genes, drop = FALSE]
    
    # Check if any genes remain after filtering
    if (ncol(counts_sub) == 0) {
      cat("Group removed: no genes passed the filtering threshold\n")
      return(NULL)
    }
    
    # Calculate the outlier threshold based on hormone values
    outlier_concentration <- quantile(metadata_sub[[hormone]], 0.75, na.rm = TRUE) + 1.5 * IQR(metadata_sub[[hormone]], na.rm = TRUE)
    cat("Outlier concentration threshold:", outlier_concentration, "\n")
    
    # Filter out the outliers based on hormone concentration
    non_outliers <- metadata_sub[[hormone]] < outlier_concentration
    counts_sub_non_outliers <- counts_sub[, non_outliers, drop = FALSE]
    metadata_sub_non_outliers <- metadata_sub[non_outliers, , drop = FALSE]
    
    cat("Samples after outlier removal:", nrow(metadata_sub_non_outliers), "\n")
    
    # Check if enough samples remain after removing outliers
    if (nrow(metadata_sub_non_outliers) < overall_min_samples) {
      cat("Group removed: insufficient samples after outlier removal", nrow(metadata_sub_non_outliers), "\n")
      return(NULL)
    }
    
    # Mask for non-zero and non-NA hormone values
    nonzero_hormone_mask <- metadata_sub_non_outliers[[hormone]] != 0 & !is.na(metadata_sub_non_outliers[[hormone]])
    cat("Samples with non-zero hormone:", sum(nonzero_hormone_mask), "\n")
    
    # Check if enough samples have non-zero hormone values
    if (sum(nonzero_hormone_mask) < overall_min_samples) {
      cat("Group removed: insufficient non-zero hormone values", sum(nonzero_hormone_mask), "\n")
      return(NULL)
    }
    
    # Subset counts and metadata to only those with non-zero hormone values
    #final_counts_sub <- counts_sub_non_outliers[, nonzero_hormone_mask, drop = FALSE]
    #final_metadata_sub <- metadata_sub_non_outliers[nonzero_hormone_mask, , drop = FALSE]
    
    cat("Final sample size:", nrow(metadata_sub_non_outliers), "\n")
    
    # Run the modeling function
    result <- model(counts_sub_non_outliers, metadata_sub_non_outliers, hormone)
    
    # Verify if the result is NULL
    if (is.null(result)) {
      cat("Model returned NULL result\n")
      return(NULL)
    }
    
    # Append additional metadata to the result
    result$cell_type <- metadata_sub_non_outliers$cell_type[1]
    result$sex <- metadata_sub_non_outliers$Sex[1]
    
    # Return the result including the filtered data used for modeling
    list(
      limma_results = result,
      counts_sub = counts_sub_non_outliers,
      metadata_sub = metadata_sub_non_outliers
    )
  })
  
  # Remove NULL groups from results_by_group
  results_by_group <- results_by_group[!sapply(results_by_group, is.null)]
  
  # Combine all limma results into one data.frame
  combined_results <- purrr::map_dfr(results_by_group, ~ .x$limma_results)
  
  # Check if combined results are empty
  if (nrow(combined_results) == 0) {
    cat("No significant results found\n")
  }
  
  # Return the combined results and data by group
  list(
    combined_results = combined_results,
    by_group_data = results_by_group
  )
}

analyze_hormone <- function(pb_counts, metadata_merged, hormone, overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, model = run_voom_lm) {
   # Split the metadata based on cell type
  split_ids <- split(rownames(metadata_merged), metadata_merged$cell_type)
  
  results_by_group <- purrr::map(split_ids, function(ids) {
    
    # Subset counts and metadata based on ids
    counts_sub <- t(pb_counts)[, ids, drop = FALSE]
    metadata_sub <- metadata_merged[ids, , drop = FALSE]
    
    initial_sample_size <- nrow(metadata_sub)
    cat("\nGroup:", metadata_sub$cell_type[1], "\n")
    cat("Initial number of samples:", initial_sample_size, "\n")
    
    # Check if the group has a minimum number of samples
    if (nrow(metadata_sub) < overall_min_samples) {
      cat("Group removed: insufficient samples", nrow(metadata_sub), "\n")
      return(NULL)
    }
    
    # Gene-level filtering to retain genes with counts above certain thresholds
    keep_genes <- colSums(counts_sub >= count_threshold) >= gene_sample_threshold
    
    counts_sub <- counts_sub[, keep_genes, drop = FALSE]
    
    # Check if any genes remain after filtering
    if (ncol(counts_sub) == 0) {
      cat("Group removed: no genes passed the filtering threshold\n")
      return(NULL)
    }
    
    # Mask for non-zero and non-NA hormone values
    nonzero_hormone_mask <- metadata_sub[[hormone]] != 0 & !is.na(metadata_sub[[hormone]])
    cat("Samples with non-zero hormone:", sum(nonzero_hormone_mask), "\n")
    
    # Check if enough samples have non-zero hormone values
    if (sum(nonzero_hormone_mask) < overall_min_samples) {
      cat("Group removed: insufficient non-zero hormone values", sum(nonzero_hormone_mask), "\n")
      return(NULL)
    }
    
    # Subset counts and metadata to only those with non-zero hormone values
    #final_counts_sub <- counts_sub[, nonzero_hormone_mask, drop = FALSE]
    #final_metadata_sub <- metadata_sub[nonzero_hormone_mask, , drop = FALSE]
    
    cat("Final sample size:", nrow(metadata_sub), "\n")
    
    # Run the modeling function
    result <- model(counts_sub, metadata_sub, hormone)
    
    # Verify if the result is NULL
    if (is.null(result)) {
      cat("Model returned NULL result\n")
      return(NULL)
    }
    
    # Append additional metadata to the result
    result$cell_type <- metadata_sub$cell_type[1]
    
    # Return the result including the filtered data used for modeling
    list(
      limma_results = result,
      counts_sub = counts_sub,
      metadata_sub = metadata_sub
    )
  })
  
  # Remove NULL groups from results_by_group
  results_by_group <- results_by_group[!sapply(results_by_group, is.null)]
  
  # Combine all limma results into one data.frame
  combined_results <- purrr::map_dfr(results_by_group, ~ .x$limma_results)
  
  # Check if combined results are empty
  if (nrow(combined_results) == 0) {
    cat("No significant results found\n")
  }
  
  # Return the combined results and data by group
  list(
    combined_results = combined_results,
    by_group_data = results_by_group
  )
}
```

### Plotting Function
```{r plotting-function, echo = TRUE, message = FALSE}
plot_top_genes_from_subsets <- function(by_group_data, output_dir, hormone, top_n = 6) {
  dir.create(output_dir, showWarnings = FALSE)
  
  total_groups <- length(by_group_data)
  message("Number of groups to plot: ", total_groups)
  flush.console()
  
  purrr::walk(names(by_group_data), function(group_name) {
    group_list <- by_group_data[[group_name]]
    
    message("Processing group: ", group_name)
    flush.console()
    
    if (is.null(group_list)) {
      message("  Skipping NULL group", group_name)
      flush.console()
      return()
    }
    
    limma_res <- group_list$limma_results
    counts_sub <- group_list$counts_sub  # genes x samples
    metadata_sub <- group_list$metadata_sub

    if (nrow(metadata_sub) < 1 || ncol(counts_sub) < 1) {
      message("  Skipping empty subset: ", group_name, ifelse(!is.null(metadata_sub$cell_type), paste0(", cell type: ", metadata_sub$cell_type[1]), ""))
      flush.console()
      return()
    }

    message("  Processing subset: ", group_name, ", cell type: ", metadata_sub$cell_type[1])
    flush.console()
    
    # Compute log1p(CPM)
    counts_cpm <- edgeR::cpm(counts_sub, log = FALSE)
    counts_log <- log1p(counts_cpm)
    
    # Transpose counts: samples x genes
    counts_log <- t(counts_log)

    top_genes <- tryCatch({
      limma_res %>%
        arrange(-abs(t)) %>%
        slice_head(n = top_n) %>%
        filter(adj.P.Val < 0.05) %>%
        pull(gene)
    }, error = function(e) {
      message("  Error encountered while extracting top genes for group: ", group_name)
      message("  Error: ", e)
      flush.console()
      return(character(0))
    })

    top_genes_pvals <- tryCatch({
      limma_res %>%
        filter(gene %in% top_genes) %>%
        select(gene, adj.P.Val) %>%
        as.data.frame()
    }, error = function(e) {
      message("  Error encountered while extracting p-values for group: ", group_name)
      message("  Error: ", e)
      flush.console()
      return(data.frame(gene = character(0), adj.P.Val = numeric(0)))
    })
    
    if (length(top_genes) == 0) {
      message("  No top genes found for subset: ", group_name)
      flush.console()
      return()
    }

    message("  Plotting group: ", group_name, ", cell type: ", metadata_sub$cell_type[1])
    flush.console()
    message("  Top genes: ", paste(top_genes, collapse = ", "))
    flush.console()

    plot_list <- purrr::map(seq_len(nrow(top_genes_pvals)), function(i) {
      gene <- top_genes_pvals$gene[i]
      adj_pval <- top_genes_pvals$adj.P.Val[i]
      
      if (!gene %in% colnames(counts_log)) {
        message("    Gene ", gene, " not in counts — skipping")
        flush.console()
        return(NULL)
      }

      expr_vals <- counts_log[, gene]
      if (all(is.na(expr_vals)) || all(expr_vals == 0)) {
        message("    Gene ", gene, " has all NA or zero expression — skipping")
        flush.console()
        return(NULL)
      }

      message("    Plotting gene: ", gene, " (adj. p-value: ", format(adj_pval, digits = 3), ")")
      flush.console()
      
      df <- metadata_sub %>% mutate(expr = expr_vals)
      
      p <- ggplot(df, aes(x = !!sym(hormone), y = expr, color = cell_count)) +
        geom_point(size = 2, alpha = 0.8) +
        geom_smooth(method = "lm", se = TRUE) +
        scale_color_viridis_c() +
        labs(title = paste0(gene, " (adj. p-value: ", format(adj_pval, digits = 3), ") in ", metadata_sub$cell_type[1]),
             x = hormone, y = "log1p(CPM)") +
        theme_minimal(base_size = 14)
      
      p
    }) %>% compact()

    if (length(plot_list) == 0) {
      message("    No valid plots for group ", group_name, ", cell type: ")
      flush.console()
      return()
    }

    combined_plot <- tryCatch({
      wrap_plots(plotlist = plot_list, ncol = 2)
    }, error = function(e) {
      message("  Error encountered while wrapping plots: ", e)
      flush.console()
      return(NULL)
    })

    if (!is.null(combined_plot)) {
      clean_name <- gsub("/", "-", metadata_sub$cell_type[1])
      tryCatch({
        ggsave(file.path(output_dir, paste0(clean_name, "_both_sex_top_genes.png")),
               plot = combined_plot, width = 10, height = 6 + ceiling(length(plot_list) / 2), dpi = 300)
        message("  ✅ Saved plot for ", clean_name)
        flush.console()
      }, error = function(e) {
        message("  Error encountered while saving plot for ", clean_name, ": ", e)
        flush.console()
      })
    } else {
      message("  Failed to wrap plots for group: ", group_name)
    }
  })
}

```

### Summary Functions
```{r summarize and save results}
summarize_sig_genes <- function(results, fdr_cutoff = 0.05, logfc_cutoff = 0.25) {
  sig <- results %>% filter(adj.P.Val < fdr_cutoff)
  if (!is.null(logfc_cutoff)) {
    sig <- sig %>% filter(abs(logFC) >= logfc_cutoff)
  }
  sig %>%
    count(cell_type, name = "num_sig_genes") %>%
    arrange(desc(num_sig_genes))
}

plot_sig_gene_summary <- function(summary_df, title) {
  ggplot(summary_df, aes(x = cell_type, y = num_sig_genes)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(title = title, x = "Cell Type", y = "# Significant Genes") +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```


# Run DE analyses with limma and filtering parameters
## Testosterone
```{r testosterone analysis}
# TT - Run analysis
res_TT_1 <- analyze_hormone(pb_TT$pb$counts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, model = run_voom_lm_cc_gw)

res_TT_2 <- analyze_hormone(pb_TT$pb$counts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, model = run_voom_lm_cc)

res_TT_3 <- analyze_hormone(pb_TT$pb$counts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, model = run_voom_lm)

res_TT_4 <- analyze_hormone(pb_TT$pb$counts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 5, gene_sample_threshold = 3, model = run_voom_lm)

res_TT_5 <- add_outlier_analyze_hormone(pb_TT$pb$counts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 5, gene_sample_threshold = 3, model = run_voom_lm)

res_TT_6 <- add_outlier_analyze_hormone(pb_TT$pb$counts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 50, gene_sample_threshold = 10, model = run_voom_lm)

# TT - Plot top genes
plot_top_genes_from_subsets(res_TT_1$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group1", hormone = "TT")
plot_top_genes_from_subsets(res_TT_2$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group2", hormone = "TT")
plot_top_genes_from_subsets(res_TT_3$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group3", hormone = "TT")
plot_top_genes_from_subsets(res_TT_4$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group4", hormone = "TT")
plot_top_genes_from_subsets(res_TT_5$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group5", hormone = "TT")

#Summary
group1_summary_TT <- summarize_sig_genes(res_TT_1$combined_results)
group2_summary_TT <- summarize_sig_genes(res_TT_2$combined_results)
group3_summary_TT <- summarize_sig_genes(res_TT_3$combined_results)
group4_summary_TT <- summarize_sig_genes(res_TT_4$combined_results)
group5_summary_TT <- summarize_sig_genes(res_TT_5$combined_results)
group6_summary_TT <- summarize_sig_genes(res_TT_6$combined_results)


#Gene Summary Plot
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group1/both_sex_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group1_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group2/both_sex_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group2_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group3/both_sex_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group3_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group4/both_sex_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group4_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group5/both_sex_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group5_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group6/both_sex_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group6_summary_TT, "Testosterone"), width = 8, height = 5)

#Save Result CSV
write_csv(res_TT_1$combined_results, "../../results/targeted/gene_hormone_csv/Group1_both_sex_TT_results.csv")
write_csv(res_TT_2$combined_results, "../../results/targeted/gene_hormone_csv/Group2_both_sex_TT_results.csv")
write_csv(res_TT_3$combined_results, "../../results/targeted/gene_hormone_csv/Group3_both_sex_TT_results.csv")
write_csv(res_TT_4$combined_results, "../../results/targeted/gene_hormone_csv/Group4_both_sex_TT_results.csv")
write_csv(res_TT_5$combined_results, "../../results/targeted/gene_hormone_csv/Group5_both_sex_TT_results.csv")

#Save Summary CSV
write_csv(group1_summary_TT, "../../results/targeted/gene_hormone_csv/Group1_both_sex_TT_summary.csv")
write_csv(group2_summary_TT, "../../results/targeted/gene_hormone_csv/Group2_both_sex_TT_summary.csv")
write_csv(group3_summary_TT, "../../results/targeted/gene_hormone_csv/Group3_both_sex_TT_summary.csv")
write_csv(group4_summary_TT, "../../results/targeted/gene_hormone_csv/Group4_both_sex_TT_summary.csv")
write_csv(group5_summary_TT, "../../results/targeted/gene_hormone_csv/Group5_both_sex_TT_summary.csv")


```

## Estradiol
```{r estradiol analysis}
# E2 - Run analysis
res_E2_1 <- analyze_hormone(pb_E2$pb$counts, pb_E2$pb$logcounts, pb_E2$metadata, hormone = "E2", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc_gw)

res_E2_2 <- analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc)

res_E2_3 <- analyze_hormone(pb_E2$pb$counts, pb_E2$pb$logcounts, pb_E2$metadata, hormone = "E2", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm)

res_E2_4 <- analyze_hormone(pb_E2$pb$counts, pb_E2$pb$logcounts, pb_E2$metadata, hormone = "E2", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

res_E2_5 <- add_outlier_analyze_hormone(pb_E2$pb$counts, pb_E2$pb$logcounts, pb_E2$metadata, hormone = "E2", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

# E2 - Plot top genes
plot_top_genes_from_subsets(res_E2_1$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group1", hormone = "E2")
plot_top_genes_from_subsets(res_E2_2$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group2", hormone = "E2")
plot_top_genes_from_subsets(res_E2_3$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group3", hormone = "E2")
plot_top_genes_from_subsets(res_E2_4$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group4", hormone = "E2")
plot_top_genes_from_subsets(res_E2_5$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group5", hormone = "E2")

#Summary
group1_summary_E2 <- summarize_sig_genes(res_E2_1$combined_results)
group2_summary_E2 <- summarize_sig_genes(res_E2_2$combined_results)
group3_summary_E2 <- summarize_sig_genes(res_E2_3$combined_results)
group4_summary_E2 <- summarize_sig_genes(res_E2_4$combined_results)
group5_summary_E2 <- summarize_sig_genes(res_E2_5$combined_results)

#Gene Summary Plot
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group1/both_sex_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group1_summary_E2, "Estradiol"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group2/both_sex_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group2_summary_E2, "Estradiol"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group3/both_sex_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group3_summary_E2, "Estradiol"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group4/both_sex_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group4_summary_E2, "Estradiol"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group5/both_sex_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group5_summary_E2, "Estradiol"), width = 8, height = 5)

#Save Result CSV
write_csv(res_E2_1$combined_results, "../../results/targeted/gene_hormone_csv/Group1_both_sex_E2_results.csv")
write_csv(res_E2_2$combined_results, "../../results/targeted/gene_hormone_csv/Group2_both_sex_E2_results.csv")
write_csv(res_E2_3$combined_results, "../../results/targeted/gene_hormone_csv/Group3_both_sex_E2_results.csv")
write_csv(res_E2_4$combined_results, "../../results/targeted/gene_hormone_csv/Group4_both_sex_E2_results.csv")
write_csv(res_E2_5$combined_results, "../../results/targeted/gene_hormone_csv/Group5_both_sex_E2_results.csv")

#Save Summary CSV
write_csv(group1_summary_E2, "../../results/targeted/gene_hormone_csv/Group1_both_sex_E2_summary.csv")
write_csv(group2_summary_E2, "../../results/targeted/gene_hormone_csv/Group2_both_sex_E2_summary.csv")
write_csv(group3_summary_E2, "../../results/targeted/gene_hormone_csv/Group3_both_sex_E2_summary.csv")
write_csv(group4_summary_E2, "../../results/targeted/gene_hormone_csv/Group4_both_sex_E2_summary.csv")
write_csv(group5_summary_E2, "../../results/targeted/gene_hormone_csv/Group5_both_sex_E2_summary.csv")

```

## Progesterone
```{r progesterone analysis}
# P4 - Run analysis
res_P4_1 <- analyze_hormone(pb_P4$pb$counts, pb_P4$pb$logcounts, pb_P4$metadata, hormone = "P4", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc_gw)

res_P4_2 <- analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc)

res_P4_3 <- analyze_hormone(pb_P4$pb$counts, pb_P4$pb$logcounts, pb_P4$metadata, hormone = "P4", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm)

res_P4_4 <- analyze_hormone(pb_P4$pb$counts, pb_P4$pb$logcounts, pb_P4$metadata, hormone = "P4", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

res_P4_5 <- add_outlier_analyze_hormone(pb_P4$pb$counts, pb_P4$pb$logcounts, pb_P4$metadata, hormone = "P4", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

# P4 - Plot top genes
plot_top_genes_from_subsets(res_P4_1$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group1", hormone = "P4")
plot_top_genes_from_subsets(res_P4_2$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group2", hormone = "P4")
plot_top_genes_from_subsets(res_P4_3$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group3", hormone = "P4")
plot_top_genes_from_subsets(res_P4_4$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group4", hormone = "P4")
plot_top_genes_from_subsets(res_P4_5$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group5", hormone = "P4")

#Summary
group1_summary_P4 <- summarize_sig_genes(res_P4_1$combined_results)
group2_summary_P4 <- summarize_sig_genes(res_P4_2$combined_results)
group3_summary_P4 <- summarize_sig_genes(res_P4_3$combined_results)
group4_summary_P4 <- summarize_sig_genes(res_P4_4$combined_results)
group5_summary_P4 <- summarize_sig_genes(res_P4_5$combined_results)

#Gene Summary Plot
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group1/both_sex_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group1_summary_P4, "Progesterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group2/both_sex_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group2_summary_P4, "Progesterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group3/both_sex_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group3_summary_P4, "Progesterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group4/both_sex_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group4_summary_P4, "Progesterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group5/both_sex_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group5_summary_P4, "Progesterone"), width = 8, height = 5)

#Save Result CSV
write_csv(res_P4_1$combined_results, "../../results/targeted/gene_hormone_csv/Group1_both_sex_P4_results.csv")
write_csv(res_P4_2$combined_results, "../../results/targeted/gene_hormone_csv/Group2_both_sex_P4_results.csv")
write_csv(res_P4_3$combined_results, "../../results/targeted/gene_hormone_csv/Group3_both_sex_P4_results.csv")
write_csv(res_P4_4$combined_results, "../../results/targeted/gene_hormone_csv/Group4_both_sex_P4_results.csv")
write_csv(res_P4_5$combined_results, "../../results/targeted/gene_hormone_csv/Group5_both_sex_P4_results.csv")

#Save Summary CSV
write_csv(group1_summary_P4, "../../results/targeted/gene_hormone_csv/Group1_both_sex_P4_summary.csv")
write_csv(group2_summary_P4, "../../results/targeted/gene_hormone_csv/Group2_both_sex_P4_summary.csv")
write_csv(group3_summary_P4, "../../results/targeted/gene_hormone_csv/Group3_both_sex_P4_summary.csv")
write_csv(group4_summary_P4, "../../results/targeted/gene_hormone_csv/Group4_both_sex_P4_summary.csv")
write_csv(group5_summary_P4, "../../results/targeted/gene_hormone_csv/Group5_both_sex_P4_summary.csv")
```

# End
