---
title: "Gonadal Hormone and Gene Expression Relationship Analysis - Seperate by Chromosomal Sex"
author: "Emma Jones, Sneha Chaturvedi"
date: "Last updated 10.28.25"
output: 
  html_document:
    toc: TRUE
---
The purpose of this script is to 1) explore single cell RNA-seq results and 2) analyze for relationships between gene expression and testosterone, progesterone, and estradiol concentrations from developing human brain samples. This script analyzes males and females separately. 

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Packages <- c("tidyverse", "Seurat", "Signac", "glue", "readxl", "cowplot", "edgeR", "limma", "patchwork", "svglite", "Matrix", "ggpubr")

lapply(Packages, library, character.only=TRUE)

set.seed(123)
```

## Load Data
```{r load data}
SSD_data_v5 <- readRDS("../../data/SSD_data_v5.rds")
SSD_data_v5 <- subset(SSD_data_v5, subset = Droplet == "SNG")

targeted_comparison_raw <- read_xlsx("../../doc/targeted/targeted_hormones.xlsx")
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`
```

## Remove Outliers
```{r remove outliers}
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) + 1.5 * IQR(data[[variable]], na.rm = TRUE)
  data %>% filter(.data[[variable]] <= outlier_limit)
}

targeted_comparison_filtered_TT <- detect_outliers(targeted_comparison_raw, "TT")
targeted_comparison_filtered_E2 <- detect_outliers(targeted_comparison_raw, "E2")
targeted_comparison_filtered_P4 <- detect_outliers(targeted_comparison_raw, "P4")
```


# Pseudobulk
## Metadata and Pseudobulk Functions

```{r metadata and pseudobulk}
prepare_metadata_for_analysis <- function(pb_metadata, targeted_comparison_filtered) {
  targeted_comparison_filtered$Sample <- paste0("tissue", targeted_comparison_filtered$Sample)
  pb_metadata %>%
    as.data.frame() %>%
    rownames_to_column("pseudobulk_id") %>%
    left_join(targeted_comparison_filtered %>% select(Sample, Sex, TT, E2, P4, GW),
              by = c("sample_id" = "Sample")) %>%
    filter(!is.na(Sex), cell_count >= 10) %>%
    mutate(group = paste(cell_type, Sex, sep = "_")) %>%
    column_to_rownames("pseudobulk_id")
}

pseudobulk_data <- function(seurat_obj, assay = "RNA", layer = "counts", sample_col = "sample_id", celltype_col = "cell_type") {
  counts <- GetAssayData(seurat_obj, assay = assay, slot = layer)
  meta <- seurat_obj@meta.data
  meta$cell_type <- meta$cluster_call
  group_ids <- interaction(meta[[sample_col]], meta[[celltype_col]], drop = TRUE)
  group_levels <- levels(group_ids)
  aggregated_list <- list()
  metadata_list <- list()
  
  for (gid in group_levels) {
  group_cells <- which(group_ids == gid)
  if (length(group_cells) < 1) next
  group_counts <- Matrix::rowSums(counts[, group_cells, drop = FALSE])
  aggregated_list[[gid]] <- group_counts
    metadata_list[[gid]] <- data.frame(
      sample_id = meta[[sample_col]][group_cells[1]],
      cell_type = meta[[celltype_col]][group_cells[1]],
      cell_count = length(group_cells),
      row.names = gid
    )
  }
  
  pseudobulk_mat <- do.call(rbind, aggregated_list)
  metadata_df <- do.call(rbind, metadata_list)
  
  pseudobulk_mat_log <- log1p(as.matrix(pseudobulk_mat))
  
  list(counts = pseudobulk_mat, logcounts = pseudobulk_mat_log, metadata = metadata_df)
}
```

## Generate and Filter Pseudobulk Data

```{r generate and filter pseudobulk data, warnings = FALSE}

generate_and_filter_pseudobulk <- function(seurat_obj, targeted_filtered) {
  seurat_obj$cell_type <- seurat_obj$cluster_call  # ensure column exists
  seurat_obj$sample_id <- seurat_obj$sample        # ensure column exists
  
  pb <- pseudobulk_data(seurat_obj)
  pb_metadata <- prepare_metadata_for_analysis(pb$metadata, targeted_filtered)
  
  common_ids <- intersect(rownames(pb$logcounts), rownames(pb_metadata))
  pb$logcounts <- pb$logcounts[common_ids, , drop = FALSE]
  pb$counts <- pb$counts[common_ids, , drop = FALSE]
  pb$metadata <- pb_metadata[common_ids, , drop = FALSE]
  
  list(pb = pb, metadata = pb_metadata)
}

pb_TT <- generate_and_filter_pseudobulk(SSD_data_v5, targeted_comparison_filtered_TT)
pb_E2 <- generate_and_filter_pseudobulk(SSD_data_v5, targeted_comparison_filtered_E2)
pb_P4 <- generate_and_filter_pseudobulk(SSD_data_v5, targeted_comparison_filtered_P4)
```

# Differential Gene Expression Analysis

## Define Functions
### DE analysis wrapper functions
```{r limma analysis}
run_voom_lm_cc <- function(counts, metadata, predictor) {
  stopifnot(ncol(counts) == nrow(metadata))

  # Safely quote variables for formula
  safe_vars <- function(x) {
    ifelse(make.names(x) != x, paste0("`", x, "`"), x)
  }

  # Build model matrix
  formula_vars <- safe_vars(c(predictor, "cell_count"))
  design <- model.matrix(reformulate(formula_vars), data = metadata)

  # Fit voom-limma model
  v <- limma::voom(counts, design)
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)

  # Figure out correct coefficient name or index
  predictor_col <- colnames(design)[2]  # assumes predictor is 1st after intercept
  limma::topTable(fit, coef = predictor_col, number = Inf, sort.by = "none") %>%
    rownames_to_column("gene")
}

run_voom_lm_cc_gw <- function(counts, metadata, predictor) {
  stopifnot(ncol(counts) == nrow(metadata))
  if (!"GW" %in% colnames(metadata)) metadata$GW <- NA

  # Safely quote variables for formula
  safe_vars <- function(x) {
    ifelse(make.names(x) != x, paste0("`", x, "`"), x)
  }

  # Build model matrix
  formula_vars <- safe_vars(c(predictor, "cell_count", "GW"))
  design <- model.matrix(reformulate(formula_vars), data = metadata)

  # Fit voom-limma model
  v <- limma::voom(counts, design)
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)

  # Figure out correct coefficient name or index
  predictor_col <- colnames(design)[2]  # assumes predictor is 1st after intercept
  limma::topTable(fit, coef = predictor_col, number = Inf, sort.by = "none") %>%
    rownames_to_column("gene")
}

run_voom_lm <- function(counts, metadata, predictor) {
  stopifnot(ncol(counts) == nrow(metadata))

  # Safely quote variables for formula
  safe_vars <- function(x) {
    ifelse(make.names(x) != x, paste0("`", x, "`"), x)
  }

  # Build model matrix
  formula_vars <- safe_vars(c(predictor))
  design <- model.matrix(reformulate(formula_vars), data = metadata)

  # Fit voom-limma model
  v <- limma::voom(counts, design)
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)

  # Figure out correct coefficient name or index
  predictor_col <- colnames(design)[2]  # assumes predictor is 1st after intercept
  limma::topTable(fit, coef = predictor_col, number = Inf, sort.by = "none") %>%
    rownames_to_column("gene")
}

```


### Filtering Function
```{r filtering functions}
add_outlier_analyze_hormone <- function(pb_counts, pb_logcounts, metadata_merged, hormone, overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm) {
  # Split the metadata based on cell type and sex
  split_ids <- split(rownames(metadata_merged), interaction(metadata_merged$cell_type, metadata_merged$Sex, drop = TRUE))
  
  results_by_group <- purrr::map(names(split_ids), function(group_name) {
    ids <- split_ids[[group_name]]
    
    # Subset counts and metadata based on ids
    counts_sub <- t(pb_counts)[, ids, drop = FALSE]
    metadata_sub <- metadata_merged[ids, , drop = FALSE]
    logcounts_sub <- t(pb_logcounts)[, ids, drop = FALSE]
    
    initial_sample_size <- nrow(metadata_sub)
    cat("\nGroup:", group_name, "\n")
    cat("Initial number of samples:", initial_sample_size, "\n")
    
    # Check if the group has a minimum number of samples
    if (nrow(metadata_sub) < overall_min_samples) {
      cat("Group removed: insufficient samples", nrow(metadata_sub), "\n")
      return(NULL)
    }
    
    # Gene-level filtering to retain genes with counts above certain thresholds
    keep_genes <- colSums(counts_sub >= count_threshold & logcounts_sub >= logcount_threshold) >= gene_sample_threshold
    
    counts_sub <- counts_sub[, keep_genes, drop = FALSE]

    # Check if any genes remain after filtering
    if (ncol(counts_sub) == 0) {
      cat("Group removed: no genes passed the filtering threshold\n")
      return(NULL)
    }
    
    # Calculate the outlier threshold based on hormone values
    outlier_concentration <- quantile(metadata_sub[[hormone]], 0.75, na.rm = TRUE) + 1.5 * IQR(metadata_sub[[hormone]], na.rm = TRUE)
    cat("Outlier concentration threshold:", outlier_concentration, "\n")
    
    # Filter out the outliers based on hormone concentration
    non_outliers <- metadata_sub[[hormone]] < outlier_concentration
    counts_sub_non_outliers <- counts_sub[, non_outliers, drop = FALSE]
    metadata_sub_non_outliers <- metadata_sub[non_outliers, , drop = FALSE]
    
    cat("Samples after outlier removal:", nrow(metadata_sub_non_outliers), "\n")
    
    # Check if enough samples remain after removing outliers
    if (nrow(metadata_sub_non_outliers) < overall_min_samples) {
      cat("Group removed: insufficient samples after outlier removal", nrow(metadata_sub_non_outliers), "\n")
      return(NULL)
    }
    
    # Mask for non-zero and non-NA hormone values
    nonzero_hormone_mask <- metadata_sub_non_outliers[[hormone]] != 0 & !is.na(metadata_sub_non_outliers[[hormone]])
    cat("Samples with non-zero hormone:", sum(nonzero_hormone_mask), "\n")
    
    # Check if enough samples have non-zero hormone values
    if (sum(nonzero_hormone_mask) < overall_min_samples) {
      cat("Group removed: insufficient non-zero hormone values", sum(nonzero_hormone_mask), "\n")
      return(NULL)
    }
    
    # Subset counts and metadata to only those with non-zero hormone values
    final_counts_sub <- counts_sub_non_outliers[, nonzero_hormone_mask, drop = FALSE]
    final_metadata_sub <- metadata_sub_non_outliers[nonzero_hormone_mask, , drop = FALSE]
    
    cat("Final sample size:", nrow(final_metadata_sub), "\n")
    
    # Run the modeling function
    result <- model(final_counts_sub, final_metadata_sub, hormone)
    
    # Verify if the result is NULL
    if (is.null(result)) {
      cat("Model returned NULL result\n")
      return(NULL)
    }
    
    # Append additional metadata to the result
    result$cell_type <- final_metadata_sub$cell_type[1]
    result$sex <- final_metadata_sub$Sex[1]
    
    # Return the result including the filtered data used for modeling
    list(
      limma_results = result,
      counts_sub = final_counts_sub,
      metadata_sub = final_metadata_sub
    )
  })
  
  # Remove NULL groups from results_by_group
  results_by_group <- results_by_group[!sapply(results_by_group, is.null)]
  
  # Combine all limma results into one data.frame
  combined_results <- purrr::map_dfr(results_by_group, ~ .x$limma_results)
  
  # Check if combined results are empty
  if (nrow(combined_results) == 0) {
    cat("No significant results found\n")
  }
  
  # Return the combined results and data by group
  list(
    combined_results = combined_results,
    by_group_data = results_by_group
  )
}

analyze_hormone <- function(pb_counts, pb_logcounts, metadata_merged, hormone, overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm) {
  # Split the metadata based on cell type and sex
  split_ids <- split(rownames(metadata_merged), interaction(metadata_merged$cell_type, metadata_merged$Sex, drop = TRUE))
  
  results_by_group <- purrr::map(names(split_ids), function(group_name) {
    ids <- split_ids[[group_name]]
    
    # Subset counts and metadata based on ids
    counts_sub <- t(pb_counts)[, ids, drop = FALSE]
    metadata_sub <- metadata_merged[ids, , drop = FALSE]
    logcounts_sub <- t(pb_logcounts)[, ids, drop = FALSE]
    
    initial_sample_size <- nrow(metadata_sub)
    cat("\nGroup:", group_name, "\n")
    cat("Initial number of samples:", initial_sample_size, "\n")
    
    # Check if the group has a minimum number of samples
    if (nrow(metadata_sub) < overall_min_samples) {
      cat("Group removed: insufficient samples", nrow(metadata_sub), "\n")
      return(NULL)
    }
    
    # Gene-level filtering to retain genes with counts above certain thresholds
    keep_genes <- colSums(counts_sub >= count_threshold & logcounts_sub >= logcount_threshold) >= gene_sample_threshold
    
    counts_sub <- counts_sub[, keep_genes, drop = FALSE]

    # Check if any genes remain after filtering
    if (ncol(counts_sub) == 0) {
      cat("Group removed: no genes passed the filtering threshold\n")
      return(NULL)
    }
    
    # Mask for non-zero and non-NA hormone values
    nonzero_hormone_mask <- metadata_sub[[hormone]] != 0 & !is.na(metadata_sub[[hormone]])
    cat("Samples with non-zero hormone:", sum(nonzero_hormone_mask), "\n")
    
    # Check if enough samples have non-zero hormone values
    if (sum(nonzero_hormone_mask) < overall_min_samples) {
      cat("Group removed: insufficient non-zero hormone values", sum(nonzero_hormone_mask), "\n")
      return(NULL)
    }
    
    # Subset counts and metadata to only those with non-zero hormone values
    final_counts_sub <- counts_sub[, nonzero_hormone_mask, drop = FALSE]
    final_metadata_sub <- metadata_sub[nonzero_hormone_mask, , drop = FALSE]
    
    cat("Final sample size:", nrow(final_metadata_sub), "\n")
    
    # Run the modeling function
    result <- model(final_counts_sub, final_metadata_sub, hormone)
    
    # Verify if the result is NULL
    if (is.null(result)) {
      cat("Model returned NULL result\n")
      return(NULL)
    }
    
    # Append additional metadata to the result
    result$cell_type <- final_metadata_sub$cell_type[1]
    result$sex <- final_metadata_sub$Sex[1]
    
    # Return the result including the filtered data used for modeling
    list(
      limma_results = result,
      counts_sub = final_counts_sub,
      metadata_sub = final_metadata_sub
    )
  })
  
  # Remove NULL groups from results_by_group
  results_by_group <- results_by_group[!sapply(results_by_group, is.null)]
  
  # Combine all limma results into one data.frame
  combined_results <- purrr::map_dfr(results_by_group, ~ .x$limma_results)
  
  # Check if combined results are empty
  if (nrow(combined_results) == 0) {
    cat("No significant results found\n")
  }
  
  # Return the combined results and data by group
  list(
    combined_results = combined_results,
    by_group_data = results_by_group
  )
}
```

### Plotting Function
```{r plotting-function, echo = TRUE, message = FALSE}
## `geom_smooth()` using formula = 'y ~ x'
plot_top_genes_from_subsets <- function(by_group_data, output_dir, hormone, top_n = 6) {
  dir.create(output_dir, showWarnings = FALSE)
  
  message("Number of groups to plot: ", length(by_group_data))
  
  purrr::walk(by_group_data, function(group_list) {
    if (is.null(group_list)) {
      message("- Skipping NULL group")
      return()
    }
    
    limma_res <- group_list$limma_results
    counts_sub <- group_list$counts_sub  # genes x samples
    metadata_sub <- group_list$metadata_sub

    if (nrow(metadata_sub) < 1 || ncol(counts_sub) < 1) {
      message("- Skipping empty subset: ", metadata_sub$cell_type[1], "_", metadata_sub$Sex[1])
      return()
    }

    # Compute log1p(CPM)
    counts_cpm <- edgeR::cpm(counts_sub, log = FALSE)
    counts_log <- log1p(counts_cpm)
    
    # Transpose counts: samples x genes
    counts_log <- t(counts_log)

    top_genes <- limma_res %>%
      arrange(-abs(t)) %>%
      slice_head(n = top_n) %>%
      filter(adj.P.Val < 0.05) %>%
      pull(gene)
    
    top_genes_pvals <- limma_res %>%
        filter(gene %in% top_genes) %>%
        select(gene, adj.P.Val) %>%
        as.data.frame()

    message(" Plotting group: ", metadata_sub$cell_type[1], "_", metadata_sub$Sex[1])
    message("- Top genes: ", paste(top_genes, collapse = ", "))

    plot_list <- purrr::map(seq_len(nrow(top_genes_pvals)), function(i) {
      gene <- top_genes_pvals$gene[i]
      adj_pval <- top_genes_pvals$adj.P.Val[i]
      
      if (!gene %in% colnames(counts_log)) {
        message("    Gene ", gene, " not in counts — skipping")
        return(NULL)
      }

      expr_vals <- counts_log[, gene]
      if (all(is.na(expr_vals)) || all(expr_vals == 0)) {
        message("    Gene ", gene, " has all NA or zero expression — skipping")
        return(NULL)
      }
      
      message("    Plotting gene: ", gene, " (adj. p-value: ", format(adj_pval, digits = 3), ")")
      flush.console()
 
      df <- metadata_sub %>% mutate(expr = expr_vals)

      p <- ggplot(df, aes(x = !!sym(hormone), y = expr, color = cell_count, shape = Sex)) +
        geom_point(size = 2, alpha = 0.8) +
        geom_smooth(method = "lm", se = TRUE) +
        scale_color_viridis_c() +
        scale_shape_manual(values = c("M" = 16, "F" = 17)) +
        labs(title = paste0(gene, " (adj. p-value: ", format(adj_pval, digits = 3), ") in ", metadata_sub$cell_type[1]),
             x = hormone, y = "log1p(CPM)") +
        theme_minimal(base_size = 14)
      stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top", size = 5)
      
      p
    }) %>% compact()

    if (length(plot_list) == 0) {
      message("No valid plots for group")
      return()
    }

    combined_plot <- patchwork::wrap_plots(plotlist = plot_list, ncol = 2)
    clean_name <- paste0(metadata_sub$cell_type[1], "_", metadata_sub$Sex[1])
    clean_name <- gsub("/", "-", clean_name)

    ggsave(file.path(output_dir, paste0(clean_name, "_top_genes.png")),
           plot = combined_plot, width = 10, height = 6 + ceiling(length(plot_list)/2), dpi = 300)
    message("  ✅ Saved plot for ", clean_name)
  })
}  

```

### Summary Functions
```{r summarize}
summarize_sig_genes <- function(results, fdr_cutoff = 0.05, logfc_cutoff = 0.25) {
  sig <- results %>% filter(adj.P.Val < fdr_cutoff)
  if (!is.null(logfc_cutoff)) {
    sig <- sig %>% filter(abs(logFC) >= logfc_cutoff)
  }
  sig %>%
    count(cell_type, sex, name = "num_sig_genes") %>%
    arrange(desc(num_sig_genes))
}

plot_sig_gene_summary <- function(summary_df, title) {
  ggplot(summary_df, aes(x = cell_type, y = num_sig_genes, fill = sex)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    labs(title = title, x = "Cell Type", y = "# Significant Genes") +
    theme_minimal(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("M" = "steelblue", "F" = "violetred3"))
}
```

## Run DE analyses with limma and filtering parameters

### Testosterone
```{r testosterone analysis}
# TT - Run analysis
res_TT_1 <- analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc_gw)

res_TT_2 <- analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc)

res_TT_3 <- analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm)

res_TT_4 <- analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

res_TT_5 <- add_outlier_analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 8, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

# TT - Plot top genes
plot_top_genes_from_subsets(res_TT_1$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group1", hormone = "TT")
plot_top_genes_from_subsets(res_TT_2$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group2", hormone = "TT")
plot_top_genes_from_subsets(res_TT_3$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group3", hormone = "TT")
plot_top_genes_from_subsets(res_TT_4$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group4", hormone = "TT")
plot_top_genes_from_subsets(res_TT_5$by_group_data, "../../results/targeted/pseudobulk_testosterone_plots/Group5", hormone = "TT")

#Summary
group1_summary_TT <- summarize_sig_genes(res_TT_1$combined_results)
group2_summary_TT <- summarize_sig_genes(res_TT_2$combined_results)
group3_summary_TT <- summarize_sig_genes(res_TT_3$combined_results)
group4_summary_TT <- summarize_sig_genes(res_TT_4$combined_results)
group5_summary_TT <- summarize_sig_genes(res_TT_5$combined_results)

#Gene Summary Plot
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group1/sex_separate_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group1_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group2/sex_separate_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group2_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group3/sex_separate_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group3_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group4/sex_separate_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group4_summary_TT, "Testosterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_testosterone_plots/Group5/sex_separate_TT_sig_gene_counts.png", plot = plot_sig_gene_summary(group5_summary_TT, "Testosterone"), width = 8, height = 5)

#Save Result CSV
write_csv(res_TT_1$combined_results, "../../results/targeted/gene_hormone_csv/Group1_sex_separate_TT_results.csv")
write_csv(res_TT_2$combined_results, "../../results/targeted/gene_hormone_csv/Group2_sex_separate_TT_results.csv")
write_csv(res_TT_3$combined_results, "../../results/targeted/gene_hormone_csv/Group3_sex_separate_TT_results.csv")
write_csv(res_TT_4$combined_results, "../../results/targeted/gene_hormone_csv/Group4_sex_separate_TT_results.csv")
write_csv(res_TT_5$combined_results, "../../results/targeted/gene_hormone_csv/Group5_sex_separate_TT_results.csv")

#Save Summary CSV
write_csv(group1_summary_TT, "../../results/targeted/gene_hormone_csv/Group1_sex_separate_TT_summary.csv")
write_csv(group2_summary_TT, "../../results/targeted/gene_hormone_csv/Group2_sex_separate_TT_summary.csv")
write_csv(group3_summary_TT, "../../results/targeted/gene_hormone_csv/Group3_sex_separate_TT_summary.csv")
write_csv(group4_summary_TT, "../../results/targeted/gene_hormone_csv/Group4_sex_separate_TT_summary.csv")
write_csv(group5_summary_TT, "../../results/targeted/gene_hormone_csv/Group5_sex_separate_TT_summary.csv")


```

### Estradiol
```{r estradiol analysis}
# E2 - Run analysis
res_E2_1 <- analyze_hormone(pb_E2$pb$counts, pb_E2$pb$logcounts, pb_E2$metadata, hormone = "E2", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc_gw)

res_E2_2 <- analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc)

res_E2_3 <- analyze_hormone(pb_E2$pb$counts, pb_E2$pb$logcounts, pb_E2$metadata, hormone = "E2", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm)

res_E2_4 <- analyze_hormone(pb_E2$pb$counts, pb_E2$pb$logcounts, pb_E2$metadata, hormone = "E2", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

res_E2_5 <- add_outlier_analyze_hormone(pb_E2$pb$counts, pb_E2$pb$logcounts, pb_E2$metadata, hormone = "E2", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

# E2 - Plot top genes
plot_top_genes_from_subsets(res_E2_1$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group1", hormone = "E2")
plot_top_genes_from_subsets(res_E2_2$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group2", hormone = "E2")
plot_top_genes_from_subsets(res_E2_3$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group3", hormone = "E2")
plot_top_genes_from_subsets(res_E2_4$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group4", hormone = "E2")
plot_top_genes_from_subsets(res_E2_5$by_group_data, "../../results/targeted/pseudobulk_estrogen_plots/Group5", hormone = "E2")

#Summary
group1_summary_E2 <- summarize_sig_genes(res_E2_1$combined_results)
group2_summary_E2 <- summarize_sig_genes(res_E2_2$combined_results)
group3_summary_E2 <- summarize_sig_genes(res_E2_3$combined_results)
group4_summary_E2 <- summarize_sig_genes(res_E2_4$combined_results)
group5_summary_E2 <- summarize_sig_genes(res_E2_5$combined_results)

#Gene Summary Plot
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group1/sex_separate_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group1_summary_E2, "Estradiol"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group2/sex_separate_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group2_summary_E2, "Estradiol"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group3/sex_separate_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group3_summary_E2, "Estradiol"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group4/sex_separate_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group4_summary_E2, "Estradiol"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_estrogen_plots/Group5/sex_separate_E2_sig_gene_counts.png", plot = plot_sig_gene_summary(group5_summary_E2, "Estradiol"), width = 8, height = 5)

#Save Result CSV
write_csv(res_E2_1$combined_results, "../../results/targeted/gene_hormone_csv/Group1_sex_separate_E2_results.csv")
write_csv(res_E2_2$combined_results, "../../results/targeted/gene_hormone_csv/Group2_sex_separate_E2_results.csv")
write_csv(res_E2_3$combined_results, "../../results/targeted/gene_hormone_csv/Group3_sex_separate_E2_results.csv")
write_csv(res_E2_4$combined_results, "../../results/targeted/gene_hormone_csv/Group4_sex_separate_E2_results.csv")
write_csv(res_E2_5$combined_results, "../../results/targeted/gene_hormone_csv/Group5_sex_separate_E2_results.csv")

#Save Summary CSV
write_csv(group1_summary_E2, "../../results/targeted/gene_hormone_csv/Group1_sex_separate_E2_summary.csv")
write_csv(group2_summary_E2, "../../results/targeted/gene_hormone_csv/Group2_sex_separate_E2_summary.csv")
write_csv(group3_summary_E2, "../../results/targeted/gene_hormone_csv/Group3_sex_separate_E2_summary.csv")
write_csv(group4_summary_E2, "../../results/targeted/gene_hormone_csv/Group4_sex_separate_E2_summary.csv")
write_csv(group5_summary_E2, "../../results/targeted/gene_hormone_csv/Group5_sex_separate_E2_summary.csv")

```

### Progesterone
```{r progesterone analysis}
# P4 - Run analysis
res_P4_1 <- analyze_hormone(pb_P4$pb$counts, pb_P4$pb$logcounts, pb_P4$metadata, hormone = "P4", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc_gw)

res_P4_2 <- analyze_hormone(pb_TT$pb$counts, pb_TT$pb$logcounts, pb_TT$metadata, hormone = "TT", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm_cc)

res_P4_3 <- analyze_hormone(pb_P4$pb$counts, pb_P4$pb$logcounts, pb_P4$metadata, hormone = "P4", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0, model = run_voom_lm)

res_P4_4 <- analyze_hormone(pb_P4$pb$counts, pb_P4$pb$logcounts, pb_P4$metadata, hormone = "P4", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

res_P4_5 <- add_outlier_analyze_hormone(pb_P4$pb$counts, pb_P4$pb$logcounts, pb_P4$metadata, hormone = "P4", overall_min_samples = 10, count_threshold = 2, gene_sample_threshold = 3, logcount_threshold = 0.5, model = run_voom_lm)

# P4 - Plot top genes
plot_top_genes_from_subsets(res_P4_1$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group1", hormone = "P4")
plot_top_genes_from_subsets(res_P4_2$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group2", hormone = "P4")
plot_top_genes_from_subsets(res_P4_3$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group3", hormone = "P4")
plot_top_genes_from_subsets(res_P4_4$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group4", hormone = "P4")
plot_top_genes_from_subsets(res_P4_5$by_group_data, "../../results/targeted/pseudobulk_progesterone_plots/Group5", hormone = "P4")

#Summary
group1_summary_P4 <- summarize_sig_genes(res_P4_1$combined_results)
group2_summary_P4 <- summarize_sig_genes(res_P4_2$combined_results)
group3_summary_P4 <- summarize_sig_genes(res_P4_3$combined_results)
group4_summary_P4 <- summarize_sig_genes(res_P4_4$combined_results)
group5_summary_P4 <- summarize_sig_genes(res_P4_5$combined_results)

#Gene Summary Plot
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group1/sex_separate_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group1_summary_P4, "Progesterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group2/sex_separate_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group2_summary_P4, "Progesterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group3/sex_separate_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group3_summary_P4, "Progesterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group4/sex_separate_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group4_summary_P4, "Progesterone"), width = 8, height = 5)
ggsave("../../results/targeted/pseudobulk_progesterone_plots/Group5/sex_separate_P4_sig_gene_counts.png", plot = plot_sig_gene_summary(group5_summary_P4, "Progesterone"), width = 8, height = 5)

#Save Result CSV
write_csv(res_P4_1$combined_results, "../../results/targeted/gene_hormone_csv/Group1_sex_separate_P4_results.csv")
write_csv(res_P4_2$combined_results, "../../results/targeted/gene_hormone_csv/Group2_sex_separate_P4_results.csv")
write_csv(res_P4_3$combined_results, "../../results/targeted/gene_hormone_csv/Group3_sex_separate_P4_results.csv")
write_csv(res_P4_4$combined_results, "../../results/targeted/gene_hormone_csv/Group4_sex_separate_P4_results.csv")
write_csv(res_P4_5$combined_results, "../../results/targeted/gene_hormone_csv/Group5_sex_separate_P4_results.csv")

#Save Summary CSV
write_csv(group1_summary_P4, "../../results/targeted/gene_hormone_csv/Group1_sex_separate_P4_summary.csv")
write_csv(group2_summary_P4, "../../results/targeted/gene_hormone_csv/Group2_sex_separate_P4_summary.csv")
write_csv(group3_summary_P4, "../../results/targeted/gene_hormone_csv/Group3_sex_separate_P4_summary.csv")
write_csv(group4_summary_P4, "../../results/targeted/gene_hormone_csv/Group4_sex_separate_P4_summary.csv")
write_csv(group5_summary_P4, "../../results/targeted/gene_hormone_csv/Group5_sex_separate_P4_summary.csv")
```


# End