---
title: "Targeted Hormone Sex Comparison"
author: "Sneha Chaturvedi"
date: "Updated 11.6.25"
output: 
  html_document:
    toc: TRUE
---

The purpose of this analysis is to 1) to find outliers in the distribution of hormone concentrations and 2) compare concentrations between males and females for significant differences. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Packages <- c("tidyverse", "readxl", "stats", "ggpubr", "svglite", "knitr")

lapply(Packages, library, character.only=TRUE)
svg_path <- "../../results/targeted/sex_comparison_plots"
```

# Set Up
```{r LoadDataset, echo=TRUE}
#Load Dataset
targeted = read_excel('../../doc/targeted/targeted_hormones.xlsx',sheet="DATA",range=NULL)
targeted <- targeted %>%
  rename(`TT (pg/mg)`     = `TT (ng/mL)`,
         `17B-E2 (pg/mg)` = `17B-E2 (ng/mL)`,
         `P4 (pg/mg)`     = `P4 (ng/mL)`)
#Create separate datasets by sex
female_targeted <- filter(targeted, Sex == 'F')
male_targeted <- filter(targeted, Sex == 'M')

#convert sex column to factor
targeted$Sex <- as.factor(targeted$Sex)
targeted$`TT (pg/mg)`<- as.numeric(targeted$`TT (pg/mg)`)
targeted$`P4 (pg/mg)`<- as.numeric(targeted$`P4 (pg/mg)`)
targeted$`17B-E2 (pg/mg)`<- as.numeric(targeted$`17B-E2 (pg/mg)`)

#Load Functions
detect_outliers <- function(data, variables) {
  for (variable in variables) {
  #Calculate the outlier limit
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) + 1.5 * IQR(data[[variable]], na.rm = TRUE)
  #Find outliers
  outliers <- data %>% filter(.data[[variable]] > outlier_limit)
  #Print out outliers
  print(paste("Outliers for:", variable))
  print(outliers)
  #Create dataframe with outliers removed
  filtered_data <- data %>% filter(.data[[variable]] <= outlier_limit)
  }
  return(filtered_data)
}
```
# Hormone QC
## Raw Data Distribution
```{r visualize_raw, echo=TRUE}
#convert sex column to factor and ensure hormone values are numeric
targeted$Sex <- as.factor(targeted$Sex)
targeted$`TT (pg/mg)`<- as.numeric(targeted$`TT (pg/mg)`)
targeted$`P4 (pg/mg)`<- as.numeric(targeted$`P4 (pg/mg)`)
targeted$`17B-E2 (pg/mg)`<- as.numeric(targeted$`17B-E2 (pg/mg)`)

# Plot the distribution of TT (pg/mg) for the entire dataset
ggplot(targeted, aes(x = `TT (pg/mg)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of TT (pg/mg) - Total') +
  xlab('TT (pg/mg)') +
  ylab('Count')
ggsave("tt_distribution.svg", path = svg_path)

# QQ plot for total distribution
ggqqplot(targeted$`TT (pg/mg)`) +
  labs(title = "QQ Plot of TT (pg/mg) - Total")
ggsave("tt_QQ.svg", path = svg_path)

# Plot the distribution of TT (pg/mg) by Sex
ggplot(targeted, aes(x = `TT (pg/mg)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of TT (pg/mg) by Sex') +
  xlab('TT (pg/mg)') +
  ylab('Count')
ggsave("tt_distribution_sex.svg", path = svg_path)


# Plot the distribution of 17B-E2 (pg/mg) for the entire dataset
ggplot(targeted, aes(x = `17B-E2 (pg/mg)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of 17B-E2 (pg/mg) - Total') +
  xlab('17B-E2 (pg/mg)') +
  ylab('Count')
ggsave("e2_distribution.svg", path = svg_path)

# Generate QQ plots
# QQ plot for total distribution
ggqqplot(targeted$`17B-E2 (pg/mg)`) +
  labs(title = "QQ Plot of 17B-E2 (pg/mg) - Total")
ggsave("e2_QQ.svg", path = svg_path)

# Check the distribution of 17B-E2 (pg/mg) by Sex
ggplot(targeted, aes(x = `17B-E2 (pg/mg)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of 17B-E2 (pg/mg) by Sex') +
  xlab('17B-E2 (pg/mg)') +
  ylab('Count')
ggsave("e2_distribution_sex.svg", path = svg_path)


# Plot the distribution of P4 (pg/mg) for the entire dataset
ggplot(targeted, aes(x = `P4 (pg/mg)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of P4 (pg/mg) - Total') +
  xlab('P4 (pg/mg)') +
  ylab('Count')
ggsave("p4_distribution.svg", path = svg_path)

# Generate QQ plots
# QQ plot for total distribution
ggqqplot(targeted$`P4 (pg/mg)`) +
  labs(title = "QQ Plot of P4 (pg/mg) - Total")
ggsave("p4_QQ.svg", path = svg_path)

# Check the distribution of P4 (pg/mg) by Sex
ggplot(targeted, aes(x = `P4 (pg/mg)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of P4 (pg/mg) by Sex') +
  xlab('P4 (pg/mg)') +
  ylab('Count')
ggsave("p4_distribution_sex.svg", path = svg_path)
```

## Tests of Normality
```{r Normality, echo=TRUE}
# Perform Shapiro-Wilk normality test for each sex group
  all_shapiro_tt <- shapiro.test(targeted$`TT (pg/mg)`)
  m_shapiro_tt <- shapiro.test(male_targeted$`TT (pg/mg)`)
  f_shapiro_tt <- shapiro.test(female_targeted$`TT (pg/mg)`)
  
all_shapiro_e2 <- shapiro.test(targeted$`17B-E2 (pg/mg)`)
  m_shapiro_e2 <- shapiro.test(male_targeted$`17B-E2 (pg/mg)`)
  f_shapiro_e2 <- shapiro.test(female_targeted$`17B-E2 (pg/mg)`)

all_shapiro_p4 <- shapiro.test(targeted$`P4 (pg/mg)`)
  m_shapiro_p4 <- shapiro.test(male_targeted$`P4 (pg/mg)`)
  f_shapiro_p4 <- shapiro.test(female_targeted$`P4 (pg/mg)`)
  

# Create a summary data frame with the results
shapiro_results <- data.frame(
  Group = c('All', 'Male', 'Female'),
  TT_p_value = formatC(c(all_shapiro_tt$p.value, m_shapiro_tt$p.value, f_shapiro_tt$p.value), format = "e", digits = 2),
  E2_p_value = formatC(c(all_shapiro_e2$p.value, m_shapiro_e2$p.value, f_shapiro_e2$p.value), format = "e", digits = 2),
  P4_p_value = formatC(c(all_shapiro_p4$p.value, m_shapiro_p4$p.value, f_shapiro_p4$p.value), format = "e", digits = 2)
)

# Print the results in a table
kable(shapiro_results, caption = "Shapiro-Wilk Normality Test Results (Raw)")

```
# Data Filtering and Clean Up

## Outlier Detection + Removal
```{r outlier, echo=TRUE}
# Create a new dataset with outliers removed
hormones <- c("TT (pg/mg)", "17B-E2 (pg/mg)", "P4 (pg/mg)")
clean_targeted <- detect_outliers(targeted, hormones)
```

```{r re-visualize, echo=TRUE}
# Plot the distribution of TT (pg/mg) for the entire dataset
ggplot(clean_targeted, aes(x = `TT (pg/mg)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of TT (pg/mg) - Total') +
  xlab('TT (pg/mg)') +
  ylab('Count')
ggsave("clean_tt_distribution.svg", path = svg_path)

# QQ plot for total distribution
ggqqplot(clean_targeted$`TT (pg/mg)`) +
  labs(title = "QQ Plot of TT (pg/mg) - Total")
ggsave("clean_tt_QQ.svg", path = svg_path)

# Check the distribution of TT (pg/mg) by Sex
ggplot(clean_targeted, aes(x = `TT (pg/mg)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of TT (pg/mg) by Sex') +
  xlab('TT (pg/mg)') +
  ylab('Count')
ggsave("clean_tt_distribution_sex.svg", path = svg_path)


# Plot the distribution of 17B-E2 (pg/mg) for the entire dataset
ggplot(clean_targeted, aes(x = `17B-E2 (pg/mg)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of 17B-E2 (pg/mg) - Total') +
  xlab('17B-E2 (pg/mg)') +
  ylab('Count')
ggsave("clean_e2_distribution.svg", path = svg_path)

# QQ plot for total distribution
ggqqplot(clean_targeted$`17B-E2 (pg/mg)`) +
  labs(title = "QQ Plot of 17B-E2 (pg/mg) - Total")
ggsave("clean_e2_QQ.svg", path = svg_path)

# Check the distribution of 17B-E2 (pg/mg) Sex
ggplot(clean_targeted, aes(x = `17B-E2 (pg/mg)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of 17B-E2 (pg/mg) by Sex') +
  xlab('17B-E2 (pg/mg)') +
  ylab('Count')
ggsave("clean_e2_distribution_sex.svg", path = svg_path)

# Plot the distribution of P4 (pg/mg) for the entire dataset
ggplot(clean_targeted, aes(x = `P4 (pg/mg)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of P4 (pg/mg) - Total') +
  xlab('P4 (pg/mg)') +
  ylab('Count')
ggsave("clean_p4_distribution.svg", path = svg_path)

# QQ plot for total distribution
ggqqplot(clean_targeted$`P4 (pg/mg)`) +
  labs(title = "QQ Plot of P4 (pg/mg) - Total")
ggsave("clean_p4_QQ.svg", path = svg_path)

# Check the distribution of P4 (pg/mg) by Sex
ggplot(clean_targeted, aes(x = `P4 (pg/mg)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of P4 (pg/mg) by Sex') +
  xlab('P4 (pg/mg)') +
  ylab('Count')
ggsave("clean_p4_distribution_sex.svg", path = svg_path)
```

```{r re-normality, echo = TRUE}
#Create separate datasets by sex
female_clean_targeted <- filter(clean_targeted, Sex == 'F')
male_clean_targeted <- filter(clean_targeted, Sex == 'M')

# Perform Shapiro-Wilk normality test for each sex group
  clean_all_shapiro_tt <- shapiro.test(clean_targeted$`TT (pg/mg)`)
  clean_m_shapiro_tt <- shapiro.test(male_clean_targeted$`TT (pg/mg)`)
  clean_f_shapiro_tt <- shapiro.test(female_clean_targeted$`TT (pg/mg)`)

clean_all_shapiro_e2 <- shapiro.test(clean_targeted$`17B-E2 (pg/mg)`)
clean_m_shapiro_e2 <- shapiro.test(male_clean_targeted$`17B-E2 (pg/mg)`)
clean_f_shapiro_e2 <- shapiro.test(female_clean_targeted$`17B-E2 (pg/mg)`)

clean_all_shapiro_p4 <- shapiro.test(clean_targeted$`P4 (pg/mg)`)
clean_m_shapiro_p4 <- shapiro.test(male_clean_targeted$`P4 (pg/mg)`)
clean_f_shapiro_p4 <- shapiro.test(female_clean_targeted$`P4 (pg/mg)`)
  

# Create a summary data frame with the results
clean_shapiro_results <- data.frame(
  Group = c('All', 'Male', 'Female'),
  TT_p_value = formatC(c(clean_all_shapiro_tt$p.value, clean_m_shapiro_tt$p.value, clean_f_shapiro_tt$p.value), format = "e", digits = 2),
  E2_p_value = formatC(c(clean_all_shapiro_e2$p.value, clean_m_shapiro_e2$p.value, clean_f_shapiro_e2$p.value), format = "e", digits = 2),
  P4_p_value = formatC(c(clean_all_shapiro_p4$p.value, clean_m_shapiro_p4$p.value, clean_f_shapiro_p4$p.value), format = "e", digits = 2)
)

# Print the results in a table
kable(clean_shapiro_results, caption = "Shapiro-Wilk Normality Test Results (Clean)")


```

# Testosterone 
```{r TT_SexComparison}

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (all_shapiro_tt$p.value > 0.05) {
  # Perform t-test
  tt_t_test <- t.test(`TT (pg/mg)` ~ Sex, data = targeted)
  tt_test_type <- "t-test"
  tt_test_statistic <- tt_t_test$statistic
  tt_p_value <- tt_t_test$p.value
  print(tt_t_test)
} else {
  # Perform Wilcoxon rank-sum test
  tt_wilcox_test <- wilcox.test(`TT (pg/mg)` ~ Sex, data = targeted)
  tt_test_type <- "Wilcoxon rank-sum test"
  tt_test_statistic <- tt_wilcox_test$statistic
  tt_p_value <- tt_wilcox_test$p.value
  print(tt_wilcox_test)
}

# Visualize results with boxplot
ggboxplot(targeted, x = "Sex", y = "`TT (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "TT (pg/mg)", xlab = "Sex",
          add = "jitter") +
  labs(title = paste(tt_test_type, ": statistic =", round(tt_test_statistic, 2), ": pvalue =", formatC(tt_p_value, format = "e", digits = 2)))
ggsave("tt_statistics_sex.svg", path = svg_path)

# Visualize results with boxplot + log y scale
ggboxplot(targeted, x = "Sex", y = "`TT (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "Log 10 TT (pg/mg)", xlab = "Sex",
          add = "jitter") +
  scale_y_log10() +
  labs(title = paste(tt_test_type, ": statistic =", round(tt_test_statistic, 2), ": pvalue =", formatC(tt_p_value, format = "e", digits = 2)))
ggsave("log_tt_statistics_sex.svg", path = svg_path)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (clean_all_shapiro_tt$p.value > 0.05) {
  # Perform t-test
  clean_tt_t_test <- t.test(`TT (pg/mg)` ~ Sex, data = clean_targeted)
  clean_tt_test_type <- "t-test"
  clean_tt_test_statistic <- clean_tt_t_test$statistic
  clean_tt_p_value <- clean_tt_t_test$p.value
  print(clean_tt_t_test)
} else {
  # Perform Wilcoxon rank-sum test
  clean_tt_wilcox_test <- wilcox.test(`TT (pg/mg)` ~ Sex, data = clean_targeted)
  clean_tt_test_type <- "Wilcoxon rank-sum test"
  clean_tt_test_statistic <- clean_tt_wilcox_test$statistic
  clean_tt_p_value <- clean_tt_wilcox_test$p.value
  print(clean_tt_wilcox_test)
}

# Visualize results with boxplot
ggboxplot(clean_targeted, x = "Sex", y = "`TT (pg/mg)`",
          color = "Sex", width = 0.5, 
          ylab = "TT (pg/mg)", xlab = "Sex",
          add = "jitter") + 
  labs(title = paste(clean_tt_test_type, ": statistic =", round(clean_tt_test_statistic, 2), ": pvalue =", formatC(clean_tt_p_value, format = "e", digits = 2)))

# Visualize results with boxplot
ggboxplot(clean_targeted, x = "Sex", y = "`TT (pg/mg)`",
          color = "Sex", width = 0.5, 
          ylab = "Log 10 TT (pg/mg)", xlab = "Sex",
          add = "jitter") + 
  scale_y_log10() +
  labs(title = paste(clean_tt_test_type, ": statistic =", round(clean_tt_test_statistic, 2), ": pvalue =", formatC(clean_tt_p_value, format = "e", digits = 2)))

ggsave("log_clean_tt_statistics_sex.svg", path = svg_path)

```

# Estrogen
```{r E2_SexComparison, echo=TRUE}
# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (all_shapiro_e2$p.value > 0.05) {
  # Perform t-test
  e2_t_test <- t.test(`17B-E2 (pg/mg)` ~ Sex, data = targeted)
  e2_test_type <- "t-test"
  e2_test_statistic <- e2_t_test$statistic
  e2_p_value <- e2_t_test$p.value
  print(e2_t_test)
} else {
  # Perform Wilcoxon rank-sum test
  e2_wilcox_test <- wilcox.test(`17B-E2 (pg/mg)` ~ Sex, data = targeted)
  e2_test_type <- "Wilcoxon rank-sum test"
  e2_test_statistic <- e2_wilcox_test$statistic
  e2_p_value <- e2_wilcox_test$p.value
  print(e2_wilcox_test)
}

# Visualize results with boxplot
ggboxplot(targeted, x = "Sex", y = "`17B-E2 (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "17B-E2 (pg/mg)", xlab = "Sex",
          add = "jitter") +
  labs(title = paste(e2_test_type, ": statistic =", round(e2_test_statistic, 2), ": pvalue =", formatC(e2_p_value, format = "e", digits = 2)))
ggsave("e2_statistics_sex.svg", path = svg_path)

ggboxplot(targeted, x = "Sex", y = "`17B-E2 (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "Log 10 17B-E2 (pg/mg)", xlab = "Sex",
          add = "jitter") +
  scale_y_log10() +
  labs(title = paste(e2_test_type, ": statistic =", round(e2_test_statistic, 2), ": pvalue =", formatC(e2_p_value, format = "e", digits = 2)))
ggsave("log_e2_statistics_sex.svg", path = svg_path)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (clean_all_shapiro_e2$p.value > 0.05) {
  # Perform t-test
  clean_e2_t_test <- t.test(`17B-E2 (pg/mg)` ~ Sex, data = clean_targeted)
  clean_e2_test_type <- "t-test"
  clean_e2_test_statistic <- clean_e2_t_test$statistic
  clean_e2_p_value <- clean_e2_t_test$p.value
  print(clean_e2_t_test)
} else {
  # Perform Wilcoxon rank-sum test
  clean_e2_wilcox_test <- wilcox.test(`17B-E2 (pg/mg)` ~ Sex, data = clean_targeted)
  clean_e2_test_type <- "Wilcoxon rank-sum test"
  clean_e2_test_statistic <- clean_e2_wilcox_test$statistic
  clean_e2_p_value <- clean_e2_wilcox_test$p.value
  print(clean_e2_wilcox_test)
}

# Visualize results with boxplot
ggboxplot(clean_targeted, x = "Sex", y = "`17B-E2 (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "17B-E2 (pg/mg)", xlab = "Sex",
          add = "jitter") +
  labs(title = paste(clean_e2_test_type, ": statistic =", round(clean_e2_test_statistic, 2), ": pvalue =", formatC(clean_e2_p_value, format = "e", digits = 2)))
ggsave("clean_e2_statistics_sex.svg", path = svg_path)


ggboxplot(clean_targeted, x = "Sex", y = "`17B-E2 (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "Log 10 17B-E2 (pg/mg)", xlab = "Sex",
          add = "jitter") +
  scale_y_log10() +
  labs(title = paste(clean_e2_test_type, ": statistic =", round(clean_e2_test_statistic, 2), ": pvalue =", formatC(clean_e2_p_value, format = "e", digits = 2)))
ggsave("log_clean_e2_statistics_sex.svg", path = svg_path)
```

# Progesterone

```{r Progesterone, echo=TRUE}
# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (all_shapiro_p4$p.value > 0.05) {
  # Perform t-test
  p4_t_test <- t.test(`P4 (pg/mg)` ~ Sex, data = targeted)
  p4_test_type <- "t-test"
  p4_test_statistic <- p4_t_test$statistic
  p4_p_value <- p4_t_test$p.value
  print(p4_t_test)
} else {
  # Perform Wilcoxon rank-sum test
  p4_wilcox_test <- wilcox.test(`P4 (pg/mg)` ~ Sex, data = targeted)
  p4_test_type <- "Wilcoxon rank-sum test"
  p4_test_statistic <- p4_wilcox_test$statistic
  p4_p_value <- p4_wilcox_test$p.value
  print(p4_wilcox_test)
}

# Visualize results with boxplot
ggboxplot(targeted, x = "Sex", y = "`P4 (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "P4 (pg/mg)", xlab = "Sex",
          add = "jitter") +
  labs(title = paste(p4_test_type, ": statistic =", round(p4_test_statistic, 2), ": pvalue =", formatC(p4_p_value, format = "e", digits = 2)))
ggsave("p4_statistics_sex.svg", path = svg_path)

ggboxplot(targeted, x = "Sex", y = "`P4 (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "Log 10 P4 (pg/mg)", xlab = "Sex",
          add = "jitter") +
  scale_y_log10() +
  labs(title = paste(p4_test_type, ": statistic =", round(p4_test_statistic, 2), ": pvalue =", formatC(p4_p_value, format = "e", digits = 2)))
ggsave("log_p4_statistics_sex.svg", path = svg_path)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (clean_all_shapiro_p4$p.value > 0.05) {
  # Perform t-test
  clean_p4_t_test <- t.test(`P4 (pg/mg)` ~ Sex, data = clean_targeted)
  clean_p4_test_type <- "t-test"
  clean_p4_test_statistic <- clean_p4_t_test$statistic
  clean_p4_p_value <- clean_p4_t_test$p.value
  print(clean_p4_t_test)
} else {
  # Perform Wilcoxon rank-sum test
  clean_p4_wilcox_test <- wilcox.test(`P4 (pg/mg)` ~ Sex, data = clean_targeted)
  clean_p4_test_type <- "Wilcoxon rank-sum test"
  clean_p4_test_statistic <- clean_p4_wilcox_test$statistic
  clean_p4_p_value <- clean_p4_wilcox_test$p.value
  print(clean_p4_wilcox_test)
}

# Visualize results with boxplot
ggboxplot(clean_targeted, x = "Sex", y = "`P4 (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "P4 (pg/mg)", xlab = "Sex",
          add = "jitter") +
  labs(title = paste(clean_p4_test_type, ": statistic =", round(clean_p4_test_statistic, 2), ": pvalue =", formatC(clean_p4_p_value, format = "e", digits = 2)))
ggsave("clean_p4_statistics_sex.svg", path = svg_path)

ggboxplot(clean_targeted, x = "Sex", y = "`P4 (pg/mg)`",
          color = "Sex", width = 0.5,
          ylab = "Log 10 P4 (pg/mg)", xlab = "Sex",
          add = "jitter") +
  scale_y_log10() +
  labs(title = paste(clean_p4_test_type, ": statistic =", round(clean_p4_test_statistic, 2), ": pvalue =", formatC(clean_p4_p_value, format = "e", digits = 2)))
ggsave("log_clean_p4_statistics_sex.svg", path = svg_path)

```
# Summary Table
``` {r summary_table}
# Enhanced statistics table with descriptive stats (outlier-removed only)
stats_summary_enhanced <- data.frame(
  Hormone = c("`TT (pg/mg)`", "`17B-E2 (pg/mg)`", "`P4 (pg/mg)`"),
  N_Female = c(
    sum(clean_targeted$Sex == "F"),
    sum(clean_targeted$Sex == "F"),
    sum(clean_targeted$Sex == "F")
  ),
  N_Male = c(
    sum(clean_targeted$Sex == "M"),
    sum(clean_targeted$Sex == "M"),
    sum(clean_targeted$Sex == "M")
  ),
  Mean_Female = c(
    mean(clean_targeted$`TT (pg/mg)`[clean_targeted$Sex == "F"], na.rm = TRUE),
    mean(clean_targeted$`17B-E2 (pg/mg)`[clean_targeted$Sex == "F"], na.rm = TRUE),
    mean(clean_targeted$`P4 (pg/mg)`[clean_targeted$Sex == "F"], na.rm = TRUE)
  ),
  Mean_Male = c(
    mean(clean_targeted$`TT (pg/mg)`[clean_targeted$Sex == "M"], na.rm = TRUE),
    mean(clean_targeted$`17B-E2 (pg/mg)`[clean_targeted$Sex == "M"], na.rm = TRUE),
    mean(clean_targeted$`P4 (pg/mg)`[clean_targeted$Sex == "M"], na.rm = TRUE)
  ),
  Test_Type = c(
    clean_tt_test_type,
    clean_e2_test_type,
    clean_p4_test_type
  ),
  Test_Statistic = c(
    clean_tt_test_statistic,
    clean_e2_test_statistic,
    clean_p4_test_statistic
  ),
  P_Value = c(
    clean_tt_p_value,
    clean_e2_p_value,
    clean_p4_p_value
  ),
  Significant = c(
    clean_tt_p_value < 0.05,
    clean_e2_p_value < 0.05,
    clean_p4_p_value < 0.05
  )
)

# Round numeric columns
stats_summary_enhanced$Mean_Female <- round(stats_summary_enhanced$Mean_Female, 4)
stats_summary_enhanced$Mean_Male <- round(stats_summary_enhanced$Mean_Male, 4)
stats_summary_enhanced$Test_Statistic <- round(stats_summary_enhanced$Test_Statistic, 3)
stats_summary_enhanced$P_Value <- formatC(stats_summary_enhanced$P_Value, format = "e", digits = 3)

# Print and save
print(stats_summary_enhanced)
write.csv(stats_summary_enhanced, "../../results/targeted/hormone_csv/hormone_sex_comparison_stats_detailed.csv", row.names = FALSE)
```

# Literature Comparison

Compares the targeted hormone concentrations measured in fetal cortical tissue
(pg/mg) against values reported in the literature. Note that with one exception
these published values are **culture media exposure concentrations**, not tissue
measurements: only the Kelava et al. ELISA readout is a measured concentration.
Exposure and measured values are distinguished by point shape below.

```{r lit_setup, echo=TRUE}
library(patchwork)

# Literature values recorded as originally reported (pg/mL), then converted to
# tissue-equivalent pg/mg using the same fetal brain density constant as the Kd
# conversion, so all three scales in the figure are directly comparable.
# type: "Exposure" = concentration applied to culture; "Measured" = assayed readout.
BRAIN_DENSITY <- 1120  # mg/mL; see doi:10.1002/uog.70140

lit_hormones <- tribble(
  ~study,     ~hormone, ~pg_per_mL, ~type,
  "Supakul",  "E2",      27240,     "Exposure",

  "Smeeth",   "E2",         10,     "Exposure",
  "Smeeth",   "E2",        100,     "Exposure",
  "Smeeth",   "E2",       1000,     "Exposure",
  "Smeeth",   "E2",       5000,     "Exposure",
  "Smeeth",   "E2",      25000,     "Exposure",
  "Smeeth",   "E2",      75000,     "Exposure",

  "Smeeth",   "TT",         10,     "Exposure",
  "Smeeth",   "TT",        100,     "Exposure",
  "Smeeth",   "TT",       1000,     "Exposure",
  "Smeeth",   "TT",       5000,     "Exposure",
  "Smeeth",   "TT",      10000,     "Exposure",

  "Kelava",   "E2",      27240,     "Exposure",
  "Kelava",   "E2",       4358.4,   "Measured",
  "Kelava",   "TT",      28840,     "Exposure",
  "Kelava",   "TT",       4614.4,   "Measured",
  "Kelava",   "P4",      31450,     "Exposure",
  "Kelava",   "P4",       5032.0,   "Measured"
) %>%
  mutate(
    value   = pg_per_mL / BRAIN_DENSITY,
    hormone = factor(hormone, levels = c("TT", "E2", "P4")),
    type    = factor(type, levels = c("Exposure", "Measured"))
  )

# Our measured values, long format, outlier-removed set
our_hormones <- clean_targeted %>%
  select(Sex, `TT (pg/mg)`, `17B-E2 (pg/mg)`, `P4 (pg/mg)`) %>%
  pivot_longer(-Sex, names_to = "col", values_to = "value") %>%
  mutate(
    hormone = case_when(
      str_detect(col, "TT")  ~ "TT",
      str_detect(col, "E2")  ~ "E2",
      str_detect(col, "P4")  ~ "P4"
    ),
    hormone = factor(hormone, levels = c("TT", "E2", "P4")),
    study   = "Our data",
    type    = "Measured"
  ) %>%
  filter(!is.na(value))

# Log10 panels drop non-positive values; report how many
n_dropped <- our_hormones %>%
  group_by(hormone) %>%
  summarise(n_total = n(), n_nonpositive = sum(value <= 0), .groups = "drop")
kable(n_dropped, caption = "Values excluded from log10 panels (<= 0)")

our_hormones_log <- filter(our_hormones, value > 0)

hormone_labels <- c(TT = "TT (pg/mg)", E2 = "17B-E2 (pg/mg)", P4 = "P4 (pg/mg)")

study_shapes <- c("Supakul" = 21, "Smeeth" = 22, "Kelava" = 24)
type_fills   <- c("Exposure" = "white", "Measured" = "black")
```

## Standalone literature panels (right-hand axis)

These are sized to sit flush beside the existing sex-comparison plots. The axis
is on the right so the two panels read as a single figure with independent scales.

```{r lit_standalone, echo=TRUE, fig.width=3, fig.height=4}
plot_lit_panel <- function(h) {
  d <- filter(lit_hormones, hormone == h)
  if (nrow(d) == 0) return(NULL)

  ggplot(d, aes(x = study, y = value)) +
    geom_point(aes(shape = study, fill = type),
               size = 3, stroke = 0.6, colour = "black",
               position = position_jitter(width = 0.08, height = 0)) +
    scale_shape_manual(values = study_shapes, guide = "none") +
    scale_fill_manual(values = type_fills, name = NULL,
                      guide = guide_legend(override.aes = list(shape = 21))) +
    scale_y_log10(position = "right") +
    labs(x = NULL, y = hormone_labels[[h]]) +
    theme_pubr(legend = "bottom") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

for (h in c("TT", "E2", "P4")) {
  p <- plot_lit_panel(h)
  if (is.null(p)) next
  print(p)
  ggsave(paste0("lit_", tolower(h), "_standalone.svg"),
         plot = p, path = svg_path, width = 3, height = 4)
  ggsave(paste0("lit_", tolower(h), "_standalone.png"),
         plot = p, path = svg_path, width = 3, height = 4, dpi = 300)
}
```

## Combined panels: our distribution with literature overlaid

Preferred for the manuscript — a single log10 axis removes the need for the
reader to align two independent scales.

```{r lit_combined, echo=TRUE, fig.width=4.5, fig.height=4.5}
plot_lit_combined <- function(h) {
  d_ours <- filter(our_hormones_log, hormone == h)
  d_lit  <- filter(lit_hormones, hormone == h)

  ggplot() +
    geom_boxplot(data = d_ours, aes(x = "Our data", y = value),
                 width = 0.5, outlier.shape = NA, colour = "grey30") +
    geom_jitter(data = d_ours, aes(x = "Our data", y = value),
                width = 0.12, height = 0, alpha = 0.5, size = 1.4,
                colour = "grey30") +
    geom_point(data = d_lit, aes(x = study, y = value,
                                 shape = study, fill = type),
               size = 3, stroke = 0.6, colour = "black",
               position = position_jitter(width = 0.08, height = 0)) +
    scale_x_discrete(limits = c("Our data",
                                intersect(c("Supakul", "Smeeth", "Kelava"),
                                          unique(d_lit$study)))) +
    scale_shape_manual(values = study_shapes, guide = "none") +
    scale_fill_manual(values = type_fills, name = NULL,
                      guide = guide_legend(override.aes = list(shape = 21))) +
    scale_y_log10() +
    labs(x = NULL, y = hormone_labels[[h]]) +
    theme_pubr(legend = "bottom") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

combined_plots <- purrr::map(c("TT", "E2", "P4"), plot_lit_combined)
names(combined_plots) <- c("TT", "E2", "P4")

for (h in names(combined_plots)) {
  print(combined_plots[[h]])
  ggsave(paste0("lit_", tolower(h), "_combined.svg"),
         plot = combined_plots[[h]], path = svg_path, width = 4.5, height = 4.5)
  ggsave(paste0("lit_", tolower(h), "_combined.png"),
         plot = combined_plots[[h]], path = svg_path, width = 4.5, height = 4.5,
         dpi = 300)
}
```

```{r lit_patchwork, echo=TRUE, fig.width=12, fig.height=4.5}
lit_grid <- (combined_plots$TT | combined_plots$E2 | combined_plots$P4) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")

print(lit_grid)
ggsave("lit_all_hormones_combined.svg", plot = lit_grid, path = svg_path,
       width = 12, height = 4.5)
ggsave("lit_all_hormones_combined.png", plot = lit_grid, path = svg_path,
       width = 12, height = 4.5, dpi = 300)
```
# Receptor Kd Reference Lines

Adds horizontal reference lines showing the tissue-equivalent dissociation
constant for each hormone's cognate receptor(s).

Kd values are free-ligand affinities converted to a tissue concentration:

    Kd (nM) x MW (g/mol) = pg/mL, then / brain density (mg/mL) = pg/mg

Note that measured tissue concentrations are *total* (protein-bound plus free
plus intracellular), whereas Kd describes free ligand at the receptor. These
lines therefore mark a permissive floor -- concentrations above the line are in
a range where receptor binding is possible -- not an occupancy estimate.

```{r kd_setup, echo=TRUE}
# Single source of truth. Fetal brain density estimate appropriate to the
# GW16-24 window (Ultrasound Obstet Gynecol, doi:10.1002/uog.70140).
# Change here to propagate to every panel and to the literature conversions.
BRAIN_DENSITY <- 1120  # mg/mL

kd_receptors <- tribble(
  ~hormone, ~receptor, ~kd_nM, ~mw_g_mol,
  "TT",     "AR",       0.26,   272.4,
  "E2",     "ERa",      0.13,   288.4,
  "E2",     "ERB1",     0.15,   288.4,
  "E2",     "ERB2",     1.84,   288.4,
  "P4",     "PR",       0.38,   314.5
) %>%
  mutate(
    pg_per_mL = kd_nM * mw_g_mol,
    kd_pg_mg  = pg_per_mL / BRAIN_DENSITY,
    hormone   = factor(hormone, levels = c("TT", "E2", "P4")),
    # ERB2 is ~10x weaker than the other estrogen receptors; distinguish it
    linetype  = if_else(receptor == "ERB2", "dotted", "dashed")
  )

kable(
  kd_receptors %>%
    select(hormone, receptor, kd_nM, mw_g_mol, pg_per_mL, kd_pg_mg) %>%
    mutate(across(c(pg_per_mL, kd_pg_mg), ~round(.x, 4))),
  caption = paste0("Receptor Kd converted to tissue-equivalent pg/mg ",
                   "(brain density = ", BRAIN_DENSITY, " mg/mL)")
)

#' Overlay receptor Kd reference lines on an existing hormone plot.
#'
#' @param p       a ggplot / ggpubr object with hormone concentration on y
#' @param h       one of "TT", "E2", "P4"
#' @param label   whether to annotate each line with the receptor name
#' @param x_label x position (in discrete-axis units) for the annotation
#' @param lt      optional linetype applied to all lines, overriding the
#'                per-receptor default. Use "solid" on linear axes where the
#'                lines sit close to zero and dashes are hard to resolve.
add_kd_lines <- function(p, h, label = TRUE, x_label = Inf, lt = NULL) {
  d <- filter(kd_receptors, hormone == h)
  if (nrow(d) == 0) return(p)
  if (!is.null(lt)) d$linetype <- lt

  p <- p +
    geom_hline(data = d,
               aes(yintercept = kd_pg_mg, linetype = linetype),
               colour = "grey40", linewidth = 0.5,
               inherit.aes = FALSE, show.legend = FALSE) +
    scale_linetype_identity()

  if (label) {
    p <- p +
      geom_text(data = d,
                aes(x = x_label, y = kd_pg_mg, label = receptor),
                hjust = 1.1, vjust = -0.4, size = 3, colour = "grey40",
                inherit.aes = FALSE)
  }
  p
}
```

## Applying to the sex-comparison panels

`add_kd_lines()` returns a modified plot, so assign the existing `ggboxplot()`
call to an object first rather than piping straight into `ggsave()`.

### Linear axis

Kd values are small relative to the measured ranges for E2 and P4, so on a
linear axis these lines sit very close to zero. They are drawn solid here for
visibility; see the log10 versions below for the readable comparison.

```{r kd_sex_panels_linear, echo=TRUE}
# --- TT, outlier-removed, linear scale ---
p_tt_lin <- ggboxplot(clean_targeted, x = "Sex", y = "`TT (pg/mg)`",
                      color = "Sex", width = 0.5,
                      ylab = "TT (pg/mg)", xlab = "Sex",
                      add = "jitter") +
  labs(title = paste(clean_tt_test_type,
                     ": statistic =", round(clean_tt_test_statistic, 2),
                     ": pvalue =", formatC(clean_tt_p_value, format = "e", digits = 2)))

p_tt_lin <- add_kd_lines(p_tt_lin, "TT", lt = "solid")
print(p_tt_lin)
ggsave("clean_tt_statistics_sex_kd.svg", plot = p_tt_lin, path = svg_path)
ggsave("clean_tt_statistics_sex_kd.png", plot = p_tt_lin, path = svg_path, dpi = 300)

# --- E2, outlier-removed, linear scale ---
p_e2_lin <- ggboxplot(clean_targeted, x = "Sex", y = "`17B-E2 (pg/mg)`",
                      color = "Sex", width = 0.5,
                      ylab = "17B-E2 (pg/mg)", xlab = "Sex",
                      add = "jitter") +
  labs(title = paste(clean_e2_test_type,
                     ": statistic =", round(clean_e2_test_statistic, 2),
                     ": pvalue =", formatC(clean_e2_p_value, format = "e", digits = 2)))

p_e2_lin <- add_kd_lines(p_e2_lin, "E2", lt = "solid")
print(p_e2_lin)
ggsave("clean_e2_statistics_sex_kd.svg", plot = p_e2_lin, path = svg_path)
ggsave("clean_e2_statistics_sex_kd.png", plot = p_e2_lin, path = svg_path, dpi = 300)

# --- P4, outlier-removed, linear scale ---
p_p4_lin <- ggboxplot(clean_targeted, x = "Sex", y = "`P4 (pg/mg)`",
                      color = "Sex", width = 0.5,
                      ylab = "P4 (pg/mg)", xlab = "Sex",
                      add = "jitter") +
  labs(title = paste(clean_p4_test_type,
                     ": statistic =", round(clean_p4_test_statistic, 2),
                     ": pvalue =", formatC(clean_p4_p_value, format = "e", digits = 2)))

p_p4_lin <- add_kd_lines(p_p4_lin, "P4", lt = "solid")
print(p_p4_lin)
ggsave("clean_p4_statistics_sex_kd.svg", plot = p_p4_lin, path = svg_path)
ggsave("clean_p4_statistics_sex_kd.png", plot = p_p4_lin, path = svg_path, dpi = 300)
```

### Log10 axis

```{r kd_sex_panels, echo=TRUE}
# --- TT, outlier-removed, log scale ---
p_tt <- ggboxplot(clean_targeted, x = "Sex", y = "`TT (pg/mg)`",
                  color = "Sex", width = 0.5,
                  ylab = "Log10 TT (pg/mg)", xlab = "Sex",
                  add = "jitter") +
  scale_y_log10() +
  labs(title = paste(clean_tt_test_type,
                     ": statistic =", round(clean_tt_test_statistic, 2),
                     ": pvalue =", formatC(clean_tt_p_value, format = "e", digits = 2)))

p_tt <- add_kd_lines(p_tt, "TT")
print(p_tt)
ggsave("log_clean_tt_statistics_sex_kd.svg", plot = p_tt, path = svg_path)
ggsave("log_clean_tt_statistics_sex_kd.png", plot = p_tt, path = svg_path, dpi = 300)

# --- E2, outlier-removed, log scale ---
p_e2 <- ggboxplot(clean_targeted, x = "Sex", y = "`17B-E2 (pg/mg)`",
                  color = "Sex", width = 0.5,
                  ylab = "Log10 17B-E2 (pg/mg)", xlab = "Sex",
                  add = "jitter") +
  scale_y_log10() +
  labs(title = paste(clean_e2_test_type,
                     ": statistic =", round(clean_e2_test_statistic, 2),
                     ": pvalue =", formatC(clean_e2_p_value, format = "e", digits = 2)))

p_e2 <- add_kd_lines(p_e2, "E2")
print(p_e2)
ggsave("log_clean_e2_statistics_sex_kd.svg", plot = p_e2, path = svg_path)
ggsave("log_clean_e2_statistics_sex_kd.png", plot = p_e2, path = svg_path, dpi = 300)

# --- P4, outlier-removed, log scale ---
p_p4 <- ggboxplot(clean_targeted, x = "Sex", y = "`P4 (pg/mg)`",
                  color = "Sex", width = 0.5,
                  ylab = "Log10 P4 (pg/mg)", xlab = "Sex",
                  add = "jitter") +
  scale_y_log10() +
  labs(title = paste(clean_p4_test_type,
                     ": statistic =", round(clean_p4_test_statistic, 2),
                     ": pvalue =", formatC(clean_p4_p_value, format = "e", digits = 2)))

p_p4 <- add_kd_lines(p_p4, "P4")
print(p_p4)
ggsave("log_clean_p4_statistics_sex_kd.svg", plot = p_p4, path = svg_path)
ggsave("log_clean_p4_statistics_sex_kd.png", plot = p_p4, path = svg_path, dpi = 300)
```

## Applying to the literature-comparison panels

Requires the `lit_combined` chunk to have run first.

```{r kd_lit_panels, echo=TRUE, fig.width=4.5, fig.height=4.5}
combined_plots_kd <- purrr::imap(combined_plots, ~add_kd_lines(.x, .y))

for (h in names(combined_plots_kd)) {
  print(combined_plots_kd[[h]])
  ggsave(paste0("lit_", tolower(h), "_combined_kd.svg"),
         plot = combined_plots_kd[[h]], path = svg_path,
         width = 4.5, height = 4.5)
  ggsave(paste0("lit_", tolower(h), "_combined_kd.png"),
         plot = combined_plots_kd[[h]], path = svg_path,
         width = 4.5, height = 4.5, dpi = 300)
}
```

```{r kd_lit_patchwork, echo=TRUE, fig.width=12, fig.height=4.5}
lit_grid_kd <- (combined_plots_kd$TT | combined_plots_kd$E2 | combined_plots_kd$P4) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")

print(lit_grid_kd)
ggsave("lit_all_hormones_combined_kd.svg", plot = lit_grid_kd, path = svg_path,
       width = 12, height = 4.5)
ggsave("lit_all_hormones_combined_kd.png", plot = lit_grid_kd, path = svg_path,
       width = 12, height = 4.5, dpi = 300)
```

## Fold change over Kd

Summarises how far measured concentrations sit above the binding threshold.

```{r kd_fold, echo=TRUE}
kd_fold <- our_hormones %>%
  group_by(hormone) %>%
  summarise(median_pg_mg = median(value, na.rm = TRUE),
            max_pg_mg    = max(value, na.rm = TRUE),
            .groups = "drop") %>%
  left_join(select(kd_receptors, hormone, receptor, kd_pg_mg), by = "hormone") %>%
  mutate(fold_median = median_pg_mg / kd_pg_mg,
         fold_max    = max_pg_mg / kd_pg_mg,
         across(c(median_pg_mg, max_pg_mg, kd_pg_mg), ~round(.x, 4)),
         across(c(fold_median, fold_max), ~round(.x, 1)))

kable(kd_fold, caption = "Measured concentration relative to receptor Kd")
write.csv(kd_fold, "../../results/targeted/hormone_csv/hormone_kd_fold.csv",
          row.names = FALSE)
```
