---
title: "Prenatal Targeted Analysis"
author: "Sneha Chaturvedi"
date: "2025-05-02"
output: html_document
---
#hello
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) {
  install.packages("pacman")
  library(pacman)
} 
p_load(tidyverse,readxl,dplyr,stats,ggplot2,ggpubr)
```

```{r LoadDataset, echo=TRUE}
targeted = read_excel('./Datasets/Targeted Comparison.xlsx',sheet="Targeted Comparison",range=NULL)
#Check dataset class
print(sapply(targeted,class))
```
Outlier Detection
```{r outlier, echo=TRUE}
#calculate z-scores for each hormone and attach to dataset
targeted$z_scores_tt <- scale(targeted$`TT (ng/mL)`)
targeted$z_scores_e2 <- scale(targeted$`17B-E2 (ng/mL)`)
targeted$z_scores_p4 <- scale(targeted$`P4 (ng/mL)`)
print(which(abs(targeted$z_scores_tt) > 3))
print(which(abs(targeted$z_scores_e2) > 3))
print(which(abs(targeted$z_scores_p4) > 3))
```

Create dataset with no outliers > manual 
Load
```{r LoadxOutlierDataset, echo=TRUE}
targeted_Xoutliers = read_excel('./Datasets/Targeted Comparison.xlsx',sheet="Outliers Removed",range=NULL)

```

Testosterone sex Analysis (full dataset)
```{r TT_SexComparison}

#convert sex column to factor
targeted$Sex <- as.factor(targeted$Sex)
targeted$`TT (ng/mL)`<- as.numeric(targeted$`TT (ng/mL)`)
targeted$`P4 (ng/mL)`<- as.numeric(targeted$`P4 (ng/mL)`)
targeted$`17B-E2 (ng/mL)`<- as.numeric(targeted$`17B-E2 (ng/mL)`)

#check distribution of hormones for normality
# Plot the distribution of TT (ng/mL) for the entire dataset
ggplot(targeted, aes(x = `TT (ng/mL)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of TT (ng/mL) - Total') +
  xlab('TT (ng/mL)') +
  ylab('Count')

# Check the distribution of TT (ng/mL) for normality by Sex
ggplot(targeted, aes(x = `TT (ng/mL)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of TT (ng/mL) by Sex') +
  xlab('TT (ng/mL)') +
  ylab('Count')

# Generate QQ plots
# QQ plot for total distribution
ggqqplot(targeted$`TT (ng/mL)`, ggtitle = "QQ Plot of TT (ng/mL) - Total")


# Perform Shapiro-Wilk normality test for each sex group
shapiro_tt <- shapiro.test(targeted$`TT (ng/mL)`)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (shapiro_tt$p.value > 0.05) {
  # Perform t-test
  t_test <- t.test(`TT (ng/mL)` ~ Sex, data = targeted)
  print(t_test)
} else {
  # Perform Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(`TT (ng/mL)` ~ Sex, data = targeted)
  print(wilcox_test)
}

# Visualize results with boxplot
ggboxplot(targeted, x = "Sex", y = "`TT (ng/mL)`",
          color = "Sex", 
          ylab = "TT (ng/mL)", xlab = "Sex",
          add = "jitter")
```

Testosterone sex Analysis (outlier dataset)
```{r SexComparison}

#convert sex column to factor
targeted_Xoutliers$Sex <- as.factor(targeted_Xoutliers$Sex)
targeted_Xoutliers$`TT (ng/mL)`<- as.numeric(targeted_Xoutliers$`TT (ng/mL)`)
targeted_Xoutliers$`P4 (ng/mL)`<- as.numeric(targeted_Xoutliers$`P4 (ng/mL)`)
targeted_Xoutliers$`17B-E2 (ng/mL)`<- as.numeric(targeted_Xoutliers$`17B-E2 (ng/mL)`)

#check distribution of hormones for normality
# Plot the distribution of TT (ng/mL) for the entire dataset
ggplot(targeted_Xoutliers, aes(x = `TT (ng/mL)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of TT (ng/mL) - Total') +
  xlab('TT (ng/mL)') +
  ylab('Count')

# Check the distribution of TT (ng/mL) for normality by Sex
ggplot(targeted_Xoutliers, aes(x = `TT (ng/mL)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of TT (ng/mL) by Sex') +
  xlab('TT (ng/mL)') +
  ylab('Count')

# Generate QQ plots
# QQ plot for total distribution
ggqqplot(targeted_Xoutliers$`TT (ng/mL)`, ggtitle = "QQ Plot of TT (ng/mL) - Total")


# Perform Shapiro-Wilk normality test for each sex group
shapiro_tt_xoutlier <- shapiro.test(targeted_Xoutliers$`TT (ng/mL)`)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (shapiro_tt_xoutlier$p.value > 0.05) {
  # Perform t-test
  t_test_xoutlier <- t.test(`TT (ng/mL)` ~ Sex, data = targeted_Xoutliers)
  print(t_test_xoutlier)
} else {
  # Perform Wilcoxon rank-sum test
  wilcox_test_xoutlier <- wilcox.test(`TT (ng/mL)` ~ Sex, data = targeted_Xoutliers)
  print(wilcox_test_xoutlier)
}

# Visualize results with boxplot
ggboxplot(targeted_Xoutliers, x = "Sex", y = "`TT (ng/mL)`",
          color = "Sex", 
          ylab = "TT (ng/mL)", xlab = "Sex",
          add = "jitter")
```

Estrogen
```{r Estrogen, echo=TRUE}
#check distribution of hormones for normality
# Plot the distribution of 17B-E2 (ng/mL) for the entire dataset
ggplot(targeted, aes(x = `17B-E2 (ng/mL)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of 17B-E2 (ng/mL) - Total') +
  xlab('17B-E2 (ng/mL)') +
  ylab('Count')

# Check the distribution of 17B-E2 (ng/mL) for normality by Sex
ggplot(targeted, aes(x = `17B-E2 (ng/mL)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of 17B-E2 (ng/mL) by Sex') +
  xlab('TT (ng/mL)') +
  ylab('Count')

# Generate QQ plots
# QQ plot for total distribution
ggqqplot(targeted$`17B-E2 (ng/mL)`, ggtitle = "QQ Plot of 17B-E2 (ng/mL) - Total")


# Perform Shapiro-Wilk normality test for each sex group
shapiro_e2 <- shapiro.test(targeted$`17B-E2 (ng/mL)`)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (shapiro_e2$p.value > 0.05) {
  # Perform t-test
  t_test <- t.test(`17B-E2 (ng/mL)` ~ Sex, data = targeted)
  print(t_test)
} else {
  # Perform Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(`17B-E2 (ng/mL)` ~ Sex, data = targeted)
  print(wilcox_test)
}

# Visualize results with boxplot
ggboxplot(targeted, x = "Sex", y = "`17B-E2 (ng/mL)`",
          color = "Sex", 
          ylab = "17B-E2 (ng/mL)", xlab = "Sex",
          add = "jitter")

##Outlier Removal
#check distribution of hormones for normality
# Plot the distribution of 17B-E2 (ng/mL) for the entire dataset
ggplot(targeted_Xoutliers, aes(x = `17B-E2 (ng/mL)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of 17B-E2 (ng/mL) - Total') +
  xlab('17B-E2 (ng/mL)') +
  ylab('Count')

# Check the distribution of 17B-E2 (ng/mL) for normality by Sex
ggplot(targeted_Xoutliers, aes(x = `17B-E2 (ng/mL)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of 17B-E2 (ng/mL) by Sex') +
  xlab('TT (ng/mL)') +
  ylab('Count')

# Generate QQ plots
# QQ plot for total distribution
ggqqplot(targeted_Xoutliers$`17B-E2 (ng/mL)`, ggtitle = "QQ Plot of 17B-E2 (ng/mL) - Total")


# Perform Shapiro-Wilk normality test for each sex group
shapiro_e2 <- shapiro.test(targeted_Xoutliers$`17B-E2 (ng/mL)`)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (shapiro_e2$p.value > 0.05) {
  # Perform t-test
  t_test <- t.test(`17B-E2 (ng/mL)` ~ Sex, data = targeted_Xoutliers)
  print(t_test)
} else {
  # Perform Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(`17B-E2 (ng/mL)` ~ Sex, data = targeted_Xoutliers)
  print(wilcox_test)
}

# Visualize results with boxplot
ggboxplot(targeted_Xoutliers, x = "Sex", y = "`17B-E2 (ng/mL)`",
          color = "Sex", 
          ylab = "17B-E2 (ng/mL)", xlab = "Sex",
          add = "jitter")
```

Progesterone

```{r Progesterone, echo=TRUE}
##Outlier Removal
#check distribution of hormones for normality
# Plot the distribution of 17B-E2 (ng/mL) for the entire dataset
ggplot(targeted_Xoutliers, aes(x = `P4 (ng/mL)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of P4 (ng/mL) - Total') +
  xlab('P4 (ng/mL)') +
  ylab('Count')

# Check the distribution of P4 (ng/mL) for normality by Sex
ggplot(targeted_Xoutliers, aes(x = `P4 (ng/mL)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of P4 (ng/mL) by Sex') +
  xlab('TT (ng/mL)') +
  ylab('Count')

# Generate QQ plots
# QQ plot for total distribution
ggqqplot(targeted_Xoutliers$`P4 (ng/mL)`, ggtitle = "QQ Plot of P4 (ng/mL) - Total")


# Perform Shapiro-Wilk normality test for each sex group
shapiro_e2 <- shapiro.test(targeted_Xoutliers$`P4 (ng/mL)`)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (shapiro_e2$p.value > 0.05) {
  # Perform t-test
  t_test <- t.test(`P4 (ng/mL)` ~ Sex, data = targeted_Xoutliers)
  print(t_test)
} else {
  # Perform Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(`P4 (ng/mL)` ~ Sex, data = targeted_Xoutliers)
  print(wilcox_test)
}

# Visualize results with boxplot
ggboxplot(targeted_Xoutliers, x = "Sex", y = "`P4 (ng/mL)`",
          color = "Sex", 
          ylab = "P4 (ng/mL)", xlab = "Sex",
          add = "jitter")

#full dataset
#check distribution of hormones for normality
# Plot the distribution of P4 (ng/mL) for the entire dataset
ggplot(targeted, aes(x = `P4 (ng/mL)`)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30, fill = 'blue') +
  ggtitle('Distribution of P4 (ng/mL) - Total') +
  xlab('P4 (ng/mL)') +
  ylab('Count')

# Check the distribution of P4 (ng/mL) for normality by Sex
ggplot(targeted, aes(x = `P4 (ng/mL)`, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = 'identity', bins = 30) +
  facet_wrap(~ Sex) +
  ggtitle('Distribution of P4 (ng/mL) by Sex') +
  xlab('P4 (ng/mL)') +
  ylab('Count')

# Generate QQ plots
# QQ plot for total distribution
ggqqplot(targeted$`P4 (ng/mL)`, ggtitle = "QQ Plot of P4 (ng/mL) - Total")


# Perform Shapiro-Wilk normality test for each sex group
shapiro_p4 <- shapiro.test(targeted$`P4 (ng/mL)`)

# Check the p-values
# If  > 0.05, we assume normality and can use a parametric test (t-test)
# Otherwise, use a non-parametric test (Wilcoxon rank-sum test)

if (shapiro_p4$p.value > 0.05) {
  # Perform t-test
  t_test <- t.test(`P4 (ng/mL)` ~ Sex, data = targeted)
  print(t_test)
} else {
  # Perform Wilcoxon rank-sum test
  wilcox_test <- wilcox.test(`P4 (ng/mL)` ~ Sex, data = targeted)
  print(wilcox_test)
}

# Visualize results with boxplot
ggboxplot(targeted, x = "Sex", y = "`P4 (ng/mL)`",
          color = "Sex", 
          ylab = "P4 (ng/mL)", xlab = "Sex",
          add = "jitter")

```

Relationship with gestational week
Here is a script that will graph TT (ng/mL) over gestational week (GW) and calculate if there is a significant effect of gestational week (GW) on TT concentration using linear regression:

```{R GestationalWeek_TT, echo=TRUE}

targeted_Xoutliers$GW <- as.numeric(targeted_Xoutliers$GW)

# Scatter plot of TT (ng/mL) over gestational week (GW) with a linear regression line
ggplot(targeted_Xoutliers, aes(x = GW, y = `TT (ng/mL)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  ggtitle("TT (ng/mL) over Gestational Week") +
  xlab("Gestational Week (GW)") +
  ylab("TT (ng/mL)")

# Perform linear regression to test the effect of gestational week on TT concentration
lm_model <- lm(`TT (ng/mL)` ~ GW, data = targeted_Xoutliers)
summary(lm_model)
```
Estrogen and Progesterone time
```{R GestationalWeek, echo=TRUE}

# Scatter plot of P4 (ng/mL) over gestational week (GW) with a linear regression line
ggplot(targeted_Xoutliers, aes(x = GW, y = `P4 (ng/mL)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  ggtitle("P4 (ng/mL) over Gestational Week") +
  xlab("Gestational Week (GW)") +
  ylab("P4 (ng/mL)")

# Perform linear regression to test the effect of gestational week on P4 concentration
lm_model_p4 <- lm(`P4 (ng/mL)` ~ GW, data = targeted_Xoutliers)
summary(lm_model_p4)

# Scatter plot of 17B-E2 (ng/mL) over gestational week (GW) with a linear regression line
ggplot(targeted_Xoutliers, aes(x = GW, y = `17B-E2 (ng/mL)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  ggtitle("17B-E2 (ng/mL) over Gestational Week") +
  xlab("Gestational Week (GW)") +
  ylab("17B-E2 (ng/mL)")

# Perform linear regression to test the effect of gestational week on 17B-E2 concentration
lm_model_e2 <- lm(`17B-E2 (ng/mL)` ~ GW, data = targeted_Xoutliers)
summary(lm_model_e2)
```

Binned testosterone analysis

```{r BinnedTT, echo=TRUE}

# Split GW into three bins
targeted_Xoutliers$GW_bin <- cut(targeted$GW, breaks = 3, labels = c("Early", "Mid", "Late"))

# Check the distribution of samples in each bin
table(targeted_Xoutliers$GW_bin)

# Boxplot to visualize TT (ng/mL) across the three GW bins
ggplot(targeted_Xoutliers, aes(x = GW_bin, y = `TT (ng/mL)`)) +
  geom_boxplot(fill = "blue", alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  ggtitle("TT (ng/mL) across GW Bins") +
  xlab("GW Bins") +
  ylab("TT (ng/mL)")

# Perform one-way ANOVA
aov_model <- aov(`TT (ng/mL)` ~ GW_bin, data = targeted)
summary(aov_model)

# If significant, perform post-hoc tests
if (summary(aov_model)[[1]][["Pr(>F)"]][1] < 0.05) {
  post_hoc <- TukeyHSD(aov_model)
  print(post_hoc)
}
```
