---
title: "Sex and Gestational Week Analysis"
author: "Sneha M. Chaturvedi"
date: "Last updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

Statistical analysis of metabolite variation by sex and gestational week.
Tests include: GW Spearman correlation, GW limma, sex Wilcoxon rank-sum,
sex limma, sex PLS-DA (with permutation testing), GW PLS-DA, GW-adjusted sex
limma, sex-adjusted GW limma, sex x GW interaction limma, and UpSet plots
summarizing cross-method overlap of significant metabolites.

| | |
|---|---|
| **Inputs** | `results/untargeted/peak_area_clean.csv`, `results/untargeted/metabolite_annotations.csv`, `results/untargeted/sample_metadata.csv` |
| **Outputs** | Result CSVs in `results/untargeted/untargeted_clean/csv/`, figures in `results/untargeted/untargeted_clean/` |
| **Upstream** | `Exogenous_Flagging.Rmd` |
| **Downstream** | `Pathway_Analysis.Rmd`, `Analysis_Summary.Rmd` |

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Packages <- c("readxl", "stats", "ggfortify", "ggrepel", "svglite", "Biobase", "ComplexHeatmap", "limma", "tidyverse", "ropls", "cowplot", "UpSetR")

lapply(Packages, library, character.only = TRUE)
```

```{r source_function, include=FALSE}
# Source Functions
source("functions.R")
source("missingness_visualization_functions.R")
```

```{r set_path, include=FALSE}
root <- find_project_root()
png_path <- file.path(root, "results/untargeted/untargeted_clean")
svg_path <- file.path(root, "results/untargeted/untargeted_clean/svg")
csv_path <- file.path(root, "results/untargeted/untargeted_clean/csv")
```

## Load Dataset(s)

```{r LoadDataset, include=FALSE}
# Peak Area Values
peak_area <- read_csv(file.path(root, "results/untargeted/peak_area_clean.csv"), show_col_types = FALSE)

# Metabolite Annotations
compound_annotation <- read_csv(file.path(root, "results/untargeted/metabolite_annotations.csv"), show_col_types = FALSE)

# Sample Metadata
sample_metadata <- read_csv(file.path(root, "results/untargeted/sample_metadata.csv"), show_col_types = FALSE)
```


``` {r prep_dataframe}
# Convert metabolites data to matrix format
final_matrix <- peak_area %>%
  dplyr::select(starts_with("Sample"))

# Filter annotation to only include compounds present in peak area
compounds_in_peak <- unique(peak_area$Compound.ID)

compound_annotation <- compound_annotation %>%
  filter(Compound.ID %in% compounds_in_peak) %>%
  arrange(match(Name, rownames(final_matrix)))


metabolite_names <- compound_annotation$Name
final_matrix <- apply(final_matrix, 2, as.numeric)
rownames(final_matrix) <- metabolite_names

# Filter metadata to match metabolites samples
sample_metadata <- sample_metadata %>%
  filter(Sample %in% colnames(final_matrix)) %>%
  # Ensure the order matches the metabolites data
  arrange(match(Sample, colnames(final_matrix)))

# Create the metadata dataframe
sample_metadata_df <- as.data.frame(sample_metadata)
rownames(sample_metadata_df) <- sample_metadata_df$Sample

# Create Biobase ExpressionSet with the matched data
sample_phenoData <- AnnotatedDataFrame(sample_metadata_df)

# Create the metabolite metadata dataframe
metabolite_metadata_df <- as.data.frame(compound_annotation)
rownames(metabolite_metadata_df) <- metabolite_metadata_df$Name

metabolite_featureData <- AnnotatedDataFrame(metabolite_metadata_df)

# Create the ExpressionSet with both phenodata and featuredata
metabolomics <- ExpressionSet(
  assayData = final_matrix,
  phenoData = sample_phenoData,
  featureData = metabolite_featureData
)
```

```{r log_transform}
# Log2-transform peak intensities (log2(x + 1) to handle any zeros)
exprs(metabolomics) <- log2(exprs(metabolomics) + 1)

cat("Log2 transformation applied.\n")
cat("Intensity range before: check raw matrix if needed\n")
cat("Intensity range after:", range(exprs(metabolomics), na.rm = TRUE), "\n")
```
# GW Analysis

## GW Correlation
```{r gw_correlation}
# Calculate correlations and p-values between metabolites and metadata
metadata <- pData(metabolomics)
GW <- metadata$GW

# Function to calculate correlation and p-value
correlation_test <- function(x, y) {
  result <- cor.test(x, y, method = "spearman", exact = FALSE)
  c(correlation = result$estimate, p_value = result$p.value)
}

# Calculate correlations and p-values for each metabolite
metabolite_samplemetadata_stats <- data.frame(
  Name = fData(metabolomics)$Name,
  GW_corr_pval = t(apply(exprs(metabolomics), 1, function(x) correlation_test(x, GW)))
)

metabolite_samplemetadata_stats <- metabolite_samplemetadata_stats %>%
  mutate(
    GW_FDR = p.adjust(GW_corr_pval.p_value, method = "fdr")
  )


# For GW
# Scatter plot for significant GW correlations
significant_GW <- metabolite_samplemetadata_stats %>% filter(GW_FDR < 0.1)

significant_GW <- significant_GW %>% mutate(Name_display = truncate_labels(Name))

scatter_GW <- ggplot(significant_GW, aes(x = reorder(Name_display, GW_corr_pval.correlation.rho), y = GW_corr_pval.correlation.rho)) +
  geom_point(aes(size = -log10(GW_corr_pval.p_value), color = GW_corr_pval.correlation.rho)) +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "Metabolite",
    y = "Correlation with GW",
    title = "Significant Metabolite Correlations with Gestational Weeks (GW)"
  ) +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right", axis.text.x = element_text(angle = 90, hjust = 1))

save_dual_format(scatter_GW, "significant_GW",
  png_path = png_path, svg_path = svg_path,
  label_lookup = build_label_lookup(significant_GW$Name)
)

print(scatter_GW)

# Print top correlations
top_GW_correlations <- metabolite_samplemetadata_stats %>%
  arrange(desc(abs(GW_corr_pval.correlation.rho))) %>%
  head(10) %>%
  dplyr::select(Name, GW_corr_pval.correlation.rho, GW_corr_pval.p_value, GW_FDR)

print(top_GW_correlations)
```

## GW Limma
```{r gw_limma}
# limma linear model for GW associations
# Design uses ~ GW (with intercept)

metadata_limma <- pData(metabolomics)
peak_matrix_limma <- exprs(metabolomics) # features × samples

design_gw <- model.matrix(~GW, data = metadata_limma)

fit_gw <- lmFit(peak_matrix_limma, design_gw)
fit_gw <- eBayes(fit_gw)
gw_limma_results <- topTable(fit_gw, coef = "GW", number = Inf, adjust.method = "BH") %>%
  rownames_to_column("Name")

# Merge with annotation
gw_limma_results <- gw_limma_results %>%
  left_join(fData(metabolomics) %>% dplyr::select(Name, Compound.ID, Super.Class, Class),
    by = "Name"
  )

# Top hits bar chart - effect size (slope per GW week)
# Always show all FDR-significant hits; fill remaining slots up to 30 with
# top nominal-only hits ranked by absolute effect size
n_fdr_gw <- sum(gw_limma_results$adj.P.Val < 0.1)
top_gw <- bind_rows(
  gw_limma_results %>% filter(adj.P.Val < 0.1) %>%
    arrange(desc(abs(logFC))),
  gw_limma_results %>% filter(P.Value < 0.05, adj.P.Val >= 0.1) %>%
    arrange(desc(abs(logFC))) %>%
    head(max(0L, 30L - n_fdr_gw))
) %>%
  mutate(
    Direction = ifelse(logFC > 0, "Increases with GW", "Decreases with GW"),
    FDR_label = case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01 ~ "**",
      adj.P.Val < 0.1 ~ "*",
      TRUE ~ ""
    )
  )

gw_bar_lookup <- build_label_lookup(top_gw$Name)
top_gw <- top_gw %>% mutate(Name_display = truncate_labels(Name))

gw_bar <- ggplot(top_gw, aes(x = reorder(Name, logFC), y = logFC, fill = Direction)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(
    aes(
      label = Name_display,
      y     = logFC * 0.98,
      hjust = ifelse(logFC > 0, 1, 0)
    ),
    size = 2.8, color = "white"
  ) +
  geom_text(
    aes(
      label = FDR_label,
      hjust = ifelse(logFC > 0, -0.2, 1.2)
    ),
    size = 4, fontface = "bold"
  ) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Increases with GW" = "#1B5E20",
    "Decreases with GW" = "#C8E6C9"
  )) +
  theme_minimal() +
  labs(
    x       = NULL,
    y       = "Slope (log2 peak area per GW week)",
    title   = "Top GW-Associated Metabolites (limma, p < 0.05)",
    caption = "* FDR < 0.1   ** FDR < 0.01   *** FDR < 0.001"
  ) +
  theme(
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title   = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0)
  )

save_dual_format(gw_bar, "gw_limma_bar",
  png_path = png_path, svg_path = svg_path,
  label_lookup = gw_bar_lookup
)
print(gw_bar)

# Trajectory plots for all GW FDR-significant hits (shown immediately after bar chart)
gw_fdr_traj_hits <- gw_limma_results %>%
  filter(adj.P.Val < 0.1) %>%
  arrange(desc(abs(logFC))) %>%
  mutate(label = paste0(
    "FDR=", formatC(adj.P.Val, format = "e", digits = 2),
    "  slope=", round(logFC, 3)
  ))

if (nrow(gw_fdr_traj_hits) > 0) {
  gw_fdr_traj_plots <- build_trajectory_plots(
    metabolite_names = gw_fdr_traj_hits$Name,
    eset             = metabolomics,
    title_suffixes   = paste0("GW FDR significant | ", gw_fdr_traj_hits$label)
  )
  save_trajectory_grids(gw_fdr_traj_plots, "gw_trajectory_fdr", png_path, svg_path)
} else {
  cat("No FDR-significant GW hits to plot\n")
}

# Ranked dot plot - all metabolites
gw_dot <- ggplot(
  gw_limma_results %>%
    arrange(logFC) %>%
    mutate(rank = row_number()),
  aes(x = rank, y = logFC, color = adj.P.Val < 0.1)
) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(
    values = c("FALSE" = "grey70", "TRUE" = "red"),
    name = "FDR < 0.1"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  theme_minimal() +
  labs(
    x = "Metabolite rank",
    y = "Slope (log2 peak area per GW week)",
    title = "GW Association Effect Sizes - All Metabolites"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(gw_dot, "gw_limma_dot", png_path = png_path, svg_path = svg_path)
print(gw_dot)

# Save results
write_csv(gw_limma_results, file.path(csv_path, "gw_limma_results.csv"))

cat("GW limma results summary:\n")
cat("  Nominal p < 0.05:", sum(gw_limma_results$P.Value < 0.05), "\n")
cat("  FDR < 0.1:", sum(gw_limma_results$adj.P.Val < 0.1), "\n")
```


# Sex Differential Analysis
## Univariate Testing
```{r sex_analysis}
# Calculate p-values for sex differences using Wilcoxon rank-sum test
metadata <- pData(metabolomics)
sex_groups <- metadata$Sex

# Function to calculate Wilcoxon rank-sum test p-value and statistic
wilcoxon_test <- function(x, sex) {
  # Remove missing values
  valid_idx <- !is.na(x)
  x_clean <- x[valid_idx]
  sex_clean <- sex[valid_idx]

  # Only test if we have both groups
  if (length(unique(sex_clean)) < 2) {
    return(c(statistic = NA, p_value = NA))
  }

  result <- wilcox.test(x_clean[sex_clean == "F"],
    x_clean[sex_clean == "M"],
    exact = FALSE
  )
  c(statistic = as.numeric(result$statistic), p_value = result$p.value)
}

# Perform Wilcoxon rank-sum test for each metabolite
wilcoxon_results <- t(apply(exprs(metabolomics), 1, wilcoxon_test, sex = sex_groups))

# Ensure column names are set
colnames(wilcoxon_results) <- c("statistic", "p_value")

# Calculate mean values by sex for fold change calculation
mean_male <- apply(exprs(metabolomics)[, sex_groups == "M"], 1, mean, na.rm = TRUE)
mean_female <- apply(exprs(metabolomics)[, sex_groups == "F"], 1, mean, na.rm = TRUE)

# Create comprehensive statistics data frame
metabolite_sex_stats <- data.frame(
  Name = fData(metabolomics)$Name,
  Compound.ID = fData(metabolomics)$Compound.ID,
  Wilcoxon_statistic = wilcoxon_results[, "statistic"],
  p_value = wilcoxon_results[, "p_value"],
  Mean_Female = mean_female,
  Mean_Male = mean_male,
  logFC = log2(mean_male / mean_female)
) %>%
  mutate(
    FDR = p.adjust(p_value, method = "fdr"),
    Bonferroni = p.adjust(p_value, method = "bonferroni"),
    Significant_FDR = FDR < 0.1,
    Significant_Bonferroni = Bonferroni < 0.05,
    Direction = case_when(
      logFC > 0 ~ "Higher in Males",
      logFC < 0 ~ "Higher in Females",
      TRUE ~ "No difference"
    )
  )
metabolite_sex_stats <- metabolite_sex_stats %>% mutate(Name_display = truncate_labels(Name))

sig_metabolites_fdr <- metabolite_sex_stats %>% filter(p_value < 0.05)

# Volcano plot with FDR correction - NO bubble size
volcano_FDR <- ggplot(metabolite_sex_stats, aes(x = logFC, y = -log10(p_value))) +
  geom_point(aes(color = Significant_FDR), alpha = 0.6) +
  geom_text_repel(
    data = sig_metabolites_fdr,
    aes(label = Name_display),
    size = 3, max.overlaps = 15
  ) +
  scale_color_manual(
    values = c("FALSE" = "grey", "TRUE" = "red"),
    name = "FDR < 0.1"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_minimal() +
  labs(
    x = "Log2 Fold Change (Male/Female)",
    y = "-log10(p-value)",
    title = "Sex Differential Analysis - Volcano Plot (FDR)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )

save_dual_format(volcano_FDR, "sex_volcano_FDR",
  png_path = png_path, svg_path = svg_path,
  label_lookup = build_label_lookup(metabolite_sex_stats$Name)
)
print(volcano_FDR)

# Volcano plot with Bonferroni correction - NO bubble size
volcano_Bonferroni <- ggplot(metabolite_sex_stats, aes(x = logFC, y = -log10(p_value))) +
  geom_point(aes(color = Significant_Bonferroni), alpha = 0.6) +
  geom_text_repel(
    data = sig_metabolites_fdr,
    aes(label = Name_display),
    size = 3, max.overlaps = 15
  ) +
  scale_color_manual(
    values = c("FALSE" = "grey", "TRUE" = "red"),
    name = "Bonferroni < 0.05"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_minimal() +
  labs(
    x = "Log2 Fold Change (Male/Female)",
    y = "-log10(p-value)",
    title = "Sex Differential Analysis - Volcano Plot (Bonferroni)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )

save_dual_format(volcano_Bonferroni, "sex_volcano_Bonferroni",
  png_path = png_path, svg_path = svg_path,
  label_lookup = build_label_lookup(metabolite_sex_stats$Name)
)
print(volcano_Bonferroni)

# Summary statistics
cat("\n=== Sex Differential Analysis Summary ===\n")
cat("Total metabolites analyzed:", nrow(metabolite_sex_stats), "\n")
cat("Significant at FDR < 0.1:", sum(metabolite_sex_stats$Significant_FDR, na.rm = TRUE), "\n")
cat("Significant at Bonferroni < 0.05:", sum(metabolite_sex_stats$Significant_Bonferroni, na.rm = TRUE), "\n")

# Direction summary
cat("\nAmong FDR-significant metabolites:\n")
direction_summary <- metabolite_sex_stats %>%
  filter(Significant_FDR) %>%
  count(Direction)
print(direction_summary)

# Top metabolites by significance
cat("\n=== Top 10 Most Significant Metabolites ===\n")
top_metabolites <- metabolite_sex_stats %>%
  arrange(p_value) %>%
  head(10) %>%
  dplyr::select(Name, logFC, Mean_Female, Mean_Male, p_value, FDR, Bonferroni, Direction)
print(top_metabolites)

# Top metabolites with largest fold changes (FDR significant only)
cat("\n=== Top 10 Largest Fold Changes (FDR < 0.1) ===\n")
top_FC <- metabolite_sex_stats %>%
  filter(Significant_FDR) %>%
  arrange(desc(abs(logFC))) %>%
  head(10) %>%
  dplyr::select(Name, logFC, Mean_Female, Mean_Male, p_value, FDR, Direction)
print(top_FC)

# Save results to CSV
write_csv(
  metabolite_sex_stats,
  file.path(csv_path, "sex_differential_analysis.csv")
)

cat("\nResults saved to:", file.path(csv_path, "sex_differential_analysis.csv"), "\n")
```
### By Annotation Group
```{r group_sex_analysis}
# Function to collapse metabolites by annotation category
collapse_metabolites_by_group <- function(eset, group_col) {
  # Get expression data and annotations
  expr_data <- exprs(eset)
  annotations <- fData(eset)

  # Check if column exists
  if (!group_col %in% colnames(annotations)) {
    warning(paste("Column", group_col, "not found in annotations"))
    return(NULL)
  }

  # Remove rows with missing group information
  valid_idx <- !is.na(annotations[[group_col]]) &
    annotations[[group_col]] != "" &
    annotations[[group_col]] != "NA" &
    !is.na(annotations[[group_col]])

  expr_data_clean <- expr_data[valid_idx, ]
  groups <- annotations[[group_col]][valid_idx]

  cat("  Valid metabolites with", group_col, "annotation:", sum(valid_idx), "\n")

  # Collapse by taking median of each group
  unique_groups <- unique(groups)
  collapsed_data <- matrix(NA, nrow = length(unique_groups), ncol = ncol(expr_data))
  rownames(collapsed_data) <- unique_groups
  colnames(collapsed_data) <- colnames(expr_data)

  for (i in seq_along(unique_groups)) {
    group_idx <- groups == unique_groups[i]
    if (sum(group_idx) == 1) {
      collapsed_data[i, ] <- expr_data_clean[group_idx, ]
    } else {
      collapsed_data[i, ] <- apply(expr_data_clean[group_idx, , drop = FALSE], 2, median, na.rm = TRUE)
    }
  }

  # Create summary annotation
  collapsed_annotation <- data.frame(
    Group = unique_groups,
    N_metabolites = as.numeric(table(groups)[unique_groups]),
    Category = group_col
  )
  rownames(collapsed_annotation) <- unique_groups

  # Create new ExpressionSet
  collapsed_phenoData <- phenoData(eset)
  collapsed_featureData <- AnnotatedDataFrame(collapsed_annotation)

  collapsed_eset <- ExpressionSet(
    assayData = collapsed_data,
    phenoData = collapsed_phenoData,
    featureData = collapsed_featureData
  )

  return(collapsed_eset)
}

# Function to perform sex analysis on collapsed data
perform_collapsed_sex_analysis <- function(eset, category_name) {
  metadata <- pData(eset)
  sex_groups <- metadata$Sex

  # Wilcoxon test function
  wilcoxon_test <- function(x, sex) {
    valid_idx <- !is.na(x)
    x_clean <- x[valid_idx]
    sex_clean <- sex[valid_idx]

    if (length(unique(sex_clean)) < 2) {
      return(c(statistic = NA, p_value = NA))
    }

    result <- wilcox.test(x_clean[sex_clean == "F"],
      x_clean[sex_clean == "M"],
      exact = FALSE
    )
    c(statistic = as.numeric(result$statistic), p_value = result$p.value)
  }

  # Perform tests
  wilcoxon_results <- t(apply(exprs(eset), 1, wilcoxon_test, sex = sex_groups))
  colnames(wilcoxon_results) <- c("statistic", "p_value")

  # Calculate means
  mean_male <- apply(exprs(eset)[, sex_groups == "M"], 1, mean, na.rm = TRUE)
  mean_female <- apply(exprs(eset)[, sex_groups == "F"], 1, mean, na.rm = TRUE)

  # Create results data frame
  results <- data.frame(
    Group = fData(eset)$Group,
    N_metabolites = fData(eset)$N_metabolites,
    Category = category_name,
    Wilcoxon_statistic = wilcoxon_results[, "statistic"],
    p_value = wilcoxon_results[, "p_value"],
    Mean_Female = mean_female,
    Mean_Male = mean_male,
    logFC = log2(mean_male / mean_female)
  ) %>%
    mutate(
      Group_display = truncate_labels(Group),
      FDR = p.adjust(p_value, method = "fdr"),
      Bonferroni = p.adjust(p_value, method = "bonferroni"),
      Significant_FDR = FDR < 0.1,
      Significant_Bonferroni = Bonferroni < 0.05,
      Significant_nominal = p_value < 0.05,
      Direction = case_when(
        logFC > 0 ~ "Higher in Males",
        logFC < 0 ~ "Higher in Females",
        TRUE ~ "No difference"
      )
    ) %>%
    arrange(p_value)

  return(results)
}

# Analyze each available grouping category
grouping_cols <- c("Super.Class", "Class", "Sub.Class", "Pathways")

all_collapsed_results <- list()

for (group_col in grouping_cols) {
  cat("\n", rep("=", 80), "\n", sep = "")
  cat("Analyzing by", group_col, "\n")
  cat(rep("=", 80), "\n\n", sep = "")

  # Collapse metabolites
  collapsed_eset <- collapse_metabolites_by_group(metabolomics, group_col)

  if (!is.null(collapsed_eset)) {
    cat("  Number of", group_col, "groups:", nrow(collapsed_eset), "\n\n")

    # Perform sex analysis
    results <- perform_collapsed_sex_analysis(collapsed_eset, group_col)
    all_collapsed_results[[group_col]] <- results

    # Print summary
    cat("Summary:\n")
    cat("  Total groups:", nrow(results), "\n")
    cat("  Significant at p < 0.05:", sum(results$Significant_nominal, na.rm = TRUE), "\n")
    cat("  Significant at FDR < 0.1:", sum(results$Significant_FDR, na.rm = TRUE), "\n")
    cat("  Significant at Bonferroni < 0.05:", sum(results$Significant_Bonferroni, na.rm = TRUE), "\n\n")

    # Print top results
    cat("Top 10 groups by p-value:\n")
    print(results %>%
      head(10) %>%
      dplyr::select(Group, N_metabolites, logFC, p_value, FDR, Direction))
    cat("\n")

    results_sig <- results %>% filter(p_value < 0.05)

    # Create volcano plot - SIZE = N_metabolites
    volcano_plot <- ggplot(results, aes(x = logFC, y = -log10(p_value))) +
      geom_point(aes(color = Significant_FDR, size = N_metabolites), alpha = 0.6) +
      geom_text_repel(
        data = results_sig,
        aes(label = Group_display),
        size = 3, max.overlaps = 15
      ) +
      scale_color_manual(
        values = c("FALSE" = "grey", "TRUE" = "red"),
        name = "FDR < 0.1"
      ) +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
      geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
      theme_minimal() +
      labs(
        x = "Log2 Fold Change (Male/Female)",
        y = "-log10(p-value)",
        title = paste("Sex Differences by", group_col),
        size = "N Metabolites"
      ) +
      theme(
        plot.title = element_text(hjust = 0.5),
        legend.position = "right"
      )

    save_dual_format(volcano_plot,
      paste0("sex_volcano_", tolower(gsub("\\.", "_", group_col))),
      png_path = png_path,
      svg_path = svg_path
    )
    print(volcano_plot)

    # Save results
    write_csv(
      results,
      file.path(csv_path, paste0("sex_analysis_by_", tolower(gsub("\\.", "_", group_col)), ".csv"))
    )
  }
}
```

```{r combine_results}
# Combine all results for overview
if (length(all_collapsed_results) > 0) {
  combined_results <- bind_rows(all_collapsed_results)

  # Summary plot across all categories
  summary_sig <- combined_results %>%
    group_by(Category) %>%
    summarise(
      Total = n(),
      Nominal_sig = sum(Significant_nominal, na.rm = TRUE),
      FDR_sig = sum(Significant_FDR, na.rm = TRUE),
      Bonferroni_sig = sum(Significant_Bonferroni, na.rm = TRUE)
    ) %>%
    pivot_longer(
      cols = c(Nominal_sig, FDR_sig, Bonferroni_sig),
      names_to = "Significance_Level",
      values_to = "Count"
    ) %>%
    mutate(Significance_Level = factor(Significance_Level,
      levels = c("Nominal_sig", "FDR_sig", "Bonferroni_sig"),
      labels = c("p < 0.05", "FDR < 0.1", "Bonferroni < 0.05")
    ))

  summary_plot <- ggplot(summary_sig, aes(x = Category, y = Count, fill = Significance_Level)) +
    geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
    geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
    theme_minimal() +
    labs(
      title = "Significant Sex Differences by Annotation Category",
      x = "Annotation Category",
      y = "Number of Significant Groups",
      fill = "Significance Level"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )

  save_dual_format(summary_plot, "sex_collapsed_summary", png_path = png_path, svg_path = svg_path)
  print(summary_plot)

  # Save combined results
  write_csv(
    combined_results,
    file.path(csv_path, "sex_analysis_all_collapsed.csv")
  )

  # Print overall summary
  cat("\n", rep("=", 80), "\n", sep = "")
  cat("OVERALL SUMMARY\n")
  cat(rep("=", 80), "\n\n", sep = "")

  overall_summary <- combined_results %>%
    group_by(Category) %>%
    summarise(
      Total_Groups = n(),
      Groups_with_nominal_sig = sum(Significant_nominal, na.rm = TRUE),
      Groups_with_FDR_sig = sum(Significant_FDR, na.rm = TRUE),
      Groups_with_Bonferroni_sig = sum(Significant_Bonferroni, na.rm = TRUE),
      Median_logFC = median(abs(logFC), na.rm = TRUE)
    )

  print(overall_summary)

  # Most significant groups across all categories
  cat("\n\nTop 20 Most Significant Groups (all categories combined):\n")
  top_overall <- combined_results %>%
    arrange(p_value) %>%
    head(20) %>%
    dplyr::select(Category, Group, N_metabolites, logFC, p_value, FDR, Bonferroni, Direction)
  print(top_overall)
}
```

## Sex Limma
```{r sex_limma}
# limma linear model for sex differences
# Design uses ~ Sex (with intercept), reference level is Female
# logFC = log2(Male/Female)
metadata_limma <- pData(metabolomics)
peak_matrix_limma <- exprs(metabolomics)

# Ensure Sex is a factor with Female as reference
metadata_limma$Sex <- factor(metadata_limma$Sex, levels = c("F", "M"))

design_sex <- model.matrix(~Sex, data = metadata_limma)

fit_sex <- lmFit(peak_matrix_limma, design_sex)
fit_sex <- eBayes(fit_sex)
sex_limma_results <- topTable(fit_sex,
  coef = "SexM", number = Inf,
  adjust.method = "BH"
) %>%
  rownames_to_column("Name") %>%
  left_join(
    fData(metabolomics) %>%
      dplyr::select(Name, Compound.ID, Super.Class, Class),
    by = "Name"
  ) %>%
  mutate(
    Direction = ifelse(logFC > 0, "Higher in Males", "Higher in Females"),
    FDR_label = case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01 ~ "**",
      adj.P.Val < 0.1 ~ "*",
      TRUE ~ ""
    )
  )

cat("Sex limma results summary:\n")
cat("  Nominal p < 0.05:", sum(sex_limma_results$P.Value < 0.05), "\n")
cat("  FDR < 0.1:", sum(sex_limma_results$adj.P.Val < 0.1), "\n")

# Volcano plot 
volcano_sex_limma <- ggplot(
  sex_limma_results,
  aes(x = logFC, y = -log10(P.Value))
) +
  geom_point(aes(color = adj.P.Val < 0.1), alpha = 0.6) +
  scale_color_manual(
    values = c("FALSE" = "grey", "TRUE" = "red"),
    name = "FDR < 0.1"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_minimal() +
  labs(
    x = "Log2 Fold Change (Male/Female)",
    y = "-log10(p-value)",
    title = "Sex Differential Analysis - limma Volcano Plot",
    caption = "Positive logFC = higher in males"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(volcano_sex_limma, "sex_limma_volcano",
  png_path = png_path, svg_path = svg_path
)
print(volcano_sex_limma)

# Bar chart of top hits with FDR labels
# Always show all FDR-significant hits; fill remaining slots up to 30 with
# top nominal-only hits ranked by absolute effect size
n_fdr_sex <- sum(sex_limma_results$adj.P.Val < 0.1)
top_sex <- bind_rows(
  sex_limma_results %>% filter(adj.P.Val < 0.1) %>%
    arrange(desc(abs(logFC))),
  sex_limma_results %>% filter(P.Value < 0.05, adj.P.Val >= 0.1) %>%
    arrange(desc(abs(logFC))) %>%
    head(max(0L, 30L - n_fdr_sex))
)

sex_bar_lookup <- build_label_lookup(top_sex$Name)
top_sex <- top_sex %>% mutate(Name_display = truncate_labels(Name))

sex_bar <- ggplot(top_sex, aes(
  x = reorder(Name, logFC), y = logFC,
  fill = Direction
)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(
    aes(
      label = Name_display,
      y     = logFC * 0.98,
      hjust = ifelse(logFC > 0, 1, 0)
    ),
    size = 2.8, color = "white"
  ) +
  geom_text(
    aes(
      label = FDR_label,
      hjust = ifelse(logFC > 0, -0.2, 1.2)
    ),
    size = 4, fontface = "bold"
  ) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Higher in Males" = "#571C90",
    "Higher in Females" = "#FFC20A"
  )) +
  theme_minimal() +
  labs(
    x       = NULL,
    y       = "Log2 Fold Change (Male/Female)",
    title   = "Top Sex-Differential Metabolites (limma, p < 0.05)",
    caption = "* FDR < 0.1   ** FDR < 0.01   *** FDR < 0.001"
  ) +
  theme(
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title   = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0)
  )

save_dual_format(sex_bar, "sex_limma_bar",
  png_path = png_path, svg_path = svg_path,
  label_lookup = sex_bar_lookup
)
print(sex_bar)

write_csv(sex_limma_results, file.path(csv_path, "sex_limma_results.csv"))
```
# Interaction limma
```{r interaction_limma}
# limma interaction model: Sex * GW

metadata_int <- pData(metabolomics)
peak_matrix_int <- exprs(metabolomics)

metadata_int$Sex <- factor(metadata_int$Sex, levels = c("F", "M"))

design_int <- model.matrix(~ Sex * GW, data = metadata_int)
cat("Design matrix columns:\n")
print(colnames(design_int))

fit_int <- lmFit(peak_matrix_int, design_int)
fit_int <- eBayes(fit_int)

# --- Extract all three coefficients ---

# 1. Sex main effect (adjusted for GW)
sex_adj_results <- topTable(fit_int, coef = "SexM", number = Inf,
                             adjust.method = "BH") %>%
  rownames_to_column("Name") %>%
  left_join(fData(metabolomics) %>%
              dplyr::select(Name, Compound.ID, Super.Class, Class),
            by = "Name") %>%
  mutate(
    Direction = ifelse(logFC > 0, "Higher in Males", "Higher in Females"),
    FDR_label = case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01  ~ "**",
      adj.P.Val < 0.1  ~ "*",
      TRUE              ~ ""
    )
  )

cat("\nSex effect (adjusted for GW):\n")
cat("  Nominal p < 0.05:", sum(sex_adj_results$P.Value < 0.05), "\n")
cat("  FDR < 0.1:", sum(sex_adj_results$adj.P.Val < 0.1), "\n")

# 2. GW main effect (adjusted for Sex)
gw_adj_results <- topTable(fit_int, coef = "GW", number = Inf,
                            adjust.method = "BH") %>%
  rownames_to_column("Name") %>%
  left_join(fData(metabolomics) %>%
              dplyr::select(Name, Compound.ID, Super.Class, Class),
            by = "Name") %>%
  mutate(
    Direction = ifelse(logFC > 0, "Increases with GW", "Decreases with GW"),
    FDR_label = case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01  ~ "**",
      adj.P.Val < 0.1  ~ "*",
      TRUE              ~ ""
    )
  )

cat("\nGW effect (adjusted for Sex):\n")
cat("  Nominal p < 0.05:", sum(gw_adj_results$P.Value < 0.05), "\n")
cat("  FDR < 0.1:", sum(gw_adj_results$adj.P.Val < 0.1), "\n")

# 3. Sex:GW interaction
int_results <- topTable(fit_int, coef = "SexM:GW", number = Inf,
                         adjust.method = "BH") %>%
  rownames_to_column("Name") %>%
  left_join(fData(metabolomics) %>%
              dplyr::select(Name, Compound.ID, Super.Class, Class),
            by = "Name") %>%
  mutate(
    Direction = ifelse(logFC > 0, "Steeper in Males", "Steeper in Females"),
    FDR_label = case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01  ~ "**",
      adj.P.Val < 0.1  ~ "*",
      TRUE              ~ ""
    )
  )

cat("\nSex:GW interaction:\n")
cat("  Nominal p < 0.05:", sum(int_results$P.Value < 0.05), "\n")
cat("  FDR < 0.1:", sum(int_results$adj.P.Val < 0.1), "\n")

# --- Plots ---

plot_limma_bar <- function(results, title, ylab, fill_values, filename,
                           fdr_cut = 0.1, n = 30L) {
  n_fdr <- sum(results$adj.P.Val < fdr_cut, na.rm = TRUE)
  top <- bind_rows(
    results %>% filter(adj.P.Val < fdr_cut) %>% arrange(desc(abs(logFC))),
    results %>% filter(P.Value < 0.05, adj.P.Val >= fdr_cut) %>%
      arrange(desc(abs(logFC))) %>% head(max(0L, n - n_fdr))
  ) %>%
    mutate(
      FDR_label = case_when(
        adj.P.Val < 0.001   ~ "***",
        adj.P.Val < 0.01    ~ "**",
        adj.P.Val < fdr_cut ~ "*",
        TRUE                ~ ""
      )
    )

  if (nrow(top) == 0) {
    cat("No nominal hits for:", title, "\n")
    return(invisible(NULL))
  }

  label_lookup <- build_label_lookup(top$Name)
  top <- top %>% mutate(Name_display = truncate_labels(Name))

  p <- ggplot(top, aes(x = reorder(Name, logFC), y = logFC, fill = Direction)) +
    geom_bar(stat = "identity", alpha = 0.8) +
    geom_text(
      aes(
        label = Name_display,
        y     = logFC * 0.98,
        hjust = ifelse(logFC > 0, 1, 0)
      ),
      size = 2.8, color = "white"
    ) +
    geom_text(
      aes(
        label = FDR_label,
        hjust = ifelse(logFC > 0, -0.2, 1.2)
      ),
      size = 4, fontface = "bold"
    ) +
    coord_flip() +
    scale_fill_manual(values = fill_values) +
    theme_minimal() +
    labs(
      x       = NULL,
      y       = ylab,
      title   = title,
      caption = paste0("* FDR < ", fdr_cut, "   ** FDR < 0.01   *** FDR < 0.001")
    ) +
    theme(
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
      plot.title   = element_text(hjust = 0.5),
      plot.caption = element_text(hjust = 0)
    )

  save_dual_format(p, filename, png_path = png_path, svg_path = svg_path,
                   label_lookup = label_lookup)
  p
}

# Sex adjusted bar
p_sex_adj <- plot_limma_bar(
  sex_adj_results,
  title       = "Sex Effect Adjusted for GW (limma, p < 0.05)",
  ylab        = "Log2 Fold Change (Male/Female, GW-adjusted)",
  fill_values = c("Higher in Males" = "#571C90", "Higher in Females" = "#FFC20A"),
  filename    = "sex_adj_limma_bar"
)
if (!is.null(p_sex_adj)) print(p_sex_adj)

# GW adjusted bar
p_gw_adj <- plot_limma_bar(
  gw_adj_results,
  title       = "GW Effect Adjusted for Sex (limma, p < 0.05)",
  ylab        = "Slope (log2 peak area per GW week, Sex-adjusted)",
  fill_values = c("Increases with GW" = "#1B5E20", "Decreases with GW" = "#C8E6C9"),
  filename    = "gw_adj_limma_bar"
)
if (!is.null(p_gw_adj)) print(p_gw_adj)

# Interaction bar
p_int <- plot_limma_bar(
  int_results,
  title       = "Sex:GW Interaction (limma, p < 0.05)",
  ylab        = "Interaction coefficient (difference in GW slope by sex)",
  fill_values = c("Steeper in Males" = "#1A5276", "Steeper in Females" = "#AED6F1"),
  filename    = "interaction_limma_bar"
)
if (!is.null(p_int)) print(p_int)

# Volcano for sex adjusted (discrete comparison, volcano appropriate)
volcano_sex_adj <- ggplot(sex_adj_results,
                           aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = adj.P.Val < 0.1), alpha = 0.6) +
  scale_color_manual(values = c("FALSE" = "grey", "TRUE" = "red"),
                     name = "FDR < 0.1") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_minimal() +
  labs(x = "Log2 Fold Change (Male/Female)",
       y = "-log10(p-value)",
       title = "Sex Effect Adjusted for GW - Volcano Plot",
       caption = "Positive logFC = higher in males") +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(volcano_sex_adj, "sex_adj_limma_volcano",
                 png_path = png_path, svg_path = svg_path)
print(volcano_sex_adj)

# Ranked dot plot for GW adjusted (continuous, dot plot appropriate)
gw_adj_dot <- ggplot(gw_adj_results %>%
                       arrange(logFC) %>%
                       mutate(rank = row_number()),
                     aes(x = rank, y = logFC, color = adj.P.Val < 0.1)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red"),
                     name = "FDR < 0.1") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  theme_minimal() +
  labs(x = "Metabolite rank",
       y = "Slope (log2 peak area per GW week, Sex-adjusted)",
       title = "GW Effect Sizes Adjusted for Sex - All Metabolites") +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(gw_adj_dot, "gw_adj_limma_dot",
                 png_path = png_path, svg_path = svg_path)
print(gw_adj_dot)

# --- Save results ---
write_csv(sex_adj_results, file.path(csv_path, "sex_adjusted_limma_results.csv"))
write_csv(gw_adj_results,  file.path(csv_path, "gw_adjusted_limma_results.csv"))
write_csv(int_results,     file.path(csv_path, "interaction_limma_results.csv"))
```

## limma trajectory plots
```{r trajectory_plots}
# Trajectory plots for top GW and sex limma hits
# Points only, colored by sex, x = GW (continuous)
metadata_traj <- pData(metabolomics)

# --- GW trajectory plots ---
# Note: FDR-significant GW trajectories are plotted in the gw_limma chunk above.
# Top 20 nominal only (not FDR significant)
gw_nominal_hits <- gw_limma_results %>%
  filter(P.Value < 0.05 & adj.P.Val >= 0.1) %>%
  arrange(desc(abs(logFC))) %>%
  head(20) %>%
  mutate(label = paste0(
    "p=", formatC(P.Value, format = "e", digits = 2),
    "  slope=", round(logFC, 3)
  ))

# Plot GW nominal hits
if (nrow(gw_nominal_hits) > 0) {
  gw_nominal_plots <- build_trajectory_plots(
    metabolite_names = gw_nominal_hits$Name,
    eset             = metabolomics,
    title_suffixes   = paste0("GW nominal only | ", gw_nominal_hits$label)
  )
  save_trajectory_grids(gw_nominal_plots, "gw_trajectory_nominal", png_path, svg_path)
}

# --- Sex trajectory plots ---
# Top 20 FDR-significant by effect size
sex_fdr_hits <- sex_limma_results %>%
  filter(adj.P.Val < 0.1) %>%
  arrange(desc(abs(logFC))) %>%
  head(20) %>%
  mutate(label = paste0(
    "FDR=", formatC(adj.P.Val, format = "e", digits = 2),
    "  logFC=", round(logFC, 3)
  ))

# Top 20 nominal only
sex_nominal_hits <- sex_limma_results %>%
  filter(P.Value < 0.05 & adj.P.Val >= 0.1) %>%
  arrange(desc(abs(logFC))) %>%
  head(20) %>%
  mutate(label = paste0(
    "p=", formatC(P.Value, format = "e", digits = 2),
    "  logFC=", round(logFC, 3)
  ))

# Plot sex FDR hits
if (nrow(sex_fdr_hits) > 0) {
  sex_fdr_plots <- build_trajectory_plots(
    metabolite_names = sex_fdr_hits$Name,
    eset             = metabolomics,
    title_suffixes   = paste0("Sex FDR significant | ", sex_fdr_hits$label)
  )
  save_trajectory_grids(sex_fdr_plots, "sex_trajectory_fdr", png_path, svg_path)
} else {
  cat("No FDR-significant sex hits to plot\n")
}

# Plot sex nominal hits
if (nrow(sex_nominal_hits) > 0) {
  sex_nominal_plots <- build_trajectory_plots(
    metabolite_names = sex_nominal_hits$Name,
    eset             = metabolomics,
    title_suffixes   = paste0("Sex nominal only | ", sex_nominal_hits$label)
  )
  save_trajectory_grids(sex_nominal_plots, "sex_trajectory_nominal", png_path, svg_path)
}
```

# Sex PLS-DA
```{r sex_plsda}
# PLS-DA using ropls (predI = safe_ncomp, no orthoI = standard PLS-DA)
plsda_matrix <- t(exprs(metabolomics)) # samples × features
sex_labels <- factor(pData(metabolomics)$Sex)

# Impute for PLS-DA only 
plsda_matrix_imputed <- apply(plsda_matrix, 2, min_value_impute)

# Safe ncomp based on smallest group size
min_group_size <- min(table(sex_labels))
safe_ncomp <- min(5, min_group_size - 1)
cat("Group sizes:", table(sex_labels), "\n")
cat("Using ncomp =", safe_ncomp, "\n")

set.seed(42)
plsda_fit <- opls(
  x         = plsda_matrix_imputed,
  y         = sex_labels,
  predI     = safe_ncomp,
  orthoI    = 0, # 0 orthogonal components = standard PLS-DA
  permI     = 0, # permutation in dedicated chunk
  fig.pdfC  = "none",
  info.txtC = "none"
)

# Score plot (comp 1 vs comp 2)
plsda_scores <- data.frame(
  Comp1 = plsda_fit@scoreMN[, 1],
  Comp2 = plsda_fit@scoreMN[, 2],
  Sex   = sex_labels
)

score_plot <- ggplot(plsda_scores, aes(x = Comp1, y = Comp2, color = Sex)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("F" = "#FFC20A", "M" = "#571C90"), name = "Sex") +
  theme_minimal() +
  labs(
    x = "Component 1", y = "Component 2",
    title = "PLS-DA Score Plot - Sex"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(score_plot, "sex_plsda_scores", png_path = png_path, svg_path = svg_path)
print(score_plot)

# VIP scores
vip_df <- data.frame(
  Name = colnames(plsda_matrix_imputed),
  VIP_comp1 = as.numeric(getVipVn(plsda_fit))
) %>%
  arrange(desc(VIP_comp1)) %>%
  left_join(
    fData(metabolomics) %>%
      dplyr::select(Name, Compound.ID, Super.Class, Class),
    by = "Name"
  )

vip_plot_data <- vip_df %>% head(30)
vip_lookup <- build_label_lookup(vip_plot_data$Name)
vip_plot_data <- vip_plot_data %>% mutate(Name_display = truncate_labels(Name))

vip_plot <- ggplot(
  vip_plot_data,
  aes(x = reorder(Name_display, VIP_comp1), y = VIP_comp1)
) +
  geom_bar(stat = "identity", fill = "#571C90", alpha = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "Metabolite", y = "VIP Score (Component 1)",
    title = "PLS-DA VIP Scores for Sex Classification (Top 30)",
    caption = "Red dashed line = VIP > 1"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(vip_plot, "sex_plsda_vip",
  png_path = png_path, svg_path = svg_path,
  label_lookup = vip_lookup
)
print(vip_plot)

write_csv(vip_df, file.path(csv_path, "sex_plsda_vip.csv"))
cat("Metabolites with VIP > 1:", sum(vip_df$VIP_comp1 > 1), "\n")
```
## Permutation
```{r plsda_permutation}
# PLS-DA permutation test via ropls
set.seed(42)
plsda_perm <- opls(
  x         = plsda_matrix_imputed,
  y         = sex_labels,
  predI     = safe_ncomp,
  orthoI    = 0,
  permI     = 1000,
  fig.pdfC  = "none",
  info.txtC = "none"
)

cat("\nPLS-DA permutation test results (1000 permutations):\n")
print(plsda_perm@summaryDF)

# Plot permuted R2Y and Q2 distributions vs observed
perm_data <- data.frame(
  R2Y = plsda_perm@suppLs$permMN[, "R2Y(cum)"],
  Q2  = plsda_perm@suppLs$permMN[, "Q2(cum)"]
)

observed_R2Y <- plsda_fit@summaryDF[1, "R2Y(cum)"]
observed_Q2 <- plsda_fit@summaryDF[1, "Q2(cum)"]

perm_plot <- ggplot(perm_data) +
  geom_histogram(aes(x = R2Y, fill = "R2Y"), alpha = 0.5, bins = 20) +
  geom_histogram(aes(x = Q2, fill = "Q2"), alpha = 0.5, bins = 20) +
  geom_vline(
    xintercept = observed_R2Y, color = "#571C90", linewidth = 1,
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = observed_Q2, color = "#1B5E20", linewidth = 1,
    linetype = "dashed"
  ) +
  scale_fill_manual(
    values = c("R2Y" = "#571C90", "Q2" = "#1B5E20"),
    name = "Metric"
  ) +
  theme_minimal() +
  labs(
    x = "Permuted value", y = "Count",
    title = "PLS-DA Permutation Test (1000 permutations)",
    caption = "Dashed lines = observed model values; histograms = permuted null distribution"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(perm_plot, "sex_plsda_permutation", png_path = png_path, svg_path = svg_path)
print(perm_plot)
```

```{r upset_overlap}
# --- Sex analyses overlap ---
# Collect significant hits from each sex method
sex_wilcoxon_fdr <- metabolite_sex_stats %>%
  filter(FDR < 0.1) %>%
  pull(Name)
sex_wilcoxon_nominal <- metabolite_sex_stats %>%
  filter(p_value < 0.05 & FDR >= 0.1) %>%
  pull(Name)
sex_limma_fdr <- sex_limma_results %>%
  filter(adj.P.Val < 0.1) %>%
  pull(Name)
sex_limma_nominal <- sex_limma_results %>%
  filter(P.Value < 0.05 & adj.P.Val >= 0.1) %>%
  pull(Name)
sex_plsda_vip <- vip_df %>%
  filter(VIP_comp1 > 1) %>%
  pull(Name)
all_sex_names <- unique(c(
  sex_wilcoxon_fdr, sex_wilcoxon_nominal,
  sex_limma_fdr, sex_limma_nominal,
  sex_plsda_vip
))

sex_upset_df <- data.frame(
  Name                 = all_sex_names,
  Wilcoxon_FDR         = as.integer(all_sex_names %in% sex_wilcoxon_fdr),
  Wilcoxon_nominal     = as.integer(all_sex_names %in% sex_wilcoxon_nominal),
  limma_FDR            = as.integer(all_sex_names %in% sex_limma_fdr),
  limma_nominal        = as.integer(all_sex_names %in% sex_limma_nominal),
  PLSDA_VIP            = as.integer(all_sex_names %in% sex_plsda_vip)
)

png(file.path(png_path, "sex_upset_overlap.png"),
  width = 12, height = 7, units = "in", res = 300
)
upset(
  sex_upset_df %>% dplyr::select(-Name),
  order.by = "freq",
  decreasing = TRUE,
  main.bar.color = "grey30",
  text.scale = 1.3,
  mainbar.y.label = "Metabolites in Intersection",
  sets.x.label = "Metabolites per Method"
)
dev.off()

# Print to Rmd output as well
upset(
  sex_upset_df %>% dplyr::select(-Name),
  order.by        = "freq",
  decreasing      = TRUE,
  main.bar.color  = "grey30",
  text.scale      = 1.3,
  mainbar.y.label = "Metabolites in Intersection",
  sets.x.label    = "Metabolites per Method"
)

# Save intersection membership for downstream use
write_csv(sex_upset_df, file.path(csv_path, "sex_upset_membership.csv"))

# Identify metabolites significant in ALL FDR methods
sex_all_fdr <- sex_upset_df %>%
  filter(Wilcoxon_FDR == 1 & limma_FDR == 1) %>%
  pull(Name)
cat(
  "Metabolites FDR-significant in both Wilcoxon and limma (sex):",
  length(sex_all_fdr), "\n"
)
if (length(sex_all_fdr) > 0) print(sex_all_fdr)

# --- GW analyses overlap ---
gw_corr_fdr <- metabolite_samplemetadata_stats %>%
  filter(GW_FDR < 0.1) %>%
  pull(Name)
gw_corr_nominal <- metabolite_samplemetadata_stats %>%
  filter(GW_corr_pval.p_value < 0.05 & GW_FDR >= 0.1) %>%
  pull(Name)
gw_limma_fdr <- gw_limma_results %>%
  filter(adj.P.Val < 0.1) %>%
  pull(Name)
gw_limma_nominal <- gw_limma_results %>%
  filter(P.Value < 0.05 & adj.P.Val >= 0.1) %>%
  pull(Name)

all_gw_names <- unique(c(
  gw_corr_fdr, gw_corr_nominal,
  gw_limma_fdr, gw_limma_nominal
))

gw_upset_df <- data.frame(
  Name = all_gw_names,
  Correlation_FDR = as.integer(all_gw_names %in% gw_corr_fdr),
  Correlation_nominal = as.integer(all_gw_names %in% gw_corr_nominal),
  limma_FDR = as.integer(all_gw_names %in% gw_limma_fdr),
  limma_nominal = as.integer(all_gw_names %in% gw_limma_nominal)
)

png(file.path(png_path, "gw_upset_overlap.png"),
  width = 10, height = 6, units = "in", res = 300
)
upset(
  gw_upset_df %>% dplyr::select(-Name),
  order.by        = "freq",
  decreasing      = TRUE,
  main.bar.color  = "grey30",
  text.scale      = 1.3,
  mainbar.y.label = "Metabolites in Intersection",
  sets.x.label    = "Metabolites per Method"
)
dev.off()

upset(
  gw_upset_df %>% dplyr::select(-Name),
  order.by        = "freq",
  decreasing      = TRUE,
  main.bar.color  = "grey30",
  text.scale      = 1.3,
  mainbar.y.label = "Metabolites in Intersection",
  sets.x.label    = "Metabolites per Method"
)

write_csv(gw_upset_df, file.path(csv_path, "gw_upset_membership.csv"))

gw_all_fdr <- gw_upset_df %>%
  filter(Correlation_FDR == 1 & limma_FDR == 1) %>%
  pull(Name)
cat(
  "Metabolites FDR-significant in both correlation and limma (GW):",
  length(gw_all_fdr), "\n"
)
if (length(gw_all_fdr) > 0) print(gw_all_fdr)
```

# GW PLS-DA
```{r gw_plsda}
# Add GW_bucket to phenoData if not present
if (!"GW_bucket" %in% colnames(pData(metabolomics))) {
  pData(metabolomics)$GW_bucket <- cut(
    pData(metabolomics)$GW,
    breaks = c(15, 18, 21, 24),
    labels = c("16-18", "19-21", "22-24"),
    include.lowest = TRUE
  )
}

# PLS-DA for GW buckets (3-level categorical: 16-18, 19-21, 22-24)
# Uses same imputed matrix as sex PLS-DA - local copy only
# GW bucket labels pulled from sample metadata
gw_labels <- factor(pData(metabolomics)$GW_bucket,
  levels = c("16-18", "19-21", "22-24")
)

cat("GW bucket distribution:\n")
print(table(gw_labels))

# ncomp ceiling: number of groups - 1 = 2 for 3 GW buckets
min_gw_group <- min(table(gw_labels))
safe_ncomp_gw <- min(2, min_gw_group - 1)
cat("Using ncomp =", safe_ncomp_gw, "\n")

set.seed(42)
gw_plsda_fit <- opls(
  x         = plsda_matrix_imputed,
  y         = gw_labels,
  predI     = safe_ncomp_gw,
  orthoI    = 0,
  permI     = 0,
  fig.pdfC  = "none",
  info.txtC = "none"
)

cat("\nGW PLS-DA model summary:\n")
print(gw_plsda_fit@summaryDF)

# Score plot (comp 1 vs comp 2)
gw_plsda_scores <- data.frame(
  Comp1 = gw_plsda_fit@scoreMN[, 1],
  Comp2 = gw_plsda_fit@scoreMN[, 2],
  GW_bucket = gw_labels
)

gw_score_plot <- ggplot(
  gw_plsda_scores,
  aes(x = Comp1, y = Comp2, color = GW_bucket)
) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("16-18" = "#C8E6C9", "19-21" = "#66BB6A", "22-24" = "#1B5E20"), name = "GW Bucket") +
  theme_minimal() +
  labs(
    x = "Component 1", y = "Component 2",
    title = "PLS-DA Score Plot - Gestational Week Buckets"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(gw_score_plot, "gw_plsda_scores",
  png_path = png_path, svg_path = svg_path
)
print(gw_score_plot)

# VIP scores
gw_vip_df <- data.frame(
  Name      = colnames(plsda_matrix_imputed),
  VIP_comp1 = as.numeric(getVipVn(gw_plsda_fit))
) %>%
  arrange(desc(VIP_comp1)) %>%
  left_join(
    fData(metabolomics) %>%
      dplyr::select(Name, Compound.ID, Super.Class, Class),
    by = "Name"
  )

gw_vip_plot <- ggplot(
  gw_vip_df %>% head(30),
  aes(x = reorder(Name, VIP_comp1), y = VIP_comp1)
) +
  geom_bar(stat = "identity", fill = "#1B5E20", alpha = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "Metabolite", y = "VIP Score (Component 1)",
    title = "PLS-DA VIP Scores for GW Classification (Top 30)",
    caption = "Red dashed line = VIP > 1"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(gw_vip_plot, "gw_plsda_vip",
  png_path = png_path, svg_path = svg_path
)
print(gw_vip_plot)

write_csv(gw_vip_df, file.path(csv_path, "gw_plsda_vip.csv"))
cat("Metabolites with VIP > 1:", sum(gw_vip_df$VIP_comp1 > 1), "\n")
```
## Permutation
```{r gw_plsda_permutation}
# GW PLS-DA permutation test via ropls
set.seed(42)
gw_plsda_perm <- opls(
  x         = plsda_matrix_imputed,
  y         = gw_labels,
  predI     = safe_ncomp_gw,
  orthoI    = 0,
  permI     = 1000,
  fig.pdfC  = "none",
  info.txtC = "none"
)

cat("\nGW PLS-DA permutation test results (1000 permutations):\n")
print(gw_plsda_perm@summaryDF)

perm_data_gw <- data.frame(
  R2Y = gw_plsda_perm@suppLs$permMN[, "R2Y(cum)"],
  Q2  = gw_plsda_perm@suppLs$permMN[, "Q2(cum)"]
)

observed_R2Y_gw <- gw_plsda_fit@summaryDF[1, "R2Y(cum)"]
observed_Q2_gw <- gw_plsda_fit@summaryDF[1, "Q2(cum)"]

gw_perm_plot <- ggplot(perm_data_gw) +
  geom_histogram(aes(x = R2Y, fill = "R2Y"), alpha = 0.5, bins = 20) +
  geom_histogram(aes(x = Q2, fill = "Q2"), alpha = 0.5, bins = 20) +
  geom_vline(
    xintercept = observed_R2Y_gw, color = "#571C90",
    linewidth = 1, linetype = "dashed"
  ) +
  geom_vline(
    xintercept = observed_Q2_gw, color = "#1B5E20",
    linewidth = 1, linetype = "dashed"
  ) +
  scale_fill_manual(
    values = c("R2Y" = "#571C90", "Q2" = "#1B5E20"),
    name = "Metric"
  ) +
  theme_minimal() +
  labs(
    x = "Permuted value", y = "Count",
    title = "GW PLS-DA Permutation Test (1000 permutations)",
    caption = "Dashed lines = observed model values; histograms = permuted null distribution"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(gw_perm_plot, "gw_plsda_permutation",
  png_path = png_path, svg_path = svg_path
)
print(gw_perm_plot)
```

# AI Assistance Disclosure

Code in this script was developed with assistance from Claude (Anthropic). All
AI-generated code was reviewed, validated, and adapted by the author.

# Session Info

```{r session-info}
sessionInfo()
```
