---
title: "Metabolomics Visualization"
author: "Sneha M. Chaturvedi"
date: "Last Updated December 9, 2025"
output: 
  html_document:
    toc: TRUE
---

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
 
Packages <- c("tidyverse", "readxl", "stats", "ggfortify", "ggrepel", "svglite")

lapply(Packages, library, character.only=TRUE)

#Functions
#function to convert to numeric
convert_to_numeric <- function(x) {
  x <- as.character(x)
  x[x == "N/A"] <- NA
  as.numeric(x)}

# Imputation function to replace NaN with 1/5th of the minimum value in each row
min_value_impute <- function(column) {
  min_value <- min(column, na.rm = TRUE)
  column[is.na(column)] <- min_value / 5
  return(column)
}

# Function to calculate percentage of samples each peak is present in
calculate_missingness <- function(data) {
  # Calculate proportion of NA values for each metabolite
  missing_prop <- rowMeans(is.na(data[,-1]))
  
  # Add missingness to metabolite names
  missing_df <- data.frame(
    metabolite = data[,1],
    missing_proportion = missing_prop
  )
  
  return(missing_df)
}

svg_path <- "../../results/untargeted/Batch 2"
```

## Load Dataset
```{r LoadDataset, echo=TRUE}
#Peak Area Values
full_dataset = read_excel('../../doc/untargeted/24_untargeted_developing_2025.xlsx',sheet="Area",range=NULL)

#Metabolite Annotations
compound_annotation = read_excel('../../doc/untargeted/24_untargeted_developing_2025.xlsx',sheet="Annotation", range=NULL)

#Sample Metadata
sample_metadata = read_excel('../../doc/untargeted/24_metadata.xlsx',sheet="metadata",range=NULL)
#Rename SampleName to Sample in metadata if needed
sample_metadata <- sample_metadata %>%
  rename(Sample = SampleName)
```

## Format Datafile
1. Remove asterisks from names
  MTAC Key: *indicates isomers detected
2. Make data set completely numeric
3. Remove rows that are NaN across all samples - esp. relevant if annotated with other groups not present in current analysis

```{r CleanUp, echo=TRUE}
# Clean the data
full_dataset <- full_dataset %>%
  # First clean names with asterisks
  mutate(Name = gsub("\\*$", "", Name)) %>%
  # Then convert numeric columns and filter complete NA rows
  mutate(across(4:ncol(.), convert_to_numeric)) %>%
  filter(if_any(4:ncol(.), ~!is.na(.)))

compound_annotation <- compound_annotation %>%
  # First clean names with asterisks
  mutate(Name = gsub("\\*$", "", Name))

#create complete numeric data set for functions by replacing N/A with NaN
full_dataset <- cbind(full_dataset[1:3], full_dataset[,(4:27)] %>% replace(is.na(.), NaN))

# Modify sample names for columns 4 to 27
new_col_names <- names(full_dataset)[4:27] %>%
  sub("Area\\. ", "Sample", .)

# Assign the new column names back to the data frame
names(full_dataset)[4:27] <- new_col_names
```

# Data Visualization and QC
```{r Distribution_Plots_One}

#Peak Intensity Histogram Across Samples
# Reshape the data
data_long <- full_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Merge metadata with peak data for stratified analyses
sample_metadata_filtered <- sample_metadata %>%
  filter(Sample %in% names(full_dataset)[4:27])

# Add metadata to long format data
data_long <- data_long %>%
  left_join(sample_metadata_filtered %>% select(Sample, Sex, GW), by = "Sample")

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Peaks")
ggsave("raw_intensity_histogram.svg", path = svg_path)

#Peak Intensity Violin Plot per Sample
# Plot the violin plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_violin(fill = "darkorange", color = "black") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("raw_intensity_violin.svg", path = svg_path)

# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(full_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Peaks per Sample",
       x = "Sample", y = "Peak Count")
ggsave("raw_peak_count.svg", path = svg_path)

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(full_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Peaks Across Samples",
       x = "Percentage of Samples Missing Peak",
       y = "Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
ggsave("raw_missing_histogram.svg", path = svg_path)

# MOST ABUNDANT METABOLITES
# Calculate mean intensity across all samples for each metabolite
metabolite_abundance <- full_dataset %>%
  mutate(mean_intensity = rowMeans(select(., starts_with("Sample")), na.rm = TRUE)) %>%
  select(Compound.ID, Name, Formula, mean_intensity) %>%
  arrange(desc(mean_intensity)) %>%
  # Create display name: use Name if available, otherwise ID
  mutate(display_name = ifelse(is.na(Name) | Name == "", Compound.ID, Name))

# Top 20 most abundant metabolites
top20_metabolites <- metabolite_abundance %>%
  slice_head(n = 20)

# Print table
print("Top 20 Most Abundant Metabolites:")
print(top20_metabolites %>% select(display_name, mean_intensity, Formula))

# Visualize top 20
ggplot(top20_metabolites, aes(x = reorder(display_name, mean_intensity), y = mean_intensity)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top 20 Most Abundant Metabolites",
       x = "Metabolite",
       y = "Mean Peak Intensity") +
  theme(axis.text.y = element_text(size = 8))
ggsave("top20_abundant_metabolites.svg", path = svg_path, width = 10, height = 8)

# Summary statistics
cat("\nSummary of metabolite annotations:\n")
cat("Total metabolites:", nrow(full_dataset), "\n")
cat("Named metabolites:", sum(!is.na(full_dataset$Name) & full_dataset$Name != ""), "\n")
cat("Unannotated (ID only):", sum(is.na(full_dataset$Name) | full_dataset$Name == ""), "\n")

# Export top metabolites table
write.csv(metabolite_abundance %>% select(Compound.ID, Name, Formula, mean_intensity, display_name), 
          file.path(svg_path, "metabolite_abundance_ranking.csv"), 
          row.names = FALSE)
```

## Sex Stratified
```{r sex-stratified-visualization, echo=TRUE}
# SEX-STRATIFIED VISUALIZATIONS
#Peak Intensity Histogram by Sex
ggplot(data_long, aes(x = log10, fill = Sex)) +
  geom_histogram(binwidth = 0.4, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(title = "Distribution of Peak Intensity by Sex",
       x = "Log10 Intensity",
       y = "Number of Peaks") +
  facet_wrap(~Sex, ncol = 1)
ggsave("raw_intensity_histogram_by_sex.svg", path = svg_path)

#Peak Intensity Violin Plot per Sample colored by Sex
ggplot(data_long, aes(x = as.factor(Sample), y = log10, fill = Sex)) +
  geom_violin(color = "black") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample by Sex",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("raw_intensity_violin_by_sex.svg", path = svg_path)

# Peak count by sex
peak_count_by_sex <- peak_count_full %>%
  left_join(sample_metadata_filtered %>% select(Sample, Sex), by = "Sample")

ggplot(data = peak_count_by_sex, aes(x = Sample, y = Count, fill = Sex)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Peaks per Sample by Sex",
       x = "Sample", y = "Peak Count")
ggsave("raw_peak_count_by_sex.svg", path = svg_path)

# GESTATIONAL WEEK RELATIONSHIPS
# Calculate mean intensity per sample
mean_intensity_per_sample <- data_long %>%
  group_by(Sample, Sex, GW) %>%
  summarise(mean_log10_intensity = mean(log10, na.rm = TRUE),
            .groups = "drop")

# Peak intensity vs GW
ggplot(mean_intensity_per_sample, aes(x = GW, y = mean_log10_intensity, color = Sex)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
  theme_minimal() +
  labs(title = "Mean Peak Intensity vs Gestational Week",
       x = "Gestational Week",
       y = "Mean Log10 Intensity")
ggsave("intensity_vs_GW.svg", path = svg_path)

# Peak count vs GW
peak_count_with_metadata <- peak_count_full %>%
  left_join(sample_metadata_filtered %>% select(Sample, Sex, GW), by = "Sample")

ggplot(peak_count_with_metadata, aes(x = GW, y = Count, color = Sex)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
  theme_minimal() +
  labs(title = "Peak Count vs Gestational Week",
       x = "Gestational Week",
       y = "Number of Distinct Peaks")
ggsave("peak_count_vs_GW.svg", path = svg_path)

#ABUNDANT METABOLITES
# Calculate mean intensity per metabolite by sex
metabolite_abundance_by_sex <- data_long %>%
  group_by(Compound.ID, Name, Sex) %>%
  summarise(mean_intensity = mean(Intensity, na.rm = TRUE),
            .groups = "drop") %>%
  # Merge with formula from full_dataset
  left_join(full_dataset %>% select(Compound.ID, Formula), by = "Compound.ID") %>%
  # Create display name
  mutate(display_name = ifelse(is.na(Name) | Name == "", Compound.ID, Name)) %>%
  arrange(Sex, desc(mean_intensity))

# Get top 20 for each sex
top20_by_sex <- metabolite_abundance_by_sex %>%
  group_by(Sex) %>%
  slice_head(n = 20) %>%
  ungroup()

# Visualize top 20 by sex (faceted)
ggplot(top20_by_sex, aes(x = reorder(display_name, mean_intensity), y = mean_intensity, fill = Sex)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~Sex, scales = "free_y") +
  theme_minimal() +
  labs(title = "Top 20 Most Abundant Metabolites by Sex",
       x = "Metabolite",
       y = "Mean Peak Intensity") +
  theme(axis.text.y = element_text(size = 7),
        strip.text = element_text(size = 12, face = "bold"))
ggsave("top20_abundant_metabolites_by_sex.svg", path = svg_path, width = 12, height = 10)

# Print tables for each sex
cat("\nTop 20 Most Abundant Metabolites - Female:\n")
print(top20_by_sex %>% filter(Sex == "F") %>% select(display_name, mean_intensity, Formula))

cat("\nTop 20 Most Abundant Metabolites - Male:\n")
print(top20_by_sex %>% filter(Sex == "M") %>% select(display_name, mean_intensity, Formula))

# Export sex-stratified rankings
write.csv(metabolite_abundance_by_sex %>% select(Compound.ID, Name, Formula, Sex, mean_intensity, display_name), 
          file.path(svg_path, "metabolite_abundance_ranking_by_sex.csv"), 
          row.names = FALSE)
```

## GW Buckets
```{r GW_buckets, echo=TRUE}
# Create GW buckets in metadata
sample_metadata_filtered <- sample_metadata_filtered %>%
  mutate(GW_bucket = case_when(
    GW >= 16 & GW <= 18 ~ "16-18",
    GW >= 19 & GW <= 21 ~ "19-21",
    GW >= 22 & GW <= 24 ~ "22-24",
    TRUE ~ NA_character_
  ))

# Add GW_bucket to long format data
data_long <- data_long %>%
  left_join(sample_metadata_filtered %>% select(Sample, GW_bucket), by = "Sample")

# Calculate mean intensity per metabolite by GW bucket
metabolite_abundance_by_GW <- data_long %>%
  filter(!is.na(GW_bucket)) %>%
  group_by(Compound.ID, Name, GW_bucket) %>%
  summarise(mean_intensity = mean(Intensity, na.rm = TRUE),
            .groups = "drop") %>%
  # Merge with formula from full_dataset
  left_join(full_dataset %>% select(Compound.ID, Formula), by = "Compound.ID") %>%
  # Create display name
  mutate(display_name = ifelse(is.na(Name) | Name == "", Compound.ID, Name)) %>%
  arrange(GW_bucket, desc(mean_intensity))

# Get top 20 for each GW bucket
top20_by_GW <- metabolite_abundance_by_GW %>%
  group_by(GW_bucket) %>%
  slice_head(n = 20) %>%
  ungroup()

# Visualize top 20 by GW bucket (faceted)
ggplot(top20_by_GW, aes(x = reorder(display_name, mean_intensity), y = mean_intensity, fill = GW_bucket)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~GW_bucket, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(title = "Top 20 Most Abundant Metabolites by Gestational Week",
       x = "Metabolite",
       y = "Mean Peak Intensity") +
  theme(axis.text.y = element_text(size = 6),
        strip.text = element_text(size = 11, face = "bold"),
        plot.margin = margin(10, 10, 10, 10))
ggsave("top20_abundant_metabolites_by_GW.svg", path = svg_path, width = 18, height = 10)

# Print tables for each GW bucket
cat("\nTop 20 Most Abundant Metabolites - GW 16-18:\n")
print(top20_by_GW %>% filter(GW_bucket == "16-18") %>% select(display_name, mean_intensity, Formula))

cat("\nTop 20 Most Abundant Metabolites - GW 19-21:\n")
print(top20_by_GW %>% filter(GW_bucket == "19-21") %>% select(display_name, mean_intensity, Formula))

cat("\nTop 20 Most Abundant Metabolites - GW 22-24:\n")
print(top20_by_GW %>% filter(GW_bucket == "22-24") %>% select(display_name, mean_intensity, Formula))

# Export GW-stratified rankings
write.csv(metabolite_abundance_by_GW %>% select(Compound.ID, Name, Formula, GW_bucket, mean_intensity, display_name), 
          file.path(svg_path, "metabolite_abundance_ranking_by_GW.csv"), 
          row.names = FALSE)

# Summary of samples per GW bucket
cat("\nSamples per GW bucket:\n")
print(sample_metadata_filtered %>% 
        count(GW_bucket) %>% 
        arrange(GW_bucket))
```

## Initial PCA
```{r PCA_One, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- full_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Merge sample scores with metadata
sample_scores <- sample_scores %>%
  left_join(sample_metadata, by = "Sample")

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained by Samples") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("raw_sample_elbow.svg", path = svg_path)

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]), title = "PCA of Sample Variation")
ggsave("raw_sample_pca.svg", path = svg_path)

# Sex PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("raw_sample_pca1x2_sex.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("raw_sample_pca2x3_sex.svg", path = svg_path)

# GW PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("raw_sample_pca1x2_gw.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("raw_sample_pca2x3_gw.svg", path = svg_path)

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- full_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- full_dataset$Compound.ID

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained by Metabolites") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("raw_metabolite_elbow.svg", path = svg_path)

# Calculate distance from origin for each metabolite
metabolite_scores <- metabolite_scores %>%
  mutate(distance = sqrt(PC1^2 + PC2^2))

# Define a threshold for labeling outliers.
# This threshold can be adjusted based on the spread of your data.
threshold <- quantile(metabolite_scores$distance, 0.99)

# Create a logical column to indicate if the metabolite is an outlier
metabolite_scores <- metabolite_scores %>%
  mutate(is_outlier = distance > threshold)

# Plot PCA for metabolites with outlier labels
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(aes(label = ifelse(is_outlier, Compound.ID, "")), vjust = -1, hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
       title = "PCA of Metabolite Variation") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("raw_metabolite_pca1x2.svg", path = svg_path)

ggplot(metabolite_scores, aes(x = PC2, y = PC3)) +
  geom_point() +
  geom_text(aes(label = ifelse(is_outlier, Compound.ID, "")), vjust = -1, hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_metabolites[3]), 
       title = "PCA of Metabolite Variation") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("raw_metabolite_pca2x3.svg", path = svg_path)
```

# Outlier Removal
```{r OutlierSamples, echo=TRUE}
identify_outlier_samples <- function(data, threshold = 3, i = c(4:27)) {
  # Calculate total intensity per sample
  sample_sums <- colSums(data[,i], na.rm = TRUE)
  
  # Calculate z-scores for sample intensities
  z_scores <- scale(sample_sums)
  
  # Identify outliers (samples with z-scores above threshold)
  outliers <- which(abs(z_scores) > threshold)
  
  return(list(
    outlier_indices = outliers,
    sample_sums = sample_sums,
    z_scores = z_scores
  ))
}

# Run function and save output
outliers_info <- identify_outlier_samples(full_dataset)

# Convert outlier indices to actual column names
outlier_columns <- colnames(full_dataset)[outliers_info$outlier_indices + 3] # Adjusted for the correct indexing
print(outlier_columns)

# Remove the outliers from the data set
clean_dataset <- full_dataset %>% select(-all_of(outlier_columns))
```

## Outlier Visualization

```{r Distribution_Plots_Two, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- clean_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Peaks")
ggsave("outlier_intensity_histogram.svg", path = svg_path)

# Plot the violin plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_violin(fill = "darkorange", color = "black") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("outlier_intensity_violin.svg", path = svg_path)

# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(clean_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Peaks per Sample",
       x = "Sample", y = "Peak Count")
ggsave("outlier_peak_count.svg", path = svg_path)

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(clean_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Peaks Across Samples",
       x = "Percentage of Samples Missing Peak",
       y = "Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
ggsave("outlier_missing_histogram.svg", path = svg_path)  
```
## Outlier PCA
```{r PCA_Two, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- clean_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Merge sample scores with metadata
sample_scores <- sample_scores %>%
  left_join(sample_metadata, by = "Sample")

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained by Samples") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("outlier_sample_elbow.svg", path = svg_path)

# Sex PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("outlier_sample_pca1x2_sex.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("outlier_sample_pca2x3_sex.svg", path = svg_path)

# GW PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("outlier_sample_pca1x2_gw.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("outlier_sample_pca2x3_gw.svg", path = svg_path)

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- clean_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- clean_dataset$Compound.ID

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained by Metabolites") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("outlier_metabolite_elbow.svg", path = svg_path)

# Calculate distance from origin for each metabolite
metabolite_scores <- metabolite_scores %>%
  mutate(distance = sqrt(PC1^2 + PC2^2))

# Define a threshold for labeling outliers.
# This threshold can be adjusted based on the spread of your data.
threshold <- quantile(metabolite_scores$distance, 0.99)

# Create a logical column to indicate if the metabolite is an outlier
metabolite_scores <- metabolite_scores %>%
  mutate(is_outlier = distance > threshold)

# Plot PCA for metabolites with outlier labels
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(aes(label = ifelse(is_outlier, Compound.ID, "")), vjust = -1, hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
       title = "PCA of Metabolite Variation") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("outlier_metabolite_pca1x2.svg", path = svg_path)

ggplot(metabolite_scores, aes(x = PC2, y = PC3)) +
  geom_point() +
  geom_text(aes(label = ifelse(is_outlier, Compound.ID, "")), vjust = -1, hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_metabolites[3]), 
       title = "PCA of Metabolite Variation") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("outlier_metabolite_pca2x3.svg", path = svg_path)
```
  
# Data Split
```{r Dataframe_Split}
annotated_dataset <- clean_dataset[!is.na(clean_dataset$Name) & clean_dataset$Name != "", ]
unannotated_dataset <- clean_dataset[is.na(clean_dataset$Name) | clean_dataset$Name == "", ]
```

## Annotated Visualization
```{r Distribution_Plots_Annotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- annotated_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Annotated Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Annotated Peaks")
ggsave("annotated_intensity_histogram.svg", path = svg_path)

# Plot the violin plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_violin(fill = "darkorange", color = "black") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("annotated_intensity_violin.svg", path = svg_path)


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(annotated_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Annotated Peaks per Sample",
       x = "Sample", y = "Peak Count")
ggsave("annotated_peak_count.svg", path = svg_path)

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(annotated_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Annotated Peaks Across Samples",
       x = "Percentage of Samples Missing Annotated Peak",
       y = "Annotated Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
ggsave("annotated_missing_histogram.svg", path = svg_path)
  
```
## Annotated PCA
```{r PCA_Annotated, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- annotated_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Merge sample scores with metadata
sample_scores <- sample_scores %>%
  left_join(sample_metadata, by = "Sample")

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_sample_elbow.svg", path = svg_path)

# Sex PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_sample_pca1x2_sex.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_sample_pca2x3_sex.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC3, y = PC4, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_sample_pca3x4_sex.svg", path = svg_path)

# GW PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_sample_pca1x2_gw.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_sample_pca2x3_gw.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC3, y = PC4, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_sample_pca3x4_gw.svg", path = svg_path)

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- annotated_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Name <- annotated_dataset$Name

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_metabolite_elbow.svg", path = svg_path)

# Calculate distance from origin for each metabolite
metabolite_scores <- metabolite_scores %>%
  mutate(distance = sqrt(PC1^2 + PC2^2))

# Define a threshold for labeling outliers.
# This threshold can be adjusted based on the spread of your data.
threshold <- quantile(metabolite_scores$distance, 0.99)

# Create a logical column to indicate if the metabolite is an outlier
metabolite_scores <- metabolite_scores %>%
  mutate(is_outlier = distance > threshold)

# Plot PCA for metabolites with outlier labels
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(aes(label = ifelse(is_outlier, Name, "")), vjust = -1, hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
       title = "PCA of Metabolite Variation") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_metabolite_pca1x2.svg", path = svg_path)

ggplot(metabolite_scores, aes(x = PC2, y = PC3)) +
  geom_point() +
  geom_text(aes(label = ifelse(is_outlier, Name, "")), vjust = -1, hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_metabolites[3]), 
       title = "PCA of Metabolite Variation") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("annotated_metabolite_pca2x3.svg", path = svg_path)

```

##Unannotated Visualization
```{r Distribution_Plots_Unannotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- unannotated_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Annotated Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Annotated Peaks")
ggsave("unannotated_intensity_histogram.svg", path = svg_path)

# Plot the violin plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_violin(fill = "darkorange", color = "black") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("unannotated_intensity_violin.svg", path = svg_path)

# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(unannotated_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Unannotated Peaks per Sample",
       x = "Sample", y = "Peak Count")
ggsave("unannotated_peak_count.svg", path = svg_path)

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(unannotated_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Unannotated Peaks Across Samples",
       x = "Percentage of Samples Missing Unannotated Peak",
       y = "Unannotated Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
ggsave("unannotated_missing_histogram.svg", path = svg_path)  
```

## Unannotated PCA
```{r PCA_Unannotated, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- unannotated_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Merge sample scores with metadata
sample_scores <- sample_scores %>%
  left_join(sample_metadata, by = "Sample")

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_sample_elbow.svg", path = svg_path)

# Sex PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_sample_pca1x2_sex.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_sample_pca2x3_sex.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC3, y = PC4, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_sample_pca3x4_sex.svg", path = svg_path)

# GW PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_sample_pca1x2_gw.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_sample_pca2x3_gw.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC3, y = PC4, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_sample_pca3x4_gw.svg", path = svg_path)


# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- unannotated_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Name <- unannotated_dataset$Compound.ID

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_metabolite_elbow.svg", path = svg_path)

# Calculate distance from origin for each metabolite
metabolite_scores <- metabolite_scores %>%
  mutate(distance = sqrt(PC1^2 + PC2^2))

# Define a threshold for labeling outliers.
# This threshold can be adjusted based on the spread of your data.
threshold <- quantile(metabolite_scores$distance, 0.99)

# Create a logical column to indicate if the metabolite is an outlier
metabolite_scores <- metabolite_scores %>%
  mutate(is_outlier = distance > threshold)

# Plot PCA for metabolites with outlier labels
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(aes(label = ifelse(is_outlier, Name, "")), vjust = -1, hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
       title = "PCA of Metabolite Variation") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unannotated_metabolite_pca1x2.svg", path = svg_path)
```

# Duplicate Removal
## Load Annotation

```{r LoadAnnotation, echo=TRUE}
compound_annotation = read_excel('../../doc/untargeted/32_untargeted_developing_2025.xlsx',sheet="Annotation", range=NULL)

compound_annotation <- compound_annotation %>%
  # First clean names with asterisks
  mutate(Name = gsub("\\*$", "", Name))

# split file
annotated_compounds <- compound_annotation[!is.na(compound_annotation$InChiKey) & compound_annotation$InChiKey != "", ]
unannotated_compounds <- compound_annotation[is.na(compound_annotation$InChiKey) | compound_annotation$InChiKey == "", ]
```

## Identify Duplicates
```{r Duplicate_Identification}

# Find duplicate names in the column "InChiKey"
duplicates <- annotated_compounds %>%
  group_by(InChiKey) %>%
  filter(n() > 1) %>%
  arrange(Name)

# Calculate the count of occurrences for each duplicate name
duplicate_counts <- duplicates %>%
  group_by(InChiKey) %>%
  summarise(count = n())

# Create a histogram
ggplot(duplicate_counts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  ggtitle("Histogram of Duplicate InChiKey") +
  xlab("Count of Duplicates") +
  ylab("Frequency")
ggsave("duplicate_histogram.svg", path = svg_path)
```
## Compare Duplicate Types
```{r Compare_Duplicates, echo=TRUE}

# Calculate for each row - RSD, STD Dev, highest intensity, median intensity, mean value
annotated_dataset <- annotated_dataset %>%
  rowwise() %>%
  mutate(
    mean_value = mean(c_across(4:27), na.rm = TRUE),
    std_dev = sd(c_across(4:27), na.rm = TRUE),
    percent_RSD = (std_dev / mean_value) * 100,
    highest_intensity = max(c_across(4:27), na.rm = TRUE),
    median_intensity = median(c_across(4:27), na.rm = TRUE),
  ) %>%
  ungroup()

annotated_dataset <- annotated_dataset %>%
  left_join(annotated_compounds %>% select(Compound.ID, InChiKey, Super.Class, Class, Sub.Class, Pathways, CID, CHEBI, HMDB, KEGG, CAS, Molecular.Weight, Best.Match, Library.Name, Library.ID, 'Annot.DeltaMass[ppm]', 'm/z', 'RT[min]', Background, Reference.Ion, Smiles), by = "Compound.ID")

duplicates <- duplicates %>%
  left_join(annotated_dataset %>% select(Compound.ID, mean_value, std_dev, percent_RSD, highest_intensity, median_intensity), by = "Compound.ID")

# Add columns for liquid chromatography and ionization
duplicates <- duplicates %>%
  mutate(
    LC_column = str_extract(Compound.ID, "^[^\\-]+"),
    ionization = str_extract(Compound.ID, "(?<=\\-)[^\\-]+(?=\\-)")
  )

#Identify duplicates
duplicates  <- duplicates %>%
  group_by(InChiKey) %>%
  filter(n() > 1) %>%
  mutate(same_column = if_else(n_distinct(LC_column, ionization) > 1, 1, 0)) %>%
  ungroup()

write.csv(duplicates, "../../results/untargeted/duplicate_metabolites.csv", row.names = FALSE)

# Filter duplicates with same_column == 0
filtered_duplicates <- duplicates %>%
  filter(same_column == 0)

# Calculate the count of occurrences for each duplicate name
same_duplicate_counts <- filtered_duplicates %>%
  group_by(InChiKey) %>%
  summarise(count = n())

# Create a histogram
ggplot(same_duplicate_counts, aes(x = count)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  ggtitle("Histogram of Duplicate InChiKey on Same Column") +
  xlab("Count of Duplicates") +
  ylab("Frequency")
ggsave("same_column_duplicate_histogram.svg", path = svg_path)

# Determine the duplicates with the lowest % RSD, ensuring ties are handled by selecting the first occurrence
lowest_RSD_duplicates <- filtered_duplicates %>%
  group_by(InChiKey) %>%
  arrange(percent_RSD) %>%
  slice(1) %>%
  ungroup()

# Determine the duplicates with the highest metabolite intensity, ensuring ties are handled by selecting the first occurrence
highest_intensity_duplicates <- filtered_duplicates %>%
  group_by(InChiKey) %>%
  arrange(desc(highest_intensity)) %>%
  slice(1) %>%
  ungroup()

# Compare lowest % RSD and highest metabolite intensity to see if they overlap
overlapping_duplicates <- inner_join(lowest_RSD_duplicates, highest_intensity_duplicates, by = c("Compound.ID", "InChiKey", "Name"))


write.csv(lowest_RSD_duplicates, "../../results/untargeted/lowest_RSD_duplicates.csv", row.names = FALSE)
write.csv(highest_intensity_duplicates, "../../results/untargeted/highest_intensity_duplicates.csv", row.names = FALSE)
write.csv(overlapping_duplicates, "../../results/untargeted/overlapping_duplicates.csv", row.names = FALSE)

```

### Same Column
```{r TieredDuplicateRemoval, echo=TRUE}
#Tiered Duplicate Selection for Metabolites on Same Column
## First step: Resolve duplicates by highest median_intensity
first_step <- filtered_duplicates %>%
  group_by(InChiKey) %>%
  arrange(desc(median_intensity)) %>%
  slice(1) %>%
  ungroup()

resolved_first_step <- first_step %>%
  distinct(InChiKey) %>%
  nrow()

remaining_first_step <- filtered_duplicates %>%
  filter(!InChiKey %in% first_step$InChiKey) %>%
  distinct(InChiKey) %>%
  nrow()

print(paste("Resolved after first step (highest median_intensity):", resolved_first_step))
print(paste("Remaining duplicates after first step:", remaining_first_step))

# Second step: Resolve remaining duplicates by lowest % RSD
second_step <- filtered_duplicates %>%
  filter(!InChiKey %in% first_step$InChiKey) %>%
  group_by(InChiKey) %>%
  arrange(percent_RSD) %>%
  slice(1) %>%
  ungroup()

resolved_second_step <- second_step %>%
  distinct(InChiKey) %>%
  nrow()

remaining_second_step <- filtered_duplicates %>%
  filter(!InChiKey %in% c(first_step$InChiKey, second_step$InChiKey)) %>%
  distinct(InChiKey) %>%
  nrow()

print(paste("Resolved after second step (lowest % RSD):", resolved_second_step))
print(paste("Remaining duplicates after second step:", remaining_second_step))

# Combine resolved duplicates
resolved_duplicates <- bind_rows(first_step, second_step) %>%
  distinct(Compound.ID, .keep_all = TRUE)


# View the results
print("Resolved duplicates by tiers:")
print(resolved_duplicates)

# Optionally, save the results to CSV files
write.csv(resolved_duplicates, "../../results/untargeted/resolved_duplicates_by_tiers.csv", row.names = FALSE)

```

## Same Column Duplicate Removal
```{r UniqueDataframeCreation, echo = TRUE}
resolved_same_column_0 <- duplicates %>%
  filter(same_column == 0) %>%
  group_by(InChiKey) %>%
  arrange(desc(median_intensity)) %>%
  slice(1) %>%
  ungroup()

#Keep duplicates with same_column == 1
same_column_1 <- duplicates %>%
  filter(same_column == 1)

# Extract InChiKeys of all duplicates for filtering non-duplicates
duplicate_inchikeys <- unique(duplicates$InChiKey)

# Combine resolved duplicates (same_column == 0) and non-duplicates
non_duplicates <- annotated_dataset %>%
  filter(!InChiKey %in% duplicate_inchikeys)

almost_unique_dataset <- bind_rows(non_duplicates, resolved_same_column_0, same_column_1)

#save the final dataset to CSV files
write.csv(almost_unique_dataset, "../../results/untargeted/almost_unique_dataset.csv", row.names = FALSE)

```

## Different Column

```{r DifferentColumn, echo = TRUE}
# Filter duplicates with same_column == 1
different_column_duplicates <- duplicates %>%
  filter(same_column == 1)

# Calculate the count of occurrences for each duplicate name
different_duplicate_counts <- different_column_duplicates %>%
  group_by(InChiKey) %>%
  summarise(count = n())

# Create a histogram
ggplot(different_duplicate_counts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  ggtitle("Histogram of Duplicate InChiKey on Different Columns") +
  xlab("Count of Duplicates") +
  ylab("Frequency")
ggsave("different_column_duplicate_histogram.svg", path = svg_path)
```
## Different Column Duplicate Removal

```{r DifferentDuplicate, echo = TRUE}
#Tiered Duplicate Selection for Metabolites on Different Columns
## First step: Resolve duplicates by highest median_intensity
first_step <- different_column_duplicates %>%
  group_by(InChiKey) %>%
  arrange(desc(median_intensity)) %>%
  slice(1) %>%
  ungroup()

resolved_first_step <- first_step %>%
  distinct(InChiKey) %>%
  nrow()

remaining_first_step <- different_column_duplicates %>%
  filter(!InChiKey %in% first_step$InChiKey) %>%
  distinct(InChiKey) %>%
  nrow()

print(paste("Resolved after first step (highest median_intensity):", resolved_first_step))
print(paste("Remaining duplicates after first step:", remaining_first_step))

# Second step: Resolve remaining duplicates by lowest % RSD
second_step <- different_column_duplicates %>%
  filter(!InChiKey %in% first_step$InChiKey) %>%
  group_by(InChiKey) %>%
  arrange(percent_RSD) %>%
  slice(1) %>%
  ungroup()

resolved_second_step <- second_step %>%
  distinct(InChiKey) %>%
  nrow()

remaining_second_step <- different_column_duplicates %>%
  filter(!InChiKey %in% c(first_step$InChiKey, second_step$InChiKey)) %>%
  distinct(InChiKey) %>%
  nrow()

print(paste("Resolved after second step (lowest % RSD):", resolved_second_step))
print(paste("Remaining duplicates after second step:", remaining_second_step))

# Combine resolved duplicates
different_resolved_duplicates <- bind_rows(first_step, second_step) %>%
  distinct(Compound.ID, .keep_all = TRUE)


# View the results
print("Resolved duplicates by tiers:")
print(different_resolved_duplicates)

# Optionally, save the results to CSV files
write.csv(different_resolved_duplicates, "../../results/untargeted/different_resolved_duplicates_by_tiers.csv", row.names = FALSE)

#do it based on median intensity for now
resolved_different_column_0 <- duplicates %>%
  filter(same_column == 1) %>%
  group_by(InChiKey) %>%
  arrange(desc(median_intensity)) %>%
  slice(1) %>%
  ungroup()

# Extract InChiKeys of all duplicates for filtering non-duplicates
duplicate_inchikeys <- unique(duplicates$InChiKey)

# Combine resolved duplicates (same_column == 1) and non-duplicates
non_duplicates <- almost_unique_dataset %>%
  filter(!InChiKey %in% duplicate_inchikeys)

unique_dataset <- bind_rows(non_duplicates, resolved_different_column_0)

#save the final dataset to CSV files
write.csv(unique_dataset, "../../results/untargeted/almost_unique_dataset.csv", row.names = FALSE)
```
# Final Dataset

## Unique Visualization

```{r Distribution_Plots_Unique_Annotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- unique_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Unique Metabolite Intensity",
       x = "Log10 Intensity",
       y = "Number of Unique Metablites")
ggsave("unique_intensity_histogram.svg", path = svg_path)

# Plot the violin plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_violin(fill = "darkorange", color = "black") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("unique_intensity_violin.svg", path = svg_path)

# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(unique_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Unique Metabolites per Sample",
       x = "Sample", y = "Metabolite Count")
ggsave("unique_peak_count.svg", path = svg_path)

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(unique_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Metabolites Across Samples",
       x = "Percentage of Samples Missing Metabolite",
       y = "Unique Metabolite Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
ggsave("unique_missing_histogram.svg", path = svg_path)
  
```

## Unique PCA
```{r PCA_Unique_Annotated, echo = TRUE}

# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- unique_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

variable_columns <- apply(samples_matrix, 2, function(x) length(unique(x)) > 1)
samples_matrix <- samples_matrix[, variable_columns]

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Merge sample scores with metadata
sample_scores <- sample_scores %>%
  left_join(sample_metadata, by = "Sample")

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_sample_elbow.svg", path = svg_path)

# Sex PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_sample_pca1x2_sex.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_sample_pca2x3_sex.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC3, y = PC4, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_sample_pca3x4_sex.svg", path = svg_path)

# GW PCA
ggplot(sample_scores, aes(x = PC1, y = PC2, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_sample_pca1x2_gw.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC2, y = PC3, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_sample_pca2x3_gw.svg", path = svg_path)

ggplot(sample_scores, aes(x = PC3, y = PC4, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_sample_pca3x4_gw.svg", path = svg_path)

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- unique_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Name <- unique_dataset$Name

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_metabolite_elbow.svg", path = svg_path)

# Calculate distance from origin for each metabolite
metabolite_scores <- metabolite_scores %>%
  mutate(distance = sqrt(PC1^2 + PC2^2))

# Define a threshold for labeling outliers.
# This threshold can be adjusted based on the spread of your data.
threshold <- quantile(metabolite_scores$distance, 0.99)

# Create a logical column to indicate if the metabolite is an outlier
metabolite_scores <- metabolite_scores %>%
  mutate(is_outlier = distance > threshold)

# Plot PCA for metabolites with outlier labels
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(aes(label = ifelse(is_outlier, Name, "")), vjust = -1, hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
       title = "PCA of Metabolite Variation") +
  theme(plot.title = element_text(hjust = 0.5))
ggsave("unique_metabolite_pca1x2.svg", path = svg_path)

```

# Assign Metadata
```{r Metadata, echo=TRUE}
# Convert metabolites data to matrix format
final_matrix <- unique_dataset %>%
  select(starts_with("Sample"))
  
metabolite_names <- unique_dataset$InChiKey
final_matrix <- apply(final_matrix, 2, as.numeric)
rownames(final_matrix) <- metabolite_names

# Filter `annotated_compounds` to contain only rows from `unique_dataset`
unique_annotated_compounds <- annotated_compounds %>%
  filter(Compound.ID %in% unique_dataset$Compound.ID)

# Ensure final_matrix rows match metabolite metadata
metabolite_metadata_filtered <- unique_annotated_compounds %>%
  filter(InChiKey %in% rownames(final_matrix)) %>%
  arrange(match(InChiKey, rownames(final_matrix)))

# Create the metabolite metadata dataframe
metabolite_metadata_df <- as.data.frame(metabolite_metadata_filtered)
rownames(metabolite_metadata_df) <- metabolite_metadata_df$InChiKey

# Ensure sample_metadata matches the columns of final_matrix
sample_metadata_filtered <- sample_metadata %>%
  filter(Sample %in% colnames(final_matrix)) %>%
  arrange(match(Sample, colnames(final_matrix)))

# Create the metadata dataframe
sample_metadata_df <- as.data.frame(sample_metadata_filtered)
rownames(sample_metadata_df) <- sample_metadata_df$Sample

# Create Biobase AnnotatedDataFrame objects
sample_phenoData <- AnnotatedDataFrame(sample_metadata_df)
metabolite_featureData <- AnnotatedDataFrame(metabolite_metadata_df)

# Create the ExpressionSet object, ensuring rownames match between assayData and featureData
final_annotated_df <- ExpressionSet(
  assayData = final_matrix,
  phenoData = sample_phenoData,
  featureData = metabolite_featureData
)

#Create Biobase ExpressionSet with imputed values for PCA
imputed_matrix <- apply(final_matrix, 2, min_value_impute)
imputed_annotated_df <- ExpressionSet(
  assayData = imputed_matrix,
  phenoData = sample_phenoData,
  featureData = metabolite_featureData
)
```

## Imputed Correlation for statistical testing
```{r Metabolite_Correlations, echo=TRUE}
# Calculate correlations and p-values between metabolites and metadata
metadata <- pData(imputed_annotated_df)
sex_numeric <- as.numeric(metadata$Sex == "M")  # Convert sex to numeric (0=F, 1=M)
GW <- metadata$GW

# Function to calculate correlation and p-value
correlation_test <- function(x, y) {
  result <- cor.test(x, y)
  c(correlation = result$estimate, p_value = result$p.value)
}

# Calculate correlations and p-values for each metabolite
metabolite_samplemetadata_stats <- data.frame(
  InChiKey = fData(imputed_annotated_df)$InChiKey,
  Name = fData(imputed_annotated_df)$Name,
  GW_corr_pval = t(apply(exprs(imputed_annotated_df), 1, function(x) correlation_test(x, GW)))
)

metabolite_samplemetadata_stats <- metabolite_samplemetadata_stats %>%
  mutate(
    GW_FDR = p.adjust(GW_corr_pval.p_value, method = "fdr")
  )

# Scatter plot for significant GW correlations
significant_GW <- metabolite_samplemetadata_stats %>% filter(GW_FDR < 0.05)
scatter_GW <- ggplot(significant_GW, aes(x = reorder(Name, GW_corr_pval.correlation.cor), y = GW_corr_pval.correlation.cor)) +
  geom_point(aes(size = -log10(GW_corr_pval.p_value), color = GW_corr_pval.correlation.cor)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Correlation with GW",
       title = "Significant Metabolite Correlations with Gestational Weeks (GW)") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right", axis.text.x = element_text(angle = 90, hjust = 1))
ggsave("GW_correlation.svg", path = svg_path)

# Heatmap of correlations
correlation_matrix <- metabolite_samplemetadata_stats %>%
  select( GW_corr_pval.correlation.cor) %>%
  as.matrix()

rownames(correlation_matrix) <- metabolite_samplemetadata_stats$Name
colnames(correlation_matrix) <- c("GW")

heatmap_plot <- Heatmap(correlation_matrix, name = "Correlation",
                        show_row_names = F, show_column_names = T)

# Plot the results
print(scatter_GW)
heatmap_plot

# Print top correlations with significance details
top_GW_correlations <- metabolite_samplemetadata_stats %>%
  arrange(desc(abs(GW_corr_pval.correlation.cor))) %>%
  head(10) %>%
  select(Name, GW_corr_pval.correlation.cor, GW_corr_pval.p_value, GW_FDR)
print(top_GW_correlations)
```


## Sex Differential Analysis
# Univariate Testing for Sex Differences with Volcano Plots
```{r Univariate_Testing, echo=TRUE, message=FALSE, warning=FALSE}

# Calculate p-values for sex differences using the Wilcoxon rank-sum test
metadata <- pData(final_annotated_df)
sex_numeric <- as.numeric(metadata$Sex == "M")  # Convert sex to numeric (0=F, 1=M)

# Function to calculate Wilcoxon rank-sum test p-value
wilcoxon_test <- function(x, sex) {
  result <- wilcox.test(x[sex == 0], x[sex == 1], exact = FALSE)
  c(statistic = result$statistic, p_value = result$p.value)
}

# Perform Wilcoxon rank-sum test for each metabolite
metabolite_sex_stats <- data.frame(
  InChiKey = fData(imputed_annotated_df)$InChiKey[apply(exprs(imputed_annotated_df), 1, function(x) all(is.finite(x)))],
  Name = fData(imputed_annotated_df)$Name[apply(exprs(imputed_annotated_df), 1, function(x) all(is.finite(x)))],
  Wilcoxon_stat_pval = t(apply(imputed_annotated_df, 1, function(x) wilcoxon_test(x, sex_numeric)))
)

# Calculate log fold changes for volcano plot
metabolite_sex_stats <- metabolite_sex_stats %>%
  mutate(
    logFC = log2(apply(imputed_annotated_df[, sex_numeric == 1], 1, mean) / apply(imputed_annotated_df[, sex_numeric == 0], 1, mean)),
    FDR = p.adjust(Wilcoxon_stat_pval.p_value, method = "fdr"),
    Bonferroni = p.adjust(Wilcoxon_stat_pval.p_value, method = "bonferroni")
  )

# Volcano plot with FDR correction
volcano_FDR <- ggplot(metabolite_sex_stats, aes(x = logFC, y = -log10(Wilcoxon_stat_pval.p_value))) +
  geom_point(aes(color = FDR < 0.05, size = -log10(Wilcoxon_stat_pval.p_value))) +
  theme_minimal() +
  labs(x = "Log2 Fold Change",
       y = "-log10(p-value)",
       title = "Volcano Plot with FDR Correction") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right")

# Volcano plot with Bonferroni correction
volcano_Bonferroni <- ggplot(metabolite_sex_stats, aes(x = logFC, y = -log10(Wilcoxon_stat_pval.p_value))) +
  geom_point(aes(color = Bonferroni < 0.05, size = -log10(Wilcoxon_stat_pval.p_value))) +
  theme_minimal() +
  labs(x = "Log2 Fold Change",
       y = "-log10(p-value)",
       title = "Volcano Plot with Bonferroni Correction") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right")

# Plot the results
print(volcano_FDR)
print(volcano_Bonferroni)

# Print top metabolites with significance details
top_metabolites <- metabolite_sex_stats %>%
  arrange(desc(-log10(Wilcoxon_stat_pval.p_value))) %>%
  head(10) %>%
  select(Name, logFC, Wilcoxon_stat_pval.p_value, FDR, Bonferroni)
print(top_metabolites)

```
# Multivariate testing
```{r Multivariate_Model, echo=TRUE}
metadata <- pData(final_annotated_df)
metadata$Sex <- as.factor(metadata$Sex)
sex_numeric <- as.numeric(metadata$Sex == "M")  # Convert sex to numeric (0=F, 1=M)
GW <- metadata$GW

metabolite_matrix <- exprs(final_annotated_df)

# Fit linear model and collect coefficients and p-values
lm_results <- lapply(1:nrow(metabolite_matrix), function(i) {
  metabolite <- metabolite_matrix[i, ]
  tryCatch({
    model <- lm(metabolite ~ Sex + GW, data = metadata)
    summary(model)
  }, error = function(e) {
    warning_message <- paste("Model fitting failed for row", i, "due to error:", e$message)
    message(warning_message)
    NULL
  })
})

# Extract coefficients and p-values
coefficients_list <- lapply(lm_results, function(res) {
  if (is.null(res) || !inherits(res, "summary.lm")) {
    return(c(Sex_estimate = NA, Sex_p_value = NA, GW_estimate = NA, GW_p_value = NA))
  }
  coef_table <- res$coefficients
  if (all(c("SexM", "GW") %in% rownames(coef_table))) {
    c(Sex_estimate = coef_table["SexM", "Estimate"],
      Sex_p_value = coef_table["SexM", "Pr(>|t|)"],
      GW_estimate = coef_table["GW", "Estimate"],
      GW_p_value = coef_table["GW", "Pr(>|t|)"])
  } else {
    c(Sex_estimate = NA, Sex_p_value = NA, GW_estimate = NA, GW_p_value = NA)
  }
})

coefficients_df <- do.call(rbind, coefficients_list)
metabolite_lm_stats <- data.frame(
  InChiKey = fData(imputed_annotated_df)$InChiKey[1:nrow(coefficients_df)],
  Name = fData(imputed_annotated_df)$Name[1:nrow(coefficients_df)],
  coefficients_df
)

# Adjust p-values for multiple testing
metabolite_lm_stats <- metabolite_lm_stats %>%
  mutate(
    Sex_FDR = p.adjust(Sex_p_value, method = "fdr"),
    GW_FDR = p.adjust(GW_p_value, method = "fdr"),
    Sex_Bonferroni = p.adjust(Sex_p_value, method = "bonferroni"),
    GW_Bonferroni = p.adjust(GW_p_value, method = "bonferroni")
  )

# Scatter plot for significant Sex effects
significant_sex <- metabolite_lm_stats %>% filter(Sex_FDR < 0.05)
scatter_sex <- ggplot(significant_sex, aes(x = reorder(Name, Sex_estimate), y = Sex_estimate)) +
  geom_point(aes(size = -log10(Sex_p_value), color = Sex_FDR < 0.05)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Effect Size of Sex",
       title = "Significant Metabolite Effects due to Sex") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right", axis.text.x = element_text(angle = 90, hjust = 1))


# Scatter plot for significant GW effects
significant_GW <- metabolite_lm_stats %>% filter(GW_FDR < 0.05)
scatter_GW <- ggplot(significant_GW, aes(x = reorder(Name, GW_estimate), y = GW_estimate)) +
  geom_point(aes(size = -log10(GW_p_value), color = GW_FDR < 0.05)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Effect Size of GW",
       title = "Significant Metabolite Effects due to Gestational Weeks (GW)") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right", axis.text.x = element_text(angle = 90, hjust = 1))

# Plot the results
print(scatter_sex)
print(scatter_GW)

# Print top metabolites with significance details
top_sex_metabolites <- metabolite_lm_stats %>%
  arrange(Sex_p_value) %>%
  head(10) %>%
  select(Name, Sex_estimate, Sex_p_value, Sex_FDR, Sex_Bonferroni)
print(top_sex_metabolites)

top_GW_metabolites <- metabolite_lm_stats %>%
  arrange(GW_p_value) %>%
  head(10) %>%
  select(Name, GW_estimate, GW_p_value, GW_FDR, GW_Bonferroni)
print(top_GW_metabolites)
```

```{r MetaboliteGraphs, echo = TRUE}
# Function to plot metabolite concentration across samples
plot_metabolite <- function(dataset, metabolite_id) {
  # Filter data for the selected metabolite
  metabolite_data <- dataset %>%
    filter(Name == metabolite_id) %>%
    select(starts_with("Sample"))
  
  # Gather data into long format for ggplot2
  metabolite_data_long <- metabolite_data %>%
    gather(Sample, Concentration)
  
  # Create the plot
  p <- ggplot(metabolite_data_long, aes(x = Sample, y = Concentration, group = 1)) +
    geom_point() +
    geom_line() +
    labs(title = paste("Concentration of Metabolite", metabolite_id, "across Samples"),
         x = "Sample",
         y = "Concentration") +
    theme_minimal()
  
  print(p)
}


# Function to plot metabolite intensity by sex
plot_metabolite_sex <- function(dataset, sample_metadata, metabolite_id) {
  # Filter data for the selected metabolite
  metabolite_data <- dataset %>%
    filter(Name == metabolite_id) %>%
    select(starts_with("Sample")) %>%
    gather(Sample, Intensity) %>%
    left_join(sample_metadata, by = "Sample")
  
  # Create the plot
  p_sex <- ggplot(metabolite_data, aes(x = Sex, y = Intensity, fill = Sex)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = paste("Intensity of Metabolite", metabolite_id, "by Sex"),
         x = "Sex",
         y = "Intensity") +
    theme_minimal()
  
  print(p_sex)
}

# Function to plot metabolite intensity by gestational weeks
plot_metabolite_gw <- function(dataset, sample_metadata, metabolite_id) {
  # Filter data for the selected metabolite
  metabolite_data <- dataset %>%
    filter(Name == metabolite_id) %>%
    select(starts_with("Sample")) %>%
    gather(Sample, Intensity) %>%
    left_join(sample_metadata, by = "Sample")
  
  # Create the plot
  p_gw <- ggplot(metabolite_data, aes(x = factor(GW), y = Intensity, fill = factor(GW))) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = paste("Intensity of Metabolite", metabolite_id, "by Gestational Weeks"),
         x = "Gestational Weeks",
         y = "Intensity") +
    theme_minimal()
  
  print(p_gw)
}
```
``` {r metabolite-plotting, echo=TRUE}
# Example usage: Plot Metabolite 'C3' across samples
plot_metabolite(unique_dataset, "7α-Hydroxytestosterone")
plot_metabolite_sex(unique_dataset, sample_metadata, "7α-Hydroxytestosterone")
plot_metabolite_gw(unique_dataset, sample_metadata, "7α-Hydroxytestosterone")

```