---
title: "Batch2 Exogenous Compound Flagging"
author: "Sneha M. Chaturvedi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---



# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r set_paths}
annotation_path <- "../../results/untargeted/batch2_metabolite_annotations.csv"
output_csv_path <- "../../results/untargeted/csv"
output_path     <- "../../results/untargeted/batch2_metabolite_annotations_flagged.csv"
candidates_path <- file.path(output_csv_path, "batch2_candidate_exogenous_flags.csv")

dir.create(output_csv_path, recursive = TRUE, showWarnings = FALSE)
```

```{r load_annotation}
annotation <- read_csv(annotation_path, show_col_types = FALSE)
cat("Compounds loaded:", nrow(annotation), "\n")
```

```{r source}
source("functions.R")
```
---

# Flag Definitions

Two flag levels are used:

- **`instrument_contaminant`**: Procedural artifacts from lab consumables, plastics, solvents, or industrial chemicals. These have no biological origin and are excluded from all downstream analyses.
- **`environmental_exposure`**: Pharmaceuticals, pesticides, dietary compounds, or synthetic hormones that may reflect genuine maternal/fetal exposure. These are retained in the main analysis but reported separately.
- **`review_needed`**: NIST14 spectral match entries where compound identity is uncertain. Resolved manually (see below).

Compounds with `NA` are presumed endogenous.

---

# Known Compounds

Compounds previously identified in project notes or resolved from manual NIST14 review.

```{r known_compounds}
known_instrument_contaminants <- c(
  # From project notes
  "erucamide",                    # C22:1 fatty acid amide; plastic slip agent
  "octinoxate",                   # UV filter / sunscreen agent
  "n,n'-diphenylguanidine",       # rubber accelerator / plasticizer additive
  "levoglucosan",                 # combustion product (biomass burning)
  "2-methylbenzoic acid",         # industrial solvent / toluene oxidation product
  "muscone",                      # synthetic fragrance (macrocyclic musk)
  # Resolved from NIST14 spectral match review
  "decaethylene glycol",          # PEG-10; synthetic polymer / lab consumable leachate
  "decanedioic acid, bis(2-ethylhexyl) ester",  # dioctyl sebacate; plasticizer
  "bis(2-ethylhexyl) sebacate",   # alias for above; same compound, different name
  "didodecyl 3,3'-thiodipropionate",              # plastic antioxidant / stabilizer
  "Irgafos 168",                  # plastic antioxidant / stabilizer
  "Victoria Pure Blue Bo",        # standard
  # Identified in metabolite-gene and metabolite-hormone review
  "methanesulfonic acid",         # industrial acid; common HPLC mobile-phase additive
  "meglumine",                    # N-methyl-D-glucamine; pharma excipient in contrast agents
  "triethyl citrate",             # synthetic citrate triester plasticizer/pharma excipient (≠ endogenous citrate)
  "dipropyleneglycol dibenzoate", # industrial plasticizer / coalescent solvent
  "peg n7",                       # polyethylene glycol oligomer; synthetic polymer leachate
  "peg n8",                       # polyethylene glycol oligomer; synthetic polymer leachate
  "peg n11",                      # polyethylene glycol oligomer; synthetic polymer leachate
  "peg n13",                      # polyethylene glycol oligomer; synthetic polymer leachate
  "peg n14",                      # polyethylene glycol oligomer; synthetic polymer leachate
  "6-chloro-5-methyl-1h-1,2,3-benzotriazole",              # chlorinated benzotriazole UV stabilizer / biocide
  "2-(2h-benzotriazol-2-yl)-4,6-bis(1-methyl-1-phenylethyl)phenol", # Tinuvin 234; UV absorber in plastics
  "octadecyl 3-(3,5-di-tert-butyl-4-hydroxyphenyl)propionate",      # Irganox 1076; polyolefin antioxidant
  "bis(2,2,6,6-tetramethyl-4-piperidyl)sebacate",                   # Tinuvin 770; hindered amine light stabilizer
  "bis(2-ethylhexyl) adipate",     # DEHA plasticizer
  # ── Audit round 3: remaining exogenous compounds in clean peak area ──
  # Plasticizers / solvents
  "citroflex 2",                   # triethyl citrate by trade name (chemical name already listed above)
  "citroflex 4",                   # tributyl citrate; plasticizer
  "citroflex a-4",                 # acetyltributyl citrate; plasticizer
  "acetyltributyl citrate",        # IUPAC name for Citroflex A-4 (duplicate entry in peak area)
  "tetraglyme",                    # tetraethylene glycol dimethyl ether; industrial solvent
  "triethylene glycol monobutyl ether",  # industrial solvent
  "triethyleneglycol bis(2-ethylhexanoate)", # synthetic plasticizer
  "triethyl phosphate",            # organophosphate flame retardant / plasticizer
  # Surfactants / detergents
  "4-dodecylbenzenesulfonic acid", # LAS anionic surfactant
  "dodecyl sulfate",               # SDS/SLS; lab and industrial detergent
  "cocamidopropyl betaine",        # synthetic amphoteric surfactant (cosmetics/lab)
  # UV filters / absorbers
  "octabenzone",                   # 2-hydroxy-4-octyloxybenzophenone; UV filter
  "dibenzoylmethane",              # UV absorber / chelating agent
  # Plastic additives
  "bis(4-ethylbenzylidene)sorbitol",  # DBS; plastic clarifying / nucleating agent
  "bis(methylbenzylidene)sorbitol",   # MDBS; same class
  "caprolactam",                   # epsilon-caprolactam; nylon-6 monomer
  # Polymer leachates
  "ppg n4",                        # polypropylene glycol oligomer; synthetic polymer leachate
  "ppg n5",                        # polypropylene glycol oligomer; synthetic polymer leachate
  # Lab reagents / buffers
  "trometamol",                    # tris(hydroxymethyl)aminomethane; laboratory buffer
  "\u03b1-cyano-3-hydroxycinnamic acid", # CHCA; MALDI matrix / MS standard
  # Industrial amines
  "cyclohexylamine",               # industrial amine / rubber accelerator metabolite
  "n-methylcyclohexylamine",       # industrial amine
  "3-isopropoxypropylamine",       # industrial ether-amine
  "3-(ethenyloxy)propan-1-amine",  # vinyl ether amine; industrial
  "n-methyl-2-phenylpropan-1-amine", # synthetic amine; no endogenous pathway
  # Fragrances / cosmetics
  "galaxolidone",                  # oxidation product of Galaxolide synthetic musk
  "triethanolamine",               # industrial/cosmetic tertiary amine
  "geranyllinalool"                # plant diterpene / fragrance compound; humans lack diterpene synthase
)

known_environmental_exposures <- c(
  "norgestrel",      # synthetic progestin (oral contraceptive)
  "methasterone",    # anabolic-androgenic synthetic steroid
  "minoxidil",       # vasodilator; topical/oral hair-loss or antihypertensive medication
  # Identified in metabolite-gene and metabolite-hormone review
  "ketamine",        # dissociative anesthetic; required for surgical tissue procurement
  "lidocaine",       # local anesthetic amide; required for surgical tissue procurement
  "midazolam",       # benzodiazepine sedative; required for surgical tissue procurement
  "fentanyl",        # synthetic opioid analgesic; required for surgical tissue procurement
  "piperine",         # alkaloid from black pepper (Piper nigrum); maternal dietary exposure
  # ── Audit round 3: remaining exogenous compounds in clean peak area ──
  # Pharmaceuticals
  "penicillin g",                  # beta-lactam antibiotic
  "allopurinol",                   # synthetic xanthine oxidase inhibitor (gout drug)
  "phthalylsulfathiazole",         # sulfonamide antibiotic (GI-targeted)
  "risedronic acid",               # bisphosphonate (osteoporosis drug)
  "6-aminocaproic acid",           # antifibrinolytic drug (Amicar); not a lysine catabolite
  "amifampridine",                 # 3,4-diaminopyridine; neuromuscular junction drug
  # Pesticides / herbicides
  "diethyl phosphate",             # organophosphate pesticide metabolite (not endogenous)
  "3-ketocarbofuran",              # carbofuran insecticide metabolite
  "propham",                       # isopropyl carbanilate; carbamate herbicide
  "6-chloropicolinic acid",        # clopyralid herbicide metabolite
  "maleic hydrazide",              # plant growth regulator / herbicide
  # Dietary plant compounds (no human biosynthetic pathway)
  "proline betaine",               # stachydrine; citrus-derived
  "solanidine",                    # steroidal alkaloid from Solanaceae (nightshades)
  "chalcone",                      # plant polyphenol; requires chalcone synthase (plant-only)
  "diosgenin",                     # steroidal sapogenin; plant-only biosynthesis (Dioscorea)
  "hypaphorine"                    # N,N,N-trimethyltryptophan; plant alkaloid (Erythrina)
)

cat("Known instrument contaminants:", length(known_instrument_contaminants), "\n")
cat("Known environmental exposures:", length(known_environmental_exposures), "\n")
```

---

# Keyword Patterns

```{r keyword_patterns}
# Instrument contaminants: plasticizers, industrial chemicals, slip agents,
# surfactants, PFAS, UV stabilizers, flame retardants, synthetic musks
instrument_contaminant_patterns <- paste(
  "phthalate", "diethylhexyl", "dioctyl", "dibutyl",
  "bisphenol",
  "tributyl phosphate", "triphenyl phosphate",
  "siloxane", "dimethylsilyl", "polydimethyl",
  "erucamide", "oleamide", "stearamide", "behenamide",
  "triton", "tween", "brij", "span [0-9]",
  "nonylphenol", "octylphenol",
  "perfluoro", "fluorotelomer",
  "benzophenone", "octinoxate", "avobenzone", "oxybenzone",
  "octocrylene", "homosalate",
  "tris(2-chloroethyl)", "hexabromocyclododecane",
  "levoglucosan", "mannosan", "galactosan",
  "diphenylguanidine","benzothiazole",
  "muscone", "galaxolide", "tonalide", "\\bmusk\\b", "Irgafos", "Pure Blue",
  "butylated hydroxytoluene", "\\bBHT\\b",
  "butylated hydroxyanisole", "\\bBHA\\b",
  # ── Audit round 3 patterns ──
  "citroflex",                     # catches all Citroflex variants (2, 4, A-4, A-2, etc.)
  "caprolactam",                   # nylon monomers
  "cocamidopropyl",                # amphoteric surfactant class
  "dodecylbenzenesulfonic",        # LAS surfactant class
  "\\bppg\\b",                     # polypropylene glycol oligomers
  "tetraglyme|triglyme",           # glyme solvents
  "trometamol|tromethamine",       # Tris buffer synonyms
  "triethanolamine",               # industrial/cosmetic amine
  "cyano.*hydroxycinnamic",        # MALDI matrix compounds (CHCA and variants)
  "galaxolidone",                  # Galaxolide oxidation product (not caught by "galaxolide" substring in all cases)
  sep = "|"
)

# Environmental exposures: synthetic steroids, pharmaceuticals, pesticides,
# preservatives, dietary stimulants.
# NOTE: levothyroxine/thyroxine (identical to endogenous T4) and progesterone
# (endogenous and biologically critical in fetal brain) are intentionally excluded.
environmental_exposure_patterns <- paste(
  # Synthetic hormones (contraceptives, anabolic steroids, SERMs)
  "norgestrel", "levonorgestrel", "medroxyprogesterone", "megestrol",
  "norethindrone", "gestodene", "desogestrel", "etonogestrel",
  "methasterone", "nandrolone", "stanozolol", "trenbolone",
  "boldenone", "oxandrolone", "methandrostenolone",
  "ethinylestradiol", "diethylstilbestrol", "\\bDES\\b",
  "tamoxifen", "clomiphene",
  # Synthetic glucocorticoids (distinct from endogenous cortisol/cortisone by name)
  "betamethasone", "dexamethasone", "\\bprednisone\\b", "prednisolone",
  # NSAIDs / analgesics
  "ibuprofen", "naproxen", "acetaminophen", "paracetamol",
  "\\baspirin\\b", "acetylsalicyl",
  # Cardiovascular / metabolic
  "metformin", "atorvastatin", "simvastatin", "lovastatin",
  "lisinopril", "amlodipine", "metoprolol",
  "labetalol", "hydralazine", "nifedipine", "methyldopa", "minoxidil",
  # CNS (SSRIs, SNRIs, benzodiazepines, antiepileptics)
  "fluoxetine", "sertraline", "paroxetine", "venlafaxine",
  "lorazepam", "diazepam", "alprazolam",
  "lamotrigine", "valproic acid", "\\bvalproate\\b", "valproex",
  # Antibiotics
  "amoxicillin", "azithromycin", "doxycycline",
  # Antiemetic
  "ondansetron",
  # Antihistamine
  "diphenhydramine",
  # Dietary stimulants
  "caffeine", "theobromine", "theophylline",
  # Pesticides
  "atrazine", "glyphosate", "chlorpyrifos", "permethrin",
  "cypermethrin", "deltamethrin", "malathion",
  # Preservatives
  "methylparaben", "ethylparaben", "propylparaben", "butylparaben",
  "\\bparaben\\b",
  # ── Audit round 3 patterns ──
  "penicillin",                    # beta-lactam antibiotics
  "allopurinol",                   # xanthine oxidase inhibitors
  "sulfathiazole",                 # sulfonamide antibiotics
  "risedronic|alendronic|zoledronic", # bisphosphonate class
  "aminocaproic",                  # antifibrinolytic drugs
  "carbofuran",                    # carbamate insecticide + metabolites (3-keto-, 3-hydroxy-)
  "chloropicolinic|clopyralid",    # picolinic acid herbicides
  "solanidine|\\bsolanine\\b",     # Solanaceae alkaloids
  "diosgenin",                     # plant steroidal sapogenins
  sep = "|"
)

# ClassyFire taxonomy classes that are predominantly synthetic/exogenous
exogenous_superclass <- c("Organohalogens")
exogenous_class      <- c(
  "Organochlorides", "Organobromides",
  "Chlorofluorocarbons", "Synthetic dyes", "Anabolic steroids"
)
```

---

# Apply Flags

```{r apply_flags}
# Compounds requiring manual spectrum review before a final exogenous decision.
# These are retained in analyses but excluded from downstream reports until resolved.
review_needed_compounds <- c(
  "dl-p-hydroxyphenyllactic acid"  # High score (95.1) but SMILES is a hydroperoxide
                                    # inconsistent with the endogenous tyrosine catabolite
                                    # p-HPLA (HMDB0000755); no HMDB/CAS in record.
                                    # Requires manual spectrum verification.
)

flag_df <- annotation %>%
  mutate(Name_lower = str_to_lower(Name)) %>%
  # Step 1: boolean flag columns
  mutate(
    flag_known_instrument    = Name_lower %in% known_instrument_contaminants,
    flag_known_exposure      = Name_lower %in% known_environmental_exposures,
    flag_review_needed       = Name_lower %in% review_needed_compounds,
    flag_pattern_instrument  = str_detect(Name_lower,
                                 regex(instrument_contaminant_patterns, ignore_case = TRUE)),
    flag_pattern_exposure    = str_detect(Name_lower,
                                 regex(environmental_exposure_patterns, ignore_case = TRUE)),
    flag_taxonomy_instrument = (Super.Class %in% exogenous_superclass) |
                               (Class %in% exogenous_class)
  ) %>%
  # Step 2: Exogenous_Flag
  # review_needed is checked first so it is not overridden by pattern matches
  mutate(
    Exogenous_Flag = case_when(
      flag_review_needed
        ~ "review_needed",
      flag_known_instrument | flag_pattern_instrument | flag_taxonomy_instrument
        ~ "instrument_contaminant",
      flag_known_exposure | flag_pattern_exposure
        ~ "environmental_exposure",
      TRUE ~ NA_character_
    )
  ) %>%
  # Step 3: Flag_Reason
  mutate(
    Flag_Reason = case_when(
      flag_review_needed       ~ "SMILES inconsistent with stated identity; manual spectrum review required before exclusion",
      flag_known_instrument    ~ "Known instrument contaminant (project notes or NIST14 review)",
      flag_known_exposure      ~ "Known synthetic pharmaceutical or steroid (project notes)",
      flag_pattern_instrument  ~ "Name matches industrial or contaminant keyword pattern",
      flag_pattern_exposure    ~ "Name matches pharmaceutical or pesticide keyword pattern",
      flag_taxonomy_instrument ~ "ClassyFire taxonomy flags synthetic or halogenated class",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-Name_lower, -starts_with("flag_", ignore.case = FALSE))
```

---

# Summary

```{r summary}
annotation_flagged <- flag_df %>%
  # Annotation confidence tiers based on Best.Match score
  # Tiers: 80+ high, 70-79 medium, 60-69 low
  # High-confidence (80+) compounds are bolded in tables throughout downstream reports
  mutate(
    Annotation_Confidence = case_when(
      Best.Match >= 80 ~ "high",
      Best.Match >= 70 ~ "medium",
      Best.Match >= 60 ~ "low",
      TRUE             ~ NA_character_
    )
  )

cat("=== Exogenous Flagging Summary ===\n")
cat("Total compounds:", nrow(annotation_flagged), "\n")
cat("Presumed endogenous (NA):",
    sum(is.na(annotation_flagged$Exogenous_Flag)), "\n\n")

flag_counts <- annotation_flagged %>%
  filter(!is.na(Exogenous_Flag)) %>%
  count(Exogenous_Flag)
print(flag_counts)
```

## Instrument Contaminants

```{r show_instrument}
annotation_flagged %>%
  filter(Exogenous_Flag == "instrument_contaminant") %>%
  select(Compound.ID, Name, Best.Match, Annotation_Confidence, Super.Class, Class, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Instrument Contaminants — excluded from all analyses. **Bold** = high confidence (score ≥ 80)")
```

## Environmental Exposures

```{r show_exposures}
annotation_flagged %>%
  filter(Exogenous_Flag == "environmental_exposure") %>%
  select(Compound.ID, Name, Best.Match, Annotation_Confidence, Super.Class, Class, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Environmental Exposures — removed from main analysis. **Bold** = high confidence (score ≥ 80)")
```

## Remaining Review Needed

```{r show_review}
remaining_review <- annotation_flagged %>%
  filter(Exogenous_Flag == "review_needed")

if (nrow(remaining_review) == 0) {
  cat("No compounds remaining in review_needed.\n")
} else {
  remaining_review %>%
    select(Compound.ID, Name, Super.Class, Class, Flag_Reason) %>%
    knitr::kable(caption = "Still requires manual review")
}
```

---

# Annotation Confidence Tiers

Compounds are assigned a confidence tier based on `Best.Match` spectral score. The score
distribution is flat from 60–88 then rises sharply at ≥90, with no natural sub-break
in the low-medium range. Tiers are defined as:

| Tier | Score | Interpretation |
|---|---|---|
| `high` | ≥ 80 | Strong spectral match; high identification confidence; **bolded in tables** |
| `medium` | 70–79 | Acceptable for untargeted metabolomics |
| `low` | 60–69 | Borderline; interpret with caution |

```{r confidence_tier_summary}
tier_counts <- annotation_flagged %>%
  count(Annotation_Confidence) %>%
  mutate(Annotation_Confidence = factor(Annotation_Confidence,
                                        levels = c("high", "medium", "low"))) %>%
  arrange(Annotation_Confidence)

print(tier_counts)

ggplot(annotation_flagged, aes(x = Best.Match, fill = Annotation_Confidence)) +
  geom_histogram(binwidth = 2, color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c(high = "#2166ac", medium = "#92c5de", low = "#f4a582"),
    na.value = "grey70",
    name = "Confidence"
  ) +
  geom_vline(xintercept = c(75, 90), linetype = "dashed", color = "grey30") +
  theme_minimal() +
  labs(
    title = "Annotation Score Distribution with Confidence Tiers",
    x = "Best Match Score",
    y = "Compound Count"
  )
```

---

# Save Outputs

```{r save}
# Full annotation with Exogenous_Flag, Flag_Reason, and Annotation_Confidence columns
write_csv(annotation_flagged, output_path)
cat("Flagged annotation saved to:", output_path, "\n")

# Candidate review table (flagged only)
candidates <- annotation_flagged %>%
  filter(!is.na(Exogenous_Flag)) %>%
  select(Compound.ID, Name, Super.Class, Class, Sub.Class,
         Best.Match, Annotation_Confidence,
         Exogenous_Flag, Flag_Reason) %>%
  arrange(Exogenous_Flag, Name)

write_csv(candidates, candidates_path)
cat("Candidate review table saved to:", candidates_path, "\n")
cat("Total flagged compounds:", nrow(candidates), "\n")
```

---

# Save Clean Peak Area Dataset

Removes all exogenous compounds (both `instrument_contaminant` and `environmental_exposure`)
from the peak area matrix and saves `batch2_peak_area_clean.csv`. This is the file read
by all downstream analysis scripts — no filtering is needed there.

```{r save_clean_peak_area}
peak_area_path  <- "../../results/untargeted/batch2_peak_area_values.csv"
clean_peak_path <- "../../results/untargeted/batch2_peak_area_clean.csv"

batch2_peak_area <- read_csv(peak_area_path, show_col_types = FALSE)

# review_needed compounds are retained in the peak area file until manual spectrum
# review is complete; only confirmed instrument_contaminant and environmental_exposure are removed.
exogenous_ids <- annotation_flagged %>%
  filter(Exogenous_Flag %in% c("instrument_contaminant", "environmental_exposure")) %>%
  pull(Compound.ID)

n_before <- nrow(batch2_peak_area)
batch2_peak_area_clean <- batch2_peak_area %>%
  filter(!Compound.ID %in% exogenous_ids)
n_after <- nrow(batch2_peak_area_clean)

cat("=== Clean Peak Area Dataset ===\n")
cat("Compounds before removal:", n_before, "\n")
cat("Exogenous removed:       ", n_before - n_after, "\n")
cat("  instrument_contaminant:", sum(annotation_flagged$Exogenous_Flag == "instrument_contaminant", na.rm = TRUE), "\n")
cat("  environmental_exposure:", sum(annotation_flagged$Exogenous_Flag == "environmental_exposure", na.rm = TRUE), "\n")
cat("  review_needed (retained):", sum(annotation_flagged$Exogenous_Flag == "review_needed", na.rm = TRUE), "\n")
cat("Compounds remaining:     ", n_after, "\n\n")

cat("--- Removed environmental exposures ---\n")
annotation_flagged %>%
  filter(Exogenous_Flag == "environmental_exposure") %>%
  select(Compound.ID, Name, Best.Match, Annotation_Confidence, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Environmental exposures removed. **Bold** = high confidence (score >= 80)")

write_csv(batch2_peak_area_clean, clean_peak_path)
cat("Clean peak area saved to:", clean_peak_path, "\n")
```
