---
title: "New Batch 2 Pathway Analysis (MSEA)"
author: "Sneha Michaela Chaturvedi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
---

# Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

Packages <- c(
  "readxl", "tidyverse", "fgsea", "KEGGREST", "knitr",
  "kableExtra", "svglite", "ggrepel"
)

lapply(Packages, library, character.only = TRUE)
```

```{r source_functions, include=FALSE}
source("functions.R")
```

```{r paths, include=FALSE}
png_path <- "../../results/untargeted/batch2_clean"
svg_path <- "../../results/untargeted/batch2_clean/svg"
csv_path <- "../../results/untargeted/batch2_clean/csv"

# Input files referenced by name so the KEGGREST cache can key off them
annotation_file <- "../../results/untargeted/batch2_metabolite_annotations.csv"
peak_area_file  <- "../../results/untargeted/batch2_peak_area_clean.csv"
```

```{r cache_control, include=FALSE}
# The keggrest_remap chunk issues ~80 rate-limited KEGGREST queries
# (~30 s per render), so it is cached. The cache invalidates automatically
# when the annotation or peak-area file changes (see cache.extra below).
#
# Set refresh_kegg <- TRUE to force a fresh pull from KEGG — do this when
# you want current KEGG pathway assignments even though the input files
# have not changed (KEGG is updated on their end, not ours).
refresh_kegg <- FALSE

cache_dir <- "Batch2_Pathway_Analysis_cache"

if (refresh_kegg && dir.exists(cache_dir)) {
  unlink(cache_dir, recursive = TRUE)
  message("Cleared knitr cache: ", cache_dir)
}
```

## Load Data

```{r load_annotation}
# Full annotation file — needed for KEGG IDs and Pathways column
batch2_compound_annotation_full <- read_csv(
  annotation_file,
  show_col_types = FALSE
)

# Peak area file defines the compounds that survived filtering
batch2_peak_area <- read_csv(
  peak_area_file,
  show_col_types = FALSE
)

# Restrict annotation to measured compounds only
batch2_compound_annotation <- batch2_compound_annotation_full %>%
  filter(Compound.ID %in% batch2_peak_area$Compound.ID)

cat("Total annotation rows:   ", nrow(batch2_compound_annotation_full), "\n")
cat("Measured compounds:      ", nrow(batch2_compound_annotation), "\n")

# Pathway ID → name lookup sheet from the combined annotation Excel file.
# Explicit rename by name — robust to column order and extra columns.
pathway_info <- read_excel(
  "../../doc/untargeted/combined_annotations_untargeted_developing_2026.xlsx",
  sheet = "Pathways"
) %>%
  rename(
    pathway_id     = PathwayId,
    pathway_name   = pathwayName,
    pathway_source = pathwaySource
  ) %>%
  dplyr::select(pathway_id, pathway_name, pathway_source)
```

```{r load_limma}
# Limma results from batch2_Sex_and_GW_Analysis.Rmd
# topTable output includes t-statistic, which is the ranking variable for MSEA
gw_limma <- read_csv(file.path(csv_path, "batch2_gw_limma_results.csv"),
  show_col_types = FALSE
)
sex_limma <- read_csv(file.path(csv_path, "batch2_sex_limma_results.csv"),
  show_col_types = FALSE
)

cat("GW limma:  ", nrow(gw_limma), "metabolites\n")
cat("Sex limma: ", nrow(sex_limma), "metabolites\n")
```

---

# Pathway Annotation

## Parse Existing Pathways Column

```{r parse_pathways}
# Pathways column format in new combined annotation: bare IDs separated by ";"
# e.g. "map00120;R-HSA-156587;WP100"
# Source is inferred from ID pattern — no source prefix present.

pathways_parsed <- batch2_compound_annotation %>%
  filter(!is.na(Pathways) & Pathways != "") %>%
  dplyr::select(Compound.ID, Name, KEGG, Pathways) %>%
  separate_rows(Pathways, sep = ";") %>%
  mutate(
    pathway_id     = str_trim(Pathways),
    pathway_source = case_when(
      str_detect(pathway_id, "^map\\d+")  ~ "kegg",
      str_detect(pathway_id, "^R-HSA-")   ~ "reactome",
      str_detect(pathway_id, "^WP\\d+$")  ~ "wiki",
      TRUE                                ~ NA_character_
    )
  ) %>%
  dplyr::select(-Pathways) %>%
  filter(!is.na(pathway_id) & pathway_id != "" & !is.na(pathway_source))

cat(
  "Metabolites with any pathway annotation (existing):",
  n_distinct(pathways_parsed$Name), "of", nrow(batch2_compound_annotation), "\n"
)
cat("Unique pathways (existing):", n_distinct(pathways_parsed$pathway_id), "\n")
cat("\nBreakdown by source:\n")
print(count(pathways_parsed, pathway_source))
```

## KEGGREST Re-Mapping

```{r keggrest_remap, cache=TRUE, cache.extra=tools::md5sum(c(annotation_file, peak_area_file))}
# Use the KEGG column (KEGG compound IDs, e.g. "C00031") to fetch
# additional pathway annotations for metabolites missing from Pathways column.
# With only ~80 compounds to query, per-compound keggLink() calls are feasible.

metabolites_missing_pathways <- batch2_compound_annotation %>%
  filter(is.na(Pathways) | Pathways == "") %>%
  filter(!is.na(KEGG) & KEGG != "") %>%
  dplyr::select(Compound.ID, Name, KEGG)

cat("Metabolites missing pathway annotation:", nrow(batch2_compound_annotation %>%
  filter(is.na(Pathways) | Pathways == "")), "\n")
cat("Of those, with KEGG ID available:", nrow(metabolites_missing_pathways), "\n")

# Per-compound KEGG pathway lookup.
# keggLink(target, source) with target = "pathway", source = "cpd:CXXXXX"
# returns map-level pathway IDs (e.g. "path:map00010") for that compound.
# Sys.sleep(0.35) stays within KEGG's rate limit.
# Some KEGG cells contain semicolon-separated multiple IDs (e.g. "C05282;C20948")
# Expand these so each ID is queried individually, then re-join to Compound.ID/Name
metabolites_missing_pathways <- metabolites_missing_pathways %>%
  separate_rows(KEGG, sep = ";") %>%
  mutate(KEGG = str_trim(KEGG)) %>%
  filter(!is.na(KEGG) & KEGG != "")

# Validate: KEGG compound IDs must be "C" + exactly 5 digits
metabolites_missing_pathways <- metabolites_missing_pathways %>%
  mutate(KEGG_valid = str_detect(KEGG, "^C\\d{5}$"))

cat("KEGG IDs passing format check:", sum(metabolites_missing_pathways$KEGG_valid), "\n")
cat("KEGG IDs skipped (malformed): ", sum(!metabolites_missing_pathways$KEGG_valid), "\n")

valid_ids <- metabolites_missing_pathways %>%
  filter(KEGG_valid) %>%
  pull(KEGG) %>%
  unique()

cat("\nQuerying KEGGREST per compound (", length(valid_ids), "queries)...\n")
flush.console()

query_kegg_pathways <- function(kegg_ids) {
  results <- vector("list", length(kegg_ids))
  for (i in seq_along(kegg_ids)) {
    kid <- kegg_ids[i]
    results[[i]] <- tryCatch(
      {
        links <- keggLink("pathway", paste0("cpd:", kid))
        # links: names = "cpd:CXXXXX", values = "path:mapXXXXX"
        tibble(KEGG = kid, pathway_id = str_remove(links, "path:"))
      },
      error = function(e) {
        message("  Query failed for ", kid, ": ", e$message)
        tibble(KEGG = character(), pathway_id = character())
      }
    )
    if (i %% 10 == 0) {
      cat("  ", i, "/", length(kegg_ids), "\n")
      flush.console()
    }
    Sys.sleep(0.35)
  }
  bind_rows(results)
}

kegg_link_df <- query_kegg_pathways(valid_ids)
```

```{r build_kegg_new, cache=TRUE, dependson="keggrest_remap"}
if (nrow(kegg_link_df) > 0) {
  kegg_new <- metabolites_missing_pathways %>%
    inner_join(kegg_link_df, by = "KEGG") %>%
    mutate(pathway_source = "kegg") %>%
    dplyr::select(Compound.ID, Name, KEGG, pathway_source, pathway_id)

  cat("New KEGG pathway assignments retrieved:", nrow(kegg_new), "\n")
  cat("Metabolites newly covered:", n_distinct(kegg_new$Name), "\n")
} else {
  kegg_new <- tibble(
    Compound.ID = character(), Name = character(),
    KEGG = character(), pathway_source = character(),
    pathway_id = character()
  )
  cat("No KEGG pathway links returned — proceeding with existing annotations only.\n")
}
```

## Combine and Map Pathway Names

```{r combine_pathway_annotations}
# Stack existing + KEGG-remapped annotations
# For metabolites that already had pathway annotations, keep them as-is;
# KEGG new rows only add metabolites that had no annotations before.
pathways_combined <- bind_rows(
  pathways_parsed,
  kegg_new %>% dplyr::select(Compound.ID, Name, pathway_source, pathway_id)
)

# Attach human-readable pathway names from the core's Pathways sheet
# pathway_info has: pathway_name, pathway_source, pathway_id
pathways_combined <- pathways_combined %>%
  left_join(
    pathway_info %>% dplyr::select(pathway_id, pathway_name),
    by = "pathway_id"
  ) %>%
  mutate(
    # For pathway IDs without a name in the sheet, construct label from source + ID
    pathway_label = case_when(
      !is.na(pathway_name) & pathway_name != "" ~ pathway_name,
      pathway_source == "kegg"     ~ paste0("KEGG:", pathway_id),
      pathway_source == "reactome" ~ paste0("Reactome:", pathway_id),
      pathway_source == "wiki"     ~ paste0("WikiPathways:", pathway_id),
      pathway_source == "pfocr"    ~ paste0("PFOCR:", pathway_id),
      TRUE ~ pathway_id
    )
  )

cat("\n=== Final Pathway Coverage ===\n")
cat(
  "Metabolites with any pathway annotation:",
  n_distinct(pathways_combined$Name), "of", nrow(batch2_compound_annotation), "\n"
)
cat("Unique pathways:", n_distinct(pathways_combined$pathway_id), "\n\n")
cat("Breakdown by source:\n")
print(count(pathways_combined, pathway_source))
```

## Build Pathway Sets

```{r build_pathway_sets}
# MSEA requires: list(pathway_label = character vector of metabolite Names)
# Filter: minimum 5 and maximum 500 members in the measured background
# Use pathway_label (human-readable) as the set name

# For duplicate pathway IDs across sources, keep distinct Name-pathway_id pairs
pathway_sets_raw <- pathways_combined %>%
  distinct(Name, pathway_id, pathway_label, pathway_source) %>%
  filter(Name %in% gw_limma$Name) # restrict to metabolites in our analysis

pathway_sets <- pathway_sets_raw %>%
  mutate(pathway_source = factor(pathway_source,
    levels = c("kegg",  "reactome", "wiki",          "pfocr"),
    labels = c("KEGG",  "Reactome", "WikiPathways",  "PFOCR")
  )) %>%
  group_by(pathway_label, pathway_source) %>%
  summarise(members = list(unique(Name)), .groups = "drop") %>%
  mutate(
    n_members = lengths(members),
    # Append source to label so cross-database duplicates are distinct
    pathway_label_src = paste0(pathway_label, " (", pathway_source, ")")
  ) %>%
  filter(n_members >= 5, n_members <= 500)

# Lookup table to re-join source onto fgsea results after the fact
# Key is the unique pathway_label_src used as fgsea set name
pathway_source_lookup <- pathway_sets %>%
  dplyr::select(pathway_label_src, pathway_source)

# Convert to named list for fgsea — keyed by the source-appended label
pathway_list <- setNames(pathway_sets$members, pathway_sets$pathway_label_src)

cat("Pathway sets available for MSEA:", length(pathway_list), "\n")
cat("Median members per set:", median(lengths(pathway_list)), "\n")
cat("Range:", min(lengths(pathway_list)), "–", max(lengths(pathway_list)), "\n")
```

---

# MSEA

MSEA (Metabolite Set Enrichment Analysis) is the metabolomics analog of GSEA. Rather
than requiring a significance cutoff, it uses the **full ranked list** of metabolites
ordered by their limma t-statistic. This is appropriate here given sparse individual
hits (especially for sex), as it uses all available signal.

The ranking statistic is the limma **t-statistic**, which combines effect size and
precision. Implemented via `fgsea` with 10,000 permutations.

```{r msea_function}
run_msea <- function(limma_results, pathway_list, analysis_label,
                     nperm = 10000, seed = 42) {
  # Build named ranked vector: Name -> t-statistic
  # Remove any NAs (metabolites where model didn't converge)
  ranked <- limma_results %>%
    filter(!is.na(t)) %>%
    arrange(desc(t)) %>%
    dplyr::select(Name, t) %>%
    deframe() # named numeric vector

  cat("\n===", analysis_label, "===\n")
  cat("Metabolites in ranked list:", length(ranked), "\n")
  cat("Pathway sets to test:", length(pathway_list), "\n")

  # Restrict pathway sets to metabolites present in this ranked list
  pathway_list_filtered <- lapply(pathway_list, function(x) {
    intersect(x, names(ranked))
  })
  pathway_list_filtered <- pathway_list_filtered[lengths(pathway_list_filtered) >= 5]
  cat("Pathway sets after restricting to ranked list:", length(pathway_list_filtered), "\n")

  if (length(pathway_list_filtered) == 0) {
    cat("No pathway sets have sufficient coverage — skipping.\n")
    return(NULL)
  }

  set.seed(seed)
  results <- fgsea(
    pathways = pathway_list_filtered,
    stats = ranked,
    nPermSimple = nperm,
    minSize = 5,
    maxSize = 500
  )

  results <- results %>%
    as_tibble() %>%
    arrange(padj, pval) %>%
    mutate(analysis = analysis_label)

  cat("Pathways tested:", nrow(results), "\n")
  cat("Nominal p < 0.05:", sum(results$pval < 0.05, na.rm = TRUE), "\n")
  cat("FDR < 0.05:", sum(results$padj < 0.05, na.rm = TRUE), "\n")
  cat("FDR < 0.25:", sum(results$padj < 0.25, na.rm = TRUE), "\n")

  return(results)
}
```

## GW Analysis

```{r msea_gw}
msea_gw <- run_msea(gw_limma, pathway_list,
  analysis_label = "GW (limma t-statistic)"
)
```

## Sex Analysis

```{r msea_sex}
msea_sex <- run_msea(sex_limma, pathway_list,
  analysis_label = "Sex (limma t-statistic, positive = higher in males)"
)
```

---

# Results

## GW — Top Pathways

```{r results_gw}
if (!is.null(msea_gw)) {
  msea_gw %>%
    left_join(pathway_source_lookup, by = c("pathway" = "pathway_label_src")) %>%
    head(30) %>%
    dplyr::select(pathway, pathway_source, pval, padj, NES, size, leadingEdge) %>%
    mutate(
      pval = signif(pval, 3),
      padj = signif(padj, 3),
      NES = round(NES, 3),
      leadingEdge = map_chr(leadingEdge, ~ paste(head(.x, 5), collapse = ", "))
    ) %>%
    rename(
      Pathway = pathway,
      Source = pathway_source,
      `p-value` = pval,
      `FDR` = padj,
      `NES` = NES,
      `Set size` = size,
      `Leading edge (top 5)` = leadingEdge
    ) %>%
    kable(caption = "GW MSEA — top 30 pathways by FDR") %>%
    kable_styling(
      bootstrap_options = c("striped", "condensed"),
      full_width = FALSE, font_size = 11
    )
}
```

## Sex — Top Pathways

```{r results_sex}
if (!is.null(msea_sex)) {
  msea_sex %>%
    left_join(pathway_source_lookup, by = c("pathway" = "pathway_label_src")) %>%
    head(30) %>%
    dplyr::select(pathway, pathway_source, pval, padj, NES, size, leadingEdge) %>%
    mutate(
      pval = signif(pval, 3),
      padj = signif(padj, 3),
      NES = round(NES, 3),
      leadingEdge = map_chr(leadingEdge, ~ paste(head(.x, 5), collapse = ", "))
    ) %>%
    rename(
      Pathway = pathway,
      Source = pathway_source,
      `p-value` = pval,
      `FDR` = padj,
      `NES` = NES,
      `Set size` = size,
      `Leading edge (top 5)` = leadingEdge
    ) %>%
    kable(caption = "Sex MSEA — top 30 pathways by FDR") %>%
    kable_styling(
      bootstrap_options = c("striped", "condensed"),
      full_width = FALSE, font_size = 11
    )
}
```

---

# Visualizations

```{r plot_function}
# Companion label file for truncated pathway axes.
# The dotplots clip pathway names to 45 chars; this writes the full names
# (plus plotted statistics) next to the SVG so figures stay editable.
write_pathway_labels <- function(plot_data, file_stem, label_dir, order_by) {
  if (is.null(label_dir) || is.null(file_stem)) return(invisible(NULL))
  dir.create(label_dir, recursive = TRUE, showWarnings = FALSE)

  # Degrade to unsorted rather than aborting a long render if the sort
  # column is absent (e.g. a plot function stops computing it).
  if (!is.null(order_by) && !order_by %in% names(plot_data)) {
    warning("write_pathway_labels: '", order_by,
            "' not found; writing unsorted labels for ", file_stem, ".")
    order_by <- NULL
  }

  out <- plot_data
  if (!is.null(order_by)) out <- arrange(out, desc(.data[[order_by]]))

  out <- out %>%
    mutate(plot_order = row_number()) %>%
    dplyr::select(
      plot_order,
      axis_label   = pathway_label,   # truncated string drawn on the y-axis
      pathway_full = pathway,         # untruncated, with " (Source)" suffix
      pathway_source, stars,
      dplyr::any_of(c("NES", "size", "n_hits", "set_size",
                      "odds_ratio", "pval", "padj"))
    )

  readr::write_csv(out, file.path(label_dir, paste0(file_stem, "_labels.csv")))
  invisible(out)
}

# Dot plot: NES on x, pathway on y, color = database, size = set size
# FDR significance shown as stars appended to pathway labels on the y-axis
# * p < 0.05, ** FDR < 0.05, *** FDR < 0.01
plot_msea_dotplot <- function(results, pathway_source_lookup, analysis_label,
                              top_n = 25, file_stem = NULL, label_dir = NULL) {
  if (is.null(results) || nrow(results) == 0) {
    message("No results to plot for: ", analysis_label)
    return(NULL)
  }

  plot_data <- results %>%
    left_join(pathway_source_lookup, by = c("pathway" = "pathway_label_src")) %>%
    filter(pval < 0.05) %>%
    arrange(padj) %>%
    head(top_n) %>%
    mutate(
      stars = case_when(
        padj < 0.01 ~ "**",
        padj < 0.05 ~ "*",
        TRUE ~ ""
      ),
      # Strip the " (Source)" suffix before truncating, then re-append
      # so the source is never swallowed by str_trunc
      pathway_base = str_remove(pathway,
        " \\(KEGG\\)$| \\(Reactome\\)$| \\(WikiPathways\\)$| \\(PFOCR\\)$"),
      pathway_short = str_trunc(pathway_base, 45),
      pathway_label = paste0(
        pathway_short, " (", pathway_source, ")",
        if_else(stars != "", paste0(" ", stars), "")
      )
    )

  if (nrow(plot_data) == 0) {
    message("No nominally significant pathways for: ", analysis_label)
    return(NULL)
  }

  write_pathway_labels(plot_data, file_stem, label_dir, order_by = "NES")

  ggplot(
    plot_data,
    aes(
      x = NES,
      y = reorder(pathway_label, NES),
      color = pathway_source,
      size = size,
      shape = pathway_source
    )
  ) +
    geom_point(alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    scale_color_manual(
      name = "Database",
      values = c(
        "KEGG"         = "#E41A1C",
        "Reactome"     = "#377EB8",
        "WikiPathways" = "#4DAF4A",
        "PFOCR"        = "#FF7F00"
      )
    ) +
    scale_shape_manual(
      name = "Database",
      values = c(
        "KEGG"         = 16,
        "Reactome"     = 17,
        "WikiPathways" = 15,
        "PFOCR"        = 18
      )
    ) +
    scale_size_continuous(name = "Set size", range = c(2, 8)) +
    theme_minimal(base_size = 11) +
    labs(
      x = "Normalized Enrichment Score (NES)",
      y = NULL,
      title = paste0("MSEA — ", analysis_label),
      subtitle = paste0(
        "Top ", nrow(plot_data), " pathways by FDR (p < 0.05 shown)\n",
        "* FDR < 0.05   ** FDR < 0.01"
      )
    ) +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9),
      axis.text.y = element_text(size = 8),
      legend.position = "right"
    )
}
```

## GW Dot Plot

```{r plot_gw, fig.width=10, fig.height=8}
p_gw  <- plot_msea_dotplot(msea_gw,  pathway_source_lookup, "GW",
                           file_stem = "msea_gw_dotplot",  label_dir = svg_path)
if (!is.null(p_gw)) {
  save_dual_format(p_gw, "msea_gw_dotplot", png_path = png_path, svg_path = svg_path)
  print(p_gw)
}
```

## Sex Dot Plot

```{r plot_sex, fig.width=10, fig.height=8}
p_sex <- plot_msea_dotplot(msea_sex, pathway_source_lookup,
                           "Sex (positive NES = enriched in males)",
                           file_stem = "msea_sex_dotplot", label_dir = svg_path)
if (!is.null(p_sex)) {
  save_dual_format(p_sex, "msea_sex_dotplot", png_path = png_path, svg_path = svg_path)
  print(p_sex)
}
```

## fgsea Enrichment Plots — Top GW Pathways

```{r enrichment_plots_gw, fig.width=8, fig.height=4}
# Individual enrichment plots for top 5 GW pathways by FDR
if (!is.null(msea_gw) && nrow(msea_gw) > 0) {
  ranked_gw <- gw_limma %>%
    filter(!is.na(t)) %>%
    arrange(desc(t)) %>%
    dplyr::select(Name, t) %>%
    deframe()

  # Restrict to pathways present in filtered list
  pathway_list_gw <- lapply(pathway_list, function(x) intersect(x, names(ranked_gw)))
  pathway_list_gw <- pathway_list_gw[lengths(pathway_list_gw) >= 5]

  top_gw_pathways <- msea_gw %>%
    arrange(padj) %>%
    head(5) %>%
    pull(pathway)

  for (pw in top_gw_pathways) {
    if (pw %in% names(pathway_list_gw)) {
      p <- plotEnrichment(pathway_list_gw[[pw]], ranked_gw) +
        labs(
          title = str_trunc(pw, 70),
          subtitle = paste0(
            "NES = ", round(msea_gw$NES[msea_gw$pathway == pw], 3),
            "  |  FDR = ", signif(msea_gw$padj[msea_gw$pathway == pw], 3)
          ),
          x = "Rank in GW t-statistic",
          y = "Enrichment Score"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          plot.subtitle = element_text(size = 9, hjust = 0.5, color = "grey40")
        )
      print(p)
    }
  }
}
```

## fgsea Enrichment Plots — Top Sex Pathways

```{r enrichment_plots_sex, fig.width=8, fig.height=4}
if (!is.null(msea_sex) && nrow(msea_sex) > 0) {
  ranked_sex <- sex_limma %>%
    filter(!is.na(t)) %>%
    arrange(desc(t)) %>%
    dplyr::select(Name, t) %>%
    deframe()

  pathway_list_sex <- lapply(pathway_list, function(x) intersect(x, names(ranked_sex)))
  pathway_list_sex <- pathway_list_sex[lengths(pathway_list_sex) >= 5]

  top_sex_pathways <- msea_sex %>%
    arrange(padj) %>%
    head(5) %>%
    pull(pathway)

  for (pw in top_sex_pathways) {
    if (pw %in% names(pathway_list_sex)) {
      p <- plotEnrichment(pathway_list_sex[[pw]], ranked_sex) +
        labs(
          title = str_trunc(pw, 70),
          subtitle = paste0(
            "NES = ", round(msea_sex$NES[msea_sex$pathway == pw], 3),
            "  |  FDR = ", signif(msea_sex$padj[msea_sex$pathway == pw], 3),
            "  |  positive NES = enriched in males"
          ),
          x = "Rank in sex t-statistic (M > F)",
          y = "Enrichment Score"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 10, hjust = 0.5),
          plot.subtitle = element_text(size = 9, hjust = 0.5, color = "grey40")
        )
      print(p)
    }
  }
}
```

---

# Over-Representation Analysis (ORA)

ORA tests whether a specific hit list is enriched for particular pathways compared
to the full measured background, using Fisher's exact test. Unlike MSEA, ORA uses
a binary foreground/background split and is sensitive to the threshold chosen.
Results should be interpreted alongside the MSEA above.

The background for all ORA tests is the full set of measured metabolites.

```{r ora_function}
# Fisher's exact test ORA
# foreground: character vector of hit metabolite Names
# background: character vector of all measured metabolite Names
# pathway_list: named list of pathway member vectors (built above)
# pathway_source_lookup: data frame mapping pathway_label_src -> pathway_source
run_ora <- function(foreground, background, pathway_list,
                    pathway_source_lookup, analysis_label) {

  cat("\n===", analysis_label, "===\n")
  cat("Foreground:", length(foreground), "metabolites\n")
  cat("Background:", length(background), "metabolites\n")

  if (length(foreground) == 0) {
    cat("Empty foreground — skipping.\n")
    return(NULL)
  }

  bg_only <- background[!background %in% foreground]

  results <- map_dfr(names(pathway_list), function(pw) {
    members <- pathway_list[[pw]]

    a <- sum(foreground %in% members)   # fg in pathway
    b <- length(foreground) - a         # fg not in pathway
    c <- sum(bg_only %in% members)      # bg in pathway
    d <- length(bg_only) - c            # bg not in pathway

    ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2), alternative = "greater")

    tibble(
      pathway    = pw,
      n_hits     = a,
      set_size   = a + c,
      odds_ratio = as.numeric(ft$estimate),
      pval       = ft$p.value,
      hit_names  = paste(foreground[foreground %in% members], collapse = ";")
    )
  }) %>%
    filter(n_hits > 0) %>%
    mutate(padj = p.adjust(pval, method = "BH")) %>%
    left_join(pathway_source_lookup, by = c("pathway" = "pathway_label_src")) %>%
    arrange(padj, pval)

  cat("Pathways with \u22651 hit:", nrow(results), "\n")
  cat("Nominal p < 0.05:         ", sum(results$pval  < 0.05, na.rm = TRUE), "\n")
  cat("FDR < 0.05:               ", sum(results$padj < 0.05, na.rm = TRUE), "\n")

  results
}

# Display helper — top N by p-value, formatted kable
display_ora_table <- function(results, caption, top_n = 20) {
  if (is.null(results) || nrow(results) == 0) {
    cat("No results.\n"); return(invisible(NULL))
  }
  results %>%
    head(top_n) %>%
    mutate(
      odds_ratio = round(odds_ratio, 2),
      pval       = signif(pval, 3),
      padj       = signif(padj, 3),
      hit_names  = str_trunc(hit_names, 60)
    ) %>%
    dplyr::select(pathway, pathway_source, n_hits, set_size,
                  odds_ratio, pval, padj, hit_names) %>%
    rename(
      Pathway           = pathway,
      Source            = pathway_source,
      `Hits in set`     = n_hits,
      `Set size`        = set_size,
      `Odds ratio`      = odds_ratio,
      `p-value`         = pval,
      `FDR`             = padj,
      `Hit metabolites` = hit_names
    ) %>%
    kable(caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE, font_size = 11) %>%
    print()
}

# ORA dot plot — mirrors MSEA plot style
plot_ora_dotplot <- function(results, analysis_label, top_n = 20,
                             file_stem = NULL, label_dir = NULL) {
  if (is.null(results) || nrow(results) == 0) return(NULL)

  plot_data <- results %>%
    filter(pval < 0.05) %>%
    head(top_n) %>%
    mutate(
      stars = case_when(
        padj < 0.01 ~ "**",
        padj < 0.05 ~ "*",
        TRUE        ~ ""
      ),
      log2_or = log2(odds_ratio),
      pathway_base  = str_remove(pathway,
        " \\(KEGG\\)$| \\(Reactome\\)$| \\(WikiPathways\\)$| \\(PFOCR\\)$"),
      pathway_short = str_trunc(pathway_base, 45),
      pathway_label = paste0(pathway_short, " (", pathway_source, ")",
                             if_else(stars != "", paste0(" ", stars), ""))
    )

  if (nrow(plot_data) == 0) {
    message("No nominally significant pathways for: ", analysis_label)
    return(NULL)
  }
  
  write_pathway_labels(plot_data, file_stem, label_dir, order_by = "log2_or")

  ggplot(plot_data,
         aes(x = log2(odds_ratio),
             y = reorder(pathway_label, log2(odds_ratio)),
             color = pathway_source,
             shape = pathway_source,
             size  = set_size)) +
    geom_point(alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    scale_color_manual(name = "Database",
                       values = c(
                         "KEGG"         = "#E41A1C",
                         "Reactome"     = "#377EB8",
                         "WikiPathways" = "#4DAF4A",
                         "PFOCR"        = "#FF7F00"
                       )) +
    scale_shape_manual(name = "Database",
                       values = c(
                         "KEGG"         = 16,
                         "Reactome"     = 17,
                         "WikiPathways" = 15,
                         "PFOCR"        = 18
                       )) +
    scale_size_continuous(name = "Set size", range = c(2, 8)) +
    theme_minimal(base_size = 11) +
    labs(
      x        = "log2(Odds Ratio)",
      y        = NULL,
      title    = paste0("ORA — ", analysis_label),
      subtitle = paste0("Top ", nrow(plot_data), " pathways by p-value (p < 0.05 shown)\n",
                        "* FDR < 0.05   ** FDR < 0.01")
    ) +
    theme(
      plot.title    = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9),
      axis.text.y   = element_text(size = 8),
      legend.position = "right"
    )
}

# Background: all measured metabolite Names
background_all <- gw_limma$Name
```

## Load Hit Lists

```{r ora_hit_lists}
# Sex limma — nominally significant
sex_hits_nominal <- sex_limma %>%
  filter(P.Value < 0.05) %>%
  pull(Name)
cat("Sex limma nominal (p < 0.05):      ", length(sex_hits_nominal), "\n")

# GW limma — nominally significant
gw_hits_nominal <- gw_limma %>%
  filter(P.Value < 0.05) %>%
  pull(Name)
cat("GW limma nominal (p < 0.05):       ", length(gw_hits_nominal), "\n")

# GW limma — FDR significant
gw_hits_fdr <- gw_limma %>%
  filter(adj.P.Val < 0.05) %>%
  pull(Name)
cat("GW limma FDR (adj.P.Val < 0.05):   ", length(gw_hits_fdr), "\n")

# Sex PLS-DA VIP
plsda_vip <- read_csv(file.path(csv_path, "batch2_sex_plsda_vip.csv"),
                      show_col_types = FALSE)
sex_hits_vip1 <- plsda_vip %>% filter(VIP_comp1 > 1) %>% pull(Name)
sex_hits_vip2 <- plsda_vip %>% filter(VIP_comp1 > 2) %>% pull(Name)
cat("Sex PLS-DA VIP > 1:                ", length(sex_hits_vip1), "\n")
cat("Sex PLS-DA VIP > 2:                ", length(sex_hits_vip2), "\n")
```

## Run ORA

```{r run_ora}
ora_sex_nominal <- run_ora(sex_hits_nominal, background_all, pathway_list,
                           pathway_source_lookup, "Sex limma nominal (p < 0.05)")

ora_gw_nominal  <- run_ora(gw_hits_nominal,  background_all, pathway_list,
                           pathway_source_lookup, "GW limma nominal (p < 0.05)")

ora_gw_fdr      <- run_ora(gw_hits_fdr,      background_all, pathway_list,
                           pathway_source_lookup, "GW limma FDR (adj.P < 0.05)")

ora_vip1        <- run_ora(sex_hits_vip1,    background_all, pathway_list,
                           pathway_source_lookup, "Sex PLS-DA VIP > 1")

ora_vip2        <- run_ora(sex_hits_vip2,    background_all, pathway_list,
                           pathway_source_lookup, "Sex PLS-DA VIP > 2")
```

## Results

### Sex — Limma Nominal

```{r ora_results_sex_nominal, fig.width=10, fig.height=7}
display_ora_table(ora_sex_nominal,
                  "ORA: Sex limma nominal (p < 0.05) — top 20 by p-value")
p <- plot_ora_dotplot(ora_sex_nominal, "Sex limma nominal (p < 0.05)",
                      file_stem = "ora_sex_limma_nominal", label_dir = svg_path)
if (!is.null(p)) {
  save_dual_format(p, "ora_sex_limma_nominal",
                   png_path = png_path, svg_path = svg_path)
  print(p)
}
```

### GW — Limma Nominal

```{r ora_results_gw_nominal, fig.width=10, fig.height=7}
display_ora_table(ora_gw_nominal,
                  "ORA: GW limma nominal (p < 0.05) — top 20 by p-value")
p <- plot_ora_dotplot(ora_gw_nominal,  "GW limma nominal (p < 0.05)",
                      file_stem = "ora_gw_limma_nominal",  label_dir = svg_path)
if (!is.null(p)) {
  save_dual_format(p, "ora_gw_limma_nominal",
                   png_path = png_path, svg_path = svg_path)
  print(p)
}
```

### GW — Limma FDR

```{r ora_results_gw_fdr, fig.width=10, fig.height=7}
display_ora_table(ora_gw_fdr,
                  "ORA: GW limma FDR (adj.P < 0.05) — top 20 by p-value")
p <- plot_ora_dotplot(ora_gw_fdr,      "GW limma FDR (adj.P < 0.05)",
                      file_stem = "ora_gw_limma_fdr",      label_dir = svg_path)
if (!is.null(p)) {
  save_dual_format(p, "ora_gw_limma_fdr",
                   png_path = png_path, svg_path = svg_path)
  print(p)
}
```

### Sex — PLS-DA VIP > 1

```{r ora_results_vip1, fig.width=10, fig.height=7}
display_ora_table(ora_vip1,
                  "ORA: Sex PLS-DA VIP > 1 — top 20 by p-value")
p <- plot_ora_dotplot(ora_vip1,        "Sex PLS-DA VIP > 1",
                      file_stem = "ora_sex_plsda_vip1",    label_dir = svg_path)
if (!is.null(p)) {
  save_dual_format(p, "ora_sex_plsda_vip1",
                   png_path = png_path, svg_path = svg_path)
  print(p)
}
```

### Sex — PLS-DA VIP > 2

```{r ora_results_vip2, fig.width=10, fig.height=7}
display_ora_table(ora_vip2,
                  "ORA: Sex PLS-DA VIP > 2 — top 20 by p-value")
p <- plot_ora_dotplot(ora_vip2,        "Sex PLS-DA VIP > 2",
                      file_stem = "ora_sex_plsda_vip2",    label_dir = svg_path)
if (!is.null(p)) {
  save_dual_format(p, "ora_sex_plsda_vip2",
                   png_path = png_path, svg_path = svg_path)
  print(p)
}
```

## Save ORA Results

```{r save_ora}
ora_list <- list(
  sex_limma_nominal = ora_sex_nominal,
  gw_limma_nominal  = ora_gw_nominal,
  gw_limma_fdr      = ora_gw_fdr,
  sex_plsda_vip1    = ora_vip1,
  sex_plsda_vip2    = ora_vip2
)

for (nm in names(ora_list)) {
  df <- ora_list[[nm]]
  if (!is.null(df)) {
    write_csv(df, file.path(csv_path, paste0("batch2_ora_", nm, ".csv")))
  }
}
cat("ORA results saved to:", csv_path, "\n")
```

---

# Save Results

```{r save}
# Save pathway sets (for reproducibility / supplementary)
pathway_sets_out <- pathway_sets_raw %>%
  distinct(Name, pathway_id, pathway_label) %>%
  left_join(pathway_info %>% dplyr::select(pathway_id, pathway_source),
    by = "pathway_id"
  )
write_csv(
  pathway_sets_out,
  file.path(csv_path, "batch2_pathway_metabolite_mapping.csv")
)

# Save MSEA results
if (!is.null(msea_gw)) {
  msea_gw_out <- msea_gw %>%
    left_join(pathway_source_lookup, by = c("pathway" = "pathway_label_src")) %>%
    mutate(leadingEdge = map_chr(leadingEdge, ~ paste(.x, collapse = ";")))
  write_csv(msea_gw_out, file.path(csv_path, "batch2_msea_gw.csv"))
}

if (!is.null(msea_sex)) {
  msea_sex_out <- msea_sex %>%
    left_join(pathway_source_lookup, by = c("pathway" = "pathway_label_src")) %>%
    mutate(leadingEdge = map_chr(leadingEdge, ~ paste(.x, collapse = ";")))
  write_csv(msea_sex_out, file.path(csv_path, "batch2_msea_sex.csv"))
}

cat("Files saved to:", csv_path, "\n")
```
