---
title: "Batch 2 Metabolomics Filtering"
author: "Sneha M. Chaturvedi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
---


# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Packages <- c("tidyverse", "readxl", "stats", "ggfortify", "ggrepel", "svglite", "pheatmap", "pracma")

lapply(Packages, library, character.only = TRUE)
```

```{r source_function}
# Source Functions
source("functions.R")
source("filtering_functions.R")
```

```{r set_path}
png_path <- "../../results/untargeted/batch2_filtering"
csv_path <- "../../results/untargeted/batch2_filtering/csv"
svg_path <- "../../results/untargeted/batch2_filtering/svg"
dir.create(png_path, recursive = TRUE, showWarnings = FALSE)
dir.create(svg_path, recursive = TRUE, showWarnings = FALSE)
dir.create(csv_path, recursive = TRUE, showWarnings = FALSE)
```
## Load Dataset

```{r LoadDataset, echo=TRUE}
# Peak Area Values
full_dataset <- read_excel("../../doc/untargeted/combined_annotations_untargeted_developing_2026.xlsx", sheet = "Area", range = NULL)

# Metabolite Annotations
compound_annotation <- read_excel("../../doc/untargeted/combined_annotations_untargeted_developing_2026.xlsx", sheet = "Annotation", range = NULL)

# Sample Metadata
sample_metadata <- read_excel("../../doc/untargeted/24_metadata.xlsx", sheet = "metadata", range = NULL)
# Rename SampleName to Sample in metadata if needed
sample_metadata <- sample_metadata %>%
  rename(Sample = SampleName)
```

## Format Datafile
```{r data_format, echo=TRUE}
#Remove Batch1 From Data
full_dataset <- cbind(full_dataset[1:3], full_dataset[,(12:35)])

# Clean the data
clean_metabolomics_data(full_dataset, sample_metadata, compound_annotation, dataset_name = "batch2_Untargeted_Raw")
```

# Outlier Removal

## Remove Sample Outliers
Remove Samples with outlier mean metabolite intensity defined by z score threshold
```{r outlier_removal_1, echo=TRUE}
# Remove samples with outlier mean metabolite intensity
outlier_removal_1_results <- remove_intensity_outliers(
  data = batch2_Untargeted_Raw_peak_data,
  metadata = sample_metadata,
  threshold = 3,
  save_plot = TRUE,
  output_prefix = "01_intensity_outliers",
  png_path = png_path,
  csv_path = csv_path
)

# Extract filtered data
outlier_removal_1 <- outlier_removal_1_results$filtered_data

# Save filtered dataset
saveRDS(outlier_removal_1, file.path(csv_path, "outlier_removal_1.rds"))
write.csv(outlier_removal_1, file.path(csv_path, "outlier_removal_1.csv"), row.names = FALSE)
```

Remove Samples with outlier number of detected peaks 
with data file called outlier_removal_2

```{r outlier_removal_2, echo=TRUE}
# Remove samples with outlier number of detected peaks
outlier_removal_2_results <- remove_detection_outliers(
  data = outlier_removal_1, # Use output from step 1
  metadata = sample_metadata,
  threshold = 3,
  na_threshold = 0,
  save_plot = TRUE,
  output_prefix = "02_detection_outliers",
  png_path = png_path,
  csv_path = csv_path
)

# Extract filtered data
outlier_removal_2 <- outlier_removal_2_results$filtered_data

# Save filtered dataset
saveRDS(outlier_removal_2, file.path(csv_path, "outlier_removal_2.rds"))
write.csv(outlier_removal_2, file.path(csv_path, "outlier_removal_2.csv"), row.names = FALSE)
```

#Metabolite Filtering


## Remove Unannotated Metabolites
Create new dataframe with only annotated metabolites
```{r Dataframe_Split}
batch2_annotated_dataset <- outlier_removal_2[!is.na(outlier_removal_2$Name) & outlier_removal_2$Name != "", ]
batch2_unannotated_dataset <- outlier_removal_2[is.na(outlier_removal_2$Name) | outlier_removal_2$Name == "", ]
```

## Duplicate Check
```{r first_duplicate_check}
# Full duplicate detection with CSV output
dup_results <- detect_metabolite_duplicates(
  dataset = batch2_annotated_dataset,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation,
  name_column = "Name",
  csv_path = csv_path,
  verbose = TRUE
)
```

## Filter By Retention Time
Filter metabolites with a low retention time, found in compound annotations under RT[min]

Create visualizations to determine RT thresholds
```{r long_format}
batch2_annotated_dataset_long <- prepare_long_data(batch2_annotated_dataset, batch2_Untargeted_Raw_sample_metadata)
```
### Display Retention Time
```{r retention_time}
## RETENTION TIME ANALYSIS ##

# Compute RT thresholds here so they are available for pre-filter plot below
threshold_results <- detect_rt_threshold_from_density(
  compound_annotation = batch2_Untargeted_Raw_compound_annotation,
  png_path = "../../results/untargeted/batch2_filtering/",
  svg_path = "../../results/untargeted/batch2_filtering/svg",
  csv_path = "../../results/untargeted/batch2_filtering/csv"
)

# Calculate metabolite abundance
metabolite_abundance <- calculate_metabolite_abundance(
  dataset = batch2_annotated_dataset,
  sample_cols = 4:27,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

metabolite_abundance_by_sex <- calculate_metabolite_abundance_by_group(
  data_long = batch2_annotated_dataset_long,
  group_by = "Sex",
  dataset = batch2_annotated_dataset,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

metabolite_abundance_by_gw <- calculate_metabolite_abundance_by_group(
  data_long = batch2_annotated_dataset_long,
  group_by = "GW_bucket",
  dataset = batch2_annotated_dataset,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

# Add retention time to metabolite abundance data
metabolite_abundance_with_RT <- metabolite_abundance %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

# Distribution of retention time across samples - Histogram
RT_distribution_histogram <- ggplot(metabolite_abundance_with_RT, aes(x = RT)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time Across Metabolites",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  theme(plot.margin = margin(10, 10, 10, 10))

save_dual_format(RT_distribution_histogram, "RT_distribution_histogram", png_path = png_path, svg_path = svg_path)

# By Sex
metabolite_abundance_by_sex_with_RT <- metabolite_abundance_by_sex %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

RT_distribution_by_sex <- ggplot(metabolite_abundance_by_sex_with_RT, aes(x = RT, fill = Sex)) +
  geom_histogram(binwidth = 0.5, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time by Sex",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  facet_wrap(~Sex, ncol = 1)

save_dual_format(RT_distribution_by_sex, "RT_distribution_by_sex", png_path = png_path, svg_path = svg_path)

# By Gestational Week Bucket
metabolite_abundance_by_GW_with_RT <- metabolite_abundance_by_gw %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

RT_distribution_by_GW <- ggplot(metabolite_abundance_by_GW_with_RT, aes(x = RT, fill = GW_bucket)) +
  geom_histogram(binwidth = 0.5, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time by GW Bucket",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  facet_wrap(~GW_bucket, ncol = 1)

save_dual_format(RT_distribution_by_GW, "RT_distribution_by_GW", png_path = png_path, svg_path = svg_path)

# Time x Intensity Scatter plot per column for Super.Class
# Create dataset with metabolite with known Super.Class and extract column from Compound.ID
metabolite_RT_intensity_superclass <- metabolite_abundance_with_RT %>%
  filter(!is.na(Super.Class) & Super.Class != "" & !is.na(RT)) %>%
  mutate(log10 = log10(mean_intensity)) %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID)) %>% # Extract column from Compound.ID
  mutate(Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID))



RT_intensity_column_superclass <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~Column, scales = "free") +
  theme_minimal() +
  labs(
    title = "Retention Time vs Intensity by Column",
    x = "Retention Time (min)",
    y = "Log10 Mean Peak Intensity"
  ) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_superclass, "RT_intensity_column_superclass", png_path = png_path, svg_path = svg_path)

# Build per-facet threshold lines from density-detected thresholds
rt_thresholds_for_plot <- threshold_results$recommended_thresholds

RT_intensity_column_mode_superclass <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_vline(
    data = rt_thresholds_for_plot,
    aes(xintercept = RT_cutoff),
    color = "red", linetype = "dashed", linewidth = 0.8,
    inherit.aes = FALSE
  ) +
  facet_wrap(~ Column + Mode, scales = "free") +
  theme_minimal() +
  labs(
    title = "Retention Time vs Intensity by Column and Mode",
    subtitle = "Red dashed line = RT threshold",
    x = "Retention Time (min)",
    y = "Log10 Mean Peak Intensity"
  ) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_mode_superclass, "RT_intensity_column_mode_superclass", png_path = png_path, svg_path = svg_path)

# Zoomed version: 0-5 minutes with 30-second intervals
RT_intensity_column_mode_superclass_zoomed <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~ Column + Mode, scales = "free_y") + # Changed to free_y only, so x-axis is consistent
  scale_x_continuous(breaks = seq(0, 5, by = 0.5), limits = c(0, 5)) + # Add this line for zoom and tick marks
  theme_minimal() +
  labs(
    title = "Retention Time vs Intensity by Column and Mode (0-5 min)",
    x = "Retention Time (min)",
    y = "Log10 Mean Peak Intensity"
  ) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_mode_superclass_zoomed, "RT_intensity_column_mode_superclass_zoomed_0-5min", png_path = png_path, svg_path = svg_path)
```
### Display Column Composition
```{r column_composition}
## BAR GRAPHS: METABOLITE COUNTS BY CLASSIFICATION AND COLUMN-MODE ##

# 1. Super.Class counts by Column-Mode
superclass_counts <- metabolite_RT_intensity_superclass %>%
  group_by(Column, Mode, Super.Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

superclass_barplot <- ggplot(superclass_counts, aes(x = Super.Class, y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  facet_wrap(~Column, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Metabolite Count by Super.Class and Column-Mode",
    x = "Super.Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "bottom"
  )

save_dual_format(superclass_barplot, "metabolite_counts_by_superclass_column_mode", png_path = png_path, svg_path = svg_path)

# 2. Class counts by Column-Mode
# Need to use metabolite_abundance_with_RT since Class isn't in the superclass-filtered dataset
class_counts <- metabolite_abundance_with_RT %>%
  filter(!is.na(Class) & Class != "") %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode, Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

# show top 20 per column-mode
top_classes <- class_counts %>%
  group_by(Column, Mode) %>%
  slice_max(order_by = n_metabolites, n = 20) %>%
  ungroup()


class_barplot <- ggplot(top_classes, aes(x = reorder(Class, n_metabolites), y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Column, scales = "free") +
  coord_flip() + # Flip for readability
  theme_minimal() +
  labs(
    title = "Top 20 Classes by Metabolite Count (by Column-Mode)",
    x = "Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.y = element_text(size = 7),
    legend.position = "bottom"
  )

save_dual_format(class_barplot, "metabolite_counts_by_class_column_mode_top20", png_path = png_path, svg_path = svg_path)

# 3. Sub.Class counts by Column-Mode
subclass_counts <- metabolite_abundance_with_RT %>%
  filter(!is.na(Sub.Class) & Sub.Class != "") %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode, Sub.Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

# Sub.Class might have many categories, so let's show top 20 per column-mode
top_subclasses <- subclass_counts %>%
  group_by(Column, Mode) %>%
  slice_max(order_by = n_metabolites, n = 20) %>%
  ungroup()

subclass_barplot <- ggplot(top_subclasses, aes(x = reorder(Sub.Class, n_metabolites), y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Column, scales = "free") +
  coord_flip() + # Flip for readability
  theme_minimal() +
  labs(
    title = "Top 20 Sub.Classes by Metabolite Count (by Column-Mode)",
    x = "Sub.Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.y = element_text(size = 7),
    legend.position = "bottom"
  )

save_dual_format(subclass_barplot, "metabolite_counts_by_subclass_column_mode_top20", png_path = png_path, svg_path = svg_path)

# 4. Summary table by Column-Mode
summary_by_column_mode <- metabolite_abundance_with_RT %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode) %>%
  summarise(
    total_metabolites = n(),
    n_superclasses = n_distinct(Super.Class[!is.na(Super.Class) & Super.Class != ""]),
    n_classes = n_distinct(Class[!is.na(Class) & Class != ""]),
    n_subclasses = n_distinct(Sub.Class[!is.na(Sub.Class) & Sub.Class != ""]),
    .groups = "drop"
  )

print("Summary of metabolite diversity by Column-Mode:")
print(summary_by_column_mode)

# 5. Alternative: Stacked bar showing all column-modes side by side for each Super.Class
superclass_stacked <- ggplot(superclass_counts, aes(x = Column_Mode, y = n_metabolites, fill = Super.Class)) +
  geom_bar(stat = "identity", color = "black") +
  theme_minimal() +
  labs(
    title = "Metabolite Distribution by Super.Class across Column-Mode Combinations",
    x = "Column-Mode Combination",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

save_dual_format(superclass_stacked, "metabolite_counts_superclass_stacked_by_column_mode", png_path = png_path, svg_path = svg_path)
```
### Determine RT Threshold + Filter
```{r RT_threshold_determination}
# threshold_results already computed in retention_time chunk above
# View recommended thresholds
threshold_results$recommended_thresholds

# Apply to your dataset
filtering_results <- apply_density_detected_thresholds(
  dataset = batch2_annotated_dataset,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation,
  threshold_results = threshold_results
)

# Use filtered dataset
batch2_annotated_dataset_RT_filtered <- filtering_results$filtered_dataset

# create long version
prepare_long_data(batch2_annotated_dataset_RT_filtered, batch2_Untargeted_Raw_sample_metadata, data_name = "batch2_annotated_dataset_long_RT_filtered")
```
### Post RT Filter Duplicate Check
```{r post_RT_check}
check_duplicates(batch2_annotated_dataset_RT_filtered)
```

### Check RT Post Filtering
Re-check column composition after filtering:
```{r RT_distribution_filtered}
# Calculate metabolite abundance
metabolite_abundance <- calculate_metabolite_abundance(
  dataset = batch2_annotated_dataset_RT_filtered,
  sample_cols = 4:27,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

metabolite_abundance_by_sex <- calculate_metabolite_abundance_by_group(
  data_long = batch2_annotated_dataset_long_RT_filtered,
  group_by = "Sex",
  dataset = batch2_annotated_dataset_RT_filtered,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

metabolite_abundance_by_gw <- calculate_metabolite_abundance_by_group(
  data_long = batch2_annotated_dataset_long_RT_filtered,
  group_by = "GW_bucket",
  dataset = batch2_annotated_dataset_RT_filtered,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

# Add retention time to metabolite abundance data
metabolite_abundance_with_RT <- metabolite_abundance %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

# Distribution of retention time across samples - Histogram
RT_distribution_histogram <- ggplot(metabolite_abundance_with_RT, aes(x = RT)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time Across Metabolites",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  theme(plot.margin = margin(10, 10, 10, 10))

save_dual_format(RT_distribution_histogram, "Filtered_RT_distribution_histogram", png_path = png_path, svg_path = svg_path)

# By Sex
metabolite_abundance_by_sex_with_RT <- metabolite_abundance_by_sex %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

RT_distribution_by_sex <- ggplot(metabolite_abundance_by_sex_with_RT, aes(x = RT, fill = Sex)) +
  geom_histogram(binwidth = 0.5, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time by Sex",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  facet_wrap(~Sex, ncol = 1)

save_dual_format(RT_distribution_by_sex, "Filtered_RT_distribution_by_sex", png_path = png_path, svg_path = svg_path)

# By Gestational Week Bucket
metabolite_abundance_by_GW_with_RT <- metabolite_abundance_by_gw %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

RT_distribution_by_GW <- ggplot(metabolite_abundance_by_GW_with_RT, aes(x = RT, fill = GW_bucket)) +
  geom_histogram(binwidth = 0.5, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time by GW Bucket",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  facet_wrap(~GW_bucket, ncol = 1)

save_dual_format(RT_distribution_by_GW, "Filtered_RT_distribution_by_GW", png_path = png_path, svg_path = svg_path)

# Time x Intensity Scatter plot per column for Super.Class
# Create dataset with metabolite with known Super.Class and extract column from Compound.ID
metabolite_RT_intensity_superclass <- metabolite_abundance_with_RT %>%
  filter(!is.na(Super.Class) & Super.Class != "" & !is.na(RT)) %>%
  mutate(log10 = log10(mean_intensity)) %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID)) %>% # Extract column from Compound.ID
  mutate(Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID))



RT_intensity_column_superclass <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~Column, scales = "free") +
  theme_minimal() +
  labs(
    title = "Retention Time vs Intensity by Column",
    x = "Retention Time (min)",
    y = "Log10 Mean Peak Intensity"
  ) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_superclass, "Filtered_RT_intensity_column_superclass", png_path = png_path, svg_path = svg_path)

RT_intensity_column_mode_superclass <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~ Column + Mode, scales = "free") +
  theme_minimal() +
  labs(
    title = "Retention Time vs Intensity by Column and Mode",
    x = "Retention Time (min)",
    y = "Log10 Mean Peak Intensity"
  ) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_mode_superclass, "Filtered_RT_intensity_column_mode_superclass", png_path = png_path, svg_path = svg_path)
```
### Check Column Composition Post Filtering
```{r filtered_column_composition}
## BAR GRAPHS: METABOLITE COUNTS BY CLASSIFICATION AND COLUMN-MODE ##

# 1. Super.Class counts by Column-Mode
superclass_counts <- metabolite_RT_intensity_superclass %>%
  group_by(Column, Mode, Super.Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

superclass_barplot <- ggplot(superclass_counts, aes(x = Super.Class, y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  facet_wrap(~Column, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Metabolite Count by Super.Class and Column-Mode",
    x = "Super.Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "bottom"
  )

save_dual_format(superclass_barplot, "Filtered_metabolite_counts_by_superclass_column_mode", png_path = png_path, svg_path = svg_path)

# 2. Class counts by Column-Mode
# Need to use metabolite_abundance_with_RT since Class isn't in the superclass-filtered dataset
class_counts <- metabolite_abundance_with_RT %>%
  filter(!is.na(Class) & Class != "") %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode, Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

# show top 20 per column-mode
top_classes <- class_counts %>%
  group_by(Column, Mode) %>%
  slice_max(order_by = n_metabolites, n = 20) %>%
  ungroup()


class_barplot <- ggplot(top_classes, aes(x = reorder(Class, n_metabolites), y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Column, scales = "free") +
  coord_flip() + # Flip for readability
  theme_minimal() +
  labs(
    title = "Top 20 Classes by Metabolite Count (by Column-Mode)",
    x = "Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.y = element_text(size = 7),
    legend.position = "bottom"
  )

save_dual_format(class_barplot, "Filtered_metabolite_counts_by_class_column_mode_top20", png_path = png_path, svg_path = svg_path)

# 3. Sub.Class counts by Column-Mode
subclass_counts <- metabolite_abundance_with_RT %>%
  filter(!is.na(Sub.Class) & Sub.Class != "") %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode, Sub.Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

# Sub.Class might have many categories, so let's show top 20 per column-mode
top_subclasses <- subclass_counts %>%
  group_by(Column, Mode) %>%
  slice_max(order_by = n_metabolites, n = 20) %>%
  ungroup()

subclass_barplot <- ggplot(top_subclasses, aes(x = reorder(Sub.Class, n_metabolites), y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Column, scales = "free") +
  coord_flip() + # Flip for readability
  theme_minimal() +
  labs(
    title = "Top 20 Sub.Classes by Metabolite Count (by Column-Mode)",
    x = "Sub.Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.y = element_text(size = 7),
    legend.position = "bottom"
  )

save_dual_format(subclass_barplot, "Filtered_metabolite_counts_by_subclass_column_mode_top20", png_path = png_path, svg_path = svg_path)

# 4. Summary table by Column-Mode
summary_by_column_mode <- metabolite_abundance_with_RT %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode) %>%
  summarise(
    total_metabolites = n(),
    n_superclasses = n_distinct(Super.Class[!is.na(Super.Class) & Super.Class != ""]),
    n_classes = n_distinct(Class[!is.na(Class) & Class != ""]),
    n_subclasses = n_distinct(Sub.Class[!is.na(Sub.Class) & Sub.Class != ""]),
    .groups = "drop"
  )

print("Summary of metabolite diversity by Column-Mode:")
print(summary_by_column_mode)

# 5. Alternative: Stacked bar showing all column-modes side by side for each Super.Class
superclass_stacked <- ggplot(superclass_counts, aes(x = Column_Mode, y = n_metabolites, fill = Super.Class)) +
  geom_bar(stat = "identity", color = "black") +
  theme_minimal() +
  labs(
    title = "Metabolite Distribution by Super.Class across Column-Mode Combinations",
    x = "Column-Mode Combination",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

save_dual_format(superclass_stacked, "Filtered_metabolite_counts_superclass_stacked_by_column_mode", png_path = png_path, svg_path = svg_path)
```
## Filter by Metabolite Missingness
Filter out metabolites present in fewer than 50% of samples (threshold lowered to retain metabolites that may only be detected in one sex). 

```{r missing_filter}
# Filter metabolites
missingness_results <- filter_metabolites_by_missingness(
  dataset = batch2_annotated_dataset_RT_filtered,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation,
  detection_threshold = 0.50,
  verbose = TRUE
)

# Assign the filtered dataset
batch2_annotated_dataset_missingness_filtered <- missingness_results$filtered_dataset

# Plot with separate paths
p_missingness <- plot_missingness_distribution(
  missingness_stats = missingness_results$missingness_stats,
  detection_threshold = 0.50,
  png_path = png_path,
  svg_path = svg_path
)
```

### Post Missingness Filter Duplicate Check
```{r post_missing_check}
check_duplicates(batch2_annotated_dataset_missingness_filtered)
```

## Filter by Annotation Score
Inspect the distribution of annotation scores to determine a quality threshold. Metabolites below the chosen threshold will be removed before deduplication.

```{r annotation_score_filter}
# Join annotation scores to the missingness-filtered dataset
annotation_scores <- batch2_Untargeted_Raw_compound_annotation %>%
  select(Compound.ID, Best.Match, Library.Name) %>%
  distinct()

score_data <- batch2_annotated_dataset_missingness_filtered %>%
  left_join(annotation_scores, by = "Compound.ID")

# Histogram of annotation scores across all metabolites
annotation_score_histogram <- ggplot(score_data, aes(x = Best.Match)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black", na.rm = TRUE) +
  theme_minimal() +
  labs(
    title = "Distribution of Metabolite Annotation Scores",
    subtitle = "Use this histogram to choose a score threshold for filtering",
    x = "Best Match Score (50-100)",
    y = "Metabolite Count"
  ) +
  theme(plot.margin = margin(10, 10, 10, 10))

annotation_score_histogram
save_dual_format(annotation_score_histogram, "annotation_score_histogram", png_path = png_path, svg_path = svg_path)

cat("Annotation score summary:\n")
print(summary(score_data$Best.Match))
cat("Metabolites with missing score:", sum(is.na(score_data$Best.Match)), "\n")
```

```{r annotation_score_apply_threshold}
# SET YOUR THRESHOLD HERE based on the histogram above
annotation_score_threshold <- 80 # <-- Update this value after reviewing histogram

# Apply annotation score filter
batch2_annotated_dataset_score_filtered <- batch2_annotated_dataset_missingness_filtered %>%
  left_join(annotation_scores, by = "Compound.ID") %>%
  # LipidSearch compounds have no Best.Match score (different scoring system) — exempt only those.
  # mzCloud/mzCloud_auto compounds with NA scores are anomalous and should be removed.
  filter((is.na(Best.Match) & Library.Name == "LipidSearch") | Best.Match >= annotation_score_threshold) %>%
  select(-Best.Match, -Library.Name) # Remove temporary columns

cat("\n=== Annotation Score Filter Summary ===\n")
cat("Score threshold applied:", annotation_score_threshold, "\n")
cat("Metabolites before:", nrow(batch2_annotated_dataset_missingness_filtered), "\n")
cat("Metabolites after:", nrow(batch2_annotated_dataset_score_filtered), "\n")
cat("Metabolites removed:", nrow(batch2_annotated_dataset_missingness_filtered) - nrow(batch2_annotated_dataset_score_filtered), "\n")
cat("  of which mzCloud NaN-score (anomalous):",
    sum(score_data$Library.Name != "LipidSearch" & is.na(score_data$Best.Match), na.rm = TRUE), "\n")
```

### Post Annotation Score Filter Duplicate Check
```{r post_score_check}
check_duplicates(batch2_annotated_dataset_score_filtered)
```

## Remove Duplicates
Select metabolite with highest intensity across samples

```{r duplicate_removal}
dedup_results <- remove_duplicates_by_intensity(
  dataset = batch2_annotated_dataset_score_filtered,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation,
  name_column = "Name",
  verbose = TRUE
)

batch2_annotated_dataset_final <- dedup_results$filtered_dataset


# create long version
prepare_long_data(batch2_annotated_dataset_final, batch2_Untargeted_Raw_sample_metadata, data_name = "batch2_annotated_dataset_final_long")

# Quick check
quick_dedup(batch2_annotated_dataset_score_filtered,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)
```
### Check RT Post Filtering
Re-check column composition after filtering:
```{r final_RT_distribution_filtered}
# Calculate metabolite abundance
metabolite_abundance <- calculate_metabolite_abundance(
  dataset = batch2_annotated_dataset_final,
  sample_cols = 4:27,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

metabolite_abundance_by_sex <- calculate_metabolite_abundance_by_group(
  data_long = batch2_annotated_dataset_final_long,
  group_by = "Sex",
  dataset = batch2_annotated_dataset_final,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

metabolite_abundance_by_gw <- calculate_metabolite_abundance_by_group(
  data_long = batch2_annotated_dataset_final_long,
  group_by = "GW_bucket",
  dataset = batch2_annotated_dataset_final,
  compound_annotation = batch2_Untargeted_Raw_compound_annotation
)

# Add retention time to metabolite abundance data
metabolite_abundance_with_RT <- metabolite_abundance %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

# Distribution of retention time across samples - Histogram
RT_distribution_histogram <- ggplot(metabolite_abundance_with_RT, aes(x = RT)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time Across Metabolites",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  theme(plot.margin = margin(10, 10, 10, 10))

save_dual_format(RT_distribution_histogram, "Final_RT_distribution_histogram", png_path = png_path, svg_path = svg_path)

# By Sex
metabolite_abundance_by_sex_with_RT <- metabolite_abundance_by_sex %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

RT_distribution_by_sex <- ggplot(metabolite_abundance_by_sex_with_RT, aes(x = RT, fill = Sex)) +
  geom_histogram(binwidth = 0.5, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time by Sex",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  facet_wrap(~Sex, ncol = 1)

save_dual_format(RT_distribution_by_sex, "Final_RT_distribution_by_sex", png_path = png_path, svg_path = svg_path)

# By Gestational Week Bucket
metabolite_abundance_by_GW_with_RT <- metabolite_abundance_by_gw %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

RT_distribution_by_GW <- ggplot(metabolite_abundance_by_GW_with_RT, aes(x = RT, fill = GW_bucket)) +
  geom_histogram(binwidth = 0.5, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Distribution of Retention Time by GW Bucket",
    x = "Retention Time (min)",
    y = "Metabolite Count"
  ) +
  facet_wrap(~GW_bucket, ncol = 1)

save_dual_format(RT_distribution_by_GW, "Final_RT_distribution_by_GW", png_path = png_path, svg_path = svg_path)

# Time x Intensity Scatter plot per column for Super.Class
# Create dataset with metabolite with known Super.Class and extract column from Compound.ID
metabolite_RT_intensity_superclass <- metabolite_abundance_with_RT %>%
  filter(!is.na(Super.Class) & Super.Class != "" & !is.na(RT)) %>%
  mutate(log10 = log10(mean_intensity)) %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID)) %>% # Extract column from Compound.ID
  mutate(Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID))



RT_intensity_column_superclass <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~Column, scales = "free") +
  theme_minimal() +
  labs(
    title = "Retention Time vs Intensity by Column",
    x = "Retention Time (min)",
    y = "Log10 Mean Peak Intensity"
  ) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_superclass, "Final_RT_intensity_column_superclass", png_path = png_path, svg_path = svg_path)

RT_intensity_column_mode_superclass <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~ Column + Mode, scales = "free") +
  theme_minimal() +
  labs(
    title = "Retention Time vs Intensity by Column and Mode",
    x = "Retention Time (min)",
    y = "Log10 Mean Peak Intensity"
  ) +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_mode_superclass, "Final_RT_intensity_column_mode_superclass", png_path = png_path, svg_path = svg_path)
```
### Check Column Composition Post Filtering
```{r final_filtered_column_composition}
## BAR GRAPHS: METABOLITE COUNTS BY CLASSIFICATION AND COLUMN-MODE ##

# 1. Super.Class counts by Column-Mode
superclass_counts <- metabolite_RT_intensity_superclass %>%
  group_by(Column, Mode, Super.Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

superclass_barplot <- ggplot(superclass_counts, aes(x = Super.Class, y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  facet_wrap(~Column, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Metabolite Count by Super.Class and Column-Mode",
    x = "Super.Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "bottom"
  )

save_dual_format(superclass_barplot, "Final_metabolite_counts_by_superclass_column_mode", png_path = png_path, svg_path = svg_path)

# 2. Class counts by Column-Mode
# Need to use metabolite_abundance_with_RT since Class isn't in the superclass-filtered dataset
class_counts <- metabolite_abundance_with_RT %>%
  filter(!is.na(Class) & Class != "") %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode, Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

# show top 20 per column-mode
top_classes <- class_counts %>%
  group_by(Column, Mode) %>%
  slice_max(order_by = n_metabolites, n = 20) %>%
  ungroup()


class_barplot <- ggplot(top_classes, aes(x = reorder(Class, n_metabolites), y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Column, scales = "free") +
  coord_flip() + # Flip for readability
  theme_minimal() +
  labs(
    title = "Top 20 Classes by Metabolite Count (by Column-Mode)",
    x = "Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.y = element_text(size = 7),
    legend.position = "bottom"
  )

save_dual_format(class_barplot, "Final_metabolite_counts_by_class_column_mode_top20", png_path = png_path, svg_path = svg_path)

# 3. Sub.Class counts by Column-Mode
subclass_counts <- metabolite_abundance_with_RT %>%
  filter(!is.na(Sub.Class) & Sub.Class != "") %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode, Sub.Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

# Sub.Class might have many categories, so let's show top 20 per column-mode
top_subclasses <- subclass_counts %>%
  group_by(Column, Mode) %>%
  slice_max(order_by = n_metabolites, n = 20) %>%
  ungroup()

subclass_barplot <- ggplot(top_subclasses, aes(x = reorder(Sub.Class, n_metabolites), y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Column, scales = "free") +
  coord_flip() + # Flip for readability
  theme_minimal() +
  labs(
    title = "Top 20 Sub.Classes by Metabolite Count (by Column-Mode)",
    x = "Sub.Class",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.y = element_text(size = 7),
    legend.position = "bottom"
  )

save_dual_format(subclass_barplot, "Final_metabolite_counts_by_subclass_column_mode_top20", png_path = png_path, svg_path = svg_path)

# 4. Summary table by Column-Mode
summary_by_column_mode <- metabolite_abundance_with_RT %>%
  mutate(
    Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
    Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)
  ) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode) %>%
  summarise(
    total_metabolites = n(),
    n_superclasses = n_distinct(Super.Class[!is.na(Super.Class) & Super.Class != ""]),
    n_classes = n_distinct(Class[!is.na(Class) & Class != ""]),
    n_subclasses = n_distinct(Sub.Class[!is.na(Sub.Class) & Sub.Class != ""]),
    .groups = "drop"
  )

print("Summary of metabolite diversity by Column-Mode:")
print(summary_by_column_mode)

# 5. Alternative: Stacked bar showing all column-modes side by side for each Super.Class
superclass_stacked <- ggplot(superclass_counts, aes(x = Column_Mode, y = n_metabolites, fill = Super.Class)) +
  geom_bar(stat = "identity", color = "black") +
  theme_minimal() +
  labs(
    title = "Metabolite Distribution by Super.Class across Column-Mode Combinations",
    x = "Column-Mode Combination",
    y = "Number of Metabolites"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

save_dual_format(superclass_stacked, "Final_metabolite_counts_superclass_stacked_by_column_mode", png_path = png_path, svg_path = svg_path)
```
```{r save_files}
# Save final datasets to CSV

output_dir <- "../../results/untargeted"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

batch2_Untargeted_Raw_compound_annotation <- batch2_Untargeted_Raw_compound_annotation %>%
  mutate(Name = str_to_title(Name))

# Save peak area values
write.csv(batch2_annotated_dataset_final,
  file.path(output_dir, "batch2_peak_area_values.csv"),
  row.names = FALSE
)

# Save metabolite annotations (filtered to Compound.IDs in final peak area dataset)
write.csv(batch2_Untargeted_Raw_compound_annotation %>%
  filter(Compound.ID %in% batch2_annotated_dataset_final$Compound.ID),
  file.path(output_dir, "batch2_metabolite_annotations.csv"),
  row.names = FALSE
)

# Save sample metadata
write.csv(sample_metadata,
  file.path(output_dir, "batch2_sample_metadata.csv"),
  row.names = FALSE
)

cat("Files saved to:", output_dir, "\n")
```
