---
title: "Batch 2 Sex and GW Metabolomics Analysis (MetaboAnalystR)"
author: "Sneha Michaela Chaturvedi"
date: "Last Updated `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Install MetaboAnalystR if needed (run once):
# install.packages("devtools")
# devtools::install_github("xia-lab/MetaboAnalystR", build = TRUE, build_vignettes = FALSE)

Packages <- c("tidyverse", "ggrepel", "svglite", "MetaboAnalystR")
lapply(Packages, library, character.only = TRUE)
```

```{r source_functions, include=FALSE}
source("functions.R")
source("missingness_visualization_functions.R")
```

```{r set_paths, include=FALSE}
png_path <- "../../results/untargeted/batch2_clean"
svg_path  <- "../../results/untargeted/batch2_clean/svg"
csv_path  <- "../../results/untargeted/batch2_clean/csv"

# MetaboAnalystR writes output files to the working directory.
# We use a dedicated subdirectory so its outputs stay organised.
msa_dir <- file.path(png_path, "metaboanalyst")
dir.create(msa_dir, recursive = TRUE, showWarnings = FALSE)
```

---

## Load Datasets

```{r load_data, include=FALSE}
# Peak area values (metabolites x samples)
batch2_peak_area <- read_csv(
  '../../results/untargeted/Batch2_peak_area_values.csv',
  show_col_types = FALSE
)

# Metabolite annotations
batch2_compound_annotation <- read_csv(
  '../../results/untargeted/Batch2_metabolite_annotations.csv',
  show_col_types = FALSE
)

# Sample metadata
batch2_sample_metadata <- read_csv(
  '../../results/untargeted/Batch2_sample_metadata.csv',
  show_col_types = FALSE
)
```

---

# Prepare MetaboAnalystR Input CSVs

MetaboAnalystR reads data from CSV files on disk, so we write two sets of input files before running any analyses.

**Sex analysis** uses the `"stat"` module:  
- Format `"rowu"` / `"disc"`: row 1 = sample names, row 2 = group labels, rows 3+ = feature × sample values.

**GW analysis** uses the `"mf"` (metadata table) module:  
- A column-oriented data file (`"colu"`) plus a separate metadata CSV containing GW as a numeric column. MetaboAnalystR auto-detects numeric columns as continuous covariates and fits a limma linear model.

```{r prep_input_files}
# ── Align annotations to peak area table ────────────────────────────────────
compounds_in_peak <- unique(batch2_peak_area$Compound.ID)

annotation_filtered <- batch2_compound_annotation %>%
  filter(Compound.ID %in% compounds_in_peak) %>%
  distinct(Compound.ID, .keep_all = TRUE)

id_to_name <- annotation_filtered %>% select(Compound.ID, Name)

# Join metabolite names and keep only sample columns in metadata order
sample_cols <- batch2_sample_metadata %>%
  filter(Sample %in% colnames(batch2_peak_area)) %>%
  pull(Sample)

peak_named <- batch2_peak_area %>%
  left_join(id_to_name, by = "Compound.ID") %>%
  # batch2_peak_area already has a Name column, so left_join creates Name.x / Name.y
  mutate(FeatureName = coalesce(Name.y, Name.x, Compound.ID)) %>%
  select(FeatureName, all_of(sample_cols))

sex_meta <- batch2_sample_metadata %>%
  filter(Sample %in% sample_cols) %>%
  arrange(match(Sample, sample_cols))

# ── 1. Sex input (rowu / disc) ───────────────────────────────────────────────
# Row 1: "Label" | sample names
# Row 2: "Sex"   | group per sample
# Rows 3+: feature name | peak area values
label_row <- c("Label", sex_meta$Sample)
group_row  <- c("Sex",   sex_meta$Sex)

sex_input <- bind_rows(
  setNames(as.data.frame(t(label_row)), label_row),
  setNames(as.data.frame(t(group_row)), label_row),
  peak_named %>% rename(Label = FeatureName) %>%
    mutate(across(where(is.numeric), as.character))
)

sex_csv_path <- file.path(msa_dir, "batch2_sex_input.csv")
write.csv(sex_input, sex_csv_path, row.names = FALSE, quote = FALSE)

# ── 2. GW data file (colu) ───────────────────────────────────────────────────
# Features in rows, samples in columns; no group row needed (group info in
# the separate metadata file).
gw_data_csv_path <- file.path(msa_dir, "batch2_gw_data_input.csv")
write.csv(peak_named %>% rename(Label = FeatureName),
          gw_data_csv_path, row.names = FALSE, quote = FALSE)

# ── 3. GW metadata file ──────────────────────────────────────────────────────
# First column = sample names; GW is numeric (auto-detected as continuous).
gw_metadata <- sex_meta %>% select(Sample, GW, Sex)
gw_meta_csv_path <- file.path(msa_dir, "batch2_gw_metadata.csv")
write.csv(gw_metadata, gw_meta_csv_path, row.names = FALSE, quote = FALSE)

cat("Input files written to:", msa_dir, "\n")
cat("  Sex data :  batch2_sex_input.csv\n")
cat("  GW data  :  batch2_gw_data_input.csv\n")
cat("  GW meta  :  batch2_gw_metadata.csv\n")
```

---

# Sex Analysis — MetaboAnalystR `stat` Module

# MetaboAnalystR relies on a global `mSet` variable internally.
# We use `mSet` throughout each section, then save to a named object
# at the end so results are preserved when the GW section reinitialises `mSet`.

```{r sex_mset_init}
old_wd <- getwd()
setwd(msa_dir)  # MetaboAnalystR writes plots/tables to the working directory

# Pass default.dpi explicitly — fixes "promise already under evaluation" bug
# caused by default.dpi = default.dpi self-reference in recent MetaboAnalystR versions
mSet <- InitDataObjects("conc", "stat", FALSE, default.dpi = 72)
mSet <- Read.TextData(mSet, "batch2_sex_input.csv", "rowu", "disc")
mSet <- SanityCheckData(mSet)
# ReplaceMin() intentionally omitted — data is pre-filtered; NAs represent
# true non-detections and should not be imputed.
mSet <- PreparePrenormData(mSet)

# No normalization — data already filtered and clean
mSet <- Normalization(mSet,
                      rowNorm   = "NULL",
                      transNorm = "NULL",
                      scaleNorm = "NULL",
                      ref       = NULL,
                      ratio     = FALSE,
                      ratioNum  = 20)
```

## Univariate — Fold Change & Wilcoxon Test

```{r sex_univariate}
# Fold-change analysis (fc.thresh = 1.5 ratio; cmp.type = 0 → group2 / group1)
mSet <- FC.Anal(mSet, fc.thresh = 1.5, cmp.type = 0)

# Wilcoxon rank-sum (nonpar = TRUE); FDR-adjusted p-value threshold
mSet <- Ttests.Anal(mSet,
                    nonpar    = TRUE,
                    threshp   = 0.05,
                    paired    = FALSE,
                    equal.var = TRUE,
                    pvalType  = "fdr")

cat("\n=== Sex Differential Analysis Summary ===\n")
cat("Total metabolites tested :", nrow(mSet$dataSet$norm), "\n")
cat("Significant (FDR < 0.05) :", mSet$analSet$tt$sig.num, "\n")

if (length(mSet$analSet$tt$sig.mat) > 0 &&
    nrow(as.data.frame(mSet$analSet$tt$sig.mat)) > 0) {
  cat("\nTop significant metabolites:\n")
  print(head(as.data.frame(mSet$analSet$tt$sig.mat), 10))
}
```

## Volcano Plot

```{r sex_volcano}
mSet <- Volcano.Anal(mSet,
                     paired    = FALSE,
                     fcthresh  = 1.5,
                     cmp.type  = 0,
                     pval.type = FALSE,   # FALSE = use adjusted (FDR) p-value
                     pval      = 0.05,
                     bothSig   = TRUE,
                     pvalType  = "fdr")

# Build ggplot volcano for dual-format saving and consistent styling
volcano_data <- data.frame(
  Name        = names(mSet$analSet$volcano$p.log),
  neg_log10_p = mSet$analSet$volcano$p.log,
  logFC       = mSet$analSet$volcano$fc.log
) %>%
  mutate(
    pval_raw        = 10^(-neg_log10_p),
    FDR             = p.adjust(pval_raw, method = "fdr"),
    Significant_FDR = FDR < 0.05,
    Direction = case_when(
      Significant_FDR & logFC > 0 ~ "Higher in Males",
      Significant_FDR & logFC < 0 ~ "Higher in Females",
      TRUE ~ "Not Significant"
    )
  )

volcano_sex_FDR <- ggplot(volcano_data, aes(x = logFC, y = neg_log10_p)) +
  geom_point(aes(color = Significant_FDR), alpha = 0.6) +
  geom_text_repel(
    data = volcano_data %>% filter(Significant_FDR),
    aes(label = Name), size = 2.5, max.overlaps = 15
  ) +
  scale_color_manual(
    values = c("FALSE" = "grey70", "TRUE" = "red"),
    name   = "FDR < 0.05"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_minimal() +
  labs(
    x     = "Log2 Fold Change (Male / Female)",
    y     = "-log10(p-value)",
    title = "Sex Differential Analysis — Volcano Plot (Wilcoxon, FDR)"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(volcano_sex_FDR, "sex_volcano_FDR",
                 png_path = png_path, svg_path = svg_path)
print(volcano_sex_FDR)

# Print summary
cat("\nAmong FDR-significant metabolites:\n")
print(volcano_data %>% filter(Significant_FDR) %>% count(Direction))

cat("\n=== Top 10 Most Significant Metabolites ===\n")
print(
  volcano_data %>%
    arrange(pval_raw) %>%
    head(10) %>%
    select(Name, logFC, pval_raw, FDR, Direction)
)

write_csv(volcano_data,
          file.path(csv_path, "batch2_sex_differential_analysis.csv"))
cat("\nResults saved to:", file.path(csv_path, "batch2_sex_differential_analysis.csv"), "\n")
```

## Multivariate — PCA

```{r sex_pca}
mSet <- PCA.Anal(mSet)

mSet <- PlotPCA2DScore(mSet,
                       imgName    = "pca_score_sex_",
                       format     = "png",
                       dpi        = 150,
                       width      = NA,
                       pc1        = 1,
                       pc2        = 2,
                       reg        = 0.95,
                       show       = 1,
                       grey.scale = 0)

# ggplot version for dual-format saving
pca_scores_sex <- as.data.frame(mSet$analSet$pca$x[, 1:2]) %>%
  rownames_to_column("Sample") %>%
  left_join(batch2_sample_metadata %>% select(Sample, Sex, GW), by = "Sample")

pct_var <- round(
  mSet$analSet$pca$sdev^2 / sum(mSet$analSet$pca$sdev^2) * 100, 1
)

pca_sex_plot <- ggplot(pca_scores_sex, aes(x = PC1, y = PC2, color = Sex)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_text_repel(aes(label = Sample), size = 2.5, max.overlaps = 10) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  theme_minimal() +
  labs(
    x     = paste0("PC1 (", pct_var[1], "%)"),
    y     = paste0("PC2 (", pct_var[2], "%)"),
    title = "PCA Score Plot — Batch 2 by Sex"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(pca_sex_plot, "pca_score_sex",
                 png_path = png_path, svg_path = svg_path)
print(pca_sex_plot)
```

## Multivariate — PLS-DA

```{r sex_plsda}
mSet <- PLSDA.Anal(mSet, reg = TRUE)

mSet <- PlotPLSPairSummary(mSet,
                           imgName = "plsda_pair_sex_",
                           format  = "png",
                           dpi     = 150,
                           width   = NA,
                           pc.num  = 5)

mSet <- PlotPLS2DScore(mSet,
                       imgName    = "plsda_score_sex_",
                       format     = "png",
                       dpi        = 150,
                       width      = NA,
                       comp1      = 1,
                       comp2      = 2,
                       reg        = 0.95,
                       show       = 1,
                       grey.scale = 0)

mSet <- PlotPLSLoading(mSet,
                       imgName = "plsda_loading_sex_",
                       format  = "png",
                       dpi     = 150,
                       width   = NA,
                       comp1   = 1,
                       comp2   = 2)

mSet <- PlotPLSDA.VIP(mSet,
                      imgName = "plsda_vip_sex_",
                      format  = "png",
                      dpi     = 150,
                      width   = NA)

# Extract and save VIP scores
vip_df <- as.data.frame(mSet$analSet$plsda$vip.mat) %>%
  rownames_to_column("Name") %>%
  arrange(desc(`Comp. 1`))

cat("\nTop 20 VIP scores (Comp 1):\n")
print(head(vip_df, 20))

write_csv(vip_df, file.path(csv_path, "batch2_sex_plsda_vip.csv"))

# Save sex mSet before GW section reinitialises it
mSet_sex <- mSet
```

```{r sex_restore_wd}
setwd(old_wd)
```

---

# GW Analysis — MetaboAnalystR Metadata Table (`mf`) Module

For a continuous covariate (gestational weeks), we use the **metadata table**
module, which fits a limma linear model with GW as a numeric predictor.
MetaboAnalystR auto-detects numeric metadata columns as continuous variables.

```{r gw_mset_init}
setwd(msa_dir)

# Pass default.dpi explicitly — same fix as sex section
mSet <- InitDataObjects("conc", "mf", FALSE, default.dpi = 72)

# Column-oriented data: features in rows, samples in columns
mSet <- Read.TextDataTs(mSet, "batch2_gw_data_input.csv", "colu")

# Metadata file: sample names + GW (numeric) + Sex (categorical)
mSet <- ReadMetaData(mSet, "batch2_gw_metadata.csv")

mSet <- SanityCheckData(mSet)
mSet <- SanityCheckMeta(mSet, init = TRUE)
# ReplaceMin() intentionally omitted — data is pre-filtered; NAs represent
# true non-detections and should not be imputed.
mSet <- PreparePrenormData(mSet)

# No normalization — data already clean
mSet <- Normalization(mSet,
                      rowNorm   = "NULL",
                      transNorm = "NULL",
                      scaleNorm = "NULL",
                      ref       = NULL,
                      ratio     = FALSE,
                      ratioNum  = 20)
```

## GW Limma Linear Model

```{r gw_limma}
# Fit limma linear model: feature abundance ~ GW (continuous)
mSet <- PerformCovariateLinearModel(mSet,
                                    analysis.var = "GW",
                                    meta         = "NA",
                                    contrast.cls = "NA",
                                    equal.var    = TRUE)

# Significant features (FDR < 0.05)
gw_sig <- as.data.frame(mSet$analSet$cov$sig.mat)

cat("\n=== GW Limma Analysis Summary ===\n")
cat("Total metabolites tested  :", nrow(mSet$dataSet$norm), "\n")
cat("Significant (FDR < 0.05) :", nrow(gw_sig), "\n")

if (nrow(gw_sig) > 0) {
  cat("\nTop significant metabolites:\n")
  print(head(gw_sig, 10))
}
```

## GW Results — Lollipop Plot

```{r gw_plot}
# Full stats table from limma output
gw_all_stats <- as.data.frame(mSet$analSet$cov$stats) %>%
  rownames_to_column("Name") %>%
  rename(
    logFC   = logFC,
    p_value = P.Value,
    FDR     = adj.P.Val
  ) %>%
  mutate(
    Significant_FDR = FDR < 0.05,
    Direction = case_when(
      Significant_FDR & logFC > 0 ~ "Positive with GW",
      Significant_FDR & logFC < 0 ~ "Negative with GW",
      TRUE ~ "Not Significant"
    )
  ) %>%
  arrange(FDR)

sig_gw <- gw_all_stats %>% filter(Significant_FDR)

scatter_GW <- ggplot(sig_gw,
                     aes(x = reorder(Name, logFC), y = logFC,
                         color = Direction)) +
  geom_segment(aes(xend = reorder(Name, logFC), yend = 0),
               color = "grey70") +
  geom_point(aes(size = -log10(p_value)), alpha = 0.85) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  coord_flip() +
  scale_color_manual(values = c("Positive with GW" = "#E76F51",
                                "Negative with GW" = "#457B9D")) +
  theme_minimal() +
  labs(
    x     = "Metabolite",
    y     = "Effect Size (limma logFC per GW unit)",
    size  = "-log10(p-value)",
    title = "Metabolites Significantly Associated with Gestational Week (GW)"
  ) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.y = element_text(size = 7))

save_dual_format(scatter_GW, "significant_GW",
                 png_path = png_path, svg_path = svg_path)
print(scatter_GW)

cat("\n=== Top 10 Metabolites Associated with GW ===\n")
print(head(gw_all_stats %>%
             select(Name, logFC, p_value, FDR, Direction), 10))

write_csv(gw_all_stats,
          file.path(csv_path, "batch2_GW_limma_analysis.csv"))
cat("\nResults saved to:", file.path(csv_path, "batch2_GW_limma_analysis.csv"), "\n")
```

## GW — PCA Colored by GW

```{r gw_pca}
mSet <- PCA.Anal(mSet)

pca_scores_gw <- as.data.frame(mSet$analSet$pca$x[, 1:2]) %>%
  rownames_to_column("Sample") %>%
  left_join(batch2_sample_metadata %>% select(Sample, GW, Sex), by = "Sample")

pct_var_gw <- round(
  mSet$analSet$pca$sdev^2 / sum(mSet$analSet$pca$sdev^2) * 100, 1
)

pca_gw_plot <- ggplot(pca_scores_gw,
                      aes(x = PC1, y = PC2, color = GW)) +
  geom_point(aes(shape = Sex), size = 3, alpha = 0.9) +
  geom_text_repel(aes(label = Sample), size = 2.5, max.overlaps = 10,
                  color = "grey30") +
  scale_color_viridis_c(option = "plasma", name = "Gestational\nWeek") +
  theme_minimal() +
  labs(
    x     = paste0("PC1 (", pct_var_gw[1], "%)"),
    y     = paste0("PC2 (", pct_var_gw[2], "%)"),
    title = "PCA Score Plot — Batch 2 Colored by GW"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

save_dual_format(pca_gw_plot, "pca_score_GW",
                 png_path = png_path, svg_path = svg_path)
print(pca_gw_plot)

# Save GW mSet for reference
mSet_gw <- mSet

setwd(old_wd)
```

---

# Session Info

```{r session_info}
sessionInfo()
```
