---
title: "Batch 2 Metabolite-Metabolite Correlation Analysis"
author: "Sneha Michaela Chaturvedi"
date: "Last Updated `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
---

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Packages <- c("readxl", "stats", "ggfortify", "ggrepel", "svglite", "Biobase",
              "ComplexHeatmap", "limma", "tidyverse", "ropls", "cowplot",
              "UpSetR", "circlize", "grid")

lapply(Packages, library, character.only = TRUE)
```

```{r source_functions, include=FALSE}
source("functions.R")
source("missingness_visualization_functions.R")
```

```{r set_path, include=FALSE}
png_path <- "../../results/untargeted/batch2_clean"
svg_path <- "../../results/untargeted/batch2_clean/svg"
csv_path <- "../../results/untargeted/batch2_clean/csv"
```

```{r heatmap_save_helper, include=FALSE}
# ComplexHeatmap requires png()/svglite() + draw() + dev.off()
# save_dual_format uses ggsave and does not support ComplexHeatmap objects.
save_heatmap <- function(ht, filename, png_path = NULL, svg_path = NULL,
                         width = 10, height = 10) {
  if (!is.null(png_path)) {
    png(file.path(png_path, paste0(filename, ".png")),
        width = width, height = height, units = "in", res = 300, bg = "white")
    draw(ht)
    dev.off()
  }
  if (!is.null(svg_path)) {
    svglite::svglite(file.path(svg_path, paste0(filename, ".svg")),
                     width = width, height = height, bg = "white")
    draw(ht)
    dev.off()
  }
}
```

## Load Dataset(s)

```{r LoadDataset, include=FALSE}
# Peak Area Values
batch2_peak_area <- read_csv('../../results/untargeted/Batch2_peak_area_values.csv',
                             show_col_types = FALSE)

# Metabolite Annotations
batch2_compound_annotation <- read_csv('../../results/untargeted/Batch2_metabolite_annotations.csv',
                                       show_col_types = FALSE)

# Sample Metadata
batch2_sample_metadata <- read_csv('../../results/untargeted/Batch2_sample_metadata.csv',
                                   show_col_types = FALSE)
```

```{r prep_dataframe}
# Convert metabolites data to matrix format
final_matrix <- batch2_peak_area %>%
  dplyr::select(starts_with("Sample"))

# Filter annotation to only include compounds present in peak area
compounds_in_peak <- unique(batch2_peak_area$Compound.ID)

batch2_compound_annotation <- batch2_compound_annotation %>%
  filter(Compound.ID %in% compounds_in_peak) %>%
  arrange(match(Name, rownames(final_matrix)))

metabolite_names <- batch2_compound_annotation$Name
final_matrix <- apply(final_matrix, 2, as.numeric)
rownames(final_matrix) <- metabolite_names

# Filter metadata to match metabolite samples
batch2_sample_metadata <- batch2_sample_metadata %>%
  filter(Sample %in% colnames(final_matrix)) %>%
  arrange(match(Sample, colnames(final_matrix)))

# Create the metadata dataframe
sample_metadata_df <- as.data.frame(batch2_sample_metadata)
rownames(sample_metadata_df) <- sample_metadata_df$Sample

# Create Biobase ExpressionSet
sample_phenoData    <- AnnotatedDataFrame(sample_metadata_df)
metabolite_metadata_df <- as.data.frame(batch2_compound_annotation)
rownames(metabolite_metadata_df) <- metabolite_metadata_df$Name
metabolite_featureData <- AnnotatedDataFrame(metabolite_metadata_df)

batch2_metabolomics <- ExpressionSet(
  assayData   = final_matrix,
  phenoData   = sample_phenoData,
  featureData = metabolite_featureData
)
```

```{r log_transform}
# Log2-transform peak intensities (log2(x + 1) to handle any zeros)
# NAs are retained - no imputation outside of PCA/PLS-DA
exprs(batch2_metabolomics) <- log2(exprs(batch2_metabolomics) + 1)

cat("Log2 transformation applied.\n")
cat("Intensity range after transformation:", range(exprs(batch2_metabolomics), na.rm = TRUE), "\n")
```

---

# Correlation Matrix

```{r compute_correlation}
# Correlation computed on log2-transformed, non-imputed data.
# pairwise.complete.obs: each metabolite pair uses all samples where both are non-NA.
# The resulting correlation matrix is always complete (no NAs from the raw data),
# UNLESS two metabolites have zero overlapping non-NA samples.
cor_matrix <- cor(t(exprs(batch2_metabolomics)), use = "pairwise.complete.obs")

# Replace any NA correlations (zero overlapping samples between a pair) with 0
# before clustering. This does not affect heatmap color for non-NA pairs.
n_na_cors <- sum(is.na(cor_matrix))
if (n_na_cors > 0) {
  cat("WARNING:", n_na_cors, "NA correlations found (zero overlapping samples).",
      "Replacing with 0 for clustering.\n")
  cor_matrix[is.na(cor_matrix)] <- 0
} else {
  cat("No NA correlations - all metabolite pairs have overlapping samples.\n")
}

cat("Correlation matrix dimensions:", dim(cor_matrix), "\n")

# Save full correlation matrix
write_csv(as.data.frame(cor_matrix) %>% rownames_to_column("Metabolite"),
          file.path(csv_path, "batch2_metabolite_correlation_matrix.csv"))
cat("Full correlation matrix saved.\n")
```

```{r color_scale, include=FALSE}
# Shared diverging color scale: blue (negative) -> white (zero) -> red (positive)
cor_col_fun <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))
```

---

# 1. Full Heatmap (All 847 Metabolites)

With 847 metabolites, axis labels are suppressed — the heatmap shows global
correlation structure and cluster patterns only.

```{r full_heatmap, fig.width=14, fig.height=13}
# Suppress row/column names at this scale; dendrograms convey structure
ht_full <- Heatmap(
  cor_matrix,
  name            = "Pearson r",
  col             = cor_col_fun,
  show_row_names  = FALSE,
  show_column_names = FALSE,
  show_row_dend   = TRUE,
  show_column_dend = TRUE,
  clustering_distance_rows    = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows      = "ward.D2",
  clustering_method_columns   = "ward.D2",
  heatmap_legend_param = list(title = "Pearson r", at = c(-1, -0.5, 0, 0.5, 1)),
  column_title    = paste0("Metabolite-Metabolite Correlations — All ",
                           nrow(cor_matrix), " Metabolites (Batch 2)"),
  column_title_gp = gpar(fontsize = 13, fontface = "bold"),
  width  = unit(10, "in"),
  height = unit(10, "in")
)

draw(ht_full)

save_heatmap(ht_full, "batch2_corr_full",
             png_path = png_path, svg_path = svg_path,
             width = 14, height = 13)
```

---

# 2. Super.Class Split Heatmaps

One heatmap per Super.Class. Within-class correlations only.
Plot dimensions scale with the number of metabolites in each class.

```{r superclass_heatmaps}
super_classes <- unique(fData(batch2_metabolomics)$Super.Class)
super_classes <- super_classes[!is.na(super_classes)]

cat("Super.Classes found:", length(super_classes), "\n")
print(table(fData(batch2_metabolomics)$Super.Class))
```

```{r superclass_heatmaps_plot, results='asis'}
for (sc in sort(super_classes)) {

  sc_metabolites <- fData(batch2_metabolomics) %>%
    filter(Super.Class == sc) %>%
    pull(Name)

  n <- length(sc_metabolites)
  cat("\n\n## Super.Class:", sc, "(n =", n, ")\n\n")

  if (n < 3) {
    cat("Fewer than 3 metabolites — skipping heatmap.\n")
    next
  }

  sc_cor <- cor_matrix[sc_metabolites, sc_metabolites]

  # Scale plot size with n; cap at 20in for very large classes
  plot_size <- min(20, max(5, n * 0.08))

  # Show labels only when n is small enough to be legible
  show_labels <- n <= 80
  label_size  <- if (n <= 40) 8 else 6

  # Sanitize class name for filename
  sc_filename <- paste0("batch2_corr_superclass_",
                        gsub("[^A-Za-z0-9]", "_", sc))

  ht_sc <- Heatmap(
    sc_cor,
    name              = "Pearson r",
    col               = cor_col_fun,
    show_row_names    = show_labels,
    show_column_names = show_labels,
    row_names_gp      = gpar(fontsize = label_size),
    column_names_gp   = gpar(fontsize = label_size),
    show_row_dend     = TRUE,
    show_column_dend  = TRUE,
    clustering_distance_rows    = "euclidean",
    clustering_distance_columns = "euclidean",
    clustering_method_rows      = "ward.D2",
    clustering_method_columns   = "ward.D2",
    heatmap_legend_param = list(title = "Pearson r", at = c(-1, -0.5, 0, 0.5, 1)),
    column_title    = paste0(sc, " (n = ", n, ")"),
    column_title_gp = gpar(fontsize = 11, fontface = "bold")
  )

  draw(ht_sc)

  save_heatmap(ht_sc, sc_filename,
               png_path = png_path, svg_path = svg_path,
               width = plot_size, height = plot_size)
}
```

---

# 3. Top 100 by Mean Abundance

Top 100 metabolites ranked by mean log2 peak area across samples.
High mean abundance = consistently well-detected metabolites.

```{r top100_mean}
mean_abundance <- apply(exprs(batch2_metabolomics), 1, mean, na.rm = TRUE)
top100_mean_names <- names(sort(mean_abundance, decreasing = TRUE))[1:100]

cat("Mean abundance range (top 100):",
    round(range(mean_abundance[top100_mean_names]), 2), "\n")

top100_mean_cor <- cor_matrix[top100_mean_names, top100_mean_names]

ht_mean <- Heatmap(
  top100_mean_cor,
  name              = "Pearson r",
  col               = cor_col_fun,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_gp      = gpar(fontsize = 5),
  column_names_gp   = gpar(fontsize = 5),
  show_row_dend     = TRUE,
  show_column_dend  = TRUE,
  clustering_distance_rows    = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows      = "ward.D2",
  clustering_method_columns   = "ward.D2",
  heatmap_legend_param = list(title = "Pearson r", at = c(-1, -0.5, 0, 0.5, 1)),
  column_title    = "Top 100 Metabolites by Mean Log2 Abundance (Batch 2)",
  column_title_gp = gpar(fontsize = 11, fontface = "bold")
)

draw(ht_mean)

save_heatmap(ht_mean, "batch2_corr_top100_mean",
             png_path = png_path, svg_path = svg_path,
             width = 14, height = 13)

# Save ranked list
mean_abundance_df <- data.frame(
  Name           = names(mean_abundance),
  Mean_log2_abundance = mean_abundance
) %>%
  arrange(desc(Mean_log2_abundance)) %>%
  left_join(fData(batch2_metabolomics) %>%
              dplyr::select(Name, Compound.ID, Super.Class, Class),
            by = "Name")

write_csv(mean_abundance_df, file.path(csv_path, "batch2_top100_mean_abundance.csv"))
cat("Mean abundance rankings saved.\n")
```

---

# 4. Top 100 by Coefficient of Variation (CV)

CV = SD / mean per metabolite (on log2-transformed values).
High CV = most sample-to-sample variable metabolites.

```{r top100_cv}
cv_values <- apply(exprs(batch2_metabolomics), 1, function(x) {
  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
})

top100_cv_names <- names(sort(cv_values, decreasing = TRUE))[1:100]

cat("CV range (top 100):", round(range(cv_values[top100_cv_names]), 3), "\n")

# Overlap with top 100 mean abundance
overlap_n <- sum(top100_cv_names %in% top100_mean_names)
cat("Overlap with top 100 mean abundance:", overlap_n, "metabolites\n")

top100_cv_cor <- cor_matrix[top100_cv_names, top100_cv_names]

ht_cv <- Heatmap(
  top100_cv_cor,
  name              = "Pearson r",
  col               = cor_col_fun,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_gp      = gpar(fontsize = 5),
  column_names_gp   = gpar(fontsize = 5),
  show_row_dend     = TRUE,
  show_column_dend  = TRUE,
  clustering_distance_rows    = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows      = "ward.D2",
  clustering_method_columns   = "ward.D2",
  heatmap_legend_param = list(title = "Pearson r", at = c(-1, -0.5, 0, 0.5, 1)),
  column_title    = "Top 100 Metabolites by Coefficient of Variation (Batch 2)",
  column_title_gp = gpar(fontsize = 11, fontface = "bold")
)

draw(ht_cv)

save_heatmap(ht_cv, "batch2_corr_top100_cv",
             png_path = png_path, svg_path = svg_path,
             width = 14, height = 13)

# Save ranked list
cv_df <- data.frame(
  Name = names(cv_values),
  CV   = cv_values
) %>%
  arrange(desc(CV)) %>%
  left_join(fData(batch2_metabolomics) %>%
              dplyr::select(Name, Compound.ID, Super.Class, Class),
            by = "Name")

write_csv(cv_df, file.path(csv_path, "batch2_top100_cv.csv"))
cat("CV rankings saved.\n")
```
