---
title: "Exogenous Compound Flagging"
author: "Sneha M. Chaturvedi"
date: "Last updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

Flag and remove exogenous compounds from the filtered metabolite annotation.
Instrument contaminants (plasticizers, slip agents, industrial chemicals), environmental exposures (pharmaceuticals, pesticides, dietary plant compounds) are removed from the main analysis but reported separately. 

Outputs: flagged annotation and clean peak area file.

| | |
|---|---|
| **Inputs** | `results/untargeted/metabolite_annotations.csv`, `results/untargeted/peak_area_values.csv` |
| **Outputs** | `results/untargeted/metabolite_annotations_flagged.csv`, `results/untargeted/peak_area_clean.csv` |
| **Upstream** | `Filtering.Rmd` |
| **Downstream** | `Clean_Visualization.Rmd`, all analysis scripts |

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


```{r source}
source("functions.R")

```

```{r set_paths}
root <- find_project_root()
annotation_path <- file.path(root, "results/untargeted/metabolite_annotations.csv")
output_csv_path <- file.path(root, "results/untargeted/csv")
output_path     <- file.path(root, "results/untargeted/metabolite_annotations_flagged.csv")
candidates_path <- file.path(output_csv_path, "candidate_exogenous_flags.csv")

dir.create(output_csv_path, recursive = TRUE, showWarnings = FALSE)
```

```{r load_annotation}
annotation <- read_csv(annotation_path, show_col_types = FALSE)
cat("Compounds loaded:", nrow(annotation), "\n")
```

---

# Flag Definitions

- **`instrument_contaminant`**: Artifacts thought to be from lab consumables, plastics, solvents, or industrial chemicals. 
- **`environmental_exposure`**: Pharmaceuticals, pesticides, dietary compounds, or synthetic hormones.
- **`review_needed`**: NIST14 spectral match entries where compound identity is uncertain. Resolved manually (see below).

Compounds with `NA` are presumed endogenous.

---

# Known Compounds

Compounds previously identified in project notes or resolved from manual NIST14 review.

```{r known_compounds}
known_instrument_contaminants <- c(
  # From project notes
  "erucamide",                    # C22:1 fatty acid amide; plastic slip agent
  "octinoxate",                   # UV filter / sunscreen agent
  "n,n'-diphenylguanidine",       # rubber accelerator / plasticizer additive
  "levoglucosan",                 # combustion product (biomass burning)
  "2-methylbenzoic acid",         # industrial solvent / toluene oxidation product
  "muscone",                      # synthetic fragrance (macrocyclic musk)
  # Resolved from NIST14 spectral match review
  "decaethylene glycol",          # PEG-10; synthetic polymer / lab consumable leachate
  "decanedioic acid, bis(2-ethylhexyl) ester",  # dioctyl sebacate; plasticizer
  "bis(2-ethylhexyl) sebacate",   # alias for above; same compound, different name
  "didodecyl 3,3'-thiodipropionate",              # plastic antioxidant / stabilizer
  "Irgafos 168",                  # plastic antioxidant / stabilizer
  "Victoria Pure Blue Bo",        # standard
  # Identified in metabolite-gene and metabolite-hormone review
  "methanesulfonic acid",         # industrial acid; common HPLC mobile-phase additive
  "meglumine",                    # N-methyl-D-glucamine; pharma excipient in contrast agents
  "triethyl citrate",             # synthetic citrate triester plasticizer/pharma excipient (≠ endogenous citrate)
  "dipropyleneglycol dibenzoate", # industrial plasticizer / coalescent solvent
  "peg n7",                       # polyethylene glycol oligomer; synthetic polymer leachate
  "peg n8",                       # polyethylene glycol oligomer; synthetic polymer leachate
  "peg n11",                      # polyethylene glycol oligomer; synthetic polymer leachate
  "peg n13",                      # polyethylene glycol oligomer; synthetic polymer leachate
  "peg n14",                      # polyethylene glycol oligomer; synthetic polymer leachate
  "6-chloro-5-methyl-1h-1,2,3-benzotriazole",              # chlorinated benzotriazole UV stabilizer / biocide
  "2-(2h-benzotriazol-2-yl)-4,6-bis(1-methyl-1-phenylethyl)phenol", # Tinuvin 234; UV absorber in plastics
  "octadecyl 3-(3,5-di-tert-butyl-4-hydroxyphenyl)propionate",      # Irganox 1076; polyolefin antioxidant
  "bis(2,2,6,6-tetramethyl-4-piperidyl)sebacate",                   # Tinuvin 770; hindered amine light stabilizer
  "bis(2-ethylhexyl) adipate",     # DEHA plasticizer
  # ── Audit round 3: remaining exogenous compounds in clean peak area ──
  # Plasticizers / solvents
  "citroflex 2",                   # triethyl citrate by trade name (chemical name already listed above)
  "citroflex 4",                   # tributyl citrate; plasticizer
  "citroflex a-4",                 # acetyltributyl citrate; plasticizer
  "acetyltributyl citrate",        # IUPAC name for Citroflex A-4 (duplicate entry in peak area)
  "tetraglyme",                    # tetraethylene glycol dimethyl ether; industrial solvent
  "triethylene glycol monobutyl ether",  # industrial solvent
  "triethyleneglycol bis(2-ethylhexanoate)", # synthetic plasticizer
  "triethyl phosphate",            # organophosphate flame retardant / plasticizer
  # Surfactants / detergents
  "4-dodecylbenzenesulfonic acid", # LAS anionic surfactant
  "dodecyl sulfate",               # SDS/SLS; lab and industrial detergent
  "cocamidopropyl betaine",        # synthetic amphoteric surfactant (cosmetics/lab)
  # UV filters / absorbers
  "octabenzone",                   # 2-hydroxy-4-octyloxybenzophenone; UV filter
  "dibenzoylmethane",              # UV absorber / chelating agent
  # Plastic additives
  "bis(4-ethylbenzylidene)sorbitol",  # DBS; plastic clarifying / nucleating agent
  "bis(methylbenzylidene)sorbitol",   # MDBS; same class
  "caprolactam",                   # epsilon-caprolactam; nylon-6 monomer
  # Polymer leachates
  "ppg n4",                        # polypropylene glycol oligomer; synthetic polymer leachate
  "ppg n5",                        # polypropylene glycol oligomer; synthetic polymer leachate
  # Lab reagents / buffers
  "trometamol",                    # tris(hydroxymethyl)aminomethane; laboratory buffer
  "\u03b1-cyano-3-hydroxycinnamic acid", # CHCA; MALDI matrix / MS standard
  # Industrial amines
  "cyclohexylamine",               # industrial amine / rubber accelerator metabolite
  "n-methylcyclohexylamine",       # industrial amine
  "3-isopropoxypropylamine",       # industrial ether-amine
  "3-(ethenyloxy)propan-1-amine",  # vinyl ether amine; industrial
  "n-methyl-2-phenylpropan-1-amine", # synthetic amine; no endogenous pathway
  # Fragrances / cosmetics
  "galaxolidone",                  # oxidation product of Galaxolide synthetic musk
  "triethanolamine",               # industrial/cosmetic tertiary amine
  "geranyllinalool"                # plant diterpene / fragrance compound; humans lack diterpene synthase
)

known_environmental_exposures <- c(
  "norgestrel",      # synthetic progestin (oral contraceptive)
  "methasterone",    # anabolic-androgenic synthetic steroid
  "minoxidil",       # vasodilator; topical/oral hair-loss or antihypertensive medication
  # Identified in metabolite-gene and metabolite-hormone review
  "ketamine",        # dissociative anesthetic; required for surgical tissue procurement
  "lidocaine",       # local anesthetic amide; required for surgical tissue procurement
  "midazolam",       # benzodiazepine sedative; required for surgical tissue procurement
  "fentanyl",        # synthetic opioid analgesic; required for surgical tissue procurement
  "piperine",         # alkaloid from black pepper (Piper nigrum); maternal dietary exposure
  # ── Audit round 3: remaining exogenous compounds in clean peak area ──
  # Pharmaceuticals
  "penicillin g",                  # beta-lactam antibiotic
  "allopurinol",                   # synthetic xanthine oxidase inhibitor (gout drug)
  "phthalylsulfathiazole",         # sulfonamide antibiotic (GI-targeted)
  "risedronic acid",               # bisphosphonate (osteoporosis drug)
  "6-aminocaproic acid",           # antifibrinolytic drug (Amicar); not a lysine catabolite
  "amifampridine",                 # 3,4-diaminopyridine; neuromuscular junction drug
  # Pesticides / herbicides
  "diethyl phosphate",             # organophosphate pesticide metabolite (not endogenous)
  "3-ketocarbofuran",              # carbofuran insecticide metabolite
  "propham",                       # isopropyl carbanilate; carbamate herbicide
  "6-chloropicolinic acid",        # clopyralid herbicide metabolite
  "maleic hydrazide",              # plant growth regulator / herbicide
  # Dietary plant compounds (no human biosynthetic pathway)
  "proline betaine",               # stachydrine; citrus-derived
  "solanidine",                    # steroidal alkaloid from Solanaceae (nightshades)
  "chalcone",                      # plant polyphenol; requires chalcone synthase (plant-only)
  "diosgenin",                     # steroidal sapogenin; plant-only biosynthesis (Dioscorea)
  "hypaphorine"                    # N,N,N-trimethyltryptophan; plant alkaloid (Erythrina)
)

cat("Known instrument contaminants:", length(known_instrument_contaminants), "\n")
cat("Known environmental exposures:", length(known_environmental_exposures), "\n")
```

---

# Keyword Patterns

```{r keyword_patterns}
# Instrument contaminants: plasticizers, industrial chemicals, slip agents,
# surfactants, PFAS, UV stabilizers, flame retardants, synthetic musks
instrument_contaminant_patterns <- paste(
  "phthalate", "diethylhexyl", "dioctyl", "dibutyl",
  "bisphenol",
  "tributyl phosphate", "triphenyl phosphate",
  "siloxane", "dimethylsilyl", "polydimethyl",
  "erucamide", "oleamide", "stearamide", "behenamide",
  "triton", "tween", "brij", "span [0-9]",
  "nonylphenol", "octylphenol",
  "perfluoro", "fluorotelomer",
  "benzophenone", "octinoxate", "avobenzone", "oxybenzone",
  "octocrylene", "homosalate",
  "tris(2-chloroethyl)", "hexabromocyclododecane",
  "levoglucosan", "mannosan", "galactosan",
  "diphenylguanidine","benzothiazole",
  "muscone", "galaxolide", "tonalide", "\\bmusk\\b", "Irgafos", "Pure Blue",
  "butylated hydroxytoluene", "\\bBHT\\b",
  "butylated hydroxyanisole", "\\bBHA\\b",
  # ── Audit round 3 patterns ──
  "citroflex",                     # catches all Citroflex variants (2, 4, A-4, A-2, etc.)
  "caprolactam",                   # nylon monomers
  "cocamidopropyl",                # amphoteric surfactant class
  "dodecylbenzenesulfonic",        # LAS surfactant class
  "\\bppg\\b",                     # polypropylene glycol oligomers
  "tetraglyme|triglyme",           # glyme solvents
  "trometamol|tromethamine",       # Tris buffer synonyms
  "triethanolamine",               # industrial/cosmetic amine
  "cyano.*hydroxycinnamic",        # MALDI matrix compounds (CHCA and variants)
  "galaxolidone",                  # Galaxolide oxidation product (not caught by "galaxolide" substring in all cases)
  sep = "|"
)

# Environmental exposures: synthetic steroids, pharmaceuticals, pesticides,
# preservatives, dietary stimulants.
# NOTE: levothyroxine/thyroxine (identical to endogenous T4) and progesterone
# (endogenous and biologically critical in fetal brain) are intentionally excluded.
environmental_exposure_patterns <- paste(
  # Synthetic hormones (contraceptives, anabolic steroids, SERMs)
  "norgestrel", "levonorgestrel", "medroxyprogesterone", "megestrol",
  "norethindrone", "gestodene", "desogestrel", "etonogestrel",
  "methasterone", "nandrolone", "stanozolol", "trenbolone",
  "boldenone", "oxandrolone", "methandrostenolone",
  "ethinylestradiol", "diethylstilbestrol", "\\bDES\\b",
  "tamoxifen", "clomiphene",
  # Synthetic glucocorticoids (distinct from endogenous cortisol/cortisone by name)
  "betamethasone", "dexamethasone", "\\bprednisone\\b", "prednisolone",
  # NSAIDs / analgesics
  "ibuprofen", "naproxen", "acetaminophen", "paracetamol",
  "\\baspirin\\b", "acetylsalicyl",
  # Cardiovascular / metabolic
  "metformin", "atorvastatin", "simvastatin", "lovastatin",
  "lisinopril", "amlodipine", "metoprolol",
  "labetalol", "hydralazine", "nifedipine", "methyldopa", "minoxidil",
  # CNS (SSRIs, SNRIs, benzodiazepines, antiepileptics)
  "fluoxetine", "sertraline", "paroxetine", "venlafaxine",
  "lorazepam", "diazepam", "alprazolam",
  "lamotrigine", "valproic acid", "\\bvalproate\\b", "valproex",
  # Antibiotics
  "amoxicillin", "azithromycin", "doxycycline",
  # Antiemetic
  "ondansetron",
  # Antihistamine
  "diphenhydramine",
  # Dietary stimulants
  "caffeine", "theobromine", "theophylline",
  # Pesticides
  "atrazine", "glyphosate", "chlorpyrifos", "permethrin",
  "cypermethrin", "deltamethrin", "malathion",
  # Preservatives
  "methylparaben", "ethylparaben", "propylparaben", "butylparaben",
  "\\bparaben\\b",
  # ── Audit round 3 patterns ──
  "penicillin",                    # beta-lactam antibiotics
  "allopurinol",                   # xanthine oxidase inhibitors
  "sulfathiazole",                 # sulfonamide antibiotics
  "risedronic|alendronic|zoledronic", # bisphosphonate class
  "aminocaproic",                  # antifibrinolytic drugs
  "carbofuran",                    # carbamate insecticide + metabolites (3-keto-, 3-hydroxy-)
  "chloropicolinic|clopyralid",    # picolinic acid herbicides
  "solanidine|\\bsolanine\\b",     # Solanaceae alkaloids
  "diosgenin",                     # plant steroidal sapogenins
  sep = "|"
)

# ClassyFire taxonomy classes that are predominantly synthetic/exogenous
exogenous_superclass <- c("Organohalogens")
exogenous_class      <- c(
  "Organochlorides", "Organobromides",
  "Chlorofluorocarbons", "Synthetic dyes", "Anabolic steroids"
)
```

---

# Apply Flags

```{r apply_flags}
# Compounds requiring manual spectrum review before a final decision.
review_needed_compounds <- c(
  "dl-p-hydroxyphenyllactic acid"  # High score (95.1) but is a hydroperoxide
                                    # inconsistent with the endogenous tyrosine catabolite
                                    # p-HPLA (HMDB0000755); no HMDB/CAS in record.
)

flag_df <- annotation %>%
  mutate(Name_lower = str_to_lower(Name)) %>%
  # Step 1: boolean flag columns
  mutate(
    flag_known_instrument    = Name_lower %in% known_instrument_contaminants,
    flag_known_exposure      = Name_lower %in% known_environmental_exposures,
    flag_review_needed       = Name_lower %in% review_needed_compounds,
    flag_pattern_instrument  = str_detect(Name_lower,
                                 regex(instrument_contaminant_patterns, ignore_case = TRUE)),
    flag_pattern_exposure    = str_detect(Name_lower,
                                 regex(environmental_exposure_patterns, ignore_case = TRUE)),
    flag_taxonomy_instrument = (Super.Class %in% exogenous_superclass) |
                               (Class %in% exogenous_class)
  ) %>%
  # Step 2: Exogenous_Flag
  # review_needed is checked first so it is not overridden by pattern matches
  mutate(
    Exogenous_Flag = case_when(
      flag_review_needed
        ~ "review_needed",
      flag_known_instrument | flag_pattern_instrument | flag_taxonomy_instrument
        ~ "instrument_contaminant",
      flag_known_exposure | flag_pattern_exposure
        ~ "environmental_exposure",
      TRUE ~ NA_character_
    )
  ) %>%
  # Step 3: Flag_Reason
  mutate(
    Flag_Reason = case_when(
      flag_review_needed       ~ "SMILES inconsistent with stated identity; manual spectrum review required before exclusion",
      flag_known_instrument    ~ "Known instrument contaminant (project notes or NIST14 review)",
      flag_known_exposure      ~ "Known synthetic pharmaceutical or steroid (project notes)",
      flag_pattern_instrument  ~ "Name matches industrial or contaminant keyword pattern",
      flag_pattern_exposure    ~ "Name matches pharmaceutical or pesticide keyword pattern",
      flag_taxonomy_instrument ~ "ClassyFire taxonomy flags synthetic or halogenated class",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::select(-Name_lower, -starts_with("flag_", ignore.case = FALSE))
```

---

# Summary

```{r summary}
annotation_flagged <- flag_df %>%
  # Annotation confidence tiers based on Best.Match score
  # Tiers: 80+ high, 70-79 medium, 60-69 low
  # High-confidence (80+) compounds are bolded in tables throughout downstream reports
  mutate(
    Annotation_Confidence = case_when(
      Best.Match >= 80 ~ "high",
      Best.Match >= 70 ~ "medium",
      Best.Match >= 60 ~ "low",
      TRUE             ~ NA_character_
    )
  )

cat("=== Exogenous Flagging Summary ===\n")
cat("Total compounds:", nrow(annotation_flagged), "\n")
cat("Presumed endogenous (NA):",
    sum(is.na(annotation_flagged$Exogenous_Flag)), "\n\n")

flag_counts <- annotation_flagged %>%
  filter(!is.na(Exogenous_Flag)) %>%
  count(Exogenous_Flag)
print(flag_counts)
```

## Instrument Contaminants

```{r show_instrument}
annotation_flagged %>%
  filter(Exogenous_Flag == "instrument_contaminant") %>%
  dplyr::select(Compound.ID, Name, Best.Match, Annotation_Confidence, Super.Class, Class, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Instrument Contaminants — excluded from all analyses. **Bold** = high confidence (score ≥ 80)")
```

## Environmental Exposures

```{r show_exposures}
annotation_flagged %>%
  filter(Exogenous_Flag == "environmental_exposure") %>%
  dplyr::select(Compound.ID, Name, Best.Match, Annotation_Confidence, Super.Class, Class, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Environmental Exposures — removed from main analysis. **Bold** = high confidence (score ≥ 80)")
```

## Remaining Review Needed

```{r show_review}
remaining_review <- annotation_flagged %>%
  filter(Exogenous_Flag == "review_needed")

if (nrow(remaining_review) == 0) {
  cat("No compounds remaining in review_needed.\n")
} else {
  remaining_review %>%
    dplyr::select(Compound.ID, Name, Super.Class, Class, Flag_Reason) %>%
    knitr::kable(caption = "Still requires manual review")
}
```

# Save Outputs

```{r save}
# Full annotation with Exogenous_Flag, Flag_Reason, and Annotation_Confidence columns
write_csv(annotation_flagged, output_path)
cat("Flagged annotation saved to:", output_path, "\n")

# Candidate review table (flagged only)
candidates <- annotation_flagged %>%
  filter(!is.na(Exogenous_Flag)) %>%
  dplyr::select(Compound.ID, Name, Super.Class, Class, Sub.Class,
         Best.Match, Annotation_Confidence,
         Exogenous_Flag, Flag_Reason) %>%
  arrange(Exogenous_Flag, Name)

write_csv(candidates, candidates_path)
cat("Candidate review table saved to:", candidates_path, "\n")
cat("Total flagged compounds:", nrow(candidates), "\n")
```

---

# Save Clean Peak Area Dataset

Removes all exogenous compounds (both `instrument_contaminant` and `environmental_exposure`) from the peak area matrix and saves `peak_area_clean.csv`. This is the file read by all downstream analysis scripts.

```{r save_clean_peak_area}
peak_area_path  <- file.path(root, "results/untargeted/peak_area_values.csv")
clean_peak_path <- file.path(root, "results/untargeted/peak_area_clean.csv")

peak_area <- read_csv(peak_area_path, show_col_types = FALSE)

exogenous_ids <- annotation_flagged %>%
  filter(Exogenous_Flag %in% c("instrument_contaminant", "environmental_exposure")) %>%
  pull(Compound.ID)

n_before <- nrow(peak_area)
peak_area_clean <- peak_area %>%
  filter(!Compound.ID %in% exogenous_ids)
n_after <- nrow(peak_area_clean)

cat("=== Clean Peak Area Dataset ===\n")
cat("Compounds before removal:", n_before, "\n")
cat("Exogenous removed:       ", n_before - n_after, "\n")
cat("  instrument_contaminant:", sum(annotation_flagged$Exogenous_Flag == "instrument_contaminant", na.rm = TRUE), "\n")
cat("  environmental_exposure:", sum(annotation_flagged$Exogenous_Flag == "environmental_exposure", na.rm = TRUE), "\n")
cat("  review_needed (retained):", sum(annotation_flagged$Exogenous_Flag == "review_needed", na.rm = TRUE), "\n")
cat("Compounds remaining:     ", n_after, "\n\n")

cat("--- Removed environmental exposures ---\n")
annotation_flagged %>%
  filter(Exogenous_Flag == "environmental_exposure") %>%
  dplyr::select(Compound.ID, Name, Best.Match, Annotation_Confidence, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Environmental exposures removed. **Bold** = high confidence (score >= 80)")

write_csv(peak_area_clean, clean_peak_path)
cat("Clean peak area saved to:", clean_peak_path, "\n")
```

# AI Assistance Disclosure

Code in this script was developed with assistance from Claude (Anthropic). All
AI-generated code was reviewed, validated, and adapted by the author.

# Session Info

```{r session-info}
sessionInfo()
```
