---
title: "Metabolomics_Filtering"
author: "Sneha M. Chaturvedi"
date: "Last Updated January 8, 2026"
output: 
  html_document:
    toc: TRUE
---

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
 
Packages <- c("tidyverse", "readxl", "stats", "ggfortify", "ggrepel", "svglite", "pheatmap")

lapply(Packages, library, character.only=TRUE)
```

```{r source_function}
#Source Functions
source("visualization_functions.R")
source("filtering_functions.R")
```

```{r set_path}
png_path <- "../../results/untargeted/Batch 2"
csv_path <- "../../results/untargeted/Batch 2"
svg_path <- "../../results/untargeted/Batch 2/svg"
```
## Load Dataset

```{r LoadDataset, echo=TRUE}
#Peak Area Values
full_dataset = read_excel('../../doc/untargeted/24_untargeted_developing_2025.xlsx',sheet="Area",range=NULL)

#Metabolite Annotations
compound_annotation = read_excel('../../doc/untargeted/24_untargeted_developing_2025.xlsx',sheet="Annotation", range=NULL)

#Sample Metadata
sample_metadata = read_excel('../../doc/untargeted/24_metadata.xlsx',sheet="metadata",range=NULL)
#Rename SampleName to Sample in metadata if needed
sample_metadata <- sample_metadata %>%
  rename(Sample = SampleName)
```

## Format Datafile
```{r data_format, echo=TRUE}
# Clean the data
Batch2_Untargeted_Raw <- clean_metabolomics_data(full_dataset, sample_metadata, compound_annotation, dataset_name = "Batch2_Untargeted_Raw")
```

# Outlier Removal

## Remove Sample Outliers
Remove Samples with outlier mean metabolite intensity defined by z score threshold
```{r outlier_removal_1, echo=TRUE}
# Remove samples with outlier mean metabolite intensity
outlier_removal_1_results <- remove_intensity_outliers(
  data = Batch2_Untargeted_Raw_peak_data,
  metadata = sample_metadata,
  threshold = 3,
  save_plot = TRUE,
  output_prefix = "01_intensity_outliers",
  png_path = png_path,
  csv_path = csv_path
)

# Extract filtered data
outlier_removal_1 <- outlier_removal_1_results$filtered_data

# Save filtered dataset
saveRDS(outlier_removal_1, file.path(csv_path, "outlier_removal_1.rds"))
write.csv(outlier_removal_1, file.path(csv_path, "outlier_removal_1.csv"), row.names = FALSE)
```

Remove Samples with outlier number of detected peaks 
with data file called outlier_removal_2

```{r outlier_removal_2, echo=TRUE}
# Remove samples with outlier number of detected peaks
outlier_removal_2_results <- remove_detection_outliers(
  data = outlier_removal_1,  # Use output from step 1
  metadata = sample_metadata,
  threshold = 3,
  na_threshold = 0,
  save_plot = TRUE,
  output_prefix = "02_detection_outliers",
  png_path = png_path,
  csv_path = csv_path
)

# Extract filtered data
outlier_removal_2 <- outlier_removal_2_results$filtered_data

# Save filtered dataset
saveRDS(outlier_removal_2, file.path(csv_path, "outlier_removal_2.rds"))
write.csv(outlier_removal_2, file.path(csv_path, "outlier_removal_2.csv"), row.names = FALSE)
```

#Metabolite Filtering


# Data Split
Create new dataframe with only annotated metabolites
```{r Dataframe_Split}
Batch2_annotated_dataset <- outlier_removal_2[!is.na(outlier_removal_2$Name) & outlier_removal_2$Name != "", ]
Batch2_unannotated_dataset <- outlier_removal_2[is.na(outlier_removal_2$Name) | outlier_removal_2$Name == "", ]
```

## Low Retention Time
Filter metabolites with a low retention time, found in compound annotations under RT[min]

### Retention Time Thresholds
Create visualizations to determine RT thresholds
```{r long_format}
Batch2_annotated_dataset_long <- prepare_long_data(Batch2_annotated_dataset, Batch2_Untargeted_Raw_sample_metadata)
```

```{r retention_time}
## RETENTION TIME ANALYSIS ##

#Calculate metabolite abundance
metabolite_abundance <- calculate_metabolite_abundance(
  dataset = Batch2_annotated_dataset,
  sample_cols = 4:27,
  compound_annotation = Batch2_Untargeted_Raw_compound_annotation
)

metabolite_abundance_by_sex <- calculate_metabolite_abundance_by_group(
  data_long = Batch2_annotated_dataset_long,
  group_by = "Sex",
  dataset = Batch2_annotated_dataset,
  compound_annotation = Batch2_Untargeted_Raw_compound_annotation
)

metabolite_abundance_by_gw <- calculate_metabolite_abundance_by_group(
  data_long = Batch2_annotated_dataset_long,
  group_by = "GW_bucket",
  dataset = Batch2_annotated_dataset,
  compound_annotation = Batch2_Untargeted_Raw_compound_annotation
)

# Add retention time to metabolite abundance data
metabolite_abundance_with_RT <- metabolite_abundance %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

# Distribution of retention time across samples - Histogram
RT_distribution_histogram <- ggplot(metabolite_abundance_with_RT, aes(x = RT)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Retention Time Across Metabolites",
       x = "Retention Time (min)",
       y = "Metabolite Count") +
  theme(plot.margin = margin(10, 10, 10, 10))

save_dual_format(RT_distribution_histogram, "RT_distribution_histogram", png_path = png_path, svg_path = svg_path)

# By Sex
metabolite_abundance_by_sex_with_RT <- metabolite_abundance_by_sex %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

RT_distribution_by_sex <- ggplot(metabolite_abundance_by_sex_with_RT, aes(x = RT, fill = Sex)) +
  geom_histogram(binwidth = 0.5, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(title = "Distribution of Retention Time by Sex",
       x = "Retention Time (min)",
       y = "Metabolite Count") +
  facet_wrap(~Sex, ncol = 1)

save_dual_format(RT_distribution_by_sex, "RT_distribution_by_sex", png_path = png_path, svg_path = svg_path)

# By Gestational Week Bucket
metabolite_abundance_by_GW_with_RT <- metabolite_abundance_by_gw %>%
  left_join(compound_annotation %>% select(Compound.ID, `RT[min]`), by = "Compound.ID") %>%
  rename(RT = `RT[min]`)

RT_distribution_by_GW <- ggplot(metabolite_abundance_by_GW_with_RT, aes(x = RT, fill = GW_bucket)) +
  geom_histogram(binwidth = 0.5, color = "black", position = "identity", alpha = 0.6) +
  theme_minimal() +
  labs(title = "Distribution of Retention Time by GW Bucket",
       x = "Retention Time (min)",
       y = "Metabolite Count") +
  facet_wrap(~GW_bucket, ncol = 1)

save_dual_format(RT_distribution_by_GW, "RT_distribution_by_GW", png_path = png_path, svg_path = svg_path)

# Time x Intensity Scatter plot per column for Super.Class
# Create dataset with metabolite with known Super.Class and extract column from Compound.ID
metabolite_RT_intensity_superclass <- metabolite_abundance_with_RT %>%
  filter(!is.na(Super.Class) & Super.Class != "" & !is.na(RT)) %>%
  mutate(log10 = log10(mean_intensity)) %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID))  %>% # Extract column from Compound.ID 
  mutate(Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID))



RT_intensity_column_superclass <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~Column, scales = "free") +
  theme_minimal() +
  labs(title = "Retention Time vs Intensity by Column",
       x = "Retention Time (min)",
       y = "Log10 Mean Peak Intensity") +
  theme(legend.position = "right",axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_superclass, "RT_intensity_column_superclass", png_path = png_path, svg_path = svg_path)

RT_intensity_column_mode_superclass <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~ Column + Mode, scales = "free") +
  theme_minimal() +
  labs(title = "Retention Time vs Intensity by Column and Mode",
       x = "Retention Time (min)",
       y = "Log10 Mean Peak Intensity") +
  theme(legend.position = "right",axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_mode_superclass, "RT_intensity_column_mode_superclass", png_path = png_path, svg_path = svg_path)

# Zoomed version: 0-5 minutes with 30-second intervals
RT_intensity_column_mode_superclass_zoomed <- ggplot(metabolite_RT_intensity_superclass, aes(x = RT, y = log10, color = Super.Class)) +
  geom_point(alpha = 0.6, size = 2) +
  facet_wrap(~ Column + Mode, scales = "free_y") +  # Changed to free_y only, so x-axis is consistent
  scale_x_continuous(breaks = seq(0, 5, by = 0.5), limits = c(0, 5)) +  # Add this line for zoom and tick marks
  theme_minimal() +
  labs(title = "Retention Time vs Intensity by Column and Mode (0-5 min)",
       x = "Retention Time (min)",
       y = "Log10 Mean Peak Intensity") +
  theme(legend.position = "right", axis.text.x = element_text(angle = 45, hjust = 1))

save_dual_format(RT_intensity_column_mode_superclass_zoomed, "RT_intensity_column_mode_superclass_zoomed_0-5min", png_path = png_path, svg_path = svg_path)

```

```{r column_composition}
## BAR GRAPHS: METABOLITE COUNTS BY CLASSIFICATION AND COLUMN-MODE ##

# 1. Super.Class counts by Column-Mode
superclass_counts <- metabolite_RT_intensity_superclass %>%
  group_by(Column, Mode, Super.Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

superclass_barplot <- ggplot(superclass_counts, aes(x = Super.Class, y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  facet_wrap(~Column, scales = "free_y") +
  theme_minimal() +
  labs(title = "Metabolite Count by Super.Class and Column-Mode",
       x = "Super.Class",
       y = "Number of Metabolites") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "bottom")

save_dual_format(superclass_barplot, "metabolite_counts_by_superclass_column_mode", png_path = png_path, svg_path = svg_path)

# 2. Class counts by Column-Mode
# Need to use metabolite_abundance_with_RT since Class isn't in the superclass-filtered dataset
class_counts <- metabolite_abundance_with_RT %>%
  filter(!is.na(Class) & Class != "") %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
         Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode, Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

# show top 20 per column-mode
top_classes <- class_counts %>%
  group_by(Column, Mode) %>%
  slice_max(order_by = n_metabolites, n = 20) %>%
  ungroup()


class_barplot <- ggplot(top_classes, aes(x = reorder(Class, n_metabolites), y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Column, scales = "free") +
  coord_flip() +  # Flip for readability
  theme_minimal() +
  labs(title = "Top 20 Classes by Metabolite Count (by Column-Mode)",
       x = "Class",
       y = "Number of Metabolites") +
  theme(axis.text.y = element_text(size = 7),
        legend.position = "bottom")

save_dual_format(class_barplot, "metabolite_counts_by_class_column_mode_top20", png_path = png_path, svg_path = svg_path)

# 3. Sub.Class counts by Column-Mode
subclass_counts <- metabolite_abundance_with_RT %>%
  filter(!is.na(Sub.Class) & Sub.Class != "") %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
         Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode, Sub.Class) %>%
  summarise(n_metabolites = n(), .groups = "drop") %>%
  mutate(Column_Mode = paste(Column, Mode, sep = "_"))

# Sub.Class might have many categories, so let's show top 20 per column-mode
top_subclasses <- subclass_counts %>%
  group_by(Column, Mode) %>%
  slice_max(order_by = n_metabolites, n = 20) %>%
  ungroup()

subclass_barplot <- ggplot(top_subclasses, aes(x = reorder(Sub.Class, n_metabolites), y = n_metabolites, fill = Mode)) +
  geom_bar(stat = "identity", color = "black") +
  facet_wrap(~Column, scales = "free") +
  coord_flip() +  # Flip for readability
  theme_minimal() +
  labs(title = "Top 20 Sub.Classes by Metabolite Count (by Column-Mode)",
       x = "Sub.Class",
       y = "Number of Metabolites") +
  theme(axis.text.y = element_text(size = 7),
        legend.position = "bottom")

save_dual_format(subclass_barplot, "metabolite_counts_by_subclass_column_mode_top20", png_path = png_path, svg_path = svg_path)

# 4. Summary table by Column-Mode
summary_by_column_mode <- metabolite_abundance_with_RT %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
         Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)) %>%
  filter(!is.na(Column) & !is.na(Mode)) %>%
  group_by(Column, Mode) %>%
  summarise(
    total_metabolites = n(),
    n_superclasses = n_distinct(Super.Class[!is.na(Super.Class) & Super.Class != ""]),
    n_classes = n_distinct(Class[!is.na(Class) & Class != ""]),
    n_subclasses = n_distinct(Sub.Class[!is.na(Sub.Class) & Sub.Class != ""]),
    .groups = "drop"
  )

print("Summary of metabolite diversity by Column-Mode:")
print(summary_by_column_mode)

# 5. Alternative: Stacked bar showing all column-modes side by side for each Super.Class
superclass_stacked <- ggplot(superclass_counts, aes(x = Column_Mode, y = n_metabolites, fill = Super.Class)) +
  geom_bar(stat = "identity", color = "black") +
  theme_minimal() +
  labs(title = "Metabolite Distribution by Super.Class across Column-Mode Combinations",
       x = "Column-Mode Combination",
       y = "Number of Metabolites") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

save_dual_format(superclass_stacked, "metabolite_counts_superclass_stacked_by_column_mode", png_path = png_path, svg_path = svg_path)
```
### Filter
```{r RT_filter}
## APPLY COLUMN-MODE SPECIFIC RT CUTOFFS TO SOURCE DATASET ##

# Define RT cutoffs by column-mode combination
rt_cutoffs <- data.frame(
  Column = c("Amide", "Amide", "C18", "C18", "C8", "C8"),
  Mode = c("Neg", "Pos", "Neg", "Pos", "Neg", "Pos"),
  RT_cutoff = c(1.5, 2.0, 1.75, 1.5, 1.75, 2.0)
)

print("RT cutoffs to be applied:")
print(rt_cutoffs)

# Get RT and Column/Mode info from compound annotation
compound_annotation_with_cutoff <- Batch2_Untargeted_Raw_compound_annotation %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
         Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)) %>%
  left_join(rt_cutoffs, by = c("Column", "Mode")) %>%
  filter(`RT[min]` >= RT_cutoff) %>%
  select(Compound.ID)  # Keep only Compound.ID for filtering

# Filter Batch2_annotated_dataset to only include compounds that pass RT cutoff
Batch2_annotated_dataset_RT_filtered <- Batch2_annotated_dataset %>%
  filter(Compound.ID %in% compound_annotation_with_cutoff$Compound.ID)

# Also filter the long format dataset
Batch2_annotated_dataset_long_RT_filtered <- Batch2_annotated_dataset_long %>%
  filter(Compound.ID %in% compound_annotation_with_cutoff$Compound.ID)

# Report filtering results
cat("\n=== RT FILTERING RESULTS ===\n")
cat("Before RT filtering:", nrow(Batch2_annotated_dataset), "compounds\n")
cat("After RT filtering:", nrow(Batch2_annotated_dataset_RT_filtered), "compounds\n")
cat("Compounds removed:", nrow(Batch2_annotated_dataset) - nrow(Batch2_annotated_dataset_RT_filtered), "\n")
cat("Percent retained:", round(100 * nrow(Batch2_annotated_dataset_RT_filtered) / nrow(Batch2_annotated_dataset), 1), "%\n")

# Detailed summary by column-mode
filtering_summary <- Batch2_Untargeted_Raw_compound_annotation %>%
  mutate(Column = sub("-[^-]+-[0-9]+$", "", Compound.ID),
         Mode = sub(".*-(.*?)-.*", "\\1", Compound.ID)) %>%
  left_join(rt_cutoffs, by = c("Column", "Mode")) %>%
  group_by(Column, Mode, RT_cutoff) %>%
  summarise(
    total_compounds = n(),
    below_cutoff = sum(`RT[min]` < RT_cutoff, na.rm = TRUE),
    above_cutoff = sum(`RT[min]` >= RT_cutoff, na.rm = TRUE),
    percent_retained = round(100 * above_cutoff / total_compounds, 1),
    .groups = "drop"
  )

cat("\nFiltering summary by Column-Mode:\n")
print(filtering_summary)

```


## Standards
```{r standards}
#Filter Batch2_Annotated_Dataset for Standards
# Define Mix 1 standard names
mix1_standards <- c(
  "L-Alanine",
  "1,4-Butanediamine",
  "Creatinine",
  "Ethanolamine.HCl",
  "Guanosine.2H2O",
  "Hypoxanthine",
  "L-Leucine",
  "L-Phenylalanine",
  "Thymine",
  "L-Tryptophan",
  "L-Tyrosine",
  "Vitamin B3"
)

# Filter the main dataset for Mix 1 standards
Batch2_annotated_dataset_mix1 <- Batch2_annotated_dataset %>%
  filter(Name %in% mix1_standards)

# Also filter the long format if you have it
Batch2_annotated_dataset_long_mix1 <- Batch2_annotated_dataset_long %>%
  filter(Name %in% mix1_standards)

# Check which standards were found
found_standards <- unique(Batch2_annotated_dataset_mix1$Name)
missing_standards <- setdiff(mix1_standards, found_standards)

cat("Found", length(found_standards), "out of", length(mix1_standards), "Mix 1 standards\n")
cat("\nFound standards:\n")
print(found_standards)

if(length(missing_standards) > 0) {
  cat("\nMissing standards (may need to search by Compound Formula):\n")
  print(missing_standards)
}

# Show summary of filtered dataset
cat("\nFiltered dataset dimensions:\n")
cat("Rows:", nrow(Batch2_annotated_dataset_mix1), "\n")
cat("Unique compounds:", length(unique(Batch2_annotated_dataset_mix1$Compound.ID)), "\n")
```
