---
title: "newbatch2 Exogenous Compound Flagging"
author: "Sneha M. Chaturvedi"
date: "Last Updated `r Sys.Date()`"
output:
  html_document:
    toc: true
---

Code developed by author with assistance from Claude (Sonnet 4.6). Output checked and validated by author.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r set_paths}
annotation_path <- "../../results/untargeted/newbatch2_metabolite_annotations.csv"
output_csv_path <- "../../results/untargeted/csv"
output_path     <- "../../results/untargeted/newbatch2_metabolite_annotations_flagged.csv"
candidates_path <- file.path(output_csv_path, "newbatch2_candidate_exogenous_flags.csv")

dir.create(output_csv_path, recursive = TRUE, showWarnings = FALSE)
```

```{r load_annotation}
annotation <- read_csv(annotation_path, show_col_types = FALSE)
cat("Compounds loaded:", nrow(annotation), "\n")
```

```{r source}
source("functions.R")
```
---

# Flag Definitions

Two flag levels are used:

- **`instrument_contaminant`**: Procedural artifacts from lab consumables, plastics, solvents, or industrial chemicals. These have no biological origin and are excluded from all downstream analyses.
- **`environmental_exposure`**: Pharmaceuticals, pesticides, dietary compounds, or synthetic hormones that may reflect genuine maternal/fetal exposure. These are retained in the main analysis but reported separately.
- **`review_needed`**: NIST14 spectral match entries where compound identity is uncertain. Resolved manually (see below).

Compounds with `NA` are presumed endogenous.

---

# Known Compounds

Compounds previously identified in project notes or resolved from manual NIST14 review.

```{r known_compounds}
known_instrument_contaminants <- c(
  # From project notes
  "erucamide",                    # C22:1 fatty acid amide; plastic slip agent
  "octinoxate",                   # UV filter / sunscreen agent
  "n,n'-diphenylguanidine",       # rubber accelerator / plasticizer additive
  "levoglucosan",                 # combustion product (biomass burning)
  "2-methylbenzoic acid",         # industrial solvent / toluene oxidation product
  "muscone",                      # synthetic fragrance (macrocyclic musk)
  # Resolved from NIST14 spectral match review
  "decaethylene glycol",          # PEG-10; synthetic polymer / lab consumable leachate
  "decanedioic acid, bis(2-ethylhexyl) ester",  # dioctyl sebacate; plasticizer
  "didodecyl 3,3'-thiodipropionate",              # plastic antioxidant / stabilizer
  "Irgafos 168", # plastic antioxidant / stabilizer
  "Victoria Pure Blue Bo" # standard 
)

known_environmental_exposures <- c(
  "norgestrel",      # synthetic progestin (oral contraceptive)
  "methasterone",    # anabolic-androgenic synthetic steroid
  "minoxidil"        # vasodilator; topical/oral hair-loss or antihypertensive medication
)

cat("Known instrument contaminants:", length(known_instrument_contaminants), "\n")
cat("Known environmental exposures:", length(known_environmental_exposures), "\n")
```

---

# Keyword Patterns

```{r keyword_patterns}
# Instrument contaminants: plasticizers, industrial chemicals, slip agents,
# surfactants, PFAS, UV stabilizers, flame retardants, synthetic musks
instrument_contaminant_patterns <- paste(
  "phthalate", "diethylhexyl", "dioctyl", "dibutyl",
  "bisphenol",
  "tributyl phosphate", "triphenyl phosphate",
  "siloxane", "dimethylsilyl", "polydimethyl",
  "erucamide", "oleamide", "stearamide", "behenamide",
  "triton", "tween", "brij", "span [0-9]",
  "nonylphenol", "octylphenol",
  "perfluoro", "fluorotelomer",
  "benzophenone", "octinoxate", "avobenzone", "oxybenzone",
  "octocrylene", "homosalate",
  "tris(2-chloroethyl)", "hexabromocyclododecane",
  "levoglucosan", "mannosan", "galactosan",
  "diphenylguanidine","benzothiazole",
  "muscone", "galaxolide", "tonalide", "\\bmusk\\b", "Irgafos", "Pure Blue",
  "butylated hydroxytoluene", "\\bBHT\\b",
  "butylated hydroxyanisole", "\\bBHA\\b",
  sep = "|"
)

# Environmental exposures: synthetic steroids, pharmaceuticals, pesticides,
# preservatives, dietary stimulants.
# NOTE: levothyroxine/thyroxine (identical to endogenous T4) and progesterone
# (endogenous and biologically critical in fetal brain) are intentionally excluded.
environmental_exposure_patterns <- paste(
  # Synthetic hormones (contraceptives, anabolic steroids, SERMs)
  "norgestrel", "levonorgestrel", "medroxyprogesterone", "megestrol",
  "norethindrone", "gestodene", "desogestrel", "etonogestrel",
  "methasterone", "nandrolone", "stanozolol", "trenbolone",
  "boldenone", "oxandrolone", "methandrostenolone",
  "ethinylestradiol", "diethylstilbestrol", "\\bDES\\b",
  "tamoxifen", "clomiphene",
  # Synthetic glucocorticoids (distinct from endogenous cortisol/cortisone by name)
  "betamethasone", "dexamethasone", "\\bprednisone\\b", "prednisolone",
  # NSAIDs / analgesics
  "ibuprofen", "naproxen", "acetaminophen", "paracetamol",
  "\\baspirin\\b", "acetylsalicyl",
  # Cardiovascular / metabolic
  "metformin", "atorvastatin", "simvastatin", "lovastatin",
  "lisinopril", "amlodipine", "metoprolol",
  "labetalol", "hydralazine", "nifedipine", "methyldopa", "minoxidil",
  # CNS (SSRIs, SNRIs, benzodiazepines, antiepileptics)
  "fluoxetine", "sertraline", "paroxetine", "venlafaxine",
  "lorazepam", "diazepam", "alprazolam",
  "lamotrigine", "valproic acid", "\\bvalproate\\b", "valproex",
  # Antibiotics
  "amoxicillin", "azithromycin", "doxycycline",
  # Antiemetic
  "ondansetron",
  # Antihistamine
  "diphenhydramine",
  # Dietary stimulants
  "caffeine", "theobromine", "theophylline",
  # Pesticides
  "atrazine", "glyphosate", "chlorpyrifos", "permethrin",
  "cypermethrin", "deltamethrin", "malathion",
  # Preservatives
  "methylparaben", "ethylparaben", "propylparaben", "butylparaben",
  "\\bparaben\\b",
  sep = "|"
)

# ClassyFire taxonomy classes that are predominantly synthetic/exogenous
exogenous_superclass <- c("Organohalogens")
exogenous_class      <- c(
  "Organochlorides", "Organobromides",
  "Chlorofluorocarbons", "Synthetic dyes", "Anabolic steroids"
)
```

---

# Apply Flags

```{r apply_flags}
flag_df <- annotation %>%
  mutate(Name_lower = str_to_lower(Name)) %>%
  # Step 1: boolean flag columns
  mutate(
    flag_known_instrument    = Name_lower %in% known_instrument_contaminants,
    flag_known_exposure      = Name_lower %in% known_environmental_exposures,
    flag_pattern_instrument  = str_detect(Name_lower,
                                 regex(instrument_contaminant_patterns, ignore_case = TRUE)),
    flag_pattern_exposure    = str_detect(Name_lower,
                                 regex(environmental_exposure_patterns, ignore_case = TRUE)),
    flag_taxonomy_instrument = (Super.Class %in% exogenous_superclass) |
                               (Class %in% exogenous_class)
  ) %>%
  # Step 2: Exogenous_Flag
  mutate(
    Exogenous_Flag = case_when(
      flag_known_instrument | flag_pattern_instrument | flag_taxonomy_instrument
        ~ "instrument_contaminant",
      flag_known_exposure | flag_pattern_exposure
        ~ "environmental_exposure",
      TRUE ~ NA_character_
    )
  ) %>%
  # Step 3: Flag_Reason
  mutate(
    Flag_Reason = case_when(
      flag_known_instrument    ~ "Known instrument contaminant (project notes or NIST14 review)",
      flag_known_exposure      ~ "Known synthetic pharmaceutical or steroid (project notes)",
      flag_pattern_instrument  ~ "Name matches industrial or contaminant keyword pattern",
      flag_pattern_exposure    ~ "Name matches pharmaceutical or pesticide keyword pattern",
      flag_taxonomy_instrument ~ "ClassyFire taxonomy flags synthetic or halogenated class",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-Name_lower, -starts_with("flag_", ignore.case = FALSE))
```

---

# Summary

```{r summary}
annotation_flagged <- flag_df %>%
  # Annotation confidence tiers based on Best.Match score
  # Tiers: 80+ high, 70-79 medium, 60-69 low
  # High-confidence (80+) compounds are bolded in tables throughout downstream reports
  mutate(
    Annotation_Confidence = case_when(
      Best.Match >= 80 ~ "high",
      Best.Match >= 70 ~ "medium",
      Best.Match >= 60 ~ "low",
      TRUE             ~ NA_character_
    )
  )

cat("=== Exogenous Flagging Summary ===\n")
cat("Total compounds:", nrow(annotation_flagged), "\n")
cat("Presumed endogenous (NA):",
    sum(is.na(annotation_flagged$Exogenous_Flag)), "\n\n")

flag_counts <- annotation_flagged %>%
  filter(!is.na(Exogenous_Flag)) %>%
  count(Exogenous_Flag)
print(flag_counts)
```

## Instrument Contaminants

```{r show_instrument}
annotation_flagged %>%
  filter(Exogenous_Flag == "instrument_contaminant") %>%
  select(Compound.ID, Name, Best.Match, Annotation_Confidence, Super.Class, Class, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Instrument Contaminants — excluded from all analyses. **Bold** = high confidence (score ≥ 80)")
```

## Environmental Exposures

```{r show_exposures}
annotation_flagged %>%
  filter(Exogenous_Flag == "environmental_exposure") %>%
  select(Compound.ID, Name, Best.Match, Annotation_Confidence, Super.Class, Class, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Environmental Exposures — removed from main analysis. **Bold** = high confidence (score ≥ 80)")
```

## Remaining Review Needed

```{r show_review}
remaining_review <- annotation_flagged %>%
  filter(Exogenous_Flag == "review_needed")

if (nrow(remaining_review) == 0) {
  cat("No compounds remaining in review_needed.\n")
} else {
  remaining_review %>%
    select(Compound.ID, Name, Super.Class, Class, Flag_Reason) %>%
    knitr::kable(caption = "Still requires manual review")
}
```

---

# Annotation Confidence Tiers

Compounds are assigned a confidence tier based on `Best.Match` spectral score. The score
distribution is flat from 60–88 then rises sharply at ≥90, with no natural sub-break
in the low-medium range. Tiers are defined as:

| Tier | Score | Interpretation |
|---|---|---|
| `high` | ≥ 80 | Strong spectral match; high identification confidence; **bolded in tables** |
| `medium` | 70–79 | Acceptable for untargeted metabolomics |
| `low` | 60–69 | Borderline; interpret with caution |

```{r confidence_tier_summary}
tier_counts <- annotation_flagged %>%
  count(Annotation_Confidence) %>%
  mutate(Annotation_Confidence = factor(Annotation_Confidence,
                                        levels = c("high", "medium", "low"))) %>%
  arrange(Annotation_Confidence)

print(tier_counts)

ggplot(annotation_flagged, aes(x = Best.Match, fill = Annotation_Confidence)) +
  geom_histogram(binwidth = 2, color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c(high = "#2166ac", medium = "#92c5de", low = "#f4a582"),
    na.value = "grey70",
    name = "Confidence"
  ) +
  geom_vline(xintercept = c(75, 90), linetype = "dashed", color = "grey30") +
  theme_minimal() +
  labs(
    title = "Annotation Score Distribution with Confidence Tiers",
    x = "Best Match Score",
    y = "Compound Count"
  )
```

---

# Score Distribution Analysis: Exogenous Enrichment in Low-Confidence Zone

Are exogenous compounds disproportionately found in the 60–80 score range? If so, this
would support using annotation confidence as an additional filter criterion for
biological interpretation.

```{r score_exogenous_analysis}
# Define low-confidence zone for this analysis
score_zone_cutoff <- 80

analysis_df <- annotation_flagged %>%
  mutate(
    Score_Zone   = if_else(Best.Match < score_zone_cutoff, "low (60-79)", "higher (80-100)"),
    Is_Exogenous = if_else(!is.na(Exogenous_Flag), "exogenous", "endogenous/unknown")
  )

# --- Contingency table ---
contingency <- table(analysis_df$Score_Zone, analysis_df$Is_Exogenous)
cat("=== Contingency Table: Score Zone vs Exogenous Status ===\n")
print(contingency)

# --- Proportions ---
cat("\n=== Proportion exogenous within each score zone ===\n")
prop_table <- analysis_df %>%
  group_by(Score_Zone) %>%
  summarise(
    n_total    = n(),
    n_exogenous = sum(Is_Exogenous == "exogenous"),
    pct_exogenous = round(100 * n_exogenous / n_total, 1),
    .groups = "drop"
  )
print(prop_table)

# --- Visualisation ---
ggplot(prop_table, aes(x = Score_Zone, y = pct_exogenous, fill = Score_Zone)) +
  geom_col(color = "grey30", width = 0.5) +
  geom_text(aes(label = paste0(pct_exogenous, "%\n(n=", n_exogenous, "/", n_total, ")")),
            vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("low (60-79)" = "#f4a582", "higher (80-100)" = "#92c5de")) +
  scale_y_continuous(limits = c(0, max(prop_table$pct_exogenous) * 1.3)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Proportion of Exogenous Compounds by Score Zone",
    x = "Annotation Score Zone",
    y = "% Exogenous"
  )
```

## Breakdown by Flag Type within Score Zones

```{r score_flag_breakdown}
annotation_flagged %>%
  mutate(Score_Zone = if_else(Best.Match < score_zone_cutoff,
                              "low (60-79)", "higher (80-100)")) %>%
  filter(!is.na(Exogenous_Flag)) %>%
  count(Score_Zone, Exogenous_Flag) %>%
  pivot_wider(names_from = Score_Zone, values_from = n, values_fill = 0) %>%
  knitr::kable(caption = "Exogenous compound counts by flag type and score zone")
```

## Low-Confidence Exogenous Compounds

Exogenous compounds in the 60–79 score range warrant particular scrutiny — these may
be misidentified endogenous metabolites.

```{r low_confidence_exogenous}
annotation_flagged %>%
  filter(Best.Match < score_zone_cutoff & !is.na(Exogenous_Flag)) %>%
  select(Compound.ID, Name, Best.Match, Annotation_Confidence,
         Exogenous_Flag, Flag_Reason, Super.Class, Class) %>%
  arrange(Best.Match) %>%
  knitr::kable(caption = "Exogenous compounds with Best.Match < 80 — verify identities")
```

---

# Save Outputs

```{r save}
# Full annotation with Exogenous_Flag, Flag_Reason, and Annotation_Confidence columns
write_csv(annotation_flagged, output_path)
cat("Flagged annotation saved to:", output_path, "\n")

# Candidate review table (flagged only)
candidates <- annotation_flagged %>%
  filter(!is.na(Exogenous_Flag)) %>%
  select(Compound.ID, Name, Super.Class, Class, Sub.Class,
         Best.Match, Annotation_Confidence,
         Exogenous_Flag, Flag_Reason) %>%
  arrange(Exogenous_Flag, Name)

write_csv(candidates, candidates_path)
cat("Candidate review table saved to:", candidates_path, "\n")
cat("Total flagged compounds:", nrow(candidates), "\n")
```

---

# Save Clean Peak Area Dataset

Removes all exogenous compounds (both `instrument_contaminant` and `environmental_exposure`)
from the peak area matrix and saves `newbatch2_peak_area_clean.csv`. This is the file read
by all downstream analysis scripts — no filtering is needed there.

```{r save_clean_peak_area}
peak_area_path  <- "../../results/untargeted/newbatch2_peak_area_values.csv"
clean_peak_path <- "../../results/untargeted/newbatch2_peak_area_clean.csv"

newbatch2_peak_area <- read_csv(peak_area_path, show_col_types = FALSE)

exogenous_ids <- annotation_flagged %>%
  filter(!is.na(Exogenous_Flag)) %>%
  pull(Compound.ID)

n_before <- nrow(newbatch2_peak_area)
newbatch2_peak_area_clean <- newbatch2_peak_area %>%
  filter(!Compound.ID %in% exogenous_ids)
n_after <- nrow(newbatch2_peak_area_clean)

cat("=== Clean Peak Area Dataset ===\n")
cat("Compounds before removal:", n_before, "\n")
cat("Exogenous removed:       ", n_before - n_after, "\n")
cat("  instrument_contaminant:", sum(annotation_flagged$Exogenous_Flag == "instrument_contaminant", na.rm = TRUE), "\n")
cat("  environmental_exposure:", sum(annotation_flagged$Exogenous_Flag == "environmental_exposure", na.rm = TRUE), "\n")
cat("Compounds remaining:     ", n_after, "\n\n")

cat("--- Removed environmental exposures ---\n")
annotation_flagged %>%
  filter(Exogenous_Flag == "environmental_exposure") %>%
  select(Compound.ID, Name, Best.Match, Annotation_Confidence, Flag_Reason) %>%
  arrange(Name) %>%
  mutate(Name = bold_high_confidence(Name, Annotation_Confidence)) %>%
  knitr::kable(caption = "Environmental exposures removed. **Bold** = high confidence (score >= 80)")

write_csv(newbatch2_peak_area_clean, clean_peak_path)
cat("Clean peak area saved to:", clean_peak_path, "\n")
```
