---
title: "New Batch 2 Analysis Results Summary"
author: "Sneha Michaela Chaturvedi"
date: "Last Updated `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
---

Code developed by author with assistance from Claude (Sonnet 4.6). Output checked and validated by author.

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

Packages <- c("tidyverse", "knitr", "kableExtra")
lapply(Packages, library, character.only = TRUE)
```

```{r paths, include=FALSE}
# Input: CSVs produced by the individual analysis scripts
clean_csv <- "../../results/untargeted/newbatch2_clean/csv"
steroid_csv <- function(scope) file.path("../../results/untargeted/newbatch2_steroids", scope, "csv")
nt_csv <- function(scope) file.path("../../results/untargeted/newbatch2_neurotransmitters", scope, "csv")
med_csv <- "../../results/untargeted/newbatch2_medications/csv"

# Output
png_path <- "../../results/untargeted/newbatch2_clean"
svg_path <- "../../results/untargeted/newbatch2_clean/svg"
csv_path <- clean_csv
```

```{r helpers, include=FALSE}
# Safely read a CSV; returns NULL with a message if file not found
safe_read <- function(path) {
  if (file.exists(path)) {
    read_csv(path, show_col_types = FALSE)
  } else {
    message("File not found (analysis may not have been run): ", path)
    NULL
  }
}

# Count significant hits from a limma-style results data frame
# Expects columns: P.Value, adj.P.Val
limma_counts <- function(df) {
  if (is.null(df)) {
    return(list(n_total = NA, n_nominal = NA, n_fdr = NA))
  }
  list(
    n_total   = nrow(df),
    n_nominal = sum(df$P.Value < 0.05, na.rm = TRUE),
    n_fdr     = sum(df$adj.P.Val < 0.1, na.rm = TRUE)
  )
}

# Count significant hits from a wilcoxon-style results data frame
# Expects columns: p_value, FDR
wilcox_counts <- function(df) {
  if (is.null(df)) {
    return(list(n_total = NA, n_nominal = NA, n_fdr = NA))
  }
  list(
    n_total   = nrow(df),
    n_nominal = sum(df$p_value < 0.05, na.rm = TRUE),
    n_fdr     = sum(df$FDR < 0.1, na.rm = TRUE)
  )
}

# Count VIP hits from a PLS-DA VIP data frame
# Expects column: VIP_comp1
vip_counts <- function(df) {
  if (is.null(df)) {
    return(list(n_total = NA, n_vip1 = NA, n_vip2 = NA))
  }
  list(
    n_total = nrow(df),
    n_vip1  = sum(df$VIP_comp1 > 1, na.rm = TRUE),
    n_vip2  = sum(df$VIP_comp1 > 2, na.rm = TRUE)
  )
}

# Count metabolites in both VIP > 1 (PLS-DA) and limma nominal (p < 0.05)
# limma_df: expects column P.Value; vip_df: expects column VIP_comp1; joined on Name
overlap_vip_nominal <- function(limma_df, vip_df) {
  if (is.null(limma_df) || is.null(vip_df)) return(NA_integer_)
  nominal_names <- limma_df %>% filter(P.Value < 0.05) %>% pull(Name)
  vip_names     <- vip_df   %>% filter(VIP_comp1 > 1)  %>% pull(Name)
  length(intersect(nominal_names, vip_names))
}

# Count significant pathways from MSEA (fgsea); expects columns pval, padj
msea_counts <- function(df) {
  if (is.null(df)) return(list(n_total = NA, n_nominal = NA, n_fdr = NA))
  list(
    n_total   = nrow(df),
    n_nominal = sum(df$pval < 0.05, na.rm = TRUE),
    n_fdr     = sum(df$padj < 0.1,  na.rm = TRUE)
  )
}

# Count significant pathways from ORA (Fisher); expects columns pval, padj
ora_counts <- function(df) {
  if (is.null(df)) return(list(n_total = NA, n_nominal = NA, n_fdr = NA))
  list(
    n_total   = nrow(df),
    n_nominal = sum(df$pval < 0.05, na.rm = TRUE),
    n_fdr     = sum(df$padj < 0.1,  na.rm = TRUE)
  )
}

# Build one summary row
summary_row <- function(analysis, subgroup, method, n_total, n_nominal, n_fdr,
                        n_vip1 = NA, n_vip2 = NA, n_overlap = NA) {
  data.frame(
    Analysis  = analysis,
    Subgroup  = subgroup,
    Method    = method,
    N_Total   = n_total,
    N_Nominal = n_nominal,
    N_FDR     = n_fdr,
    N_VIP1    = n_vip1,
    N_VIP2    = n_vip2,
    N_Overlap = n_overlap,
    stringsAsFactors = FALSE
  )
}
```

# Load Results

```{r load_untargeted}
# ── Full Untargeted (newbatch2_clean) ─────────────────────────────────────────────
gw_limma <- safe_read(file.path(clean_csv, "newbatch2_gw_limma_results.csv"))
sex_limma <- safe_read(file.path(clean_csv, "newbatch2_sex_limma_results.csv"))
sex_wilcox <- safe_read(file.path(clean_csv, "newbatch2_sex_differential_analysis.csv"))
plsda_vip <- safe_read(file.path(clean_csv, "newbatch2_sex_plsda_vip.csv"))

# Interaction model (Sex * GW): three extractable coefficients
sex_adj_limma <- safe_read(file.path(clean_csv, "newbatch2_sex_adjusted_limma_results.csv"))
gw_adj_limma  <- safe_read(file.path(clean_csv, "newbatch2_gw_adjusted_limma_results.csv"))
int_limma     <- safe_read(file.path(clean_csv, "newbatch2_interaction_limma_results.csv"))
```

```{r load_pathway}
# ── Pathway Analysis ───────────────────────────────────────────────────────────
msea_gw  <- safe_read(file.path(clean_csv, "newbatch2_msea_gw.csv"))
msea_sex <- safe_read(file.path(clean_csv, "newbatch2_msea_sex.csv"))
ora_sex_nominal   <- safe_read(file.path(clean_csv, "newbatch2_ora_sex_limma_nominal.csv"))
ora_gw_nominal    <- safe_read(file.path(clean_csv, "newbatch2_ora_gw_limma_nominal.csv"))
ora_gw_fdr        <- safe_read(file.path(clean_csv, "newbatch2_ora_gw_limma_fdr.csv"))
ora_vip1          <- safe_read(file.path(clean_csv, "newbatch2_ora_sex_plsda_vip1.csv"))
ora_vip2          <- safe_read(file.path(clean_csv, "newbatch2_ora_sex_plsda_vip2.csv"))
```

```{r load_steroids}
# ── Steroid Subanalysis ────────────────────────────────────────────────────────
steroid_broad_inventory <- safe_read(file.path(steroid_csv("broad"), "newbatch2_steroid_broad_inventory.csv"))
steroid_strict_inventory <- safe_read(file.path(steroid_csv("strict"), "newbatch2_steroid_strict_inventory.csv"))

steroid_broad_gw_limma <- safe_read(file.path(steroid_csv("broad"), "newbatch2_steroid_broad_gw_limma.csv"))
steroid_broad_sex_limma <- safe_read(file.path(steroid_csv("broad"), "newbatch2_steroid_broad_sex_limma.csv"))
steroid_broad_plsda_vip <- safe_read(file.path(steroid_csv("broad"), "newbatch2_steroid_broad_plsda_vip.csv"))

steroid_strict_gw_limma <- safe_read(file.path(steroid_csv("strict"), "newbatch2_steroid_strict_gw_limma.csv"))
steroid_strict_sex_limma <- safe_read(file.path(steroid_csv("strict"), "newbatch2_steroid_strict_sex_limma.csv"))
steroid_strict_plsda_vip <- safe_read(file.path(steroid_csv("strict"), "newbatch2_steroid_strict_plsda_vip.csv"))
```

```{r load_neurotransmitters}
# ── Neurotransmitter Subanalysis ───────────────────────────────────────────────
nt_broad_inventory <- safe_read(file.path(nt_csv("broad"), "newbatch2_nt_broad_inventory.csv"))
nt_strict_inventory <- safe_read(file.path(nt_csv("strict"), "newbatch2_nt_strict_inventory.csv"))

nt_broad_gw_limma <- safe_read(file.path(nt_csv("broad"), "newbatch2_nt_broad_gw_limma.csv"))
nt_broad_sex_limma <- safe_read(file.path(nt_csv("broad"), "newbatch2_nt_broad_sex_limma.csv"))
nt_broad_plsda_vip <- safe_read(file.path(nt_csv("broad"), "newbatch2_nt_broad_plsda_vip.csv"))

nt_strict_gw_limma <- safe_read(file.path(nt_csv("strict"), "newbatch2_nt_strict_gw_limma.csv"))
nt_strict_sex_limma <- safe_read(file.path(nt_csv("strict"), "newbatch2_nt_strict_sex_limma.csv"))
nt_strict_plsda_vip <- safe_read(file.path(nt_csv("strict"), "newbatch2_nt_strict_plsda_vip.csv"))
```

```{r load_medications}
# ── Medication Subanalysis ─────────────────────────────────────────────────────
med_inventory    <- safe_read(file.path(med_csv, "newbatch2_medication_inventory.csv"))
med_gw_limma     <- safe_read(file.path(med_csv, "newbatch2_medication_gw_limma.csv"))
med_sex_limma    <- safe_read(file.path(med_csv, "newbatch2_medication_sex_limma.csv"))
med_sex_wilcoxon <- safe_read(file.path(med_csv, "newbatch2_medication_sex_wilcoxon.csv"))
```

# Results Summary

```{r build_summary_table}
# Helper to build a limma row without repeating boilerplate
limma_row <- function(analysis, subgroup, df) {
  c <- limma_counts(df)
  summary_row(analysis, subgroup, "limma",
    n_total = c$n_total, n_nominal = c$n_nominal, n_fdr = c$n_fdr
  )
}

wilcox_row <- function(analysis, subgroup, df) {
  c <- wilcox_counts(df)
  summary_row(analysis, subgroup, "Wilcoxon",
    n_total = c$n_total, n_nominal = c$n_nominal, n_fdr = c$n_fdr
  )
}

plsda_row <- function(analysis, subgroup, df, limma_df = NULL) {
  c <- vip_counts(df)
  summary_row(analysis, subgroup, "PLS-DA VIP",
    n_total = c$n_total, n_nominal = NA, n_fdr = NA,
    n_vip1 = c$n_vip1, n_vip2 = c$n_vip2,
    n_overlap = if (!is.null(limma_df)) overlap_vip_nominal(limma_df, df) else NA
  )
}

# pathway_row: N_Total = pathways tested; N_Nominal = pval < 0.05; N_FDR = padj < 0.1
pathway_row <- function(analysis, subgroup, method, df, count_fn) {
  c <- count_fn(df)
  summary_row(analysis, subgroup, method,
    n_total = c$n_total, n_nominal = c$n_nominal, n_fdr = c$n_fdr
  )
}

rows <- list(

  # ── Full Untargeted — independent models ─────────────────────────────────────
  limma_row("GW (independent)", "Full untargeted", gw_limma),
  limma_row("Sex (independent)", "Full untargeted", sex_limma),
  wilcox_row("Sex", "Full untargeted", sex_wilcox),
  plsda_row("Sex", "Full untargeted", plsda_vip, limma_df = sex_limma),

  # ── Full Untargeted — interaction model (Sex * GW) ───────────────────────────
  limma_row("Sex | GW (adjusted)", "Full untargeted", sex_adj_limma),
  limma_row("GW | Sex (adjusted)", "Full untargeted", gw_adj_limma),
  limma_row("Sex:GW interaction", "Full untargeted", int_limma),

  # ── Steroids (broad) ────────────────────────────────────────────────────────
  limma_row("GW", "Steroids (broad)", steroid_broad_gw_limma),
  limma_row("Sex", "Steroids (broad)", steroid_broad_sex_limma),
  plsda_row("Sex", "Steroids (broad)", steroid_broad_plsda_vip),

  # ── Steroids (strict) ───────────────────────────────────────────────────────
  limma_row("GW", "Steroids (strict)", steroid_strict_gw_limma),
  limma_row("Sex", "Steroids (strict)", steroid_strict_sex_limma),
  plsda_row("Sex", "Steroids (strict)", steroid_strict_plsda_vip),

  # ── Neurotransmitters (broad) ────────────────────────────────────────────────
  limma_row("GW", "Neurotransmitters (broad)", nt_broad_gw_limma),
  limma_row("Sex", "Neurotransmitters (broad)", nt_broad_sex_limma),
  plsda_row("Sex", "Neurotransmitters (broad)", nt_broad_plsda_vip),

  # ── Neurotransmitters (strict) ───────────────────────────────────────────────
  limma_row("GW", "Neurotransmitters (strict)", nt_strict_gw_limma),
  limma_row("Sex", "Neurotransmitters (strict)", nt_strict_sex_limma),
  plsda_row("Sex", "Neurotransmitters (strict)", nt_strict_plsda_vip),

  # ── Medications ──────────────────────────────────────────────────────────────
  limma_row("GW",  "Medications", med_gw_limma),
  limma_row("Sex", "Medications", med_sex_limma),
  wilcox_row("Sex", "Medications", med_sex_wilcoxon),

  # ── Pathway Analysis (MSEA) ───────────────────────────────────────────────────
  pathway_row("GW",  "Full untargeted", "MSEA (fgsea)", msea_gw,  msea_counts),
  pathway_row("Sex", "Full untargeted", "MSEA (fgsea)", msea_sex, msea_counts),

  # ── Pathway Analysis (ORA) ────────────────────────────────────────────────────
  pathway_row("GW (nominal hits)",  "Full untargeted", "ORA (Fisher)", ora_gw_nominal,  ora_counts),
  pathway_row("GW (FDR hits)",      "Full untargeted", "ORA (Fisher)", ora_gw_fdr,      ora_counts),
  pathway_row("Sex (nominal hits)", "Full untargeted", "ORA (Fisher)", ora_sex_nominal,  ora_counts),
  pathway_row("Sex (VIP > 1)",      "Full untargeted", "ORA (Fisher)", ora_vip1,         ora_counts),
  pathway_row("Sex (VIP > 2)",      "Full untargeted", "ORA (Fisher)", ora_vip2,         ora_counts)
)

summary_df <- bind_rows(rows)
```

```{r display_table}
summary_df %>%
  mutate(
    N_Nominal = ifelse(is.na(N_Nominal), "—", as.character(N_Nominal)),
    N_FDR     = ifelse(is.na(N_FDR), "—", as.character(N_FDR)),
    N_VIP1    = ifelse(is.na(N_VIP1), "—", as.character(N_VIP1)),
    N_VIP2    = ifelse(is.na(N_VIP2), "—", as.character(N_VIP2)),
    N_Overlap = ifelse(is.na(N_Overlap), "—", as.character(N_Overlap))
  ) %>%
  rename(
    `Analysis`              = Analysis,
    `Subgroup`              = Subgroup,
    `Method`                = Method,
    `N Total`               = N_Total,
    `p < 0.05`              = N_Nominal,
    `FDR < 0.1`             = N_FDR,
    `VIP > 1`               = N_VIP1,
    `VIP > 2`               = N_VIP2,
    `VIP>1 ∩ p<0.05`        = N_Overlap
  ) %>%
  kable(caption = "Batch 2 Analysis Results Summary") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  pack_rows("Full Untargeted — independent models", 1, 4) %>%
  pack_rows("Full Untargeted — interaction model", 5, 7) %>%
  pack_rows("Steroids (broad)", 8, 10) %>%
  pack_rows("Steroids (strict)", 11, 13) %>%
  pack_rows("Neurotransmitters (broad)", 14, 16) %>%
  pack_rows("Neurotransmitters (strict)", 17, 19) %>%
  pack_rows("Medications", 20, 22) %>%
  pack_rows("Pathway Analysis — MSEA", 23, 24) %>%
  pack_rows("Pathway Analysis — ORA", 25, 29)
```

# Top Hits Per Analysis

## GW — Full Untargeted (limma, top 20 by p-value)
```{r top_gw}
if (!is.null(gw_limma)) {
  gw_limma %>%
    arrange(P.Value) %>%
    head(20) %>%
    dplyr::select(Name, logFC, P.Value, adj.P.Val, Super.Class, Class) %>%
    mutate(across(where(is.numeric), ~ signif(., 3))) %>%
    rename(`log2FC` = logFC, `p-value` = P.Value, `FDR` = adj.P.Val) %>%
    kable(caption = "Top 20 GW-associated metabolites (full untargeted)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

## Sex — Full Untargeted (limma, top 20 by p-value)
```{r top_sex_limma}
if (!is.null(sex_limma)) {
  sex_limma %>%
    arrange(P.Value) %>%
    head(20) %>%
    dplyr::select(Name, logFC, P.Value, adj.P.Val, Direction, Super.Class, Class) %>%
    mutate(across(where(is.numeric), ~ signif(., 3))) %>%
    rename(`log2FC` = logFC, `p-value` = P.Value, `FDR` = adj.P.Val) %>%
    kable(caption = "Top 20 sex-differential metabolites (full untargeted, limma)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

## Sex — Full Untargeted (PLS-DA VIP > 2)
```{r top_plsda}
if (!is.null(plsda_vip)) {
  plsda_vip %>%
    filter(VIP_comp1 > 2) %>%
    arrange(desc(VIP_comp1)) %>%
    dplyr::select(Name, VIP_comp1, Super.Class, Class) %>%
    mutate(VIP_comp1 = round(VIP_comp1, 3)) %>%
    kable(caption = "PLS-DA metabolites with VIP > 2 (full untargeted)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

## Steroids — Top Hits (broad scope)
```{r top_steroid_broad}
if (!is.null(steroid_broad_gw_limma)) {
  cat("GW limma — Steroids (broad)\n")
  steroid_broad_gw_limma %>%
    arrange(P.Value) %>%
    head(10) %>%
    dplyr::select(Name, logFC, P.Value, adj.P.Val) %>%
    mutate(across(where(is.numeric), ~ signif(., 3))) %>%
    rename(`log2FC` = logFC, `p-value` = P.Value, `FDR` = adj.P.Val) %>%
    kable(caption = "Top 10 GW-associated steroids (broad)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    print()
}

if (!is.null(steroid_broad_sex_limma)) {
  steroid_broad_sex_limma %>%
    arrange(P.Value) %>%
    head(10) %>%
    dplyr::select(Name, logFC, P.Value, adj.P.Val, Direction) %>%
    mutate(across(where(is.numeric), ~ signif(., 3))) %>%
    rename(`log2FC` = logFC, `p-value` = P.Value, `FDR` = adj.P.Val) %>%
    kable(caption = "Top 10 sex-differential steroids (broad, limma)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    print()
}
```

## Neurotransmitters — Top Hits (broad scope)
```{r top_nt_broad}
if (!is.null(nt_broad_gw_limma)) {
  nt_broad_gw_limma %>%
    arrange(P.Value) %>%
    head(10) %>%
    dplyr::select(Name, logFC, P.Value, adj.P.Val) %>%
    mutate(across(where(is.numeric), ~ signif(., 3))) %>%
    rename(`log2FC` = logFC, `p-value` = P.Value, `FDR` = adj.P.Val) %>%
    kable(caption = "Top 10 GW-associated neurotransmitters (broad)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    print()
}

if (!is.null(nt_broad_sex_limma)) {
  nt_broad_sex_limma %>%
    arrange(P.Value) %>%
    head(10) %>%
    dplyr::select(Name, logFC, P.Value, adj.P.Val, Direction) %>%
    mutate(across(where(is.numeric), ~ signif(., 3))) %>%
    rename(`log2FC` = logFC, `p-value` = P.Value, `FDR` = adj.P.Val) %>%
    kable(caption = "Top 10 sex-differential neurotransmitters (broad, limma)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
    print()
}
```

## Medications — Detected Compounds and Top Hits

```{r med_inventory_table}
if (!is.null(med_inventory)) {
  med_inventory %>%
    dplyr::select(Name, Drug_Class, Class, Annotation_Confidence) %>%
    arrange(Drug_Class, Name) %>%
    kable(caption = "Detected medications by drug class") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

```{r top_med_gw}
if (!is.null(med_gw_limma)) {
  med_gw_limma %>%
    arrange(P.Value) %>%
    dplyr::select(Name, Drug_Class, logFC, P.Value, adj.P.Val, Direction) %>%
    mutate(across(where(is.numeric), ~ signif(., 3))) %>%
    rename(`log2FC` = logFC, `p-value` = P.Value, `FDR` = adj.P.Val) %>%
    kable(caption = "GW limma — all medications (sorted by p-value)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

```{r top_med_sex}
if (!is.null(med_sex_limma)) {
  med_sex_limma %>%
    arrange(P.Value) %>%
    dplyr::select(Name, Drug_Class, logFC, P.Value, adj.P.Val, Direction) %>%
    mutate(across(where(is.numeric), ~ signif(., 3))) %>%
    rename(`log2FC` = logFC, `p-value` = P.Value, `FDR` = adj.P.Val) %>%
    kable(caption = "Sex limma — all medications (sorted by p-value)") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

# Save Summary Table

```{r save_summary}
write_csv(summary_df, file.path(csv_path, "newbatch2_analysis_summary.csv"))
cat("Summary saved to:", file.path(csv_path, "newbatch2_analysis_summary.csv"), "\n")
```
