---
title: "Combine Raw Datasets (Unfiltered, With Duplicates)"
author: "Sneha Michaela Chaturvedi"
date: "Last Updated `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: TRUE
---

Made w/ assistance from Claude Sonnet 4.6. Output checked and validated by author.

# Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

Packages <- c("tidyverse", "readxl", "stats", "ggfortify", "ggrepel", "svglite", "pheatmap", "pracma")
lapply(Packages, library, character.only = TRUE)
```

```{r source_functions}
# Source Functions
source("functions.R")
source("filtering_functions.R")
```

```{r set_path}
png_path <- "../../results/untargeted/raw_combined_dataset"
csv_path <- "../../results/untargeted/raw_combined_dataset/csv"
svg_path <- "../../results/untargeted/raw_combined_dataset/svg"
data_path <- "../../data"

# Create output directories
dir.create(png_path, recursive = TRUE, showWarnings = FALSE)
dir.create(csv_path, recursive = TRUE, showWarnings = FALSE)
dir.create(svg_path, recursive = TRUE, showWarnings = FALSE)
```

# Load Raw Data

## Batch 1
```{r load-batch1-raw}
# Peak Area Values and Metabolite Annotations
batch1_full_dataset      <- read_excel('../../doc/untargeted/8_untargeted_developing_2025.xlsx', sheet = "Area",       range = NULL)
batch1_compound_annotation <- read_excel('../../doc/untargeted/8_untargeted_developing_2025.xlsx', sheet = "Annotation", range = NULL)

# Sample Metadata
batch1_sample_metadata <- read_excel('../../doc/untargeted/8_metadata.xlsx', sheet = "metadata", range = NULL)
batch1_sample_metadata <- batch1_sample_metadata %>%
  rename(Sample = SampleName)
```

## Batch 2
```{r load-batch2-raw}
# Peak Area Values and Metabolite Annotations
batch2_full_dataset        <- read_excel('../../doc/untargeted/24_untargeted_developing_2025.xlsx', sheet = "Area",       range = NULL)
batch2_compound_annotation <- read_excel('../../doc/untargeted/24_untargeted_developing_2025.xlsx', sheet = "Annotation", range = NULL)

# Sample Metadata
batch2_sample_metadata <- read_excel('../../doc/untargeted/24_metadata.xlsx', sheet = "metadata", range = NULL)
batch2_sample_metadata <- batch2_sample_metadata %>%
  rename(Sample = SampleName)
```

# Clean Raw Data

`clean_metabolomics_data()` assigns cleaned objects into the global environment using
the `dataset_name` prefix, producing `<dataset_name>_peak_data`, `_compound_annotation`,
and `_sample_metadata`.

## Batch 1
```{r clean-batch1}
clean_metabolomics_data(batch1_full_dataset, batch1_sample_metadata, batch1_compound_annotation,
                        dataset_name = "Batch1_Untargeted_Raw")
```

## Batch 2
```{r clean-batch2}
clean_metabolomics_data(batch2_full_dataset, batch2_sample_metadata, batch2_compound_annotation,
                        dataset_name = "Batch2_Untargeted_Raw")
```

# Sample Outlier Removal

## Batch 1

Remove samples with outlier mean metabolite intensity:
```{r batch1-outlier-intensity}
batch1_outlier_removal_1_results <- remove_intensity_outliers(
  data           = Batch1_Untargeted_Raw_peak_data,
  metadata       = batch1_sample_metadata,
  threshold      = 3,
  save_plot      = TRUE,
  output_prefix  = "Batch1_01_intensity_outliers",
  png_path       = png_path,
  csv_path       = csv_path
)
batch1_outlier_removal_1 <- batch1_outlier_removal_1_results$filtered_data
```

Remove samples with outlier number of detected peaks:
```{r batch1-outlier-detection}
batch1_outlier_removal_2_results <- remove_detection_outliers(
  data          = batch1_outlier_removal_1,
  metadata      = batch1_sample_metadata,
  threshold     = 3,
  na_threshold  = 0,
  save_plot     = TRUE,
  output_prefix = "Batch1_02_detection_outliers",
  png_path      = png_path,
  csv_path      = csv_path
)
batch1_outlier_removal_2 <- batch1_outlier_removal_2_results$filtered_data
```

## Batch 2

Remove samples with outlier mean metabolite intensity:
```{r batch2-outlier-intensity}
batch2_outlier_removal_1_results <- remove_intensity_outliers(
  data           = Batch2_Untargeted_Raw_peak_data,
  metadata       = batch2_sample_metadata,
  threshold      = 3,
  save_plot      = TRUE,
  output_prefix  = "Batch2_01_intensity_outliers",
  png_path       = png_path,
  csv_path       = csv_path
)
batch2_outlier_removal_1 <- batch2_outlier_removal_1_results$filtered_data
```

Remove samples with outlier number of detected peaks:
```{r batch2-outlier-detection}
batch2_outlier_removal_2_results <- remove_detection_outliers(
  data          = batch2_outlier_removal_1,
  metadata      = batch2_sample_metadata,
  threshold     = 3,
  na_threshold  = 0,
  save_plot     = TRUE,
  output_prefix = "Batch2_02_detection_outliers",
  png_path      = png_path,
  csv_path      = csv_path
)
batch2_outlier_removal_2 <- batch2_outlier_removal_2_results$filtered_data
```

# Remove Unannotated Metabolites

Keep only rows with a non-empty Name. Unannotated features are retained separately for reference but not carried forward into the combined dataset.

```{r remove-unannotated}
batch1_annotated      <- batch1_outlier_removal_2[!is.na(batch1_outlier_removal_2$Name) & batch1_outlier_removal_2$Name != "", ]
batch1_unannotated    <- batch1_outlier_removal_2[ is.na(batch1_outlier_removal_2$Name) | batch1_outlier_removal_2$Name == "", ]

batch2_annotated      <- batch2_outlier_removal_2[!is.na(batch2_outlier_removal_2$Name) & batch2_outlier_removal_2$Name != "", ]
batch2_unannotated    <- batch2_outlier_removal_2[ is.na(batch2_outlier_removal_2$Name) | batch2_outlier_removal_2$Name == "", ]

cat("Batch 1 — annotated:", nrow(batch1_annotated), "  unannotated:", nrow(batch1_unannotated), "\n")
cat("Batch 2 — annotated:", nrow(batch2_annotated), "  unannotated:", nrow(batch2_unannotated), "\n")
```

# Data Inspection

```{r inspect-dimensions}
# Within-batch duplicate check
batch1_dup_names <- batch1_annotated$Name[duplicated(batch1_annotated$Name)]
batch2_dup_names <- batch2_annotated$Name[duplicated(batch2_annotated$Name)]

cat("Duplicate compound names within Batch 1:", length(unique(batch1_dup_names)),
    " (", sum(duplicated(batch1_annotated$Name)), "duplicate rows)\n")
cat("Duplicate compound names within Batch 2:", length(unique(batch2_dup_names)),
    " (", sum(duplicated(batch2_annotated$Name)), "duplicate rows)\n")

# Cross-batch overlap based on unique Names
batch1_unique_names <- unique(batch1_annotated$Name)
batch2_unique_names <- unique(batch2_annotated$Name)

overlap     <- intersect(batch1_unique_names, batch2_unique_names)
batch1_only <- setdiff(batch1_unique_names, batch2_unique_names)
batch2_only <- setdiff(batch2_unique_names, batch1_unique_names)

cat("\nUnique compound names — Batch 1:", length(batch1_unique_names),
    "  Batch 2:", length(batch2_unique_names), "\n")
cat("In BOTH batches:", length(overlap), "\n")
cat("Batch 1 ONLY:", length(batch1_only), "\n")
cat("Batch 2 ONLY:", length(batch2_only), "\n")
```

# Batch Overlap Bar Graph
```{r overlap-barplot}
overlap_summary <- tibble(
  Category = c("Batch 1 Only", "Batch 2 Only", "Both Batches"),
  Count    = c(length(batch1_only), length(batch2_only), length(overlap))
)

p_overlap <- ggplot(overlap_summary, aes(x = Category, y = Count, fill = Category)) +
  geom_bar(stat = "identity", color = "white", width = 0.6) +
  geom_text(aes(label = Count), vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c(
    "Batch 1 Only" = "#4E79A7",
    "Batch 2 Only" = "#F28E2B",
    "Both Batches" = "#59A14F"
  )) +
  labs(
    title = "Metabolites by Batch Presence (Raw Data, Sample Outliers Removed)",
    subtitle = "Based on unique compound Name; within-batch duplicates collapsed for overlap calculation",
    x = NULL,
    y = "Number of Unique Compound Names"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank()
  )

print(p_overlap)

ggsave(file.path(png_path, "metabolite_batch_overlap.png"),
       plot = p_overlap, width = 7, height = 5, dpi = 300)
ggsave(file.path(svg_path, "metabolite_batch_overlap.svg"),
       plot = p_overlap, width = 7, height = 5)

cat("Overlap bar graph saved.\n")
```

# Prepare Data for Combining

Because this is raw data, a single compound Name may appear multiple times within a
batch (multiple peaks/features). A unique row-level ID per batch is assigned so every
peak is traceable even when the same compound Name appears multiple times.

```{r prepare-data}
batch1_annotated <- batch1_annotated %>%
  mutate(
    Compound.ID_Batch1 = Compound.ID,
    Row_ID_Batch1      = row_number()
  ) %>%
  select(Name, Compound.ID_Batch1, Row_ID_Batch1, everything(), -Compound.ID, -Formula)

batch2_annotated <- batch2_annotated %>%
  mutate(
    Compound.ID_Batch2 = Compound.ID,
    Row_ID_Batch2      = row_number()
  ) %>%
  select(Name, Compound.ID_Batch2, Row_ID_Batch2, everything(), -Compound.ID, -Formula)

cat("Batch 1 rows after preparation:", nrow(batch1_annotated), "\n")
cat("Batch 2 rows after preparation:", nrow(batch2_annotated), "\n")
```

# Combine Peak Area Data

Batches are stacked with `bind_rows()` rather than joined on Name. A Name-based join
would create a combinatorial explosion of rows because duplicates exist within and
between batches. Each row retains its batch origin, batch-specific Compound.ID, and
unique row ID for downstream grouping, filtering, or deduplication.

```{r combine-peak-area}
batch1_tagged <- batch1_annotated %>%
  mutate(
    Compound.ID_Batch2 = NA_character_,
    Row_ID_Batch2      = NA_integer_,
    Batch              = "Batch1"
  )

batch2_tagged <- batch2_annotated %>%
  mutate(
    Compound.ID_Batch1 = NA_character_,
    Row_ID_Batch1      = NA_integer_,
    Batch              = "Batch2"
  )

combined_peak_area <- bind_rows(batch1_tagged, batch2_tagged) %>%
  mutate(
    Detected_Batch1 = Batch == "Batch1",
    Detected_Batch2 = Batch == "Batch2"
  ) %>%
  select(
    Name, Batch,
    Compound.ID_Batch1, Row_ID_Batch1,
    Compound.ID_Batch2, Row_ID_Batch2,
    Detected_Batch1, Detected_Batch2,
    everything()
  )

cat("Combined peak data:", nrow(combined_peak_area), "total rows\n")
cat("  Batch 1 features:", sum(combined_peak_area$Batch == "Batch1"), "\n")
cat("  Batch 2 features:", sum(combined_peak_area$Batch == "Batch2"), "\n")
```

# Combine Metabolite Annotations
```{r combine-annotations}
batch1_annotations_tagged <- Batch1_Untargeted_Raw_compound_annotation %>%
  mutate(
    Compound.ID_Batch1 = Compound.ID,
    Batch = "Batch1"
  ) %>%
  select(-Compound.ID)

batch2_annotations_tagged <- Batch2_Untargeted_Raw_compound_annotation %>%
  mutate(
    Compound.ID_Batch2 = Compound.ID,
    Batch = "Batch2"
  ) %>%
  select(-Compound.ID)

combined_annotations <- bind_rows(batch1_annotations_tagged, batch2_annotations_tagged) %>%
  arrange(Name, Batch) %>%
  select(Compound.ID_Batch1, Compound.ID_Batch2, Batch, everything())

cat("Combined annotations:", nrow(combined_annotations), "rows\n")
cat("  Batch 1 annotations:", sum(combined_annotations$Batch == "Batch1"), "\n")
cat("  Batch 2 annotations:", sum(combined_annotations$Batch == "Batch2"), "\n")
```

# Combine Sample Metadata
```{r combine-metadata}
combined_sample_metadata <- bind_rows(
  batch1_sample_metadata %>% mutate(Batch = "Batch1"),
  batch2_sample_metadata %>% mutate(Batch = "Batch2")
) %>%
  select(Sample, Batch, GW, Sex)

cat("Combined samples:", nrow(combined_sample_metadata), "\n")
cat("  Batch 1 samples:", sum(combined_sample_metadata$Batch == "Batch1"), "\n")
cat("  Batch 2 samples:", sum(combined_sample_metadata$Batch == "Batch2"), "\n")

cat("\nSample distribution:\n")
print(combined_sample_metadata %>%
        group_by(Batch, Sex) %>%
        summarise(Count = n(), .groups = "drop"))
```

# Save Combined Raw Data
```{r save-combined}
write_csv(combined_peak_area,
          file.path(data_path, 'raw_full_dataset_peak_area_values.csv'))

write_csv(combined_annotations,
          file.path(data_path, 'raw_full_dataset_metabolite_annotations.csv'))

write_csv(combined_sample_metadata,
          file.path(data_path, 'raw_full_dataset_sample_metadata.csv'))

cat("\nAll raw combined datasets saved successfully to:", data_path, "\n")
cat("Files saved:\n")
cat("  - raw_full_dataset_peak_area_values.csv\n")
cat("  - raw_full_dataset_metabolite_annotations.csv\n")
cat("  - raw_full_dataset_sample_metadata.csv\n")
```

# End
