---
title: "Untargeted_Analysis"
author: "Sneha Michaela Chaturvedi"
date: "05-06-2025"
output: html_document
---

##Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) {
  install.packages("pacman")
  library(pacman)
} 
p_load(tidyverse,readxl,dplyr,stats,ggplot2,ggfortify)

#Functions
#function to convert to numeric
convert_to_numeric <- function(x) {
  x <- as.character(x)
  x[x == "N/A"] <- NA
  as.numeric(x)}

# Imputation function to replace NaN with 1/5th of the minimum value in each row
min_value_impute <- function(column) {
  min_value <- min(column, na.rm = TRUE)
  column[is.na(column)] <- min_value / 5
  return(column)
}

# Function to calculate percentage of samples each peak is present in
calculate_missingness <- function(data) {
  # Calculate proportion of NA values for each metabolite
  missing_prop <- rowMeans(is.na(data[,-1]))
  
  # Add missingness to metabolite names
  missing_df <- data.frame(
    metabolite = data[,1],
    missing_proportion = missing_prop
  )
  
  return(missing_df)
}
```

#Load Dataset
```{r LoadDataset, echo=TRUE}
full_dataset = read_excel('../../doc/untargeted/32_untargeted_developing_2025.xlsx',sheet="Area",range=NULL)
```

#Format Datafile
1. Remove asterisks from names 
  MTAC Key: *indicates isomers detected
2. Make data set completely numeric
3. Remove rows that are NaN across all samples - esp. relevant if annotated with other groups not present in current analysis

```{r CleanUp, echo=TRUE}
# Clean the data
full_dataset <- full_dataset %>%
  # First clean names with asterisks
  mutate(Name = gsub("\\*$", "", Name)) %>%
  # Then convert numeric columns and filter NA rows
  mutate(across(4:ncol(.), convert_to_numeric)) %>%
  filter(if_any(4:ncol(.), ~!is.na(.)))

#create complete numeric data set for functions by replacing N/A with NaN
full_dataset <- cbind(full_dataset[1:3], full_dataset[,(4:27)] %>% replace(is.na(.), NaN))

# Modify sample names for columns 4 to 27
new_col_names <- names(full_dataset)[4:27] %>%
  sub("Area\\. ", "Sample", .)

# Assign the new column names back to the data frame
names(full_dataset)[4:27] <- new_col_names
```

##QC

#Data Visualization
```{r Distribution_Plots_One}
#peak scatter plot
# Reshape the data
data_long <- full_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Peaks")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(full_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Peaks per Sample",
       x = "Sample", y = "Peak Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(full_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Peaks Across Samples",
       x = "Percentage of Samples Missing Peak",
       y = "Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```

#Initial PCA
```{r PCA_One, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- full_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- full_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- full_dataset$Compound.ID

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "PCA of Metabolite Variation")

```

#Outlier Removal
```{r OutlierSamples, echo=TRUE}
identify_outlier_samples <- function(data, threshold = 3, i = c(4:27)) {
  # Calculate total intensity per sample
  sample_sums <- colSums(data[,i], na.rm = TRUE)
  
  # Calculate z-scores for sample intensities
  z_scores <- scale(sample_sums)
  
  # Identify outliers (samples with z-scores above threshold)
  outliers <- which(abs(z_scores) > threshold)
  
  return(list(
    outlier_indices = outliers,
    sample_sums = sample_sums,
    z_scores = z_scores
  ))
}

# Run function and save output
outliers_info <- identify_outlier_samples(full_dataset)

# Convert outlier indices to actual column names
outlier_columns <- colnames(full_dataset)[outliers_info$outlier_indices + 3] # Adjusted for the correct indexing
print(outlier_columns)

# Remove the outliers from the data set
clean_dataset <- full_dataset %>% select(-all_of(outlier_columns))
```

Data Re-visualization after Outlier Exclusion

```{r Distribution_Plots_Two, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- clean_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Peaks")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(clean_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Peaks per Sample",
       x = "Sample", y = "Peak Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(clean_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Peaks Across Samples",
       x = "Percentage of Samples Missing Peak",
       y = "Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```
#Outlier PCA
```{r PCA_Two, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- clean_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- clean_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- clean_dataset$Compound.ID

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "PCA of Metabolite Variation")

```
  
##Data Split
```{r Dataframe_Split}
annotated_dataset <- clean_dataset[!is.na(clean_dataset$Name) & clean_dataset$Name != "", ]
unannotated_dataset <- clean_dataset[is.na(clean_dataset$Name) | clean_dataset$Name == "", ]
```

#Visualization per Split
```{r Distribution_Plots_Annotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- annotated_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Annotated Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Annotated Peaks")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Annotated Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(annotated_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Annotated Peaks per Sample",
       x = "Sample", y = "Peak Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(annotated_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Annotated Peaks Across Samples",
       x = "Percentage of Samples Missing Annotated Peak",
       y = "Annotated Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```

```{r PCA_Annotated, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- annotated_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- annotated_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- annotated_dataset$Compound.ID

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "PCA of Metabolite Variation")

```

```{r Distribution_Plots_Unannotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- unannotated_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Annotated Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Annotated Peaks")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Annotated Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(unannotated_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Unannotated Peaks per Sample",
       x = "Sample", y = "Peak Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(unannotated_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Unannotated Peaks Across Samples",
       x = "Percentage of Samples Missing Unannotated Peak",
       y = "Unannotated Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```
```{r PCA_Unannotated, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- unannotated_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- unannotated_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- unannotated_dataset$Compound.ID

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "PCA of Metabolite Variation")

```

##Annotated Analysis
#Duplicate Removal

```{r LoadAnnotation, echo=TRUE}
compound_annotation = read_excel('../../doc/untargeted/32_untargeted_developing_2025.xlsx',sheet="Annotation", range=NULL)

compound_annotation <- compound_annotation %>%
  # First clean names with asterisks
  mutate(Name = gsub("\\*$", "", Name))

# split file
annotated_compounds <- compound_annotation[!is.na(compound_annotation$InChiKey) & compound_annotation$InChiKey != "", ]
unannotated_compounds <- compound_annotation[is.na(compound_annotation$InChiKey) | compound_annotation$InChiKey == "", ]
```

```{r Duplicate_Identification}

# Find duplicate names in the column "InChiKey"
duplicates <- annotated_compounds %>%
  group_by(InChiKey) %>%
  filter(n() > 1) %>%
  arrange(Name)

# Create a histogram plot
library(ggplot2)

# Calculate the count of occurrences for each duplicate name
duplicate_counts <- duplicates %>%
  group_by(InChiKey) %>%
  summarise(count = n())

# Create a histogram
ggplot(duplicate_counts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  ggtitle("Histogram of Duplicate InChiKey") +
  xlab("Count of Duplicates") +
  ylab("Frequency")

```
Create non-duplicate dataframe
```{r DuplicateRemoval, echo=TRUE}

# Calculate % Relative Standard Deviation (% RSD) for each row
annotated_dataset <- annotated_dataset %>%
  rowwise() %>%
  mutate(
    mean_value = mean(c_across(4:27), na.rm = TRUE),
    std_dev = sd(c_across(4:27), na.rm = TRUE),
    percent_RSD = (std_dev / mean_value) * 100
  ) %>%
  ungroup()

annotated_dataset <- annotated_dataset %>%
  left_join(annotated_compounds %>% select(Compound.ID, InChiKey), by = "Compound.ID")

# Create a new dataframe that keeps only non-NA RSD values, and then finds the duplicate with the lower % RSD for each InChiKey
clean_annotated_dataset <- annotated_dataset %>%
  filter(!is.na(percent_RSD)) %>%
  group_by(InChiKey) %>%
  filter(percent_RSD == min(percent_RSD, na.rm = TRUE)) %>%
  ungroup()
```


#Unique Metabolite Visualization

```{r Distribution_Plots_Unique_Annotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- clean_annotated_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Unique Metabolite Intensity",
       x = "Log10 Intensity",
       y = "Number of Unique Metablites")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Unique Metabolites per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(clean_annotated_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Unique Metabolites per Sample",
       x = "Sample", y = "Metabolite Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(clean_annotated_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Metabolites Across Samples",
       x = "Percentage of Samples Missing Metabolite",
       y = "Unique Metabolite Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```
```{r PCA_Unique_Annotated, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- clean_annotated_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- clean_annotated_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- clean_annotated_dataset$Compound.ID

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "PCA of Metabolite Variation")

```

Assign Metadata
```{r Metadata, echo=TRUE}
metadata = read_excel('./Datasets/metadata.xlsx',sheet="metadata",range=NULL)

# Ensure the "SampleName" column in metadata matches these sample names
metadata$Sample <- paste0("Sample", metadata$SampleName)

# In case the metadata 'Sample' column needs to be redefined to match main_df column names
metadata <- metadata %>%
  select(Sample, Sex, GW)

# Merge metadata with the long-format main dataframe
merged_df <- normalized_clean_annotated %>%
left_join(metadata, by = "Sample")

# Pivot the long-format dataframe back to the wide format if necessary
final_untargeted_dataset <- merged_df %>%
 pivot_wider(names_from = Sample, values_from = Value)

```
