---
title: "Untargeted_Analysis"
author: "Sneha Michaela Chaturvedi"
date: "06-23-2025"
output: html_document
---

# Set Up
```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) {
  install.packages("pacman")
  library(pacman)
} 
p_load(tidyverse,readxl,dplyr,stats,ggplot2,ggfortify,ggrepel,limma,Biobase,ComplexHeatmap)

#Functions
#function to convert to numeric
convert_to_numeric <- function(x) {
  x <- as.character(x)
  x[x == "N/A"] <- NA
  as.numeric(x)}

# Imputation function to replace NaN with 1/5th of the minimum value in each row
min_value_impute <- function(column) {
  min_value <- min(column, na.rm = TRUE)
  column[is.na(column)] <- min_value / 5
  return(column)
}

# Function to calculate percentage of samples each peak is present in
calculate_missingness <- function(data) {
  # Calculate proportion of NA values for each metabolite
  missing_prop <- rowMeans(is.na(data[,-1]))
  
  # Add missingness to metabolite names
  missing_df <- data.frame(
    metabolite = data[,1],
    missing_proportion = missing_prop
  )
  
  return(missing_df)
}
```

# Load Dataset
```{r LoadDataset, echo=TRUE}
full_dataset = read_excel('../../doc/untargeted/32_untargeted_developing_2025.xlsx',sheet="Area",range=NULL)
```

# Format Datafile
1. Remove asterisks from names
  MTAC Key: *indicates isomers detected
2. Make data set completely numeric
3. Remove rows that are NaN across all samples - esp. relevant if annotated with other groups not present in current analysis

```{r CleanUp, echo=TRUE}
# Clean the data
full_dataset <- full_dataset %>%
  # First clean names with asterisks
  mutate(Name = gsub("\\*$", "", Name)) %>%
  # Then convert numeric columns and filter NA rows
  mutate(across(4:ncol(.), convert_to_numeric)) %>%
  filter(if_any(4:ncol(.), ~!is.na(.)))

#create complete numeric data set for functions by replacing N/A with NaN
full_dataset <- cbind(full_dataset[1:3], full_dataset[,(4:27)] %>% replace(is.na(.), NaN))

# Modify sample names for columns 4 to 27
new_col_names <- names(full_dataset)[4:27] %>%
  sub("Area\\. ", "Sample", .)

# Assign the new column names back to the data frame
names(full_dataset)[4:27] <- new_col_names
```

## QC

# Data Visualization
```{r Distribution_Plots_One}
#peak scatter plot
# Reshape the data
data_long <- full_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Peaks")

# Plot the violin plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_violin(fill = "darkorange", color = "black") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(full_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Peaks per Sample",
       x = "Sample", y = "Peak Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(full_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Peaks Across Samples",
       x = "Percentage of Samples Missing Peak",
       y = "Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```

# Initial PCA
```{r PCA_One, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- full_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]), title = "PCA of Sample Variation")

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC2, y = PC3, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]), title = "PCA of Sample Variation")

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC3, y = PC4, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]), title = "PCA of Sample Variation")

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC4, y = PC5, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       y = sprintf("PC5 (%.1f%%)", var_explained_samples[5]), title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- full_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- full_dataset$Compound.ID

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
                   title = "PCA of Metabolite Variation")

```

# Outlier Removal
```{r OutlierSamples, echo=TRUE}
identify_outlier_samples <- function(data, threshold = 3, i = c(4:27)) {
  # Calculate total intensity per sample
  sample_sums <- colSums(data[,i], na.rm = TRUE)
  
  # Calculate z-scores for sample intensities
  z_scores <- scale(sample_sums)
  
  # Identify outliers (samples with z-scores above threshold)
  outliers <- which(abs(z_scores) > threshold)
  
  return(list(
    outlier_indices = outliers,
    sample_sums = sample_sums,
    z_scores = z_scores
  ))
}

# Run function and save output
outliers_info <- identify_outlier_samples(full_dataset)

# Convert outlier indices to actual column names
outlier_columns <- colnames(full_dataset)[outliers_info$outlier_indices + 3] # Adjusted for the correct indexing
print(outlier_columns)

# Remove the outliers from the data set
clean_dataset <- full_dataset %>% select(-all_of(outlier_columns))
```

Data Re-visualization after Outlier Exclusion

```{r Distribution_Plots_Two, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- clean_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Peaks")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(clean_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Peaks per Sample",
       x = "Sample", y = "Peak Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(clean_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Peaks Across Samples",
       x = "Percentage of Samples Missing Peak",
       y = "Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```
# Outlier PCA
```{r PCA_Two, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- clean_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]), title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- clean_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- clean_dataset$Compound.ID

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
                   title = "PCA of Metabolite Variation")
```
  
## Data Split
```{r Dataframe_Split}
annotated_dataset <- clean_dataset[!is.na(clean_dataset$Name) & clean_dataset$Name != "", ]
unannotated_dataset <- clean_dataset[is.na(clean_dataset$Name) | clean_dataset$Name == "", ]
```

# Visualization per Split
```{r Distribution_Plots_Annotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- annotated_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Annotated Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Annotated Peaks")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Annotated Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(annotated_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Annotated Peaks per Sample",
       x = "Sample", y = "Peak Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(annotated_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Annotated Peaks Across Samples",
       x = "Percentage of Samples Missing Annotated Peak",
       y = "Annotated Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```

```{r PCA_Annotated, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- annotated_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]), title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- annotated_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- annotated_dataset$Compound.ID

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
                   title = "PCA of Metabolite Variation")
```

```{r Distribution_Plots_Unannotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- unannotated_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Unannotated Peak Intensity",
       x = "Log10 Intensity",
       y = "Number of Unannotated Peaks")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Unannotated Peak Intensity per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(unannotated_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Distinct Unannotated Peaks per Sample",
       x = "Sample", y = "Peak Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(unannotated_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Unannotated Peaks Across Samples",
       x = "Percentage of Samples Missing Unannotated Peak",
       y = "Unannotated Peak Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```
```{r PCA_Unannotated, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- unannotated_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]), title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- unannotated_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- unannotated_dataset$Compound.ID

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
                   title = "PCA of Metabolite Variation")
```

## Annotated Analysis
# Duplicate Removal

```{r LoadAnnotation, echo=TRUE}
compound_annotation = read_excel('../../doc/untargeted/32_untargeted_developing_2025.xlsx',sheet="Annotation", range=NULL)

compound_annotation <- compound_annotation %>%
  # First clean names with asterisks
  mutate(Name = gsub("\\*$", "", Name))

# split file
annotated_compounds <- compound_annotation[!is.na(compound_annotation$InChiKey) & compound_annotation$InChiKey != "", ]
unannotated_compounds <- compound_annotation[is.na(compound_annotation$InChiKey) | compound_annotation$InChiKey == "", ]
```

```{r Duplicate_Identification}

# Find duplicate names in the column "InChiKey"
duplicates <- annotated_compounds %>%
  group_by(InChiKey) %>%
  filter(n() > 1) %>%
  arrange(Name)

# Create a histogram plot
library(ggplot2)

# Calculate the count of occurrences for each duplicate name
duplicate_counts <- duplicates %>%
  group_by(InChiKey) %>%
  summarise(count = n())

# Create a histogram
ggplot(duplicate_counts, aes(x = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  ggtitle("Histogram of Duplicate InChiKey") +
  xlab("Count of Duplicates") +
  ylab("Frequency")

```
Create non-duplicate dataframe
```{r DuplicateRemoval, echo=TRUE}

# Calculate % Relative Standard Deviation (% RSD) for each row
annotated_dataset <- annotated_dataset %>%
  rowwise() %>%
  mutate(
    mean_value = mean(c_across(4:27), na.rm = TRUE),
    std_dev = sd(c_across(4:27), na.rm = TRUE),
    percent_RSD = (std_dev / mean_value) * 100
  ) %>%
  ungroup()

annotated_dataset <- annotated_dataset %>%
  left_join(annotated_compounds %>% select(Compound.ID, InChiKey), by = "Compound.ID")

# Create a new dataframe that keeps only non-NA RSD values, and then finds the duplicate with the lower % RSD for each InChiKey
clean_annotated_dataset <- annotated_dataset %>%
  filter(!is.na(percent_RSD)) %>%
  group_by(InChiKey) %>%
  filter(percent_RSD == min(percent_RSD, na.rm = TRUE)) %>%
  ungroup()

# Filter the annotated compounds to keep only the rows with Compound.ID in the unique compounds
unique_annotated_compounds = annotated_compounds %>%
  filter(Compound.ID %in% clean_annotated_dataset$Compound.ID)

```
# Unique Metabolite Visualization

```{r Distribution_Plots_Unique_Annotated, echo=TRUE}
#peak scatter plot
# Reshape the data
data_long <- clean_annotated_dataset %>% 
  pivot_longer(cols = starts_with("Sample"), names_to = "Sample", values_to = "Intensity")

#Remove non-finite values
data_long <- data_long %>%
  filter(is.finite(Intensity))

# Log-transform the metabolite values
data_long <- data_long %>%
  mutate(log10 = log10(Intensity))

# Plot intensity histogram
ggplot(data_long, aes(x = log10)) +
  geom_histogram(binwidth = 0.4, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Unique Metabolite Intensity",
       x = "Log10 Intensity",
       y = "Number of Unique Metablites")

# Plot the scatter plot
ggplot(data_long, aes(x = as.factor(Sample), y = log10)) +
  geom_point(size = 1, color = "darkorange") +
  theme_minimal() +
  labs(title = "Unique Metabolites per Sample",
       x = "Sample",
       y = "Log10 Intensity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# Calculate number of peaks per sample
  peak_count_full <- colSums(!is.na(clean_annotated_dataset[, 4:27]))
  peak_count_full <- data.frame(Sample = names(peak_count_full), Count = peak_count_full)

#Graph
ggplot(data = peak_count_full, aes(x = Sample, y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Number of Unique Metabolites per Sample",
       x = "Sample", y = "Metabolite Count")

#Run function for percentage of peaks missing across samples
 missing_stats_full <- calculate_missingness(clean_annotated_dataset)

#Plot
ggplot(missing_stats_full, aes(x = missing_proportion)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Missing Metabolites Across Samples",
       x = "Percentage of Samples Missing Metabolite",
       y = "Unique Metabolite Count") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust=-1.5, color="black", size=3.5, position = position_stack(vjust = 0.5))
  
```
```{r PCA_Unique_Annotated, echo = TRUE}
# PCA of sample variation
# Prepare data for PCA (samples as rows, metabolites as columns)
samples_matrix <- clean_annotated_dataset %>%
  select(starts_with("Sample")) %>%
  t()

# 1/5th min value imputation
samples_matrix <- apply(samples_matrix, 2, min_value_impute)

# Perform PCA
pca_samples <- prcomp(samples_matrix, scale. = TRUE)

# Convert PCA results to data frame and add Sample labels
sample_scores <- as.data.frame(pca_samples$x)
sample_scores$Sample <- rownames(sample_scores)

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for samples
ggplot(sample_scores, aes(x = PC1, y = PC2, label = Sample)) +
  geom_point(aes(color = Sample), size = 3) +
  geom_text(vjust = -1, hjust = -0.1) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]), title = "PCA of Sample Variation")

# PCA of metabolite variation
# Prepare data for PCA (metabolites as rows, samples as columns)
metabolites_matrix <- clean_annotated_dataset %>%
  select(starts_with("Sample"))

metabolites_matrix <- apply(metabolites_matrix, 2, min_value_impute)

# Perform PCA
pca_metabolites <- prcomp(metabolites_matrix, scale. = TRUE)

# Convert PCA results to data frame and add metabolite labels
metabolite_scores <- as.data.frame(pca_metabolites$x)
metabolite_scores$Compound.ID <- clean_annotated_dataset$Compound.ID

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Plot PCA for metabolites
ggplot(metabolite_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]), 
                   title = "PCA of Metabolite Variation")
```

# Assign Metadata
```{r Metadata, echo=TRUE}
# Convert metabolites data to matrix format
final_matrix <- clean_annotated_dataset %>%
  select(starts_with("Sample"))
  
metabolite_names <- clean_annotated_dataset[[31]]
final_matrix <- apply(final_matrix, 2, as.numeric)
rownames(final_matrix) <- metabolite_names

sample_metadata = read_excel('../../doc/untargeted/32_metadata.xlsx',sheet="metadata",range=NULL)

# Filter metadata to match metabolites samples
sample_metadata_filtered <- sample_metadata %>%
  filter(SampleName %in% colnames(final_matrix)) %>%
  # Ensure the order matches the metabolites data
  arrange(match(SampleName, colnames(final_matrix)))

# Create the metadata dataframe
sample_metadata_df <- as.data.frame(sample_metadata_filtered)
rownames(sample_metadata_df) <- sample_metadata_df$SampleName

# Create Biobase ExpressionSet with the matched data
sample_phenoData <- AnnotatedDataFrame(sample_metadata_df)

# Ensure the metabolite metadata matches the order of the final_matrix rows
metabolite_metadata_filtered <- unique_annotated_compounds %>%
  filter(InChiKey %in% rownames(final_matrix)) %>%
  arrange(match(InChiKey, rownames(final_matrix)))

# Create the metabolite metadata dataframe
metabolite_metadata_df <- as.data.frame(metabolite_metadata_filtered)
rownames(metabolite_metadata_df) <- metabolite_metadata_df$InChiKey

metabolite_featureData <- AnnotatedDataFrame(metabolite_metadata_df)

# Create the ExpressionSet with both phenodata and featuredata
final_annotated_df <- ExpressionSet(
  assayData = final_matrix,
  phenoData = sample_phenoData,
  featureData = metabolite_featureData
)

#Create Biobase ExpressionSet with imputed values for PCA
imputed_matrix <- apply(final_matrix, 2, min_value_impute)
imputed_annotated_df <- ExpressionSet(
  assayData = imputed_matrix,
  phenoData = sample_phenoData,
  featureData = metabolite_featureData
)
```

# PCA by Sample Variable
```{r Variable_Sample_PCA, echo = TRUE}
#PCA using imputed Biobase dataset
pca_samples <- prcomp(t(exprs(imputed_annotated_df)), scale. = TRUE)


# Create PCA data frame with metadata
pca_data <- data.frame(
  PC1 = pca_samples$x[,1],
  PC2 = pca_samples$x[,2],
  PC3 = pca_samples$x[,3],
  PC4 = pca_samples$x[,4],
  PC5 = pca_samples$x[,5],
  Sex = as.factor(pData(imputed_annotated_df)$Sex),
  GW = pData(imputed_annotated_df)$GW,
  Sample = rownames(pca_samples$x)
)

# Calculate variance explained
var_explained_samples <- pca_samples$sdev^2 / sum(pca_samples$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_samples),
  VarExp = var_explained_samples
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

# Sex PCA
ggplot(pca_data, aes(x = PC1, y = PC2, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))

# Sex PCA
ggplot(pca_data, aes(x = PC2, y = PC3, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))

# Sex PCA
ggplot(pca_data, aes(x = PC3, y = PC4, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))

# Sex PCA
ggplot(pca_data, aes(x = PC4, y = PC5, color = Sex, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       y = sprintf("PC5 (%.1f%%)", var_explained_samples[5]),
       title = "Sample PCA by Sex") +
  scale_color_manual(values = c("F" = "#FF69B4", "M" = "#4169E1")) +
  theme(plot.title = element_text(hjust = 0.5))


# GW PCA
ggplot(pca_data, aes(x = PC1, y = PC2, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_samples[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))

# GW PCA
ggplot(pca_data, aes(x = PC2, y = PC3, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC2 (%.1f%%)", var_explained_samples[2]),
       y = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))

# GW PCA
ggplot(pca_data, aes(x = PC3, y = PC4, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC3 (%.1f%%)", var_explained_samples[3]),
       y = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))

# GW PCA
ggplot(pca_data, aes(x = PC4, y = PC5, color = GW, label = Sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 10) +
  theme_minimal() +
  labs(x = sprintf("PC4 (%.1f%%)", var_explained_samples[4]),
       y = sprintf("PC5 (%.1f%%)", var_explained_samples[5]),
       title = "Sample PCA by Gestational Week") +
  scale_color_viridis_c() +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r PCA_Metabolites, echo=TRUE}

#PCA using imputed Biobase dataset
pca_metabolites <- prcomp(exprs(imputed_annotated_df), scale. = TRUE)

# Create PCA data frame with metadata
pca_data <- data.frame(
  PC1 = pca_metabolites$x[,1],
  PC2 = pca_metabolites$x[,2],
  PC3 = pca_metabolites$x[,3],
  PC4 = pca_metabolites$x[,4],
  Name = as.factor(fData(imputed_annotated_df)$Name),
  SuperClass = fData(imputed_annotated_df)$Super.Class,
  Class = fData(imputed_annotated_df)$Class,
  SubClass = fData(imputed_annotated_df)$Sub.Class,
  Pathways = fData(imputed_annotated_df)$Pathways,
  Metabolites = rownames(pca_metabolites$x)
)

# Calculate variance explained
var_explained_metabolites <- pca_metabolites$sdev^2 / sum(pca_metabolites$sdev^2) * 100

# Create Elbow Plot
elbow_data <- data.frame(
  PC = 1:length(var_explained_metabolites),
  VarExp = var_explained_metabolites
)

ggplot(elbow_data, aes(x = PC, y = VarExp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Principal Component",
       y = "Variance Explained (%)",
       title = "Scree Plot of PCA Variance Explained") +
  theme(plot.title = element_text(hjust = 0.5))

#SuperClass PCA
ggplot(pca_data, aes(x = PC1, y = PC2, color = SuperClass)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(x = sprintf("PC1 (%.1f%%)", var_explained_metabolites[1]),
       y = sprintf("PC2 (%.1f%%)", var_explained_metabolites[2]),
       title = "Metabolite PCA by SuperClass") +
  theme(plot.title = element_text(hjust = 0.5))

```
## Metabolite-Factor Correlations

Non-Imputed for Comparison
```{r nonimputed_correlations, echo=TRUE}
# Calculate correlations between metabolites and metadata
metadata <- pData(final_annotated_df)
sex_numeric <- as.numeric(metadata$Sex == "M")  # Convert sex to numeric (0=F, 1=M)
GW <- metadata$GW

# Calculate correlations for each metabolite
nonimputed_metadata_correlation <- data.frame(
  InChiKey = fData(imputed_annotated_df)$InChiKey,
  Name = fData(imputed_annotated_df)$Name,
  Sex_correlation = apply(exprs(final_annotated_df), 1, function(x) cor(x, sex_numeric)),
  GW_correlation = apply(exprs(final_annotated_df), 1, function(x) cor(x, GW))
)

# Create correlation strength plots
# For sex
top_sex_correlations <- nonimputed_metadata_correlation %>%
  arrange(desc(abs(Sex_correlation))) %>%
  head(15)

ggplot(top_sex_correlations, 
                               aes(x = reorder(Name, abs(Sex_correlation)), 
                                   y = Sex_correlation)) +
  geom_bar(stat = "identity", aes(fill = Sex_correlation > 0)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Correlation with Sex",
       title = "Top 15 Sex-Correlated Metabolites") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

# For GW
top_GW_correlations <- nonimputed_metadata_correlation %>%
  arrange(desc(abs(GW_correlation))) %>%
  head(15)

ggplot(top_GW_correlations, 
                               aes(x = reorder(Name, abs(GW_correlation)), 
                                   y = GW_correlation)) +
  geom_bar(stat = "identity", aes(fill = GW_correlation > 0)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Correlation with GW",
       title = "Top 15 GW-Correlated Metabolites") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

# Print top correlations
print(nonimputed_metadata_correlation %>% 
        arrange(desc(abs(Sex_correlation))) %>% 
        select(Name, Sex_correlation) %>% 
        head(10))

print(nonimputed_metadata_correlation %>% 
        arrange(desc(abs(GW_correlation))) %>% 
        select(Name, GW_correlation) %>% 
        head(10))
```

Imputed Correlation for statistical testing
```{r Metabolite_Correlations, echo=TRUE}
# Calculate correlations and p-values between metabolites and metadata
metadata <- pData(imputed_annotated_df)
sex_numeric <- as.numeric(metadata$Sex == "M")  # Convert sex to numeric (0=F, 1=M)
GW <- metadata$GW

# Function to calculate correlation and p-value
correlation_test <- function(x, y) {
  result <- cor.test(x, y)
  c(correlation = result$estimate, p_value = result$p.value)
}

# Calculate correlations and p-values for each metabolite
metabolite_samplemetadata_stats <- data.frame(
  InChiKey = fData(imputed_annotated_df)$InChiKey,
  Name = fData(imputed_annotated_df)$Name,
  Sex_corr_pval = t(apply(exprs(imputed_annotated_df), 1, function(x) correlation_test(x, sex_numeric))),
  GW_corr_pval = t(apply(exprs(imputed_annotated_df), 1, function(x) correlation_test(x, GW)))
)

metabolite_samplemetadata_stats <- metabolite_samplemetadata_stats %>%
  mutate(
    Sex_FDR = p.adjust(Sex_corr_pval.p_value, method = "fdr"),
    GW_FDR = p.adjust(GW_corr_pval.p_value, method = "fdr")
  )

# Scatter plot for significant Sex correlations
significant_sex <- metabolite_samplemetadata_stats %>% filter(Sex_FDR < 0.05)
scatter_sex <- ggplot(significant_sex, aes(x = reorder(Name, Sex_corr_pval.correlation.cor), y = Sex_corr_pval.correlation.cor)) +
  geom_point(aes(size = -log10(Sex_corr_pval.p_value), color = Sex_corr_pval.correlation.cor)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Correlation with Sex",
       title = "Significant Metabolite Correlations with Sex") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right", axis.text.x = element_text(angle = 90, hjust = 1))

# Scatter plot for significant GW correlations
significant_GW <- metabolite_samplemetadata_stats %>% filter(GW_FDR < 0.05)
scatter_GW <- ggplot(significant_GW, aes(x = reorder(Name, GW_corr_pval.correlation.cor), y = GW_corr_pval.correlation.cor)) +
  geom_point(aes(size = -log10(GW_corr_pval.p_value), color = GW_corr_pval.correlation.cor)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Correlation with GW",
       title = "Significant Metabolite Correlations with Gestational Weeks (GW)") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right", axis.text.x = element_text(angle = 90, hjust = 1))

# Heatmap of correlations
correlation_matrix <- metabolite_samplemetadata_stats %>%
  select(Sex_corr_pval.correlation.cor, GW_corr_pval.correlation.cor) %>%
  as.matrix()

rownames(correlation_matrix) <- metabolite_samplemetadata_stats$Name
colnames(correlation_matrix) <- c("Sex", "GW")

heatmap_plot <- Heatmap(correlation_matrix, name = "Correlation",
                        show_row_names = F, show_column_names = T)

# Plot the results
print(scatter_sex)
print(scatter_GW)
heatmap_plot

# Print top correlations with significance details
top_sex_correlations <- metabolite_samplemetadata_stats %>%
  arrange(desc(abs(Sex_corr_pval.correlation.cor))) %>%
  head(10) %>%
  select(Name, Sex_corr_pval.correlation.cor, Sex_corr_pval.p_value, Sex_FDR)
print(top_sex_correlations)

top_GW_correlations <- metabolite_samplemetadata_stats %>%
  arrange(desc(abs(GW_corr_pval.correlation.cor))) %>%
  head(10) %>%
  select(Name, GW_corr_pval.correlation.cor, GW_corr_pval.p_value, GW_FDR)
print(top_GW_correlations)
```


## Sex Differential Analysis
# Univariate Testing for Sex Differences with Volcano Plots
```{r Univariate_Testing, echo=TRUE, message=FALSE, warning=FALSE}

# Calculate p-values for sex differences using the Wilcoxon rank-sum test
metadata <- pData(imputed_annotated_df)
sex_numeric <- as.numeric(metadata$Sex == "M")  # Convert sex to numeric (0=F, 1=M)

# Function to calculate Wilcoxon rank-sum test p-value
wilcoxon_test <- function(x, sex) {
  result <- wilcox.test(x[sex == 0], x[sex == 1], exact = FALSE)
  c(statistic = result$statistic, p_value = result$p.value)
}

# Perform Wilcoxon rank-sum test for each metabolite
metabolite_sex_stats <- data.frame(
  InChiKey = fData(imputed_annotated_df)$InChiKey[apply(exprs(imputed_annotated_df), 1, function(x) all(is.finite(x)))],
  Name = fData(imputed_annotated_df)$Name[apply(exprs(imputed_annotated_df), 1, function(x) all(is.finite(x)))],
  Wilcoxon_stat_pval = t(apply(imputed_annotated_df, 1, function(x) wilcoxon_test(x, sex_numeric)))
)

# Calculate log fold changes for volcano plot
metabolite_sex_stats <- metabolite_sex_stats %>%
  mutate(
    logFC = log2(apply(imputed_annotated_df[, sex_numeric == 1], 1, mean) / apply(imputed_annotated_df[, sex_numeric == 0], 1, mean)),
    FDR = p.adjust(Wilcoxon_stat_pval.p_value, method = "fdr"),
    Bonferroni = p.adjust(Wilcoxon_stat_pval.p_value, method = "bonferroni")
  )

# Volcano plot with FDR correction
volcano_FDR <- ggplot(metabolite_sex_stats, aes(x = logFC, y = -log10(Wilcoxon_stat_pval.p_value))) +
  geom_point(aes(color = FDR < 0.05)) +
  theme_minimal() +
  labs(x = "Log2 Fold Change",
       y = "-log10(p-value)",
       title = "Volcano Plot with FDR Correction") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right")

# Volcano plot with Bonferroni correction
volcano_Bonferroni <- ggplot(metabolite_sex_stats, aes(x = logFC, y = -log10(Wilcoxon_stat_pval.p_value))) +
  geom_point(aes(color = Bonferroni < 0.05)) +
  theme_minimal() +
  labs(x = "Log2 Fold Change",
       y = "-log10(p-value)",
       title = "Volcano Plot with Bonferroni Correction") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "right")

# Plot the results
print(volcano_FDR)
print(volcano_Bonferroni)

# Print top metabolites with significance details
top_metabolites <- metabolite_sex_stats %>%
  arrange(desc(-log10(Wilcoxon_stat_pval.p_value))) %>%
  head(10) %>%
  select(Name, logFC, Wilcoxon_stat_pval.p_value, FDR, Bonferroni)
print(top_metabolites)

```
# Multivariate testing
```{r Multivariate_Model, echo=TRUE}
metadata <- pData(imputed_annotated_df)
metadata$Sex <- as.factor(metadata$Sex)
sex_numeric <- as.numeric(metadata$Sex == "M")  # Convert sex to numeric (0=F, 1=M)
GW <- metadata$GW

metabolite_matrix <- exprs(imputed_annotated_df)

# Fit linear model and collect coefficients and p-values
lm_results <- lapply(1:nrow(metabolite_matrix), function(i) {
  metabolite <- metabolite_matrix[i, ]
  tryCatch({
    model <- lm(metabolite ~ Sex + GW, data = metadata)
    summary(model)
  }, error = function(e) {
    warning_message <- paste("Model fitting failed for row", i, "due to error:", e$message)
    message(warning_message)
    NULL
  })
})

# Extract coefficients and p-values
coefficients_list <- lapply(lm_results, function(res) {
  if (is.null(res) || !inherits(res, "summary.lm")) {
    return(c(Sex_estimate = NA, Sex_p_value = NA, GW_estimate = NA, GW_p_value = NA))
  }
  coef_table <- res$coefficients
  if (all(c("SexM", "GW") %in% rownames(coef_table))) {
    c(Sex_estimate = coef_table["SexM", "Estimate"],
      Sex_p_value = coef_table["SexM", "Pr(>|t|)"],
      GW_estimate = coef_table["GW", "Estimate"],
      GW_p_value = coef_table["GW", "Pr(>|t|)"])
  } else {
    c(Sex_estimate = NA, Sex_p_value = NA, GW_estimate = NA, GW_p_value = NA)
  }
})

coefficients_df <- do.call(rbind, coefficients_list)
metabolite_lm_stats <- data.frame(
  InChiKey = fData(imputed_annotated_df)$InChiKey[1:nrow(coefficients_df)],
  Name = fData(imputed_annotated_df)$Name[1:nrow(coefficients_df)],
  coefficients_df
)

# Adjust p-values for multiple testing
metabolite_lm_stats <- metabolite_lm_stats %>%
  mutate(
    Sex_FDR = p.adjust(Sex_p_value, method = "fdr"),
    GW_FDR = p.adjust(GW_p_value, method = "fdr"),
    Sex_Bonferroni = p.adjust(Sex_p_value, method = "bonferroni"),
    GW_Bonferroni = p.adjust(GW_p_value, method = "bonferroni")
  )

# Scatter plot for significant Sex effects
significant_sex <- metabolite_lm_stats %>% filter(Sex_FDR < 0.05)
scatter_sex <- ggplot(significant_sex, aes(x = reorder(Name, Sex_estimate), y = Sex_estimate)) +
  geom_point(aes(size = -log10(Sex_p_value), color = Sex_FDR < 0.05)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Effect Size of Sex",
       title = "Significant Metabolite Effects due to Sex") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right", axis.text.x = element_text(angle = 90, hjust = 1))


# Scatter plot for significant GW effects
significant_GW <- metabolite_lm_stats %>% filter(GW_FDR < 0.05)
scatter_GW <- ggplot(significant_GW, aes(x = reorder(Name, GW_estimate), y = GW_estimate)) +
  geom_point(aes(size = -log10(GW_p_value), color = GW_FDR < 0.05)) +
  coord_flip() +
  theme_minimal() +
  labs(x = "Metabolite",
       y = "Effect Size of GW",
       title = "Significant Metabolite Effects due to Gestational Weeks (GW)") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "right", axis.text.x = element_text(angle = 90, hjust = 1))

# Plot the results
print(scatter_sex)
print(scatter_GW)

# Print top metabolites with significance details
top_sex_metabolites <- metabolite_lm_stats %>%
  arrange(Sex_p_value) %>%
  head(10) %>%
  select(Name, Sex_estimate, Sex_p_value, Sex_FDR, Sex_Bonferroni)
print(top_sex_metabolites)

top_GW_metabolites <- metabolite_lm_stats %>%
  arrange(GW_p_value) %>%
  head(10) %>%
  select(Name, GW_estimate, GW_p_value, GW_FDR, GW_Bonferroni)
print(top_GW_metabolites)
```

