---
title: "Hormone–Metabolite Correlations"
author: "Sneha Chaturvedi"
date: "`r Sys.Date()`"
format:
  revealjs:
    theme: simple
    slide-number: true
    scrollable: true
    embed-resources: false
freeze: auto
execute:
  echo: false
  warning: false
  message: false
---

## {visibility="hidden"}

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(ppcor)
  library(svglite)
  library(patchwork)
  library(knitr)
})

path_hormones <- "../../doc/targeted/targeted_hormones.xlsx"
path_peak     <- "../../results/untargeted/batch2_peak_area_clean.csv"
path_annot    <- "../../results/untargeted/batch2_metabolite_annotations_flagged.csv"

out_dir <- "../../results/hormone-metabolite"
dir.create(file.path(out_dir, "svg"), showWarnings = FALSE, recursive = TRUE)

# Analysis parameters
MIN_BEST_MATCH <- 80    # High-confidence annotations only
MIN_N_TT       <- 10    # Males only; hard floor per Bonett & Wright 2000
MIN_N_E2P4     <- 12    # Sex-combined; half of total n = 24
FDR_THRESH     <- 0.10  
RHO_THRESH     <- 0.40  # |ρ| ≥ 0.40 moderate correlation
TOP_N          <- 10    # Scatter plots per hormone
```

## {visibility="hidden"}

```{r helpers}
# Local copy — avoids loading full pseudobulk_functions.R
save_dual_format <- function(plot, output_dir, filename_base,
                              width = 10, height = 7, dpi = 300) {
  dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
  svg_d <- file.path(output_dir, "svg")
  dir.create(svg_d, showWarnings = FALSE)
  ggsave(file.path(output_dir, paste0(filename_base, ".png")),
         plot = plot, width = width, height = height, dpi = dpi)
  ggsave(file.path(svg_d,    paste0(filename_base, ".svg")),
         plot = plot, width = width, height = height,
         device = svglite::svglite)
  message("Saved: ", filename_base, ".png  +  svg/", filename_base, ".svg")
  invisible(plot)
}

theme_nature <- function(base_size = 9) {
  theme_classic(base_size = base_size) +
    theme(
      text               = element_text(family = "Helvetica", colour = "black"),
      axis.line          = element_line(colour = "black", linewidth = 0.4),
      axis.ticks         = element_line(colour = "black", linewidth = 0.4),
      axis.text          = element_text(colour = "black"),
      panel.grid         = element_blank(),
      legend.background  = element_blank(),
      legend.key         = element_blank(),
      # Subtle strip for Super.Class facets in heatmap
      strip.background   = element_rect(fill = "grey96", colour = "grey80",
                                         linewidth = 0.3),
      strip.text         = element_text(size = 7.5, colour = "black"),
      plot.title         = element_text(face = "bold", size = base_size + 1),
      plot.subtitle      = element_text(size = base_size - 1, colour = "grey40")
    )
}
```

## {visibility="hidden"}

```{r correlation-functions}
# Raw Spearman — supplementary; pairwise complete observations
run_raw_spearman <- function(df, hormone_col, met_ids, min_n = 5) {
  map_dfr(met_ids, function(met) {
    sub <- df |> dplyr::select(h = all_of(hormone_col), m = all_of(met)) |> drop_na()
    if (nrow(sub) < min_n) return(NULL)
    ct <- cor.test(sub$h, sub$m, method = "spearman", exact = FALSE)
    tibble(compound_id = met,
           rho = unname(ct$estimate),
           p   = ct$p.value,
           n   = nrow(sub))
  })
}

# Partial Spearman — primary analysis; single covariate via ppcor
#   TT:   males only, covariate = GW   (fetal TT varies with gestational age)
#   E2:   sex-combined, covariate = Sex (binary, 0/1)
#   P4:   sex-combined, covariate = Sex
run_partial_spearman <- function(df, hormone_col, met_ids, covariate_col, min_n) {
  map_dfr(met_ids, function(met) {
    sub <- df |>
      dplyr::select(h = all_of(hormone_col),
             m = all_of(met),
             z = all_of(covariate_col)) |>
      drop_na()
    if (nrow(sub) < min_n) return(NULL)
    tryCatch({
      res <- ppcor::pcor.test(sub$h, sub$m, sub$z, method = "spearman")
      tibble(compound_id = met,
             rho = res$estimate,
             p   = res$p.value,
             n   = res$n)
    }, error = function(e) NULL)
  })
}

add_fdr <- function(df) mutate(df, fdr = p.adjust(p, method = "BH"))
add_sig <- function(df) mutate(df, sig = fdr < FDR_THRESH & abs(rho) >= RHO_THRESH)
```

## Data Loading & QC

```{r load-data}
hormones_raw <- read_xlsx(path_hormones)

detect_outliers <- function(data, variable) {
  limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) +
           1.5 * IQR(data[[variable]], na.rm = TRUE)
  filter(data, .data[[variable]] <= limit)
}

# Each analysis uses its own outlier-filtered sample set so that an outlier
# in one hormone does not exclude samples from the other two analyses.
horm_TT <- detect_outliers(hormones_raw, "TT (ng/mL)") |>
  dplyr::select(Sample, Sex, GW, TT = `TT (ng/mL)`) |>
  mutate(Sample = as.character(Sample)) |>
  filter(Sex == "M")                        # TT: males only

horm_E2 <- detect_outliers(hormones_raw, "17B-E2 (ng/mL)") |>
  dplyr::select(Sample, Sex, GW, E2 = `17B-E2 (ng/mL)`) |>
  mutate(Sample  = as.character(Sample),
         sex_num = as.integer(Sex == "M"))

horm_P4 <- detect_outliers(hormones_raw, "P4 (ng/mL)") |>
  dplyr::select(Sample, Sex, GW, P4 = `P4 (ng/mL)`) |>
  mutate(Sample  = as.character(Sample),
         sex_num = as.integer(Sex == "M"))

# Metabolomics annotations
annot_clean <- read_csv(path_annot, show_col_types = FALSE)

# Peak areas: log2(x+1) transform → pivot to sample × metabolite
peak_raw  <- read_csv(path_peak, show_col_types = FALSE)
smpl_cols <- names(peak_raw)[str_detect(names(peak_raw), "^Sample")]

peak_wide <- peak_raw |>
  mutate(across(all_of(smpl_cols), ~ log2(.x + 1))) |>
  dplyr::select(Compound.ID, all_of(smpl_cols)) |>
  pivot_longer(-Compound.ID, names_to = "Sample", values_to = "log2_area") |>
  pivot_wider(names_from = "Compound.ID", values_from = "log2_area")

# One merged dataset per hormone
merged_TT <- inner_join(horm_TT, peak_wide, by = "Sample")
merged_E2 <- inner_join(horm_E2, peak_wide, by = "Sample")
merged_P4 <- inner_join(horm_P4, peak_wide, by = "Sample")

# Single met_ids; metabolite filtering handled by min_n inside the loop
met_ids <- intersect(annot_clean$Compound.ID, names(merged_E2))
```

```{r qc-table}
#| results: asis
tibble(
  Metric = c(
    "TT: samples post-outlier (males only)",
    "E2: samples post-outlier",
    "P4: samples post-outlier",
    "Metabolomics samples (peak file)",
    "TT: matched samples",
    "E2: matched samples",
    "P4: matched samples",
    "Metabolites tested (min_n filter applied per pair)"
  ),
  N = c(
    nrow(horm_TT),
    nrow(horm_E2), nrow(horm_P4),
    length(smpl_cols),
    nrow(merged_TT), nrow(merged_E2), nrow(merged_P4),
    length(met_ids)
  )
) |> knitr::kable(format = "html")
```

## Analysis Overview

:::: {.columns}

::: {.column width="55%"}
**Raw Spearman** *(supplementary only)*

- Pairwise complete observations
- FDR correction within each hormone (BH)
- Saved to CSV; not shown in figures

**Partial Spearman** *(primary)*

| Hormone | Subset | Covariate | min *n* |
|---------|--------|-----------|---------|
| TT | Males | GW | 10 |
| E2 | All | Sex | 12 |
| P4 | All | Sex | 12 |
:::

::: {.column width="45%"}
**Significance thresholds**

- FDR < 0.10 (BH within hormone)
- |ρ| ≥ 0.40 (moderate correlation)

**Outputs**

- `raw_spearman_all_hormones.csv`
- `partial_spearman_all_hormones.csv`
- `significant_hits_summary.csv`
- `significant_hits_pathway_annotated.csv`
- Heatmap + scatter plots (PNG + SVG)
:::

::::

## Correlation Results

```{r run-correlations}
message("Running raw Spearman correlations...")

raw_TT <- run_raw_spearman(merged_TT, "TT", met_ids, min_n = MIN_N_TT) |>
  add_fdr() |> dplyr::mutate(hormone = "TT", subset = "Males")
raw_E2 <- run_raw_spearman(merged_E2, "E2", met_ids, min_n = MIN_N_E2P4) |>
  add_fdr() |> dplyr::mutate(hormone = "E2", subset = "Sex-combined")
raw_P4 <- run_raw_spearman(merged_P4, "P4", met_ids, min_n = MIN_N_E2P4) |>
  add_fdr() |> dplyr::mutate(hormone = "P4", subset = "Sex-combined")

raw_all <- bind_rows(raw_TT, raw_E2, raw_P4) |>
  left_join(annot_clean |> dplyr::select(Compound.ID, Name, Super.Class, Pathways),
            by = c("compound_id" = "Compound.ID")) |>
  dplyr::select(hormone, subset, compound_id, Name, Super.Class, Pathways, rho, p, fdr, n) |>
  arrange(hormone, fdr, desc(abs(rho)))

write_csv(raw_all, file.path(out_dir, "raw_spearman_all_hormones.csv"))

message("Running partial Spearman correlations...")

adj_TT <- run_partial_spearman(merged_TT, "TT", met_ids, "GW",
                                MIN_N_TT) |>
  add_fdr() |> add_sig() |>
  mutate(hormone = "TT", covariate = "GW", subset = "Males")

adj_E2 <- run_partial_spearman(merged_E2, "E2", met_ids, "sex_num",
                                MIN_N_E2P4) |>
  add_fdr() |> add_sig() |>
  mutate(hormone = "E2", covariate = "Sex", subset = "Sex-combined")

adj_P4 <- run_partial_spearman(merged_P4, "P4", met_ids, "sex_num",
                                MIN_N_E2P4) |>
  add_fdr() |> add_sig() |>
  mutate(hormone = "P4", covariate = "Sex", subset = "Sex-combined")

adj_all <- bind_rows(adj_TT, adj_E2, adj_P4) |>
  left_join(annot_clean |> dplyr::select(Compound.ID, Name, Super.Class, Pathways),
            by = c("compound_id" = "Compound.ID")) |>
  dplyr::select(hormone, subset, covariate, compound_id, Name,
         Super.Class, Pathways, rho, p, fdr, sig, n) |>
  arrange(hormone, fdr, desc(abs(rho)))

sig_hits <- adj_all |> filter(sig)

write_csv(adj_all,  file.path(out_dir, "partial_spearman_all_hormones.csv"))
write_csv(sig_hits, file.path(out_dir, "significant_hits_summary.csv"))
```

```{r results-summary-table}
#| results: asis
tibble(
  Hormone   = c("TT (Males)", "E2", "P4"),
  Subset    = c("Males only", "Sex-combined", "Sex-combined"),
  Covariate = c("GW", "Sex", "Sex"),
  Tested    = c(nrow(adj_TT), nrow(adj_E2), nrow(adj_P4)),
  `FDR < 0.10 & |ρ| ≥ 0.40` = c(
    sum(adj_TT$sig, na.rm = TRUE),
    sum(adj_E2$sig, na.rm = TRUE),
    sum(adj_P4$sig, na.rm = TRUE)
  )
) |> knitr::kable(format = "html")
```

## Heatmap

```{r heatmap}
sig_met_ids <- unique(sig_hits$compound_id)

if (length(sig_met_ids) == 0) {

  cat("No metabolites reached significance (FDR < 0.10 and |ρ| ≥ 0.40).")

} else {

  hm_data <- adj_all |>
    filter(compound_id %in% sig_met_ids) |>
    mutate(
      hormone     = factor(hormone, levels = c("TT", "E2", "P4")),
      sig_label   = ifelse(sig, "*", ""),
      super_class = str_wrap(Super.Class, width = 22)
    )

  # Order metabolites: group by Super.Class, then by |ρ| for TT within class
  tt_order <- hm_data |>
    filter(hormone == "TT") |>
    arrange(super_class, desc(abs(rho))) |>
    pull(compound_id)

  remaining <- setdiff(sig_met_ids, tt_order)
  if (length(remaining) > 0) {
    extra <- hm_data |>
      filter(compound_id %in% remaining) |>
      distinct(compound_id, super_class) |>
      arrange(super_class) |>
      pull(compound_id)
    tt_order <- c(tt_order, extra)
  }

  name_map  <- hm_data |> distinct(compound_id, Name)       |> deframe()
  class_map <- hm_data |> distinct(compound_id, super_class) |> deframe()

  hm_data <- hm_data |>
    mutate(
      met_label  = factor(name_map[compound_id],
                           levels = rev(name_map[tt_order])),
      super_class = factor(class_map[compound_id],
                            levels = unique(class_map[tt_order]))
    )

  hm_height <- max(4.5, length(sig_met_ids) * 0.45 + 1.5)

  p_heatmap <- ggplot(hm_data,
                       aes(x = hormone, y = met_label, fill = rho)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(aes(label = sig_label), size = 5.5, colour = "black",
              vjust = 0.75) +
    facet_grid(super_class ~ ., scales = "free_y", space = "free_y") +
    scale_fill_gradient2(
      low = "#2166AC", mid = "white", high = "#D6604D",
      midpoint = 0, limits = c(-1, 1),
      name = "Partial ρ",
      breaks = c(-1, -0.5, 0, 0.5, 1)
    ) +
    scale_x_discrete(
      labels = c("TT" = "TT\n(Males)", "E2" = "E2", "P4" = "P4")
    ) +
    labs(
      x        = NULL,
      y        = NULL,
      title    = "Hormone–Metabolite Partial Spearman Correlations",
      subtitle = "* FDR < 0.10, |ρ| ≥ 0.40"
    ) +
    theme_nature(base_size = 9) +
    theme(
      axis.text.y     = element_text(size = 8),
      axis.text.x     = element_text(size = 9, face = "bold"),
      legend.position = "right"
    )

  save_dual_format(p_heatmap, out_dir, "heatmap_hormone_metabolite",
                   width = 5.5, height = hm_height)
  p_heatmap
}
```

## Scatter Plots

```{r scatter-functions}
make_scatter <- function(df, hormone_col, met_id, met_name,
                          rho_val, fdr_val, color_by_sex, hormone_label) {
  sub <- df |>
    dplyr::select(h = all_of(hormone_col), m = all_of(met_id), Sex) |>
    drop_na()

  annotation <- sprintf("Partial \u03c1 = %.2f\nFDR = %.3f", rho_val, fdr_val)

  p <- ggplot(sub, aes(x = h, y = m)) +
    geom_smooth(method = "lm", se = TRUE, colour = "grey70",
                linewidth = 0.5, alpha = 0.2) +
    { if (color_by_sex) {
        list(
          geom_point(aes(colour = Sex), size = 2.2, alpha = 0.85),
          scale_colour_manual(
            values = c("M" = "#4575B4", "F" = "#D73027"),
            labels = c("M" = "Male",    "F" = "Female"),
            name   = NULL
          )
        )
      } else {
        list(geom_point(colour = "#4575B4", size = 2.2, alpha = 0.85))
      }
    } +
    annotate("text",
             x = -Inf, y = Inf, label = annotation,
             hjust = -0.08, vjust = 1.5,
             size = 2.4, fontface = "italic", colour = "grey30") +
    labs(
      x     = paste0(hormone_label, " (ng/mL)"),
      y     = "log\u2082(peak area + 1)",
      title = met_name
    ) +
    theme_nature(base_size = 8) +
    theme(
      legend.position  = "bottom",
      legend.key.size  = unit(0.3, "cm"),
      plot.title       = element_text(size = 7.5, face = "plain")
    )
  p
}

make_scatter_panel <- function(adj_df, df, hormone_col, hormone_label,
                                color_by_sex, out_dir) {
  hits <- adj_df |>
    filter(sig) |>
    arrange(desc(abs(rho))) |>
    slice_head(n = TOP_N) |>
    left_join(annot_clean |> dplyr::select(Compound.ID, Name),
              by = c("compound_id" = "Compound.ID"))

  if (nrow(hits) == 0) {
    message("No significant hits for ", hormone_label)
    return(invisible(NULL))
  }

  plots <- purrr::map(seq_len(nrow(hits)), function(i) {
    r <- hits[i, ]
    make_scatter(df, hormone_col, r$compound_id, r$Name,
                  r$rho, r$fdr, color_by_sex, hormone_label)
  })

  n_plots <- length(plots)
  nc      <- min(2, n_plots)
  nr      <- ceiling(n_plots / nc)

  panel <- wrap_plots(plots, ncol = nc) +
    plot_annotation(
      title    = paste0("Top ", n_plots, " Metabolite Correlates: ", hormone_label),
      subtitle = "Partial Spearman | FDR < 0.10, |ρ| ≥ 0.40 | Ordered by |ρ|",
      theme    = theme(
        plot.title    = element_text(family = "Helvetica", face = "bold",
                                      size = 10, colour = "black"),
        plot.subtitle = element_text(family = "Helvetica",
                                      size = 8, colour = "grey40")
      )
    )

  fname <- paste0("scatter_top", n_plots, "_", tolower(hormone_col))
  save_dual_format(panel, out_dir, fname,
                   width = nc * 3.6, height = nr * 3.3)
  panel
}
```

## Scatter Plots — TT (Males)

```{r scatter-tt}
p_tt <- make_scatter_panel(adj_TT, merged_TT, "TT", "TT",
                            color_by_sex = FALSE, out_dir)
if (!is.null(p_tt)) p_tt else cat("No significant TT hits.")
```

## Scatter Plots — E2

```{r scatter-e2}
p_e2 <- make_scatter_panel(adj_E2, merged_E2, "E2", "E2",
                            color_by_sex = TRUE, out_dir)
if (!is.null(p_e2)) p_e2 else cat("No significant E2 hits.")
```

## Scatter Plots — P4

```{r scatter-p4}
p_p4 <- make_scatter_panel(adj_P4, merged_P4, "P4", "P4",
                            color_by_sex = TRUE, out_dir)
if (!is.null(p_p4)) p_p4 else cat("No significant P4 hits.")
```

## Pathway Summary

```{r pathway-summary}
#| results: asis

if (nrow(sig_hits) == 0) {
  cat("No significant hits to summarize.")
} else {
  pathway_tbl <- sig_hits |>
    left_join(annot_clean |> dplyr::select(Compound.ID, Class),
              by = c("compound_id" = "Compound.ID")) |>
    dplyr::select(
      Hormone       = hormone,
      Subset        = subset,
      Name,
      `Super Class` = Super.Class,
      Class,
      Pathways,
      rho, fdr, n
    ) |>
    mutate(across(c(rho, fdr), ~ round(.x, 3))) |>
    arrange(Hormone, fdr)

  knitr::kable(pathway_tbl, format = "html")

  write_csv(pathway_tbl,
            file.path(out_dir, "significant_hits_pathway_annotated.csv"))
}
```
