---
title: "Batch 2 Pathway Analysis (MSEA)"
author: "Sneha Michaela Chaturvedi"
date: "Last Updated `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
---

Code developed by author with assistance from Claude (Sonnet 4.6). Output checked and validated by author.

# Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

Packages <- c("readxl", "tidyverse", "fgsea", "KEGGREST", "knitr",
              "kableExtra", "svglite", "ggrepel")

lapply(Packages, library, character.only = TRUE)
```

```{r source_functions, include=FALSE}
source("functions.R")
```

```{r paths, include=FALSE}
png_path <- "../../results/untargeted/batch2_clean"
svg_path <- "../../results/untargeted/batch2_clean/svg"
csv_path <- "../../results/untargeted/batch2_clean/csv"
```

## Load Data

```{r load_annotation}
# Full annotation file — needed for KEGG IDs and Pathways column
batch2_compound_annotation_full <- read_csv(
  "../../results/untargeted/Batch2_metabolite_annotations.csv",
  show_col_types = FALSE
)

# Peak area file defines the compounds that survived filtering
batch2_peak_area <- read_csv(
  "../../results/untargeted/Batch2_peak_area_values.csv",
  show_col_types = FALSE
)

# Restrict annotation to measured compounds only
batch2_compound_annotation <- batch2_compound_annotation_full %>%
  filter(Compound.ID %in% batch2_peak_area$Compound.ID)

cat("Total annotation rows:   ", nrow(batch2_compound_annotation_full), "\n")
cat("Measured compounds:      ", nrow(batch2_compound_annotation), "\n")

# Pathway ID → name lookup sheet from the core's Excel file
# Columns: pathway names, pathway source, pathwayid
pathway_info <- read_excel(
  "../../doc/untargeted/24_untargeted_developing_2025.xlsx",
  sheet = "Pathways"
) %>%
  rename_with(~c("pathway_name", "pathway_source", "pathway_id"),
              everything())
```

```{r load_limma}
# Limma results from Batch2_Sex_and_GW_Analysis.Rmd
# topTable output includes t-statistic, which is the ranking variable for MSEA
gw_limma  <- read_csv(file.path(csv_path, "batch2_gw_limma_results.csv"),
                      show_col_types = FALSE)
sex_limma <- read_csv(file.path(csv_path, "batch2_sex_limma_results.csv"),
                      show_col_types = FALSE)

cat("GW limma:  ", nrow(gw_limma),  "metabolites\n")
cat("Sex limma: ", nrow(sex_limma), "metabolites\n")
```

---

# Pathway Annotation

## Parse Existing Pathways Column

```{r parse_pathways}
# Pathways column format: "kegg:map00620;reactome:R-HSA-196807;wiki:WP2456"
# Split per metabolite into one row per pathway ID, retaining source prefix

pathways_parsed <- batch2_compound_annotation %>%
  filter(!is.na(Pathways) & Pathways != "") %>%
  dplyr::select(Compound.ID, Name, KEGG, Pathways) %>%
  separate_rows(Pathways, sep = ";") %>%
  mutate(
    pathway_source = str_extract(Pathways, "^[^:]+"),   # "kegg", "reactome", "wiki"
    pathway_id     = str_remove(Pathways, "^[^:]+:")    # "map00620", "R-HSA-...", "WP..."
  ) %>%
  dplyr::select(-Pathways) %>%
  filter(!is.na(pathway_id) & pathway_id != "")

cat("Metabolites with any pathway annotation (existing):",
    n_distinct(pathways_parsed$Name), "of", nrow(batch2_compound_annotation), "\n")
cat("Unique pathways (existing):", n_distinct(pathways_parsed$pathway_id), "\n")
cat("\nBreakdown by source:\n")
print(count(pathways_parsed, pathway_source))
```

## KEGGREST Re-Mapping

```{r keggrest_remap}
# Use the KEGG column (KEGG compound IDs, e.g. "C00031") to fetch
# additional pathway annotations for metabolites missing from Pathways column.
# With only ~80 compounds to query, per-compound keggLink() calls are feasible.

metabolites_missing_pathways <- batch2_compound_annotation %>%
  filter(is.na(Pathways) | Pathways == "") %>%
  filter(!is.na(KEGG) & KEGG != "") %>%
  dplyr::select(Compound.ID, Name, KEGG)

cat("Metabolites missing pathway annotation:", nrow(batch2_compound_annotation %>%
      filter(is.na(Pathways) | Pathways == "")), "\n")
cat("Of those, with KEGG ID available:", nrow(metabolites_missing_pathways), "\n")

# Per-compound KEGG pathway lookup.
# keggLink(target, source) with target = "pathway", source = "cpd:CXXXXX"
# returns map-level pathway IDs (e.g. "path:map00010") for that compound.
# 81 queries is feasible; Sys.sleep(0.35) stays within KEGG's rate limit.
cat("\nQuerying KEGGREST per compound (", nrow(metabolites_missing_pathways), "queries)...\n")

query_kegg_pathways <- function(kegg_ids) {
  results <- vector("list", length(kegg_ids))
  for (i in seq_along(kegg_ids)) {
    kid <- kegg_ids[i]
    results[[i]] <- tryCatch({
      links <- keggLink("pathway", paste0("cpd:", kid))
      # links: names = "cpd:CXXXXX", values = "path:mapXXXXX"
      tibble(KEGG = kid, pathway_id = str_remove(links, "path:"))
    }, error = function(e) {
      tibble(KEGG = character(), pathway_id = character())
    })
    if (i %% 10 == 0) cat("  ", i, "/", length(kegg_ids), "\n")
    Sys.sleep(0.35)   # KEGG rate limit: ~3 requests/sec
  }
  bind_rows(results)
}

kegg_link_df <- query_kegg_pathways(metabolites_missing_pathways$KEGG)
```

```{r build_kegg_new, dependson="keggrest_remap"}
if (nrow(kegg_link_df) > 0) {
  kegg_new <- metabolites_missing_pathways %>%
    inner_join(kegg_link_df, by = "KEGG") %>%
    mutate(pathway_source = "kegg") %>%
    dplyr::select(Compound.ID, Name, KEGG, pathway_source, pathway_id)

  cat("New KEGG pathway assignments retrieved:", nrow(kegg_new), "\n")
  cat("Metabolites newly covered:", n_distinct(kegg_new$Name), "\n")
} else {
  kegg_new <- tibble(Compound.ID = character(), Name = character(),
                     KEGG = character(), pathway_source = character(),
                     pathway_id = character())
  cat("No KEGG pathway links returned — proceeding with existing annotations only.\n")
}
```

## Combine and Map Pathway Names

```{r combine_pathway_annotations}
# Stack existing + KEGG-remapped annotations
# For metabolites that already had pathway annotations, keep them as-is;
# KEGG new rows only add metabolites that had no annotations before.
pathways_combined <- bind_rows(
  pathways_parsed,
  kegg_new %>% dplyr::select(Compound.ID, Name, pathway_source, pathway_id)
)

# Attach human-readable pathway names from the core's Pathways sheet
# pathway_info has: pathway_name, pathway_source, pathway_id
pathways_combined <- pathways_combined %>%
  left_join(
    pathway_info %>% dplyr::select(pathway_id, pathway_name),
    by = "pathway_id"
  ) %>%
  mutate(
    # For KEGG pathway IDs without a name in the sheet, fetch label from ID
    pathway_label = case_when(
      !is.na(pathway_name) & pathway_name != "" ~ pathway_name,
      pathway_source == "kegg"      ~ paste0("KEGG:", pathway_id),
      pathway_source == "reactome"  ~ paste0("Reactome:", pathway_id),
      pathway_source == "wiki"      ~ paste0("WikiPathways:", pathway_id),
      TRUE ~ pathway_id
    )
  )

cat("\n=== Final Pathway Coverage ===\n")
cat("Metabolites with any pathway annotation:",
    n_distinct(pathways_combined$Name), "of", nrow(batch2_compound_annotation), "\n")
cat("Unique pathways:", n_distinct(pathways_combined$pathway_id), "\n\n")
cat("Breakdown by source:\n")
print(count(pathways_combined, pathway_source))
```

## Build Pathway Sets

```{r build_pathway_sets}
# MSEA requires: list(pathway_label = character vector of metabolite Names)
# Filter: minimum 5 and maximum 500 members in the measured background
# Use pathway_label (human-readable) as the set name

# For duplicate pathway IDs across sources, keep distinct Name-pathway_id pairs
pathway_sets_raw <- pathways_combined %>%
  distinct(Name, pathway_id, pathway_label) %>%
  filter(Name %in% gw_limma$Name)    # restrict to metabolites in our analysis

pathway_sets <- pathway_sets_raw %>%
  group_by(pathway_label) %>%
  summarise(members = list(unique(Name)), .groups = "drop") %>%
  mutate(n_members = lengths(members)) %>%
  filter(n_members >= 5, n_members <= 500)

# Convert to named list for fgsea
pathway_list <- setNames(pathway_sets$members, pathway_sets$pathway_label)

cat("Pathway sets available for MSEA:", length(pathway_list), "\n")
cat("Median members per set:", median(lengths(pathway_list)), "\n")
cat("Range:", min(lengths(pathway_list)), "–", max(lengths(pathway_list)), "\n")
```

---

# MSEA

MSEA (Metabolite Set Enrichment Analysis) is the metabolomics analog of GSEA. Rather
than requiring a significance cutoff, it uses the **full ranked list** of metabolites
ordered by their limma t-statistic. This is appropriate here given sparse individual
hits (especially for sex), as it uses all available signal.

The ranking statistic is the limma **t-statistic**, which combines effect size and
precision. Implemented via `fgsea` with 10,000 permutations.

```{r msea_function}
run_msea <- function(limma_results, pathway_list, analysis_label,
                     nperm = 10000, seed = 42) {
  # Build named ranked vector: Name -> t-statistic
  # Remove any NAs (metabolites where model didn't converge)
  ranked <- limma_results %>%
    filter(!is.na(t)) %>%
    arrange(desc(t)) %>%
    dplyr::select(Name, t) %>%
    deframe()                      # named numeric vector

  cat("\n===", analysis_label, "===\n")
  cat("Metabolites in ranked list:", length(ranked), "\n")
  cat("Pathway sets to test:", length(pathway_list), "\n")

  # Restrict pathway sets to metabolites present in this ranked list
  pathway_list_filtered <- lapply(pathway_list, function(x) {
    intersect(x, names(ranked))
  })
  pathway_list_filtered <- pathway_list_filtered[lengths(pathway_list_filtered) >= 5]
  cat("Pathway sets after restricting to ranked list:", length(pathway_list_filtered), "\n")

  if (length(pathway_list_filtered) == 0) {
    cat("No pathway sets have sufficient coverage — skipping.\n")
    return(NULL)
  }

  set.seed(seed)
  results <- fgsea(
    pathways = pathway_list_filtered,
    stats    = ranked,
    nPermSimple = nperm,
    minSize  = 5,
    maxSize  = 500
  )

  results <- results %>%
    as_tibble() %>%
    arrange(padj, pval) %>%
    mutate(analysis = analysis_label)

  cat("Pathways tested:", nrow(results), "\n")
  cat("Nominal p < 0.05:", sum(results$pval  < 0.05, na.rm = TRUE), "\n")
  cat("FDR < 0.05:",       sum(results$padj < 0.05, na.rm = TRUE), "\n")
  cat("FDR < 0.25:",       sum(results$padj < 0.25, na.rm = TRUE), "\n")

  return(results)
}
```

## GW Analysis

```{r msea_gw}
msea_gw <- run_msea(gw_limma, pathway_list,
                    analysis_label = "GW (limma t-statistic)")
```

## Sex Analysis

```{r msea_sex}
msea_sex <- run_msea(sex_limma, pathway_list,
                     analysis_label = "Sex (limma t-statistic, positive = higher in males)")
```

---

# Results

## GW — Top Pathways

```{r results_gw}
if (!is.null(msea_gw)) {
  msea_gw %>%
    head(30) %>%
    dplyr::select(pathway, pval, padj, NES, size, leadingEdge) %>%
    mutate(
      pval = signif(pval, 3),
      padj = signif(padj, 3),
      NES  = round(NES, 3),
      leadingEdge = map_chr(leadingEdge, ~paste(head(.x, 5), collapse = ", "))
    ) %>%
    rename(
      Pathway         = pathway,
      `p-value`       = pval,
      `FDR`           = padj,
      `NES`           = NES,
      `Set size`      = size,
      `Leading edge (top 5)` = leadingEdge
    ) %>%
    kable(caption = "GW MSEA — top 30 pathways by FDR") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE, font_size = 11)
}
```

## Sex — Top Pathways

```{r results_sex}
if (!is.null(msea_sex)) {
  msea_sex %>%
    head(30) %>%
    dplyr::select(pathway, pval, padj, NES, size, leadingEdge) %>%
    mutate(
      pval = signif(pval, 3),
      padj = signif(padj, 3),
      NES  = round(NES, 3),
      leadingEdge = map_chr(leadingEdge, ~paste(head(.x, 5), collapse = ", "))
    ) %>%
    rename(
      Pathway         = pathway,
      `p-value`       = pval,
      `FDR`           = padj,
      `NES`           = NES,
      `Set size`      = size,
      `Leading edge (top 5)` = leadingEdge
    ) %>%
    kable(caption = "Sex MSEA — top 30 pathways by FDR") %>%
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE, font_size = 11)
}
```

---

# Visualizations

```{r plot_function}
# Dot plot: NES on x, pathway on y, color = FDR, size = set size
# Shows top N by FDR, separated into enriched (NES > 0) and depleted (NES < 0)
plot_msea_dotplot <- function(results, analysis_label, top_n = 25,
                               fdr_threshold = 0.25) {
  if (is.null(results) || nrow(results) == 0) {
    message("No results to plot for: ", analysis_label)
    return(NULL)
  }

  plot_data <- results %>%
    filter(pval < 0.05) %>%           # at minimum nominal significance
    arrange(padj) %>%
    head(top_n) %>%
    mutate(
      pathway_short = str_trunc(pathway, 55),
      direction     = ifelse(NES > 0, "Enriched (+NES)", "Depleted (−NES)"),
      sig_label     = case_when(
        padj < 0.05  ~ "FDR < 0.05",
        padj < 0.25  ~ "FDR < 0.25",
        TRUE         ~ "p < 0.05"
      )
    )

  if (nrow(plot_data) == 0) {
    message("No nominally significant pathways for: ", analysis_label)
    return(NULL)
  }

  ggplot(plot_data,
         aes(x = NES,
             y = reorder(pathway_short, NES),
             color = padj,
             size  = size)) +
    geom_point(alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    scale_color_gradient(low = "firebrick", high = "steelblue",
                         name = "FDR", limits = c(0, 1),
                         breaks = c(0, 0.05, 0.25, 0.5, 1)) +
    scale_size_continuous(name = "Set size", range = c(2, 8)) +
    theme_minimal(base_size = 11) +
    labs(
      x     = "Normalized Enrichment Score (NES)",
      y     = NULL,
      title = paste0("MSEA — ", analysis_label),
      subtitle = paste0("Top ", nrow(plot_data), " pathways by FDR (p < 0.05 shown)")
    ) +
    theme(
      plot.title    = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
      axis.text.y   = element_text(size = 8),
      legend.position = "right"
    )
}
```

## GW Dot Plot

```{r plot_gw, fig.width=10, fig.height=8}
p_gw <- plot_msea_dotplot(msea_gw, "GW")
if (!is.null(p_gw)) {
  save_dual_format(p_gw, "msea_gw_dotplot", png_path = png_path, svg_path = svg_path)
  print(p_gw)
}
```

## Sex Dot Plot

```{r plot_sex, fig.width=10, fig.height=8}
p_sex <- plot_msea_dotplot(msea_sex, "Sex (positive NES = enriched in males)")
if (!is.null(p_sex)) {
  save_dual_format(p_sex, "msea_sex_dotplot", png_path = png_path, svg_path = svg_path)
  print(p_sex)
}
```

## fgsea Enrichment Plots — Top GW Pathways

```{r enrichment_plots_gw, fig.width=8, fig.height=4}
# Individual enrichment plots for top 5 GW pathways by FDR
if (!is.null(msea_gw) && nrow(msea_gw) > 0) {
  ranked_gw <- gw_limma %>%
    filter(!is.na(t)) %>%
    arrange(desc(t)) %>%
    dplyr::select(Name, t) %>%
    deframe()

  # Restrict to pathways present in filtered list
  pathway_list_gw <- lapply(pathway_list, function(x) intersect(x, names(ranked_gw)))
  pathway_list_gw <- pathway_list_gw[lengths(pathway_list_gw) >= 5]

  top_gw_pathways <- msea_gw %>%
    arrange(padj) %>%
    head(5) %>%
    pull(pathway)

  for (pw in top_gw_pathways) {
    if (pw %in% names(pathway_list_gw)) {
      p <- plotEnrichment(pathway_list_gw[[pw]], ranked_gw) +
        labs(title = str_trunc(pw, 70),
             subtitle = paste0("NES = ", round(msea_gw$NES[msea_gw$pathway == pw], 3),
                               "  |  FDR = ", signif(msea_gw$padj[msea_gw$pathway == pw], 3)),
             x = "Rank in GW t-statistic",
             y = "Enrichment Score") +
        theme_minimal() +
        theme(plot.title    = element_text(size = 10, hjust = 0.5),
              plot.subtitle = element_text(size = 9,  hjust = 0.5, color = "grey40"))
      print(p)
    }
  }
}
```

## fgsea Enrichment Plots — Top Sex Pathways

```{r enrichment_plots_sex, fig.width=8, fig.height=4}
if (!is.null(msea_sex) && nrow(msea_sex) > 0) {
  ranked_sex <- sex_limma %>%
    filter(!is.na(t)) %>%
    arrange(desc(t)) %>%
    dplyr::select(Name, t) %>%
    deframe()

  pathway_list_sex <- lapply(pathway_list, function(x) intersect(x, names(ranked_sex)))
  pathway_list_sex <- pathway_list_sex[lengths(pathway_list_sex) >= 5]

  top_sex_pathways <- msea_sex %>%
    arrange(padj) %>%
    head(5) %>%
    pull(pathway)

  for (pw in top_sex_pathways) {
    if (pw %in% names(pathway_list_sex)) {
      p <- plotEnrichment(pathway_list_sex[[pw]], ranked_sex) +
        labs(title = str_trunc(pw, 70),
             subtitle = paste0("NES = ", round(msea_sex$NES[msea_sex$pathway == pw], 3),
                               "  |  FDR = ", signif(msea_sex$padj[msea_sex$pathway == pw], 3),
                               "  |  positive NES = enriched in males"),
             x = "Rank in sex t-statistic (M > F)",
             y = "Enrichment Score") +
        theme_minimal() +
        theme(plot.title    = element_text(size = 10, hjust = 0.5),
              plot.subtitle = element_text(size = 9,  hjust = 0.5, color = "grey40"))
      print(p)
    }
  }
}
```

---

# Save Results

```{r save}
# Save pathway sets (for reproducibility / supplementary)
pathway_sets_out <- pathway_sets_raw %>%
  distinct(Name, pathway_id, pathway_label) %>%
  left_join(pathway_info %>% dplyr::select(pathway_id, pathway_source),
            by = "pathway_id")
write_csv(pathway_sets_out,
          file.path(csv_path, "batch2_pathway_metabolite_mapping.csv"))

# Save MSEA results
if (!is.null(msea_gw)) {
  msea_gw_out <- msea_gw %>%
    mutate(leadingEdge = map_chr(leadingEdge, ~paste(.x, collapse = ";")))
  write_csv(msea_gw_out, file.path(csv_path, "batch2_msea_gw.csv"))
}

if (!is.null(msea_sex)) {
  msea_sex_out <- msea_sex %>%
    mutate(leadingEdge = map_chr(leadingEdge, ~paste(.x, collapse = ";")))
  write_csv(msea_sex_out, file.path(csv_path, "batch2_msea_sex.csv"))
}

cat("Files saved to:", csv_path, "\n")
```
