---
title: "Cell Type Characterization — QC & UMAP"
author: "Sneha M. Chaturvedi"
date: "Last updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

Cell type QC and UMAP visualization of the filtered snRNA-seq dataset.
Sample composition tables, cell count summaries, and UMAP embeddings colored
by cell type, sex, gestational week, and batch.

| | |
|---|---|
| **Inputs** | `data/cache/SSD_meta_umap.rds` |
| **Outputs** | Figures in `results/gene-hormone/qc/` and `results/gene-hormone/umap/` |
| **Upstream** | `prep_data.R` |
| **Downstream** | None (visualization only) |


## Goals

Characterize cell type composition and visualize UMAP embedding of the cleaned, reclustered snRNA-seq dataset (filtering and reclustering handled by `filter_recluster.R`; pseudobulk caching by `prep_data.R`).

1. **Cell count QC** — per-sample count tables and bar charts across three classification levels (`celltype`, `subclass`, `class`), with breakdowns by sex and gestational week.
2. **UMAP visualizations** — Harmony-corrected embedding (`umap.rna`) colored by cell type classification, sex, gestational week, and batch.

## Set Up

```{r load-packages, results='hide', message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

packages <- c(
  "tidyverse", "Seurat", "patchwork", "cowplot",
  "knitr", "kableExtra", "svglite", "scales"
)
lapply(packages, library, character.only = TRUE)

set.seed(123)

find_project_root <- function(marker = ".git") {
  d <- normalizePath(getwd())
  repeat {
    if (dir.exists(file.path(d, marker))) return(d)
    parent <- dirname(d)
    if (parent == d) stop("Project root not found (no ", marker, " above ", getwd(), ")")
    d <- parent
  }
}
root <- find_project_root()
```

```{r config}
# Metadata columns
celltype_cols <- c("celltype", "subclass", "class")
sample_col    <- "Sample"
sex_col       <- "genotype" # "XX" = female, "XY" = male
gw_col        <- "age"
batch_col     <- "batch"

# Colors
sex_colors  <- c("XX" = "#FFC20A", "XY" = "#571C90")
gw_palette  <- colorRampPalette(c("#C8E6C9", "#1B5E20"))

# Absence threshold: flag cell types with fewer than this many cells in any sample
absence_threshold <- 0

# Output directories
dir_qc   <- file.path(root, "results/gene-hormone/qc")
dir_umap <- file.path(root, "results/gene-hormone/umap")
dir.create(dir_qc,                          recursive = TRUE, showWarnings = FALSE)
dir.create(dir_umap,                        recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(dir_umap, "svg"),      recursive = TRUE, showWarnings = FALSE)
```

```{r load-data}
SSD_data <- readRDS(file.path(root, "data/cache/SSD_meta_umap.rds"))

meta <- SSD_data@meta.data %>%
  as_tibble(rownames = "cell_barcode")

all_samples <- unique(meta[[sample_col]])

cat("Cells:", ncol(SSD_data), "\n")
cat("Samples:", paste(sort(all_samples), collapse = ", "), "\n")
cat("Reductions:", paste(names(SSD_data@reductions), collapse = ", "), "\n")
```

## Per-Sample Metadata Summary

```{r sample-summary}
sample_meta <- meta %>%
  distinct(.data[[sample_col]], .data[[sex_col]], .data[[gw_col]]) %>%
  rename(Sample = all_of(sample_col), Sex = all_of(sex_col), GW = all_of(gw_col)) %>%
  arrange(GW, Sex)

kable(sample_meta, caption = "Sample metadata summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

total_per_sample <- meta %>%
  count(.data[[sample_col]], .data[[sex_col]], .data[[gw_col]], name = "total_cells") %>%
  rename(Sample = all_of(sample_col), Sex = all_of(sex_col), GW = all_of(gw_col)) %>%
  arrange(GW, Sex)

kable(total_per_sample, caption = "Total cells per sample") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Cell Count QC — Helper Functions

```{r qc-helpers}
build_count_table <- function(meta, celltype_col, sample_col) {
  meta %>%
    count(.data[[sample_col]], .data[[celltype_col]], name = "n") %>%
    group_by(.data[[sample_col]]) %>%
    mutate(pct = round(100 * n / sum(n), 2)) %>%
    ungroup() %>%
    rename(CellType = all_of(celltype_col), Sample = all_of(sample_col))
}

flag_absent <- function(count_df, all_samples, threshold = 0) {
  count_df %>%
    tidyr::complete(Sample = all_samples, CellType, fill = list(n = 0, pct = 0)) %>%
    group_by(CellType) %>%
    summarise(
      min_cells        = min(n),
      n_samples_absent = sum(n <= threshold),
      absent_in        = paste(Sample[n <= threshold], collapse = "; "),
      .groups = "drop"
    ) %>%
    filter(n_samples_absent > 0) %>%
    arrange(desc(n_samples_absent))
}

plot_count_bar <- function(count_df, title, fill_palette = NULL) {
  p <- ggplot(count_df, aes(x = Sample, y = n, fill = CellType)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = title, x = "Sample", y = "Cell count", fill = "Cell type") +
    theme_cowplot(12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right")
  if (!is.null(fill_palette)) p <- p + scale_fill_manual(values = fill_palette)
  p
}

plot_prop_bar <- function(count_df, title, fill_palette = NULL) {
  p <- ggplot(count_df, aes(x = Sample, y = pct, fill = CellType)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = title, x = "Sample", y = "% of cells", fill = "Cell type") +
    scale_y_continuous(limits = c(0, 100)) +
    theme_cowplot(12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right")
  if (!is.null(fill_palette)) p <- p + scale_fill_manual(values = fill_palette)
  p
}
```

## Cell Counts by Classification Level

```{r qc-loop, results='asis', fig.width=14, fig.height=7}

cat("::: panel-tabset\n\n")

for (ct_col in celltype_cols) {
  cat("###", ct_col, "\n\n")

  ct_counts <- build_count_table(meta, ct_col, sample_col)

  ct_wide_n <- ct_counts %>%
    dplyr::select(Sample, CellType, n) %>%
    pivot_wider(names_from = Sample, values_from = n, values_fill = 0)

  ct_wide_pct <- ct_counts %>%
    dplyr::select(Sample, CellType, pct) %>%
    pivot_wider(names_from = Sample, values_from = pct, values_fill = 0)

  cat("#### Absolute counts\n\n")
  print(
    kable(ct_wide_n, caption = paste("Cell counts per sample —", ct_col)) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE, font_size = 11)
  )

  cat("\n\n#### Percentage of total cells per sample\n\n")
  print(
    kable(ct_wide_pct, caption = paste("% of cells per sample —", ct_col)) %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE, font_size = 11)
  )

  absent <- flag_absent(ct_counts, all_samples, threshold = absence_threshold)
  if (nrow(absent) > 0) {
    cat("\n\n**Cell types absent (≤", absence_threshold, "cells) in at least one sample:**\n\n")
    print(
      kable(absent,
        col.names = c("Cell type", "Min cells", "N samples absent", "Absent in"),
        caption   = paste("Absent cell types —", ct_col)
      ) %>%
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
        row_spec(0, bold = TRUE)
    )
  } else {
    cat("\n\n*All cell types present in all samples.*\n\n")
  }

  p_abs  <- plot_count_bar(ct_counts, paste("Cell counts by sample —", ct_col))
  p_prop <- plot_prop_bar(ct_counts, paste("% composition by sample —", ct_col))

  cat("\n\n#### Stacked bar charts\n\n")
  print(p_abs + p_prop + plot_layout(guides = "collect"))
  ggsave(file.path(dir_qc, paste0(ct_col, "_counts_by_sample.png")),
    plot = p_abs + p_prop + plot_layout(guides = "collect"),
    width = 16, height = 7, dpi = 150
  )

  # By sex
  ct_sex <- meta %>%
    count(.data[[sample_col]], .data[[sex_col]], .data[[ct_col]], name = "n") %>%
    group_by(.data[[sample_col]], .data[[sex_col]]) %>%
    mutate(pct = round(100 * n / sum(n), 2)) %>%
    ungroup() %>%
    rename(Sample = all_of(sample_col), Sex = all_of(sex_col), CellType = all_of(ct_col))

  p_sex_abs <- ggplot(ct_sex, aes(x = Sample, y = n, fill = CellType)) +
    geom_bar(stat = "identity", position = "stack") +
    facet_wrap(~Sex, scales = "free_x",
      labeller = labeller(Sex = c("XX" = "Female (XX)", "XY" = "Male (XY)"))) +
    labs(title = paste("Cell counts by sample and sex —", ct_col),
      x = "Sample", y = "Cell count", fill = "Cell type") +
    theme_cowplot(11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text  = element_text(face = "bold"))

  p_sex_prop <- ggplot(ct_sex, aes(x = Sample, y = pct, fill = CellType)) +
    geom_bar(stat = "identity", position = "stack") +
    facet_wrap(~Sex, scales = "free_x",
      labeller = labeller(Sex = c("XX" = "Female (XX)", "XY" = "Male (XY)"))) +
    labs(title = paste("% composition by sex —", ct_col),
      x = "Sample", y = "% of cells", fill = "Cell type") +
    scale_y_continuous(limits = c(0, 100)) +
    theme_cowplot(11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text  = element_text(face = "bold"))

  cat("\n\n#### By sex\n\n")
  print(p_sex_abs / p_sex_prop + plot_layout(guides = "collect"))
  ggsave(file.path(dir_qc, paste0(ct_col, "_counts_by_sex.png")),
    plot = p_sex_abs / p_sex_prop + plot_layout(guides = "collect"),
    width = 14, height = 12, dpi = 150
  )

  # By GW
  ct_gw <- meta %>%
    count(.data[[sample_col]], .data[[gw_col]], .data[[ct_col]], name = "n") %>%
    group_by(.data[[sample_col]], .data[[gw_col]]) %>%
    mutate(pct = round(100 * n / sum(n), 2)) %>%
    ungroup() %>%
    rename(Sample = all_of(sample_col), GW = all_of(gw_col), CellType = all_of(ct_col)) %>%
    mutate(GW = factor(GW, levels = sort(unique(GW))))

  gw_levels <- levels(ct_gw$GW)
  gw_cols   <- setNames(gw_palette(length(gw_levels)), gw_levels)

  ct_gw_total <- meta %>%
    count(.data[[gw_col]], .data[[ct_col]], name = "n") %>%
    rename(GW = all_of(gw_col), CellType = all_of(ct_col)) %>%
    mutate(GW = factor(GW, levels = sort(unique(GW))))

  p_gw_total <- ggplot(ct_gw_total, aes(x = CellType, y = n, fill = GW)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = gw_cols) +
    labs(title = paste("Total cells per cell type across GW —", ct_col),
      x = "Cell type", y = "Total cell count", fill = "GW") +
    theme_cowplot(11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  p_gw_prop <- ggplot(ct_gw, aes(x = GW, y = pct, fill = CellType)) +
    geom_bar(stat = "identity", position = "stack") +
    facet_wrap(~Sample) +
    labs(title = paste("% composition by GW —", ct_col),
      x = "GW", y = "% of cells", fill = "Cell type") +
    scale_y_continuous(limits = c(0, 100)) +
    theme_cowplot(10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text  = element_text(face = "bold"))

  cat("\n\n#### By gestational week\n\n")
  print(p_gw_total)
  print(p_gw_prop)
  ggsave(file.path(dir_qc, paste0(ct_col, "_counts_by_gw.png")),
    plot = p_gw_total / p_gw_prop + plot_layout(guides = "collect"),
    width = 14, height = 14, dpi = 150
  )

  write_csv(ct_wide_n,   file.path(dir_qc, paste0(ct_col, "_counts_wide.csv")))
  write_csv(ct_wide_pct, file.path(dir_qc, paste0(ct_col, "_pct_wide.csv")))
  if (nrow(absent) > 0) {
    write_csv(absent, file.path(dir_qc, paste0(ct_col, "_absent_celltypes.csv")))
  }
}

cat(":::\n")
```

## UMAP — Helper Functions

```{r umap-helpers}
plot_umap <- function(obj, group_by, title, cols = NULL,
                      reduction = "umap.rna", label = TRUE,
                      label_size = 3, pt_size = 0.3, na_color = "grey80") {
  DimPlot(obj,
    reduction  = reduction,
    group.by   = group_by,
    label      = label,
    label.size = label_size,
    pt.size    = pt_size,
    cols       = cols,
    repel      = TRUE,
    na.value   = na_color
  ) +
    labs(title = title) +
    theme_cowplot(12) +
    theme(legend.position = "right",
          plot.title = element_text(face = "bold", size = 13))
}

save_umap <- function(plot, filename, width = 10, height = 8) {
  ggsave(file.path(dir_umap, paste0(filename, ".png")),
    plot, width = width, height = height, dpi = 150)
  ggsave(file.path(dir_umap, "svg", paste0(filename, ".svg")),
    plot, width = width, height = height)
  invisible(plot)
}
```

## UMAP — Cell Type Classifications

```{r umap-celltypes, fig.width=10, fig.height=8}
for (ct_col in celltype_cols) {
  p <- plot_umap(SSD_data,
    group_by   = ct_col,
    title      = paste("UMAP —", ct_col),
    label      = TRUE,
    label_size = 3
  )
  print(p)
  save_umap(p, paste0("umap_", ct_col))
}
```

## UMAP — Sex & Gestational Week

```{r umap-sex-gw, fig.width=18, fig.height=8}
gw_vals   <- sort(unique(SSD_data@meta.data[[gw_col]]))
gw_colors <- setNames(gw_palette(length(gw_vals)), gw_vals)

p_sex <- plot_umap(SSD_data,
  group_by = sex_col,
  title    = "UMAP — Genotype (sex)",
  cols     = sex_colors,
  label    = FALSE
) +
  scale_color_manual(values = sex_colors,
    labels = c("XX" = "Female (XX)", "XY" = "Male (XY)"))

p_gw <- plot_umap(SSD_data,
  group_by = gw_col,
  title    = "UMAP — Gestational Week",
  cols     = gw_colors,
  label    = FALSE
)

p_sex + p_gw
save_umap(p_sex, "umap_sex")
save_umap(p_gw,  "umap_gw")
save_umap(p_sex + p_gw, "umap_sex_gw_combined", width = 18, height = 8)
```

## UMAP — Batch

```{r umap-batch, fig.width=10, fig.height=8}
batch_vals   <- sort(unique(SSD_data@meta.data[[batch_col]]))
batch_colors <- setNames(scales::hue_pal()(length(batch_vals)), batch_vals)

p_batch <- plot_umap(SSD_data,
  group_by = batch_col,
  title    = "UMAP — Batch",
  cols     = batch_colors,
  label    = FALSE
)

p_batch
save_umap(p_batch, "umap_batch")
```

# END

## AI Assistance Disclosure

Code in this script was developed with assistance from Claude (Anthropic). All
AI-generated code was reviewed, validated, and adapted by the author.

## Session Info

```{r session-info}
sessionInfo()
```
