---
title: "Gene Covariate Testing"
author: "Sneha M. Chaturvedi"
date: "Last updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

Tests whether technical and biological covariates (gestational week, cell
count, sex, batch1 fraction) predict pseudobulk gene expression across cell
types. Used to determine which covariates to include in downstream limma models.

| | |
|---|---|
| **Inputs** | `data/cache/pb_base.rds`, `data/cache/batch_props_sample.rds` |
| **Outputs** | Regression results and volcano plots in `results/gene-hormone/covariate_analysis/` |
| **Upstream** | `prep_data.R` |
| **Downstream** | Informs model specification for hormone–gene limma scripts |


## Goals

Test whether sample-level covariates predict pseudobulk gene expression at the whole-transcriptome level. Results justify covariate inclusion in downstream limma-voom hormone \~ gene models.

**Covariates tested:**

-   **Gestational Week (GW)**
-   **Cell Count** — investigated as sample covariate and confounder
-   **Sex (XX / XY)**
-   **Batch** — investigated as potential technical confound

## Set Up

### Load packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r load-packages, results='hide', message=FALSE, warning=FALSE}
Packages <- c(
  "tidyverse", "Seurat", "Signac", "glue", "readxl", "cowplot",
  "edgeR", "limma", "patchwork", "svglite", "Matrix", "ggpubr"
)
lapply(Packages, library, character.only = TRUE)

source("pseudobulk_functions.R")
root <- find_project_root()

set.seed(123)
```

### Load data

```{r load-cached-data}
pb_base <- readRDS(file.path(root, "data/cache/pb_base.rds"))
batch_props_sample <- readRDS(file.path(root, "data/cache/batch_props_sample.rds"))
```

```{r load-metadata}
targeted_comparison_raw <- read_xlsx(file.path(root, "doc/targeted/targeted_hormones.xlsx"))

targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
```

### Generate pseudobulk *(min 30 cells per cell type for a sample to be included)*

```{r set-variables}
celltype_col <- "celltype"
sample_col <- "Sample"
pb_sample_pattern <- "tissue\\d+"
```

```{r build-pseudobulk}
pb_total <- merge_hormone_metadata(
  pb_base,
  targeted_comparison_raw,
  pb_sample_pattern = pb_sample_pattern,
  min_cells         = 30
)
```

### Compute per-sample batch composition *(used in Batch section)*

```{r batch-props}
# Join batch1_frac onto pb_total$metadata for use in perform_regression()
pb_total$metadata <- pb_total$metadata %>%
  tibble::rownames_to_column("pseudobulk_id") %>%
  dplyr::left_join(batch_props_sample,
by           = c("sample_id" = "Sample"),
relationship = "many-to-one"
  ) %>%
  tibble::column_to_rownames("pseudobulk_id")
```

## Gestational Week

```{r gw-histogram, echo=FALSE}
knitr::include_graphics(file.path(root, "results/targeted/age_analysis_plots/sample_distribution_GW_histogram.svg"))
```

Does gestational week predict pseudobulk gene expression?

```{r gw-regression}
# GW is numeric (integer gestational weeks) from the hormone Excel file.
# Verify class before running; coerce if needed.
cat(
  "GW class:", class(pb_total$metadata$GW),
  "| range:", paste(range(pb_total$metadata$GW, na.rm = TRUE), collapse = "-"), "\n"
)

results_gw <- perform_regression(pb_total, "GW", "Gestational Week")
summarize_regression_results(results_gw, "Gestational Week", alpha = 0.10)

write.csv(
  results_gw,
  file.path(root, "results/gene-hormone/covariate_analysis/results_gw_regression.csv"),
  row.names = FALSE
)
```

::: panel-tabset
### Diagnostic Plots

```{r gw-summary-plot, fig.width=10, fig.height=7}
plot_covariate_summary(
  results_gw,
  covariate_label = "Gestational Week",
  alpha           = 0.10,
  output_dir      = file.path(root, "results/gene-hormone/covariate_analysis"),
  filename_base   = "GW_regression_summary"
)
```

### Top Genes

```{r gw-top-genes, fig.width=12, fig.height=8}
plot_top_genes_continuous(
  pb_total,
  results         = results_gw,
  covariate       = "GW",
  covariate_label = "Gestational Week",
  n_genes         = 6,
  output_dir      = file.path(root, "results/gene-hormone/covariate_analysis"),
  filename_base   = "GW_top_genes"
)
```

## Cell Count

Does the number of cells aggregated per pseudobulk sample predict gene expression?

::: smallest
Note that `pb_total$logcounts` is log1p of raw summed counts with no library size normalization, so some association with cell_count is expected by construction and does not indicate a meaningful biological or technical effect — voom handles depth normalization internally.

Does cell_count correlate with hormone expression or GW, making it a confound in the limma models?

::: smallest
Confounding requires correlation with both the outcome (gene expression) and the predictor (hormone/GW).

```{r cc-regression}
results_cc <- perform_regression(pb_total, "cell_count", "Cell Count")
summarize_regression_results(results_cc, "Cell Count", alpha = 0.10)

write.csv(
  results_cc,
  file.path(root, "results/gene-hormone/covariate_analysis/results_cellcount_regression.csv"),
  row.names = FALSE
)
```

```{r cc-hormone-correlation}
cor(
  pb_total$metadata[, c("cell_count", "GW", "TT", "E2", "P4")],
  use = "pairwise.complete.obs"
) %>% round(2)
```

::: fragment
**Conclusion:** All correlations between `cell_count` and hormone/GW variables are negligible (\|r\| ≤ 0.02). Cell count is not a confound and will **not** be included as a covariate in limma-voom models. Depth normalization is handled internally by voom.

::: panel-tabset
### Diagnostic Plots

```{r cc-summary-plot, fig.width=10, fig.height=7}
plot_covariate_summary(
  results_cc,
  covariate_label = "Cell Count",
  alpha           = 0.10,
  output_dir      = file.path(root, "results/gene-hormone/covariate_analysis"),
  filename_base   = "CellCount_regression_summary"
)
```

### Top Genes

```{r cc-top-genes, fig.width=12, fig.height=8}
plot_top_genes_continuous(
  pb_total,
  results         = results_cc,
  covariate       = "cell_count",
  covariate_label = "Cell Count",
  n_genes         = 6,
  output_dir      = file.path(root, "results/gene-hormone/covariate_analysis"),
  filename_base   = "CellCount_top_genes"
)
```

## Sex

Does biological sex (XX / XY) predict pseudobulk gene expression?

::: smallest
Reference level: `F` (XX). Positive estimates indicate higher expression in XY samples.

```{r sex-regression}
results_sex <- perform_regression(pb_total, "Sex", "Sex")
summarize_regression_results(results_sex, "Sex", alpha = 0.10)

write.csv(
  results_sex,
  file.path(root, "results/gene-hormone/covariate_analysis/results_sex_regression.csv"),
  row.names = FALSE
)
```

::: panel-tabset
### Diagnostic Plots

```{r sex-summary-plot, fig.width=10, fig.height=7}
plot_covariate_summary(
  results_sex,
  covariate_label = "Sex",
  alpha           = 0.10,
  output_dir      = file.path(root, "results/gene-hormone/covariate_analysis"),
  filename_base   = "Sex_regression_summary"
)
```

### Top Genes

```{r sex-top-genes, fig.width=12, fig.height=8}
plot_top_genes_categorical(
  pb_total,
  results         = results_sex,
  covariate       = "Sex",
  covariate_label = "Sex",
  n_genes         = 6,
  output_dir      = file.path(root, "results/gene-hormone/covariate_analysis"),
  filename_base   = "Sex_top_genes"
)
```

## Batch

Batch is labeled at the **cell level** in `@meta.data$batch` ("batch1" / "batch2"). The same biological samples contributed cells to both batches, so no single batch label can be assigned to a pseudobulk sample. The variable tested below is the **fraction of batch1 cells** within each pseudobulk sample (`batch1_frac`).

```{r batch-distribution, fig.width=7, fig.height=4}
# Visualise batch1_frac across all biological samples
cat("batch1_frac summary:\n")
print(summary(batch_props_sample$batch1_frac))

ggplot(batch_props_sample, aes(x = batch1_frac)) +
  geom_histogram(bins = 30, fill = "steelblue", color = NA, alpha = 0.85) +
  labs(
title    = "Batch Composition per Biological Sample",
x        = "Fraction of batch1 cells",
y        = "Samples"
  ) +
  theme_minimal(base_size = 12)

print(batch_props_sample)
```

The distribution is strongly bimodal — most samples are composed almost entirely of batch1 or batch2 cells, not a genuine mixture. Batch composition is therefore a meaningful variable to assess as a potential covariate and/or confounder.

```{r batch-correlation}
# Test whether batch1_frac correlates with biological variables of interest.
# Sex must be numeric for cor(); F=1, M=2 (alphabetical factor order).
batch_cor <- pb_total$metadata %>%
  dplyr::mutate(Sex_numeric = as.integer(factor(Sex))) %>%
  dplyr::select(GW, TT, E2, P4, Sex_numeric, batch1_frac) %>%
  cor(use = "pairwise.complete.obs") %>%
  round(2)

print(batch_cor)
```

::: fragment
**Conclusion (confounder):** `batch1_frac` correlations with all hormone concentrations and gestational week are negligible (\|r\| ≤ 0.04). The largest correlation is -0.14 with P4. Batch composition is effectively orthogonal to the biological variables of interest and is **not** a confounder.

Does `batch1_frac` predict pseudobulk gene expression at the transcriptome-wide level, justifying inclusion as a covariate?

```{r batch1-frac-regression}
results_batch1_frac <- perform_regression(pb_total, "batch1_frac", "Batch1 Fraction")
summarize_regression_results(results_batch1_frac, "Batch1 Fraction", alpha = 0.10)

write.csv(
  results_batch1_frac,
  file.path(root, "results/gene-hormone/covariate_analysis/results_batch1frac_regression.csv"),
  row.names = FALSE
)
```

::: panel-tabset
### Diagnostic Plots

```{r batch1-frac-summary-plot, fig.width=10, fig.height=7}
plot_covariate_summary(
  results_batch1_frac,
  covariate_label = "Batch1 Fraction",
  alpha           = 0.10,
  output_dir      = file.path(root, "results/gene-hormone/covariate_analysis"),
  filename_base   = "Batch1Frac_regression_summary"
)
```

### Top Genes

```{r batch1-frac-top-genes, fig.width=12, fig.height=8}
plot_top_genes_continuous(
  pb_total,
  results         = results_batch1_frac,
  covariate       = "batch1_frac",
  covariate_label = "Batch1 Fraction",
  n_genes         = 6,
  output_dir      = file.path(root, "results/gene-hormone/covariate_analysis"),
  filename_base   = "Batch1Frac_top_genes"
)
```

## Summary

```{r summary-table}
covariate_summary <- tibble::tibble(
  Covariate = c("Gestational Week", "Cell Count", "Sex", "Batch1 Fraction"),
  Type = c("Continuous", "Continuous", "Categorical", "Continuous"),
  `Genes Tested` = c(
nrow(results_gw),
nrow(results_cc),
nrow(results_sex),
nrow(results_batch1_frac)
  ),
  `Sig. (FDR < 0.10)` = c(
sum(results_gw$adj_p_value < 0.10, na.rm = TRUE),
sum(results_cc$adj_p_value < 0.10, na.rm = TRUE),
sum(results_sex$adj_p_value < 0.10, na.rm = TRUE),
sum(results_batch1_frac$adj_p_value < 0.10, na.rm = TRUE)
  )
) %>%
  dplyr::mutate(
`% Sig.` = round(100 * `Sig. (FDR < 0.10)` / `Genes Tested`, 1),
`Median R²` = round(c(
  median(results_gw$r_squared, na.rm = TRUE),
  median(results_cc$r_squared, na.rm = TRUE),
  median(results_sex$r_squared, na.rm = TRUE),
  median(results_batch1_frac$r_squared, na.rm = TRUE)
), 4)
  )

knitr::kable(covariate_summary)
```

::: fragment
**Interpretation:** Sample level covariates where a substantial fraction of genes reach FDR \< 0.10 — or where the p-value histogram shows enrichment near 0 — exert transcriptome-wide influence and should be included as additive terms in limma-voom models. Potential confounders such as Batch and Cell Count were assessed separately and determined to not be directly related to variables of interest.

## AI Assistance Disclosure

Code in this script was developed with assistance from Claude (Anthropic). All
AI-generated code was reviewed, validated, and adapted by the author.

## Session Info

```{r session-info, code.fold=TRUE}
sessionInfo()
```
