---
title: "Testosterone Gene Expression Prediction — Subclass"
author: "Sneha M. Chaturvedi, Emma Jones"
date: "Last updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

Testosterone–gene limma-voom at the subclass-level cell type annotation.

| | |
|---|---|
| **Inputs** | `data/cache/pb_base_subclass.rds`, `doc/targeted/targeted_hormones.xlsx` |
| **Outputs** | `results/gene-hormone/testosterone_genes_subclass/` |
| **Upstream** | `prep_data.R` |
| **Downstream** | None |


## Goals

Identify genes whose pseudobulk expression is predicted by testosterone (TT) concentration, by **subclass**. Analysis is **sex-separate**: XX and XY samples are modeled independently within each subclass.

**Four models tested:**

::: incremental
-   **Base** — `~ TT`
-   **Base + GW** — `~ TT + GW` (gestational week as additive covariate; justified by covariate testing)
-   **Base + Batch** — `~ TT + batch1_frac` (batch1 fraction as a continuous additive covariate; both batches draw cells from the same biological samples but differ in which cells were captured)
-   **Base + GW + Batch** — `~ TT + GW + batch1_frac` (both covariates)
:::

**Gene filter:** manual rowSums — genes with ≥ 2 counts in ≥ 3 samples per (subclass × Sex) group.\
**Sample threshold:** `floor(n_group / 2)` non-zero TT samples required per group.

## Set Up

-   Load packages

```{r load-packages, results='hide', message=FALSE, warning=FALSE}
    Packages <- c(
      "tidyverse", "Seurat", "Signac", "glue", "readxl", "cowplot",
      "edgeR", "limma", "patchwork", "svglite", "Matrix", "ggpubr"
    )
    lapply(Packages, library, character.only = TRUE)

    source("pseudobulk_functions.R")
root <- find_project_root()

    set.seed(123)
```

-   Load data```{r load-data}

```{r load-data}
    SSD_data <- readRDS(file.path(root, "data/cache/SSD_meta_umap.rds"))
    pb_base   <- readRDS(file.path(root, "data/cache/pb_base_subclass.rds"))
    targeted_comparison_raw <- read_xlsx(file.path(root, "doc/targeted/targeted_hormones.xlsx"))
    targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
    targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
    targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`
```

-   Configuration

```{r config}
    celltype_col      <- "subclass"
    sample_col        <- "Sample"
    pb_sample_pattern <- "tissue\\d+"

    out_base <- file.path(root, "results/gene-hormone/testosterone_genes_subclass")

    for (sub in c("run_voom_lm", "run_voom_lm_gw",
                  "run_voom_lm_batch", "run_voom_lm_gw_batch")) {
      dir.create(file.path(out_base, sub), recursive = TRUE, showWarnings = FALSE)
    }
```

## Outlier Removal

Samples with TT \> Q3 + 1.5 × IQR are excluded before pseudobulk construction.

```{r outlier-removal}
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) +
    1.5 * IQR(data[[variable]], na.rm = TRUE)
  data %>% dplyr::filter(.data[[variable]] <= outlier_limit)
}

targeted_comparison_filtered_TT <- detect_outliers(targeted_comparison_raw, "TT")

cat("Samples before outlier removal:", nrow(targeted_comparison_raw), "\n")
cat("Samples after outlier removal: ", nrow(targeted_comparison_filtered_TT), "\n")
cat("TT range after filtering:      ",
    paste(round(range(targeted_comparison_filtered_TT$TT, na.rm = TRUE), 3),
          collapse = " – "), "ng/mL\n")
```

## Pseudobulk

Min 30 cells per (sample × subclass) for inclusion. `add_batch_fraction()` computes the fraction of batch1 cells per biological sample and joins it onto the pseudobulk metadata.

```{r build-pseudobulk}
pb_TT <- merge_hormone_metadata(
  pb_base,
  targeted_comparison_filtered_TT,
  pb_sample_pattern = pb_sample_pattern,
  min_cells         = 30
)

pb_TT <- add_batch_fraction(pb_TT, SSD_data)

cat("Pseudobulk samples after merge:", nrow(pb_TT$metadata), "\n")
cat("batch1_frac summary:\n")
print(summary(pb_TT$metadata$batch1_frac))
cat("NAs in batch1_frac:", sum(is.na(pb_TT$metadata$batch1_frac)), "\n")
```

## Models

| Model             | Formula                      | Rationale                                                                                                                            |
|------------------------|------------------------|------------------------|
| Base              | `~ TT`                       | Direct TT–gene association                                                                                                           |
| Base + GW         | `~ TT + GW`                  | GW predicts transcriptome-wide expression (covariate testing)                                                                        |
| Base + Batch      | `~ TT + batch1_frac`      | Absorbs residual expression differences between batches; continuous encoding matches the non-bimodal batch composition |
| Base + GW + Batch | `~ TT + GW + batch1_frac` | Both covariates adjusted simultaneously                                                                                              |

All models use `analyze_hormone()`: splits by (subclass × Sex), applies manual gene filter, drops groups below sample threshold, fits limma-voom per group.

## Base Model

`~ TT` — sex-separate, no covariates.

```{r run-base}
res_TT_base <- analyze_hormone(
  pb_TT$counts, pb_TT$metadata,
  hormone               = "TT",
  min_samples           = 10,
  count_threshold       = 2,
  gene_sample_threshold = 3,
  model                 = run_voom_lm
)

summary_TT_base <- summarize_sig_genes(res_TT_base$combined_results)

plot_top_genes_from_subsets(
  res_TT_base$by_group_data,
  output_dir = file.path(out_base, "run_voom_lm"),
  hormone    = "TT"
)

save_dual_format(
  plot_sig_gene_summary(summary_TT_base, "TT — Base (~ TT) | Subclass"),
  output_dir    = file.path(out_base, "run_voom_lm"),
  filename_base = "TT_base_sig_gene_summary"
)

write_csv(res_TT_base$combined_results,
          file.path(out_base, "run_voom_lm", "TT_base_results.csv"))
write_csv(summary_TT_base,
          file.path(out_base, "run_voom_lm", "TT_base_summary.csv"))
```

::: panel-tabset
### Significant Genes

```{r base-sig-plot, fig.width=10, fig.height=5}
plot_sig_gene_summary(summary_TT_base, "TT — Base (~ TT) | Subclass")
```

### Top Results

```{r base-top-table}
res_TT_base$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — Base model")
```
:::

## Base + GW

`~ TT + GW` — gestational week as additive covariate.

```{r run-gw}
res_TT_gw <- analyze_hormone(
  pb_TT$counts, pb_TT$metadata,
  hormone               = "TT",
  min_samples           = 10,
  count_threshold       = 2,
  gene_sample_threshold = 3,
  model                 = run_voom_lm_gw
)

summary_TT_gw <- summarize_sig_genes(res_TT_gw$combined_results)

plot_top_genes_from_subsets(
  res_TT_gw$by_group_data,
  output_dir = file.path(out_base, "run_voom_lm_gw"),
  hormone    = "TT"
)

save_dual_format(
  plot_sig_gene_summary(summary_TT_gw, "TT — Base + GW (~ TT + GW) | Subclass"),
  output_dir    = file.path(out_base, "run_voom_lm_gw"),
  filename_base = "TT_gw_sig_gene_summary"
)

write_csv(res_TT_gw$combined_results,
          file.path(out_base, "run_voom_lm_gw", "TT_gw_results.csv"))
write_csv(summary_TT_gw,
          file.path(out_base, "run_voom_lm_gw", "TT_gw_summary.csv"))
```

::: panel-tabset
### Significant Genes

```{r gw-sig-plot, fig.width=10, fig.height=5}
plot_sig_gene_summary(summary_TT_gw, "TT — Base + GW (~ TT + GW) | Subclass")
```

### Top Results

```{r gw-top-table}
res_TT_gw$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — Base + GW model")
```
:::

## Base + Batch

`~ TT + batch1_frac` — batch1 fraction (continuous) as additive covariate.

```{r run-batch}
res_TT_batch <- analyze_hormone(
  pb_TT$counts, pb_TT$metadata,
  hormone               = "TT",
  min_samples           = 10,
  count_threshold       = 2,
  gene_sample_threshold = 3,
  model                 = run_voom_lm_batch
)

summary_TT_batch <- summarize_sig_genes(res_TT_batch$combined_results)

plot_top_genes_from_subsets(
  res_TT_batch$by_group_data,
  output_dir = file.path(out_base, "run_voom_lm_batch"),
  hormone    = "TT"
)

save_dual_format(
  plot_sig_gene_summary(summary_TT_batch, "TT — Base + Batch (~ TT + batch1_frac) | Subclass"),
  output_dir    = file.path(out_base, "run_voom_lm_batch"),
  filename_base = "TT_batch_sig_gene_summary"
)

write_csv(res_TT_batch$combined_results,
          file.path(out_base, "run_voom_lm_batch", "TT_batch_results.csv"))
write_csv(summary_TT_batch,
          file.path(out_base, "run_voom_lm_batch", "TT_batch_summary.csv"))
```

::: panel-tabset
### Significant Genes

```{r batch-sig-plot, fig.width=10, fig.height=5}
plot_sig_gene_summary(summary_TT_batch, "TT — Base + Batch (~ TT + batch1_frac) | Subclass")
```

### Top Results

```{r batch-top-table}
res_TT_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — Base + Batch model")
```
:::

## Base + GW + Batch

`~ TT + GW + batch1_frac` — gestational week and batch1 fraction as simultaneous additive covariates.

```{r run-gw-batch}
res_TT_gw_batch <- analyze_hormone(
  pb_TT$counts, pb_TT$metadata,
  hormone               = "TT",
  min_samples           = 10,
  count_threshold       = 2,
  gene_sample_threshold = 3,
  model                 = run_voom_lm_gw_batch
)

summary_TT_gw_batch <- summarize_sig_genes(res_TT_gw_batch$combined_results)

plot_top_genes_from_subsets(
  res_TT_gw_batch$by_group_data,
  output_dir = file.path(out_base, "run_voom_lm_gw_batch"),
  hormone    = "TT"
)

save_dual_format(
  plot_sig_gene_summary(summary_TT_gw_batch,
                        "TT — Base + GW + Batch (~ TT + GW + batch1_frac) | Subclass"),
  output_dir    = file.path(out_base, "run_voom_lm_gw_batch"),
  filename_base = "TT_gw_batch_sig_gene_summary"
)

write_csv(res_TT_gw_batch$combined_results,
          file.path(out_base, "run_voom_lm_gw_batch", "TT_gw_batch_results.csv"))
write_csv(summary_TT_gw_batch,
          file.path(out_base, "run_voom_lm_gw_batch", "TT_gw_batch_summary.csv"))
```

::: panel-tabset
### Significant Genes

```{r gw-batch-sig-plot, fig.width=10, fig.height=5}
plot_sig_gene_summary(summary_TT_gw_batch,
                      "TT — Base + GW + Batch (~ TT + GW + batch1_frac) | Subclass")
```

### Top Results

```{r gw-batch-top-table}
res_TT_gw_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — Base + GW + Batch model")
```
:::

## Summary

```{r cross-model-summary}
cross_model_TT <- dplyr::bind_rows(
  summarize_sig_genes(res_TT_base$combined_results)     %>%
    dplyr::mutate(model = "~ TT"),
  summarize_sig_genes(res_TT_gw$combined_results)       %>%
    dplyr::mutate(model = "~ TT + GW"),
  summarize_sig_genes(res_TT_batch$combined_results)    %>%
    dplyr::mutate(model = "~ TT + batch1_frac"),
  summarize_sig_genes(res_TT_gw_batch$combined_results) %>%
    dplyr::mutate(model = "~ TT + GW + batch1_frac")
)

knitr::kable(
  cross_model_TT %>%
    dplyr::select(model, cell_type, sex, num_sig_genes) %>%
    dplyr::arrange(model, desc(num_sig_genes)),
  caption = "Significant genes (FDR < 0.10, |logFC| ≥ 0.25) per model"
)
```

```{r cross-model-totals}
cross_model_TT %>%
  dplyr::group_by(model) %>%
  dplyr::summarise(total_sig_genes = sum(num_sig_genes), .groups = "drop") %>%
  knitr::kable(caption = "Total significant genes per model")
```

## Analysis UMAP

Cells included in the testosterone analysis: from (subclass × XY) groups that passed all thresholds. XY cells from included groups are colored by subclass; all XX cells and cells from excluded subclasses are shown in grey.

```{r tt-analysis-umap, fig.width=10, fig.height=8}
plot_analysis_umap(
  seurat_obj    = SSD_data,
  by_group_data = res_TT_base$by_group_data,
  title         = "Testosterone Analysis — Included Cells (XY) | Subclass",
  celltype_col  = celltype_col,
  sex_separate  = TRUE,
  output_dir    = out_base,
  filename_base = "TT_analysis_umap"
)
```

## AI Assistance Disclosure

Code in this script was developed with assistance from Claude (Anthropic). All
AI-generated code was reviewed, validated, and adapted by the author.

## Session Info

```{r session-info}
sessionInfo()
```
