---
title: "Cross-Hormone Comparison of DEGs — Summary Figures"
subtitle: "Sneha Chaturvedi, Emma Jones"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    code-tools: true
    scrollable: true
    theme: dark
    chalkboard: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

## Goals

Summarise and compare significant hormone-associated DEGs across testosterone,
estradiol, and progesterone for the manuscript.

::: incremental
- **Gene-level** overlap across hormones (ignoring cell type)
- **Gene × cell-type** overlap across hormones
- **Sign concordance** of the hormone slope for genes shared by ≥ 2 hormones
- **Integrative panel**: hormone DEGs that are also directly receptor-bound
  (promoter-proximal CUT&Tag; TT × AR, E2 × ESR1)
:::

Significance throughout: **FDR < 0.10 AND |logFC| >= 0.25** (matches the
hormone-script graphs and `SFARI_DEG.qmd`).

## Set Up

```{r}
#| label: load-packages
Packages <- c("tidyverse", "glue", "ComplexUpset", "patchwork", "svglite")
lapply(Packages, library, character.only = TRUE)

source("pseudobulk_functions.R")

set.seed(123)

out_dir <- "../../results/gene-hormone/compare_gene_results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

hormone_cols <- c(Testosterone = "#2c7fb8",
                  Estradiol    = "#f03b20",
                  Progesterone = "#31a354")
```

```{r}
#| label: load-results
# Same result files as SFARI_DEG.qmd: TT = GW model (male), E2/P4 = Sex model.
TT_results <- read_csv(
  "../../results/gene-hormone/testosterone_genes/run_voom_lm_gw/TT_gw_results.csv",
  show_col_types = FALSE
)
E2_results <- read_csv(
  "../../results/gene-hormone/estradiol_genes/run_voom_lm_sex/E2_sex_results.csv",
  show_col_types = FALSE
)
P4_results <- read_csv(
  "../../results/gene-hormone/progesterone_genes/run_voom_lm_sex/P4_sex_results.csv",
  show_col_types = FALSE
)

TT_sig <- filter_sig(TT_results, sex_filter = "M") %>% dplyr::mutate(hormone = "Testosterone")
E2_sig <- filter_sig(E2_results)                   %>% dplyr::mutate(hormone = "Estradiol")
P4_sig <- filter_sig(P4_results)                   %>% dplyr::mutate(hormone = "Progesterone")

sig_all <- dplyr::bind_rows(TT_sig, E2_sig, P4_sig) %>%
  dplyr::mutate(hormone = factor(hormone, levels = names(hormone_cols)))

sig_all %>%
  dplyr::count(hormone, name = "sig_gene_celltype_rows") %>%
  dplyr::mutate(unique_genes = purrr::map_int(
    levels(hormone),
    ~ length(unique(sig_all$gene[sig_all$hormone == .x]))
  )) %>%
  knitr::kable(caption = "Significant DE counts per hormone")
```

## Gene-Level Overlap (UpSet)

Headline comparison: which genes are significant in which hormones, collapsed
across cell types.

```{r}
#| label: gene-sets
gene_sets <- list(
  Testosterone = unique(TT_sig$gene),
  Estradiol    = unique(E2_sig$gene),
  Progesterone = unique(P4_sig$gene)
)

membership_gene <- make_membership_df(gene_sets, element_col = "gene")
```

```{r}
#| label: upset-gene
#| fig-width: 10
#| fig-height: 6
p_upset_gene <- ComplexUpset::upset(
  membership_gene,
  intersect = names(gene_sets),
  name = "Hormone",
  sort_intersections_by = "cardinality",
  base_annotations = list(
    "Intersection size" = ComplexUpset::intersection_size(text = list(size = 3.5))
  )
) +
  patchwork::plot_annotation(
    title = "Shared significant genes across hormones (gene-level)"
  )

p_upset_gene

save_dual_format(p_upset_gene, out_dir, "upset_gene_level",
                 width = 10, height = 6)
```

## Gene × Cell-Type Overlap (UpSet)

Stricter unit: a hit must match on **both** gene and cell type to be shared.

```{r}
#| label: pair-sets
pair_id <- function(df) paste(df$gene, df$cell_type, sep = "__")

pair_sets <- list(
  Testosterone = unique(pair_id(TT_sig)),
  Estradiol    = unique(pair_id(E2_sig)),
  Progesterone = unique(pair_id(P4_sig))
)

membership_pair <- make_membership_df(pair_sets, element_col = "gene_celltype")
```

```{r}
#| label: upset-pair
#| fig-width: 10
#| fig-height: 6
p_upset_pair <- ComplexUpset::upset(
  membership_pair,
  intersect = names(pair_sets),
  name = "Hormone",
  sort_intersections_by = "cardinality",
  base_annotations = list(
    "Intersection size" = ComplexUpset::intersection_size(text = list(size = 3.5))
  )
) +
  patchwork::plot_annotation(
    title = "Shared significant hits across hormones (gene × cell-type)"
  )

p_upset_pair

save_dual_format(p_upset_pair, out_dir, "upset_gene_celltype",
                 width = 10, height = 6)
```

## Per-Gene Effect Direction

Collapses each gene to one representative slope per hormone (from its most
significant cell type) for the concordance analysis and heatmap.

```{r}
#| label: per-gene-collapse
per_gene_hormone <- sig_all %>%
  dplyr::group_by(hormone, gene) %>%
  dplyr::summarise(
    n_celltypes = dplyr::n_distinct(cell_type),
    rep_logFC   = logFC[which.min(adj.P.Val)],   # from lowest-FDR cell type
    mean_logFC  = mean(logFC),
    .groups = "drop"
  )

# Wide: gene x hormone representative slope
rep_wide <- per_gene_hormone %>%
  dplyr::select(gene, hormone, rep_logFC) %>%
  tidyr::pivot_wider(names_from = hormone, values_from = rep_logFC)

# Genes significant in >= 2 hormones
shared_genes <- rep_wide %>%
  dplyr::mutate(n_hormones = rowSums(!is.na(dplyr::across(dplyr::all_of(names(hormone_cols)))))) %>%
  dplyr::filter(n_hormones >= 2)

cat("Genes significant in >= 2 hormones:", nrow(shared_genes), "\n")
```

## Sign Concordance (shared genes)

For genes significant in ≥ 2 hormones, do the representative slopes agree in
sign?

```{r}
#| label: concordance
concordance <- shared_genes %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    signs = list(sign(dplyr::c_across(dplyr::all_of(names(hormone_cols))))),
    n_pos = sum(unlist(signs) > 0, na.rm = TRUE),
    n_neg = sum(unlist(signs) < 0, na.rm = TRUE),
    direction = dplyr::case_when(
      n_pos > 0 & n_neg == 0 ~ "concordant (+)",
      n_neg > 0 & n_pos == 0 ~ "concordant (-)",
      TRUE                   ~ "discordant"
    )
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(-signs)

concordance %>%
  dplyr::count(direction, name = "n_genes") %>%
  knitr::kable(caption = "Sign concordance of representative slopes (shared genes)")

write_csv(concordance, file.path(out_dir, "shared_gene_sign_concordance.csv"))
```

## Effect-Direction Heatmap

Representative slope per shared gene across hormones. Blue = negative, red =
positive; grey = not significant for that hormone. Rows are genes significant in
≥ 2 hormones.

```{r}
#| label: heatmap
#| fig-width: 8
#| fig-height: 10
heat_df <- shared_genes %>%
  dplyr::select(gene, dplyr::all_of(names(hormone_cols))) %>%
  tidyr::pivot_longer(-gene, names_to = "hormone", values_to = "rep_logFC") %>%
  dplyr::mutate(hormone = factor(hormone, levels = names(hormone_cols)))

# Order genes by concordance then by mean slope for a readable block structure.
gene_order <- concordance %>%
  dplyr::arrange(direction, dplyr::desc(rowMeans(
    dplyr::across(dplyr::all_of(names(hormone_cols))), na.rm = TRUE
  ))) %>%
  dplyr::pull(gene)

heat_df <- heat_df %>%
  dplyr::mutate(gene = factor(gene, levels = rev(gene_order)))

p_heat <- ggplot(heat_df, aes(x = hormone, y = gene, fill = rep_logFC)) +
  geom_tile(color = "grey30") +
  scale_fill_gradient2(low = "#2c7bb6", mid = "white", high = "#d7191c",
                       midpoint = 0, na.value = "grey85",
                       name = "Representative\nlogFC (slope)") +
  labs(title = "Effect direction across hormones (genes shared by ≥ 2)",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 7),
        panel.grid = element_blank())

p_heat

save_dual_format(p_heat, out_dir, "shared_gene_direction_heatmap",
                 width = 8, height = 10)
```

## Integrative Panel — DEGs Directly Receptor-Bound

Of each hormone's significant genes, how many are also bound by the cognate
receptor at a promoter-proximal CUT&Tag peak (TT × AR, E2 × ESR1)? Promoter
sets re-derived from the same annotation files via the shared loader.

```{r}
#| label: load-ct-promoter
AR   <- load_cut_and_tag_geneset("../../doc/cut&tag/AR_consensus_anno.csv")
ESR1 <- load_cut_and_tag_geneset("../../doc/cut&tag/ESR1_consensus_anno.csv")

TT_bound <- TT_sig %>% dplyr::mutate(bound = gene %in% AR$promoter_genes)
E2_bound <- E2_sig %>% dplyr::mutate(bound = gene %in% ESR1$promoter_genes)
```

```{r}
#| label: integrative-plot
#| fig-width: 12
#| fig-height: 6
bound_bar <- function(df, title, bound_fill) {
  df %>%
    dplyr::count(cell_type, bound, name = "n") %>%
    dplyr::mutate(bound = factor(ifelse(bound, "bound", "not bound"),
                                 levels = c("not bound", "bound"))) %>%
    ggplot(aes(x = cell_type, y = n, fill = bound)) +
    geom_col() +
    scale_fill_manual(values = c("not bound" = "grey80", "bound" = bound_fill)) +
    labs(title = title, x = "Cell Type", y = "# significant genes", fill = NULL) +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

p_tt_bound <- bound_bar(TT_bound, "Testosterone DEGs — AR promoter-bound (male)",
                        bound_fill = "#2c7fb8")
p_e2_bound <- bound_bar(E2_bound, "Estradiol DEGs — ESR1 promoter-bound",
                        bound_fill = "#f03b20")

p_integrative <- p_tt_bound + p_e2_bound
p_integrative

save_dual_format(p_integrative, out_dir, "integrative_bound_by_celltype",
                 width = 12, height = 6)
```

```{r}
#| label: integrative-tables
tt_bound_genes <- TT_bound %>%
  dplyr::filter(bound) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  dplyr::arrange(adj.P.Val)

e2_bound_genes <- E2_bound %>%
  dplyr::filter(bound) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  dplyr::arrange(adj.P.Val)

write_csv(tt_bound_genes, file.path(out_dir, "tt_ar_promoter_bound_degs.csv"))
write_csv(e2_bound_genes, file.path(out_dir, "e2_esr1_promoter_bound_degs.csv"))

cat("TT DEGs AR-bound (promoter):", length(unique(tt_bound_genes$gene)), "unique genes\n")
cat("E2 DEGs ESR1-bound (promoter):", length(unique(e2_bound_genes$gene)), "unique genes\n")
```

## Master Summary

```{r}
#| label: master-summary
master <- tibble::tibble(
  Hormone = names(hormone_cols),
  `Sig genes (unique)` = c(length(gene_sets$Testosterone),
                           length(gene_sets$Estradiol),
                           length(gene_sets$Progesterone)),
  `Sig gene×cell-type` = c(length(pair_sets$Testosterone),
                           length(pair_sets$Estradiol),
                           length(pair_sets$Progesterone)),
  `Receptor-bound (promoter)` = c(length(unique(tt_bound_genes$gene)),
                                  length(unique(e2_bound_genes$gene)),
                                  NA_integer_)   # no PGR CUT&Tag track
)

write_csv(master, file.path(out_dir, "compare_gene_results_master_summary.csv"))
knitr::kable(master, caption = "Cross-hormone summary (P4 has no CUT&Tag reference)")
```
