---
title: "CUT&Tag Binding Overlap with Hormone-Associated DEGs"
subtitle: "Sneha Chaturvedi, Emma Jones"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    code-tools: true
    scrollable: true
    theme: dark
    chalkboard: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

## Goals

Determine whether significant hormone-associated DEGs are directly bound by
their cognate receptor, using CUT&Tag consensus peaks from the same samples.

**Receptor → hormone pairing**

::: incremental
- **AR** (androgen receptor) → **testosterone** (male samples)
- **ESR1** (ER⍺) → **estradiol** (both sexes)
:::

::: notes
ESR2 (ER⍺/ERβ CUT&Tag for ESR2) did not yield usable peaks and is excluded.
Progesterone is not compared here: there is no PGR CUT&Tag track (only AR and
ESR1 succeeded), so P4 DEGs have no cognate-binding reference in this assay.
:::

CUT&Tag binding is a presence/absence signal with no direction, so overlaps are
**set membership only** (no logFC sign-concordance test here).

## Set Up

```{r}
#| label: load-packages
Packages <- c("tidyverse", "glue", "patchwork", "svglite", "knitr")
lapply(Packages, library, character.only = TRUE)

source("pseudobulk_functions.R")

set.seed(123)

out_dir <- "../../results/targeted/cut_and_tag"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r}
#| label: load-de-results
# TT: sex-separate GW model; retain males only.
# E2: sex-combined Sex model.
# Significance: FDR < 0.10 AND |logFC| >= 0.25 (matches hormone-script graphs).
TT_results <- read_csv(
  "../../results/targeted/testosterone_genes/run_voom_lm_gw/TT_gw_results.csv",
  show_col_types = FALSE
)
E2_results <- read_csv(
  "../../results/targeted/estradiol_genes/run_voom_lm_sex/E2_sex_results.csv",
  show_col_types = FALSE
)

TT_sig <- filter_sig(TT_results, sex_filter = "M")
E2_sig <- filter_sig(E2_results)

cat("TT sig (male):", nrow(TT_sig), "gene-cell_type rows,",
    length(unique(TT_sig$gene)), "unique genes\n")
cat("E2 sig (combined):", nrow(E2_sig), "gene-cell_type rows,",
    length(unique(E2_sig$gene)), "unique genes\n")
```

```{r}
#| label: load-cut-and-tag
# Two gene-set variants per TF: all annotated peaks, and promoter-proximal
# (|distanceToTSS| < 3000 bp, ChIPseeker default; Yu, Wang & He 2015).
AR   <- load_cut_and_tag_geneset("../../doc/cut&tag/AR_consensus_anno.csv")
ESR1 <- load_cut_and_tag_geneset("../../doc/cut&tag/ESR1_consensus_anno.csv")

ct_sizes <- tibble::tibble(
  TF            = c("AR", "ESR1"),
  all_genes     = c(length(AR$all_genes),      length(ESR1$all_genes)),
  promoter_genes = c(length(AR$promoter_genes), length(ESR1$promoter_genes))
)

knitr::kable(ct_sizes, caption = "CUT&Tag bound-gene set sizes (all vs promoter-proximal)")
```

## Overlap Assembly — Helper

Collapses CUT&Tag peaks to one row per gene (nearest-TSS peak) for annotation,
then intersects with the significant DE table. Presentation-only; the reusable
set logic lives in `pseudobulk_functions.R`.

```{r}
#| label: overlap-helper
build_ct_overlap <- function(sig_df, ct, window = c("all", "promoter")) {
  window   <- match.arg(window)
  gene_set <- if (window == "promoter") ct$promoter_genes else ct$all_genes
  peak_tbl <- if (window == "promoter") {
    dplyr::filter(ct$table, promoter_proximal)
  } else {
    ct$table
  }

  gene_meta <- peak_tbl %>%
    dplyr::group_by(hgnc_symbol) %>%
    dplyr::summarise(
      n_peaks            = dplyr::n(),
      min_abs_dist_tss   = min(abs(distanceToTSS)),
      nearest_annotation = annotation[which.min(abs(distanceToTSS))],
      .groups = "drop"
    )

  overlap_sig_with_geneset(sig_df, gene_set) %>%
    dplyr::left_join(gene_meta, by = c("gene" = "hgnc_symbol")) %>%
    dplyr::select(gene, cell_type, sex, logFC, adj.P.Val,
                  n_peaks, min_abs_dist_tss, nearest_annotation) %>%
    dplyr::arrange(adj.P.Val)
}
```

## AR × Testosterone (Male)

::: panel-tabset
### All annotated peaks

```{r}
#| label: ar-tt-all
ar_tt_all <- build_ct_overlap(TT_sig, AR, "all")
cat("AR-bound (all) × TT sig:", length(unique(ar_tt_all$gene)), "unique genes\n")
write_csv(ar_tt_all, file.path(out_dir, "ar_tt_overlap_all.csv"))
knitr::kable(ar_tt_all, digits = 4, caption = "AR-bound (all peaks) ∩ TT DEGs (male)")
```

### Promoter-proximal

```{r}
#| label: ar-tt-promoter
ar_tt_prom <- build_ct_overlap(TT_sig, AR, "promoter")
cat("AR-bound (promoter) × TT sig:", length(unique(ar_tt_prom$gene)), "unique genes\n")
write_csv(ar_tt_prom, file.path(out_dir, "ar_tt_overlap_promoter.csv"))
knitr::kable(ar_tt_prom, digits = 4, caption = "AR-bound (promoter-proximal) ∩ TT DEGs (male)")
```
:::

## ESR1 × Estradiol

::: panel-tabset
### All annotated peaks

```{r}
#| label: esr1-e2-all
esr1_e2_all <- build_ct_overlap(E2_sig, ESR1, "all")
cat("ESR1-bound (all) × E2 sig:", length(unique(esr1_e2_all$gene)), "unique genes\n")
write_csv(esr1_e2_all, file.path(out_dir, "esr1_e2_overlap_all.csv"))
knitr::kable(esr1_e2_all, digits = 4, caption = "ESR1-bound (all peaks) ∩ E2 DEGs")
```

### Promoter-proximal

```{r}
#| label: esr1-e2-promoter
esr1_e2_prom <- build_ct_overlap(E2_sig, ESR1, "promoter")
cat("ESR1-bound (promoter) × E2 sig:", length(unique(esr1_e2_prom$gene)), "unique genes\n")
write_csv(esr1_e2_prom, file.path(out_dir, "esr1_e2_overlap_promoter.csv"))
knitr::kable(esr1_e2_prom, digits = 4, caption = "ESR1-bound (promoter-proximal) ∩ E2 DEGs")
```
:::

## Overlap Summary

Unique bound-and-associated genes per receptor × window, alongside the total
significant genes available to overlap.

```{r}
#| label: summary-table
overlap_summary <- tibble::tibble(
  Comparison = c("AR × TT (male)", "AR × TT (male)",
                 "ESR1 × E2",      "ESR1 × E2"),
  Window     = c("all", "promoter", "all", "promoter"),
  `DEGs (unique)` = c(length(unique(TT_sig$gene)), length(unique(TT_sig$gene)),
                      length(unique(E2_sig$gene)), length(unique(E2_sig$gene))),
  `Bound & assoc. (unique)` = c(
    length(unique(ar_tt_all$gene)),   length(unique(ar_tt_prom$gene)),
    length(unique(esr1_e2_all$gene)), length(unique(esr1_e2_prom$gene))
  )
) %>%
  dplyr::mutate(`% of DEGs bound` = round(
    100 * `Bound & assoc. (unique)` / `DEGs (unique)`, 1
  ))

write_csv(overlap_summary, file.path(out_dir, "cut_and_tag_overlap_summary.csv"))
knitr::kable(overlap_summary, caption = "CUT&Tag overlap summary")
```

```{r}
#| label: summary-plot
#| fig-width: 9
#| fig-height: 5
plot_df <- overlap_summary %>%
  dplyr::mutate(
    Comparison = factor(Comparison, levels = c("AR × TT (male)", "ESR1 × E2")),
    Window     = factor(Window, levels = c("all", "promoter"))
  )

p_ct <- ggplot(plot_df,
               aes(x = Comparison, y = `Bound & assoc. (unique)`, fill = Window)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = `Bound & assoc. (unique)`),
            position = position_dodge(width = 0.7), vjust = -0.4, size = 4) +
  labs(title = "Hormone DEGs directly bound by cognate receptor",
       x = NULL, y = "# bound & associated genes", fill = "Peak window") +
  scale_fill_manual(values = c("all" = "#9ecae1", "promoter" = "#3182bd")) +
  theme_minimal(base_size = 13)

p_ct

save_dual_format(p_ct, out_dir, "cut_and_tag_overlap_summary",
                 width = 9, height = 5)
```

## Notes

::: incremental
- **ESR2 (ERβ)** CUT&Tag excluded — the assay did not yield usable consensus peaks.
- **Progesterone** excluded — no PGR CUT&Tag track exists in this dataset, so P4
  DEGs have no cognate-binding reference.
- Overlaps are **membership only**; CUT&Tag binding carries no direction, so no
  sign-concordance test is applied (unlike the literature comparisons in
  `SFARI_DEG.qmd`).
- The **promoter-proximal** set (|distanceToTSS| < 3 kb) is the more conservative,
  regulatory-relevant definition and feeds the integrative panel in
  `compare_gene_results.qmd`.
:::
