---
title: "E2 & P4 Log10 Sensitivity Analysis"
subtitle: "Subclass level — log10-transformed hormone predictors"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Sensitivity analysis repeating the E2 and P4 subclass-level limma-voom
models with log10-transformed hormone predictors. All other parameters are
identical to the raw-scale analysis: gene filter (≥ 2 counts in ≥ 3 samples),
`min_samples = 21`, significance threshold FDR < 0.10 and |logFC| ≥ 0.25.

Rationale: hormone concentrations in biological fluids are log-normally
distributed (Feng et al. 2014 *Stat Med*). Modeling on the raw scale fits a
linear relationship through a right-skewed predictor, which can reduce power.
Log10 transformation linearises this relationship; the logFC is then
interpreted as log2-CPM change per 10-fold increase in hormone concentration.

Zero hormone values are set to NA before transforming (log10(0) = -Inf).
The existing `!is.na()` guard in `analyze_hormone_combined()` excludes these
samples before model fitting, identical to how non-zero filtering works on the
raw scale.

---

## Set Up

```{r}
#| label: load-packages
Packages <- c(
  "tidyverse", "Seurat", "readxl", "edgeR", "limma", "svglite", "Matrix"
)
lapply(Packages, library, character.only = TRUE)

source("pseudobulk_functions.R")

set.seed(123)
```

```{r}
#| label: load-data
SSD_data    <- readRDS("../../data/cache/SSD_meta_umap.rds")
pb_base_subclass <- readRDS("../../data/cache/pb_base_subclass.rds")

targeted_comparison_raw <- read_xlsx("../../doc/targeted/targeted_hormones.xlsx")
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
```

```{r}
#| label: config
sample_col        <- "Sample"
pb_sample_pattern <- "tissue\\d+"
```

---

## Outlier Removal

Same IQR-based criteria as the primary analysis (Q3 + 1.5 × IQR), applied
per hormone before transformation.

```{r}
#| label: outlier-removal
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) +
    1.5 * IQR(data[[variable]], na.rm = TRUE)
  data %>% dplyr::filter(.data[[variable]] <= outlier_limit)
}

targeted_comparison_filtered_E2 <- detect_outliers(targeted_comparison_raw, "E2")
targeted_comparison_filtered_P4 <- detect_outliers(targeted_comparison_raw, "P4")

cat("E2 — before:", nrow(targeted_comparison_raw),
    "  after:", nrow(targeted_comparison_filtered_E2), "\n")
cat("P4 — before:", nrow(targeted_comparison_raw),
    "  after:", nrow(targeted_comparison_filtered_P4), "\n")
cat("E2 range (raw):", paste(round(range(targeted_comparison_filtered_E2$E2, na.rm = TRUE), 3),
    collapse = " – "), "ng/mL\n")
cat("P4 range (raw):", paste(round(range(targeted_comparison_filtered_P4$P4, na.rm = TRUE), 3),
    collapse = " – "), "ng/mL\n")
```

---

## Log10 Transformation

```{r}
#| label: log10-transform
# Zeros → NA before log10 to avoid -Inf entering the model.
# New copies preserve the raw-scale objects for reference.
targeted_comparison_filtered_E2_log10 <- targeted_comparison_filtered_E2 %>%
  dplyr::mutate(E2 = dplyr::if_else(E2 == 0, NA_real_, log10(E2)))

targeted_comparison_filtered_P4_log10 <- targeted_comparison_filtered_P4 %>%
  dplyr::mutate(P4 = dplyr::if_else(P4 == 0, NA_real_, log10(P4)))

cat("E2 log10 range:", paste(round(range(targeted_comparison_filtered_E2_log10$E2, na.rm = TRUE), 3),
    collapse = " – "), "\n")
cat("P4 log10 range:", paste(round(range(targeted_comparison_filtered_P4_log10$P4, na.rm = TRUE), 3),
    collapse = " – "), "\n")
cat("E2 zeros → NA:", sum(targeted_comparison_filtered_E2$E2 == 0, na.rm = TRUE), "\n")
cat("P4 zeros → NA:", sum(targeted_comparison_filtered_P4$P4 == 0, na.rm = TRUE), "\n")
```

---

## Pseudobulk

```{r}
#| label: build-pseudobulk
pb_E2_log10 <- merge_hormone_metadata(
  pb_base_subclass,
  targeted_comparison_filtered_E2_log10,
  pb_sample_pattern = pb_sample_pattern,
  min_cells         = 30
) %>% add_batch_fraction(SSD_data)

pb_P4_log10 <- merge_hormone_metadata(
  pb_base_subclass,
  targeted_comparison_filtered_P4_log10,
  pb_sample_pattern = pb_sample_pattern,
  min_cells         = 30
) %>% add_batch_fraction(SSD_data)

cat("E2 pseudobulk samples:", nrow(pb_E2_log10$metadata),
    "  NAs in batch1_frac:", sum(is.na(pb_E2_log10$metadata$batch1_frac)), "\n")
cat("P4 pseudobulk samples:", nrow(pb_P4_log10$metadata),
    "  NAs in batch1_frac:", sum(is.na(pb_P4_log10$metadata$batch1_frac)), "\n")
```

---

## Estradiol

Four models tested. logFC = log2-CPM change per 10-fold increase in E2 (ng/mL).

### Base `~ E2`

```{r}
#| label: e2-base
res_E2_log10_base <- analyze_hormone_combined(
  pb_E2_log10$counts, pb_E2_log10$metadata,
  hormone     = "E2",
  model       = run_voom_lm,
  min_samples = 21
)

write_csv(
  res_E2_log10_base$combined_results,
  "../../results/gene-hormone/estradiol_genes_log10/run_voom_lm/E2_log10_base_results.csv"
)

res_E2_log10_base$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 — E2 log10 Base")
```

### Base + Sex `~ E2 + Sex`

```{r}
#| label: e2-sex
res_E2_log10_sex <- analyze_hormone_combined(
  pb_E2_log10$counts, pb_E2_log10$metadata,
  hormone     = "E2",
  model       = run_voom_lm_sex,
  min_samples = 21
)

write_csv(
  res_E2_log10_sex$combined_results,
  "../../results/gene-hormone/estradiol_genes_log10/run_voom_lm_sex/E2_log10_sex_results.csv"
)

res_E2_log10_sex$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 — E2 log10 Base + Sex")
```

### Base + Batch `~ E2 + batch1_frac`

```{r}
#| label: e2-batch
res_E2_log10_batch <- analyze_hormone_combined(
  pb_E2_log10$counts, pb_E2_log10$metadata,
  hormone     = "E2",
  model       = run_voom_lm_batch,
  min_samples = 21
)

write_csv(
  res_E2_log10_batch$combined_results,
  "../../results/gene-hormone/estradiol_genes_log10/run_voom_lm_batch/E2_log10_batch_results.csv"
)

res_E2_log10_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 — E2 log10 Base + Batch")
```

### Base + Sex + Batch `~ E2 + Sex + batch1_frac`

```{r}
#| label: e2-sex-batch
res_E2_log10_sex_batch <- analyze_hormone_combined(
  pb_E2_log10$counts, pb_E2_log10$metadata,
  hormone     = "E2",
  model       = run_voom_lm_sex_batch,
  min_samples = 21
)

write_csv(
  res_E2_log10_sex_batch$combined_results,
  "../../results/gene-hormone/estradiol_genes_log10/run_voom_lm_sex_batch/E2_log10_sex_batch_results.csv"
)

res_E2_log10_sex_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 — E2 log10 Base + Sex + Batch")
```

---

## Progesterone

Four models tested. logFC = log2-CPM change per 10-fold increase in P4 (ng/mL).

### Base `~ P4`

```{r}
#| label: p4-base
res_P4_log10_base <- analyze_hormone_combined(
  pb_P4_log10$counts, pb_P4_log10$metadata,
  hormone     = "P4",
  model       = run_voom_lm,
  min_samples = 21
)

write_csv(
  res_P4_log10_base$combined_results,
  "../../results/gene-hormone/progesterone_genes_log10/run_voom_lm/P4_log10_base_results.csv"
)

res_P4_log10_base$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 — P4 log10 Base")
```

### Base + Sex `~ P4 + Sex`

```{r}
#| label: p4-sex
res_P4_log10_sex <- analyze_hormone_combined(
  pb_P4_log10$counts, pb_P4_log10$metadata,
  hormone     = "P4",
  model       = run_voom_lm_sex,
  min_samples = 21
)

write_csv(
  res_P4_log10_sex$combined_results,
  "../../results/gene-hormone/progesterone_genes_log10/run_voom_lm_sex/P4_log10_sex_results.csv"
)

res_P4_log10_sex$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 — P4 log10 Base + Sex")
```

### Base + Batch `~ P4 + batch1_frac`

```{r}
#| label: p4-batch
res_P4_log10_batch <- analyze_hormone_combined(
  pb_P4_log10$counts, pb_P4_log10$metadata,
  hormone     = "P4",
  model       = run_voom_lm_batch,
  min_samples = 21
)

write_csv(
  res_P4_log10_batch$combined_results,
  "../../results/gene-hormone/progesterone_genes_log10/run_voom_lm_batch/P4_log10_batch_results.csv"
)

res_P4_log10_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 — P4 log10 Base + Batch")
```

### Base + Sex + Batch `~ P4 + Sex + batch1_frac`

```{r}
#| label: p4-sex-batch
res_P4_log10_sex_batch <- analyze_hormone_combined(
  pb_P4_log10$counts, pb_P4_log10$metadata,
  hormone     = "P4",
  model       = run_voom_lm_sex_batch,
  min_samples = 21
)

write_csv(
  res_P4_log10_sex_batch$combined_results,
  "../../results/gene-hormone/progesterone_genes_log10/run_voom_lm_sex_batch/P4_log10_sex_batch_results.csv"
)

res_P4_log10_sex_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 — P4 log10 Base + Sex + Batch")
```

---

## Summary

Significant genes defined as FDR < 0.10 and |logFC| ≥ 0.25.

```{r}
#| label: summary-table
dplyr::bind_rows(
  summarize_sig_genes(res_E2_log10_base$combined_results)      %>% dplyr::mutate(hormone = "E2", model = "~ E2"),
  summarize_sig_genes(res_E2_log10_sex$combined_results)       %>% dplyr::mutate(hormone = "E2", model = "~ E2 + Sex"),
  summarize_sig_genes(res_E2_log10_batch$combined_results)     %>% dplyr::mutate(hormone = "E2", model = "~ E2 + batch1_frac"),
  summarize_sig_genes(res_E2_log10_sex_batch$combined_results) %>% dplyr::mutate(hormone = "E2", model = "~ E2 + Sex + batch1_frac"),
  summarize_sig_genes(res_P4_log10_base$combined_results)      %>% dplyr::mutate(hormone = "P4", model = "~ P4"),
  summarize_sig_genes(res_P4_log10_sex$combined_results)       %>% dplyr::mutate(hormone = "P4", model = "~ P4 + Sex"),
  summarize_sig_genes(res_P4_log10_batch$combined_results)     %>% dplyr::mutate(hormone = "P4", model = "~ P4 + batch1_frac"),
  summarize_sig_genes(res_P4_log10_sex_batch$combined_results) %>% dplyr::mutate(hormone = "P4", model = "~ P4 + Sex + batch1_frac")
) %>%
  dplyr::select(hormone, model, cell_type, num_sig_genes) %>%
  dplyr::arrange(hormone, model, dplyr::desc(num_sig_genes)) %>%
  knitr::kable(caption = "Log10-transformed predictors: significant genes per hormone, model, and subclass")
```

```{r}
#| label: summary-totals
dplyr::bind_rows(
  summarize_sig_genes(res_E2_log10_base$combined_results)      %>% dplyr::mutate(hormone = "E2", model = "~ E2"),
  summarize_sig_genes(res_E2_log10_sex$combined_results)       %>% dplyr::mutate(hormone = "E2", model = "~ E2 + Sex"),
  summarize_sig_genes(res_E2_log10_batch$combined_results)     %>% dplyr::mutate(hormone = "E2", model = "~ E2 + batch1_frac"),
  summarize_sig_genes(res_E2_log10_sex_batch$combined_results) %>% dplyr::mutate(hormone = "E2", model = "~ E2 + Sex + batch1_frac"),
  summarize_sig_genes(res_P4_log10_base$combined_results)      %>% dplyr::mutate(hormone = "P4", model = "~ P4"),
  summarize_sig_genes(res_P4_log10_sex$combined_results)       %>% dplyr::mutate(hormone = "P4", model = "~ P4 + Sex"),
  summarize_sig_genes(res_P4_log10_batch$combined_results)     %>% dplyr::mutate(hormone = "P4", model = "~ P4 + batch1_frac"),
  summarize_sig_genes(res_P4_log10_sex_batch$combined_results) %>% dplyr::mutate(hormone = "P4", model = "~ P4 + Sex + batch1_frac")
) %>%
  dplyr::group_by(hormone, model) %>%
  dplyr::summarise(total_sig_genes = sum(num_sig_genes), .groups = "drop") %>%
  knitr::kable(caption = "Total significant genes per hormone and model")
```
