---
title: "Estradiol & Progesterone Gene Expression Prediction — Subclass"
subtitle: "Sneha Chaturvedi, Emma Jones"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    code-tools: true
    scrollable: true
    theme: dark
    embed-resources: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

## Goals

Identify genes whose pseudobulk expression is predicted by estradiol (E2) and progesterone (P4) concentration, by **subclass**. Both hormones are analyzed **sex-combined**: XX and XY samples are pooled, with Sex optionally included as an additive covariate.

**Four models tested per hormone:**

::: incremental
-   **Base** — `~ hormone`
-   **Base + Sex** — `~ hormone + Sex`
-   **Base + Batch** — `~ hormone + batch1_frac`
-   **Base + Sex + Batch** — `~ hormone + Sex + batch1_frac`
:::

All four models use `analyze_hormone_combined()`. The **sex guard** — dropping groups where fewer than 2 sexes are present after hormone filtering — applies to every model.

**Gene filter:** manual rowSums — genes with ≥ 2 counts in ≥ 3 samples per subclass.\
**Sample threshold:** `floor(n_group / 2)` non-zero hormone samples required per subclass.

## Set Up

-   Load packages

    ```{r}
    #| label: load-packages
    #| output: false
    Packages <- c(
      "tidyverse", "Seurat", "Signac", "glue", "readxl", "cowplot",
      "edgeR", "limma", "patchwork", "svglite", "Matrix", "ggpubr"
    )
    lapply(Packages, library, character.only = TRUE)

    source("pseudobulk_functions.R")

    set.seed(123)
    ```

-   Load data

    ```{r}
    #| label: load-data
    SSD_data <- readRDS("../../data/cache/SSD_meta_umap.rds")
    pb_base  <- readRDS("../../data/cache/pb_base_subclass.rds")
    targeted_comparison_raw <- read_xlsx("../../doc/targeted/targeted_hormones.xlsx")
    targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
    targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
    targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`
    ```

-   Configuration

    ```{r}
    #| label: config
    celltype_col      <- "subclass"
    sample_col        <- "Sample"
    pb_sample_pattern <- "tissue\\d+"

    out_e2 <- "../../results/gene-hormone/estradiol_genes_subclass"
    out_p4 <- "../../results/gene-hormone/progesterone_genes_subclass"

    for (sub in c("run_voom_lm", "run_voom_lm_sex",
                  "run_voom_lm_batch", "run_voom_lm_sex_batch")) {
      dir.create(file.path(out_e2, sub), recursive = TRUE, showWarnings = FALSE)
      dir.create(file.path(out_p4, sub), recursive = TRUE, showWarnings = FALSE)
    }
    ```

```{r}
#| label: create-output-dirs
dirs <- c(
  "../../results/gene-hormone/estradiol_genes_log10/run_voom_lm",
  "../../results/gene-hormone/estradiol_genes_log10/run_voom_lm_sex",
  "../../results/gene-hormone/estradiol_genes_log10/run_voom_lm_batch",
  "../../results/gene-hormone/estradiol_genes_log10/run_voom_lm_sex_batch",
  "../../results/gene-hormone/progesterone_genes_log10/run_voom_lm",
  "../../results/gene-hormone/progesterone_genes_log10/run_voom_lm_sex",
  "../../results/gene-hormone/progesterone_genes_log10/run_voom_lm_batch",
  "../../results/gene-hormone/progesterone_genes_log10/run_voom_lm_sex_batch"
)
purrr::walk(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)
```
## Outlier Removal

Samples with hormone \> Q3 + 1.5 × IQR are excluded per hormone before pseudobulk construction.

```{r}
#| label: outlier-removal
detect_outliers <- function(data, variable) {
  outlier_limit <- quantile(data[[variable]], 0.75, na.rm = TRUE) +
    1.5 * IQR(data[[variable]], na.rm = TRUE)
  data %>% dplyr::filter(.data[[variable]] <= outlier_limit)
}

targeted_comparison_filtered_E2 <- detect_outliers(targeted_comparison_raw, "E2")
targeted_comparison_filtered_P4 <- detect_outliers(targeted_comparison_raw, "P4")

cat(
  "E2 — before:", nrow(targeted_comparison_raw),
  "  after:", nrow(targeted_comparison_filtered_E2), "\n"
)
cat(
  "P4 — before:", nrow(targeted_comparison_raw),
  "  after:", nrow(targeted_comparison_filtered_P4), "\n"
)
cat("E2 range after filtering:      ",
    paste(round(range(targeted_comparison_filtered_E2$E2, na.rm = TRUE), 3),
          collapse = " – "), "ng/mL\n")
cat("P4 range after filtering:      ",
    paste(round(range(targeted_comparison_filtered_P4$P4, na.rm = TRUE), 3),
          collapse = " – "), "ng/mL\n")

outliers_E2 <- dplyr::anti_join(targeted_comparison_raw,
                                targeted_comparison_filtered_E2, by = "Sample")
outliers_P4 <- dplyr::anti_join(targeted_comparison_raw,
                                targeted_comparison_filtered_P4, by = "Sample")

cat("E2 outliers removed:",
    if (nrow(outliers_E2)) paste(outliers_E2$Sample, collapse = ", ") else "none", "\n")
cat("P4 outliers removed:",
    if (nrow(outliers_P4)) paste(outliers_P4$Sample, collapse = ", ") else "none", "\n")
```

## Pseudobulk

Min 30 cells per (sample × subclass) for inclusion. `add_batch_fraction()` computes the fraction of batch1 cells per biological sample and joins it onto the pseudobulk metadata for each hormone.

```{r}
#| label: build-pseudobulk
pb_E2 <- merge_hormone_metadata(
  pb_base, targeted_comparison_filtered_E2,
  pb_sample_pattern = pb_sample_pattern, min_cells = 30
) %>% add_batch_fraction(SSD_data)

pb_P4 <- merge_hormone_metadata(
  pb_base, targeted_comparison_filtered_P4,
  pb_sample_pattern = pb_sample_pattern, min_cells = 30
) %>% add_batch_fraction(SSD_data)

cat(
  "E2 pseudobulk samples:", nrow(pb_E2$metadata),
  "  NAs in batch1_frac:", sum(is.na(pb_E2$metadata$batch1_frac)), "\n"
)
cat("E2 batch1_frac summary:\n")
print(summary(pb_E2$metadata$batch1_frac))
cat(
  "P4 pseudobulk samples:", nrow(pb_P4$metadata),
  "  NAs in batch1_frac:", sum(is.na(pb_P4$metadata$batch1_frac)), "\n"
)
cat("P4 batch1_frac summary:\n")
print(summary(pb_P4$metadata$batch1_frac))
```

## Models

| Model | Formula | Notes |
|---|---|---|
| Base | `~ hormone` | Combined-sample estimate; no adjustment |
| Base + Sex | `~ hormone + Sex` | Controls for sex-linked expression differences |
| Base + Batch | `~ hormone + batch1_frac` | Batch1 fraction per sample (continuous); absorbs residual batch composition differences |
| Base + Sex + Batch | `~ hormone + Sex + batch1_frac` | Both adjustments |

## Estradiol

### Base

`~ E2` — sex-combined, no covariates.

```{r}
#| label: e2-base-run
res_E2_base <- analyze_hormone_combined(
  pb_E2$counts, pb_E2$metadata,
  hormone = "E2", model = run_voom_lm, min_samples = 21
)

summary_E2_base <- summarize_sig_genes(res_E2_base$combined_results)

plot_top_genes_from_subsets(
  res_E2_base$by_group_data,
  output_dir = file.path(out_e2, "run_voom_lm"),
  hormone    = "E2"
)

save_dual_format(
  plot_sig_gene_summary(summary_E2_base, "E2 — Base (~ E2) | Subclass"),
  output_dir    = file.path(out_e2, "run_voom_lm"),
  filename_base = "E2_base_sig_gene_summary"
)

write_csv(
  res_E2_base$combined_results,
  file.path(out_e2, "run_voom_lm", "E2_base_results.csv")
)
write_csv(
  summary_E2_base,
  file.path(out_e2, "run_voom_lm", "E2_base_summary.csv")
)
```

::: panel-tabset
#### Significant Genes

```{r}
#| label: e2-base-sig-plot
#| fig-width: 10
#| fig-height: 5
plot_sig_gene_summary(summary_E2_base, "E2 — Base (~ E2) | Subclass")
```

#### Top Results

```{r}
#| label: e2-base-top-table
res_E2_base$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — E2 Base")
```
:::

### Base + Sex

`~ E2 + Sex`

```{r}
#| label: e2-sex-run
res_E2_sex <- analyze_hormone_combined(
  pb_E2$counts, pb_E2$metadata,
  hormone = "E2", model = run_voom_lm_sex, min_samples = 21
)

summary_E2_sex <- summarize_sig_genes(res_E2_sex$combined_results)

plot_top_genes_from_subsets(
  res_E2_sex$by_group_data,
  output_dir = file.path(out_e2, "run_voom_lm_sex"),
  hormone    = "E2"
)

save_dual_format(
  plot_sig_gene_summary(summary_E2_sex, "E2 — Base + Sex (~ E2 + Sex) | Subclass"),
  output_dir    = file.path(out_e2, "run_voom_lm_sex"),
  filename_base = "E2_sex_sig_gene_summary"
)

write_csv(
  res_E2_sex$combined_results,
  file.path(out_e2, "run_voom_lm_sex", "E2_sex_results.csv")
)
write_csv(
  summary_E2_sex,
  file.path(out_e2, "run_voom_lm_sex", "E2_sex_summary.csv")
)
```

::: panel-tabset
#### Significant Genes

```{r}
#| label: e2-sex-sig-plot
#| fig-width: 10
#| fig-height: 5
plot_sig_gene_summary(summary_E2_sex, "E2 — Base + Sex (~ E2 + Sex) | Subclass")
```

#### Top Results

```{r}
#| label: e2-sex-top-table
res_E2_sex$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — E2 Base + Sex")
```
:::

### Base + Batch

`~ E2 + batch1_frac`

```{r}
#| label: e2-batch-run
res_E2_batch <- analyze_hormone_combined(
  pb_E2$counts, pb_E2$metadata,
  hormone = "E2", model = run_voom_lm_batch, min_samples = 21
)

summary_E2_batch <- summarize_sig_genes(res_E2_batch$combined_results)

plot_top_genes_from_subsets(
  res_E2_batch$by_group_data,
  output_dir = file.path(out_e2, "run_voom_lm_batch"),
  hormone    = "E2"
)

save_dual_format(
  plot_sig_gene_summary(summary_E2_batch, "E2 — Base + Batch (~ E2 + batch1_frac) | Subclass"),
  output_dir    = file.path(out_e2, "run_voom_lm_batch"),
  filename_base = "E2_batch_sig_gene_summary"
)

write_csv(
  res_E2_batch$combined_results,
  file.path(out_e2, "run_voom_lm_batch", "E2_batch_results.csv")
)
write_csv(
  summary_E2_batch,
  file.path(out_e2, "run_voom_lm_batch", "E2_batch_summary.csv")
)
```

::: panel-tabset
#### Significant Genes

```{r}
#| label: e2-batch-sig-plot
#| fig-width: 10
#| fig-height: 5
plot_sig_gene_summary(summary_E2_batch, "E2 — Base + Batch (~ E2 + batch1_frac) | Subclass")
```

#### Top Results

```{r}
#| label: e2-batch-top-table
res_E2_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — E2 Base + Batch")
```
:::

### Base + Sex + Batch

`~ E2 + Sex + batch1_frac`

```{r}
#| label: e2-sex-batch-run
res_E2_sex_batch <- analyze_hormone_combined(
  pb_E2$counts, pb_E2$metadata,
  hormone = "E2", model = run_voom_lm_sex_batch, min_samples = 21
)

summary_E2_sex_batch <- summarize_sig_genes(res_E2_sex_batch$combined_results)

plot_top_genes_from_subsets(
  res_E2_sex_batch$by_group_data,
  output_dir = file.path(out_e2, "run_voom_lm_sex_batch"),
  hormone    = "E2"
)

save_dual_format(
  plot_sig_gene_summary(
    summary_E2_sex_batch,
    "E2 — Base + Sex + Batch (~ E2 + Sex + batch1_frac) | Subclass"
  ),
  output_dir    = file.path(out_e2, "run_voom_lm_sex_batch"),
  filename_base = "E2_sex_batch_sig_gene_summary"
)

write_csv(
  res_E2_sex_batch$combined_results,
  file.path(out_e2, "run_voom_lm_sex_batch", "E2_sex_batch_results.csv")
)
write_csv(
  summary_E2_sex_batch,
  file.path(out_e2, "run_voom_lm_sex_batch", "E2_sex_batch_summary.csv")
)
```

::: panel-tabset
#### Significant Genes

```{r}
#| label: e2-sex-batch-sig-plot
#| fig-width: 10
#| fig-height: 5
plot_sig_gene_summary(
  summary_E2_sex_batch,
  "E2 — Base + Sex + Batch (~ E2 + Sex + batch1_frac) | Subclass"
)
```

#### Top Results

```{r}
#| label: e2-sex-batch-top-table
res_E2_sex_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — E2 Base + Sex + Batch")
```
:::

## Progesterone

### Base

`~ P4` — sex-combined, no covariates.

```{r}
#| label: p4-base-run
res_P4_base <- analyze_hormone_combined(
  pb_P4$counts, pb_P4$metadata,
  hormone = "P4", model = run_voom_lm, min_samples = 21
)

summary_P4_base <- summarize_sig_genes(res_P4_base$combined_results)

plot_top_genes_from_subsets(
  res_P4_base$by_group_data,
  output_dir = file.path(out_p4, "run_voom_lm"),
  hormone    = "P4"
)

save_dual_format(
  plot_sig_gene_summary(summary_P4_base, "P4 — Base (~ P4) | Subclass"),
  output_dir    = file.path(out_p4, "run_voom_lm"),
  filename_base = "P4_base_sig_gene_summary"
)

write_csv(
  res_P4_base$combined_results,
  file.path(out_p4, "run_voom_lm", "P4_base_results.csv")
)
write_csv(
  summary_P4_base,
  file.path(out_p4, "run_voom_lm", "P4_base_summary.csv")
)
```

::: panel-tabset
#### Significant Genes

```{r}
#| label: p4-base-sig-plot
#| fig-width: 10
#| fig-height: 5
plot_sig_gene_summary(summary_P4_base, "P4 — Base (~ P4) | Subclass")
```

#### Top Results

```{r}
#| label: p4-base-top-table
res_P4_base$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — P4 Base")
```
:::

### Base + Sex

`~ P4 + Sex`

```{r}
#| label: p4-sex-run
res_P4_sex <- analyze_hormone_combined(
  pb_P4$counts, pb_P4$metadata,
  hormone = "P4", model = run_voom_lm_sex, min_samples = 21
)

summary_P4_sex <- summarize_sig_genes(res_P4_sex$combined_results)

plot_top_genes_from_subsets(
  res_P4_sex$by_group_data,
  output_dir = file.path(out_p4, "run_voom_lm_sex"),
  hormone    = "P4"
)

save_dual_format(
  plot_sig_gene_summary(summary_P4_sex, "P4 — Base + Sex (~ P4 + Sex) | Subclass"),
  output_dir    = file.path(out_p4, "run_voom_lm_sex"),
  filename_base = "P4_sex_sig_gene_summary"
)

write_csv(
  res_P4_sex$combined_results,
  file.path(out_p4, "run_voom_lm_sex", "P4_sex_results.csv")
)
write_csv(
  summary_P4_sex,
  file.path(out_p4, "run_voom_lm_sex", "P4_sex_summary.csv")
)
```

::: panel-tabset
#### Significant Genes

```{r}
#| label: p4-sex-sig-plot
#| fig-width: 10
#| fig-height: 5
plot_sig_gene_summary(summary_P4_sex, "P4 — Base + Sex (~ P4 + Sex) | Subclass")
```

#### Top Results

```{r}
#| label: p4-sex-top-table
res_P4_sex$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — P4 Base + Sex")
```
:::

### Base + Batch

`~ P4 + batch1_frac`

```{r}
#| label: p4-batch-run
res_P4_batch <- analyze_hormone_combined(
  pb_P4$counts, pb_P4$metadata,
  hormone = "P4", model = run_voom_lm_batch, min_samples = 21
)

summary_P4_batch <- summarize_sig_genes(res_P4_batch$combined_results)

plot_top_genes_from_subsets(
  res_P4_batch$by_group_data,
  output_dir = file.path(out_p4, "run_voom_lm_batch"),
  hormone    = "P4"
)

save_dual_format(
  plot_sig_gene_summary(summary_P4_batch, "P4 — Base + Batch (~ P4 + batch1_frac) | Subclass"),
  output_dir    = file.path(out_p4, "run_voom_lm_batch"),
  filename_base = "P4_batch_sig_gene_summary"
)

write_csv(
  res_P4_batch$combined_results,
  file.path(out_p4, "run_voom_lm_batch", "P4_batch_results.csv")
)
write_csv(
  summary_P4_batch,
  file.path(out_p4, "run_voom_lm_batch", "P4_batch_summary.csv")
)
```

::: panel-tabset
#### Significant Genes

```{r}
#| label: p4-batch-sig-plot
#| fig-width: 10
#| fig-height: 5
plot_sig_gene_summary(summary_P4_batch, "P4 — Base + Batch (~ P4 + batch1_frac) | Subclass")
```

#### Top Results

```{r}
#| label: p4-batch-top-table
res_P4_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — P4 Base + Batch")
```
:::

### Base + Sex + Batch

`~ P4 + Sex + batch1_frac`

```{r}
#| label: p4-sex-batch-run
res_P4_sex_batch <- analyze_hormone_combined(
  pb_P4$counts, pb_P4$metadata,
  hormone = "P4", model = run_voom_lm_sex_batch, min_samples = 21
)

summary_P4_sex_batch <- summarize_sig_genes(res_P4_sex_batch$combined_results)

plot_top_genes_from_subsets(
  res_P4_sex_batch$by_group_data,
  output_dir = file.path(out_p4, "run_voom_lm_sex_batch"),
  hormone    = "P4"
)

save_dual_format(
  plot_sig_gene_summary(
    summary_P4_sex_batch,
    "P4 — Base + Sex + Batch (~ P4 + Sex + batch1_frac) | Subclass"
  ),
  output_dir    = file.path(out_p4, "run_voom_lm_sex_batch"),
  filename_base = "P4_sex_batch_sig_gene_summary"
)

write_csv(
  res_P4_sex_batch$combined_results,
  file.path(out_p4, "run_voom_lm_sex_batch", "P4_sex_batch_results.csv")
)
write_csv(
  summary_P4_sex_batch,
  file.path(out_p4, "run_voom_lm_sex_batch", "P4_sex_batch_summary.csv")
)
```

::: panel-tabset
#### Significant Genes

```{r}
#| label: p4-sex-batch-sig-plot
#| fig-width: 10
#| fig-height: 5
plot_sig_gene_summary(
  summary_P4_sex_batch,
  "P4 — Base + Sex + Batch (~ P4 + Sex + batch1_frac) | Subclass"
)
```

#### Top Results

```{r}
#| label: p4-sex-batch-top-table
res_P4_sex_batch$combined_results %>%
  dplyr::arrange(adj.P.Val) %>%
  dplyr::select(gene, cell_type, sex, logFC, adj.P.Val) %>%
  head(10) %>%
  knitr::kable(digits = 4, caption = "Top 10 hits — P4 Base + Sex + Batch")
```
:::

## Summary

```{r}
#| label: e2-cross-model-summary
cross_model_E2 <- dplyr::bind_rows(
  summarize_sig_genes(res_E2_base$combined_results) %>%
    dplyr::mutate(model = "~ E2"),
  summarize_sig_genes(res_E2_sex$combined_results) %>%
    dplyr::mutate(model = "~ E2 + Sex"),
  summarize_sig_genes(res_E2_batch$combined_results) %>%
    dplyr::mutate(model = "~ E2 + batch1_frac"),
  summarize_sig_genes(res_E2_sex_batch$combined_results) %>%
    dplyr::mutate(model = "~ E2 + Sex + batch1_frac")
)

knitr::kable(
  cross_model_E2 %>%
    dplyr::select(model, cell_type, num_sig_genes) %>%
    dplyr::arrange(model, desc(num_sig_genes)),
  caption = "E2: significant genes (FDR < 0.10, |logFC| ≥ 0.25) per model"
)
```

```{r}
#| label: p4-cross-model-summary
cross_model_P4 <- dplyr::bind_rows(
  summarize_sig_genes(res_P4_base$combined_results) %>%
    dplyr::mutate(model = "~ P4"),
  summarize_sig_genes(res_P4_sex$combined_results) %>%
    dplyr::mutate(model = "~ P4 + Sex"),
  summarize_sig_genes(res_P4_batch$combined_results) %>%
    dplyr::mutate(model = "~ P4 + batch1_frac"),
  summarize_sig_genes(res_P4_sex_batch$combined_results) %>%
    dplyr::mutate(model = "~ P4 + Sex + batch1_frac")
)

knitr::kable(
  cross_model_P4 %>%
    dplyr::select(model, cell_type, num_sig_genes) %>%
    dplyr::arrange(model, desc(num_sig_genes)),
  caption = "P4: significant genes (FDR < 0.10, |logFC| ≥ 0.25) per model"
)
```

```{r}
#| label: combined-totals
dplyr::bind_rows(
  cross_model_E2 %>%
    dplyr::group_by(model) %>%
    dplyr::summarise(total_sig_genes = sum(num_sig_genes), .groups = "drop") %>%
    dplyr::mutate(hormone = "E2"),
  cross_model_P4 %>%
    dplyr::group_by(model) %>%
    dplyr::summarise(total_sig_genes = sum(num_sig_genes), .groups = "drop") %>%
    dplyr::mutate(hormone = "P4")
) %>%
  dplyr::select(hormone, model, total_sig_genes) %>%
  knitr::kable(caption = "Total significant genes per hormone and model")
```

## E2 Analysis UMAP

Cells included in the estradiol analysis: from subclass groups that passed all thresholds (outlier removal, ≥ 30 cells per pseudobulk sample, gene filter, non-zero E2 sample threshold, and sex guard requiring both sexes present). All four E2 models operate on identical groups. Cells are colored by subclass.

```{r}
#| label: e2-analysis-umap
#| fig-width: 10
#| fig-height: 8
plot_analysis_umap(
  seurat_obj    = SSD_data,
  by_group_data = res_E2_base$by_group_data,
  title         = "Estradiol Analysis — Included Cells | Subclass",
  celltype_col  = celltype_col,
  sex_separate  = FALSE,
  output_dir    = out_e2,
  filename_base = "E2_analysis_umap"
)
```

## P4 Analysis UMAP

Cells included in the progesterone analysis: from subclass groups that passed all thresholds. Outlier removal for P4 is independent of E2, so the included sample set may differ between hormones. Cells are colored by subclass.

```{r}
#| label: p4-analysis-umap
#| fig-width: 10
#| fig-height: 8
plot_analysis_umap(
  seurat_obj    = SSD_data,
  by_group_data = res_P4_base$by_group_data,
  title         = "Progesterone Analysis — Included Cells | Subclass",
  celltype_col  = celltype_col,
  sex_separate  = FALSE,
  output_dir    = out_p4,
  filename_base = "P4_analysis_umap"
)
```
