---
title: "Hormone Receptor Gene Expression"
subtitle: "Sneha Chaturvedi, Emma Jones"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    code-tools: true
    scrollable: true
    theme: dark
    embed-resources: true
    chalkboard: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

## Goals

Examine expression of gonadal hormone receptor genes across cell types in the
fetal brain snRNA-seq dataset.

**Receptors examined:**

::: incremental
- **Nuclear receptors:** AR, ESR1, ESR2, PGR
- **Gonadotropin receptors:** FSHR, LHCGR, GNRHR
- **Prostaglandin E receptors:** PTGER1–4
- **Thyroid hormone receptors:** THRA, THRB
- **Membrane progesterone receptors:** PAQR5, PAQR7, PAQR8, PGRMC1, PGRMC2
:::

## Set Up

```{r}
#| label: load-packages
Packages <- c(
  "tidyverse", "Seurat", "Signac", "glue", "readxl", "cowplot",
  "edgeR", "limma", "patchwork", "svglite", "Matrix", "viridis"
)
lapply(Packages, library, character.only = TRUE)

set.seed(123)
```

```{r}
#| label: load-data
SSD_data <- readRDS("../../data/cache/receptor_expr_object.rds")
DefaultAssay(SSD_data) <- "RNA"
Idents(SSD_data)       <- "celltype"
```

```{r}
#| label: receptor-genes
receptor_genes <- c(
  "ESR1", "ESR2", "PGR", "AR",
  "FSHR", "LHCGR", "GNRHR",
  "PTGER1", "PTGER2", "PTGER3", "PTGER4",
  "THRA", "THRB",
  "PAQR7", "PAQR8", "PAQR5", "PGRMC1", "PGRMC2"
)

# Confirm which genes are present in the assay
cat("Receptor genes in assay:",
    sum(receptor_genes %in% rownames(SSD_data)), "/", length(receptor_genes), "\n")
cat("Missing:", setdiff(receptor_genes, rownames(SSD_data)), "\n")
```

## Cell Type UMAP

```{r}
#| label: dimplot
#| fig-width: 10
#| fig-height: 7
DimPlot(SSD_data, group.by = "celltype", label = TRUE, repel = TRUE) +
  theme(legend.position = "right")
```

## VlnPlots — Key Nuclear Receptors

::: panel-tabset
### AR

```{r}
#| label: vln-ar
#| fig-width: 12
#| fig-height: 5
VlnPlot(SSD_data, features = "AR", group.by = "celltype", pt.size = 0) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

### ESR1

```{r}
#| label: vln-esr1
#| fig-width: 12
#| fig-height: 5
VlnPlot(SSD_data, features = "ESR1", group.by = "celltype", pt.size = 0) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

### ESR2

```{r}
#| label: vln-esr2
#| fig-width: 12
#| fig-height: 5
VlnPlot(SSD_data, features = "ESR2", group.by = "celltype", pt.size = 0) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

### PGR

```{r}
#| label: vln-pgr
#| fig-width: 12
#| fig-height: 5
VlnPlot(SSD_data, features = "PGR", group.by = "celltype", pt.size = 0) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```
:::

## FeaturePlots — Key Nuclear Receptors

::: panel-tabset
### AR

```{r}
#| label: feat-ar
#| fig-width: 8
#| fig-height: 6
FeaturePlot(SSD_data, features = "AR",
            cols = c("lightgrey", "#440154FF"), alpha = 0.7)
```

### ESR1

```{r}
#| label: feat-esr1
#| fig-width: 8
#| fig-height: 6
FeaturePlot(SSD_data, features = "ESR1",
            cols = c("lightgrey", "#440154FF"), alpha = 0.7)
```

### ESR2

```{r}
#| label: feat-esr2
#| fig-width: 8
#| fig-height: 6
FeaturePlot(SSD_data, features = "ESR2",
            cols = c("lightgrey", "#440154FF"), alpha = 0.7)
```

### PGR

```{r}
#| label: feat-pgr
#| fig-width: 8
#| fig-height: 6
FeaturePlot(SSD_data, features = "PGR",
            cols = c("lightgrey", "#440154FF"), alpha = 0.7)
```

### All Receptors

```{r}
#| label: feat-all
#| fig-width: 20
#| fig-height: 10
FeaturePlot(SSD_data, features = receptor_genes,
            cols = c("lightgrey", "#440154FF"), alpha = 0.7,
            ncol = 6)
```
:::

## ESR2 Expression by Cell Type

Percentage of ESR2-expressing cells per cell type at three expression
thresholds (applied to the log-normalized RNA data layer).

```{r}
#| label: esr2-pct
#| fig-width: 12
#| fig-height: 8
esr2_expr <- as.numeric(
  GetAssayData(SSD_data, assay = "RNA", layer = "data")["ESR2", ]
)
SSD_data$esr2_expr <- esr2_expr

esr2_pct <- SSD_data@meta.data %>%
  dplyr::group_by(celltype) %>%
  dplyr::summarise(
    `Any (> 0)`    = mean(esr2_expr > 0)   * 100,
    `Medium (> 1)` = mean(esr2_expr > 1)   * 100,
    `High (> 1.5)` = mean(esr2_expr > 1.5) * 100,
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(cols = -celltype,
                      names_to = "threshold", values_to = "pct") %>%
  dplyr::mutate(threshold = factor(threshold,
                                   levels = c("Any (> 0)", "Medium (> 1)", "High (> 1.5)")))

ggplot(esr2_pct, aes(x = celltype, y = pct, fill = pct)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ threshold, ncol = 1, scales = "free_y") +
  scale_fill_viridis_c() +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right") +
  labs(x = "Cell Type", y = "% ESR2-expressing cells",
       fill = "% ESR2+",
       title = "ESR2 Expression by Cell Type")
```

## Save Plots

```{r}
#| label: save-plots
dir.create("../../results/targeted/single_cell_plots", showWarnings = FALSE, recursive = TRUE)

# All-receptor UMAP panel
FeaturePlot(SSD_data, features = receptor_genes,
            cols = c("lightgrey", "#440154FF"), alpha = 0.7, ncol = 6)
ggsave("../../results/targeted/single_cell_plots/hormone_receptors_umap.png",
       width = 20, height = 10, dpi = 300)

# ESR2 percentage plot
last_plot()
```
