---
title: "SFARI Gene Overlap with Hormone-Associated DEGs"
subtitle: "Sneha Chaturvedi, Emma Jones"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    code-tools: true
    scrollable: true
    theme: dark
    embed-resources: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

## Goals

Determine whether significant hormone-associated DEGs overlap with SFARI autism
risk genes (<https://gene.sfari.org/database/human-gene/>) and with published
androgen-response gene lists.

**Gene lists compared:**

::: incremental
- **SFARI** — high-confidence risk genes (score ≤ 2)
- **Human brain organoid androgen response** — [Kelava et al. 2021, *Nature*](https://www.nature.com/articles/s41586-021-04330-4)
- **Mouse embryonic neural stem/progenitor cells** — [Bramble et al. 2016, *Sci Rep*](https://www.nature.com/articles/srep36916)
- **Single-cell 17-tissue mouse androgen atlas** — [Li et al. 2024, *Nature*](https://www.nature.com/articles/s41586-024-07291-6)
:::

## Set Up

```{r}
#| label: load-packages
Packages <- c(
  "tidyverse", "Seurat", "readxl", "glue", "cowplot",
  "edgeR", "limma", "patchwork", "svglite", "Matrix"
)
lapply(Packages, library, character.only = TRUE)

source("pseudobulk_functions.R")

set.seed(123)
```

```{r}
#| label: load-gene-lists
SFARI_Gene <- read_csv(
  "../../doc/gene_lists/SFARI-Gene_genes_05-01-2026release_07-04-2026export.csv",
  col_types = cols(status = col_skip())
)

# High-confidence SFARI genes only (scores 1 and 2)
SFARI_Gene_filt <- SFARI_Gene %>% dplyr::filter(`gene-score` <= 2)

androgen_response_up   <- read_csv("../../doc/targeted/Upregulated_DEGs.csv",   col_names = FALSE)
androgen_response_down <- read_csv("../../doc/targeted/Downregulated_DEGs.csv", col_names = FALSE)

mouse_response     <- read_csv("../../doc/targeted/converted_sex_tt_mouse.csv")
singlecell_response <- read_xlsx(
  "../../doc/targeted/li_2024_sex_diff_degs.xlsx",
  sheet = "AASB-DEGs", skip = 1
)
```

```{r}
#| label: load-de-results
# TT: sex-separate GW model (run_voom_lm_gw); retain males for SFARI comparison.
# E2 / P4: sex-combined Sex model (run_voom_lm_sex); no sex filter needed.
TT_results <- read_csv(
  "../../results/targeted/testosterone_genes/run_voom_lm_gw/TT_gw_results.csv"
)
E2_results <- read_csv(
  "../../results/targeted/estradiol_genes/run_voom_lm_sex/E2_sex_results.csv"
)
P4_results <- read_csv(
  "../../results/targeted/progesterone_genes/run_voom_lm_sex/P4_sex_results.csv"
)

TT_sig <- TT_results %>%
  dplyr::filter(sex == "M", adj.P.Val < 0.10, abs(logFC) >= 0.25)

E2_sig <- E2_results %>%
  dplyr::filter(adj.P.Val < 0.10, abs(logFC) >= 0.25)

P4_sig <- P4_results %>%
  dplyr::filter(adj.P.Val < 0.10, abs(logFC) >= 0.25)

cat("TT sig (male, FDR < 0.10):", nrow(TT_sig), "gene-cell_type rows\n")
cat("E2 sig (combined, FDR < 0.10):", nrow(E2_sig), "gene-cell_type rows\n")
cat("P4 sig (combined, FDR < 0.10):", nrow(P4_sig), "gene-cell_type rows\n")
```

## SFARI Intersection — Tables

High-confidence SFARI genes (score ≤ 2) overlapping with significant DEGs.

::: panel-tabset
### Testosterone (Male)

```{r}
#| label: tt-sfari-table
TT_SFARI_shared <- TT_sig %>%
  dplyr::filter(gene %in% SFARI_Gene_filt$`gene-symbol`) %>%
  dplyr::left_join(
    SFARI_Gene_filt %>% dplyr::select(`gene-symbol`, `gene-score`),
    by = c("gene" = "gene-symbol")
  ) %>%
  dplyr::arrange(`gene-score`, adj.P.Val)

cat("TT × SFARI overlap (unique genes):", length(unique(TT_SFARI_shared$gene)), "\n")

TT_SFARI_shared %>%
  dplyr::select(gene, `gene-score`, cell_type, logFC, adj.P.Val) %>%
  knitr::kable(digits = 4, caption = "TT — SFARI-overlapping hits (male, FDR < 0.10)")
```

### Estradiol

```{r}
#| label: e2-sfari-table
E2_SFARI_shared <- E2_sig %>%
  dplyr::filter(gene %in% SFARI_Gene_filt$`gene-symbol`) %>%
  dplyr::left_join(
    SFARI_Gene_filt %>% dplyr::select(`gene-symbol`, `gene-score`),
    by = c("gene" = "gene-symbol")
  ) %>%
  dplyr::arrange(`gene-score`, adj.P.Val)

cat("E2 × SFARI overlap (unique genes):", length(unique(E2_SFARI_shared$gene)), "\n")

E2_SFARI_shared %>%
  dplyr::select(gene, `gene-score`, cell_type, logFC, adj.P.Val) %>%
  knitr::kable(digits = 4, caption = "E2 — SFARI-overlapping hits (combined, FDR < 0.10)")
```

### Progesterone

```{r}
#| label: p4-sfari-table
P4_SFARI_shared <- P4_sig %>%
  dplyr::filter(gene %in% SFARI_Gene_filt$`gene-symbol`) %>%
  dplyr::left_join(
    SFARI_Gene_filt %>% dplyr::select(`gene-symbol`, `gene-score`),
    by = c("gene" = "gene-symbol")
  ) %>%
  dplyr::arrange(`gene-score`, adj.P.Val)

cat("P4 × SFARI overlap (unique genes):", length(unique(P4_SFARI_shared$gene)), "\n")

P4_SFARI_shared %>%
  dplyr::select(gene, `gene-score`, cell_type, logFC, adj.P.Val) %>%
  knitr::kable(digits = 4, caption = "P4 — SFARI-overlapping hits (combined, FDR < 0.10)")
```
:::

## SFARI Summary — Helper Functions

Local functions for SFARI-score-stratified bar charts. Named distinctly to
avoid masking `summarize_sig_genes()` and `plot_sig_gene_summary()` from
`pseudobulk_functions.R`.

```{r}
#| label: sfari-helpers
summarize_sfari_by_score <- function(results, SFARI_Gene,
                                     fdr_cutoff   = 0.10,
                                     logfc_cutoff = 0.25) {
  results %>%
    dplyr::filter(adj.P.Val < fdr_cutoff, abs(logFC) >= logfc_cutoff) %>%
    dplyr::left_join(
      SFARI_Gene %>% dplyr::select(`gene-symbol`, `gene-score`),
      by = c("gene" = "gene-symbol")
    ) %>%
    dplyr::mutate(
      score_group = dplyr::if_else(
        is.na(`gene-score`), "Not SFARI", as.character(`gene-score`)
      ),
      score_group = factor(score_group, levels = c("1", "2", "3", "4", "Not SFARI"))
    ) %>%
    dplyr::count(cell_type, score_group, name = "num_sig_genes")
}

plot_sfari_summary <- function(summary_df, title) {
  ggplot(summary_df, aes(x = cell_type, y = num_sig_genes, fill = score_group)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_text(
      aes(label = num_sig_genes),
      position = position_stack(vjust = 0.5),
      color = "white", size = 3.5
    ) +
    labs(title = title, x = "Cell Type",
         y = "# Significant Genes", fill = "SFARI Gene Score") +
    theme_minimal(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c(
      "1"        = "#bdd7e7",
      "2"        = "#6baed6",
      "3"        = "#3182bd",
      "4"        = "#08519c",
      "Not SFARI" = "gray80"
    ))
}
```

## SFARI Summary — Charts

Stacked bars show all significant genes by cell type, coloured by SFARI gene
score. "Not SFARI" = gene not in the SFARI database.

::: panel-tabset
### Testosterone (Male)

```{r}
#| label: tt-sfari-plot
#| fig-width: 12
#| fig-height: 6
summary_tt <- summarize_sfari_by_score(
  TT_results %>% dplyr::filter(sex == "M"), SFARI_Gene
)

p_tt <- plot_sfari_summary(
  summary_tt, "TT — Significant Genes by Cell Type (male, ~ TT + GW)"
)

p_tt

dir.create("../../results/targeted/SFARI_testosterone_plots",
           showWarnings = FALSE, recursive = TRUE)
ggsave("../../results/targeted/SFARI_testosterone_plots/combined_SFARI_score_summary_male.png",
       p_tt, width = 12, height = 6, dpi = 300)
```

### Estradiol

```{r}
#| label: e2-sfari-plot
#| fig-width: 12
#| fig-height: 6
summary_e2 <- summarize_sfari_by_score(E2_results, SFARI_Gene)

p_e2 <- plot_sfari_summary(
  summary_e2, "E2 — Significant Genes by Cell Type (combined, ~ E2 + Sex)"
)

p_e2

dir.create("../../results/targeted/SFARI_estradiol_plots",
           showWarnings = FALSE, recursive = TRUE)
ggsave("../../results/targeted/SFARI_estradiol_plots/combined_SFARI_score_summary.png",
       p_e2, width = 12, height = 6, dpi = 300)
```

### Progesterone

```{r}
#| label: p4-sfari-plot
#| fig-width: 12
#| fig-height: 6
summary_p4 <- summarize_sfari_by_score(P4_results, SFARI_Gene)

p_p4 <- plot_sfari_summary(
  summary_p4, "P4 — Significant Genes by Cell Type (combined, ~ P4 + Sex)"
)

p_p4

dir.create("../../results/targeted/SFARI_progesterone_plots",
           showWarnings = FALSE, recursive = TRUE)
ggsave("../../results/targeted/SFARI_progesterone_plots/combined_SFARI_score_summary.png",
       p_p4, width = 12, height = 6, dpi = 300)
```
:::

## External Androgen-Response Gene Lists

Comparing TT male significant DEGs against two published human/mouse androgen
response gene lists. Mouse stem-cell symbols are pre-converted to human
(see `converted_sex_tt_mouse.csv`); single-cell mouse data requires biomaRt
conversion in the next section.

```{r}
#| label: external-overlap
androgen_response_all  <- c(androgen_response_up$X1, androgen_response_down$X1)
mouse_response_genes   <- mouse_response$symbol_human

# Raznahan 2021 organoid paper
ar_shared <- intersect(androgen_response_all, TT_sig$gene)
cat("TT × organoid androgen response:", length(ar_shared), "genes\n")
print(ar_shared)

# Yoon 2016 mouse neural stem cells (pre-converted)
mouse_shared <- intersect(mouse_response_genes, TT_sig$gene)
cat("\nTT × mouse neural stem cells:", length(mouse_shared), "genes\n")
print(mouse_shared)
```

## Mouse Gene Conversion (Li 2024)

Converts Li et al. 2024 mouse gene symbols to human orthologs via the Ensembl
compara homology table. **Note:** loads biomaRt and AnnotationDbi, which mask
several dplyr/tidyverse functions — run after all plotting code.

```{r}
#| label: mouse-conversion
#| cache: true
library(data.table)
library(biomaRt)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(org.Hs.eg.db)

singlecell_needs_converting <- singlecell_response$Gene_symbol

url <- "https://ftp.ensembl.org/pub/current_tsv/ensembl-compara/homologies/mus_musculus/Compara.114.protein_default.homologies.tsv.gz"
orthologs <- as.data.frame(fread(url))

# Filter 1:1 human orthologs
mouse2human_1to1 <- orthologs %>%
  dplyr::filter(homology_species == "homo_sapiens",
                homology_type    == "ortholog_one2one")

# Map Ensembl IDs → gene symbols
mouse_symbols <- mapIds(
  org.Mm.eg.db,
  keys     = mouse2human_1to1$gene_stable_id,
  column   = "SYMBOL",
  keytype  = "ENSEMBL",
  multiVals = "first"
)
human_symbols <- mapIds(
  org.Hs.eg.db,
  keys     = mouse2human_1to1$homology_gene_stable_id,
  column   = "SYMBOL",
  keytype  = "ENSEMBL",
  multiVals = "first"
)

mouse2human_1to1$mouse_symbol <- mouse_symbols
mouse2human_1to1$human_symbol <- human_symbols

converted <- mouse2human_1to1 %>%
  dplyr::filter(mouse_symbol %in% singlecell_needs_converting)

converted_human_genes <- unique(converted$human_symbol)
cat("Mouse single-cell genes converted to human:", length(converted_human_genes), "\n")
```

## Li 2024 Overlap

```{r}
#| label: singlecell-overlap
sc_shared <- intersect(converted_human_genes, TT_sig$gene)
cat("TT × Li 2024 single-cell mouse (human orthologs):", length(sc_shared), "genes\n")
print(sc_shared)
```

## Summary

```{r}
#| label: summary-table
tibble::tibble(
  `Gene List`   = c(
    "SFARI (score ≤ 2)",
    "Organoid androgen response (Raznahan 2021)",
    "Mouse neural stem cells (Yoon 2016)",
    "Single-cell mouse atlas (Li 2024)"
  ),
  `TT × List`   = c(
    length(unique(TT_SFARI_shared$gene)),
    length(ar_shared),
    length(mouse_shared),
    length(sc_shared)
  )
) %>%
  knitr::kable(caption = "TT male DEG overlap with external gene lists")
```
