---
title: "Metabolite–Gene Associations"
author: "Sneha M. Chaturvedi"
date: "Last updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

Threshold and annotate metabolite-gene association results from the limma-voom
loop. Apply global FDR correction across all metabolite x cell-type tests,
generate summary figures, and scatter plots for top hits.

| | |
|---|---|
| **Inputs** | `results/gene-metabolite/parquet-log2_na/` (from `metabolite_gene_limma.R`) |
| **Outputs** | CSVs in `results/gene-metabolite/csv-log2_na/`, figures in `results/gene-metabolite/figures-log2_na/` |
| **Upstream** | `metabolite_gene_limma.R` |
| **Downstream** | `metabolite_gene_imputation_sensitivity.qmd`, pathway QMDs |


## Overview

Reads the parquet archive written by `metabolite_gene_limma.R` and does all
thresholding, global-FDR, annotation, and figures here — so screen changes never
re-run the loop.

- **Model (per metabolite):** gene counts ~ `metab + GW + Sex + batch1_frac`, split by cell type; voom.
- **Interpretation:** `logFC` = Δ(gene log2-CPM) per **1 SD** of the standardised metabolite. These are **associations**, not directional predictions.
- **Primary screen:** within-(metabolite × cell type) `adj.P.Val < 0.10` **and** `|logFC| ≥ 0.25`.
- **Global FDR:** BH across every metabolite × cell type × gene test, carried as a reference column.

## Set Up

```{r load-packages}
Packages <- c("tidyverse", "arrow", "cowplot", "svglite", "UpSetR", "edgeR", "patchwork")
lapply(Packages, library, character.only = TRUE)

find_project_root <- function(marker = ".git") {
  d <- normalizePath(getwd())
  repeat {
    if (dir.exists(file.path(d, marker))) return(d)
    parent <- dirname(d)
    if (parent == d) stop("Project root not found (no ", marker, " above ", getwd(), ")")
    d <- parent
  }
}
root <- find_project_root()
source(file.path(root, "src/gene-hormone/pseudobulk_functions.R"))     # save_dual_format(), merge_hormone_metadata()
source(file.path(root, "src/gene-metabolite/metabolite_functions.R"))     # build_metab_covariates(), INT transform, mapping
set.seed(123)
```

```{r load-model-objects}
# METHOD selects which transform run to analyze; it sets the parquet read below
# AND the reporter's predictor transform, so they can never mismatch.
METHOD <- "log2_na"                             
suffix <- if (METHOD == "int") "" else paste0("-", METHOD)
parquet_dir <- if (METHOD == "int") {file.path(root, "results/gene-metabolite/parquet")} else {sprintf(file.path(root, "results/gene-metabolite/parquet-%s"), METHOD)}

# Rebuild the modeled objects (needed only for the LOO / Cook's reporter below).
cache_dir    <- file.path(root, "data/cache")
metab_path   <- file.path(root, "results/untargeted/peak_area_clean.csv")
hormone_xlsx <- file.path(root, "doc/targeted/targeted_hormones.xlsx")

pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))
targeted_comparison_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`

pb       <- build_metab_covariates(pb_base, targeted_comparison_raw, batch_props,
                                   min_cells = 30)
metab_mat <- read_metabolite_matrix(metab_path)
metab_tx  <- transform_metabolite_matrix(metab_mat, method = METHOD)

# Raw log2 (no imputation — panel 1 x-axis)
metab_log <- log2(metab_mat + 1)

# Log2 with IQR outliers flagged NA (panels 2–3 x-axis)
.log2_trim_only <- function(x) {
  logged  <- log2(x + 1)
  orig_na <- is.na(logged)
  obs     <- logged[!orig_na]
  if (length(obs) < 2) return(rep(NA_real_, length(x)))
  qs  <- quantile(obs, c(0.25, 0.75), names = FALSE)
  iqr <- qs[2] - qs[1]
  v   <- logged
  v[!orig_na & (logged < qs[1] - 0.75 * iqr | logged > qs[2] + 0.75 * iqr)] <- NA
  v
}
metab_log_trim <- t(apply(metab_mat, 1, .log2_trim_only))
dimnames(metab_log_trim) <- dimnames(metab_mat)

rn        <- rownames(pb$counts)
row_ct    <- sub("^[^.]+\\.(.*)$", "\\1", rn)
metab_col <- metab_sample_to_pb(colnames(metab_tx), sub("^([^.]+)\\..*$", "\\1", rn))
```

```{r config}
# parquet_dir + suffix come from the load-model-objects chunk (METHOD switch).
annot_path   <- file.path(root, "results/untargeted/metabolite_annotations_flagged.csv")
out_dir      <- file.path(root, "results/gene-metabolite")
svg_dir      <- file.path(out_dir, paste0("svg", suffix))   # method-keyed: no clobber
csv_dir      <- file.path(out_dir, paste0("csv", suffix))
for (d in c(out_dir, svg_dir, csv_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

FDR_WITHIN <- 0.10   # primary screen (within metabolite x cell type)
LFC_MIN    <- 0.25   # per-SD effect-size floor
FDR_GLOBAL <- 0.10   # reference global BH threshold
LABEL_CHARS <- 30    # metabolite-name clip length for plot labels

# Clip long metabolite names for plotting only (data keeps full Name). Names
# longer than `n` get an ellipsis; if two clips collide, the 2nd+ get "#2","#3"
# so bars stay distinct. Returns the input tibble with a `label` column added;
# write the paired label<->Name lookup beside each figure's SVG.
add_plot_label <- function(df, name_col = "Name", n = LABEL_CHARS) {
  short <- ifelse(nchar(df[[name_col]]) > n,
                  paste0(substr(df[[name_col]], 1, n), "\u2026"),
                  df[[name_col]])
  idx   <- stats::ave(seq_along(short), short, FUN = seq_along)   # 1,2,3 within group
  short[idx > 1] <- paste0(short[idx > 1], "#", idx[idx > 1])
  df$label <- short
  df
}
```

## Open the archive (lazy)

`open_dataset()` reads only the columns/rows a query touches, so the multi-GB
archive is never loaded whole.

```{r open-dataset}
ds <- open_dataset(parquet_dir)

n_tests    <- nrow(ds)                                   # total tests (cheap)
n_metab    <- ds |> distinct(Compound.ID) |> collect() |> nrow()
n_celltype <- ds |> distinct(cell_type)  |> collect() |> nrow()

cat(sprintf("Total tests: %s | metabolites: %d | cell types: %d\n",
            format(n_tests, big.mark = ","), n_metab, n_celltype))
```

## Global BH (memory-light)

Any test with global FDR < 0.10 must have a small raw p-value, so we only pull
candidates (`P.Value < 0.05`) and rank them against the full test count `m`.
This reproduces the exact BH adjustment those rows would receive over all
`n_tests` without collecting the whole archive.

```{r global-bh}
bh_global <- function(p, m) {
  o   <- order(p)
  ro  <- order(o)
  adj <- p[o] * m / seq_along(p[o])
  adj <- rev(cummin(rev(adj)))
  pmin(adj[ro], 1)
}

candidates <- ds |>
  filter(P.Value < 0.05) |>
  collect() |>
  mutate(FDR_global = bh_global(P.Value, n_tests))

cat(sprintf("Candidate rows (p < 0.05): %s\n",
            format(nrow(candidates), big.mark = ",")))
```

## Global-FDR hits + stability reporter

```{r annotate-and-filter}
annot <- read_csv(annot_path, show_col_types = FALSE) |>
  dplyr::select(Compound.ID, Super.Class, Class, Sub.Class, Pathways, KEGG, Exogenous_Flag, Annotation_Confidence)

# Primary screen = global FDR (BH across all tests) & |logFC|. Within-FDR
# (adj.P.Val) is retained as a carried column.
hits <- candidates |>
  filter(FDR_global < FDR_GLOBAL, abs(logFC) >= LFC_MIN) |>
  left_join(annot, by = "Compound.ID") |>
  arrange(cell_type, Compound.ID, FDR_global)

cat(sprintf("Global-FDR hits (FDR < %.2f & |logFC| >= %.2f): %s\n",
            FDR_GLOBAL, LFC_MIN, format(nrow(hits), big.mark = ",")))

# Reviewer/co-author question: how many associations survive a stricter global
# FDR? Reported only -- `hits` and everything downstream stay at FDR_GLOBAL.
# |logFC| is held at LFC_MIN so FDR is the only variable that moves. The
# candidate prefilter (p < 0.05) is a superset of any FDR < 0.10 cut, so
# tightening the threshold needs no rerun.
for (f in c(0.05, 0.01)) {
  h <- candidates |> filter(FDR_global < f, abs(logFC) >= LFC_MIN)
  cat(sprintf(
    "  FDR < %.2f: %s associations | %s metabolites | %s genes | %s cell types\n",
    f,
    format(nrow(h), big.mark = ","),
    format(dplyr::n_distinct(h$Compound.ID), big.mark = ","),
    format(dplyr::n_distinct(h$gene), big.mark = ","),
    dplyr::n_distinct(h$cell_type)))
}
```

```{r stability-reporter}
# Per hit, REPORTER columns (do not gate significance -- filter on them later):
#   n_nonzero    modeled samples with raw count >= 2
#   max_cooks    max Cook's distance in the lm(log1p(CPM) ~ metab + covariates)
#   loo_unstable dropping the most influential sample flips the metabolite slope or halves |slope| -> small-n fragility 
report_group <- function(cid, ct, genes) {
  idx      <- which(row_ct == ct)
  mv       <- metab_tx[cid, match(metab_col[idx], colnames(metab_tx))]
  keep     <- !is.na(mv)
  rows_use <- idx[keep]
  md       <- pb$metadata[rn[rows_use], ]
  lib      <- rowSums(pb$counts[rows_use, , drop = FALSE])     # library size / sample
  base     <- tibble(metab = as.numeric(mv[keep]),
                     GW = md$GW, Sex = md$Sex, batch1_frac = md$batch1_frac)
  covs  <- c("GW", "Sex", "batch1_frac")
  keepc <- covs[vapply(covs, function(cv) length(unique(base[[cv]])) >= 2, logical(1))]
  fml   <- reformulate(c("metab", keepc), response = "expr")

  purrr::map_dfr(genes, function(g) {
    if (!g %in% colnames(pb$counts))
      return(tibble(gene = g, n_nonzero = NA_integer_,
                    max_cooks = NA_real_, loo_unstable = NA))
    cnt <- pb$counts[rows_use, g]
    df  <- base; df$expr <- log1p(cnt / lib * 1e6)             # log1p(CPM)
    if (length(unique(df$metab)) < 2 || nrow(df) < 4)
      return(tibble(gene = g, n_nonzero = sum(cnt >= 2),
                    max_cooks = NA_real_, loo_unstable = NA))
    fit <- lm(fml, data = df)
    cd  <- cooks.distance(fit); b <- coef(fit)[["metab"]]
    b2  <- coef(lm(fml, data = df[-which.max(cd), ]))[["metab"]]
    tibble(gene = g, n_nonzero = sum(cnt >= 2), max_cooks = max(cd),
           loo_unstable = sign(b2) != sign(b) || abs(b2) < 0.5 * abs(b))
  })
}

reporter <- hits |>
  group_by(Compound.ID, cell_type) |>
  group_modify(~ report_group(.y$Compound.ID, .y$cell_type, .x$gene)) |>
  ungroup()

hits <- hits |> left_join(reporter, by = c("Compound.ID", "cell_type", "gene"))
write_csv(hits, file.path(csv_dir, "metabolite_gene_hits.csv"))

cat(sprintf("Reporter added. loo_unstable: %s of %s hits (%.1f%%) flagged fragile.\n",
            format(sum(hits$loo_unstable, na.rm = TRUE), big.mark = ","),
            format(nrow(hits), big.mark = ","),
            100 * mean(hits$loo_unstable, na.rm = TRUE)))
```

## Recurrence — genes across metabolites

Genes that track the most distinct metabolites (a gene appearing broadly may be a hub of metabolic coupling, or a covariate-driven artifact worth inspecting).

```{r gene-recurrence}
gene_recur <- hits |>
  group_by(gene) |>
  summarise(n_metabolites = n_distinct(Compound.ID),
            n_cell_types  = n_distinct(cell_type),
            n_tests       = n(), .groups = "drop") |>
  arrange(desc(n_metabolites))

top_genes <- gene_recur |> slice_head(n = 25)

p_gene <- ggplot(top_genes,
                 aes(x = reorder(gene, n_metabolites), y = n_metabolites)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  coord_flip() +
  labs(x = "Gene", y = "Distinct metabolites associated",
       title = "Top recurrent genes across metabolites") +
  theme_cowplot(12)

save_dual_format(p_gene, out_dir, "recurrence_genes", width = 8, height = 7)
print(p_gene)
```

## Recurrence — metabolites across genes

```{r metabolite-recurrence}
metab_recur <- hits |>
  group_by(Compound.ID, Name, Super.Class, Class) |>
  summarise(n_genes      = n_distinct(gene),
            n_cell_types = n_distinct(cell_type), .groups = "drop") |>
  arrange(desc(n_genes))

top_metab <- metab_recur |> slice_head(n = 25) |> add_plot_label()

# Full Name <-> clipped label lookup, paired with the SVG.
write_csv(top_metab |> dplyr::select(label, Name, Compound.ID, Super.Class, Class),
          file.path(svg_dir, "recurrence_metabolites_labels.csv"))

p_metab <- ggplot(top_metab,
                  aes(x = reorder(label, n_genes), y = n_genes, fill = Super.Class)) +
  geom_col(alpha = 0.85) +
  coord_flip() +
  labs(x = "Metabolite", y = "Distinct genes associated",
       title = "Top recurrent metabolites across genes") +
  theme_cowplot(12) +
  theme(legend.position = "right")

save_dual_format(p_metab, out_dir, "recurrence_metabolites", width = 10, height = 7)
print(p_metab)
```

## Recurrence — cell types

```{r celltype-recurrence}
ct_recur <- hits |>
  group_by(cell_type) |>
  summarise(n_associations = n(),
            n_metabolites  = n_distinct(Compound.ID),
            n_genes        = n_distinct(gene), .groups = "drop") |>
  arrange(desc(n_associations))

p_ct <- ggplot(ct_recur,
               aes(x = reorder(cell_type, n_metabolites), y = n_metabolites)) +
  geom_col(fill = "darkorange", alpha = 0.85) +
  coord_flip() +
  labs(x = "Cell type", y = "Distinct metabolites associated",
       title = "Distinct metabolites by cell type") +
  theme_cowplot(12)

save_dual_format(p_ct, out_dir, "recurrence_celltypes", width = 8, height = 6)
print(p_ct)

knitr::kable(ct_recur, caption = "Hit counts per cell type")
```

## Sharing across cell types (UpSet)

Which findings are shared vs. cell-type-specific. Three panels, because the
answer differs by unit: **genes** (does a gene associate with any metabolite in
more than one cell type), **associations** (does the same gene-metabolite pair
recur across cell types — the strictest test), and **metabolites** (does a
metabolite associate with any gene in more than one cell type).

```{r upset-sharing, fig.width=10, fig.height=6}
# UpSetR >= 1.4.0: upset() RETURNS an object of class `upset`; all drawing lives
# in print.upset(). So it cannot be passed to ggsave()/save_dual_format(), and it
# must be print()-ed explicitly — inside eval() there is no auto-printing, which
# silently yields a blank file. `expr` is passed unevaluated and forced (and
# printed) once per device.
save_dual_format_base <- function(expr, output_dir, filename_base,
                                  width = 10, height = 6, dpi = 300) {
  expr     <- substitute(expr)
  penv     <- parent.frame()
  svg_out  <- file.path(output_dir, "svg")
  dir.create(svg_out, recursive = TRUE, showWarnings = FALSE)

  png(file.path(output_dir, paste0(filename_base, ".png")),
      width = width, height = height, units = "in", res = dpi)
  print(eval(expr, envir = penv)); dev.off()   # print() is load-bearing

  svglite::svglite(file.path(svg_out, paste0(filename_base, ".svg")),
                   width = width, height = height)
  print(eval(expr, envir = penv)); dev.off()

  message("Saved: ", filename_base, ".png  +  svg/", filename_base, ".svg")
}

# One UpSet per unit of analysis. `id_cols` defines the element being counted;
# cell types are always the sets. Set sizes (left bars) = distinct elements per
# cell type, so the three panels are NOT rescalings of each other.
upset_by_celltype <- function(df, id_cols, filename_base, label) {
  wide <- df |>
    distinct(dplyr::across(dplyr::all_of(c(id_cols, "cell_type")))) |>
    tidyr::unite("element", dplyr::all_of(id_cols), sep = " ~ ", remove = TRUE) |>
    mutate(present = 1L) |>
    pivot_wider(names_from = cell_type, values_from = present, values_fill = 0L)

  ct_cols <- setdiff(colnames(wide), "element")

  n_shared <- sum(rowSums(wide[, ct_cols, drop = FALSE]) > 1)
  cat(sprintf("%s: %s total, %s in >1 cell type (%.1f%%)\n",
              label, format(nrow(wide), big.mark = ","),
              format(n_shared, big.mark = ","),
              100 * n_shared / max(nrow(wide), 1)))

  if (length(ct_cols) < 2 || nrow(wide) == 0) {
    cat("  Not enough cell types / elements for an UpSet plot.\n")
    return(invisible(NULL))
  }

  upset_call <- quote(
    UpSetR::upset(
      as.data.frame(wide), sets = ct_cols,
      nsets = length(ct_cols), nintersects = NA,   # keep all non-empty intersections
      order.by = "freq", keep.order = TRUE,
      text.scale = 1.3
    )
  )
  save_dual_format_base(eval(upset_call), out_dir, filename_base,
                        width = 10, height = 6)
  print(eval(upset_call))   # render inline in the rendered document
  invisible(wide)
}

upset_by_celltype(hits, "gene",
                  "upset_genes_by_celltype",        "Genes")
upset_by_celltype(hits, c("gene", "Compound.ID"),
                  "upset_associations_by_celltype", "Associations (gene x metabolite)")
upset_by_celltype(hits, "Compound.ID",
                  "upset_metabolites_by_celltype",  "Metabolites")
```

## Association sharing across neuronal lineages

Cell-type-level results are sparse by construction, so a gene-metabolite pair
recurring in two related cell types (e.g. IN-MGE-Immature and IN-MGE-SST) is
more informative than one appearing in a single type. Here we ask two things:
whether pairs recur *within* a lineage, and whether any pair is shared *between*
excitatory neurons, inhibitory neurons, and glia. Sharing is evaluated on the
gene x metabolite pair, which is stricter than the gene-level UpSet above.

```{r lineage-sharing}
# Lineage assignment. EN-*/IN-* are resolved by prefix (covers EN-Newborn,
# IN-Mix-LAMP5, IN-dLGE-Immature etc.); everything else is mapped explicitly
# below. Only cell types present in `hits` are used, so entries for types that
# fail filtering are simply inert.
#   - Progenitors (RG-*, IPC-*) are kept separate from mature glia: they are
#     neurogenic, and folding them in would blur an "excitatory vs glia" claim.
#     IPC-EN is an excitatory-neuron progenitor, but it is still a progenitor —
#     move it to "Excitatory" only if the claim is about lineage, not state.
#   - Microglia is grouped with Glia by convention, though it is myeloid-derived
#     rather than neuroectodermal. Split it out if a reviewer objects.
#   - Cajal-Retzius cells are glutamatergic but arise from the cortical hem, so
#     they are held separate rather than counted as excitatory.
#   - "Unknown" is its own category so it can never silently prop up a
#     cross-lineage overlap claim.
NONNEURONAL_MAP <- c(
  "Astrocyte-Immature"       = "Glia",
  "Astrocyte-Fibrous"        = "Glia",
  "Oligodendrocyte-Immature" = "Glia",
  "OPC"                      = "Glia",
  "Microglia"                = "Glia",
  "RG-oRG"                   = "Progenitor",
  "RG-tRG"                   = "Progenitor",
  "RG-vRG"                   = "Progenitor",
  "IPC-EN"                   = "Progenitor",
  "IPC-Glia"                 = "Progenitor",
  "Cajal-Retzius cell"       = "Cajal-Retzius",
  "Vascular"                 = "Other",
  "Unknown"                  = "Unknown"
)

assign_lineage <- function(ct) {
  out <- dplyr::case_when(
    grepl("^EN", ct) ~ "Excitatory",
    grepl("^IN", ct) ~ "Inhibitory",
    TRUE             ~ unname(NONNEURONAL_MAP[ct])
  )
  out[is.na(out)] <- "Unmapped"
  out
}

hits_lin <- hits |>
  mutate(lineage = assign_lineage(cell_type))

# Surface anything the map missed so the grouping is never silently wrong.
unmapped <- hits_lin |> filter(lineage == "Unmapped") |> distinct(cell_type) |> pull()
if (length(unmapped)) {
  warning("Unmapped cell types (add to NONNEURONAL_MAP): ",
          paste(unmapped, collapse = ", "))
}

cat("Cell types per lineage:\n")
print(hits_lin |> distinct(lineage, cell_type) |> count(lineage, name = "n_cell_types"))

# ---- Within-lineage recurrence: same pair in >1 cell type of the same lineage
pair_within <- hits_lin |>
  distinct(lineage, Compound.ID, gene, cell_type) |>
  count(lineage, Compound.ID, gene, name = "n_cell_types")

cat("\nGene-metabolite pairs recurring in >1 cell type of the same lineage:\n")
print(pair_within |> filter(n_cell_types > 1) |> count(lineage, name = "n_pairs"))

within_tbl <- pair_within |>
  filter(n_cell_types > 1) |>
  left_join(hits_lin |> distinct(Compound.ID, dplyr::across(dplyr::any_of(c("Name", "Super.Class", "Class")))),
            by = "Compound.ID") |>
  left_join(hits_lin |>
              distinct(lineage, Compound.ID, gene, cell_type) |>
              group_by(lineage, Compound.ID, gene) |>
              summarise(cell_types = paste(sort(cell_type), collapse = ", "), .groups = "drop"),
            by = c("lineage", "Compound.ID", "gene")) |>
  arrange(lineage, desc(n_cell_types), gene)

# Print the pairs themselves — these are what get quoted in the text.
if (nrow(within_tbl)) {
  cat("\nWithin-lineage recurrent pairs:\n")
  print(as.data.frame(within_tbl |>
          dplyr::select(dplyr::any_of(c("lineage", "gene", "Name", "Compound.ID",
                                        "Class", "n_cell_types", "cell_types")))),
        right = FALSE)
} else {
  cat("\nNo gene-metabolite pair recurred in >1 cell type of the same lineage.\n")
}

# ---- Between-lineage sharing: same pair in >1 lineage
pair_between <- hits_lin |>
  distinct(Compound.ID, gene, lineage) |>
  group_by(Compound.ID, gene) |>
  summarise(n_lineages = dplyr::n(),
            lineages   = paste(sort(unique(lineage)), collapse = " + "),
            .groups = "drop")

cat(sprintf("\nPairs in >1 lineage: %s of %s total pairs\n",
            format(sum(pair_between$n_lineages > 1), big.mark = ","),
            format(nrow(pair_between), big.mark = ",")))

between_tbl <- pair_between |>
  filter(n_lineages > 1) |>
  left_join(hits_lin |> distinct(Compound.ID, dplyr::across(dplyr::any_of(c("Name", "Super.Class", "Class")))),
            by = "Compound.ID") |>
  left_join(hits_lin |>
              distinct(Compound.ID, gene, cell_type) |>
              group_by(Compound.ID, gene) |>
              summarise(cell_types = paste(sort(cell_type), collapse = ", "), .groups = "drop"),
            by = c("Compound.ID", "gene")) |>
  arrange(desc(n_lineages), lineages, gene)

if (nrow(between_tbl)) {
  cat("\nCross-lineage pairs:\n")
  print(as.data.frame(between_tbl |>
          dplyr::select(dplyr::any_of(c("gene", "Name", "Compound.ID", "Class",
                                        "n_lineages", "lineages", "cell_types")))),
        right = FALSE)
} else {
  cat("\nNo gene-metabolite pair was shared across lineages.\n")
}

# ---- Gene-level sharing between lineages, for comparison with the UpSet above
gene_between <- hits_lin |>
  distinct(gene, lineage) |>
  count(gene, name = "n_lineages")

cat(sprintf("\nGenes appearing in >1 lineage: %s of %s hit genes\n",
            format(sum(gene_between$n_lineages > 1), big.mark = ","),
            format(nrow(gene_between), big.mark = ",")))

write_csv(within_tbl,  file.path(csv_dir, "lineage_pairs_within.csv"))
write_csv(between_tbl, file.path(csv_dir, "lineage_pairs_between.csv"))
cat("\nWrote lineage_pairs_within.csv and lineage_pairs_between.csv\n")
```

## Save recurrence tables

```{r save-tables}
write_csv(gene_recur,  file.path(csv_dir, "recurrence_genes.csv"))
write_csv(metab_recur, file.path(csv_dir, "recurrence_metabolites.csv"))
write_csv(ct_recur,    file.path(csv_dir, "recurrence_celltypes.csv"))
cat("Recurrence tables written to", csv_dir, "\n")
```

## Top-20 global-FDR scatters

Five-panel scatter: raw log2 (all samples) | outlier-trimmed log2 |
covariate-adjusted log2 || outlier-trimmed z-score | covariate-adjusted z-score.
Log2 panels use free x-axis per metabolite; z-score panels share a fixed axis
across all plots for cross-metabolite comparison. One pooled regression line per
panel; points colored by GW and shaped by Sex.

```{r scatter-builder}
scatter_dir <- file.path(csv_dir, "scatter-plots")
dir.create(scatter_dir, recursive = TRUE, showWarnings = FALSE)

make_pair_plot <- function(cid, gene, ct, meta_name, z_x_lim) {
  idx      <- which(row_ct == ct)
  pos      <- match(metab_col[idx], colnames(metab_tx))
  present  <- !is.na(pos) & gene %in% colnames(pb$counts)
  rows_use <- idx[present]; pos_p <- pos[present]
  if (length(rows_use) < 3) return(NULL)

  expr <- as.numeric(log1p(edgeR::cpm(t(pb$counts[rows_use, , drop = FALSE]), log = FALSE)[gene, ]))
  md   <- pb$metadata[rn[rows_use], ]

  df <- tibble(
    metab_raw     = as.numeric(metab_log[cid, pos_p]),
    metab_logtrim = as.numeric(metab_log_trim[cid, pos_p]),
    metab_trim    = as.numeric(metab_tx[cid, pos_p]),
    expr          = expr,
    Sex           = md$Sex,
    GW            = md$GW,
    batch1_frac   = md$batch1_frac
  )

  covs <- c("GW", "Sex", "batch1_frac")

  # ---- log2 panels (free x-axis per metabolite) ----
  x_range <- range(df$metab_raw, na.rm = TRUE)
  x_pad   <- diff(x_range) * 0.05
  x_lim   <- c(x_range[1] - x_pad, x_range[2] + x_pad)
  x_lab   <- paste0(meta_name, " [log2 peak area]")

  bl_log <- list(
    geom_point(aes(shape = Sex), size = 2.5, alpha = 0.85),
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE, linewidth = 0.8),
    scale_color_viridis_c(name = "GW"),
    scale_shape_manual(values = c("M" = 16, "F" = 17)),
    coord_cartesian(xlim = x_lim),
    theme_minimal(base_size = 13)
  )

  df_raw <- df |> filter(!is.na(metab_raw))
  p_raw <- ggplot(df_raw, aes(metab_raw, expr, color = GW)) + bl_log +
    labs(title = "Raw log2 (all samples)", x = x_lab, y = "log1p(CPM)")

  dft <- df |> filter(!is.na(metab_logtrim))
  keepc <- covs[vapply(covs, function(cv) length(unique(dft[[cv]])) >= 2, logical(1))]
  p_mid <- ggplot(dft, aes(metab_logtrim, expr, color = GW)) + bl_log +
    labs(title = "Outlier-trimmed (log2)", x = x_lab, y = "log1p(CPM)")

  fit_log  <- lm(reformulate(c("metab_logtrim", keepc), response = "expr"), data = dft)
  dft$part <- resid(fit_log) + coef(fit_log)[["metab_logtrim"]] * dft$metab_logtrim
  p_adj <- ggplot(dft, aes(metab_logtrim, part, color = GW)) + bl_log +
    labs(title = "Covariate-adjusted (log2)", x = x_lab, y = "log1p(CPM), adjusted")

  # ---- z-score panels (fixed x-axis across all plots) ----
  z_lab <- paste0(meta_name, " [z-score]")

  bl_z <- list(
    geom_point(aes(shape = Sex), size = 2.5, alpha = 0.85),
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE, linewidth = 0.8),
    scale_color_viridis_c(name = "GW"),
    scale_shape_manual(values = c("M" = 16, "F" = 17)),
    coord_cartesian(xlim = z_x_lim),
    theme_minimal(base_size = 13)
  )

  dft_z   <- df |> filter(!is.na(metab_trim))
  keepc_z <- covs[vapply(covs, function(cv) length(unique(dft_z[[cv]])) >= 2, logical(1))]
  p_zmid  <- ggplot(dft_z, aes(metab_trim, expr, color = GW)) + bl_z +
    labs(title = "Outlier-trimmed (z-score)", x = z_lab, y = "log1p(CPM)")

  fit_z        <- lm(reformulate(c("metab_trim", keepc_z), response = "expr"), data = dft_z)
  dft_z$part_z <- resid(fit_z) + coef(fit_z)[["metab_trim"]] * dft_z$metab_trim
  p_zadj <- ggplot(dft_z, aes(metab_trim, part_z, color = GW)) + bl_z +
    labs(title = "Covariate-adjusted (z-score)", x = z_lab, y = "log1p(CPM), adjusted")

  (p_raw | p_mid | p_adj) / (plot_spacer() | p_zmid | p_zadj) +
    plot_annotation(title = paste0(gene, "  ~  ", meta_name, "   (", ct, ")")) +
    plot_layout(guides = "collect")
}
```

```{r scatter-generate, results='asis'}
top20 <- hits |>
  arrange(FDR_global, desc(abs(logFC))) |>
  distinct(Compound.ID, .keep_all = TRUE) |>
  slice_head(n = 20)

# Global z-score x-axis: symmetric range covering 99.8% of all z-scores across
# all metabolites. Fixed once so z-score panels are directly comparable.
z_vals   <- as.vector(metab_tx)
z_abs    <- quantile(abs(z_vals), 0.999, na.rm = TRUE)
z_x_lim  <- c(-z_abs, z_abs)
cat(sprintf("Global z-score x-axis: [%.2f, %.2f]\n", z_x_lim[1], z_x_lim[2]))

for (i in seq_len(nrow(top20))) {
  r  <- top20[i, ]
  pl <- make_pair_plot(r$Compound.ID, r$gene, r$cell_type, r$Name, z_x_lim)
  if (is.null(pl)) next
  base <- gsub("[^A-Za-z0-9]+", "-", paste(r$Compound.ID, r$gene, r$cell_type, sep = "_"))
  save_dual_format(pl, scatter_dir, base, width = 16, height = 10)
  print(pl)
}
cat("\nSaved", nrow(top20), "global-FDR scatters to", scatter_dir, "\n")
```


## Total associations by cell type

```{r associations-bar, fig.width=8, fig.height=6}
p_assoc <- ggplot(ct_recur,
                  aes(x = reorder(cell_type, n_associations), y = n_associations)) +
  geom_col(fill = "#4E79A7", alpha = 0.85) +
  coord_flip() +
  labs(x = NULL, y = "Total associations",
       title = "Gene\u2013metabolite associations by cell type") +
  theme_cowplot(12)

save_dual_format(p_assoc, out_dir, "associations_by_celltype", width = 8, height = 6)
print(p_assoc)
```

## Distinct metabolites by cell type

```{r distinct-metab-bar, fig.width=8, fig.height=6}
p_ct_metab <- ggplot(ct_recur,
                     aes(x = reorder(cell_type, n_metabolites), y = n_metabolites)) +
  geom_col(fill = "#F28E2B", alpha = 0.85) +
  coord_flip() +
  labs(x = NULL, y = "Distinct metabolites",
       title = "Distinct metabolites associated per cell type") +
  theme_cowplot(12)

save_dual_format(p_ct_metab, out_dir, "distinct_metabolites_by_celltype", width = 8, height = 6)
print(p_ct_metab)
```

## Distinct genes by cell type

```{r distinct-genes-bar, fig.width=8, fig.height=6}
p_ct_genes <- ggplot(ct_recur,
                     aes(x = reorder(cell_type, n_genes), y = n_genes)) +
  geom_col(fill = "#59A14F", alpha = 0.85) +
  coord_flip() +
  labs(x = NULL, y = "Distinct genes",
       title = "Distinct genes associated per cell type") +
  theme_cowplot(12)

save_dual_format(p_ct_genes, out_dir, "distinct_genes_by_celltype", width = 8, height = 6)
print(p_ct_genes)
```


## Spotlight scatter — helper function

Simple 1-panel scatter showing the covariate-adjusted partial residual
(the actual model relationship: gene ~ metabolite after adjusting for
GW + Sex + batch1_frac). Points colored by GW, shaped by Sex. One pooled
regression line matching model structure.

```{r spotlight-scatter-fn}
make_spotlight_scatter <- function(cid, gene, ct, meta_name,
                                   label = NULL) {
  idx      <- which(row_ct == ct)
  pos      <- match(metab_col[idx], colnames(metab_tx))
  present  <- !is.na(pos) & gene %in% colnames(pb$counts)
  rows_use <- idx[present]; pos_p <- pos[present]
  if (length(rows_use) < 3) { message("Too few samples for ", gene, " ~ ", meta_name); return(NULL) }

  # Gene expression: log1p(CPM) computed per cell type
  expr <- as.numeric(log1p(
    edgeR::cpm(t(pb$counts[rows_use, , drop = FALSE]), log = FALSE)[gene, ]
  ))
  md <- pb$metadata[rn[rows_use], ]

  df <- tibble(
    metab = as.numeric(metab_tx[cid, pos_p]),
    expr  = expr,
    Sex   = md$Sex,
    GW    = md$GW,
    batch1_frac = md$batch1_frac
  ) |> filter(!is.na(metab))

  if (nrow(df) < 4) { message("Too few non-NA for ", gene, " ~ ", meta_name); return(NULL) }

  # Partial residual: remove covariate effects, keep metabolite slope
  covs   <- c("GW", "Sex", "batch1_frac")
  keepc  <- covs[vapply(covs, function(cv) length(unique(df[[cv]])) >= 2, logical(1))]
  fit    <- lm(reformulate(c("metab", keepc), response = "expr"), data = df)
  df$partial <- resid(fit) + coef(fit)[["metab"]] * df$metab

  # Stats annotation
  hit_row <- hits |> filter(Compound.ID == cid, gene == !!gene, cell_type == ct)
  lfc_txt <- if (nrow(hit_row) > 0) sprintf("logFC = %.2f", hit_row$logFC[1]) else ""
  fdr_txt <- if (nrow(hit_row) > 0) sprintf("FDR = %.3f",   hit_row$FDR_global[1]) else ""
  stat_label <- paste(lfc_txt, fdr_txt, sep = "  |  ")

  if (is.null(label)) label <- paste0(gene, " ~ ", meta_name, "  (", ct, ")")

  p <- ggplot(df, aes(metab, partial, color = GW)) +
    geom_point(aes(shape = Sex), size = 3, alpha = 0.85) +
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE, linewidth = 0.9,
                color = "grey30") +
    scale_color_viridis_c(name = "GW") +
    scale_shape_manual(values = c("M" = 16, "F" = 17)) +
    annotate("text", x = Inf, y = -Inf, hjust = 1.05, vjust = -0.5,
             label = stat_label, size = 3.5, color = "grey40") +
    labs(x = paste0(meta_name, " [", METHOD, " transformed]"),
         y = "Gene expression [log1p(CPM), adjusted]",
         title = label) +
    theme_cowplot(13)

  p
}
```


## OPC lipid spotlight (Figure 5C)

OPC has the most distinct metabolites (n = `r ct_recur |> filter(cell_type == "OPC") |> pull(n_metabolites)`),
with lipids representing the largest metabolite class. Spotlight pairs connect
myelination-associated genes to lipid metabolites.

```{r opc-lipid-spotlight, fig.width=8, fig.height=6}
opc_spotlight_dir <- file.path(csv_dir, "spotlight-opc")
dir.create(opc_spotlight_dir, recursive = TRUE, showWarnings = FALSE)

# ---- BCAS1 ~ Pc(19:5cooh_25:8): myelination marker + phospholipid ----
p_bcas1 <- make_spotlight_scatter(
  cid  = hits |> filter(gene == "BCAS1", cell_type == "OPC",
                         grepl("Glycerophospholipid", Class)) |> pull(Compound.ID) |> first(),
  gene = "BCAS1", ct = "OPC",
  meta_name = hits |> filter(gene == "BCAS1", cell_type == "OPC",
                              grepl("Glycerophospholipid", Class)) |> pull(Name) |> first()
)
if (!is.null(p_bcas1)) {
  save_dual_format(p_bcas1, opc_spotlight_dir, "OPC_BCAS1_phospholipid", width = 8, height = 6)
  print(p_bcas1)
}

# ---- PLCG2 ~ Palmitic Acid: phospholipase + fatty acid substrate ----
p_plcg2 <- make_spotlight_scatter(
  cid  = hits |> filter(gene == "PLCG2", Name == "Palmitic Acid", cell_type == "OPC") |>
           pull(Compound.ID) |> first(),
  gene = "PLCG2", ct = "OPC",
  meta_name = "Palmitic Acid"
)
if (!is.null(p_plcg2)) {
  save_dual_format(p_plcg2, opc_spotlight_dir, "OPC_PLCG2_palmitic", width = 8, height = 6)
  print(p_plcg2)
}

# ---- NRG1 ~ Pc(O-10:0_23:2): OPC differentiation signal + phospholipid ----
p_nrg1 <- make_spotlight_scatter(
  cid  = hits |> filter(gene == "NRG1", cell_type == "OPC") |> pull(Compound.ID) |> first(),
  gene = "NRG1", ct = "OPC",
  meta_name = hits |> filter(gene == "NRG1", cell_type == "OPC") |> pull(Name) |> first()
)
if (!is.null(p_nrg1)) {
  save_dual_format(p_nrg1, opc_spotlight_dir, "OPC_NRG1_phospholipid", width = 8, height = 6)
  print(p_nrg1)
}
```


## Interpretation spotlight scatters

Three categories of gene–metabolite association for the manuscript narrative:

1. **Direct / expected** — enzyme whose protein directly processes the associated
   metabolite class (PLCG2, a phospholipase, with lipid metabolites in OPC).
2. **Ribosomal / expected** — ribosomal protein genes whose expression covaries
   broadly with metabolic activity rather than specific metabolites.
3. **Indirect / transcription factor** — TF with no enzymatic metabolite link;
   association implies coordinated developmental programs.

```{r interpretation-spotlights, fig.width=8, fig.height=6}
interp_dir <- file.path(csv_dir, "spotlight-interpretation")
dir.create(interp_dir, recursive = TRUE, showWarnings = FALSE)

# ---- 1. Direct: PLCG2 ~ Cer(D15:2_7:0) in OPC ----
# Phospholipase C gamma 2 cleaves lipid substrates; ceramide association is
# a direct enzyme-substrate class relationship.
p_direct <- make_spotlight_scatter(
  cid  = hits |> filter(gene == "PLCG2", cell_type == "OPC",
                         grepl("Cer", Name)) |> pull(Compound.ID) |> first(),
  gene = "PLCG2", ct = "OPC",
  meta_name = hits |> filter(gene == "PLCG2", cell_type == "OPC",
                              grepl("Cer", Name)) |> pull(Name) |> first(),
  label    = "Direct: PLCG2 ~ ceramide (OPC)"
)
if (!is.null(p_direct)) {
  save_dual_format(p_direct, interp_dir, "direct_PLCG2_ceramide", width = 8, height = 6)
  print(p_direct)
}

# ---- 2. Ribosomal: RPL37A ~ Pyrrole-2-Carboxylic Acid in EN-Non-IT-Immature ----
# RPL37A has the most metabolite associations among ribosomal genes (5).
# 73 RPL/RPS genes collectively associate with Pyrrole-2-Carboxylic Acid,
# reflecting coordinated translational machinery and metabolic activity.
p_ribo <- make_spotlight_scatter(
  cid  = hits |> filter(gene == "RPL37A",
                         Name == "Pyrrole-2-Carboxylic Acid") |>
           pull(Compound.ID) |> first(),
  gene = "RPL37A", ct = "EN-Non-IT-Immature",
  meta_name = "Pyrrole-2-Carboxylic Acid",
  label    = "Ribosomal: RPL37A ~ Pyrrole-2-Carboxylic Acid (EN-Non-IT-Immature)"
)
if (!is.null(p_ribo)) {
  save_dual_format(p_ribo, interp_dir, "ribosomal_RPL37A_pyrrole", width = 8, height = 6)
  print(p_ribo)
}

# ---- 3. Indirect: TSHZ2 ~ Palmitic Acid in IN-MGE-SST ----
# TSHZ2 is a zinc-finger TF for cortical neuron specification (Caubit et al.
# 2016, Cerebral Cortex). No enzymatic link to fatty acid metabolism; suggests
# coordinated developmental gene expression programs that covary with metabolic
# state.
p_tf <- make_spotlight_scatter(
  cid  = hits |> filter(gene == "TSHZ2", Name == "Palmitic Acid",
                         cell_type == "IN-MGE-SST") |>
           pull(Compound.ID) |> first(),
  gene = "TSHZ2", ct = "IN-MGE-SST",
  meta_name = "Palmitic Acid",
  label    = "Indirect: TSHZ2 ~ Palmitic Acid (IN-MGE-SST)"
)
if (!is.null(p_tf)) {
  save_dual_format(p_tf, interp_dir, "indirect_TSHZ2_palmitic", width = 8, height = 6)
  print(p_tf)
}
```


## IN-MGE-SST top 10 associations

IN-MGE-SST has the highest total association count. Spotlight scatters for
the 10 strongest associations (ranked by global FDR, then |logFC|).

```{r inmge-sst-top10, fig.width=8, fig.height=6}
sst_dir <- file.path(csv_dir, "spotlight-IN-MGE-SST")
dir.create(sst_dir, recursive = TRUE, showWarnings = FALSE)

sst_top10 <- hits |>
  filter(cell_type == "IN-MGE-SST") |>
  arrange(FDR_global, desc(abs(logFC))) |>
  slice_head(n = 10)

for (i in seq_len(nrow(sst_top10))) {
  r  <- sst_top10[i, ]
  pl <- make_spotlight_scatter(
    cid       = r$Compound.ID,
    gene      = r$gene,
    ct        = "IN-MGE-SST",
    meta_name = r$Name
  )
  if (is.null(pl)) next
  base <- gsub("[^A-Za-z0-9]+", "-", paste("SST", r$gene, r$Name, sep = "_"))
  save_dual_format(pl, sst_dir, base, width = 8, height = 6)
  print(pl)
}
cat("Saved", nrow(sst_top10), "IN-MGE-SST spotlight scatters to", sst_dir, "\n")
```


# AI Assistance Disclosure

Code in this script was developed with assistance from Claude (Anthropic). All
AI-generated code was reviewed, validated, and adapted by the author.

# Session Info

```{r session-info}
sessionInfo()
```
