---
title: "Gene–Metabolite: Gene Pathway Analysis"
author: "Sneha M. Chaturvedi"
date: "Last updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

Gene-set enrichment analysis (fgsea) on the gene-side results from metabolite-gene associations. 

| | |
|---|---|
| **Inputs** | Association CSVs from `metabolite_gene_associations.Rmd` |
| **Outputs** | Enrichment CSVs and figures in `results/gene-metabolite/gene-pathway/` |
| **Upstream** | `metabolite_gene_associations.Rmd` |
| **Downstream** | None |


## Overview

ORA + GSEA on the **genes** from all gene–metabolite associations. The companion script
`gene_metabolite_pathway_analysis.Rmd` does the same for metabolites.

- **ORA**: Fisher's exact test; foreground = genes with ≥1 significant association;
  background = all genes tested in the limma-voom loop.
- **GSEA**: fgsea ranked by −log10(min raw *P*-value per gene) across all
  metabolite × cell type pairs. Unsigned — NES > 0 means the gene set's members
  are enriched for metabolite associations.

Three analyses:

1. **All cell types (union)** — genes significant in ≥1 cell type
2. **Cell-type-specific** — ORA + GSEA per cell type
3. **Glia vs. neurons** — class-specific gene sets and shared pathway comparison

**Gene sets** from `msigdbr` (MSigDB v2023.2): GO:BP, GO:MF, GO:CC, Hallmark, KEGG, Reactome. 

## Set Up

```{r setup}
Packages <- c(
  "tidyverse", "arrow", "fgsea", "msigdbr",
  "knitr", "kableExtra", "svglite", "ggrepel", "cowplot"
)
lapply(Packages, library, character.only = TRUE)

find_project_root <- function(marker = ".git") {
  d <- normalizePath(getwd())
  repeat {
    if (dir.exists(file.path(d, marker))) return(d)
    parent <- dirname(d)
    if (parent == d) stop("Project root not found (no ", marker, " above ", getwd(), ")")
    d <- parent
  }
}
root <- find_project_root()
source(file.path(root, "src/gene-hormone/pseudobulk_functions.R"))   # save_dual_format()

set.seed(42)

find_project_root <- function(marker = ".git") {
  d <- normalizePath(getwd())
  repeat {
    if (dir.exists(file.path(d, marker))) return(d)
    parent <- dirname(d)
    if (parent == d) stop("Could not locate project root (no ", marker, " found above ", getwd(), ")")
    d <- parent
  }
}
results_dir <- file.path(find_project_root(), "results")
```

```{r config}

METHOD <- "log2_na"   
suffix <- if (METHOD == "int") "" else paste0("-", METHOD)

# ---- paths -------------------------------------------------------------------
parquet_dir <- if (METHOD == "int") {
  file.path(results_dir, "gene-metabolite/parquet")
} else {
  file.path(results_dir, sprintf("gene-metabolite/parquet-%s", METHOD))
}

iter_dir  <- file.path(results_dir, "gene-metabolite",
                       paste0("stability-imputation", suffix))
hits_path <- file.path(iter_dir, "metabolite_gene_hits_imputation_stable.csv")

out_dir  <- file.path(iter_dir, "gene-pathway")
csv_path <- file.path(out_dir, "csv")
for (d in c(out_dir, csv_path)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

# ---- gene set parameters -----------------------------------------------------
GSEA_MIN_SIZE  <- 10    # min genes in set after restricting to tested background
GSEA_MAX_SIZE  <- 500
MIN_CT_HITS    <- 5     # min distinct gene hits to run ORA for a cell type

# ---- cell type groupings (update after inspecting "Available Cell Types") ----
GLIAL_TYPES <- c(
  "Astrocyte-Immature", "OPC"
)
NEURONAL_TYPES <- c(
  "EN-IT-Immature", "EN-L6-IT", "EN-L6b", "EN-Newborn", "EN-Non-IT-Immature",
  "IN-CGE-Immature", "IN-MGE-Immature", "IN-MGE-SST", "IPC-EN"
)
```

## Load Hits

```{r load-hits}
hits <- read_csv(hits_path, show_col_types = FALSE)

cat(sprintf("Reading: %s\n", hits_path))
cat(sprintf(
  "Hits: %s rows | %d metabolites | %d cell types | %d genes\n",
  format(nrow(hits), big.mark = ","),
  n_distinct(hits$Compound.ID),
  n_distinct(hits$cell_type),
  n_distinct(hits$gene)
))
```

## Available Cell Types

Inspect output below to set `GLIAL_TYPES` and `NEURONAL_TYPES` in config.

```{r cell-types-print}
cat("Cell types present in hits:\n")
print(sort(unique(hits$cell_type)))
```

## Gene Background (Parquet)

Background = all genes tested in the limma-voom loop, regardless of significance.


```{r gene-background}
ds <- open_dataset(parquet_dir)

background_genes <- ds |>
  distinct(gene) |>
  collect() |>
  pull(gene)

cat(sprintf("Total genes in background: %s\n",
            format(length(background_genes), big.mark = ",")))
```

## GSEA Rankings (Parquet)

Per-gene −log10(min raw *P*-value) across all metabolite × cell type pairs.
Genes with no association anywhere receive score 0 (min_P = 1).

```{r gsea-ranking}
# All-cell-types ranking
ranking_all_raw <- ds |>
  group_by(gene) |>
  summarise(min_P = min(P.Value, na.rm = TRUE), .groups = "drop") |>
  collect()

# Cell-type-specific rankings
ranking_ct_raw <- ds |>
  group_by(cell_type, gene) |>
  summarise(min_P = min(P.Value, na.rm = TRUE), .groups = "drop") |>
  collect()

cat(sprintf(
  "All-gene ranking rows: %s | Cell-type ranking rows: %s\n",
  format(nrow(ranking_all_raw), big.mark = ","),
  format(nrow(ranking_ct_raw),  big.mark = ",")
))
```

---

# Gene Set Annotation

Fetches six MSigDB collections via `msigdbr`. 

## Fetch Gene Sets

```{r fetch-gene-sets}
fetch_msigdb_sets <- function(category, subcategory = NULL, label) {
  df <- if (is.null(subcategory)) {
    msigdbr(species = "Homo sapiens", category = category)
  } else {
    msigdbr(species = "Homo sapiens", category = category,
            subcategory = subcategory)
  }
  df |>
    dplyr::select(gs_name, gene_symbol) |>
    mutate(collection = label)
}

gene_set_collections <- list(
  fetch_msigdb_sets("H",  NULL,           "Hallmark"),
  fetch_msigdb_sets("C5", "GO:BP",        "GO_BP"),
  fetch_msigdb_sets("C5", "GO:MF",        "GO_MF"),
  fetch_msigdb_sets("C5", "GO:CC",        "GO_CC"),
  fetch_msigdb_sets("C2", "CP:KEGG", "KEGG"),
  fetch_msigdb_sets("C2", "CP:REACTOME",  "Reactome")
)

gene_sets_raw <- bind_rows(gene_set_collections)

cat(sprintf(
  "MSigDB gene sets loaded: %s gene-set members | %d unique sets across %d collections\n",
  format(nrow(gene_sets_raw), big.mark = ","),
  n_distinct(gene_sets_raw$gs_name),
  n_distinct(gene_sets_raw$collection)
))
print(count(gene_sets_raw, collection))
```

## Build Gene Set List

```{r build-gene-set-list}
# Append collection to set name to keep cross-database duplicates distinct,
# then size-filter against the tested background.
gene_sets <- gene_sets_raw |>
  filter(gene_symbol %in% background_genes) |>
  group_by(gs_name, collection) |>
  summarise(members = list(unique(gene_symbol)), .groups = "drop") |>
  mutate(
    n_members    = lengths(members),
    gs_label     = paste0(gs_name, " (", collection, ")")
  ) |>
  filter(n_members >= GSEA_MIN_SIZE, n_members <= GSEA_MAX_SIZE)

gene_set_source_lookup <- gene_sets |>
  dplyr::select(gs_label, collection)

gene_set_list <- setNames(gene_sets$members, gene_sets$gs_label)

cat(sprintf(
  "Gene sets after size filter [%d–%d]: %d | median members: %d\n",
  GSEA_MIN_SIZE, GSEA_MAX_SIZE, length(gene_set_list),
  median(lengths(gene_set_list))
))
print(count(gene_sets, collection))
```

---

# Analysis Functions

```{r functions}

# ---- ORA (Fisher's exact, one-sided greater) ---------------------------------
run_gene_ora <- function(foreground, background, gene_set_list,
                         gene_set_source_lookup, analysis_label) {
  cat("\n===", analysis_label, "===\n")
  cat("Foreground genes:", length(foreground),
      "| Background genes:", length(background), "\n")

  if (length(foreground) < 1) { cat("Empty foreground — skipping.\n"); return(NULL) }

  bg_only <- background[!background %in% foreground]

  results <- map_dfr(names(gene_set_list), function(gs) {
    members <- gene_set_list[[gs]]
    a <- sum(foreground %in% members)
    b <- length(foreground) - a
    c <- sum(bg_only %in% members)
    d <- length(bg_only) - c
    ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2), alternative = "greater")
    tibble(
      gene_set   = gs,
      n_hits     = a,
      set_size   = a + c,
      odds_ratio = as.numeric(ft$estimate),
      pval       = ft$p.value,
      hit_genes  = paste(foreground[foreground %in% members], collapse = ";")
    )
  }) |>
    filter(n_hits > 0) |>
    mutate(padj = p.adjust(pval, method = "BH")) |>
    left_join(gene_set_source_lookup, by = c("gene_set" = "gs_label")) |>
    arrange(padj, pval)

  cat("Sets with \u22651 hit:", nrow(results),
      "| p < 0.05:", sum(results$pval  < 0.05, na.rm = TRUE),
      "| FDR < 0.05:", sum(results$padj < 0.05, na.rm = TRUE), "\n")
  results
}

# ---- GSEA (fgsea on −log10 min P per gene) -----------------------------------
run_gene_gsea <- function(ranking_df, background_genes, gene_set_list,
                          analysis_label, nperm = 10000, seed = 42) {
  cat("\n===", analysis_label, "===\n")

  ranked <- tibble(gene = background_genes) |>
    left_join(ranking_df |> dplyr::select(gene, min_P), by = "gene") |>
    mutate(score = -log10(pmax(replace_na(min_P, 1), 1e-300))) |>
    arrange(desc(score)) |>
    dplyr::select(gene, score) |>
    deframe()

  cat("Genes in ranked list:", length(ranked), "\n")

  gs_filt <- lapply(gene_set_list, function(x) intersect(x, names(ranked)))
  gs_filt <- gs_filt[lengths(gs_filt) >= GSEA_MIN_SIZE]
  cat("Gene sets after intersection:", length(gs_filt), "\n")

  if (length(gs_filt) == 0) {
    cat("No gene sets with sufficient coverage — skipping.\n")
    return(NULL)
  }

  set.seed(seed)
  results <- fgsea(
    pathways    = gs_filt,
    stats       = ranked,
    nPermSimple = nperm,
    minSize     = GSEA_MIN_SIZE,
    maxSize     = GSEA_MAX_SIZE
  ) |>
    as_tibble() |>
    arrange(padj, pval) |>
    mutate(analysis = analysis_label)

  cat("Sets tested:", nrow(results),
      "| p < 0.05:", sum(results$pval  < 0.05, na.rm = TRUE),
      "| FDR < 0.05:", sum(results$padj < 0.05, na.rm = TRUE), "\n")
  results
}

# ---- Plots -------------------------------------------------------------------
# Collection colour palette (6 collections)
COLL_COLORS <- c(
  Hallmark  = "#984EA3",
  GO_BP     = "#377EB8",
  GO_MF     = "#4DAF4A",
  GO_CC     = "#FF7F00",
  KEGG      = "#E41A1C",
  Reactome  = "#A65628"
)

plot_ora_dotplot <- function(results, analysis_label, top_n = 20,
                             collections = NULL) {
  if (is.null(results) || nrow(results) == 0) return(NULL)

  plot_data <- results
  if (!is.null(collections)) plot_data <- dplyr::filter(plot_data, collection %in% collections)
  plot_data <- plot_data |>
    dplyr::filter(pval < 0.05) |>
    head(top_n) |>
    mutate(
      stars     = case_when(padj < 0.01 ~ "**", padj < 0.05 ~ "*", TRUE ~ ""),
      gs_base   = str_remove(gene_set, " \\([^)]+\\)$"),
      gs_short  = str_trunc(gs_base, 45),
      gs_label  = paste0(gs_short, " (", collection, ")",
                         if_else(stars != "", paste0(" ", stars), ""))
    )

  if (nrow(plot_data) == 0) { message("No nominal hits for: ", analysis_label); return(NULL) }

  ggplot(plot_data,
         aes(x = log2(odds_ratio),
             y = reorder(gs_label, log2(odds_ratio)),
             color = collection, shape = collection, size = set_size)) +
    geom_point(alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    scale_color_manual(name = "Collection", values = COLL_COLORS) +
    scale_shape_manual(name = "Collection",
                       values = c(Hallmark = 18, GO_BP = 16, GO_MF = 17,
                                  GO_CC = 15, KEGG = 8, Reactome = 4)) +
    scale_size_continuous(name = "Set size", range = c(2, 8)) +
    theme_minimal(base_size = 11) +
    labs(x = "log2(Odds Ratio)", y = NULL,
         title    = paste0("ORA \u2014 ", analysis_label),
         subtitle = paste0("Top ", nrow(plot_data), " by p-value (p < 0.05)\n",
                           "* FDR < 0.05   ** FDR < 0.01")) +
    theme(plot.title    = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9),
          axis.text.y   = element_text(size = 8),
          legend.position = "right")
}

plot_gsea_dotplot <- function(results, gene_set_source_lookup,
                              analysis_label, top_n = 25,
                              collections = NULL) {
  if (is.null(results) || nrow(results) == 0) return(NULL)

  plot_data <- results |>
    left_join(gene_set_source_lookup, by = c("pathway" = "gs_label"))
  if (!is.null(collections)) plot_data <- dplyr::filter(plot_data, collection %in% collections)
  plot_data <- plot_data |>
    dplyr::filter(pval < 0.05) |>
    arrange(padj) |>
    head(top_n) |>
    mutate(
      stars    = case_when(padj < 0.01 ~ "**", padj < 0.05 ~ "*", TRUE ~ ""),
      gs_base  = str_remove(pathway, " \\([^)]+\\)$"),
      gs_short = str_trunc(gs_base, 45),
      gs_label = paste0(gs_short, " (", collection, ")",
                        if_else(stars != "", paste0(" ", stars), ""))
    )

  if (nrow(plot_data) == 0) { message("No nominal hits for: ", analysis_label); return(NULL) }

  ggplot(plot_data,
         aes(x = NES,
             y = reorder(gs_label, NES),
             color = collection, shape = collection, size = size)) +
    geom_point(alpha = 0.85) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    scale_color_manual(name = "Collection", values = COLL_COLORS) +
    scale_shape_manual(name = "Collection",
                       values = c(Hallmark = 18, GO_BP = 16, GO_MF = 17,
                                  GO_CC = 15, KEGG = 8, Reactome = 4)) +
    scale_size_continuous(name = "Set size", range = c(2, 8)) +
    theme_minimal(base_size = 11) +
    labs(x = "Normalized Enrichment Score (NES)", y = NULL,
         title    = paste0("GSEA \u2014 ", analysis_label),
         subtitle = paste0("Top ", nrow(plot_data), " by FDR (p < 0.05)\n",
                           "* FDR < 0.05   ** FDR < 0.01\nNES > 0: set enriched for metabolite associations")) +
    theme(plot.title    = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9),
          axis.text.y   = element_text(size = 8),
          legend.position = "right")
}

# ---- Table helper ------------------------------------------------------------
display_results_table <- function(results, caption, top_n = 20,
                                  type = c("ora", "gsea"),
                                  gene_set_source_lookup = NULL) {
  type <- match.arg(type)
  if (is.null(results) || nrow(results) == 0) { cat("No results.\n"); return(invisible(NULL)) }

  if (type == "ora") {
    results |>
      head(top_n) |>
      mutate(odds_ratio = round(odds_ratio, 2),
             pval = signif(pval, 3), padj = signif(padj, 3),
             hit_genes = str_trunc(hit_genes, 60)) |>
      dplyr::select(gene_set, collection, n_hits, set_size,
                    odds_ratio, pval, padj, hit_genes) |>
      rename(`Gene set` = gene_set, Collection = collection,
             Hits = n_hits, `Set size` = set_size,
             OR = odds_ratio, `p-value` = pval, FDR = padj,
             `Hit genes` = hit_genes) |>
      kable(caption = caption) |>
      kable_styling(bootstrap_options = c("striped", "condensed"),
                    full_width = FALSE, font_size = 11) |>
      print()
  } else {
    df <- results
    if (!is.null(gene_set_source_lookup))
      df <- df |> left_join(gene_set_source_lookup, by = c("pathway" = "gs_label"))
    df |>
      head(top_n) |>
      mutate(pval = signif(pval, 3), padj = signif(padj, 3),
             NES  = round(NES, 3),
             leadingEdge = map_chr(leadingEdge, ~ paste(head(.x, 5), collapse = ", "))) |>
      dplyr::select(pathway, collection, pval, padj, NES, size, leadingEdge) |>
      rename(`Gene set` = pathway, Collection = collection,
             `p-value` = pval, FDR = padj, `Set size` = size,
             `Leading edge (top 5)` = leadingEdge) |>
      kable(caption = caption) |>
      kable_styling(bootstrap_options = c("striped", "condensed"),
                    full_width = FALSE, font_size = 11) |>
      print()
  }
}

flatten_leading_edge <- function(df) {
  if (!is.null(df) && "leadingEdge" %in% colnames(df))
    df |> mutate(leadingEdge = map_chr(leadingEdge, ~ paste(.x, collapse = ";")))
  else df
}
```

---

# 1. All Cell Types (Union)

Foreground: genes with ≥1 significant association in any cell type.

## ORA and GSEA

```{r run-all}
genes_all <- hits |> distinct(gene) |> pull(gene)
cat("Union gene hits:", length(genes_all), "\n")

ora_all  <- run_gene_ora(
  genes_all, background_genes, gene_set_list,
  gene_set_source_lookup, "All cell types (union)"
)
gsea_all <- run_gene_gsea(
  ranking_all_raw, background_genes, gene_set_list,
  "All cell types"
)
```

## ORA — Table

```{r results-ora-all-table}
display_results_table(ora_all, "ORA: All cell types — top 20", type = "ora")
```

## ORA — GO:BP Plot

```{r results-ora-all-gobp, fig.width=10, fig.height=7}
p <- plot_ora_dotplot(ora_all, "All cell types — GO:BP", collections = "GO_BP")
if (!is.null(p)) { save_dual_format(p, out_dir, "ora_all_GO_BP"); print(p) }
```

## ORA — Hallmark + KEGG + Reactome Plot

```{r results-ora-all-curated, fig.width=10, fig.height=7}
p <- plot_ora_dotplot(ora_all, "All cell types — Curated",
                      collections = c("Hallmark", "KEGG", "Reactome"))
if (!is.null(p)) { save_dual_format(p, out_dir, "ora_all_curated"); print(p) }
```

## GSEA — Table

```{r results-gsea-all-table}
display_results_table(gsea_all, "GSEA: All cell types — top 20",
                      type = "gsea", gene_set_source_lookup = gene_set_source_lookup)
```

## GSEA — GO:BP Plot

```{r results-gsea-all-gobp, fig.width=10, fig.height=8}
p <- plot_gsea_dotplot(gsea_all, gene_set_source_lookup,
                       "All cell types — GO:BP", collections = "GO_BP")
if (!is.null(p)) { save_dual_format(p, out_dir, "gsea_all_GO_BP"); print(p) }
```

## GSEA — Hallmark + KEGG + Reactome Plot

```{r results-gsea-all-curated, fig.width=10, fig.height=8}
p <- plot_gsea_dotplot(gsea_all, gene_set_source_lookup,
                       "All cell types — Curated",
                       collections = c("Hallmark", "KEGG", "Reactome"))
if (!is.null(p)) { save_dual_format(p, out_dir, "gsea_all_curated"); print(p) }
```

---

# 2. Cell-Type-Specific

ORA and GSEA run independently per cell type. Cell types with fewer than
`r MIN_CT_HITS` gene hits are skipped for ORA.

```{r celltype-loop}
cell_types   <- sort(unique(hits$cell_type))
ora_ct_list  <- list()
gsea_ct_list <- list()

for (ct in cell_types) {
  ct_genes <- hits |> filter(cell_type == ct) |> distinct(gene) |> pull(gene)
  cat(sprintf("\n--- %s | %d gene hits ---\n", ct, length(ct_genes)))

  if (length(ct_genes) >= MIN_CT_HITS) {
    ora_ct_list[[ct]] <- run_gene_ora(
      ct_genes, background_genes, gene_set_list, gene_set_source_lookup, ct
    )
  } else {
    cat("  ORA skipped: fewer than", MIN_CT_HITS, "hits\n")
  }

  ct_ranking <- ranking_ct_raw |> filter(cell_type == ct)
  if (nrow(ct_ranking) > 0)
    gsea_ct_list[[ct]] <- run_gene_gsea(ct_ranking, background_genes, gene_set_list, ct)
}

cat(sprintf(
  "\nCompleted: ORA for %d cell types | GSEA for %d cell types\n",
  length(ora_ct_list), length(gsea_ct_list)
))
```

## Cell-Type ORA — Summary Table

Top 3 gene sets per cell type (p < 0.05).

```{r ct-ora-summary}
ct_ora_summary <- map_dfr(names(ora_ct_list), function(ct) {
  res <- ora_ct_list[[ct]]
  if (is.null(res) || nrow(res) == 0) return(NULL)
  res |>
    filter(pval < 0.05) |>
    slice_head(n = 3) |>
    mutate(cell_type = ct,
           gs_short  = str_trunc(str_remove(gene_set, " \\([^)]+\\)$"), 40))
})

if (nrow(ct_ora_summary) > 0) {
  ct_ora_summary |>
    dplyr::select(cell_type, gs_short, collection,
                  n_hits, set_size, odds_ratio, pval, padj) |>
    mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
    rename(`Cell type` = cell_type, `Gene set` = gs_short,
           Collection = collection, Hits = n_hits,
           `Set size` = set_size, OR = odds_ratio,
           `p-value` = pval, FDR = padj) |>
    kable(caption = "Top 3 ORA gene sets per cell type (p < 0.05)") |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE, font_size = 10)
}
```

## Cell-Type ORA — Dot Plots

```{r ct-ora-plots, fig.width=10, fig.height=7}
for (ct in names(ora_ct_list)) {
  p <- plot_ora_dotplot(ora_ct_list[[ct]], ct, top_n = 15)
  if (!is.null(p)) {
    ct_safe <- gsub("[^A-Za-z0-9_]", "_", ct)
    save_dual_format(p, out_dir, paste0("ora_ct_", ct_safe))
    print(p)
  }
}
```

## Cell-Type GSEA — Summary Table

```{r ct-gsea-summary}
ct_gsea_summary <- map_dfr(names(gsea_ct_list), function(ct) {
  res <- gsea_ct_list[[ct]]
  if (is.null(res) || nrow(res) == 0) return(NULL)
  res |>
    left_join(gene_set_source_lookup, by = c("pathway" = "gs_label")) |>
    filter(pval < 0.05) |>
    slice_head(n = 3) |>
    mutate(cell_type = ct,
           gs_short  = str_trunc(str_remove(pathway, " \\([^)]+\\)$"), 40))
})

if (nrow(ct_gsea_summary) > 0) {
  ct_gsea_summary |>
    dplyr::select(cell_type, gs_short, collection, pval, padj, NES, size) |>
    mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
    rename(`Cell type` = cell_type, `Gene set` = gs_short,
           Collection = collection,
           `p-value` = pval, FDR = padj, `Set size` = size) |>
    kable(caption = "Top 3 GSEA gene sets per cell type (p < 0.05)") |>
    kable_styling(bootstrap_options = c("striped", "condensed"),
                  full_width = FALSE, font_size = 10)
}
```

## Cell-Type GSEA — Dot Plots

```{r ct-gsea-plots, fig.width=10, fig.height=8}
for (ct in names(gsea_ct_list)) {
  p <- plot_gsea_dotplot(gsea_ct_list[[ct]], gene_set_source_lookup, ct, top_n = 15)
  if (!is.null(p)) {
    ct_safe <- gsub("[^A-Za-z0-9_]", "_", ct)
    save_dual_format(p, out_dir, paste0("gsea_ct_", ct_safe))
    print(p)
  }
}
```

---

# 3. Glia vs. Neurons

## Define Groups

```{r define-groups}
glia_genes   <- hits |> filter(cell_type %in% GLIAL_TYPES)    |> distinct(gene) |> pull(gene)
neuron_genes <- hits |> filter(cell_type %in% NEURONAL_TYPES) |> distinct(gene) |> pull(gene)

glia_only   <- setdiff(glia_genes,   neuron_genes)
neuron_only <- setdiff(neuron_genes, glia_genes)
shared_both <- intersect(glia_genes, neuron_genes)

cat(sprintf(
  "Glia genes: %d | Neuron genes: %d\nGlia-only: %d | Neuron-only: %d | Shared: %d\n",
  length(glia_genes), length(neuron_genes),
  length(glia_only), length(neuron_only), length(shared_both)
))

unmatched_glia   <- setdiff(GLIAL_TYPES,    unique(hits$cell_type))
unmatched_neuron <- setdiff(NEURONAL_TYPES, unique(hits$cell_type))
if (length(unmatched_glia)   > 0) warning("GLIAL_TYPES not in hits: ",    paste(unmatched_glia,   collapse = ", "))
if (length(unmatched_neuron) > 0) warning("NEURONAL_TYPES not in hits: ", paste(unmatched_neuron, collapse = ", "))
```

## Gene Overlap

```{r overlap-bar, fig.width=6, fig.height=3}
tibble(
  group = factor(c("Glia-only", "Shared", "Neuron-only"),
                 levels = c("Glia-only", "Shared", "Neuron-only")),
  n     = c(length(glia_only), length(shared_both), length(neuron_only))
) |>
  ggplot(aes(x = group, y = n, fill = group)) +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_text(aes(label = n), vjust = -0.4, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c(
    "Glia-only"   = "#4E9AF1",
    "Shared"      = "#9B59B6",
    "Neuron-only" = "#E67E22"
  )) +
  theme_minimal(base_size = 12) +
  labs(x = NULL, y = "Distinct genes",
       title = "Gene overlap: glia vs. neurons") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```

## ORA — Glia and Neurons

```{r gn-ora}
ora_glia        <- run_gene_ora(glia_genes,   background_genes, gene_set_list, gene_set_source_lookup, "Glia (union)")
ora_neuron      <- run_gene_ora(neuron_genes, background_genes, gene_set_list, gene_set_source_lookup, "Neurons (union)")
ora_glia_only   <- run_gene_ora(glia_only,    background_genes, gene_set_list, gene_set_source_lookup, "Glia-exclusive")
ora_neuron_only <- run_gene_ora(neuron_only,  background_genes, gene_set_list, gene_set_source_lookup, "Neuron-exclusive")
```

## GSEA — Glia and Neurons

Rankings collapse to min P.Value within each cell class.

```{r gn-gsea}
glia_ranking <- ranking_ct_raw |>
  filter(cell_type %in% GLIAL_TYPES) |>
  group_by(gene) |>
  summarise(min_P = min(min_P, na.rm = TRUE), .groups = "drop")

neuron_ranking <- ranking_ct_raw |>
  filter(cell_type %in% NEURONAL_TYPES) |>
  group_by(gene) |>
  summarise(min_P = min(min_P, na.rm = TRUE), .groups = "drop")

gsea_glia   <- run_gene_gsea(glia_ranking,   background_genes, gene_set_list, "Glia")
gsea_neuron <- run_gene_gsea(neuron_ranking, background_genes, gene_set_list, "Neurons")
```

## Results — ORA Glia (Union)

```{r plot-ora-glia, fig.width=10, fig.height=7}
display_results_table(ora_glia, "ORA: Glia (union) — top 20", type = "ora")
p <- plot_ora_dotplot(ora_glia, "Glia (union)")
if (!is.null(p)) { save_dual_format(p, out_dir, "ora_glia"); print(p) }
```

## Results — ORA Neurons (Union)

```{r plot-ora-neuron, fig.width=10, fig.height=7}
display_results_table(ora_neuron, "ORA: Neurons (union) — top 20", type = "ora")
p <- plot_ora_dotplot(ora_neuron, "Neurons (union)")
if (!is.null(p)) { save_dual_format(p, out_dir, "ora_neurons"); print(p) }
```

## Results — ORA Glia-Exclusive

```{r plot-ora-glia-only, fig.width=10, fig.height=7}
display_results_table(ora_glia_only, "ORA: Glia-exclusive genes — top 20", type = "ora")
p <- plot_ora_dotplot(ora_glia_only, "Glia-exclusive")
if (!is.null(p)) { save_dual_format(p, out_dir, "ora_glia_only"); print(p) }
```

## Results — ORA Neuron-Exclusive

```{r plot-ora-neuron-only, fig.width=10, fig.height=7}
display_results_table(ora_neuron_only, "ORA: Neuron-exclusive genes — top 20", type = "ora")
p <- plot_ora_dotplot(ora_neuron_only, "Neuron-exclusive")
if (!is.null(p)) { save_dual_format(p, out_dir, "ora_neuron_only"); print(p) }
```

## Results — GSEA Glia

```{r plot-gsea-glia, fig.width=10, fig.height=8}
display_results_table(gsea_glia, "GSEA: Glia — top 20", type = "gsea",
                      gene_set_source_lookup = gene_set_source_lookup)
p <- plot_gsea_dotplot(gsea_glia, gene_set_source_lookup, "Glia")
if (!is.null(p)) { save_dual_format(p, out_dir, "gsea_glia"); print(p) }
```

## Results — GSEA Neurons

```{r plot-gsea-neuron, fig.width=10, fig.height=8}
display_results_table(gsea_neuron, "GSEA: Neurons — top 20", type = "gsea",
                      gene_set_source_lookup = gene_set_source_lookup)
p <- plot_gsea_dotplot(gsea_neuron, gene_set_source_lookup, "Neurons")
if (!is.null(p)) { save_dual_format(p, out_dir, "gsea_neurons"); print(p) }
```

## Shared Gene Sets: Glia vs. Neurons (ORA)

Gene sets significant in both (p < 0.05); scatter shows which class has the
stronger enrichment (above diagonal = stronger in neurons).

```{r gn-geneset-overlap, fig.width=9, fig.height=7}
if (!is.null(ora_glia) && !is.null(ora_neuron)) {
  shared_sets <- inner_join(
    ora_glia   |> filter(pval < 0.05) |>
      dplyr::select(gene_set, collection, or_glia   = odds_ratio, pval_glia   = pval),
    ora_neuron |> filter(pval < 0.05) |>
      dplyr::select(gene_set,             or_neuron = odds_ratio, pval_neuron = pval),
    by = "gene_set"
  ) |>
    mutate(
      gs_short       = str_trunc(str_remove(gene_set, " \\([^)]+\\)$"), 38),
      log2_or_glia   = log2(or_glia),
      log2_or_neuron = log2(or_neuron)
    )

  cat(sprintf("Gene sets significant in both (p < 0.05): %d\n", nrow(shared_sets)))

  if (nrow(shared_sets) > 0) {
    p_overlap <- ggplot(shared_sets,
                        aes(x = log2_or_glia, y = log2_or_neuron,
                            color = collection, label = gs_short)) +
      geom_point(alpha = 0.8, size = 3) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
      geom_hline(yintercept = 0, linetype = "dotted", color = "grey70") +
      geom_vline(xintercept = 0, linetype = "dotted", color = "grey70") +
      ggrepel::geom_text_repel(size = 3, max.overlaps = 15) +
      scale_color_manual(name = "Collection", values = COLL_COLORS) +
      theme_minimal(base_size = 11) +
      labs(x = "log2(OR) in Glia", y = "log2(OR) in Neurons",
           title    = "Shared gene sets: Glia vs. Neurons (ORA)",
           subtitle = "Above diagonal = stronger in neurons | Below = stronger in glia") +
      theme(plot.title    = element_text(hjust = 0.5),
            plot.subtitle = element_text(hjust = 0.5, color = "grey40", size = 9))

    save_dual_format(p_overlap, out_dir, "glia_neuron_geneset_overlap")
    print(p_overlap)
  }
}
```

---

# Metabolic Hub Genes (≥ 5 distinct metabolites)

Genes associated with ≥5 distinct metabolites 

## Identify hub genes

```{r hub-genes}
MIN_METAB_RECURRENCE <- 5

hub_recur <- hits |>
  group_by(gene) |>
  summarise(n_metabolites = n_distinct(Compound.ID),
            n_cell_types  = n_distinct(cell_type), .groups = "drop") |>
  arrange(desc(n_metabolites))

hub_genes <- hub_recur |>
  filter(n_metabolites >= MIN_METAB_RECURRENCE) |>
  pull(gene)

cat(sprintf(
  "Metabolite recurrence distribution (all %d genes):\n", nrow(hub_recur)))
print(summary(hub_recur$n_metabolites))
cat(sprintf(
  "\nHub genes (>= %d distinct metabolites): %d of %d (%.1f%%)\n",
  MIN_METAB_RECURRENCE, length(hub_genes), nrow(hub_recur),
  100 * length(hub_genes) / nrow(hub_recur)))

knitr::kable(
  hub_recur |> filter(n_metabolites >= MIN_METAB_RECURRENCE) |> head(30),
  caption = sprintf("Top hub genes (>= %d distinct metabolites)", MIN_METAB_RECURRENCE)
)
```

## Hub gene ORA

```{r hub-ora}
ora_hub <- run_gene_ora(
  foreground              = hub_genes,
  background              = background_genes,
  gene_set_list           = gene_set_list,
  gene_set_source_lookup  = gene_set_source_lookup,
  analysis_label          = sprintf("Hub genes (>= %d metabolites)", MIN_METAB_RECURRENCE)
)

if (!is.null(ora_hub) && sum(ora_hub$pval < 0.05) > 0) {
  p_ora_hub <- plot_ora_dotplot(ora_hub,
    sprintf("Hub genes (>= %d distinct metabolites)", MIN_METAB_RECURRENCE))
  if (!is.null(p_ora_hub)) {
    save_dual_format(p_ora_hub, out_dir, "ora_hub_genes", width = 10, height = 8)
    print(p_ora_hub)
  }
} else {
  cat("No nominally significant ORA results for hub genes.\n")
}
```

## Hub gene GSEA (ranked by metabolite recurrence)

Ranking = number of distinct metabolites per gene (not p-value). Tests which
pathways are enriched for metabolically broad genes.

```{r hub-gsea}
hub_ranking <- tibble(gene = background_genes) |>
  left_join(hub_recur, by = "gene") |>
  mutate(min_P = 10^(-replace_na(n_metabolites, 0)))  # synthetic P for fgsea interface

# Use the existing run_gene_gsea with this recurrence-derived ranking
gsea_hub <- run_gene_gsea(
  ranking_df       = hub_ranking,
  background_genes = background_genes,
  gene_set_list    = gene_set_list,
  analysis_label   = "GSEA: metabolite recurrence ranking"
)

if (!is.null(gsea_hub) && sum(gsea_hub$padj < 0.05) > 0) {
  p_gsea_hub <- plot_gsea_dotplot(gsea_hub, gene_set_source_lookup,
    "GSEA: pathways enriched for metabolic hub genes")
  if (!is.null(p_gsea_hub)) {
    save_dual_format(p_gsea_hub, out_dir, "gsea_hub_genes", width = 10, height = 8)
    print(p_gsea_hub)
  }
} else {
  cat("No FDR-significant GSEA results for hub gene ranking.\n")
  # Show top nominal results anyway
  if (!is.null(gsea_hub) && nrow(gsea_hub) > 0) {
    cat("\nTop 10 nominal results:\n")
    gsea_hub |>
      head(10) |>
      dplyr::select(pathway, NES, pval, padj, size) |>
      knitr::kable(digits = 4)
  }
}
```

---

# Save Results

```{r save-results}

# All cell types
if (!is.null(ora_all))  write_csv(ora_all,                       file.path(csv_path, "ora_all_celltypes.csv"))
if (!is.null(gsea_all)) write_csv(flatten_leading_edge(gsea_all), file.path(csv_path, "gsea_all_celltypes.csv"))

# Cell-type-specific
for (ct in names(ora_ct_list)) {
  ct_safe <- gsub("[^A-Za-z0-9_]", "_", ct)
  if (!is.null(ora_ct_list[[ct]]))
    write_csv(ora_ct_list[[ct]], file.path(csv_path, paste0("ora_ct_", ct_safe, ".csv")))
}
for (ct in names(gsea_ct_list)) {
  ct_safe <- gsub("[^A-Za-z0-9_]", "_", ct)
  if (!is.null(gsea_ct_list[[ct]]))
    write_csv(flatten_leading_edge(gsea_ct_list[[ct]]),
              file.path(csv_path, paste0("gsea_ct_", ct_safe, ".csv")))
}
if (exists("ct_ora_summary")  && nrow(ct_ora_summary)  > 0)
  write_csv(ct_ora_summary,  file.path(csv_path, "ora_celltype_summary.csv"))
if (exists("ct_gsea_summary") && nrow(ct_gsea_summary) > 0)
  write_csv(ct_gsea_summary, file.path(csv_path, "gsea_celltype_summary.csv"))

# Glia vs. neurons
if (!is.null(ora_glia))        write_csv(ora_glia,                           file.path(csv_path, "ora_glia.csv"))
if (!is.null(ora_neuron))      write_csv(ora_neuron,                         file.path(csv_path, "ora_neuron.csv"))
if (!is.null(ora_glia_only))   write_csv(ora_glia_only,                      file.path(csv_path, "ora_glia_only.csv"))
if (!is.null(ora_neuron_only)) write_csv(ora_neuron_only,                    file.path(csv_path, "ora_neuron_only.csv"))
if (!is.null(gsea_glia))       write_csv(flatten_leading_edge(gsea_glia),    file.path(csv_path, "gsea_glia.csv"))
if (!is.null(gsea_neuron))     write_csv(flatten_leading_edge(gsea_neuron),  file.path(csv_path, "gsea_neuron.csv"))

# Hub genes
if (!is.null(ora_hub))  write_csv(ora_hub,  file.path(csv_path, "ora_hub_genes.csv"))
if (!is.null(gsea_hub)) write_csv(flatten_leading_edge(gsea_hub), file.path(csv_path, "gsea_hub_genes.csv"))

cat("All results saved to:", csv_path, "\n")
```

# AI Assistance Disclosure

Code in this script was developed with assistance from Claude (Anthropic). All
AI-generated code was reviewed, validated, and adapted by the author.

# Session Info

```{r session-info}
sessionInfo()
```
