---
title: "Metabolite–Gene Scatter Check (Top 20)"
subtitle: "Sneha Chaturvedi"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    scrollable: true
    theme: dark
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Purpose

First-pass QC on the top-20 pairs, three panels per pair:
**raw** (`log2(x+1)`, all samples — outliers still present), **outlier-trimmed**
(the `zscore_trim` predictor; trimmed samples drop out, so the panel shows the
modeled set), and **outlier-trimmed adjusted** (partial residual on the trimmed
subset). Comparing panel 1 vs 2 shows exactly which samples the IQR trim removes
and how the slope responds. The adjusted slope is the `lm` approximation of the
modeled effect (close to, not identical to, limma's precision-weighted logFC).

## Setup — rebuild from cached objects

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(tidyverse); library(edgeR); library(patchwork)
})
source("pseudobulk_functions.R")     # merge_hormone_metadata, save_dual_format
source("metabolite_functions.R")     # build_metab_covariates, transforms, mapping
set.seed(123)

# METHOD sets the middle/right-panel transform. "zscore_trim" = IQR outlier
# removal; keep "int" to view the INT run instead.
METHOD    <- "zscore_trim"           # "int" | "zscore" | "zscore_trim"
suffix    <- if (METHOD == "int") "" else paste0("-", METHOD)

cache_dir    <- "../../data/cache"
metab_path   <- "../../results/untargeted/batch2_peak_area_clean.csv"
hormone_xlsx <- "../../doc/targeted/targeted_hormones.xlsx"
pairs_csv    <- "../../results/gene-metabolite/csv/top20_pairs.csv"   # pairs to plot
out_dir      <- paste0("../../results/gene-metabolite/scatter-plots", suffix)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

MIN_CELLS <- 30
```

```{r}
#| label: load
pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))

targeted_comparison_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`

pb <- build_metab_covariates(pb_base, targeted_comparison_raw, batch_props,
                             min_cells = MIN_CELLS)

metab_mat <- read_metabolite_matrix(metab_path)
metab_tx  <- transform_metabolite_matrix(metab_mat, method = METHOD)   # trimmed predictor
# raw-scale value for panel 1: log2(x+1) -> min/5 impute, NO trimming (all samples)
metab_log <- t(apply(log2(metab_mat + 1), 1, min_value_impute))
dimnames(metab_log) <- dimnames(metab_mat)

pairs <- readr::read_csv(pairs_csv, show_col_types = FALSE)

rn        <- rownames(pb$counts)
row_ct    <- sub("^[^.]+\\.(.*)$",  "\\1", rn)
metab_col <- metab_sample_to_pb(colnames(metab_tx), sub("^([^.]+)\\..*$", "\\1", rn))

cc_col <- intersect(c("cell_count", "n_cells", "ncells"), colnames(pb$metadata))[1]
```

## Plot builder

```{r}
#| label: builder
make_pair_plot <- function(cid, gene, ct, meta_name) {
  idx     <- which(row_ct == ct)
  pos     <- match(metab_col[idx], colnames(metab_tx))
  present <- !is.na(pos) & gene %in% colnames(pb$counts)   # metabolite measured/imputed
  rows_use <- idx[present]; pos_p <- pos[present]
  if (length(rows_use) < 3) return(NULL)

  csub <- t(pb$counts[rows_use, , drop = FALSE])           # genes x samples
  expr <- as.numeric(log1p(edgeR::cpm(csub, log = FALSE)[gene, ]))

  md <- pb$metadata[rn[rows_use], ]
  df <- tibble(
    metab_raw  = as.numeric(metab_log[cid, pos_p]),        # all present samples
    metab_trim = as.numeric(metab_tx[cid,  pos_p]),        # NA = outlier trimmed out
    expr       = expr,
    cell_count = if (!is.na(cc_col)) md[[cc_col]] else NA_real_,
    Sex        = md$Sex,
    GW         = md$GW,
    batch1_frac = md$batch1_frac
  )

  base_layers <- list(
    geom_point(size = 2, alpha = 0.8),
    geom_smooth(method = "lm", se = TRUE),
    scale_color_viridis_c(),
    scale_shape_manual(values = c("M" = 16, "F" = 17)),
    theme_minimal(base_size = 13)
  )

  # PANEL 1: raw log2(x+1), ALL samples -- outliers still present
  p_raw <- ggplot(df, aes(metab_raw, expr, color = cell_count, shape = Sex)) +
    base_layers + labs(title = "Raw (log2, all samples)", x = meta_name, y = "log1p(CPM)")

  # PANEL 2: outlier-trimmed predictor -- trimmed samples drop out (NA)
  dft <- df |> filter(!is.na(metab_trim))
  p_mid <- ggplot(dft, aes(metab_trim, expr, color = cell_count, shape = Sex)) +
    base_layers + labs(title = "Outlier-trimmed",
                       x = paste0(meta_name, " (trimmed z)"), y = "log1p(CPM)")

  # PANEL 3: partial residual on the trimmed subset + that group's covariates
  if (nrow(dft) < 4 || length(unique(dft$metab_trim)) < 2) {
    combined <- (p_raw | p_mid)
  } else {
    covs  <- c("GW", "Sex", "batch1_frac")
    keepc <- covs[vapply(covs, function(cv) length(unique(dft[[cv]])) >= 2, logical(1))]
    fit   <- lm(reformulate(c("metab_trim", keepc), response = "expr"), data = dft)
    dft$part <- resid(fit) + coef(fit)[["metab_trim"]] * dft$metab_trim
    p_adj <- ggplot(dft, aes(metab_trim, part, color = cell_count, shape = Sex)) +
      base_layers +
      labs(title = "Outlier-trimmed (adjusted)",
           x = paste0(meta_name, " (trimmed z)"), y = "log1p(CPM), adjusted")
    combined <- (p_raw | p_mid | p_adj)
  }

  combined +
    plot_annotation(title = paste0(gene, "  ~  ", meta_name, "   (", ct, ")")) +
    plot_layout(guides = "collect")
}
```

## Generate one file per pair

```{r}
#| label: generate
#| results: asis
for (i in seq_len(nrow(pairs))) {
  r  <- pairs[i, ]
  pl <- make_pair_plot(r$Compound.ID, r$gene, r$cell_type, r$Name)
  if (is.null(pl)) { message("Skipping ", r$gene, " / ", r$Name); next }

  base <- gsub("[^A-Za-z0-9]+", "-",
               paste(r$Compound.ID, r$gene, r$cell_type, sep = "_"))
  save_dual_format(pl, out_dir, base, width = 16, height = 5)
  print(pl)
}
cat("\nSaved", nrow(pairs), "scatter pairs to", out_dir, "\n")
```
