---
title: "Metabolite–Gene Scatter Check (Top 20)"
subtitle: "Sneha Chaturvedi"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    scrollable: true
    theme: dark
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Purpose

First-pass QC on the top-20 global-FDR metabolite–gene associations: for each
pair, a side-by-side scatter of **raw** expression vs. the metabolite (left) and
the **covariate-adjusted** partial-residual view (right). The right panel's slope
is the `lm` approximation of the modeled effect; because limma fit
precision-weighted counts, it is close to but not identical to the reported
logFC. Reads the shared `top20_pairs.csv` so the figures match `check_hit_nonzero.R`.

## Setup — rebuild from cached objects

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(tidyverse); library(edgeR); library(patchwork)
})
source("pseudobulk_functions.R")     # merge_hormone_metadata, save_dual_format
source("metabolite_functions.R")     # build_metab_covariates, metabolite transform helpers
set.seed(123)

cache_dir    <- "../../data/cache"
metab_path   <- "../../results/untargeted/batch2_peak_area_clean.csv"
hormone_xlsx <- "../../doc/targeted/targeted_hormones.xlsx"
pairs_csv    <- "../../results/gene-metabolite/csv/top20_pairs.csv"
out_dir      <- "../../results/gene-metabolite/scatter-plots"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

MIN_CELLS <- 30
```

```{r}
#| label: load
pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))

targeted_comparison_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`

pb <- build_metab_covariates(pb_base, targeted_comparison_raw, batch_props,
                             min_cells = MIN_CELLS)

metab_mat <- read_metabolite_matrix(metab_path)
metab_z   <- transform_metabolite_matrix(metab_mat)
# pre-standardization value (log2(x+1) -> min/5 impute), for the raw-scale panel
metab_log <- t(apply(log2(metab_mat + 1), 1, min_value_impute))
dimnames(metab_log) <- dimnames(metab_mat)

pairs <- readr::read_csv(pairs_csv, show_col_types = FALSE)

# per-row keys for pb$counts (rownames = "sample_id.cell_type")
rn        <- rownames(pb$counts)
row_samp  <- sub("^([^.]+)\\..*$", "\\1", rn)
row_ct    <- sub("^[^.]+\\.(.*)$",  "\\1", rn)
metab_col <- metab_sample_to_pb(colnames(metab_z), row_samp)

# cell_count column name can vary across pipeline versions; resolve once
cc_col <- intersect(c("cell_count", "n_cells", "ncells"), colnames(pb$metadata))[1]
```

## Plot builder

```{r}
#| label: builder
make_pair_plot <- function(cid, gene, ct, meta_name) {
  idx  <- which(row_ct == ct)
  mcol <- metab_col[idx]
  pos  <- match(mcol, colnames(metab_z))
  mval <- metab_z[cid, pos]                                 # standardised metabolite
  keep <- !is.na(mval) & gene %in% colnames(pb$counts)
  rows_use <- idx[keep]
  if (length(rows_use) < 3) return(NULL)

  # log1p(CPM) within the modeled subset (matches plot_top_genes_from_subsets)
  csub <- t(pb$counts[rows_use, , drop = FALSE])            # genes x samples
  expr <- as.numeric(log1p(edgeR::cpm(csub, log = FALSE)[gene, ]))

  md <- pb$metadata[rn[rows_use], ]
  df <- tibble(
    metab      = as.numeric(mval[keep]),                    # z-scored (model scale)
    metab_raw  = as.numeric(metab_log[cid, pos[keep]]),     # log2(x+1), pre-z
    expr       = expr,
    cell_count = if (!is.na(cc_col)) md[[cc_col]] else NA_real_,
    Sex        = md$Sex,
    GW         = md$GW,
    batch1_frac = md$batch1_frac
  )

  base_layers <- list(
    geom_point(size = 2, alpha = 0.8),
    geom_smooth(method = "lm", se = TRUE),
    scale_color_viridis_c(),
    scale_shape_manual(values = c("M" = 16, "F" = 17)),
    theme_minimal(base_size = 13)
  )

  # PANEL 1: raw metabolite scale, log2(x+1) -- shows the true x-distribution
  #          (a lone low-abundance sample sits far out here, revealing leverage)
  p_raw <- ggplot(df, aes(metab_raw, expr, color = cell_count, shape = Sex)) +
    base_layers + labs(title = "Metabolite log2(x+1)", x = meta_name, y = "log1p(CPM)")

  # PANEL 2: INT-transformed metabolite (model scale), marginal
  p_mid <- ggplot(df, aes(metab, expr, color = cell_count, shape = Sex)) +
    base_layers + labs(title = "INT (raw expr)",
                       x = paste0(meta_name, " (INT)"), y = "log1p(CPM)")

  # PANEL 3: partial residual on the covariates this group actually modeled
  covs  <- c("GW", "Sex", "batch1_frac")
  keepc <- covs[vapply(covs, function(cv) length(unique(df[[cv]])) >= 2, logical(1))]
  fit   <- lm(reformulate(c("metab", keepc), response = "expr"), data = df)
  df$part <- resid(fit) + coef(fit)[["metab"]] * df$metab

  p_adj <- ggplot(df, aes(metab, part, color = cell_count, shape = Sex)) +
    base_layers +
    labs(title = "INT (adjusted)",
         x = paste0(meta_name, " (INT)"), y = "log1p(CPM), adjusted")

  (p_raw | p_mid | p_adj) +
    plot_annotation(title = paste0(gene, "  ~  ", meta_name, "   (", ct, ")")) +
    plot_layout(guides = "collect")
}
```

## Generate one file per pair

```{r}
#| label: generate
#| results: asis
for (i in seq_len(nrow(pairs))) {
  r  <- pairs[i, ]
  pl <- make_pair_plot(r$Compound.ID, r$gene, r$cell_type, r$Name)
  if (is.null(pl)) { message("Skipping ", r$gene, " / ", r$Name); next }

  base <- gsub("[^A-Za-z0-9]+", "-",
               paste(r$Compound.ID, r$gene, r$cell_type, sep = "_"))
  save_dual_format(pl, out_dir, base, width = 16, height = 5)
  print(pl)
}
cat("\nSaved", nrow(pairs), "scatter pairs to", out_dir, "\n")
```
