---
title: "Metabolite–Gene Imputation Sensitivity"
subtitle: "Sneha Chaturvedi"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    scrollable: true
    theme: dark
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Purpose

Diagnose whether global-FDR hits are driven by imputed metabolite values.
Imputation: originally-missing metabolites (NaN in raw peak area) receive
`min(log2 values)/5` before z-scoring, placing them as a cluster at the low
end of the predictor. Two-to-three such points could inflate an association
without being caught by DFBETAS (which only flags single-sample influence).

This document:

1. Annotates scatter plots with a **red ring** on imputed samples.
2. Re-fits each association dropping all imputed samples; flags pairs where
   significance is lost (`p > 0.05`) or direction reverses — the
   **imputation sensitivity** test.
3. Cross-tabulates imputation sensitivity with DFBETAS stability to show overlap.
4. Produces a **clean hit set**: global-FDR hits that are imputation-insensitive
   (primary analysis), with DFBETAS stability retained as an annotation column.
5. Summary recurrence figures (top genes, metabolites, cell types) on the clean set.

## Setup

```{r}
#| label: setup
Packages <- c("tidyverse", "arrow", "cowplot", "svglite", "edgeR", "patchwork")
lapply(Packages, library, character.only = TRUE)
source("pseudobulk_functions.R")
source("metabolite_functions.R")
set.seed(123)

METHOD <- "zscore_trim"
suffix <- if (METHOD == "int") "" else paste0("-", METHOD)

parquet_dir  <- sprintf("../../results/gene-metabolite/parquet-%s", METHOD)
annot_path   <- "../../results/untargeted/batch2_metabolite_annotations_flagged.csv"
dfbetas_path <- file.path(
  "../../results/gene-metabolite",
  paste0("stability-dfbetas", suffix),
  "metabolite_gene_hits_dfbetas_stable.csv"
)

out_dir     <- "../../results/gene-metabolite"
iter_dir    <- file.path(out_dir, paste0("stability-imputation", suffix))
csv_dir     <- iter_dir
fig_dir     <- iter_dir
scatter_dir <- file.path(iter_dir, "scatter-plots")
for (d in c(iter_dir, scatter_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

FDR_GLOBAL  <- 0.10
LFC_MIN     <- 0.25
LABEL_CHARS <- 30
# Minimum non-imputed samples required to attempt the drop-imputed re-fit.
# Below this, the pair is flagged sensitive conservatively.
MIN_OBS <- 5

add_plot_label <- function(df, name_col = "Name", n = LABEL_CHARS) {
  short <- ifelse(nchar(df[[name_col]]) > n,
                  paste0(substr(df[[name_col]], 1, n), "\u2026"), df[[name_col]])
  idx <- stats::ave(seq_along(short), short, FUN = seq_along)
  short[idx > 1] <- paste0(short[idx > 1], "#", idx[idx > 1])
  df$label <- short
  df
}

bh_global <- function(p, m) {
  o <- order(p); ro <- order(o)
  adj <- p[o] * m / seq_along(p[o])
  adj <- rev(cummin(rev(adj)))
  pmin(adj[ro], 1)
}
```

## Load data

```{r}
#| label: load-data
cache_dir    <- "../../data/cache"
metab_path   <- "../../results/untargeted/batch2_peak_area_clean.csv"
hormone_xlsx <- "../../doc/targeted/targeted_hormones.xlsx"

pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))

targeted_comparison_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`

pb        <- build_metab_covariates(pb_base, targeted_comparison_raw, batch_props, min_cells = 30)
metab_mat <- read_metabolite_matrix(metab_path)
metab_tx  <- transform_metabolite_matrix(metab_mat, method = METHOD)
metab_log <- t(apply(log2(metab_mat + 1), 1, min_value_impute))
dimnames(metab_log) <- dimnames(metab_mat)

# Imputation mask: TRUE = compound was absent (NaN) in the raw peak area matrix.
# These positions receive min(log2)/5 before z-scoring and appear as a low
# cluster in the transformed predictor. is.na() catches both NA and NaN.
imputed_mask <- is.na(metab_mat)   # features x samples (all metabolite columns)

rn        <- rownames(pb$counts)
row_ct    <- sub("^[^.]+\\.(.*)$", "\\1", rn)
metab_col <- metab_sample_to_pb(colnames(metab_tx), sub("^([^.]+)\\..*$", "\\1", rn))
cc_col    <- intersect(c("cell_count", "n_cells", "ncells"), colnames(pb$metadata))[1]
```

## Global-FDR hits (pre-DFBETAS)

```{r}
#| label: hits
ds      <- open_dataset(parquet_dir)
n_tests <- nrow(ds)

annot <- read_csv(annot_path, show_col_types = FALSE) |>
  dplyr::select(Compound.ID, Super.Class, Class, Sub.Class,
                Pathways, KEGG, Exogenous_Flag, Annotation_Confidence)

hits <- ds |>
  dplyr::filter(P.Value < 0.05) |>
  collect() |>
  mutate(FDR_global = bh_global(P.Value, n_tests)) |>
  dplyr::filter(FDR_global < FDR_GLOBAL, abs(logFC) >= LFC_MIN) |>
  dplyr::left_join(annot, by = "Compound.ID")

dfbetas_stable <- read_csv(dfbetas_path, show_col_types = FALSE)

cat(sprintf("Global-FDR hits (pre-DFBETAS): %s\nDFBETAS-stable hits:           %s\n",
            format(nrow(hits),           big.mark = ","),
            format(nrow(dfbetas_stable), big.mark = ",")))
```

## Imputation sensitivity test

A pair is **imputation-sensitive** if, after dropping all imputed samples for
that metabolite in that cell type, either:

- the **slope direction reverses** (sign of `metab_trim` coefficient flips), or
- the association **loses nominal significance** (`p > 0.05`).

Pairs with fewer than `r MIN_OBS` non-imputed samples after the drop are flagged
`too_few_observed = TRUE` and treated as sensitive (conservative).

The re-fit uses `lm` on `log1p(CPM)` — the same response/predictor as the
DFBETAS check — rather than voom, which is appropriate for a sensitivity
diagnostic (not the primary model).

```{r}
#| label: imputation-sensitivity
imputation_sensitivity_group <- function(cid, ct, genes) {
  idx  <- which(row_ct == ct)
  pos  <- match(metab_col[idx], colnames(metab_tx))
  keep <- !is.na(pos)
  rows_use <- idx[keep]
  pos_p    <- pos[keep]

  mv      <- as.numeric(metab_tx[cid, pos_p])
  obs_ok  <- !is.na(mv)           # NA in metab_tx = IQR-trimmed outlier; drop these
  imp_flg <- as.logical(imputed_mask[cid, pos_p])

  rows_use <- rows_use[obs_ok]
  imp_flg  <- imp_flg[obs_ok]
  mv       <- mv[obs_ok]

  if (length(rows_use) < 3)
    return(tibble(gene = genes, n_imputed = NA_integer_, n_observed = NA_integer_,
                  too_few_observed = NA, slope_full = NA_real_, p_full = NA_real_,
                  slope_noimpute = NA_real_, p_noimpute = NA_real_,
                  direction_same = NA, sig_noimpute = NA, imputation_sensitive = NA))

  md  <- pb$metadata[rn[rows_use], ]
  lib <- rowSums(pb$counts[rows_use, , drop = FALSE])

  base <- tibble(metab_trim  = mv,
                 imputed     = imp_flg,
                 GW          = md$GW,
                 Sex         = md$Sex,
                 batch1_frac = md$batch1_frac)

  covs  <- c("GW", "Sex", "batch1_frac")
  keepc <- covs[vapply(covs, function(cv) length(unique(base[[cv]])) >= 2, logical(1))]
  fml   <- reformulate(c("metab_trim", keepc), response = "expr")

  n_imp <- sum(imp_flg)
  n_obs <- sum(!imp_flg)

  purrr::map_dfr(genes, function(g) {
    empty <- tibble(gene = g, n_imputed = n_imp, n_observed = n_obs,
                    too_few_observed = NA, slope_full = NA_real_, p_full = NA_real_,
                    slope_noimpute = NA_real_, p_noimpute = NA_real_,
                    direction_same = NA, sig_noimpute = NA, imputation_sensitive = NA)

    if (!g %in% colnames(pb$counts)) return(empty)

    df      <- base
    df$expr <- log1p(pb$counts[rows_use, g] / lib * 1e6)

    # ---- Full model (all samples, imputed included) ----
    fit_full <- tryCatch(lm(fml, data = df), error = function(e) NULL)
    if (is.null(fit_full)) return(empty)

    coef_full  <- summary(fit_full)$coefficients
    slope_full <- coef_full["metab_trim", "Estimate"]
    p_full     <- coef_full["metab_trim", "Pr(>|t|)"]

    # ---- Drop-imputed model ----
    df_obs  <- dplyr::filter(df, !imputed)
    too_few <- nrow(df_obs) < MIN_OBS || length(unique(df_obs$metab_trim)) < 2

    if (too_few) {
      return(tibble(gene = g, n_imputed = n_imp, n_observed = n_obs,
                    too_few_observed = TRUE,
                    slope_full = slope_full, p_full = p_full,
                    slope_noimpute = NA_real_, p_noimpute = NA_real_,
                    direction_same = NA, sig_noimpute = NA,
                    imputation_sensitive = TRUE))
    }

    # Covariates may become single-level after removing imputed rows
    keepc_obs <- covs[vapply(covs, function(cv) length(unique(df_obs[[cv]])) >= 2, logical(1))]
    fml_obs   <- reformulate(c("metab_trim", keepc_obs), response = "expr")
    fit_obs   <- tryCatch(lm(fml_obs, data = df_obs), error = function(e) NULL)

    if (is.null(fit_obs)) {
      return(tibble(gene = g, n_imputed = n_imp, n_observed = n_obs,
                    too_few_observed = FALSE,
                    slope_full = slope_full, p_full = p_full,
                    slope_noimpute = NA_real_, p_noimpute = NA_real_,
                    direction_same = NA, sig_noimpute = NA,
                    imputation_sensitive = TRUE))
    }

    coef_obs  <- summary(fit_obs)$coefficients
    slope_obs <- coef_obs["metab_trim", "Estimate"]
    p_obs     <- coef_obs["metab_trim", "Pr(>|t|)"]
    dir_same  <- sign(slope_full) == sign(slope_obs)
    sig_obs   <- p_obs < 0.05

    tibble(gene = g, n_imputed = n_imp, n_observed = n_obs,
           too_few_observed = FALSE,
           slope_full = slope_full, p_full = p_full,
           slope_noimpute = slope_obs, p_noimpute = p_obs,
           direction_same = dir_same, sig_noimpute = sig_obs,
           imputation_sensitive = !dir_same | !sig_obs)
  })
}

reporter_imp <- hits |>
  dplyr::group_by(Compound.ID, cell_type) |>
  dplyr::group_modify(~ imputation_sensitivity_group(
    .y$Compound.ID, .y$cell_type, .x$gene
  )) |>
  dplyr::ungroup()

hits <- hits |>
  dplyr::left_join(reporter_imp, by = c("Compound.ID", "cell_type", "gene"))

cat(sprintf(
  "Imputation-insensitive (direction held & p<0.05 without imputed): %s / %s (%.1f%%)\n",
  format(sum(!hits$imputation_sensitive, na.rm = TRUE), big.mark = ","),
  format(nrow(hits), big.mark = ","),
  100 * mean(!hits$imputation_sensitive, na.rm = TRUE)
))
write_csv(hits, file.path(csv_dir, "metabolite_gene_hits_imputation_annotated.csv"))
```

## Sensitivity gate summary

```{r}
#| label: summary-bar
counts <- hits |>
  mutate(status = dplyr::case_when(
    too_few_observed %in% TRUE     ~ "too few observed",
    imputation_sensitive %in% TRUE ~ "imputation-sensitive",
    TRUE                           ~ "insensitive"
  )) |>
  dplyr::count(status)

p_cnt <- ggplot(counts, aes(status, n, fill = status)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.3) +
  scale_fill_manual(values = c(
    insensitive              = "steelblue",
    `imputation-sensitive`   = "firebrick",
    `too few observed`       = "grey60"
  )) +
  labs(x = NULL, y = "Hits", title = "Imputation sensitivity gate") +
  theme_cowplot(13)
save_dual_format(p_cnt, fig_dir, "imputation_gate_counts", width = 6, height = 5)
print(p_cnt)
```

## Why sensitive? (direction vs. significance)

```{r}
#| label: sensitivity-breakdown
sens <- hits |>
  dplyr::filter(imputation_sensitive %in% TRUE, !too_few_observed %in% TRUE)

cat(sprintf(
  "Direction flip only:        %d\nLost significance only:    %d\nBoth:                      %d\n",
  sum(!sens$direction_same &  sens$sig_noimpute,  na.rm = TRUE),
  sum( sens$direction_same & !sens$sig_noimpute,  na.rm = TRUE),
  sum(!sens$direction_same & !sens$sig_noimpute,  na.rm = TRUE)
))

# Scatter: full-model slope vs. drop-imputed slope, colored by sensitivity reason
p_slope <- hits |>
  dplyr::filter(!is.na(slope_full), !is.na(slope_noimpute)) |>
  mutate(reason = dplyr::case_when(
    !direction_same & !sig_noimpute ~ "flip + lost sig",
    !direction_same                 ~ "direction flip",
    !sig_noimpute                   ~ "lost significance",
    TRUE                            ~ "insensitive"
  )) |>
  ggplot(aes(slope_full, slope_noimpute, color = reason)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c(
    "insensitive"       = "steelblue",
    "direction flip"    = "firebrick",
    "lost significance" = "darkorange",
    "flip + lost sig"   = "purple"
  )) +
  labs(x = "Slope (full model)", y = "Slope (drop-imputed)",
       title = "Full vs. drop-imputed slope",
       color = "Sensitivity reason") +
  theme_cowplot(12)
save_dual_format(p_slope, fig_dir, "slope_comparison", width = 7, height = 6)
print(p_slope)
```

## n_imputed distribution

```{r}
#| label: imputed-counts
p_nimp <- hits |>
  dplyr::filter(!is.na(n_imputed)) |>
  mutate(sens_label = ifelse(imputation_sensitive %in% TRUE,
                             "sensitive", "insensitive")) |>
  ggplot(aes(n_imputed, fill = sens_label)) +
  geom_histogram(binwidth = 1, position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c(insensitive = "steelblue",
                               sensitive   = "firebrick")) +
  labs(x = "Number of imputed samples per pair",
       y = "Pairs", fill = NULL,
       title = "Imputed-sample count vs. sensitivity status") +
  theme_cowplot(12)
save_dual_format(p_nimp, fig_dir, "n_imputed_histogram", width = 7, height = 5)
print(p_nimp)
```

## Overlap with DFBETAS instability

```{r}
#| label: overlap
dfbetas_key <- dfbetas_stable |>
  dplyr::select(Compound.ID, cell_type, gene) |>
  dplyr::mutate(dfbetas_stable = TRUE)

hits_overlap <- hits |>
  dplyr::left_join(dfbetas_key, by = c("Compound.ID", "cell_type", "gene")) |>
  dplyr::mutate(
    dfbetas_stable  = tidyr::replace_na(dfbetas_stable, FALSE),
    imp_insensitive = !imputation_sensitive %in% TRUE & !too_few_observed %in% TRUE
  )

ov <- hits_overlap |>
  dplyr::count(dfbetas_stable, imp_insensitive) |>
  mutate(
    x_lab = factor(dfbetas_stable,  levels = c(FALSE, TRUE),
                   labels = c("DFBETAS\nunstable", "DFBETAS\nstable")),
    y_lab = factor(imp_insensitive, levels = c(FALSE, TRUE),
                   labels = c("Imputation\nsensitive", "Imputation\ninsensitive"))
  )

cat("Overlap table (n = pairs):\n")
print(dplyr::select(ov, dfbetas_stable, imp_insensitive, n))

p_ov <- ggplot(ov, aes(x_lab, y_lab, fill = n, label = scales::comma(n))) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(size = 6, color = "white", fontface = "bold") +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "Pairs") +
  labs(x = NULL, y = NULL, title = "DFBETAS \u00d7 Imputation sensitivity overlap") +
  theme_cowplot(13) + theme(legend.position = "right")
save_dual_format(p_ov, fig_dir, "overlap_dfbetas_imputation", width = 6, height = 5)
print(p_ov)
```

## Clean hit set (imputation-insensitive)

```{r}
#| label: clean-hits
# Clean set = imputation-insensitive only.
# dfbetas_stable is retained as a column so scatter annotations show which
# pairs also pass the DFBETAS gate, enabling visual comparison.
clean_hits <- hits_overlap |>
  dplyr::filter(imp_insensitive)

write_csv(clean_hits, file.path(csv_dir, "metabolite_gene_hits_clean.csv"))

cat(sprintf(
  "Global-FDR hits:                      %s\nDFBETAS-stable:                       %s\nImputation-insensitive (clean set):   %s\n  of which also DFBETAS-stable:        %s\n  of which DFBETAS-unstable:           %s\n",
  format(nrow(hits),                              big.mark = ","),
  format(sum(hits_overlap$dfbetas_stable),        big.mark = ","),
  format(nrow(clean_hits),                        big.mark = ","),
  format(sum(clean_hits$dfbetas_stable),          big.mark = ","),
  format(sum(!clean_hits$dfbetas_stable),         big.mark = ",")
))
```

## Recurrence on clean hits

```{r}
#| label: recurrence
gene_recur <- clean_hits |>
  dplyr::group_by(gene) |>
  dplyr::summarise(n_metabolites = n_distinct(Compound.ID), .groups = "drop") |>
  dplyr::slice_max(n_metabolites, n = 20)

p_g <- ggplot(gene_recur, aes(reorder(gene, n_metabolites), n_metabolites)) +
  geom_col(fill = "steelblue", alpha = 0.85) + coord_flip() +
  labs(x = "Gene", y = "Distinct metabolites",
       title = "Clean hits: top genes by metabolite associations") +
  theme_cowplot(12)
save_dual_format(p_g, fig_dir, "clean_recurrence_genes", width = 8, height = 7)
print(p_g)

metab_recur <- clean_hits |>
  dplyr::group_by(Compound.ID, Name, Super.Class) |>
  dplyr::summarise(n_genes = n_distinct(gene), .groups = "drop") |>
  dplyr::slice_max(n_genes, n = 25) |>
  add_plot_label()
write_csv(metab_recur |> dplyr::select(label, Name, Compound.ID),
          file.path(fig_dir, "clean_recurrence_metabolites_labels.csv"))

p_m <- ggplot(metab_recur, aes(reorder(label, n_genes), n_genes, fill = Super.Class)) +
  geom_col(alpha = 0.85) + coord_flip() +
  labs(x = "Metabolite", y = "Distinct genes",
       title = "Clean hits: top metabolites by gene associations") +
  theme_cowplot(12)
save_dual_format(p_m, fig_dir, "clean_recurrence_metabolites", width = 10, height = 7)
print(p_m)

ct_recur <- clean_hits |>
  dplyr::group_by(cell_type) |>
  dplyr::summarise(n_metabolites = n_distinct(Compound.ID), .groups = "drop")

p_c <- ggplot(ct_recur, aes(reorder(cell_type, n_metabolites), n_metabolites)) +
  geom_col(fill = "darkorange", alpha = 0.85) + coord_flip() +
  labs(x = "Cell type", y = "Distinct metabolites",
       title = "Clean hits: metabolites per cell type") +
  theme_cowplot(12)
save_dual_format(p_c, fig_dir, "clean_recurrence_celltypes", width = 8, height = 6)
print(p_c)
```

## Scatter plots — top 20 clean hits with imputed labels

Imputed samples shown with a **red open ring** overlay. All three panels
(raw log2, outlier-trimmed z, covariate-adjusted) carry the same overlay so
you can see the imputed cluster's position before and after transform.

```{r}
#| label: scatter-builder
make_pair_plot <- function(cid, gene, ct, meta_name, dfbetas_stable = NA) {
  idx     <- which(row_ct == ct)
  pos     <- match(metab_col[idx], colnames(metab_tx))
  present <- !is.na(pos) & gene %in% colnames(pb$counts)
  if (!any(present)) return(NULL)
  rows_use <- idx[present]
  pos_p    <- pos[present]
  if (length(rows_use) < 3) return(NULL)

  expr    <- as.numeric(log1p(
    edgeR::cpm(t(pb$counts[rows_use, , drop = FALSE]), log = FALSE)[gene, ]
  ))
  md      <- pb$metadata[rn[rows_use], ]
  imp_flg <- as.logical(imputed_mask[cid, pos_p])

  df <- tibble(
    metab_raw   = as.numeric(metab_log[cid, pos_p]),
    metab_trim  = as.numeric(metab_tx[cid,  pos_p]),
    expr        = expr,
    cell_count  = if (!is.na(cc_col)) md[[cc_col]] else NA_real_,
    Sex         = md$Sex,
    GW          = md$GW,
    batch1_frac = md$batch1_frac,
    imputed     = imp_flg
  )

  n_imp <- sum(imp_flg)

  # Red open-ring overlay on imputed points (layered above filled points so
  # it's always visible regardless of cell_count color scale).
  imputed_overlay <- geom_point(
    data   = ~ dplyr::filter(.x, imputed == TRUE),
    shape  = 21, size = 4.5, color = "red", fill = NA, stroke = 1.5
  )

  bl <- list(
    geom_point(size = 2, alpha = 0.8),
    imputed_overlay,
    geom_smooth(method = "lm", se = TRUE),
    scale_color_viridis_c(),
    scale_shape_manual(values = c("M" = 16, "F" = 17)),
    theme_minimal(base_size = 13)
  )

  p_raw <- ggplot(df, aes(metab_raw, expr, color = cell_count, shape = Sex)) + bl +
    labs(title = "Raw (log2, all samples)",
         x     = meta_name,
         y     = "log1p(CPM)",
         caption = sprintf("%d of %d samples imputed (red ring)", n_imp, nrow(df)))

  dft <- dplyr::filter(df, !is.na(metab_trim))

  p_mid <- ggplot(dft, aes(metab_trim, expr, color = cell_count, shape = Sex)) + bl +
    labs(title = "Outlier-trimmed",
         x     = paste0(meta_name, " (trimmed z)"),
         y     = "log1p(CPM)")

  covs  <- c("GW", "Sex", "batch1_frac")
  keepc <- covs[vapply(covs, function(cv) length(unique(dft[[cv]])) >= 2, logical(1))]
  fit   <- lm(reformulate(c("metab_trim", keepc), response = "expr"), data = dft)
  dft$part <- resid(fit) + coef(fit)[["metab_trim"]] * dft$metab_trim

  p_adj <- ggplot(dft, aes(metab_trim, part, color = cell_count, shape = Sex)) + bl +
    labs(title = "Outlier-trimmed (adjusted)",
         x     = paste0(meta_name, " (trimmed z)"),
         y     = "log1p(CPM), adjusted")

  dfb_label <- dplyr::case_when(
    isTRUE(dfbetas_stable)  ~ "DFBETAS: stable",
    isFALSE(dfbetas_stable) ~ "DFBETAS: unstable",
    TRUE                    ~ "DFBETAS: unknown"
  )

  (p_raw | p_mid | p_adj) +
    plot_annotation(
      title    = paste0(gene, "  ~  ", meta_name, "   (", ct, ")"),
      subtitle = dfb_label
    ) +
    plot_layout(guides = "collect")
}
```

```{r}
#| label: scatter-generate
#| results: asis
top20 <- clean_hits |>
  dplyr::arrange(FDR_global, dplyr::desc(abs(logFC))) |>
  dplyr::distinct(Compound.ID, .keep_all = TRUE) |>
  dplyr::slice_head(n = 20)

for (i in seq_len(nrow(top20))) {
  r  <- top20[i, ]
  pl <- make_pair_plot(r$Compound.ID, r$gene, r$cell_type, r$Name,
                       dfbetas_stable = r$dfbetas_stable)
  if (is.null(pl)) next
  base_name <- gsub("[^A-Za-z0-9]+", "-",
                    paste(r$Compound.ID, r$gene, r$cell_type, sep = "_"))
  save_dual_format(pl, scatter_dir, base_name, width = 16, height = 5)
  print(pl)
}
cat("\nSaved", nrow(top20), "clean-hit scatters to", scatter_dir, "\n")
```
