---
title: "Metabolite Transform QC"
subtitle: "Comparing z-score, median normalization, and combined transforms"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    code-tools: true
    scrollable: true
    theme: dark
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Purpose

Three questions this QC answers:

1. **Is sample-level variation present?** Per-sample log2 distributions ordered by GW — do samples look systematically different in overall metabolite load?
2. **Is median normalization safe?** If the sample median correlates with GW or Sex, median normalization would flatten real biology before the model sees it.
3. **What does each transform do?** Global distributions and example metabolite–GW relationships under four transforms side-by-side.

## Setup

```{r}
#| label: setup
Packages <- c("tidyverse", "cowplot", "svglite", "patchwork")
lapply(Packages, library, character.only = TRUE)

source("pseudobulk_functions.R")
source("metabolite_functions.R")
set.seed(123)

metab_path   <- "../../results/untargeted/batch2_peak_area_clean.csv"
hormone_xlsx <- "../../doc/targeted/targeted_hormones.xlsx"
cache_dir    <- "../../data/cache"
stable_csv   <- "../../results/gene-metabolite/stability-dfbetas-zscore_trim/metabolite_gene_hits_dfbetas_stable.csv"
out_dir      <- "../../results/gene-metabolite/transform-qc"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data + compute transforms

```{r}
#| label: transforms
metab_mat <- read_metabolite_matrix(metab_path)

# Rebuild pb for GW/Sex metadata
pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))
targeted_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_raw$E2 <- targeted_raw$`17B-E2 (ng/mL)`
targeted_raw$TT <- targeted_raw$`TT (ng/mL)`
targeted_raw$P4 <- targeted_raw$`P4 (ng/mL)`
pb <- build_metab_covariates(pb_base, targeted_raw, batch_props, min_cells = 30)

# Map metabolite sample columns to GW/Sex via tissue number
# "Sample07" -> "tissue7" <-> "SSD07_tissue7" (mirrors metab_sample_to_pb logic)
pb_meta_unique <- pb$metadata |>
  tibble::rownames_to_column("pb_row") |>
  mutate(tissue_id = stringr::str_extract(pb_row, "tissue\\d+")) |>
  distinct(tissue_id, .keep_all = TRUE) |>
  dplyr::select(tissue_id, GW, Sex, batch1_frac)

metab_sample_meta <- tibble(
  metab_col = colnames(metab_mat),
  tissue_id = paste0("tissue", as.integer(sub("^Sample", "", colnames(metab_mat))))
) |>
  left_join(pb_meta_unique, by = "tissue_id")

# ---- Four transforms ----
# 1. log2 + imputed (baseline)
metab_log <- t(apply(log2(metab_mat + 1), 1, min_value_impute))
dimnames(metab_log) <- dimnames(metab_mat)

# 2. z-score per metabolite, IQR trim (current pipeline)
metab_zscore <- transform_metabolite_matrix(metab_mat, method = "zscore_trim")

# 3. Median normalization within sample on log2 scale (subtract per-sample median)
#    Corrects systematic differences in total metabolite load between samples.
#    Does NOT correct per-metabolite variance — z-score still needed for that.
metab_log2_raw <- log2(metab_mat + 1)
samp_med       <- apply(metab_log2_raw, 2, median, na.rm = TRUE)
metab_mednorm  <- sweep(metab_log2_raw, 2, samp_med, FUN = "-")
metab_mednorm_imp <- t(apply(metab_mednorm, 1, min_value_impute))
dimnames(metab_mednorm_imp) <- dimnames(metab_mat)

# 4. Median norm + z-score per metabolite (combined)
metab_mednorm_z <- t(apply(metab_mednorm_imp, 1, function(x) {
  s <- sd(x, na.rm = TRUE)
  if (is.na(s) || s == 0) return(rep(NA_real_, length(x)))
  (x - mean(x, na.rm = TRUE)) / s
}))
dimnames(metab_mednorm_z) <- dimnames(metab_mat)

cat(sprintf("Metabolites: %d | Samples in matrix: %d | Samples with GW/Sex: %d\n",
            nrow(metab_mat), ncol(metab_mat), sum(!is.na(metab_sample_meta$GW))))
```

## Per-sample distribution

Log2 peak areas per sample, ordered by gestational week. Each violin = one
sample; median line shown. Systematic shifts (e.g. widening IQR with GW) would
indicate sample-level variation worth correcting.

```{r}
#| label: per-sample-violin
#| fig-width: 14
#| fig-height: 6
sample_long <- metab_log2_raw |>
  as_tibble(rownames = "Compound.ID") |>
  pivot_longer(-Compound.ID, names_to = "metab_col", values_to = "log2_value") |>
  left_join(metab_sample_meta, by = "metab_col") |>
  filter(!is.na(GW), !is.na(log2_value), is.finite(log2_value))

# Order samples by GW then sample name for ties
sample_order <- metab_sample_meta |>
  filter(!is.na(GW)) |>
  arrange(GW, metab_col) |>
  pull(metab_col)

sample_long <- sample_long |>
  mutate(metab_col = factor(metab_col, levels = sample_order))

p_violin <- ggplot(sample_long, aes(metab_col, log2_value, fill = GW)) +
  geom_violin(scale = "width", alpha = 0.85, draw_quantiles = 0.5, linewidth = 0.3) +
  scale_fill_viridis_c(name = "GW") +
  labs(x = "Sample (ordered by GW)", y = "log2(peak area + 1)",
       title = "Per-sample distribution — log2 raw") +
  theme_cowplot(11) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))

save_dual_format(p_violin, out_dir, "per_sample_distribution", width = 14, height = 6)
print(p_violin)
```

## Sample median diagnostics

Key diagnostic: if the sample median (across all metabolites) correlates with
GW or Sex, median normalization would absorb biological signal before the model
sees it. GW and Sex are already covariates in the limma model, so a modest
correlation is tolerable — a strong one (|ρ| > 0.4) would be a red flag.

```{r}
#| label: median-diagnostics
#| fig-width: 14
#| fig-height: 6
samp_med_df <- tibble(
  metab_col   = colnames(metab_mat),
  log2_median = apply(metab_log2_raw, 2, median, na.rm = TRUE),
  log2_mean   = apply(metab_log2_raw, 2, mean,   na.rm = TRUE),
  log2_sd     = apply(metab_log2_raw, 2, sd,     na.rm = TRUE)
) |>
  left_join(metab_sample_meta, by = "metab_col") |>
  filter(!is.na(GW))

cor_gw  <- cor(samp_med_df$log2_median, samp_med_df$GW,
               use = "complete.obs", method = "spearman")
cor_sex <- cor(as.integer(factor(samp_med_df$Sex)), samp_med_df$log2_median,
               use = "complete.obs", method = "spearman")

p_med_gw <- ggplot(samp_med_df, aes(GW, log2_median, color = Sex, shape = Sex)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_smooth(method = "lm", se = TRUE, color = "white", linewidth = 0.8) +
  scale_color_manual(values = c("M" = "#4EC0E4", "F" = "#F88080")) +
  scale_shape_manual(values = c("M" = 16, "F" = 17)) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
           label = sprintf("Spearman ρ = %.2f", cor_gw), size = 4.5, color = "grey80") +
  labs(x = "Gestational Week", y = "Sample median [log2]",
       title = "Sample median ~ GW",
       subtitle = "Strong correlation → median norm removes biology") +
  theme_cowplot(12)

p_med_sex <- ggplot(samp_med_df, aes(Sex, log2_median, fill = Sex)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21, width = 0.5) +
  geom_jitter(width = 0.1, size = 2.5, alpha = 0.8, aes(color = Sex)) +
  scale_fill_manual(values  = c("M" = "#4EC0E4", "F" = "#F88080")) +
  scale_color_manual(values = c("M" = "#4EC0E4", "F" = "#F88080")) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
           label = sprintf("Spearman ρ = %.2f", cor_sex), size = 4.5, color = "grey80") +
  labs(x = "Sex", y = "Sample median [log2]",
       title = "Sample median ~ Sex") +
  theme_cowplot(12) +
  theme(legend.position = "none")

p_diag <- p_med_gw | p_med_sex
save_dual_format(p_diag, out_dir, "sample_median_diagnostics", width = 14, height = 6)
print(p_diag)
```

## Global value distributions

All four transforms across all metabolites × samples. The key comparisons:
`log2 raw` vs `median-norm` shows whether within-sample centering shifts the
population; `z-score` vs `median-norm + z-score` shows whether the final
predictor distribution changes after sample-level correction.

```{r}
#| label: global-distributions
#| fig-width: 12
#| fig-height: 8
make_long <- function(mat, label) {
  tibble(value = as.vector(mat), transform = label) |>
    filter(!is.na(value), is.finite(value))
}

dist_long <- bind_rows(
  make_long(metab_log,        "1. log2 + imputed"),
  make_long(metab_zscore,     "2. z-score, IQR-trim (current)"),
  make_long(metab_mednorm_imp,"3. median-norm + imputed"),
  make_long(metab_mednorm_z,  "4. median-norm + z-score")
) |>
  mutate(transform = factor(transform, levels = c(
    "1. log2 + imputed", "2. z-score, IQR-trim (current)",
    "3. median-norm + imputed", "4. median-norm + z-score"
  )))

p_dist <- ggplot(dist_long, aes(value, fill = transform, color = transform)) +
  geom_density(alpha = 0.25, linewidth = 0.7) +
  facet_wrap(~transform, scales = "free", ncol = 2) +
  labs(x = "Value", y = "Density",
       title = "Global value distributions — all metabolites × samples") +
  theme_cowplot(12) +
  theme(legend.position = "none")

save_dual_format(p_dist, out_dir, "global_distributions", width = 12, height = 8)
print(p_dist)
```

## Sample-level spread comparison

Box per sample, ordered by GW, under log2-raw vs median-norm. Medians should
align horizontally after median normalization. If the log2-raw medians are
already well-aligned, normalization adds little.

```{r}
#| label: sample-spread-comparison
#| fig-width: 14
#| fig-height: 8
spread_long <- bind_rows(
  as_tibble(metab_log,         rownames = "Compound.ID") |>
    pivot_longer(-Compound.ID, names_to = "metab_col", values_to = "value") |>
    mutate(transform = "log2 + imputed"),
  as_tibble(metab_mednorm_imp, rownames = "Compound.ID") |>
    pivot_longer(-Compound.ID, names_to = "metab_col", values_to = "value") |>
    mutate(transform = "median-norm + imputed")
) |>
  left_join(metab_sample_meta, by = "metab_col") |>
  filter(!is.na(GW), !is.na(value), is.finite(value)) |>
  mutate(metab_col = factor(metab_col, levels = sample_order))

p_spread <- ggplot(spread_long, aes(metab_col, value, fill = GW)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3, linewidth = 0.3) +
  scale_fill_viridis_c(name = "GW") +
  facet_wrap(~transform, ncol = 1, scales = "free_y") +
  labs(x = "Sample (ordered by GW)", y = "Value",
       title = "Per-sample spread: log2 vs median-normalized") +
  theme_cowplot(10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6))

save_dual_format(p_spread, out_dir, "sample_spread_comparison", width = 14, height = 8)
print(p_spread)
```

## Example metabolites — value ~ GW under each transform

Four metabolites from the top stable hits shown across all transforms. Each row
= one metabolite; each column = one transform. Look for whether the slope
direction and apparent relationship change across transforms.

```{r}
#| label: example-metabolites
#| fig-width: 16
#| fig-height: 12
# Use top stable hits if available; otherwise first 4 detected metabolites
if (file.exists(stable_csv)) {
  example_meta <- read_csv(stable_csv, show_col_types = FALSE) |>
    arrange(FDR_global) |>
    distinct(Compound.ID, Name) |>
    slice_head(n = 4)
} else {
  example_meta <- tibble(
    Compound.ID = rownames(metab_mat)[1:4],
    Name        = rownames(metab_mat)[1:4]
  )
  message("stable_csv not found — using first 4 metabolites as examples.")
}

example_long <- purrr::map_dfr(seq_len(nrow(example_meta)), function(i) {
  cid  <- example_meta$Compound.ID[i]
  name <- example_meta$Name[i]
  bind_rows(
    tibble(value = as.numeric(metab_log[cid,        colnames(metab_mat)]),
           transform = "1. log2 + imputed"),
    tibble(value = as.numeric(metab_zscore[cid,     colnames(metab_mat)]),
           transform = "2. z-score (current)"),
    tibble(value = as.numeric(metab_mednorm_imp[cid,colnames(metab_mat)]),
           transform = "3. median-norm"),
    tibble(value = as.numeric(metab_mednorm_z[cid,  colnames(metab_mat)]),
           transform = "4. median-norm + z-score")
  ) |>
    mutate(metab_col = rep(colnames(metab_mat), 4),
           Compound.ID = cid,
           Name_short  = ifelse(nchar(name) > 28, paste0(substr(name, 1, 28), "\u2026"), name))
}) |>
  left_join(metab_sample_meta, by = "metab_col") |>
  filter(!is.na(GW), !is.na(value), is.finite(value)) |>
  mutate(transform = factor(transform, levels = c(
    "1. log2 + imputed", "2. z-score (current)",
    "3. median-norm", "4. median-norm + z-score"
  )))

p_examples <- ggplot(example_long, aes(GW, value, color = Sex, shape = Sex)) +
  geom_point(size = 2, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = c("M" = "#4EC0E4", "F" = "#F88080")) +
  scale_shape_manual(values = c("M" = 16, "F" = 17)) +
  facet_grid(Name_short ~ transform,
             scales = "free_y",
             labeller = labeller(Name_short = label_wrap_gen(25))) +
  labs(x = "Gestational Week", y = "Metabolite value",
       title = "Top 4 stable metabolites: value ~ GW under each transform") +
  theme_cowplot(10) +
  theme(
    strip.text.y = element_text(size = 8),
    strip.text.x = element_text(size = 9),
    panel.spacing = unit(0.8, "lines")
  )

save_dual_format(p_examples, out_dir, "example_metabolites_by_transform",
                 width = 16, height = max(8, nrow(example_meta) * 3))
print(p_examples)
```

## Summary

```{r}
#| label: summary-table
tibble(
  Transform              = c("log2 + imputed", "z-score IQR-trim (current)",
                             "median-norm + imputed", "median-norm + z-score"),
  `Corrects sample load` = c("No",  "No",  "Yes", "Yes"),
  `Comparable logFC`     = c("No",  "Yes", "No",  "Yes"),
  `Interpretable axis`   = c("Yes", "No",  "Yes", "No"),
  `MSEA ranking valid`   = c("No",  "Yes", "No",  "Yes"),
  `Model predictor`      = c("—",   "Current", "—", "Option if median needed")
) |>
  knitr::kable(caption = "Transform comparison — properties relevant to this pipeline")
```
