---
title: "Metabolite–Gene Associations"
subtitle: "Sneha Chaturvedi, Emma Jones"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    code-tools: true
    scrollable: true
    theme: dark
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Overview

Reads the parquet archive written by `metabolite_gene_limma.R` and does all
thresholding, global-FDR, annotation, and figures here — so screen changes never
re-run the loop.

- **Model (per metabolite):** gene counts ~ `metab + GW + Sex + batch1_frac`, split by cell type; voom.
- **Interpretation:** `logFC` = Δ(gene log2-CPM) per **1 SD** of the standardised metabolite. These are **associations**, not directional predictions.
- **Primary screen:** within-(metabolite × cell type) `adj.P.Val < 0.10` **and** `|logFC| ≥ 0.25`.
- **Global FDR:** BH across every metabolite × cell type × gene test, carried as a reference column.

## Set Up

```{r}
#| label: load-packages
Packages <- c("tidyverse", "arrow", "cowplot", "svglite", "UpSetR", "edgeR")
lapply(Packages, library, character.only = TRUE)

source("pseudobulk_functions.R")     # save_dual_format(), merge_hormone_metadata()
source("metabolite_functions.R")     # build_metab_covariates(), INT transform, mapping
set.seed(123)
```

```{r}
#| label: load-model-objects
# METHOD selects which transform run to analyze; it sets the parquet read below
# AND the reporter's predictor transform, so they can never mismatch.
METHOD <- "zscore_trim"                        # "int" | "zscore" | "zscore_trim"
suffix <- if (METHOD == "int") "" else paste0("-", METHOD)
parquet_dir <- if (METHOD == "int") {"../../results/gene-metabolite/parquet"} else {sprintf("../../results/gene-metabolite/parquet-%s", METHOD)}

# Rebuild the modeled objects (needed only for the LOO / Cook's reporter below).
cache_dir    <- "../../data/cache"
metab_path   <- "../../results/untargeted/batch2_peak_area_clean.csv"
hormone_xlsx <- "../../doc/targeted/targeted_hormones.xlsx"

pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))
targeted_comparison_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`

pb       <- build_metab_covariates(pb_base, targeted_comparison_raw, batch_props,
                                   min_cells = 30)
metab_tx <- transform_metabolite_matrix(read_metabolite_matrix(metab_path),
                                        method = METHOD)   # matches the parquet run

rn        <- rownames(pb$counts)
row_ct    <- sub("^[^.]+\\.(.*)$", "\\1", rn)
metab_col <- metab_sample_to_pb(colnames(metab_tx), sub("^([^.]+)\\..*$", "\\1", rn))
```

```{r}
#| label: config
# parquet_dir + suffix come from the load-model-objects chunk (METHOD switch).
annot_path   <- "../../results/untargeted/batch2_metabolite_annotations_flagged.csv"
out_dir      <- "../../results/gene-metabolite"
svg_dir      <- file.path(out_dir, paste0("svg", suffix))   # method-keyed: no clobber
csv_dir      <- file.path(out_dir, paste0("csv", suffix))
for (d in c(out_dir, svg_dir, csv_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

FDR_WITHIN <- 0.10   # primary screen (within metabolite x cell type)
LFC_MIN    <- 0.25   # per-SD effect-size floor
FDR_GLOBAL <- 0.10   # reference global BH threshold
LABEL_CHARS <- 30    # metabolite-name clip length for plot labels

# Clip long metabolite names for plotting only (data keeps full Name). Names
# longer than `n` get an ellipsis; if two clips collide, the 2nd+ get "#2","#3"
# so bars stay distinct. Returns the input tibble with a `label` column added;
# write the paired label<->Name lookup beside each figure's SVG.
add_plot_label <- function(df, name_col = "Name", n = LABEL_CHARS) {
  short <- ifelse(nchar(df[[name_col]]) > n,
                  paste0(substr(df[[name_col]], 1, n), "\u2026"),
                  df[[name_col]])
  idx   <- stats::ave(seq_along(short), short, FUN = seq_along)   # 1,2,3 within group
  short[idx > 1] <- paste0(short[idx > 1], "#", idx[idx > 1])
  df$label <- short
  df
}
```

## Open the archive (lazy)

`open_dataset()` reads only the columns/rows a query touches, so the multi-GB
archive is never loaded whole.

```{r}
#| label: open-dataset
ds <- open_dataset(parquet_dir)

n_tests    <- nrow(ds)                                   # total tests (cheap)
n_metab    <- ds |> distinct(Compound.ID) |> collect() |> nrow()
n_celltype <- ds |> distinct(cell_type)  |> collect() |> nrow()

cat(sprintf("Total tests: %s | metabolites: %d | cell types: %d\n",
            format(n_tests, big.mark = ","), n_metab, n_celltype))
```

## Global BH (memory-light)

Any test with global FDR < 0.10 must have a small raw p-value, so we only pull
candidates (`P.Value < 0.05`) and rank them against the full test count `m`.
This reproduces the exact BH adjustment those rows would receive over all
`n_tests` without collecting the whole archive.

```{r}
#| label: global-bh
bh_global <- function(p, m) {
  o   <- order(p)
  ro  <- order(o)
  adj <- p[o] * m / seq_along(p[o])
  adj <- rev(cummin(rev(adj)))
  pmin(adj[ro], 1)
}

candidates <- ds |>
  filter(P.Value < 0.05) |>
  collect() |>
  mutate(FDR_global = bh_global(P.Value, n_tests))

cat(sprintf("Candidate rows (p < 0.05): %s\n",
            format(nrow(candidates), big.mark = ",")))
```

## Global-FDR hits + stability reporter

```{r}
#| label: annotate-and-filter
annot <- read_csv(annot_path, show_col_types = FALSE) |>
  dplyr::select(Compound.ID, Super.Class, Class, Sub.Class, Pathways, KEGG, Exogenous_Flag, Annotation_Confidence)

# Primary screen = global FDR (BH across all tests) & |logFC|. Within-FDR
# (adj.P.Val) is retained as a carried column.
hits <- candidates |>
  filter(FDR_global < FDR_GLOBAL, abs(logFC) >= LFC_MIN) |>
  left_join(annot, by = "Compound.ID") |>
  arrange(cell_type, Compound.ID, FDR_global)

cat(sprintf("Global-FDR hits (FDR < %.2f & |logFC| >= %.2f): %s\n",
            FDR_GLOBAL, LFC_MIN, format(nrow(hits), big.mark = ",")))
```

```{r}
#| label: stability-reporter
# Per hit, REPORTER columns (do not gate significance -- filter on them later):
#   n_nonzero    modeled samples with raw count >= 2
#   max_cooks    max Cook's distance in the lm(log1p(CPM) ~ metab + covariates)
#   loo_unstable dropping the most influential sample flips the metabolite slope
#                or halves |slope| -> small-n fragility (Squair et al. 2021,
#                Nat Commun 12:5692). 
report_group <- function(cid, ct, genes) {
  idx      <- which(row_ct == ct)
  mv       <- metab_tx[cid, match(metab_col[idx], colnames(metab_tx))]
  keep     <- !is.na(mv)
  rows_use <- idx[keep]
  md       <- pb$metadata[rn[rows_use], ]
  lib      <- rowSums(pb$counts[rows_use, , drop = FALSE])     # library size / sample
  base     <- tibble(metab = as.numeric(mv[keep]),
                     GW = md$GW, Sex = md$Sex, batch1_frac = md$batch1_frac)
  covs  <- c("GW", "Sex", "batch1_frac")
  keepc <- covs[vapply(covs, function(cv) length(unique(base[[cv]])) >= 2, logical(1))]
  fml   <- reformulate(c("metab", keepc), response = "expr")

  purrr::map_dfr(genes, function(g) {
    if (!g %in% colnames(pb$counts))
      return(tibble(gene = g, n_nonzero = NA_integer_,
                    max_cooks = NA_real_, loo_unstable = NA))
    cnt <- pb$counts[rows_use, g]
    df  <- base; df$expr <- log1p(cnt / lib * 1e6)             # log1p(CPM)
    if (length(unique(df$metab)) < 2 || nrow(df) < 4)
      return(tibble(gene = g, n_nonzero = sum(cnt >= 2),
                    max_cooks = NA_real_, loo_unstable = NA))
    fit <- lm(fml, data = df)
    cd  <- cooks.distance(fit); b <- coef(fit)[["metab"]]
    b2  <- coef(lm(fml, data = df[-which.max(cd), ]))[["metab"]]
    tibble(gene = g, n_nonzero = sum(cnt >= 2), max_cooks = max(cd),
           loo_unstable = sign(b2) != sign(b) || abs(b2) < 0.5 * abs(b))
  })
}

reporter <- hits |>
  group_by(Compound.ID, cell_type) |>
  group_modify(~ report_group(.y$Compound.ID, .y$cell_type, .x$gene)) |>
  ungroup()

hits <- hits |> left_join(reporter, by = c("Compound.ID", "cell_type", "gene"))
write_csv(hits, file.path(csv_dir, "metabolite_gene_hits.csv"))

cat(sprintf("Reporter added. loo_unstable: %s of %s hits (%.1f%%) flagged fragile.\n",
            format(sum(hits$loo_unstable, na.rm = TRUE), big.mark = ","),
            format(nrow(hits), big.mark = ","),
            100 * mean(hits$loo_unstable, na.rm = TRUE)))
```

## Recurrence — genes across metabolites

Genes that track the most distinct metabolites (a gene appearing broadly may be
a hub of metabolic coupling, or a covariate-driven artefact worth inspecting).

```{r}
#| label: gene-recurrence
gene_recur <- hits |>
  group_by(gene) |>
  summarise(n_metabolites = n_distinct(Compound.ID),
            n_cell_types  = n_distinct(cell_type),
            n_tests       = n(), .groups = "drop") |>
  arrange(desc(n_metabolites))

top_genes <- gene_recur |> slice_head(n = 25)

p_gene <- ggplot(top_genes,
                 aes(x = reorder(gene, n_metabolites), y = n_metabolites)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  coord_flip() +
  labs(x = "Gene", y = "Distinct metabolites associated",
       title = "Top recurrent genes across metabolites") +
  theme_cowplot(12)

save_dual_format(p_gene, out_dir, "recurrence_genes", width = 8, height = 7)
print(p_gene)
```

## Recurrence — metabolites across genes

```{r}
#| label: metabolite-recurrence
metab_recur <- hits |>
  group_by(Compound.ID, Name, Super.Class, Class) |>
  summarise(n_genes      = n_distinct(gene),
            n_cell_types = n_distinct(cell_type), .groups = "drop") |>
  arrange(desc(n_genes))

top_metab <- metab_recur |> slice_head(n = 25) |> add_plot_label()

# Full Name <-> clipped label lookup, paired with the SVG.
write_csv(top_metab |> dplyr::select(label, Name, Compound.ID, Super.Class, Class),
          file.path(svg_dir, "recurrence_metabolites_labels.csv"))

p_metab <- ggplot(top_metab,
                  aes(x = reorder(label, n_genes), y = n_genes, fill = Super.Class)) +
  geom_col(alpha = 0.85) +
  coord_flip() +
  labs(x = "Metabolite", y = "Distinct genes associated",
       title = "Top recurrent metabolites across genes") +
  theme_cowplot(12) +
  theme(legend.position = "right")

save_dual_format(p_metab, out_dir, "recurrence_metabolites", width = 10, height = 7)
print(p_metab)
```

## Recurrence — cell types

```{r}
#| label: celltype-recurrence
ct_recur <- hits |>
  group_by(cell_type) |>
  summarise(n_associations = n(),
            n_metabolites  = n_distinct(Compound.ID),
            n_genes        = n_distinct(gene), .groups = "drop") |>
  arrange(desc(n_associations))

p_ct <- ggplot(ct_recur,
               aes(x = reorder(cell_type, n_metabolites), y = n_metabolites)) +
  geom_col(fill = "darkorange", alpha = 0.85) +
  coord_flip() +
  labs(x = "Cell type", y = "Distinct metabolites associated",
       title = "Distinct metabolites by cell type") +
  theme_cowplot(12)

save_dual_format(p_ct, out_dir, "recurrence_celltypes", width = 8, height = 6)
print(p_ct)

knitr::kable(ct_recur, caption = "Hit counts per cell type")
```

## Gene sharing across cell types (UpSet)

Which genes are shared vs. cell-type-specific in their metabolite associations.

```{r}
#| label: upset-genes
#| fig-width: 10
#| fig-height: 6
gene_by_ct <- hits |>
  distinct(gene, cell_type) |>
  mutate(present = 1L) |>
  pivot_wider(names_from = cell_type, values_from = present, values_fill = 0L)

ct_cols <- setdiff(colnames(gene_by_ct), "gene")

if (length(ct_cols) >= 2 && nrow(gene_by_ct) > 0) {
  UpSetR::upset(
    as.data.frame(gene_by_ct), sets = ct_cols,
    nsets = length(ct_cols), nintersects = NA,   # keep all non-empty intersections
    order.by = "freq", keep.order = TRUE
  )
} else {
  cat("Not enough cell types / genes for an UpSet plot.\n")
}
```

## Save recurrence tables

```{r}
#| label: save-tables
write_csv(gene_recur,  file.path(csv_dir, "recurrence_genes.csv"))
write_csv(metab_recur, file.path(csv_dir, "recurrence_metabolites.csv"))
write_csv(ct_recur,    file.path(csv_dir, "recurrence_celltypes.csv"))
cat("Recurrence tables written to", csv_dir, "\n")
```
