---
title: "Metabolite–Gene Stability Filter — DFBETAS (outlier-trimmed run)"
subtitle: "Sneha Chaturvedi"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    scrollable: true
    theme: dark
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Purpose

Second-gate filter on the outlier-trimmed global-FDR hits: keep an association
only if no single sample has outsized *influence* on the metabolite slope. Uses
DFBETAS — the standardized single-deletion change in the coefficient, in units of
its own SE — with the standard cutoff |DFBETAS| > 2/sqrt(n) (Belsley, Kuh &
Welsch 1980). This answers "does dropping one point *matter*", not merely whether
a high-leverage point *exists*, so a far-x sample consistent with the trend is
kept. Outputs to its own subdirectory to keep this iteration separate.

## Setup

```{r}
#| label: setup
Packages <- c("tidyverse", "arrow", "cowplot", "svglite", "edgeR", "patchwork")
lapply(Packages, library, character.only = TRUE)
source("pseudobulk_functions.R")
source("metabolite_functions.R")
set.seed(123)

METHOD <- "zscore_trim"
suffix <- if (METHOD == "int") "" else paste0("-", METHOD)

parquet_dir <- if (METHOD == "int"){ "../../results/gene-metabolite/parquet"} else {sprintf("../../results/gene-metabolite/parquet-%s", METHOD)}
annot_path  <- "../../results/untargeted/batch2_metabolite_annotations_flagged.csv"
out_dir     <- "../../results/gene-metabolite"
# dedicated subdirectory for this iteration (dfbetas gate)
iter_dir    <- file.path(out_dir, paste0("stability-dfbetas", suffix))
csv_dir     <- iter_dir
fig_dir     <- iter_dir
scatter_dir <- file.path(iter_dir, "scatter-plots")
for (d in c(iter_dir, scatter_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

FDR_GLOBAL <- 0.10
LFC_MIN    <- 0.25
LABEL_CHARS <- 30

add_plot_label <- function(df, name_col = "Name", n = LABEL_CHARS) {
  short <- ifelse(nchar(df[[name_col]]) > n,
                  paste0(substr(df[[name_col]], 1, n), "\u2026"), df[[name_col]])
  idx <- stats::ave(seq_along(short), short, FUN = seq_along)
  short[idx > 1] <- paste0(short[idx > 1], "#", idx[idx > 1])
  df$label <- short; df
}
bh_global <- function(p, m) {
  o <- order(p); ro <- order(o)
  adj <- p[o] * m / seq_along(p[o]); adj <- rev(cummin(rev(adj)))
  pmin(adj[ro], 1)
}
```

```{r}
#| label: load-model-objects
cache_dir    <- "../../data/cache"
metab_path   <- "../../results/untargeted/batch2_peak_area_clean.csv"
hormone_xlsx <- "../../doc/targeted/targeted_hormones.xlsx"

pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))
targeted_comparison_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`

pb        <- build_metab_covariates(pb_base, targeted_comparison_raw, batch_props, min_cells = 30)
metab_mat <- read_metabolite_matrix(metab_path)
metab_tx  <- transform_metabolite_matrix(metab_mat, method = METHOD)
metab_log <- t(apply(log2(metab_mat + 1), 1, min_value_impute)); dimnames(metab_log) <- dimnames(metab_mat)

rn        <- rownames(pb$counts)
row_ct    <- sub("^[^.]+\\.(.*)$", "\\1", rn)
metab_col <- metab_sample_to_pb(colnames(metab_tx), sub("^([^.]+)\\..*$", "\\1", rn))
cc_col    <- intersect(c("cell_count", "n_cells", "ncells"), colnames(pb$metadata))[1]
```

## Global-FDR hits

```{r}
#| label: hits
ds <- open_dataset(parquet_dir); n_tests <- nrow(ds)
annot <- read_csv(annot_path, show_col_types = FALSE) |>
  dplyr::select(Compound.ID, Super.Class, Class, Sub.Class, Pathways, KEGG,
                Exogenous_Flag, Annotation_Confidence)

hits <- ds |> filter(P.Value < 0.05) |> collect() |>
  mutate(FDR_global = bh_global(P.Value, n_tests)) |>
  filter(FDR_global < FDR_GLOBAL, abs(logFC) >= LFC_MIN) |>
  left_join(annot, by = "Compound.ID")

cat(sprintf("Global-FDR hits: %s\n", format(nrow(hits), big.mark = ",")))
```

## DFBETAS influence gate

```{r}
#| label: dfbetas
# stable = no single sample moves the metabolite coefficient by more than
# 2/sqrt(n) of its SE. max_dfb = worst |DFBETAS| across samples (carried column).
dfbetas_group <- function(cid, ct, genes) {
  idx  <- which(row_ct == ct)
  mv   <- metab_tx[cid, match(metab_col[idx], colnames(metab_tx))]
  keep <- !is.na(mv); rows_use <- idx[keep]
  md   <- pb$metadata[rn[rows_use], ]
  lib  <- rowSums(pb$counts[rows_use, , drop = FALSE])
  base <- tibble(metab_trim = as.numeric(mv[keep]),
                 GW = md$GW, Sex = md$Sex, batch1_frac = md$batch1_frac)
  covs  <- c("GW", "Sex", "batch1_frac")
  keepc <- covs[vapply(covs, function(cv) length(unique(base[[cv]])) >= 2, logical(1))]
  fml   <- reformulate(c("metab_trim", keepc), response = "expr")

  purrr::map_dfr(genes, function(g) {
    if (!g %in% colnames(pb$counts))
      return(tibble(gene = g, max_dfb = NA_real_, dfb_cut = NA_real_, stable = NA))
    df <- base; df$expr <- log1p(pb$counts[rows_use, g] / lib * 1e6)
    n  <- nrow(df)
    if (length(unique(df$metab_trim)) < 2 || n < length(keepc) + 3)
      return(tibble(gene = g, max_dfb = NA_real_, dfb_cut = NA_real_, stable = NA))
    fit    <- lm(fml, data = df)
    dfb    <- stats::dfbetas(fit)[, "metab_trim"]      # SE-scaled single-deletion change
    cutoff <- 2 / sqrt(n)                              # standard DFBETAS cutoff
    tibble(gene = g, max_dfb = max(abs(dfb)), dfb_cut = cutoff,
           stable = max(abs(dfb)) <= cutoff)
  })
}

reporter <- hits |> group_by(Compound.ID, cell_type) |>
  group_modify(~ dfbetas_group(.y$Compound.ID, .y$cell_type, .x$gene)) |> ungroup()
hits <- hits |> left_join(reporter, by = c("Compound.ID", "cell_type", "gene"))

stable_hits <- hits |> filter(stable %in% TRUE)          # FILTER, not flag
write_csv(stable_hits, file.path(csv_dir, "metabolite_gene_hits_dfbetas_stable.csv"))

cat(sprintf("DFBETAS-stable (no |DFBETAS| > 2/sqrt(n)): %s / %s (%.1f%%)\n",
            format(nrow(stable_hits), big.mark = ","),
            format(nrow(hits), big.mark = ","),
            100 * mean(hits$stable %in% TRUE)))
```

## Stable vs. unstable

```{r}
#| label: count-bar
counts <- hits |> mutate(status = ifelse(stable %in% TRUE, "stable", "unstable")) |> count(status)
p_cnt <- ggplot(counts, aes(status, n, fill = status)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.3) +
  scale_fill_manual(values = c(stable = "steelblue", unstable = "grey60")) +
  labs(x = NULL, y = "Hits", title = "DFBETAS influence gate") + theme_cowplot(13)
save_dual_format(p_cnt, fig_dir, "dfbetas_counts", width = 5, height = 5); print(p_cnt)
```

## Recurrence on stable hits

```{r}
#| label: recurrence
gene_recur <- stable_hits |> group_by(gene) |>
  summarise(n_metabolites = n_distinct(Compound.ID), .groups = "drop") |>
  slice_max(n_metabolites, n = 25)
p_g <- ggplot(gene_recur, aes(reorder(gene, n_metabolites), n_metabolites)) +
  geom_col(fill = "steelblue", alpha = 0.85) + coord_flip() +
  labs(x = "Gene", y = "Distinct metabolites", title = "Stable: top genes") + theme_cowplot(12)
save_dual_format(p_g, fig_dir, "stable_recurrence_genes", width = 8, height = 7); print(p_g)

metab_recur <- stable_hits |> group_by(Compound.ID, Name, Super.Class) |>
  summarise(n_genes = n_distinct(gene), .groups = "drop") |>
  slice_max(n_genes, n = 25) |> add_plot_label()
write_csv(metab_recur |> dplyr::select(label, Name, Compound.ID),
          file.path(fig_dir, "stable_recurrence_metabolites_labels.csv"))
p_m <- ggplot(metab_recur, aes(reorder(label, n_genes), n_genes, fill = Super.Class)) +
  geom_col(alpha = 0.85) + coord_flip() +
  labs(x = "Metabolite", y = "Distinct genes", title = "Stable: top metabolites") + theme_cowplot(12)
save_dual_format(p_m, fig_dir, "stable_recurrence_metabolites", width = 10, height = 7); print(p_m)

ct_recur <- stable_hits |> group_by(cell_type) |>
  summarise(n_metabolites = n_distinct(Compound.ID), .groups = "drop")
p_c <- ggplot(ct_recur, aes(reorder(cell_type, n_metabolites), n_metabolites)) +
  geom_col(fill = "darkorange", alpha = 0.85) + coord_flip() +
  labs(x = "Cell type", y = "Distinct metabolites", title = "Stable: by cell type") + theme_cowplot(12)
save_dual_format(p_c, fig_dir, "stable_recurrence_celltypes", width = 8, height = 6); print(p_c)
```

## Top-20 stable scatters (outlier-trimmed panels)

```{r}
#| label: scatter-builder
make_pair_plot <- function(cid, gene, ct, meta_name) {
  idx <- which(row_ct == ct); pos <- match(metab_col[idx], colnames(metab_tx))
  present <- !is.na(pos) & gene %in% colnames(pb$counts)
  rows_use <- idx[present]; pos_p <- pos[present]
  if (length(rows_use) < 3) return(NULL)
  expr <- as.numeric(log1p(edgeR::cpm(t(pb$counts[rows_use, , drop = FALSE]), log = FALSE)[gene, ]))
  md <- pb$metadata[rn[rows_use], ]
  df <- tibble(metab_raw = as.numeric(metab_log[cid, pos_p]),
               metab_trim = as.numeric(metab_tx[cid, pos_p]), expr = expr,
               cell_count = if (!is.na(cc_col)) md[[cc_col]] else NA_real_,
               Sex = md$Sex, GW = md$GW, batch1_frac = md$batch1_frac)
  bl <- list(geom_point(size = 2, alpha = 0.8), geom_smooth(method = "lm", se = TRUE),
             scale_color_viridis_c(), scale_shape_manual(values = c("M" = 16, "F" = 17)),
             theme_minimal(base_size = 13))
  p_raw <- ggplot(df, aes(metab_raw, expr, color = cell_count, shape = Sex)) + bl +
    labs(title = "Raw (log2, all samples)", x = meta_name, y = "log1p(CPM)")
  dft <- df |> filter(!is.na(metab_trim))
  p_mid <- ggplot(dft, aes(metab_trim, expr, color = cell_count, shape = Sex)) + bl +
    labs(title = "Outlier-trimmed", x = paste0(meta_name, " (trimmed z)"), y = "log1p(CPM)")
  covs <- c("GW", "Sex", "batch1_frac")
  keepc <- covs[vapply(covs, function(cv) length(unique(dft[[cv]])) >= 2, logical(1))]
  fit <- lm(reformulate(c("metab_trim", keepc), response = "expr"), data = dft)
  dft$part <- resid(fit) + coef(fit)[["metab_trim"]] * dft$metab_trim
  p_adj <- ggplot(dft, aes(metab_trim, part, color = cell_count, shape = Sex)) + bl +
    labs(title = "Outlier-trimmed (adjusted)", x = paste0(meta_name, " (trimmed z)"),
         y = "log1p(CPM), adjusted")
  (p_raw | p_mid | p_adj) +
    plot_annotation(title = paste0(gene, "  ~  ", meta_name, "   (", ct, ")")) +
    plot_layout(guides = "collect")
}
```

```{r}
#| label: scatter-generate
#| results: asis
top20 <- stable_hits |>
  arrange(FDR_global, desc(abs(logFC))) |>
  distinct(Compound.ID, .keep_all = TRUE) |>
  slice_head(n = 20)

for (i in seq_len(nrow(top20))) {
  r  <- top20[i, ]
  pl <- make_pair_plot(r$Compound.ID, r$gene, r$cell_type, r$Name)
  if (is.null(pl)) next
  base <- gsub("[^A-Za-z0-9]+", "-", paste(r$Compound.ID, r$gene, r$cell_type, sep = "_"))
  save_dual_format(pl, scatter_dir, base, width = 16, height = 5)
  print(pl)
}
cat("\nSaved", nrow(top20), "stable scatters to", scatter_dir, "\n")
```
