---
title: "Metabolite–Gene Stability Filter — DFBETAS (log2_na run)"
subtitle: "Sneha Chaturvedi"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    scrollable: true
    theme: dark
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Purpose

Second-gate filter producing the **most robust** gene–metabolite associations.
Input is the global-FDR hit set from `metabolite_gene_associations.qmd`
(the `log2_na` run: no imputation, IQR outlier trimming, z-scored predictor).
Because `log2_na` carries no imputed values, the imputation-sensitivity gate is
not needed and this script reads global-FDR hits directly.

Associations are kept only if no single sample has outsized *influence* on the
metabolite slope. Uses DFBETAS — the standardized single-deletion change in the
coefficient, in units of its own SE — with the standard cutoff
|DFBETAS| > 2/sqrt(n) (Belsley, Kuh & Welsch 1980). This answers "does
dropping one point *matter*", not merely whether a high-leverage point *exists*,
so a far-x sample consistent with the trend is kept.

Pairs that survive the DFBETAS gate are written as
`metabolite_gene_hits_robust.csv` and represent the highest-confidence set for
manuscript-level reporting.

## Setup

```{r}
#| label: setup
Packages <- c("tidyverse", "arrow", "cowplot", "svglite", "edgeR", "patchwork", "UpSetR")
lapply(Packages, library, character.only = TRUE)
source("pseudobulk_functions.R")
source("metabolite_functions.R")
set.seed(123)

METHOD <- "log2_na"
suffix <- if (METHOD == "int") "" else paste0("-", METHOD)

out_dir          <- "../../results/gene-metabolite"
# Input: global-FDR hits from metabolite_gene_associations.qmd.
# log2_na has no imputation, so the imputation-sensitivity gate is skipped;
# DFBETAS reads the global-FDR set directly.
assoc_csv_dir    <- file.path(out_dir, paste0("csv", suffix))
hits_path        <- file.path(assoc_csv_dir, "metabolite_gene_hits.csv")
# Output: dedicated subdirectory for this (DFBETAS) gate
iter_dir         <- file.path(out_dir, paste0("stability-dfbetas", suffix))
csv_dir     <- iter_dir
fig_dir     <- iter_dir
scatter_dir <- file.path(iter_dir, "scatter-plots")
for (d in c(iter_dir, scatter_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

FDR_GLOBAL <- 0.10
LFC_MIN    <- 0.25
LABEL_CHARS <- 30

add_plot_label <- function(df, name_col = "Name", n = LABEL_CHARS) {
  short <- ifelse(nchar(df[[name_col]]) > n,
                  paste0(substr(df[[name_col]], 1, n), "\u2026"), df[[name_col]])
  idx <- stats::ave(seq_along(short), short, FUN = seq_along)
  short[idx > 1] <- paste0(short[idx > 1], "#", idx[idx > 1])
  df$label <- short; df
}
bh_global <- function(p, m) {
  o <- order(p); ro <- order(o)
  adj <- p[o] * m / seq_along(p[o]); adj <- rev(cummin(rev(adj)))
  pmin(adj[ro], 1)
}
```

```{r}
#| label: load-model-objects
cache_dir    <- "../../data/cache"
metab_path   <- "../../results/untargeted/batch2_peak_area_clean.csv"
hormone_xlsx <- "../../doc/targeted/targeted_hormones.xlsx"

pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))
targeted_comparison_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`

pb        <- build_metab_covariates(pb_base, targeted_comparison_raw, batch_props, min_cells = 30)
metab_mat <- read_metabolite_matrix(metab_path)
metab_tx  <- transform_metabolite_matrix(metab_mat, method = METHOD)
metab_log <- log2(metab_mat + 1)   # raw log2; NAs stay NA (no imputation)

# Log2 values with IQR outliers flagged NA — used for scatter x-axis only.
# Mirrors the fence from zscore_trim (Q1 - 0.75*IQR / Q3 + 0.75*IQR) but
# skips imputation and z-scoring so the axis stays in interpretable log2 units.
.log2_trim_only <- function(x) {
  logged  <- log2(x + 1)
  orig_na <- is.na(logged)
  obs     <- logged[!orig_na]
  if (length(obs) < 2) return(rep(NA_real_, length(x)))
  qs  <- quantile(obs, c(0.25, 0.75), names = FALSE)
  iqr <- qs[2] - qs[1]
  v   <- logged
  v[!orig_na & (logged < qs[1] - 0.75 * iqr | logged > qs[2] + 0.75 * iqr)] <- NA
  v
}
metab_log_trim <- t(apply(metab_mat, 1, .log2_trim_only))
dimnames(metab_log_trim) <- dimnames(metab_mat)

rn        <- rownames(pb$counts)
row_ct    <- sub("^[^.]+\\.(.*)$", "\\1", rn)
metab_col <- metab_sample_to_pb(colnames(metab_tx), sub("^([^.]+)\\..*$", "\\1", rn))
```

## Global-FDR hits (input)

```{r}
#| label: hits
# Load global-FDR hits from metabolite_gene_associations.qmd (log2_na run).
# Annotations and reporter columns are already present.
hits <- read_csv(hits_path, show_col_types = FALSE)

cat(sprintf(
  "Reading: %s\nGlobal-FDR hits: %s rows | %d metabolites | %d cell types | %d genes\n",
  hits_path,
  format(nrow(hits), big.mark = ","),
  n_distinct(hits$Compound.ID),
  n_distinct(hits$cell_type),
  n_distinct(hits$gene)
))
```

## DFBETAS influence gate

```{r}
#| label: dfbetas
# stable = no single sample moves the metabolite coefficient by more than
# 2/sqrt(n) of its SE. max_dfb = worst |DFBETAS| across samples (carried column).
dfbetas_group <- function(cid, ct, genes) {
  idx  <- which(row_ct == ct)
  mv   <- metab_tx[cid, match(metab_col[idx], colnames(metab_tx))]
  keep <- !is.na(mv); rows_use <- idx[keep]
  md   <- pb$metadata[rn[rows_use], ]
  lib  <- rowSums(pb$counts[rows_use, , drop = FALSE])
  base <- tibble(metab_trim = as.numeric(mv[keep]),
                 GW = md$GW, Sex = md$Sex, batch1_frac = md$batch1_frac)
  covs  <- c("GW", "Sex", "batch1_frac")
  keepc <- covs[vapply(covs, function(cv) length(unique(base[[cv]])) >= 2, logical(1))]
  fml   <- reformulate(c("metab_trim", keepc), response = "expr")

  purrr::map_dfr(genes, function(g) {
    if (!g %in% colnames(pb$counts))
      return(tibble(gene = g, max_dfb = NA_real_, dfb_cut = NA_real_, stable = NA))
    df <- base; df$expr <- log1p(pb$counts[rows_use, g] / lib * 1e6)
    n  <- nrow(df)
    if (length(unique(df$metab_trim)) < 2 || n < length(keepc) + 3)
      return(tibble(gene = g, max_dfb = NA_real_, dfb_cut = NA_real_, stable = NA))
    fit    <- lm(fml, data = df)
    dfb    <- stats::dfbetas(fit)[, "metab_trim"]      # SE-scaled single-deletion change
    cutoff <- 2 / sqrt(n)                              # standard DFBETAS cutoff
    tibble(gene = g, max_dfb = max(abs(dfb)), dfb_cut = cutoff,
           stable = max(abs(dfb)) <= cutoff)
  })
}

reporter <- hits |> group_by(Compound.ID, cell_type) |>
  group_modify(~ dfbetas_group(.y$Compound.ID, .y$cell_type, .x$gene)) |> ungroup()
hits <- hits |> left_join(reporter, by = c("Compound.ID", "cell_type", "gene"))

stable_hits <- hits |> filter(stable %in% TRUE)          # FILTER, not flag
write_csv(stable_hits, file.path(csv_dir, "metabolite_gene_hits_robust.csv"))

cat(sprintf(
  "Robust hits (imputation-insensitive + DFBETAS-stable): %s / %s (%.1f%%)\n",
  format(nrow(stable_hits), big.mark = ","),
  format(nrow(hits), big.mark = ","),
  100 * mean(hits$stable %in% TRUE)
))
```

## DFBETAS gate on global-FDR hits

```{r}
#| label: count-bar
counts <- hits |> mutate(status = ifelse(stable %in% TRUE, "robust", "DFBETAS-unstable")) |> count(status)
p_cnt <- ggplot(counts, aes(status, n, fill = status)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.3) +
  scale_fill_manual(values = c(robust = "steelblue", `DFBETAS-unstable` = "grey60")) +
  labs(x = NULL, y = "Hits",
       title = "DFBETAS gate (input: global-FDR hits, log2_na)") + theme_cowplot(13)
save_dual_format(p_cnt, fig_dir, "dfbetas_counts", width = 5, height = 5); print(p_cnt)
```

## Recurrence on robust hits

```{r}
#| label: recurrence
gene_recur <- stable_hits |> group_by(gene) |>
  summarise(n_metabolites = n_distinct(Compound.ID), .groups = "drop") |>
  slice_max(n_metabolites, n = 20)
p_g <- ggplot(gene_recur, aes(reorder(gene, n_metabolites), n_metabolites)) +
  geom_col(fill = "steelblue", alpha = 0.85) + coord_flip() +
  labs(x = "Gene", y = "Distinct metabolites", title = "Robust: top genes") + theme_cowplot(12)
save_dual_format(p_g, fig_dir, "robust_recurrence_genes", width = 8, height = 7); print(p_g)

metab_recur <- stable_hits |> group_by(Compound.ID, Name, Super.Class) |>
  summarise(n_genes = n_distinct(gene), .groups = "drop") |>
  slice_max(n_genes, n = 25) |> add_plot_label()
write_csv(metab_recur |> dplyr::select(label, Name, Compound.ID),
          file.path(fig_dir, "robust_recurrence_metabolites_labels.csv"))
p_m <- ggplot(metab_recur, aes(reorder(label, n_genes), n_genes, fill = Super.Class)) +
  geom_col(alpha = 0.85) + coord_flip() +
  labs(x = "Metabolite", y = "Distinct genes", title = "Robust: top metabolites") + theme_cowplot(12)
save_dual_format(p_m, fig_dir, "robust_recurrence_metabolites", width = 10, height = 7); print(p_m)

ct_recur <- stable_hits |> group_by(cell_type) |>
  summarise(n_metabolites = n_distinct(Compound.ID), .groups = "drop")
p_c <- ggplot(ct_recur, aes(reorder(cell_type, n_metabolites), n_metabolites)) +
  geom_col(fill = "darkorange", alpha = 0.85) + coord_flip() +
  labs(x = "Cell type", y = "Distinct metabolites", title = "Robust: by cell type") + theme_cowplot(12)
save_dual_format(p_c, fig_dir, "robust_recurrence_celltypes", width = 8, height = 6); print(p_c)
```

## Robust gene sharing across cell types (UpSet)

Which genes are shared vs. cell-type-specific among robust (DFBETAS-stable) associations.

```{r}
#| label: upset-genes
#| fig-width: 10
#| fig-height: 6
gene_by_ct <- stable_hits |>
  distinct(gene, cell_type) |>
  mutate(present = 1L) |>
  pivot_wider(names_from = cell_type, values_from = present, values_fill = 0L)

ct_cols <- setdiff(colnames(gene_by_ct), "gene")

if (length(ct_cols) >= 2 && nrow(gene_by_ct) > 0) {
  UpSetR::upset(
    as.data.frame(gene_by_ct), sets = ct_cols,
    nsets = length(ct_cols), nintersects = NA,
    order.by = "freq", keep.order = TRUE
  )
} else {
  cat("Not enough cell types / genes for an UpSet plot.\n")
}
```

## Top-20 robust scatters

Five-panel scatter: raw log2 (all samples) | outlier-trimmed log2 |
covariate-adjusted log2 || outlier-trimmed z-score | covariate-adjusted z-score.
Log2 panels use free x-axis per metabolite; z-score panels share a fixed axis
across all plots for cross-metabolite comparison. One pooled regression line per
panel (the model includes Sex as a covariate, not a grouping variable); points
colored by GW and shaped by Sex.

```{r}
#| label: scatter-builder
make_pair_plot <- function(cid, gene, ct, meta_name, z_x_lim) {
  idx      <- which(row_ct == ct)
  pos      <- match(metab_col[idx], colnames(metab_tx))
  present  <- !is.na(pos) & gene %in% colnames(pb$counts)
  rows_use <- idx[present]; pos_p <- pos[present]
  if (length(rows_use) < 3) return(NULL)

  expr <- as.numeric(log1p(edgeR::cpm(t(pb$counts[rows_use, , drop = FALSE]), log = FALSE)[gene, ]))
  md   <- pb$metadata[rn[rows_use], ]

  df <- tibble(
    metab_raw     = as.numeric(metab_log[cid, pos_p]),       # raw log2 (NAs for missing)
    metab_logtrim = as.numeric(metab_log_trim[cid, pos_p]),  # log2, outliers = NA
    metab_trim    = as.numeric(metab_tx[cid, pos_p]),        # z-score (model predictor)
    expr          = expr,
    Sex           = md$Sex,
    GW            = md$GW,
    batch1_frac   = md$batch1_frac
  )

  covs <- c("GW", "Sex", "batch1_frac")

  # ---- log2 panels (free x-axis per metabolite) ----
  # Compute x-limits from raw values (widest range, includes outliers)
  x_range <- range(df$metab_raw, na.rm = TRUE)
  x_pad   <- diff(x_range) * 0.05
  x_lim   <- c(x_range[1] - x_pad, x_range[2] + x_pad)
  x_lab   <- paste0(meta_name, " [log2 peak area]")

  # Shared layer list: shape in geom_point only → single regression line
  bl_log <- list(
    geom_point(aes(shape = Sex), size = 2.5, alpha = 0.85),
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE, linewidth = 0.8),
    scale_color_viridis_c(name = "GW"),
    scale_shape_manual(values = c("M" = 16, "F" = 17)),
    coord_cartesian(xlim = x_lim),
    theme_minimal(base_size = 13)
  )

  # Panel 1: raw log2 — all observed samples, before any trimming
  df_raw <- df |> filter(!is.na(metab_raw))
  p_raw <- ggplot(df_raw, aes(metab_raw, expr, color = GW)) + bl_log +
    labs(title = "Raw log2 (all samples)", x = x_lab, y = "log1p(CPM)")

  # Panel 2: outlier-trimmed log2
  dft <- df |> filter(!is.na(metab_logtrim))
  keepc <- covs[vapply(covs, function(cv) length(unique(dft[[cv]])) >= 2, logical(1))]
  p_mid <- ggplot(dft, aes(metab_logtrim, expr, color = GW)) + bl_log +
    labs(title = "Outlier-trimmed (log2)", x = x_lab, y = "log1p(CPM)")

  # Panel 3: covariate-adjusted log2 (partial regression)
  fit_log  <- lm(reformulate(c("metab_logtrim", keepc), response = "expr"), data = dft)
  dft$part <- resid(fit_log) + coef(fit_log)[["metab_logtrim"]] * dft$metab_logtrim
  p_adj <- ggplot(dft, aes(metab_logtrim, part, color = GW)) + bl_log +
    labs(title = "Covariate-adjusted (log2)", x = x_lab, y = "log1p(CPM), adjusted")

  # ---- z-score panels (fixed x-axis across all plots) ----
  z_lab <- paste0(meta_name, " [z-score]")

  bl_z <- list(
    geom_point(aes(shape = Sex), size = 2.5, alpha = 0.85),
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE, linewidth = 0.8),
    scale_color_viridis_c(name = "GW"),
    scale_shape_manual(values = c("M" = 16, "F" = 17)),
    coord_cartesian(xlim = z_x_lim),
    theme_minimal(base_size = 13)
  )

  # Panel 4: outlier-trimmed z-score (model predictor scale)
  dft_z   <- df |> filter(!is.na(metab_trim))
  keepc_z <- covs[vapply(covs, function(cv) length(unique(dft_z[[cv]])) >= 2, logical(1))]
  p_zmid  <- ggplot(dft_z, aes(metab_trim, expr, color = GW)) + bl_z +
    labs(title = "Outlier-trimmed (z-score)", x = z_lab, y = "log1p(CPM)")

  # Panel 5: covariate-adjusted z-score
  fit_z        <- lm(reformulate(c("metab_trim", keepc_z), response = "expr"), data = dft_z)
  dft_z$part_z <- resid(fit_z) + coef(fit_z)[["metab_trim"]] * dft_z$metab_trim
  p_zadj <- ggplot(dft_z, aes(metab_trim, part_z, color = GW)) + bl_z +
    labs(title = "Covariate-adjusted (z-score)", x = z_lab, y = "log1p(CPM), adjusted")

  # ---- 2-row layout: log2 row / z-score row ----
  (p_raw | p_mid | p_adj) / (plot_spacer() | p_zmid | p_zadj) +
    plot_annotation(title = paste0(gene, "  ~  ", meta_name, "   (", ct, ")")) +
    plot_layout(guides = "collect")
}
```

```{r}
#| label: scatter-generate
#| results: asis
top20 <- stable_hits |>
  arrange(FDR_global, desc(abs(logFC))) |>
  distinct(Compound.ID, .keep_all = TRUE) |>
  slice_head(n = 20)

# Global z-score x-axis: symmetric range covering 99.8% of all z-scores across
# all metabolites. Fixed once so z-score panels are directly comparable across plots.
z_vals   <- as.vector(metab_tx)
z_abs    <- quantile(abs(z_vals), 0.999, na.rm = TRUE)
z_x_lim  <- c(-z_abs, z_abs)
cat(sprintf("Global z-score x-axis: [%.2f, %.2f]\n", z_x_lim[1], z_x_lim[2]))

for (i in seq_len(nrow(top20))) {
  r  <- top20[i, ]
  pl <- make_pair_plot(r$Compound.ID, r$gene, r$cell_type, r$Name, z_x_lim)
  if (is.null(pl)) next
  base <- gsub("[^A-Za-z0-9]+", "-", paste(r$Compound.ID, r$gene, r$cell_type, sep = "_"))
  save_dual_format(pl, scatter_dir, base, width = 16, height = 10)
  print(pl)
}
cat("\nSaved", nrow(top20), "stable scatters to", scatter_dir, "\n")
```
