---
title: "MOFA Integration — First Pass"
subtitle: "Sneha Chaturvedi"
date: today
format:
  revealjs:
    slide-number: "c/t"
    code-fold: true
    code-tools: true
    scrollable: true
    theme: dark
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Overview

First-pass multi-omics factor analysis (MOFA2; Argelaguet et al. 2020,
Genome Biology 21:111) integrating:
 
- **11 gene expression views** — one per cell type, pseudobulk counts collapsed
  to the biological-sample level, top 2,000 HVGs per cell type by log-CPM
  variance.
- **1 metabolite view** — `log2_na` transform (log2, IQR outlier trim, z-score;
  NAs = missing/outlier, handled natively by MOFA).

Shared sample axis = biological samples with **both** assays (n ≈ 28).

Cell types with < 80% sample completeness will have NAs for absent samples —
MOFA handles this via its probabilistic framework. Cell types failing
`MIN_SAMPLES` are excluded entirely.

Reference: Argelaguet R, Arnol D, Ber D, et al. (2020). MOFA+: a statistical
framework for comprehensive integration of multi-modal single-cell data.
*Genome Biology* 21:111. doi:10.1186/s13059-020-02015-1

## One-time install

Remove this chunk after `.sif` rebuild.

```{r}
#| label: install
#| eval: false

# MOFA2 from Bioconductor (pulls basilisk for auto-managed Python env)
if (!requireNamespace("MOFA2", quietly = TRUE)) {
  BiocManager::install("MOFA2")
}
```

## Setup

```{r}
#| label: setup

suppressPackageStartupMessages({
  library(tidyverse)
  library(limma)
  library(edgeR)
  library(MOFA2)
  library(cowplot)
  library(svglite)
})

source("pseudobulk_functions.R")
source("metabolite_functions.R")

set.seed(42)

# ---- paths (from src/gene-metabolite/) ---------------------------------------
cache_dir    <- "../../data/cache"
metab_path   <- "../../results/untargeted/batch2_peak_area_clean.csv"
hormone_xlsx <- "../../doc/targeted/targeted_hormones.xlsx"
out_dir      <- "../../results/mofa"

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(out_dir, "svg"), recursive = TRUE, showWarnings = FALSE)

# ---- parameters --------------------------------------------------------------
MIN_CELLS     <- 30L    # min cells per (sample x cell type), matches limma pipeline
MIN_SAMPLES   <- 10L    # min samples per cell type to include as a view
N_HVG         <- 2000L  # top variable genes per cell type
METAB_METHOD  <- "log2_na"
```

## Load cached data

```{r}
#| label: load-data

pb_base     <- readRDS(file.path(cache_dir, "pb_base.rds"))
batch_props <- readRDS(file.path(cache_dir, "batch_props_sample.rds"))

targeted_comparison_raw <- readxl::read_xlsx(hormone_xlsx)
targeted_comparison_raw$E2 <- targeted_comparison_raw$`17B-E2 (ng/mL)`
targeted_comparison_raw$TT <- targeted_comparison_raw$`TT (ng/mL)`
targeted_comparison_raw$P4 <- targeted_comparison_raw$`P4 (ng/mL)`

# Build covariates (Sex, GW, batch1_frac) — identical gate to limma pipeline
pb <- build_metab_covariates(
  pb_base, targeted_comparison_raw, batch_props,
  min_cells = MIN_CELLS
)

cat(sprintf("Pseudobulk rows (sample x cell_type): %d\n", nrow(pb$metadata)))
cat(sprintf("Distinct biological samples: %d\n", n_distinct(pb$metadata$sample_id)))
cat(sprintf("Cell types: %s\n", paste(sort(unique(pb$metadata$cell_type)), collapse = ", ")))
```

## Identify shared samples (both assays)

```{r}
#| label: shared-samples

metab_mat <- read_metabolite_matrix(metab_path)

# Map metabolite columns to pseudobulk sample IDs
metab_col_for_pb <- metab_sample_to_pb(colnames(metab_mat), pb$metadata$sample_id)

# Biological samples that have metabolomics data
samples_with_metab <- unique(pb$metadata$sample_id[!is.na(metab_col_for_pb)])

# The shared sample axis for MOFA: biological samples with both assays
shared_samples <- sort(samples_with_metab)
n_shared <- length(shared_samples)

cat(sprintf("Shared samples (both assays): %d\n", n_shared))

# Sample metadata for covariates (one row per biological sample)
sample_meta <- pb$metadata %>%
  tibble::rownames_to_column("pseudobulk_id") %>%
  filter(sample_id %in% shared_samples) %>%
  group_by(sample_id) %>%
  summarise(
    GW          = first(GW),
    Sex         = first(Sex),
    batch1_frac = first(batch1_frac),
    .groups     = "drop"
  ) %>%
  arrange(match(sample_id, shared_samples))

print(sample_meta)
```

## Build per-cell-type gene views

For each cell type, we:

1. Extract the pseudobulk count matrix at the **biological sample** level
   (rows from `pb$counts` matching this cell type)
2. Filter to shared samples
3. Gene-filter: `rowSums(counts >= 2) >= 3` (matches limma pipeline)
4. Select top `r N_HVG` HVGs by variance of `log2(CPM + 1)`
5. Normalise to `log2(CPM + 1)` for MOFA input

Cell types present in < `MIN_SAMPLES` of shared samples are excluded.
Cell types present in < 100% of samples produce NAs for absent samples.

```{r}
#| label: gene-views

cell_types <- sort(unique(pb$metadata$cell_type))

gene_views <- list()
view_info  <- tibble(
  cell_type    = character(),
  n_samples    = integer(),
  n_genes_pass = integer(),
  n_genes_hvg  = integer(),
  frac_present = numeric()
)

for (ct in cell_types) {

  # Rows for this cell type
  ct_rows <- which(pb$metadata$cell_type == ct)
  ct_meta <- pb$metadata[ct_rows, , drop = FALSE]
  ct_counts <- pb$counts[ct_rows, , drop = FALSE]   # samples x genes

  # Restrict to shared samples
  keep <- ct_meta$sample_id %in% shared_samples
  ct_meta   <- ct_meta[keep, , drop = FALSE]
  ct_counts <- ct_counts[keep, , drop = FALSE]

  n_ct <- nrow(ct_counts)

  if (n_ct < MIN_SAMPLES) {
    cat(sprintf("  SKIP %s: %d samples < MIN_SAMPLES=%d\n", ct, n_ct, MIN_SAMPLES))
    next
  }

  # Transpose to genes x samples for edgeR/limma conventions
  counts_gxs <- t(ct_counts)   # genes x samples

  # Gene filter: >= 2 counts in >= 3 samples (manual, matches limma pipeline)
  gene_pass <- rowSums(counts_gxs >= 2) >= 3
  counts_gxs <- counts_gxs[gene_pass, , drop = FALSE]

  # log2 CPM for HVG selection
  lib_sizes <- colSums(counts_gxs)
  log2cpm   <- edgeR::cpm(counts_gxs, lib.size = lib_sizes, log = TRUE)
  
  # Top N_HVG by variance across samples
  gene_var <- apply(log2cpm, 1, var)
  top_idx  <- order(gene_var, decreasing = TRUE)[seq_len(min(N_HVG, length(gene_var)))]
  log2cpm_hvg <- log2cpm[top_idx, , drop = FALSE]
  
  # Regress out cell count from each gene
  ct_cell_count <- ct_meta$cell_count
  log_cc <- log(ct_cell_count)
  log2cpm_hvg <- t(apply(log2cpm_hvg, 1, function(g) {
    fit <- lm(g ~ log_cc)
    residuals(fit)
  }))
  colnames(log2cpm_hvg) <- rownames(ct_meta)

  # Build the view matrix: features x shared_samples (NA where cell type absent)
  view_mat <- matrix(NA_real_,
                     nrow = nrow(log2cpm_hvg),
                     ncol = n_shared,
                     dimnames = list(rownames(log2cpm_hvg), shared_samples))

  # Fill in the samples that have this cell type
  for (j in seq_len(ncol(log2cpm_hvg))) {
    sid <- ct_meta$sample_id[j]
    view_mat[, sid] <- log2cpm_hvg[, j]
  }

  view_name <- ct
  gene_views[[view_name]] <- view_mat

  view_info <- bind_rows(view_info, tibble(
    cell_type    = ct,
    n_samples    = n_ct,
    n_genes_pass = sum(gene_pass),
    n_genes_hvg  = nrow(log2cpm_hvg),
    frac_present = n_ct / n_shared
  ))

  cat(sprintf("  %s: %d/%d samples, %d genes pass filter, %d HVGs selected\n",
              ct, n_ct, n_shared, sum(gene_pass), nrow(log2cpm_hvg)))
}

knitr::kable(view_info, digits = 2, caption = "Gene views summary")
```

## Build metabolite view

```{r}
#| label: metab-view

# Apply log2_na transform (log2 -> IQR trim -> z-score, NAs stay NA)
metab_tx <- transform_metabolite_matrix(metab_mat, method = METAB_METHOD)
name_lookup <- attr(metab_tx, "name_lookup")

cat(sprintf("Metabolite matrix: %d metabolites x %d samples\n",
            nrow(metab_tx), ncol(metab_tx)))

# Build the view matrix: metabolites x shared_samples
# Need to map shared biological sample IDs -> metabolite column names
metab_view <- matrix(NA_real_,
                     nrow = nrow(metab_tx),
                     ncol = n_shared,
                     dimnames = list(rownames(metab_tx), shared_samples))

for (i in seq_along(shared_samples)) {
  sid <- shared_samples[i]
  # Find the metabolite column for this sample
  # pb rows for this sample (any cell type)
  pb_rows_sid <- which(pb$metadata$sample_id == sid)
  if (length(pb_rows_sid) == 0) next
  mcol <- metab_col_for_pb[pb_rows_sid[1]]
  if (!is.na(mcol) && mcol %in% colnames(metab_tx)) {
    metab_view[, i] <- metab_tx[, mcol]
  }
}

# Replace Compound.ID rownames with metabolite names
if (!is.null(name_lookup)) {
  new_names <- name_lookup[rownames(metab_view)]
  # Fall back to Compound.ID where no name exists
  new_names[is.na(new_names)] <- rownames(metab_view)[is.na(new_names)]
  
  # Transliterate non-ASCII to ASCII equivalents
  new_names <- gsub("\u00b1", "+/-", new_names)
  new_names <- gsub("\u03b1|\u0391", "alpha-", new_names)
  new_names <- gsub("\u03b2|\u0392", "beta-", new_names)
  new_names <- gsub("\u03b3|\u0393", "gamma-", new_names)
  new_names <- gsub("\u03b4|\u0394", "delta-", new_names)
  new_names <- gsub("\u03c5|\u03a5", "upsilon-", new_names)
  # Catch any remaining non-ASCII
  new_names <- iconv(new_names, from = "UTF-8", to = "ASCII//TRANSLIT", sub = "")

  # Handle duplicates (append Compound.ID to disambiguate)
  dupes <- duplicated(new_names) | duplicated(new_names, fromLast = TRUE)
  new_names[dupes] <- paste0(new_names[dupes], " (", rownames(metab_view)[dupes], ")")
  rownames(metab_view) <- new_names
}


n_metab_obs <- sum(!is.na(metab_view))
n_metab_tot <- prod(dim(metab_view))
cat(sprintf("Metabolite view: %d features x %d samples\n",
            nrow(metab_view), ncol(metab_view)))
cat(sprintf("Observed entries: %d / %d (%.1f%%)\n",
            n_metab_obs, n_metab_tot, 100 * n_metab_obs / n_metab_tot))
```

## Create MOFA object

```{r}
#| label: create-mofa

# Combine all views into a named list
all_views <- c(gene_views, list(metabolites = metab_view))

cat(sprintf("Total views: %d\n", length(all_views)))
cat(sprintf("  Gene views: %s\n",
            paste(names(gene_views), collapse = ", ")))
cat("  Metabolite view: metabolites\n")

# Create the MOFA object
# MOFA2 expects a list of matrices (features x samples), one per view
mofa_obj <- create_mofa(all_views)

cat("\nMOFA object summary:\n")
print(mofa_obj)
```

## Set MOFA options

```{r}
#| label: mofa-options

# Data options
data_opts <- get_default_data_options(mofa_obj)
data_opts$scale_views <- TRUE 
cat("Data options:\n")
str(data_opts)

# Model options — start with 15 factors (MOFA will drop inactive ones)
model_opts <- get_default_model_options(mofa_obj)
model_opts$num_factors <- 5

cat("\nModel options:\n")
str(model_opts)

# Training options
train_opts <- get_default_training_options(mofa_obj)
train_opts$convergence_mode <- "slow"   # more thorough for a first look
train_opts$seed <- 42

cat("\nTraining options:\n")
str(train_opts)

# Apply options
mofa_obj <- prepare_mofa(
  mofa_obj,
  data_options    = data_opts,
  model_options   = model_opts,
  training_options = train_opts
)
```

## Train MOFA

```{r}
#| label: train
#| cache: true

mofa_trained <- run_mofa(mofa_obj, 
                         outfile = file.path(out_dir, "mofa_model.hdf5"),
                         use_basilisk = TRUE)

cat(sprintf("\nTraining complete. Active factors: %d\n",
            get_dimensions(mofa_trained)$K))
```
## Factor orthogonality

MOFA's ARD prior encourages sparsity but does not enforce orthogonality.
Pairwise Pearson correlations between factor values test whether factors
capture independent sources of variation (|r| > 0.3 warrants scrutiny).

```{r}
#| label: factor-orthogonality
#| fig-width: 6
#| fig-height: 5

factors_wide_orth <- get_factors(mofa_trained, as.data.frame = TRUE) %>%
  pivot_wider(names_from = factor, values_from = value)

factor_cols_orth <- grep("^Factor", colnames(factors_wide_orth), value = TRUE)
n_factors <- length(factor_cols_orth)

if (n_factors >= 2) {
  factor_cor <- cor(
    factors_wide_orth[, factor_cols_orth, drop = FALSE],
    use = "pairwise.complete.obs"
  )
  
  cor_long_orth <- as.data.frame(as.table(factor_cor)) %>%
    rename(Factor1 = Var1, Factor2 = Var2, r = Freq) %>%
    filter(as.integer(Factor1) < as.integer(Factor2))
  
  p_orth <- ggplot(as.data.frame(as.table(factor_cor)),
                   aes(x = Var1, y = Var2, fill = Freq)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", Freq)), size = 3) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         limits = c(-1, 1)) +
    labs(title = "Factor–factor correlations (orthogonality check)",
         x = NULL, y = NULL, fill = "r") +
    theme_minimal()
  
  print(p_orth)
  save_dual_format(p_orth, out_dir, "factor_orthogonality", width = 6, height = 5)
  
  # Flag non-orthogonal pairs
  flagged_orth <- cor_long_orth %>% filter(abs(r) > 0.3)
  if (nrow(flagged_orth) > 0) {
    cat("\nFactor pairs with |r| > 0.3:\n")
    print(as.data.frame(flagged_orth), row.names = FALSE)
  } else {
    cat("\nAll factor pairs have |r| <= 0.3 — factors appear orthogonal.\n")
  }
} else {
  cat("Only 1 active factor — orthogonality check not applicable.\n")
}
```

```{r}
#| label: feature-count-diagnostic
#| fig-width: 10
#| fig-height: 12

# For each view, count non-NA features per sample
feature_counts <- map_dfr(names(all_views), function(v) {
  mat <- all_views[[v]]
  tibble(
    view   = v,
    sample = colnames(mat),
    n_expressed = colSums(!is.na(mat) & mat != 0)
  )
})

# Merge with factor values
factors_long <- get_factors(mofa_trained, as.data.frame = TRUE)

diag_df <- factors_long %>%
  left_join(feature_counts, by = "sample") %>%
  filter(!is.na(value), !is.na(n_expressed))

# Correlations: factor x view
diag_cor <- diag_df %>%
  group_by(factor, view) %>%
  summarise(
    r    = cor(value, n_expressed, use = "pairwise.complete.obs"),
    pval = cor.test(value, n_expressed)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    sig  = ifelse(pval < 0.05, "*", ""),
    view_type = ifelse(view == "metabolites", "metabolites", "gene")
  )

# Heatmap
ggplot(diag_cor, aes(x = view, y = factor, fill = r)) +
  geom_tile() +
  geom_text(aes(label = paste0(sprintf("%.2f", r), sig)), size = 2.5) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       limits = c(-1, 1)) +
  labs(title = "Factor correlation with # expressed features per view",
       subtitle = "* = p < 0.05",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_featcount <- last_plot()
save_dual_format(p_featcount, out_dir, "feature_count_diagnostic",
                 width = 10, height = 12)

# Summary: which factors are flagged and where
flagged <- diag_cor %>% filter(abs(r) > 0.5 & pval < 0.05)
if (nrow(flagged) > 0) {
  cat("\nStrongly correlated factor-view pairs (|r| > 0.5, p < 0.05):\n")
  print(as.data.frame(flagged), row.names = FALSE)
} else {
  cat("\nNo factor-view pairs exceed |r| > 0.5 at p < 0.05.\n")
}
```

## Variance explained

```{r}
#| label: variance-explained
#| fig-width: 12
#| fig-height: 8

# Per view x factor variance explained
p_var <- plot_variance_explained(mofa_trained, plot_total = TRUE)

# The function returns a list of ggplots
if (is.list(p_var)) {
  for (i in seq_along(p_var)) {
    print(p_var[[i]])
    save_dual_format(p_var[[i]], out_dir,
                     paste0("variance_explained_", i),
                     width = 12, height = 8)
  }
} else {
  print(p_var)
  save_dual_format(p_var, out_dir, "variance_explained", width = 12, height = 8)
}
```

## Factor values

```{r}
#| label: factor-values
#| fig-width: 10
#| fig-height: 8

# Extract factor values and annotate with covariates
factors_df <- get_factors(mofa_trained, as.data.frame = TRUE) %>%
  left_join(sample_meta, by = c("sample" = "sample_id"))

# Plot first few factors coloured by GW and Sex
n_factors <- get_dimensions(mofa_trained)$K

p_list <- list()
for (f in seq_len(min(n_factors, 6))) {
  fac_name <- paste0("Factor", f)

  p_gw <- ggplot(factors_df %>% filter(factor == fac_name),
                 aes(x = GW, y = value)) +
    geom_point(aes(colour = Sex), size = 3, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "grey50", linewidth = 0.5) +
    labs(title = paste(fac_name, "vs GW"), y = "Factor value") +
    theme_minimal()

  p_list[[f]] <- p_gw
}

p_factors_gw <- plot_grid(plotlist = p_list, ncol = 2)
print(p_factors_gw)
save_dual_format(p_factors_gw, out_dir, "factors_vs_GW", width = 10, height = 8)
```

## Factor correlations with covariates

```{r}
#| label: factor-covariates
#| fig-width: 8
#| fig-height: 6

# Correlate factors with GW, Sex, batch1_frac
factors_wide <- get_factors(mofa_trained, as.data.frame = TRUE) %>%
  pivot_wider(names_from = factor, values_from = value) %>%
  left_join(sample_meta, by = c("sample" = "sample_id")) %>%
  mutate(Sex_num = as.numeric(factor(Sex)))

factor_cols <- grep("^Factor", colnames(factors_wide), value = TRUE)
cov_cols    <- c("GW", "Sex_num", "batch1_frac")

cor_mat <- cor(
  factors_wide[, factor_cols, drop = FALSE],
  factors_wide[, cov_cols, drop = FALSE],
  use = "pairwise.complete.obs"
)

# Simple heatmap
cor_long <- as.data.frame(as.table(cor_mat)) %>%
  rename(Factor = Var1, Covariate = Var2, r = Freq)

ggplot(cor_long, aes(x = Covariate, y = Factor, fill = r)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       limits = c(-1, 1)) +
  labs(title = "Factor–covariate correlations (Pearson r)") +
  theme_minimal()

p_fac_cov <- last_plot()
save_dual_format(p_fac_cov, out_dir, "factor_covariate_correlations",
                 width = 8, height = 6)
```

## Sample clustering in factor space

Pairwise scatter plots of factor values reveal how samples cluster in the
latent space learned by MOFA. Coloured by GW and shaped by Sex.

```{r}
#| label: factor-scatter
#| fig-width: 10
#| fig-height: 10

factors_scatter <- get_factors(mofa_trained, as.data.frame = TRUE) %>%
  pivot_wider(names_from = factor, values_from = value) %>%
  left_join(sample_meta, by = c("sample" = "sample_id"))

factor_cols_sc <- grep("^Factor", colnames(factors_scatter), value = TRUE)

if (length(factor_cols_sc) >= 2) {
  pairs_list <- combn(factor_cols_sc, 2, simplify = FALSE)
  
  pair_plots <- purrr::map(pairs_list, function(pair) {
    ggplot(factors_scatter, aes(x = .data[[pair[1]]], y = .data[[pair[2]]])) +
      geom_point(aes(colour = GW, shape = Sex), size = 3, alpha = 0.8) +
      scale_colour_viridis_c() +
      labs(x = pair[1], y = pair[2]) +
      theme_minimal(base_size = 10)
  })
  
  p_pairs <- plot_grid(plotlist = pair_plots,
                       ncol = min(3, length(pair_plots)))
  print(p_pairs)
  save_dual_format(p_pairs, out_dir, "factor_scatter_pairs",
                   width = min(15, 5 * min(3, length(pair_plots))),
                   height = 5 * ceiling(length(pair_plots) / 3))
} else {
  cat("Only 1 factor — pairwise scatter not applicable.\n")
}
```

```{r}
#| label: factors-by-sex
#| fig-width: 10
#| fig-height: 8

factors_df <- get_factors(mofa_trained, as.data.frame = TRUE) %>%
  left_join(sample_meta, by = c("sample" = "sample_id"))

ggplot(factors_df, aes(x = Sex, y = value, fill = Sex)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(aes(colour = Sex), width = 0.15, size = 2.5, alpha = 0.8) +
  facet_wrap(~ factor, scales = "free_y", nrow = 1) +
  labs(title = "Factor values by sex", y = "Factor value") +
  theme_minimal() +
  theme(legend.position = "bottom")

p_sex <- last_plot()
save_dual_format(p_sex, out_dir, "factors_by_sex", width = 10, height = 8)
```

## Top weights — which features drive each factor?

For each factor, the standard MOFA `plot_top_weights()` lollipop plot is
shown per view (each cell type and metabolites), gridded together.

```{r}
#| label: top-weights
#| fig-width: 16
#| fig-height: 14

n_top <- 10
view_names <- names(all_views)

for (f in seq_len(min(n_factors, 5))) {
  fac_name <- paste0("Factor", f)
  
  weight_plots <- list()
  for (v in view_names) {
    tryCatch({
      p <- plot_top_weights(mofa_trained,
                            view = v,
                            factor = f,
                            nfeatures = n_top) +
        ggtitle(v) +
        theme(plot.title = element_text(size = 10, face = "bold"))
      weight_plots[[v]] <- p
      
      # Save each view individually
      save_dual_format(p, out_dir,
                       paste0("weights_", fac_name, "_", gsub("[^A-Za-z0-9]", "_", v)),
                       width = 8, height = 6)
    }, error = function(e) {
      cat(sprintf("  Skipping %s for %s: %s\n", v, fac_name, e$message))
    })
  }
  
  if (length(weight_plots) > 0) {
    p_grid <- plot_grid(plotlist = weight_plots,
                        ncol = 3,
                        labels = NULL)
    title_gg <- cowplot::ggdraw() +
      cowplot::draw_label(paste(fac_name, "— top feature weights by view"),
                          fontface = "bold", size = 14)
    p_final <- plot_grid(title_gg, p_grid, ncol = 1, rel_heights = c(0.04, 1))
    
    print(p_final)
    save_dual_format(p_final, out_dir,
                     paste0("weights_by_view_", fac_name),
                     width = 16, height = 14)
  }
}
```

## Post-hoc hormone–factor associations

Correlate MOFA factors with targeted hormone levels (TT, E2, P4) post-hoc
rather than including hormones as a MOFA view. With only 3 features, a
hormone view would give the ARD prior nothing to regularize, and keeping
hormones external allows clean statements about factors independently
tracking hormone variation (Bost et al. 2021, Genome Biol 22:232).

```{r}
#| label: hormone-factor-correlation
#| fig-width: 8
#| fig-height: 6

# Pull hormone values from pb$metadata (already merged via build_metab_covariates)
hormone_meta <- pb$metadata %>%
  tibble::rownames_to_column("pb_id") %>%
  filter(sample_id %in% shared_samples) %>%
  group_by(sample_id) %>%
  summarise(
    E2 = first(E2),
    TT = first(TT),
    P4 = first(P4),
    .groups = "drop"
  )

factors_hormone <- get_factors(mofa_trained, as.data.frame = TRUE) %>%
  pivot_wider(names_from = factor, values_from = value) %>%
  left_join(hormone_meta, by = c("sample" = "sample_id"))

factor_cols_h <- grep("^Factor", colnames(factors_hormone), value = TRUE)
hormone_cols  <- c("E2", "TT", "P4")

# Spearman correlations + BH-adjusted p-values
hormone_cor_results <- expand.grid(
  Factor  = factor_cols_h,
  Hormone = hormone_cols,
  stringsAsFactors = FALSE
) %>%
  rowwise() %>%
  mutate(
    test = list(tryCatch(
      cor.test(factors_hormone[[Factor]], factors_hormone[[Hormone]],
               method = "spearman", exact = FALSE),
      error = function(e) NULL
    )),
    rho  = if (!is.null(test)) test$estimate else NA_real_,
    pval = if (!is.null(test)) test$p.value  else NA_real_
  ) %>%
  ungroup() %>%
  select(-test) %>%
  mutate(padj = p.adjust(pval, method = "BH"))

# Heatmap
p_hormone <- ggplot(hormone_cor_results,
                    aes(x = Hormone, y = Factor, fill = rho)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f\n%s",
                                rho,
                                ifelse(padj < 0.05, sprintf("q=%.2g", padj), "ns"))),
            size = 3) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       limits = c(-1, 1)) +
  labs(title = "Factor–hormone associations (Spearman rho, BH-adjusted)",
       x = NULL, y = NULL, fill = "rho") +
  theme_minimal()

print(p_hormone)
save_dual_format(p_hormone, out_dir, "factor_hormone_correlations",
                 width = 8, height = 6)

# Table
knitr::kable(hormone_cor_results %>%
               mutate(across(c(rho, pval, padj), ~ round(., 4))),
             caption = "Factor–hormone Spearman correlations")
```

## Save outputs

```{r}
#| label: save

# Factor values with covariates
factors_out <- get_factors(mofa_trained, as.data.frame = TRUE) %>%
  left_join(sample_meta, by = c("sample" = "sample_id"))

write_csv(factors_out, file.path(out_dir, "mofa_factor_values.csv"))

# Variance explained
var_exp <- get_variance_explained(mofa_trained)
saveRDS(var_exp, file.path(out_dir, "mofa_variance_explained.rds"))

# All weights (all views)
weights_all <- get_weights(mofa_trained, as.data.frame = TRUE)
write_csv(weights_all, file.path(out_dir, "mofa_all_weights.csv"))

# Metabolite weights only (convenience)
weights_metab <- weights_all %>% filter(view == "metabolites")
write_csv(weights_metab, file.path(out_dir, "mofa_metabolite_weights.csv"))

# Gene weights only (convenience)
weights_genes <- weights_all %>% filter(view != "metabolites")
write_csv(weights_genes, file.path(out_dir, "mofa_gene_weights.csv"))

# Hormone correlation results
write_csv(hormone_cor_results, file.path(out_dir, "mofa_hormone_correlations.csv"))

cat(sprintf("Outputs saved to %s\n", out_dir))
cat("  mofa_model.hdf5                 — trained model\n")
cat("  mofa_factor_values.csv          — factors + covariates\n")
cat("  mofa_variance_explained.rds     — R2 per view x factor\n")
cat("  mofa_all_weights.csv            — feature weights (all views)\n")
cat("  mofa_metabolite_weights.csv     — feature weights (metabolites)\n")
cat("  mofa_gene_weights.csv           — feature weights (genes)\n")
cat("  mofa_hormone_correlations.csv   — factor–hormone Spearman rho\n")
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
